### 2024-11-01 paper.Rmd
convert the paper back to rmd format

---
title: "Automated Assessment of Residual Plots with Computer Vision Models"
author:
  - name: Weihao Li
    affil: a
    email: weihao.li@monash.edu
  - name: Dianne Cook
    affil: a
    email: dicook@monash.edu
  - name: Emi Tanaka
    affil: a, b, c
    email: emi.tanaka@anu.edu.au
  - name: Susan VanderPlas
    affil: d
    email: susan.vanderplas@unl.edu
  - name: Klaus Ackermann
    affil: a
    email: klaus.ackermann@monash.edu 
affiliation:
  - num: a
    address: |
      Department of Econometrics and Business Statistics, Monash University, Clayton, VIC, Australia
  - num: b
    address: |
      Biological Data Science Institute, Australian National University, Acton, ACT, Australia
  - num: c
    address: |
      Research School of Finance, Actuarial Studies and Statistics, Australian National University, Acton, ACT, Australia
  - num: d
    address: |
      Department of Statistics, University of Nebraska, Lincoln, Nebraska, USA
bibliography: bibliography.bib
abstract: |
  Plotting the residuals is a recommended procedure to diagnose deviations from linear model assumptions, such as non-linearity, heteroscedasticity, and non-normality. The presence of structure in residual plots can be tested using the lineup protocol to do visual inference. There are a variety of conventional residual tests, but the lineup protocol, used as a statistical test, performs better for diagnostic purposes because it is less sensitive and applies more broadly to different types of departures. However, the lineup protocol relies on human judgment which limits its scalability. This work presents a solution by providing a computer vision model to automate the assessment of residual plots. It is trained to predict a distance measure that quantifies the disparity between the residual distribution of a fitted classical normal linear regression model and the reference distribution, based on Kullback-Leibler divergence. From extensive simulation studies, the computer vision model exhibits lower sensitivity than conventional tests but higher sensitivity than human visual tests. It is slightly less effective on non-linearity patterns. Several examples from classical papers and contemporary data illustrate the new procedures, highlighting its usefulness in automating the diagnostic process and supplementing existing methods.
keywords: |
  statistical graphics; data visualization; visual inference; computer vision; machine learning; hypothesis testing; reression analysis; cognitive perception; simulation; practical significance
header-includes: |
  \usepackage{lscape}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \usepackage{setspace}
  \doublespacing
output: rticles::tf_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  echo = FALSE,
  fig.width = 8,
  fig.height = 6,
  out.width = "100%", 
  fig.align = "center")
```

```{r}
# Visual inference models and p-value calculation
# remotes::install_github("TengMCing/visage")
library(tidyverse)
library(visage)
library(glue)

# To control the simulation in this file
set.seed(10086)
```

# Introduction {#sec-model-introduction}

Plotting residuals is commonly regarded as a standard practice in linear regression diagnostics [@belsley1980regression; @cook1982residuals]. This visual assessment plays a crucial role in identifying whether model assumptions, such as linearity, homoscedasticity, and normality, are reasonable. It also helps in understanding the goodness of fit and various unexpected characteristics of the model.

Generating a residual plot in most statistical software is often as straightforward as executing a line of code or clicking a button. However, accurately interpreting a residual plot can be challenging. A residual plot can exhibit various visual features, but it is crucial to recognize that some may arise from the characteristics of predictors and the natural stochastic variation of the observational unit, rather than indicating a violation of model assumptions [@li2024plot]. Consider Figure \ref{fig:false-finding} as an example, the residual plot displays a triangular left-pointing shape. The distinct difference in the spread of the residuals across the fitted values may result in the analyst suggesting that there may be heteroskedasticity, however, it is important to avoid over-interpreting this visual pattern. In this case, the fitted regression model is correctly specified, and the triangular shape is actually a result of the skewed distribution of the predictors, rather than indicating a flaw in the model.

The concept of visual inference, as proposed by @buja2009statistical, provides an inferential framework to assess whether residual plots indeed contain visual patterns inconsistent with the model assumptions. The fundamental idea involves testing whether the true residual plot visually differs significantly from null plots, where null plots are plotted with residuals generated from the residual rotation distribution [@langsrud2005rotation], which is a distribution consistent with the null hypothesis $H_0$ that the linear regression model is correctly specified. Typically, the visual test is accomplished through the lineup protocol, where the true residual plot is embedded within a lineup alongside several null plots. If the true residual plot can be distinguished from the lineup, it provides evidence for rejecting $H_0$.

The practice of delivering a residual plot as a lineup is generally regarded as a valuable approach. Beyond its application in residual diagnostics, the lineup protocol has been integrated into the analysis of diverse subjects. For instance, Loy and Hofmann [-@loy2013diagnostic; -@loy2014hlmdiag; -@loy2015you] illustrated its applicability in diagnosing hierarchical linear models. Additionally, @widen2016graphical and @fieberg2024using demonstrated its utility in geographical and ecology research respectively, while @krishnan2021hierarchical explored its effectiveness in forensic examinations.

A practical limitation of the lineup protocol lies in its reliance on human judgements [see @li2024plot about the practical limitations]. Unlike conventional statistical tests that can be performed computationally in statistical software, the lineup protocol requires human evaluation of images. This characteristic makes it less suitable for large-scale applications, given the associated high labour costs and time requirements. There is a substantial need to develop an approach to substitute these human judgement with an automated reading of data plots using machines.

The utilization of computers to interpret data plots has a rich history, with early efforts such as "Scagnostics" by @tukey1985computer, a set of numerical statistics that summarize features of scatter plots. @wilkinson2005graph expanded on this work, introducing scagnostics based on computable measures applied to planar proximity graphs. These measures, including, but not limited to, "Outlying", "Skinny", "Stringy", "Straight", "Monotonic", "Skewed", "Clumpy", and "Striated",  aimed to characterize outliers, shape, density, trend, coherence and other characteristics of the data. While this approach has been inspiring, there is a recognition [@buja2009statistical] that it may not capture all the necessary visual features that differentiate true residual plots from null plots. A more promising alternative entails enabling machines to learn the function for extracting visual features from residual plots. Essentially, this means empowering computers to discern the crucial visual features for residual diagnostics and determining the method to extract them.

Modern computer vision models are well-suited for addressing this challenge. They rely on deep neural networks with convolutional layers [@fukushima1982neocognitron]. These layers use small, sliding windows to scan the image, performing a dot product to extract local features and patterns. Numerous studies have demonstrated the efficacy of convolutional layers in addressing various vision tasks, including image recognition [@rawat2017deep]. Despite the widespread use of computer vision models in fields like computer-aided diagnosis [@lee2015image], pedestrian detection [@brunetti2018computer], and facial recognition [@emami2012facial], their application in reading data plots remains limited. While some studies have explored the use of computer vision models for tasks such as reading recurrence plots for time series regression [@ojeda2020multivariate], time series classification [@chu2019automatic; @hailesilassie2019financial; @hatami2018classification; @zhang2020encoding], anomaly detection [@chen2020convolutional], and pairwise causality analysis [@singh2017deep], the application of reading residual plots with computer vision models is a new field of study.

In this paper, we develop computer vision models and integrate them into the residual plots diagnostics workflow, addressing the need for an automated visual inference. The paper is structured as follows. Section \ref{sec-model-specifications} discusses various specifications of the computer vision models. Section \ref{sec-model-distance-between-residual-plots} defines the distance measure used to detect model violations, while Section \ref{sec-model-distance-estimation} explains how the computer vision models estimate this distance measure. Section \ref{sec-model-statistical-testing} covers the statistical tests based on the estimated distance, and Section \ref{sec-model-violations-index} introduces a Model Violations Index, which offers a quicker and more convenient assessment. Sections \ref{sec-model-data-generation}, \ref{sec-model-architecture}, and  \ref{sec-model-training} detail the data preparation, model architecture, and training process, respectively. The results are presented in Section \ref{sec-model-results}. Example dataset applications are discussed in Section \ref{sec-examples}. Finally, we conclude with a discussion of our findings and propose ideas for future research directions.

```{r false-finding, fig.pos="!h", fig.width = 4, fig.height = 4*4/5, fig.pos="!h", fig.cap="An example residual vs fitted values plot (red line indicates 0 corresponds to the x-intercept, i.e. $y=0$). The vertical spread of the data points varies with the fitted values. This often indicates the existence of heteroskedasticity, however, here the result is due to skewed distribution of the predictors rather than heteroskedasticity. The Breusch-Pagan test rejects this residual plot at 95\\% significance level ($p\\text{-value} = 0.046$)."}
set.seed(452)
ori_x <- rand_lognormal()
mod <- heter_model(b = 0, x = closed_form(~-ori_x))
ori_dat <- mod$gen(300)

ori_dat %>%
  VI_MODEL$plot(theme = theme_light(), remove_grid_line = TRUE) +
  xlab("Fitted values") +
  ylab("Residuals")
```

# Model Specifications {#sec-model-specifications}

There are various specifications of the computer vision model that can be used to assess residual plots. We discuss these specifications below focusing on two key components of the model formula: the input and the output format.

## Input Formats

Deep learning models are in general very sensitive to the input data. The quality and relevance of the input data greatly influence the model's capacity to generate insightful and meaningful results. There are several designs of the input format that can be considered.

A straightforward architecture of the input layer involves feeding a vector of residuals along with a vector of fitted values, essentially providing all the necessary information for creating a residuals vs fitted values plot. However, a drawback of this method is the dynamic input size, which changes based on the number of observations. For modern computer vision models implemented in mainstream software like TensorFlow [@abadi2016tensorflow], the input shape is typically fixed. One solution is to pad the input vectors with leading or trailing zeros when the input tensor expects longer vectors, but it may fail if the input vector surpasses the designed length.

Another strategy is to summarize the residuals and fitted values separately using histograms and utilize the counts as the input. By controlling the number of bins in the histograms, it becomes possible to provide fixed-length input vectors. Still, since histograms only capture the marginal distribution of residuals and fitted values respectively, they can not be used to differentiate visual patterns with same marginal distributions but different joint distributions.

Another architecture of the input layer involves using an image as input. The primary advantage of this design, as opposed to the vector format, is the availability of the existing and sophisticated image processing architectures developed over the years, such as the VGG16 architecture proposed in @simonyan2014very. These architectures can effectively capture and summarize spatial information from nearby pixels, which is less straightforward with vector input. While encoding data points as pixels might lead to some loss of detail, this approach still provides a more comprehensive representation than the histogram-based strategy because it captures the joint distribution of fitted values and residuals. The main considerations are the image resolution and the aesthetics of the residual plot. In general, a higher resolution provides more information to the model but comes with a trade-off of increased complexity and greater difficulty in training. As for the aesthetics of the residual plot, a practical solution is to consistently present residual plots in the same style to the model. This implies that the model can not accept arbitrary images as input but requires the use of the same pre-processing pipeline to convert residuals and fitted values into a standardized-style residual plot.

Providing multiple residual plots to the model, such as a pair of plots, a triplet or a lineup is also a possible option. @chopra2005learning have shown that computer vision models designed for image comparison can assess whether a pair of images are similar or dissimilar. Applied to our specific problem, we can define null plots of a fitted regression model to be similar to each other, while considering true residual plots to be distinct from null plots of any fitted regression model. A triplet constitutes a set of three images, denoted as $image_1$, $image_2$ and $image_3$. It is often used to predict whether $image_2$ or $image_3$ is more similar to $image_1$, proving particularly useful for establishing rankings between samples. For this setup, we can apply the same criteria to define similarity between images. However, it is important to note that these two approaches usually require additional considerations regarding the loss function and, at times, non-standard training processes due to shared weights between different convolutional blocks.

Presenting a lineup to a model aligns closely with the lineup protocol. However, as the number of residual plots in a lineup increases, the resolution of the input image grows rapidly, posing challenges in training the model. We experimented with this approach in a pilot study, but the performance of the trained model was sub-optimal.

Taking into account the implementation cost and the need for model interpretability, we used the single residual plot input format in this paper.

## Output Formats

Given that the input is a single residual plot represented as a fixed-resolution image, we can choose the output from the computer vision model to be either binary (classification) or numeric (regression).

The binary outcome can represent whether the input image is consistent with a null plot as determined by either (1) the data generating process or (2) the result of a visual test based on human judgement. Training a model following the latter option requires data from prior human subject experiments, presenting difficulties in controlling the quality of data due to variations in experimental settings across different studies. Additionally, some visual inference experiments are unrelated to linear regression models or residual plot diagnostics, resulting in a limited amount of available training data.

Alternatively, the output could be a meaningful and interpretable numerical measure useful for assessing residual plots, such as the strength of suspicious visual patterns reflecting the extent of model violations, or the difficulty index for identifying whether a residual plot has no issues. However, these numeric measures are often informally used in daily communication but are not typically formalized or rigorously defined. For the purpose of training a model, this numeric measure has to be quantifiable.

In this study, we chose to define and use a distance between a true residual plot and a theoretically "good" residual plot. This is further explained in Section \ref{sec-model-distance-between-residual-plots}. @vo2016localizing have also demonstrated that defining a proper distance between images can enhance the matching accuracy in image search compared to a binary outcome model.

## Auxiliary Information with Scagnostics

In Section \ref{sec-model-introduction}, we mention that scagnostics consist of a set of manually designed visual feature extraction functions. While our computer vision model will learn its own feature extraction function during training, leveraging additional information from scagnostics can enhance the model's predictive accuracy.

For each residual plot used as an input image, we calculated four scagnostics — "Monotonic", "Sparse", "Splines", and "Striped" – using the `cassowaryr` R package [@mason2022cassowaryr]. These computed measures, along with the number of observations from the fitted model, were provided as the second input for the computer vision model. We selected these scagnostics due to their reliability and efficiency, as other scagnostics occasionally caused R process crashes (approximately 5% of the time) during training data preparation, due to a bug in the `interp` R package [@Albrecht2023interp]. Although the package maintainer later fixed this bug at our request, the fix came too late to retrain the model, and additionally, their high computational costs make them unsuitable for rapid inference, which was a critical factor in our choice.

# Distance from a Theoretically "Good" Residual Plot  {#sec-model-distance-between-residual-plots}

To develop a computer vision model for assessing residual plots within the visual inference framework, it is important to precisely define a numerical measure of "difference" or "distance" between plots. This distance can take the form of a basic statistical operation on pixels, such as the sum of square differences, however, a pixel-to-pixel comparison makes little sense in comparing residual plots where the main interest would be structural patterns. Alternatively, it could involve established image similarity metrics like the Structural Similarity Index Measure [@wang2004image] which compares images by integrating three perception features of an image: contrast, luminance, and structure (related to average, standard deviation and correlation of pixel values over a window, respectively). These image similarity metrics are tailored for image comparison in vastly different tasks to evaluating data plots, where only essential plot elements require assessment [@chowdhury2018measuring]. We can alternatively define a notion of distance by integrating key plot elements (instead of key perception features like luminance, contrast, and structure), such as those captured by scagnostics mentioned in Section \ref{sec-model-introduction}, but the functional form still needs to be carefully refined to accurately reflect the extent of the violations.

In this section, we introduce a distance measure between a true residual plot and a theoretically 'good' residual plot. This measure quantifies the divergence between the residual distribution of a given fitted regression model and that of a correctly specified model. The computation assumes knowledge of the data generating processes for predictors and response variables. Since these processes are often unknown in practice, we will discuss a method to estimate this distance using a computer vision model in Section \ref{sec-model-distance-estimation}.

## Residual Distribution

For a classical normal linear regression model, $\boldsymbol{y} = \boldsymbol{X}\boldsymbol{\beta} + \boldsymbol{e}$, the residual $\hat{\boldsymbol{e}}$ are derived as the difference of the fitted values and observed values $\boldsymbol{y}$. Suppose the data generating process is known and the regression model is correctly specified, by the Frisch-Waugh-Lowell theorem [@frisch1933partial], residuals $\hat{\boldsymbol{e}}$ can also be treated as random variables and written as a linear transformation of the error $\boldsymbol{e}$ formulated as $\hat{\boldsymbol{e}} = \boldsymbol{R}\boldsymbol{e}$, where $\boldsymbol{R}=\boldsymbol{I}_n -\boldsymbol{X}(\boldsymbol{X}^\top\boldsymbol{X})^{-1}\boldsymbol{X}^\top$ is the residual operator, $\boldsymbol{I}_n$ is a $n$ by $n$ identity matrix, and $n$ is the number of observations.

One of the assumptions of the classical normal linear regression model is that the error $\boldsymbol{e}$ follows a multivariate normal distribution with zero mean and constant variance, i.e., $\boldsymbol{e} \sim N(\boldsymbol{0}_n,\sigma^2\boldsymbol{I}_n)$. It follows that the distribution of residuals $\hat{\boldsymbol{e}}$ can be characterized by a certain probability distribution, denoted as $Q$, which is transformed from the multivariate normal distribution. This reference distribution $Q$ summarizes what "good" residuals should follow given the design matrix $\boldsymbol{X}$ is known and fixed.

Suppose the design matrix $\boldsymbol{X}$ has linearly independent columns, the trace of the hat matrix $\boldsymbol{H} = \boldsymbol{X}(\boldsymbol{X}^\top\boldsymbol{X})^{-1}\boldsymbol{X}^\top$ will equal to the number of columns in $\boldsymbol{X}$ denoted as $k$. As a result, the rank of $\boldsymbol{R}$ is $n - k$, and $Q$ is a degenerate multivariate distribution. To capture the characteristics of $Q$, such as moments, we can simulate a large numbers of $\boldsymbol{\varepsilon}$ and transform it to $\boldsymbol{e}$ to get the empirical estimates. For simplicity, in this study, we replaced the variance-covariance matrix of residuals $\text{cov}(\boldsymbol{e}, \boldsymbol{e}) = \boldsymbol{R}\sigma^2\boldsymbol{R}^\top = \boldsymbol{R}\sigma^2$ with a full-rank diagonal matrix $\text{diag}(\boldsymbol{R}\sigma^2)$, where $\text{diag}(.)$ sets the non-diagonal entries of a matrix to zeros. The resulting distribution for $Q$ is $N(\boldsymbol{0}_n, \text{diag}(\boldsymbol{R}\sigma^2))$.

Distribution $Q$ is derived from the correctly specified model. However, if the model is misspecified, then the actual distribution of residuals denoted as $P$, will be different from $Q$. For example, if the data generating process contains variables correlated with any column of $\boldsymbol{X}$ but missing from $\boldsymbol{X}$, causing an omitted variable problem, $P$ will be different from $Q$ because the residual operator obtained from the fitted regression model will not be the same as $\boldsymbol{R}$. Besides, if the $\boldsymbol{\varepsilon}$ follows a non-normal distribution such as a multivariate lognormal distribution, $P$ will usually be skewed and has a long tail.

## Distance of $P$ from $Q$

Defining a proper distance between distributions is usually easier than defining a proper distance between data plots. Given the true residual distribution $Q$ and the reference residual distribution $P$, we used a distance measure based on Kullback-Leibler divergence [@kullback1951information] to quantify the difference between two distributions as

\begin{equation} \label{eq:kl-0}
D = \log\left(1 + D_{KL}\right),
\end{equation}

where $D_{KL}$ is defined as

\begin{equation} \label{eq:kl-1}
D_{KL} = \int_{\mathbb{R}^{n}}\log\frac{p(\boldsymbol{e})}{q(\boldsymbol{e})}p(\boldsymbol{e})d\boldsymbol{e},
\end{equation}

\noindent and $p(.)$ and $q(.)$ are the probability density functions for distribution $P$ and distribution $Q$, respectively.

This distance measure was first proposed in @li2024plot. It was mainly designed for measuring the effect size of non-linearity and heteroskedasticity in a residual plot. @li2024plot have derived that, for a classical normal linear regression model that omits necessary higher-order predictors $\boldsymbol{Z}$ and the corresponding parameter $\boldsymbol{\beta}_z$, and incorrectly assumes $\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n,\sigma^2\boldsymbol{I}_n)$ while in fact $\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n, \boldsymbol{V})$ where $\boldsymbol{V}$ is an arbitrary symmetric positive semi-definite matrix, $Q$ can be represented as $N(\boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z, \text{diag}(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}))$. Note that the variance-covariance matrix is replaced with the diagonal matrix to ensure it is a full-rank matrix.

Since both $P$ and $Q$ are adjusted to be multivariate normal distributions, Equation \ref{eq:kl-1} can be further expanded to

\begin{equation} \label{eq:kl-2}
D_{KL} = \frac{1}{2}\left(\log\frac{|\boldsymbol{W}|}{|\text{diag}(\boldsymbol{R}\sigma^2)|} - n + \text{tr}(\boldsymbol{W}^{-1}\text{diag}(\boldsymbol{R}\sigma^2)) + \boldsymbol{\mu}_z^\top\boldsymbol{W}^{-1}\boldsymbol{\mu}_z\right),
\end{equation}

\noindent where $\boldsymbol{\mu}_z = \boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z$, and $\boldsymbol{W} = \text{diag}(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R})$. The assumed error variance $\sigma^2$ is set to be $\text{tr}(\boldsymbol{V})/n$, which is the expectation of the estimated variance.

## Non-normal $P$

For non-normal error $\boldsymbol{\varepsilon}$, the true residual distribution $P$ is unlikely to be a multivariate normal distribution. Thus, Equation \ref{eq:kl-2} given in @li2024plot will not be applicable to models violating the normality assumption.

To evaluate the Kullback-Leibler divergence of non-normal $P$ from $Q$, the fallback is to solve Equation \ref{eq:kl-1} numerically. However, since $\boldsymbol{e}$ is a linear transformation of non-normal random variables, it is very common that the general form of $P$ is unknown, meaning that we can not easily compute $p(\boldsymbol{e})$ using a well-known probability density function. Additionally, even if $p(\boldsymbol{e})$ can be calculated for any $\boldsymbol{e} \in \mathbb{R}^n$, it will be very difficult to do numerical integration over the $n$-dimensional space, because $n$ could be potentially very large.

In order to approximate $D_{KL}$ in a practically computable manner, the elements of $\boldsymbol{e}$ are assumed to be independent of each other. This assumption solves both of the issues mentioned above. First, we no longer need to integrate over $n$ random variables. The result of Equation \ref{eq:kl-1} is now the sum of the Kullback-Leibler divergence evaluated for each individual residual due to the assumption of independence between observations. Second, it is not required to know the joint probability density $p(\boldsymbol{e})$ any more. Instead, the evaluation of Kullback-Leibler divergence for an individual residual relies on the knowledge of the marginal density $p_i(e_i)$, where $e_i$ is the $i$-th residual for $i = 1, ..., n$. This is much easier to approximate through simulation. It is also worth mentioning that this independence assumption generally will not hold if $\text{cov}(e_i, e_j) \neq 0$ for any $1 \leq i < j \leq n$, but its existence is essential for reducing the computational cost.

Given $\boldsymbol{X}$ and $\boldsymbol{\beta}$, the algorithm for approximating Equation \ref{eq:kl-1} starts from simulating $m$ sets of observed values $\boldsymbol{y}$ according to the data generating process. The observed values are stored in a matrix $\boldsymbol{A}$ with $n$ rows and $m$ columns, where each column of $\boldsymbol{A}$ is a set of observed values. Then, we can get $m$ sets of realized values of $\boldsymbol{e}$ stored in the matrix $\boldsymbol{B}$ by applying the residual operator $\boldsymbol{B} = \boldsymbol{R}\boldsymbol{A}$. Furthermore, kernel density estimation (KDE) with Gaussian kernel and optimal bandwidth selected by the Silverman's rule of thumb [@silverman2018density] is applied on each row of $\boldsymbol{B}$ to estimate $p_i(e_i)$ for $i = 1, ..., n$. The KDE computation can be done by the `density` function in R.

Since the Kullback-Leibler divergence can be viewed as the expectation of the log-likelihood ratio between distribution $P$ and distribution $Q$ evaluated on distribution $P$, we can reuse the simulated residuals in matrix $\boldsymbol{B}$ to estimate the expectation by the sample mean. With the independence assumption, for non-normal $P$, $D_{KL}$ can be approximated by

\begin{align*} \label{eq:kl-3}
D_{KL} &\approx \sum_{i = 1}^{n} \hat{D}_{KL}^{(i)}, \\
\hat{D}_{KL}^{(i)} &= \frac{1}{m}\sum_{j = 1}^{m} \log\frac{\hat{p}_i(B_{ij})}{q(B_{ij})},
\end{align*}

\noindent where $\hat{D}_{KL}^{(i)}$ is the estimator of the Kullback-Leibler divergence for an individual residual $e_i$, $\boldsymbol{B}_{ij}$ is the $i$-th row and $j$-th column entry of the matrix $\boldsymbol{B}$, $\hat{p}_i(.)$ is the kernel density estimator of $p_i(.)$, $q(.)$ is the normal density function with mean zero and an assumed variance estimated as $\hat{\sigma}^2 = \sum_{b \in vec(\boldsymbol{B})}(b - \sum_{b \in vec(\boldsymbol{B})} b/nm)^2/(nm - 1)$, and $vec(.)$ is the vectorization operator which turns a $n \times m$ matrix into a $nm \times 1$ column vector by stacking the columns of the matrix on top of each other.

# Distance Estimation {#sec-model-distance-estimation}

In the previous sections, we have defined a distance measure given in Equation \ref{eq:kl-0} for quantifying the difference between the true residual distribution $P$ and an ideal reference distribution $Q$. However, this distance measure can only be computed when the data generating process is known. In reality, we often have no knowledge about the data generating process, otherwise we do not need to do a residual diagnostic in the first place.

We use a computer vision model to estimate this distance measure for a residual plot. Let $D$ be the result of Equation \ref{eq:kl-0}, and our estimator $\hat{D}$ is formulated as

\begin{equation} \label{eq:d-approx}
\hat{D} = f_{CV}(V_{h \times w}(\boldsymbol{e}, \hat{\boldsymbol{y}})),
\end{equation}

\noindent where $V_{h \times w}(.)$ is a plotting function that saves a residuals vs fitted values plot with fixed aesthetic as an image with $h \times w$ pixels in three channels (RGB), $f_{CV}(.)$ is a computer vision model which takes an $h \times w$ image as input and predicts the distance in the domain $[0, +\infty)$.

With the estimated distance $\hat{D}$, we can compare the underlying distribution of the residuals to a theoretically "good" residual distribution. $\hat{D}$ can also be used as an index of the model violations indicating the strength of the visual signal embedded in the residual plot.

It is not expected that $\hat{D}$ will be equal to original distance $D$. This is largely because information contained in a single residual plot is limited and it may not be able to summarize all the important characteristics of the residual distribution. For a given residual distribution $P$, many different residual plots can be simulated, where many will share similar visual patterns, but some of them could be visually very different from the rest, especially for regression models with small $n$. This suggests the error of the estimation will vary depends on whether the input residual plot is representative or not.

# Statistical Testing {#sec-model-statistical-testing}

## Lineup Evaluation {#sec-model-lineup-evaluation}

Theoretically, the distance $D$ for a correctly specified model is $0$, because $P$ will be the same as $Q$. However, the computer vision model may not necessary predict $0$ for a null plot. Using Figure \ref{fig:false-finding} as an example, it contains a visual pattern which is an indication of heteroskedasticity. We would not expect the model to be able to magically tell if the suspicious pattern is caused by the skewed distribution of the fitted values or the existence of heteroskedasticity. Additionally, some null plots could have outliers or strong visual patterns due to randomness, and a reasonable model will try to summarize those information into the prediction, resulting in $\hat{D} > 0$.

This property is not an issue if $\hat{D} \gg 0$ for which the visual signal of the residual plot is very strong, and we usually do not need any further examination of the significance of the result. However, if the visual pattern is weak or moderate, having $\hat{D}$ will not be sufficient to determine if $H_0$ should be rejected.

To address this issue we can adhere to the paradigm of visual inference, by comparing the estimated distance $\hat{D}$ to the estimated distances for the null plots in a lineup. Specifically, if a lineup comprises 20 plots, the null hypothesis $H_0$ will be rejected if $\hat{D}$ exceeds the maximum estimated distance among the $m - 1$ null plots, denoted as $\max\limits_{1 \leq i \leq m-1} {\hat{D}_{null}^{(i)}}$, where $\hat{D}_{null}^{(i)}$ represents the estimated distance for the $i$-th null plot. This approach is equivalent to the typical lineup protocol requiring a 95% significance level, where $H_0$ is rejected if the data plot is identified as the most distinct plot by the sole observer. The estimated distance serves as a metric to quantify the difference between the data plot and the null plots, as intended.

<!-- For lineups consisting of more than 20 plots, the 95% significance level can be maintained if the number of plots is a multiple of 20. Specifically, for lineups comprising $20t$ plots, where $t$ is a positive integer, we reject $H_0$ if $\hat{D}$ exceeds 95% ${\hat{D}_{null}^{(i)}}$ for $i = 1, \ldots, 20t-1$. The $p$-value in this case is given by $\frac{1}{20t} + \frac{1}{20t}\sum_{i=1}^{20t-1} I\left(\hat{D}_{null}^{(i)} > \hat{D}\right)$, where $I(\cdot)$ is the indicator function. -->

Moreover, if the number of plots in a lineup, denoted by $m$, is sufficiently large, the empirical distribution of ${\hat{D}_{null}^{(i)}}$ can be viewed as an approximation of the null distribution of the estimated distance. Consequently, quantiles of the null distribution can be estimated using the sample quantiles, and these quantiles can be utilized for decision-making purposes. The details of the sample quantile computation can be found in @hyndman1996sample. For instance, if $\hat{D}$ is greater than or equal to the 95% sample quantile, denoted as $Q_{null}(0.95)$, we can conclude that the estimated distance for the true residual plot is significantly different from the estimated distance for null plots with a 95% significance level. Based on our experience, to obtain a stable estimate of the 95% quantile, the number of null plots, $n_{null}$, typically needs to be at least 100. However, if the null distribution exhibits a long tail, a larger number of null plots may be required. Alternatively, a $p$-value is the probability of observing a distance equally or greater than $\hat{D}$ under the null hypothesis $H_0$, and it can be estimated by $\frac{1}{m} + \frac{1}{m}\sum_{i=1}^{m-1}I\left(\hat{D}_{null}^{(i)} \geq \hat{D}\right)$.

To alleviate computation burden, a lattice of quantiles for $\hat{D}$ under $H_0$ with specified sample sizes can be precomputed. We can then map the $\hat{D}$ and sample size to the closet quantile and sample size in lattice to calculate the corresponding $p$-value. This approach lose precision in $p$-value calculation, however, significantly improves computational efficiency.

## Bootstrapping

Bootstrap is often employed in linear regression when conducting inference for estimated parameters [see @davison1997bootstrap and @efron1994introduction]. It is typically done by sampling individual cases with replacement and refitting the regression model. If the observed data accurately reflects the true distribution of the population, the bootstrapped estimates can be used to measure the variability of the parameter estimate without making strong distributional assumptions about the data generating process.

Similarly, bootstrap can be applied on the estimated distance $\hat{D}$. For each refitted model $M_{boot}^{(i)}$, there will be an associated residual plot $V_{boot}^{(i)}$ which can be fed into the computer vision model to obtain $\hat{D}_{boot}^{(i)}$, where $i = 1,...,n_{boot}$, and $n_{boot}$ is the number of bootstrapped samples. If we are interested in the variation of $\hat{D}$, we can use $\hat{D}_{boot}^{(i)}$ to estimate a confidence interval.

Alternatively, since each $M_{boot}^{(i)}$ has a set of estimated coefficients $\hat{\boldsymbol{\beta}}_{boot}^{(i)}$ and an estimated variance $\hat{\sigma^2}_{boot}^{(i)}$, a new approximated null distribution can be construed and the corresponding 95% sample quantile $Q_{boot}^{(i)}(0.95)$ can be computed. Then, if $\hat{D}_{boot}^{(i)} \geq Q_{boot}^{(i)}(0.95)$, $H_0$ will be rejected for $M_{boot}^{(i)}$. The ratio of rejected $M_{boot}^{(i)}$ among all the refitted models provides an indication of how often the assumed regression model are considered to be incorrect if the data can be obtained repetitively from the same data generating process. But this approach is computationally very expensive because it requires $n_{boot} \times n_{null}$ times of residual plot assessment. In practice, $Q_{null}(0.95)$ can be used to replace $Q_{boot}^{(i)}(0.95)$ in the computation.

# Model Violations Index {#sec-model-violations-index}

While statistical testing is a powerful tool for detecting model violations, it can become cumbersome and time-consuming when quick decisions are needed, particularly due to the need to evaluate numerous null plots. In practice, a more convenient and immediate method for assessing model performance is often required. This is where an index, such as the Model Violations Index (MVI), becomes valuable. It offers a straightforward way to quantify deviations from model assumptions, enabling rapid assessment and easier comparison across models.

The estimator $\hat{D}$ measures the difference between the true residual distribution and the reference residual distribution, a difference primarily arises from deviations in model assumptions. The magnitude of $D$ directly reflects the degree of these deviations, thus making $\hat{D}$ instrumental in forming a model violations index (MVI).

Note that if more observations are used for estimating the linear regression, the result of Equation \ref{eq:kl-1} will increase, as the integration will be performed with larger $n$. For a given data generating process, $D$ typically increases logarithmically with the number of observations. This behavior comes from the relationship $D = \text{log}(1 + D_{KL})$, where $D_{KL} = \sum_{i=1}^{n}D_{KL}^{(i)}$ under the assumption of independence.

Since $\hat{D}$ is an estimate of $D$, it is expected that a larger number of observations will also lead to a higher $\hat{D}$. However, this does not imply that $\hat{D}$ fails to accurately represent the extent of model violations. In fact, when examining residual plots with more observations, we often observe a stronger visual signal strength, as the underlying patterns are more likely to be revealed, except in cases of significant overlapping.

Therefore, the Model Violations Index (MVI) can be proposed as

\begin{equation} \label{eq:mvi}
\text{MVI} = C + \hat{D} - \log(n),
\end{equation}

\noindent where $C$ is a large enough constant keeping the result positive and the term $-\log(n)$ is used to offset the increase in $D$ due to sample size.

Figure \ref{fig:poly-heter-index} displays the residual plots for fitted models exhibiting varying degrees of non-linearity and heteroskedasticity. Each residual plot's MVI is computed using Equation \ref{eq:mvi} with $C = 10$. When $\text{MVI} > 8$, the visual patterns are notably strong and easily discernible by humans. In the range $6 < \text{MVI} < 8$, the visibility of the visual pattern diminishes as MVI decreases. Conversely, when $\text{MVI} < 6$, the visual pattern tends to become relatively faint and challenging to observe. Table \ref{tab:mvi} provides a summary of the MVI usage and it is applicable to other linear regression models.

```{r}
data.frame(degree = c("Strong", "Moderate", "Weak"),
           range = c("$\\text{MVI} > 8$", "$6 < \\text{MVI} < 8$", "$\\text{MVI} < 6$")) %>%
  kableExtra::kable(booktabs = TRUE,
                    label = "mvi",
                    caption = "Degree of model violations or the strength of the visual signals according to the Model Violations Index (MVI). The constant $C$ is set to be 10.",
                    escape = FALSE,
                    col.names = c("Degree of model violations", "Range ($C$ = 10)"),
                    align = "lc")
```

```{r}
set.seed(1495)
if (!file.exists(here::here("cached_data/poly_index.rds"))) {
  e_sigma_factor <- 3
  keras_mod <- autovi::get_keras_model("vss_phn_32")
  poly_dgp <- phn_model(j = 3, sigma = 0.1^e_sigma_factor)
  ori_dat <- poly_dgp$gen(300)

  vss <- c()
  actual_ss <- c()
  dat_combined <- tibble()

  for (e_sigma in seq(0.7, 1.8, length.out = 20)^e_sigma_factor) {
    poly_dgp <- phn_model(j = 3, sigma = e_sigma)
    dat <- poly_dgp$gen(300, computed = list(x1 = ori_dat$x1, e = ori_dat$e, k = ori_dat$k * e_sigma / 0.1^e_sigma_factor))
    mod <- lm(y ~ x1, data = dat)
    my_vi <- autovi::auto_vi(fitted_mod = mod, keras_mod = keras_mod)
    vss <- c(vss, my_vi$vss()$vss)
    actual_ss <- c(actual_ss, log(poly_dgp$sample_effect_size(dat) + 1))

    dat_combined <- bind_rows(dat_combined, mutate(dat, e_sigma = e_sigma))
  }

  saveRDS(dat_combined, here::here("cached_data/poly_index.rds"))
  saveRDS(vss, here::here("cached_data/poly_index_vss.rds"))
}

dat_combined <- readRDS(here::here("cached_data/poly_index.rds"))
vss <- readRDS(here::here("cached_data/poly_index_vss.rds"))

p1 <- dat_combined %>%
  autovi::AUTO_VI$plot_resid() +
  facet_wrap(~e_sigma, scales = "free", 
             labeller = function(...){ 
               tibble(nb = map(glue("'MVI' == {format(vss - log(300) + 10, digits = 3)}"), str2expression))
               },
             ncol = 5) +
  theme(strip.text = element_text(size = 10), plot.title = element_text(size = 10)) +
  ggtitle("(A) Non-linearity")
```

```{r}
set.seed(1495)
if (!file.exists(here::here("cached_data/heter_index.rds"))) {
  b_factor <- 2
  keras_mod <- autovi::get_keras_model("vss_phn_32")
  heter_dgp <- phn_model(include_z = FALSE, b = 0)
  ori_dat <- poly_dgp$gen(300)

  vss <- c()
  actual_ss <- c()
  dat_combined <- tibble()

  for (b in rev(seq(0.2, 4, length.out = 20))^b_factor) {
    heter_dgp <- phn_model(include_z = FALSE, b = b)
    dat <- heter_dgp$gen(300, computed = list(x1 = ori_dat$x1, e = ori_dat$e))
    mod <- lm(y ~ x1, data = dat)
    my_vi <- autovi::auto_vi(fitted_mod = mod, keras_mod = keras_mod)
    vss <- c(vss, my_vi$vss()$vss)
    actual_ss <- c(actual_ss, log(heter_dgp$sample_effect_size(dat) + 1))

    dat_combined <- bind_rows(dat_combined, mutate(dat, nb = -b))
  }

  saveRDS(dat_combined, here::here("cached_data/heter_index.rds"))
  saveRDS(vss, here::here("cached_data/heter_index_vss.rds"))
}

dat_combined <- readRDS(here::here("cached_data/heter_index.rds"))
vss <- readRDS(here::here("cached_data/heter_index_vss.rds"))

p2 <- dat_combined %>%
  autovi::AUTO_VI$plot_resid() +
  facet_wrap(~nb, scales = "free", 
             labeller = function(...){ 
               tibble(nb = map(glue("'MVI' == {format(vss - log(300) + 10, digits = 3)}"), str2expression))
               },
             ncol = 5) +
  theme(strip.text = element_text(size = 10), plot.title = element_text(size = 10)) +
  ggtitle("(B) Heteroskedasticity")
```

```{r poly-heter-index, fig.pos = "!h", fig.cap = "Residual plots generated from fitted models exhibiting varying degrees of (A) non-linearity and (B) heteroskedasticity violations. The model violations index (MVI) is displayed atop each residual plot. The non-linearity patterns are relatively strong for $MVI > 8$, and relatively weak for $MVI < 6$, while the heteroskedasticity patterns are relatively strong for $MVI > 8$, and relatively weak for $MVI < 6$.", fig.height = 10}
patchwork::wrap_plots(p1, p2, ncol = 1)
```

# Data Generation {#sec-model-data-generation}

## Simulation Scheme

While observational data is frequently employed in training models for real-world applications, the data generating process of observational data often remains unknown, making computation for our target variable $D$ unattainable. Consequently, the computer vision models developed in this study were trained using synthetic data, including 80,000 training images and 8,000 test images. This approach provided us with precise label annotations. Additionally, it ensured a large and diverse training dataset, as we had control over the data generating process, and the simulation of the training data was relatively cost-effective.

We have incorporated three types of residual departures of linear regression model in the training data, including non-linearity, heteroskedasticity and non-normality. All three departures can be summarized by the data generating process formulated as

\begin{align} \label{eq:data-sim}
\boldsymbol{y} &= \boldsymbol{1}_n + \boldsymbol{x}_1 + \beta_1\boldsymbol{x}_2 + \beta_2(\boldsymbol{z} + \beta_1\boldsymbol{w}) + \boldsymbol{k} \odot \boldsymbol{\varepsilon}, \\
\boldsymbol{z} &= \text{He}_j(g(\boldsymbol{x}_1, 2)), \\
\boldsymbol{w} &= \text{He}_j(g(\boldsymbol{x}_2, 2)), \\
\boldsymbol{k} &= \left[\boldsymbol{1}_n + b(2 - |a|)(\boldsymbol{x}_1 + \beta_1\boldsymbol{x}_2 - a\boldsymbol{1}_n)^{\circ2}\right]^{\circ1/2},
\end{align}

\noindent where $\boldsymbol{y}$, $\boldsymbol{x}_1$, $\boldsymbol{x}_2$, $\boldsymbol{z}$, $\boldsymbol{w}$, $\boldsymbol{k}$ and $\boldsymbol{\varepsilon}$ are vectors of size $n$, $\boldsymbol{1}_n$ is a vector of ones of size $n$, $\boldsymbol{x}_1$ and $\boldsymbol{x}_2$ are two independent predictors, $\text{He}_j(.)$ is the $j$th-order probabilist's Hermite polynomials [@hermite1864nouveau], $(.)^{\circ2}$ and $(.)^{\circ1/2}$ are Hadamard square and square root, $\odot$ is the Hadamard product, and $g(\boldsymbol{x}, k)$ is a scaling function to enforce the support of the random vector to be $[-k, k]^n$ defined as

$$g(\boldsymbol{x}, k) = 2k \cdot \frac{\boldsymbol{x} - x_{\min}\boldsymbol{1}_n}{x_{\max} - x_{\min}} - k\boldsymbol{1}_n,~for~k > 0,$$
\noindent where $x_{\min} = \underset{i \in \{ 1,...,n\}}{\min} x_i$, $x_{\max} = \underset{i \in \{ 1,...,n\}}{\max} x_i$ and $x_i$ is the $i$-th entry of $\boldsymbol{x}$.

```{r}
tibble(Factor = c("j", "a", "b", "$\\beta_1$", "$\\beta_2$", " $\\text{dist}_\\varepsilon$", "$\\text{dist}_{x1}$", "$\\text{dist}_{x2}$", "$\\sigma_{\\varepsilon}$", "$\\sigma_{X1}$", "$\\sigma_{X2}$", "n"),
       Domain = c("\\{2, 3, ..., 18\\}", "[-1, 1]", "[0, 100]", "\\{0, 1\\}", "\\{0, 1\\}", "\\{discrete, uniform, normal, lognormal\\}", "\\{discrete, uniform, normal, lognormal\\}", "\\{discrete, uniform, normal, lognormal\\}", "[0.0625, 9]", "[0.3, 0.6]", "[0.3, 0.6]", "[50, 500]")) %>%
  kableExtra::kable(escape = FALSE,
                    label = "factor",
                    caption = "Factors used in the data generating process for synthetic data simulation. Factor $j$ and $a$ controls the non-linearity shape and the heteroskedasticity shape respectively. Factor $b$, $\\sigma_\\varepsilon$ and $n$ control the signal strength. Factor $\\text{dist}_\\varepsilon$, $\\text{dist}_{x1}$ and $\\text{dist}_{x2}$ specifies the distribution of $\\varepsilon$, $X_1$ and $X_2$ respectively.",
                    booktabs = TRUE)
```

The residuals and fitted values of the fitted model were obtained by regressing $\boldsymbol{y}$ on $\boldsymbol{x}_1$. If $\beta_1 \neq 0$, $\boldsymbol{x}_2$ was also included in the design matrix. This data generation process was adapted from @li2024plot, where it was utilized to simulate residual plots exhibiting non-linearity and heteroskedasticity visual patterns for human subject experiments. A summary of the factors utilized in Equation \ref{eq:data-sim} is provided in Table \ref{tab:factor}.

In Equation \ref{eq:data-sim}, $\boldsymbol{z}$ and $\boldsymbol{w}$ represent higher-order terms of $\boldsymbol{x}_1$ and $\boldsymbol{x}_2$, respectively. If $\beta_2 \neq 0$, the regression model will encounter non-linearity issues. Parameter $j$ serves as a shape parameter that controls the number of tuning points in the non-linear pattern. Typically, higher values of $j$ lead to an increase in the number of tuning points, as illustrated in Figure \ref{fig:different-j}.

```{r different-j, fig.pos="!h", fig.cap = "Non-linearity forms generated for the synthetic data simulation. The 17 shapes are generated by varying the order of polynomial given by $j$ in $He_j(.)$."}

set.seed(10086)

# Data for shape 1
dat_shape_1 <- phn_model(j = 2, include_x2 = FALSE, sigma = 0.05)$gen(500) %>%
  mutate(j = 2)

# Generate data for shape 2, 3 and 4. Reuse x and e.
map_df(3:18, function(j) {
  phn_model(j = j, include_x2 = FALSE, sigma = 0.05)$
    gen(500, computed = select(dat_shape_1, x1, e)) %>%
  mutate(j = j)
}) %>%

  # Combined with data for shape 1
  bind_rows(dat_shape_1) %>%
  mutate(j = factor(j)) %>%
  VI_MODEL$plot(remove_axis = TRUE, remove_grid_line = TRUE, theme = theme_light()) +
  facet_wrap(~j, scales = "free", labeller = label_parsed, ncol = 5)
```

Additionally, scaling factor $\boldsymbol{k}$ directly affects the error distribution and it is correlated with $\boldsymbol{x}_1$ and $\boldsymbol{x}_2$. If $b \neq 0$ and $\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n, \sigma^2\boldsymbol{I}_n)$, the constant variance assumption will be violated. Parameter $a$ is a shape parameter controlling the location of the smallest variance in a residual plot as shown in Figure \ref{fig:different-a}.

```{r different-a, fig.width=8/5*3*2, fig.height=6/4*3*2, fig.pos="!h", fig.cap = 'Heteroskedasticity forms generated for the synthetic data simulation. Different shapes are controlled by the continuous factor $a$ between -1 and 1. For $a = -1$, the residual plot exhibits a "left-triangle" shape. And for $a = 1$, the residual plot exhibits a "right-triangle" shape. '}

set.seed(10085)

# Generate data for a = -1
dat_a_n1 <- phn_model(include_z = FALSE,
                      include_x2 = FALSE,
                      a = -1,
                      b = 100)$gen(500) %>%
  mutate(a = -1)

# Generate data for other a
map(c(-0.75, -0.5, -0.25, 0, 0.25, 0.5, 0.75, 1), function(a) {
  phn_model(include_z = FALSE,
            include_x2 = FALSE,
            a = a,
            b = 100)$gen(500) %>%
  mutate(a = a)
}) %>%

  # Combined with data for a = -1
  bind_rows(dat_a_n1) %>%
  mutate(a = factor(a)) %>%
  VI_MODEL$plot(remove_axis = TRUE, remove_grid_line = TRUE, theme = theme_light()) +
  facet_wrap(~a, scales = "free", ncol = 3) +
  xlab("Fitted values") +
  ylab("Residuals")
```

```{r different-e, fig.pos="!h", fig.cap = 'Non-normality forms generated for the synthetic data simulation. Four different error distributions including discrete, lognormal, normal and uniform are considered.'}

set.seed(10086)

# Data for shape 1
dat_shape_1 <- phn_model(include_z = FALSE, include_x2 = FALSE, e = rand_uniform(-1.4, 1.4))$gen(500) %>%
  mutate(e_dist = "uniform")

dat_shape_2 <- phn_model(include_z = FALSE, include_x2 = FALSE, sigma = 0.8)$gen(500) %>%
  mutate(e_dist = "normal")

dat_shape_3 <- phn_model(include_z = FALSE, include_x2 = FALSE, e = rand_lognormal(sigma = 0.6))$gen(500) %>%
  mutate(e_dist = "lognormal")

dat_shape_4 <- phn_model(include_z = FALSE, include_x2 = FALSE, e = rand_uniform_d(-1.4, 1.4, even = TRUE))$gen(500) %>%
  mutate(e_dist = "discrete")

# Generate data for shape 2, 3 and 4. Reuse x and e.
bind_rows(dat_shape_1, dat_shape_2, dat_shape_3, dat_shape_4) %>%
  VI_MODEL$plot(remove_axis = TRUE, remove_grid_line = TRUE, theme = theme_light()) +
  facet_wrap(~e_dist, scales = "free", labeller = label_parsed, ncol = 2)
```

Non-normality violations arise from specifying a non-normal distribution for $\boldsymbol{\varepsilon}$. In the synthetic data simulation, four distinct error distributions are considered, including discrete, uniform, normal, and lognormal distributions, as presented in Figure \ref{fig:different-e}. Each distribution imparts unique characteristics in the residual plot. The discrete error distribution introduces discrete clusters in residuals, while the lognormal distribution typically yields outliers. Uniform error distribution may result in residuals filling the entire space of the residual plot. All of these distributions exhibit visual distinctions from the normal error distribution.

```{r different-j-x2, fig.pos="!h", fig.cap = "Residual plots of multiple linear regression models with non-linearity issues. The 17 shapes are generated by varying the order of polynomial given by $j$ in $He_j(.)$. A second predictor $\\boldsymbol{x}_2$ is introduced to the regression model to create complex shapes."}

set.seed(10086)

# Data for shape 1
dat_shape_1 <- phn_model(j = 2, include_x2 = TRUE, sigma = 0.05)$gen(500) %>%
  mutate(j = 2)

# Generate data for shape 2, 3 and 4. Reuse x and e.
map_df(3:18, function(j) {
  phn_model(j = j, include_x2 = TRUE, sigma = 0.05)$
    gen(500, computed = select(dat_shape_1, x1, e)) %>%
  mutate(j = j)
}) %>%

  # Combined with data for shape 1
  bind_rows(dat_shape_1) %>%
  mutate(j = factor(j)) %>%
  VI_MODEL$plot(remove_axis = TRUE, remove_grid_line = TRUE, theme = theme_light()) +
  facet_wrap(~j, scales = "free", labeller = label_parsed, ncol = 5)
```

```{r different-j-heter, fig.pos="!h", fig.cap = 'Residual plots of models violating both the non-linearity and the heteroskedasticity assumptions. The 17 shapes are generated by varying the order of polynomial given by $j$ in $He_j(.)$, and the "left-triangle" shape is introduced by setting $a = -1$.'}

set.seed(10086)

# Data for shape 1
dat_shape_1 <- phn_model(j = 2, a = -1, b = 100, include_x2 = FALSE, sigma = 0.05)$gen(500) %>%
  mutate(j = 2)

# Generate data for shape 2, 3 and 4. Reuse x and e.
map_df(3:18, function(j) {
  phn_model(j = j, a = -1, b = 100, include_x2 = FALSE, sigma = 0.05)$
    gen(500, computed = select(dat_shape_1, x1, e)) %>%
  mutate(j = j)
}) %>%

  # Combined with data for shape 1
  bind_rows(dat_shape_1) %>%
  mutate(j = factor(j)) %>%
  VI_MODEL$plot(remove_axis = TRUE, remove_grid_line = TRUE, theme = theme_light()) +
  facet_wrap(~j, scales = "free", labeller = label_parsed, ncol = 5)
```

```{r different-e-heter, fig.pos="!h", fig.cap = 'Residual plots of models violating both the non-normality and the heteroskedasticity assumptions. The four shapes are generated by using four different error distributions including discrete, lognormal, normal and uniform, and the "left-triangle" shape is introduced by setting $a = -1$. '}

set.seed(10085)

# Data for shape 1
dat_shape_1 <- phn_model(a = -1, b = 100, include_z = FALSE, include_x2 = FALSE, e = rand_uniform(-1.4, 1.4))$gen(500) %>%
  mutate(e_dist = "uniform")

dat_shape_2 <- phn_model(a = -1, b = 100, include_z = FALSE, include_x2 = FALSE, sigma = 0.8)$gen(500) %>%
  mutate(e_dist = "normal")

dat_shape_3 <- phn_model(a = -1, b = 100, include_z = FALSE, include_x2 = FALSE, e = rand_lognormal(sigma = 0.6))$gen(500) %>%
  mutate(e_dist = "lognormal")

dat_shape_4 <- phn_model(a = -1, b = 100, include_z = FALSE, include_x2 = FALSE, e = rand_uniform_d(-1.4, 1.4, even = TRUE))$gen(500) %>%
  mutate(e_dist = "discrete")

# Generate data for shape 2, 3 and 4. Reuse x and e.
bind_rows(dat_shape_1, dat_shape_2, dat_shape_3, dat_shape_4) %>%
  VI_MODEL$plot(remove_axis = TRUE, remove_grid_line = TRUE, theme = theme_light()) +
  facet_wrap(~e_dist, scales = "free", labeller = label_parsed, ncol = 2)
```

Equation \ref{eq:data-sim} accommodates the incorporation of the second predictor $\boldsymbol{x}_2$. Introducing it into the data generation process by setting $\beta_1 = 1$ significantly enhances the complexity of the shapes, as illustrated in Figure \ref{fig:different-j-x2}. In comparison to Figure \ref{fig:different-j}, Figure \ref{fig:different-j-x2} demonstrates that the non-linear shape resembles a surface rather than a single curve. This augmentation can facilitate the computer vision model in learning visual patterns from residual plots of the multiple linear regression model.

In real-world analysis, it is not uncommon to encounter instances where multiple model violations coexist. In such cases, the residual plots often exhibit a mixed pattern of visual anomalies corresponding to different types of model violations. Figure \ref{fig:different-j-heter} and Figure \ref{fig:different-e-heter} show the visual patterns of models with multiple model violations.

## Balanced Dataset

To train a robust computer vision model, we deliberately controlled the distribution of the target variable $D$ in the training data. We ensured that it followed a uniform distribution between $0$ and $7$. This was achieved by organizing $50$ buckets, each exclusively accepting training samples with $D$ falling within the range $[7(i - 1)/49, 7i/49)$ for $i < 50$, where $i$ represents the index of the $i$-th bucket. For the $50$-th bucket, any training samples with $D \geq 7$ were accepted.

With 80,000 training images prepared, each bucket accommodated a maximum of $80000/ 50 = 1600$ training samples. The simulator iteratively sampled parameter values from the parameter space, generated residuals and fitted values using the data generation process, computed the distance, and checked if the sample fitted within the corresponding bucket. This process continued until all buckets were filled.

Similarly, we adopted the same methodology to prepare 8,000 test images for performance evaluation and model diagnostics.

# Model Architecture {#sec-model-architecture}

```{r cnn-diag, fig.cap = "Diagram of the architecture of the optimized computer vision model. Numbers at the bottom of each box show the shape of the output of each layer. The band of each box drawn in a darker color indicates the use of the rectified linear unit activation function.  Yellow boxes are 2D convolutional layers, orange boxes are pooling layers, the grey box is the concatenation layer, and the purple boxes are dense layers.", out.width = "100%"}
magick::image_read_pdf("figures/cnn.pdf")
```

The architecture of the computer vision model is adapted from the well-established VGG16 architecture [@simonyan2014very]. While more recent architectures like ResNet [@he2016deep] and DenseNet[@huang2017densely], have achieved even greater performance, VGG16 remains a solid choice for many applications due to its simplicity and effectiveness. Our decision to use VGG16 aligns with our goal of starting with a proven and straightforward model. Figure \ref{fig:cnn-diag} provides a diagram of the architecture. More details about the neural network layers used in this study are provided in Appendix A.

The model begins with an input layer of shape $n \times h \times w \times 3$, capable of handling $n$ RGB images. This is followed by a grayscale conversion layer utilizing the luma formula under the Rec. 601 standard [@series2011studio], which converts the color image to grayscale. Grayscale suffices for our task since data points are plotted in black. We experiment with three combinations of $h$ and $w$: $32 \times 32$, $64 \times 64$, and $128 \times 128$, aiming to achieve sufficiently high image resolution for the problem at hand.

The processed image is used as the input for the first convolutional block. The model comprises at most five consecutive convolutional blocks, mirroring the original VGG16 architecture. Within each block, there are two 2D convolutional layers followed by two activation layers, respectively. Subsequently, a 2D max-pooling layer follows the second activation layer. The 2D convolutional layer convolves the input with a fixed number of $3 \times 3$ convolution filters, while the 2D max-pooling layer downsamples the input along its spatial dimensions by taking the maximum value over a $2 \times 2$ window for each channel of the input. The activation layer employs the rectified linear unit (ReLU) activation function, a standard practice in deep learning, which introduces a non-linear transformation of the output of the 2D convolutional layer. Additionally, to regularize training, a batch normalization layer is added after each 2D convolutional layer and before the activation layer. Finally, a dropout layer is appended at the end of each convolutional block to randomly set some inputs to zero during training, further aiding in regularization.

The output of the last convolutional block is summarized by either a global max pooling layer or a global average pooling layer, resulting in a two-dimensional tensor. To leverage the information contained in scagnostics, this tensor is concatenated with an additional $n \times 5$ tensor, which contains the "Monotonic", "Sparse", "Splines", and "Striped" measures, along with the number of observations for $n$ residual plots.

The concatenated tensor is then fed into the final prediction block. This block consists of two fully-connected layers. The first layer contains at least $128$ units, followed by a dropout layer. Occasionally, a batch normalization layer is inserted between the fully-connected layer and the dropout layer for regularization purposes. The second fully-connected layer consists of only one unit, serving as the output of the model.

The model weights $\boldsymbol{\theta}$ were randomly initialized and they were optimized by the Adam optimizer [@kingma2014adam] with the mean square error loss function

$$\hat{\boldsymbol{\theta}} = \underset{\boldsymbol{\theta}}{\text{arg min}}\frac{1}{n_{\text{train}}}\sum_{i=1}^{n_{\text{train}}}(D_i - f_{\boldsymbol{\theta}}(V_i, S_i))^2,$$

\noindent where $n_{\text{train}}$ is the number of training samples, $V_i$ is the $i$-th residual plot and $S_i$ is the additional information about the $i$-th residual plot including four scagnostics and the number of observations.

# Model Training {#sec-model-training}

To achieve a near-optimal deep learning model, hyperparameters like the learning rate often need to be fine-tuned using a tuner. In our study, we utilized the Bayesian optimization tuner from the `KerasTuner` Python library [@omalley2019kerastuner] for this purpose. A comprehensive list of hyperparameters is provided in Table \ref{tab:hyperparameter}.

The number of base filters determines the number of filters for the first 2D convolutional layer. In the VGG16 architecture, the number of filters for the 2D convolutional layer in a block is typically twice the number in the previous block, except for the last block, which maintains the same number of convolution filters as the previous one. This hyperparameter aids in controlling the complexity of the computer vision model. A higher number of base filters results in more trainable parameters. Likewise, the number of units for the fully-connected layer determines the complexity of the final prediction block. Increasing the number of units enhances model complexity, resulting in more trainable parameters.

The dropout rate and batch normalization are flexible hyperparameters that work in conjunction with other parameters to facilitate smooth training. A higher dropout rate is necessary when the model tends to overfit the training dataset by learning too much noise [@srivastava2014dropout]. Conversely, a lower dropout rate is preferred when the model is complex and challenging to converge. Batch normalization, on the other hand, addresses the internal covariate shift problem arising from the randomness in weight initialization and input data [@goodfellow2016deep]. It helps stabilize and accelerate the training process by normalizing the activations of each layer.

Additionally, incorporating additional inputs such as scagnostics and the number of observations can potentially enhance prediction accuracy. Therefore, we allow the tuner to determine whether these inputs were necessary for optimal model performance.

The learning rate is a crucial hyperparameter, as it dictates the step size of the optimization algorithm. A high learning rate can help the model avoid local minima but risks overshooting and missing the global minimum. Conversely, a low learning rate smoothens the training process but makes the convergence time longer and increases the likelihood of getting trapped in local minima.

Our model was trained on the MASSIVE M3 high-performance computing platform [@goscinski2014multi], using TensorFlow [@abadi2016tensorflow] and Keras [@chollet2015keras]. During training, 80% of the training data was utilized for actual training, while the remaining 20% was used as validation data. The Bayesian optimization tuner conducted 100 trials to identify the best hyperparameter values based on validation root mean square error. The tuner then restored the best epoch of the best model from the trials. Additionally, we applied early stopping, terminating the training process if the validation root mean square error fails to improve for 50 epochs. The maximum allowed epochs was set at 2,000, although no models reached this threshold.

```{r}
data.frame(`Hyperparameter` = c("Number of base filters", "Dropout rate for convolutional blocks", "Batch normalization for convolutional blocks", "Type of global pooling", "Ignore additional inputs", "Number of units for the fully-connected layer", "Batch normalization for the fully-connected layer", "Dropout rate for the fully-connected layer", "Learning rate"),
           Domain = c("\\{4, 8, 16, 32, 64\\}", "[0.1, 0.6]", "\\{false, true\\}", "\\{max, average\\}", "\\{false, true\\}", "\\{128, 256, 512, 1024, 2048\\}", "\\{false, true\\}", "[0.1, 0.6]", "[$10^{-8}$, $10^{-1}$]")) %>%
  kableExtra::kable(booktabs = TRUE, 
                    label = "hyperparameter",
                    caption = "Name of hyperparameters and their correspoding domain for the computer vision model.",
                    escape = FALSE)
```

Based on the tuning process described above, the optimized hyperparameter values are presented in Table \ref{tab:best-hyperparameter}. It was observable that a minimum of $32$ base filters was necessary, with the preferable choice being $64$ base filters for both the $64 \times 64$ and $128 \times 128$ models, mirroring the original VGG16 architecture. The optimized dropout rate for convolutional blocks hovered around $0.4$, and incorporating batch normalization for convolutional blocks proved beneficial for performance.

All optimized models chose to retain the additional inputs, contributing to the reduction of validation error. The number of units required for the fully-connected layer was $256$, a relatively modest number compared to the VGG16 classifier, suggesting that the problem at hand was less complex. The optimized learning rates were higher for models with higher resolution input, likely because models with more parameters are more prone to getting stuck in local minima, requiring a higher learning rate.

```{r}
data.frame(`Hyperparameter` = c("Number of base filters", "Dropout rate for convolutional blocks", "Batch normalization for convolutional blocks", "Type of global pooling", "Ignore additional inputs", "Number of units for the fully-connected layer", "Batch normalization for the fully-connected layer", "Dropout rate for the fully-connected layer", "Learning rate"),
           `32` = c("32", "0.4", "true", "max", "false", "256", "false", "0.2", "0.0003"),
           `64` = c("64", "0.3", "true", "average", "false", "256", "true", "0.4", "0.0006"),
           `128` = c("64", "0.4", "true", "average", "false", "256", "true", "0.1", "0.0052")) %>%
  kableExtra::kable(booktabs = TRUE, 
                    label = "best-hyperparameter",
                    caption = "Hyperparameters values for the optimized computer vision models with different input sizes.",
                    col.names = c("Hyperparameter", "$32 \\times 32$", "$64 \\times 64$", "$128 \\times 128$"),
                    escape = FALSE) %>%
  kableExtra::kable_styling(latex_options = c("scale_down"))
```

# Results {#sec-model-results}

## Model Performance

The test performance for the optimized models with three different input sizes are summarized in Table \ref{tab:performance}. Among these models, the $32 \times 32$ model consistently exhibited the best test performance. The mean absolute error of the $32 \times 32$ model indicated that the difference between $\hat{D}$ and $D$ was approximately $0.43$ on the test set, a negligible deviation considering the normal range of $D$ typically falls between $0$ and $7$. The high $R^2$ values also suggested that the predictions were largely linearly correlated with the target.

Figure \ref{fig:model-performance} presents a hexagonal heatmap for $D - \hat{D}$ versus $D$. The brown smoothing curves, fitted by generalized additive models [@hastie2017generalized], demonstrate that all the optimized models perform admirably on the test sets when $1.5 < D < 6$, where no structural issues are noticeable. However, over-predictions occurred when $D < 1.5$, while under-predictions occurred predominantly when $\hat{D} > 6$.

For input images representing null plots where $D = 0$, it was expected that the models will over-predict the distance, as explained in Section \ref{sec-model-lineup-evaluation}. However, it can not explain the under-prediction issue. Therefore, we analysed the relationship between residuals and all the factors involved in the data generating process. We found that most issues actually arose from non-linearity problems and the presence of a second predictor in the regression model as illustrated in Figure \ref{fig:over-under}. When the variance for the error distribution was small, the optimized model tended to under-predict the distance. Conversely, when the error distribution had a large variance, the model tended to over-predict the distance.

Since most of the deviation stemmed from the presence of non-linearity violations, to further investigate this, we split the test set based on violation types and re-evaluated the performance, as detailed in Table \ref{tab:performance-sub}. It was evident that metrics for null plots were notably worse compared to other categories. Furthermore, residual plots solely exhibiting non-normality issues were the easiest to predict, with very low test root mean square error (RMSE) at around $0.3$. Residual plots with non-linearity issues were more challenging to assess than those with heteroskedasticity or non-normality issues. When multiple violations were introduced to a residual plot, the performance metrics typically lay between the metrics for each individual violation.

Based on the model performance metrics, we chose to use the best-performing model evaluated on the test set, namely the $32 \times 32$ model, for the subsequent analysis.

```{r}
test_pred <- readr::read_csv(here::here("data/test_pred.csv"))
train_pred <- readr::read_csv(here::here("data/train_pred.csv"))
meta <- readr::read_csv(here::here("data/meta.csv"))

test_summary <- test_pred %>% 
  left_join(meta) %>%
  group_by(res) %>%
  summarise(RMSE = yardstick::rmse_vec(effect_size, vss),
            R2 = yardstick::rsq_vec(effect_size, vss),
            MAE = yardstick::mae_vec(effect_size, vss),
            HUBER = yardstick::huber_loss_vec(effect_size, vss)) %>%
  mutate(type = "test")

train_summary <- train_pred %>% 
  left_join(meta) %>%
  group_by(res) %>%
  summarise(RMSE = yardstick::rmse_vec(effect_size, vss),
            R2 = yardstick::rsq_vec(effect_size, vss),
            MAE = yardstick::mae_vec(effect_size, vss),
            HUBER = yardstick::huber_loss_vec(effect_size, vss)) %>%
  mutate(type = "train")
```

```{r}
test_summary %>%
  select(res, RMSE, R2, MAE, HUBER) %>%
  mutate(res = glue::glue("${res} \\times {res}$")) %>%
  mutate(across(RMSE:MAE, ~format(.x, digits = 3))) %>%
  mutate(across(HUBER, ~format(.x, digits = 2))) %>%
  kableExtra::kable(booktabs = TRUE, 
                    label = "performance",
                    caption = "The test performance of three optimized models with different input sizes.",
                    escape = FALSE,
                    align = "lrrrr",
                    col.names = c("", "RMSE", "$R^2$", "MAE", "Huber loss"))
```

```{r model-performance, fig.pos="!h", fig.cap = "Hexagonal heatmap for difference in $D$ and $\\hat{D}$ vs $D$ on test data for three optimized models with different input sizes. The brown lines are smoothing curves produced by fitting generalized additive models. The area over the zero line in light yellow indicates under-prediction, and the area under the zero line in light green indicates over-prediction.", dev = 'png', dpi = 300}
model_pred <- train_pred %>% 
  left_join(meta) %>%
  mutate(type = "train") %>%
  bind_rows(test_pred %>% 
              left_join(meta) %>%
              mutate(type = "test"))

model_pred %>%
  filter(type == "test") %>%
  ggplot() +
  geom_hline(yintercept = 0, alpha = 0.5) +
  annotate("rect", 
           ymin = min(model_pred$effect_size - model_pred$vss),
           ymax = 0,
           xmin = min(model_pred$effect_size),
           xmax = max(model_pred$effect_size),
            fill = "#40B0A6",
            alpha = 0.3) +
  geom_point(data = NULL, aes(5, 0, col = "over-prediction"), shape = 15) +
  annotate("rect", 
           ymin = 0,
           ymax = max(model_pred$effect_size - model_pred$vss),
           xmin = min(model_pred$effect_size),
           xmax = max(model_pred$effect_size),
           fill = "#E1BE6A",
           alpha = 0.3) +
  geom_point(data = NULL, aes(5, 0, col = "under-prediction"), shape = 15) +
  geom_hex(aes(effect_size, effect_size - vss), bins = 20) +
  geom_smooth(aes(effect_size, effect_size - vss), se = FALSE, col = "#994F00") +
  facet_wrap(~ res) +
  coord_fixed() +
  ylab(expression(D - hat(D))) +
  xlab(expression(D)) +
  theme_light() +
  scale_fill_continuous(limits = c(1, 6000), trans = "log10", low = "#56B1F7", high = "#132B43") +
  scale_color_manual(values = c("under-prediction" = scales::alpha("#E1BE6A", 0.3),
                                "over-prediction" = scales::alpha("#40B0A6", 0.3)),
                     breaks = c("under-prediction", "over-prediction")) +
  guides(col = guide_legend(title = NULL, override.aes = list(size = 8), order = 1))
```

```{r over-under, fig.pos="!h", fig.cap = "Scatter plots for difference in $D$ and $\\hat{D}$ vs $\\sigma$ on test data for the $32 \\times 32$ optimized model. The data is grouped by whether the regression has only non-linearity violation, and whether it includes a second predictor in the regression formula. The brown lines are smoothing curves produced by fitting generalized additive models. The area over the zero line in light yellow indicates under-prediction, and the area under the zero line in light green indicates over-prediction.", dev = 'png', dpi = 300}
model_pred %>%
  filter(res == 32L) %>%
  filter(type == "test") %>%
  mutate(only_z = include_z & !include_heter & !include_non_normal) %>%
  rename(`Only has non-linearity violation` = only_z, `Has second predictor` = include_x2) %>%
  ggplot() +
  geom_hline(yintercept = 0, alpha = 0.5) +
  annotate("rect", 
           ymin = min(model_pred$effect_size[model_pred$type == "test"] - model_pred$vss[model_pred$type == "test"]),
                ymax = 0,
                xmin = min(model_pred$e_sigma),
                xmax = max(model_pred$e_sigma),
            fill = "#40B0A6",
            alpha = 0.3) +
  annotate("rect", 
           ymin = 0,
                ymax = max(model_pred$effect_size[model_pred$type == "test"] - model_pred$vss[model_pred$type == "test"]),
                xmin = min(model_pred$e_sigma),
                xmax = max(model_pred$e_sigma),
            fill = "#E1BE6A",
            alpha = 0.3) +
  geom_point(aes(e_sigma, effect_size - vss), alpha = 0.2) +
  geom_smooth(aes(e_sigma, effect_size - vss), se = FALSE, col = "#994F00") +
  facet_grid(`Has second predictor` ~ `Only has non-linearity violation`,
             labeller = label_both) +
  theme_light() +
  ylab(expression(D - hat(D))) +
  xlab(expression(sigma))
```

```{r}
model_pred %>%
  filter(res == 32L) %>%
  group_by(include_non_normal, include_heter, include_z) %>%
  summarise(train_n = sum(type == "train"),
            train_RMSE = yardstick::rmse_vec(effect_size[type == "train"], vss[type == "train"]),
            test_n = sum(type == "test"),
            test_RMSE = yardstick::rmse_vec(effect_size[type == "test"], vss[type == "test"])) %>%
  mutate(violations = ifelse(include_z, "non-linearity", "null")) %>%
  mutate(violations = ifelse(include_heter, glue("{violations} + heteroskedasticity"), violations)) %>%
  mutate(violations = ifelse(include_non_normal, glue("{violations} + non-normality"), violations)) %>%
  mutate(violations = gsub("null \\+ ", "", violations)) %>%
  ungroup() %>%
  select(violations, train_n, train_RMSE, test_n, test_RMSE) %>%
  mutate(train_n = format(train_n)) %>%
  mutate(test_n = format(test_n)) %>%
  mutate(across(c(train_RMSE, test_RMSE), ~format(.x, digits = 3))) %>%
  mutate(violations = ifelse(violations == "null", "no violations", violations)) %>%
  select(-train_n, -train_RMSE) %>%
  kableExtra::kable(booktabs = TRUE,
                    label = "performance-sub",
                    caption = "The training and test performance of the $32 \\times 32$ model presented with different model violations.",
                    escape = FALSE,
                    linesep = "", 
                    align = "lrr",
                    col.names = c("Violations", "\\#samples", "RMSE"))
```

<!-- - check if j has an impact on residuals more carefully -->
<!-- - the RMSE has very little to do with j if x2 is not included. And its impact is not as large as sigma. -->

## Comparison with Human Visual Inference and Conventional Tests

### Overview of the Human Subject Experiment

In order to check the validity of the proposed computer vision model, residual plots presented in the human subject experiment conducted by @li2024plot will be assessed.

<!-- The experiment revealed that conventional tests are more sensitive to weak departures from model assumptions than visual test, and they often reject the null hypothesis when departures are not visibly different from null residual plots. -->

This study has collected 7,974 human responses to 1,152 lineups. Each lineup contains one randomly placed true residual plot and 19 null plots. Among the 1,152 lineups, 24 are attention check lineups in which the visual patterns are designed to be extremely obvious and very different from the corresponding to null plots, 36 are null lineups where all the lineups consist of only null plots, 279 are lineups with uniform predictor distribution evaluated by 11 participants, and the remaining 813 are lineups with discrete, skewed or normal predictor distribution evaluated by 5 participants. Attention check lineups and null lineups will not be assessed in the following analysis.

In @li2024plot, the residual plots are simulated from a data generating process which is a special case of Equation \ref{eq:data-sim}. The main characteristic is the model violations are introduced separately, meaning non-linearity and heteroskedasticity will not co-exist in one lineup but assigned uniformly to all lineups. Additionally, non-normality and multiple predictors are not considered in the experimental design.

### Model Performance on the Human-evaluated Data

```{r rd-human}
vss_32 <- readRDS(here::here("data/vss_32.rds"))

experiment <- vi_survey %>%
  group_by(unique_lineup_id) %>%
  summarise(across(everything(), first)) %>%
  select(unique_lineup_id, attention_check, null_lineup, prop_detect,
         answer, effect_size, conventional_p_value,
         p_value, type, shape, a, b, x_dist, e_dist,
         e_sigma, include_z, k, n)

experiment <- vss_32 %>%
  left_join(experiment) %>%
  mutate(conventional_reject = conventional_p_value <= 0.05) %>%
  mutate(reject = p_value <= 0.05) %>%
  mutate(model_reject = vss_p_value <= 0.05) %>%
  mutate(model_boot_reject = vss_boot_p_value <= 0.05)

experiment <- experiment %>%
  left_join(readRDS(here::here("data/actual_ss.rds")))
```

```{r}
experiment %>%
  filter(!attention_check) %>%
  filter(!null_lineup) %>%
  mutate(type = ifelse(type == "polynomial", "non-linearity", type)) %>%
  group_by(type) %>%
  summarise(RMSE = yardstick::rmse_vec(actual_ss, vss),
            R2 = yardstick::rsq_vec(actual_ss, vss),
            MAE = yardstick::mae_vec(actual_ss, vss),
            HUBER = yardstick::huber_loss_vec(actual_ss, vss)) %>%
  mutate(across(RMSE:HUBER, ~format(.x, digits = 3))) %>%
  kableExtra::kable(booktabs = TRUE, 
                    label = "experiment-performance",
                    caption = "The performance of the $32 \\times 32$ model on the data used in the human subject experiment.",
                    escape = FALSE,
                    align = "lrrrr",
                    col.names = c("Violation", "RMSE", "$R^2$", "MAE", "Huber loss"))
```

For each lineup used in @li2024plot, there is one true residual plot and 19 null plots. While the distance $D$ for the true residual plot depends on the underlying data generating process, the distance $D$ for the null plots is zero. We have used our optimized computer vision model to estimate distance for both the true residual plots and the null plots. To have a fair comparison, $H_0$ will be rejected if the true residual plot has the greatest estimated distance among all plots in a lineup. Additionally, the appropriate conventional tests including the Ramsey Regression Equation Specification Error Test (RESET) [@ramsey1969tests] for non-linearity and the Breusch-Pagan test  [@breusch1979simple] for heteroskedasticity were applied on the same data for comparison.

The performance metrics of $\hat{D}$ for true residual plots are outlined in Table \ref{tab:experiment-performance}. It is notable that all performance metrics are slightly worse than those evaluated on the test data. Nevertheless, the mean absolute error remains at a low level, and the linear correlation between the prediction and the true value remains very high. Consistent with results in Table \ref{tab:performance-sub}, lineups with non-linearity issues are more challenging to predict than those with heteroskedasticity issues.

Table \ref{tab:human-conv-table} provides a summary of the agreement between decisions made by the computer vision model and conventional tests. The agreement rates between conventional tests and the computer vision model are 85.95% and 79.69% for residual plots containing heteroskedasticity and non-linearity patterns, respectively. These figures are higher than those calculated for visual tests conducted by human, indicating that the computer vision model exhibits behavior more akin to the best available conventional tests. However, Figure \ref{fig:conv-mosaic} shows that the computer vision model does not always reject when the conventional tests reject. And a small number of plots will be rejected by computer vision model but not by conventional tests.  This suggests that conventional tests are more sensitive than the computer vision model.

Figure \ref{fig:pcp} further illustrates the decisions made by visual tests conducted by human, computer vision models, and conventional tests, using a parallel coordinate plots. It can be observed that all three tests will agree with each other for around 50% of the cases. When visual tests conducted by human do not reject, there are substantial amount of cases where computer vision model also do not reject but conventional tests reject. There are much fewer cases that do not reject by visual tests and conventional tests, but is rejected by computer vision models. This indicates computer vision model can behave like visual tests conducted by human better than conventional tests. Moreover, there are great proportion of cases where visual tests conducted by human is the only test who does not reject.

When plotting the decision against the distance, as illustrated in Figure \ref{fig:power}, several notable observations emerge. Firstly, compared to conventional tests, the computer vision model tends to have fewer rejected cases when $D < 2$ and fewer non-rejected cases when $2< D < 4$. This suggests tests based on the computer vision model are less sensitive to small deviations from model assumptions than conventional tests but more sensitive to moderate deviations. Additionally, visual tests demonstrate the lowest sensitivity to residual plots with small distances where not many residual plots are rejected when $D < 2$. Similarly, for large distances where $D > 4$, almost all residual plots are rejected by the computer vision model and conventional tests, but for visual tests conducted by humans, the threshold is higher with $D > 5$.

In Figure \ref{fig:power}, rejection decisions are fitted by logistic regression models with no intercept terms and an offset equals to $\text{log}(0.05/0.95)$. The fitted curves for the computer vision model fall between those of conventional tests and visual tests for both non-linearity and heteroskedasticity, which means there is still potential to refine the computer vision model to better align its behavior with visual tests conducted by humans.

In the experiment conducted in @li2024plot, participants were allowed to make multiple selections for a lineup. The weighted detection rate was computed by assigning weights to each detection. If the participant selected zero plots, a weight of 0.05 was assigned; otherwise, if the true residual plot was detected, the weight was 1 divided by the number of selections. This weighted detection rate allow us to assess the quality of the distance measure purposed in this paper, by using the $\delta$-difference statistic. The $\delta$-difference is originally defined by @chowdhury2018measuring, is given by

$$
\delta = \bar{d}_{\text{true}} - \underset{j}{\text{max}}\left(\bar{d}_{\text{null}}^{(j)}\right) \quad \text{for}~j = 1,...,m-1,
$$

where $\bar{d}_{\text{null}}^{(j)}$ is the mean distance between the $j$-th null plot and the other null plots, $\bar{d}_{\text{true}}$ is the mean distance between the true residual plot and null plots, and $m$ is the number of plots in a lineup. These mean distances are used because, as noted by @chowdhury2018measuring, the distances can vary depending on which data plot is used for comparison. For instance, with three null plots, A, B and C, the distance between A and B may differ from the distance between A and C. To obtain a consistent distance for null plot A, averaging is necessary. However, this approach is not applicable to the distance proposed in this paper, as we only compare the residual plot against a theoretically good residual plot. Consequently, the statistic must be adjusted to evaluate our distance measure effectively.

One important aspect that the $\delta$-difference was designed to capture is the empirical distribution of distances for null plot. If we were to replace the mean distances $\bar{d}_{\text{null}}^{(j)}$ directly with $D_{\text{null}}^{(j)}$, the distance of the $j$-th null plot, the resulting distribution would be degenerate, since $D_{null}$ equals zero by definition. Additionally, $D$ can not be derived from an image, meaning it falls outside the scope of the distances considered by @chowdhury2018measuring. Instead, the focus should be on the empirical distribution of $\hat{D}$, as it influences decision-making. Therefore, the adjusted $\delta$-different is defined as

$$
\delta_{\text{adj}} = \hat{D} - \underset{j}{\text{max}}\left(\hat{D}_{\text{null}}^{(j)}\right) \quad \text{for}~j = 1,...,m-1,
$$

\noindent where $\hat{D}_{\text{null}}^{(j)}$ is the estimated distance for the $j$-th null plot, and $m$ is the number of plots in a lineup.

Figure \ref{fig:delta} displays the scatter plot of the weighted detection rate vs the adjusted $\delta$-difference. It indicates that the weighted detection rate increases as the adjusted $\delta$-difference increases, particularly when the adjusted $\delta$-difference is greater than zero. A negative adjusted $\delta$-difference suggests that there is at least one null plot in the lineup with a stronger visual signal than the true residual plot. In some instances, the weighted detection rate is close to one, yet the adjusted $\delta$-difference is negative. This discrepancy implies that the distance measure, or the estimated distance, may not perfectly reflect actual human behavior.

```{r hist-null-human}
lineup_vss <- readRDS(here::here("data/lineup_vss.rds"))
lineup_vss <- lineup_vss %>%
  group_by(unique_lineup_id) %>%
  mutate(delta_diff = vss[!null] - max(vss[null])) %>%
  mutate(gamma_diff = sum(vss[null] > vss[!null]))

experiment <- experiment %>%
  left_join(lineup_vss %>%
  filter(null == FALSE) %>%
  mutate(model_reject_20 = rank == 1) %>%
  select(unique_lineup_id, rank, model_reject_20))
```

```{r cache = TRUE}
vi_lineup <- get_vi_lineup()
```

```{r}
bind_rows(
  experiment %>%
    filter(!attention_check) %>%
    filter(!null_lineup) %>%
    mutate(type = ifelse(type == "polynomial", "non-linearity", type)) %>%
    group_by(type) %>%
    summarise(n = format(n()), agree = format(sum(model_reject == conventional_reject)), rate = format(mean(conventional_reject == model_reject), digits = 4)),
  experiment %>%
    filter(!attention_check) %>%
    filter(!null_lineup) %>%
    mutate(type = ifelse(type == "polynomial", "non-linearity", type)) %>%
    group_by(type) %>%
    summarise(n = format(n()), agree = format(sum(model_reject == reject)), rate = format(mean(model_reject == reject), digits = 4))
  ) %>%
  kableExtra::kable(booktabs = TRUE,
                    label = "human-conv-table",
                    caption = "Summary of the comparison of decisions made by computer vision model with decisions made by conventional tests and visual tests conducted by human.",
                    escape = FALSE,
                    align = "lrrr",
                    col.names = c("Violations", "\\#Samples", "\\#Agreements", "Agreement rate")) %>%
  kableExtra::pack_rows("Compared with conventional tests", 1, 2) %>%
  kableExtra::pack_rows("Compared with visual tests conducted by human", 3, 4)
```

```{r conv-mosaic, fig.pos = "!h", fig.cap = "Rejection rate ($p$-value $\\leq0.05$) of computer vision models conditional on conventional tests on non-linearity (left) and heteroskedasticity (right) lineups displayed using a mosaic plot. When the conventional test fails to reject, the computer vision mostly fails to reject the same plot as well as indicated by the height of the top right yellow rectangle, but there are non negliable amount of plots where the conventional test rejects but the computer vision model fails to reject as indicated by the width of the top left yellow rectangle."}
library(ggmosaic)

experiment %>%
  filter(!attention_check) %>%
  filter(!null_lineup) %>%
  mutate(type = ifelse(type == "polynomial", "non-linearity", type)) %>%
  mutate(type = ifelse(type == "non-linearity", "Non-linearity", "Heteroskedasticity")) %>%
  mutate(type = factor(type, levels = c("Non-linearity", "Heteroskedasticity"))) %>%
  mutate(model_reject_20 = ifelse(model_reject_20, "Reject", "No")) %>%
  mutate(conventional_reject = ifelse(conventional_reject, "Reject", "No")) %>%
  mutate(across(c(model_reject_20, conventional_reject), ~factor(.x, levels = c("Reject", "No")))) %>%
  ggplot() +
  geom_mosaic(aes(x = ggmosaic::product(model_reject_20, conventional_reject), 
                  fill = model_reject_20)) +
  facet_grid(~type) +
  ylab("Computer vision model rejects") +
  xlab("Conventional tests reject ") +
  # labs(fill = "Visual tests reject") +
  scale_fill_manual(values = c("#40B0A6", "#E1BE6A")) + 
  theme_bw() +
  theme(legend.position = "none") +
  coord_fixed()
```

```{r power, fig.pos = "!h", fig.cap = "Comparison of power of visual tests, conventional tests and the computer vision model. Marks along the x-axis at the bottom of the plot represent rejections made by each type of test. Marks at the top of the plot represent acceptances. Power curves are fitted by logistic regression models with no intercept but an offset equals to $\\text{log}(0.05/0.95)$."}

max_actual_ss <- max(filter(experiment, !attention_check, !null_lineup)$actual_ss)
max_poly_ss <- max(filter(experiment, !attention_check, !null_lineup, type == "polynomial")$actual_ss)
max_heter_ss <- max(filter(experiment, !attention_check, !null_lineup, type != "polynomial")$actual_ss)

model_glm_pred_poly <- glm(model_reject_20 ~ actual_ss - 1, 
    data = filter(experiment, !attention_check, !null_lineup, type == "polynomial") %>% mutate(offset0 = log(0.05/0.95)), 
    offset = offset0,
    family = binomial()) %>%
  predict(data.frame(actual_ss = seq(0, max_poly_ss, 0.01), offset0 = log(0.05/0.95)),
          type = "response") %>%
  data.frame(d = seq(0, max_poly_ss, 0.01), type = "Non-linearity", name = "Computer vision model", value = .) %>%
  mutate(type = factor(type, levels = c("Non-linearity", "Heteroskedasticity")))

visual_glm_pred_poly <- glm(reject ~ actual_ss - 1, 
    data = filter(experiment, !attention_check, !null_lineup, type == "polynomial") %>% mutate(offset0 = log(0.05/0.95)), 
    offset = offset0,
    family = binomial()) %>%
  predict(data.frame(actual_ss = seq(0, max_poly_ss, 0.01), offset0 = log(0.05/0.95)),
          type = "response") %>%
  data.frame(d = seq(0, max_poly_ss, 0.01), type = "Non-linearity", name = "Visual test", value = .) %>%
  mutate(type = factor(type, levels = c("Non-linearity", "Heteroskedasticity")))

conv_glm_pred_poly <- glm(conventional_reject ~ actual_ss - 1, 
    data = filter(experiment, !attention_check, !null_lineup, type == "polynomial") %>% mutate(offset0 = log(0.05/0.95)), 
    offset = offset0,
    family = binomial()) %>%
  predict(data.frame(actual_ss = seq(0, max_poly_ss, 0.01), offset0 = log(0.05/0.95)),
          type = "response") %>%
  data.frame(d = seq(0, max_poly_ss, 0.01), type = "Non-linearity", name = "Conventional test", value = .) %>%
  mutate(type = factor(type, levels = c("Non-linearity", "Heteroskedasticity")))

model_glm_pred_heter <- glm(model_reject_20 ~ actual_ss - 1, 
    data = filter(experiment, !attention_check, !null_lineup, type != "polynomial") %>% mutate(offset0 = log(0.05/0.95)), 
    offset = offset0,
    family = binomial()) %>%
  predict(data.frame(actual_ss = seq(0, max_heter_ss, 0.01), offset0 = log(0.05/0.95)),
          type = "response") %>%
  data.frame(d = seq(0, max_heter_ss, 0.01), type = "Heteroskedasticity", name = "Computer vision model", value = .) %>%
  mutate(type = factor(type, levels = c("Non-linearity", "Heteroskedasticity")))

visual_glm_pred_heter <- glm(reject ~ actual_ss - 1, 
    data = filter(experiment, !attention_check, !null_lineup, type != "polynomial") %>% mutate(offset0 = log(0.05/0.95)), 
    offset = offset0,
    family = binomial()) %>%
  predict(data.frame(actual_ss = seq(0, max_heter_ss, 0.01), offset0 = log(0.05/0.95)),
          type = "response") %>%
  data.frame(d = seq(0, max_heter_ss, 0.01), type = "Heteroskedasticity", name = "Visual test", value = .) %>%
  mutate(type = factor(type, levels = c("Non-linearity", "Heteroskedasticity")))

conv_glm_pred_heter <- glm(conventional_reject ~ actual_ss - 1, 
    data = filter(experiment, !attention_check, !null_lineup, type != "polynomial") %>% mutate(offset0 = log(0.05/0.95)), 
    offset = offset0,
    family = binomial()) %>%
  predict(data.frame(actual_ss = seq(0, max_heter_ss, 0.01), offset0 = log(0.05/0.95)),
          type = "response") %>%
  data.frame(d = seq(0, max_heter_ss, 0.01), type = "Heteroskedasticity", name = "Conventional test", value = .) %>%
  mutate(type = factor(type, levels = c("Non-linearity", "Heteroskedasticity")))

mutate(experiment, type = ifelse(type == "polynomial", "Non-linearity", "Heteroskedasticity")) %>%
  mutate(type = factor(type, levels = c("Non-linearity", "Heteroskedasticity"))) %>%
  filter(!null_lineup, !attention_check) %>%
ggplot() +
  geom_line(data = model_glm_pred_poly, aes(d, value, col = "Computer vision model"), linewidth = 0.8) +
  geom_line(data = model_glm_pred_heter, aes(d, value, col = "Computer vision model"), linewidth = 0.8) +
  geom_line(data = visual_glm_pred_poly, aes(d, value, col = "Visual test"), linewidth = 0.8) +
  geom_line(data = visual_glm_pred_heter, aes(d, value, col = "Visual test"), linewidth = 0.8) +
  geom_line(data = conv_glm_pred_poly, aes(d, value, col = "Conventional test"), linewidth = 0.8) +
  geom_line(data = conv_glm_pred_heter, aes(d, value, col = "Conventional test"), linewidth = 0.8) +
  geom_segment(data = ~filter(.x, !reject), aes(x = actual_ss, xend = actual_ss, y = -0.1, yend = -0.1 + 0.03, col = "Visual test"), alpha = 0.4) +
  geom_segment(data = ~filter(.x, reject), aes(x = actual_ss, xend = actual_ss, y = 1.1 - 0.03 * 2, yend = 1.1 - 0.03 * 3, col = "Visual test"), alpha = 0.4) +
  geom_segment(data = ~filter(.x, !model_reject_20), aes(x = actual_ss, xend = actual_ss, y = -0.1 + 0.03, yend = -0.1 + 0.03 * 2, col = "Computer vision model"), alpha = 0.4) +
  geom_segment(data = ~filter(.x, model_reject_20), aes(x = actual_ss, xend = actual_ss, y = 1.1 - 0.03, yend = 1.1 - 0.03 * 2, col = "Computer vision model"), alpha = 0.4) +
  geom_segment(data = ~filter(.x, !conventional_reject), aes(x = actual_ss, xend = actual_ss, y = -0.1 + 0.03 * 2, yend = -0.1 + 0.03 * 3, col = "Conventional test"), alpha = 0.4) +
  geom_segment(data = ~filter(.x, conventional_reject), aes(x = actual_ss, xend = actual_ss, y = 1.1, yend = 1.1 - 0.03, col = "Conventional test"), alpha = 0.4) +
  facet_grid(~type, scales = "free_x") +
  theme_light() +
  scale_color_brewer("", palette = "Dark2") +
  theme(legend.position = "bottom") +
  xlab("D") +
  ylab("Reject")
```

```{r pcp, fig.cap = "Parallel coordinate plots of decisions made by computer vision model, conventional tests and visual tests made by human.", eval = TRUE, fig.height = 4}
library(ggpcp)
mutate(experiment, type = ifelse(type == "polynomial", "Non-linearity", "Heteroskedasticity")) %>%
  mutate(type = factor(type, levels = c("Non-linearity", "Heteroskedasticity"))) %>%
  filter(!null_lineup, !attention_check) %>%
  select(type, actual_ss, model_reject_20, reject, conventional_reject) %>%
  mutate(test = rnorm(n())) %>%
  mutate(across(c(model_reject_20, conventional_reject, reject), ~ifelse(.x, "Reject", "Not reject"))) %>%
  mutate(agree = model_reject_20 == reject) %>%
  pcp_select(c(reject, model_reject_20, conventional_reject)) %>%
  group_by(type) %>%
  pcp_scale() %>%
  ungroup() %>%
  pcp_arrange(method = "from-right") %>%
  ggplot() +
  geom_pcp(aes_pcp(), linewidth = 0.1) +
  geom_pcp_labels(aes_pcp()) +
  geom_pcp_boxes(aes_pcp(), boxwidth = 0.1) +
  facet_wrap(~type, scales = "free_y") +
  theme_light() +
  scale_x_discrete(labels = c("Visual\ntest", "Computer\nvision\nmodel", "Conventional\ntest")) +
  labs(col = "D") +
  ylab("") +
  xlab("") +
  theme(axis.text.y = element_blank(), 
        axis.ticks.y = element_blank(), 
        axis.title = element_blank())
```

```{r delta, fig.pos = "!h", fig.cap = "A weighted detection rate vs adjusted $\\delta$-difference plot. The brown line is smoothing curve produced by fitting generalized additive models.", fig.height = 4}
experiment %>%
  left_join(lineup_vss %>%
              group_by(unique_lineup_id) %>%
              summarise(delta_diff = first(delta_diff), gamma_diff = first(gamma_diff))) %>%
  ggplot() +
  geom_vline(xintercept = 0, alpha = 0.4, linetype = 2) +
  geom_point(aes(delta_diff, prop_detect), alpha = 0.3) +
  geom_smooth(aes(delta_diff, prop_detect), se = FALSE, col = "#994F00") +
  ylab("Weighted detection rate") +
  xlab(expression(delta[adj])) +
  theme_light()
```

# Examples {#sec-examples}

In this section, we present the performance of trained computer vision model on three example datasets. These include the dataset associated with the residual plot displaying a "left-triangle" shape, as displayed in Figure \ref{fig:false-finding}, along with the Boston housing dataset [@harrison1978hedonic], and the "dino" datasets from the `datasauRus` R package [@datasaurus].

The first example illustrates a scenario where both the computer vision model and human visual inspection successfully avoid rejecting $H_0$ when $H_0$ is true, contrary to conventional tests. This underscores the necessity of visually examining the residual plot.

In the second example, we encounter a more pronounced violation of the model, resulting in rejection of $H_0$ by all three tests. This highlights the practicality of the computer vision model, particularly for less intricate tasks.

The third example presents a situation where the model deviation is non-typical. Here, the computer vision model and human visual inspection reject $H_0$, whereas some commonly used conventional tests do not. This emphasizes the benefits of visual inspection and the unique advantage of the computer vision model, which, like humans, makes decisions based on visual discoveries.

## Left-triangle

In Section \ref{sec-model-introduction}, we presented an example residual plot showcased in Figure \ref{fig:false-finding}, illustrating how humans might misinterpret the "left-triangle" shape as indicative of heteroskedasticity. Additionally, the Breusch-Pagan test yielded a rejection with a $p$-value of 0.046, despite the residuals originating from a correctly specified model. Figure \ref{fig:false-lineup} offers a lineup for this fitted model, showcasing various degrees of "left-triangle" shape across all residual plots. This phenomenon is evidently caused by the skewed distribution of the fitted values. Notably, if the residual plot were evaluated through a visual test, it would not be rejected since the true residual plot positioned at 10 can not be distinguished from the others.

Figure \ref{fig:false-check} presents the results of the assessment by the computer vision model. Notably, the observed visual signal strength is considerably lower than the 95% sample quantile of the null distribution. Moreover, the bootstrapped distribution suggests that it is highly improbable for the fitted model to be misspecified as the majority of bootstrapped fitted models will not be rejected. Thus, for this particular fitted model, both the visual test and the computer vision model will not reject $H_0$. However, the Breusch-Pagan test will reject $H_0$ because it can not effectively utilize information from null plots.

The attention map at Figure \ref{fig:false-check}B suggests that the estimation is highly influenced by the top-right and bottom-right part of the residual plot, as it forms two vertices of the triangular shape. A principal component analysis (PCA) is also performed on the output of the global pooling layer of the computer vision model. As mentioned in @simonyan2014very, a computer vision model built upon the convolutional blocks can be viewed as a feature extractor. For the $32 \times 32$ model, there are 256 features outputted from the global pooling layer, which can be further used for different visual tasks not limited to distance prediction. To see if these features can be effectively used for distinguishing null plots and true residual plot, we linearly project them into the first and second principal components space as shown in Figure \ref{fig:false-check}D. It can be observed that because the bootstrapped plots are mostly similar to the null plots, the points drawn in different colors are mixed together. The true residual plot is also covered by both the cluster of null plots and cluster of bootstrapped plots. This accurately reflects our understanding of Figure \ref{fig:false-lineup}.

```{r}
set.seed(452)
ori_x <- rand_lognormal()
dat <- heter_model(b = 0, x = closed_form(~-ori_x))$gen(300)

mod <- lm(y ~ x, data = dat)
my_vi <- autovi::auto_vi(fitted_model = mod)
```

```{r}
if (!file.exists(here::here("cached_data/fig1_check.rds"))) {
  keras_model <- autovi::get_keras_model("vss_phn_32")
  my_vi$check(null_draws = 200L, boot_draws = 200L, keras_model = keras_model, extract_feature_from_layer = "global_max_pooling2d")
  saveRDS(my_vi$check_result, here::here("cached_data/fig1_check.rds"))
}

```

```{r}
my_vi$check_result <- readRDS(here::here("cached_data/fig1_check.rds"))
```

```{r}
define_gradient <- function() {
  reticulate::py_run_string(glue::glue(
"
import tensorflow as tf
import PIL
import numpy as np

def get_smooth_gradient(keras_mod, plot_path, target_size, input_auxiliary = None, noise_level = 0.1, n = 50):
    im = PIL.Image.open(plot_path)
    im = im.resize(target_size)
    input_im = tf.keras.utils.img_to_array(im)
    input_im = np.reshape(input_im, tuple([1]) + tuple(target_size) + tuple([3]))
    if input_auxiliary is not None:
        input_auxiliary = np.reshape(np.array(input_auxiliary), (1, 5))

    if input_auxiliary is not None:
        input_auxiliary = tf.Variable(input_auxiliary)

    im_grad = None
    sigma = noise_level * (np.max(input_im) - np.min(input_im))

    for i in range(n):    
        new_input_im = input_im + np.random.normal(0, sigma, input_im.shape)
        new_input_im = tf.Variable(new_input_im)

        with tf.GradientTape() as tape:
            if input_auxiliary is not None:
                pred = keras_mod([new_input_im, input_auxiliary])
            else:
                pred = keras_mod(new_input_im)
        new_im_grad = tape.gradient(pred, new_input_im)
        new_im_grad = tf.image.rgb_to_grayscale(new_im_grad)[0].numpy()
        if im_grad is None:
            im_grad = new_im_grad
        else:
            im_grad = im_grad + new_im_grad

    im_grad = im_grad / n

    input_im = tf.Variable(input_im)
    if input_auxiliary is not None:
        with tf.GradientTape() as tape:
            pred = keras_mod([input_im, input_auxiliary])
        auxiliary_grad = tape.gradient(pred, input_auxiliary)
    if input_auxiliary is not None:
        return (im_grad, auxiliary_grad[0].numpy())
    else:
        return im_grad

def get_raw_gradient(keras_mod, plot_path, target_size, input_auxiliary = None):
    im = PIL.Image.open(plot_path)
    im = im.resize(target_size)
    input_im = tf.keras.utils.img_to_array(im)
    input_im = np.reshape(input_im, tuple([1]) + tuple(target_size) + tuple([3]))
    if input_auxiliary is not None:
        input_auxiliary = np.reshape(np.array(input_auxiliary), (1, 5))

    input_im = tf.Variable(input_im)
    if input_auxiliary is not None:
        input_auxiliary = tf.Variable(input_auxiliary)

    with tf.GradientTape() as tape:
        if input_auxiliary is not None:
            pred = keras_mod([input_im, input_auxiliary])
        else:
            pred = keras_mod(input_im)
    im_grad = tape.gradient(pred, input_im)

    if input_auxiliary is not None:
        with tf.GradientTape() as tape:
            pred = keras_mod([input_im, input_auxiliary])
        auxiliary_grad = tape.gradient(pred, input_auxiliary)

    im_grad = tf.image.rgb_to_grayscale(im_grad)
    if input_auxiliary is not None:
        return (im_grad[0].numpy(), auxiliary_grad[0].numpy())
    else:
        return im_grad[0].numpy()
"
))
}

```

```{r}
if (!file.exists(here::here("cached_data/fig1_attention_map.rds"))) {
  reticulate::py_set_seed(10086, disable_hash_randomization = TRUE)
  define_gradient()
  keras_mod <- autovi::get_keras_model("vss_phn_32")
  plot_path <- my_vi$plot_resid() %>%
    autovi::save_plot()
  input_auxiliary <- my_vi$auxiliary()

  reticulate::py$get_smooth_gradient(keras_mod = keras_mod, 
                                     plot_path = plot_path, 
                                     target_size = c(32L, 32L), 
                                     input_auxiliary = input_auxiliary,
                                     noise_level = 0.5,
                                     n = 1000L) -> smooth_g

  x_range <- ggplot2::ggplot_build(my_vi$plot_resid())$layout$panel_params[[1]]$x.range
  y_range <- ggplot2::ggplot_build(my_vi$plot_resid())$layout$panel_params[[1]]$y.range

  smooth_att_map <- smooth_g[[1]] |> 
    as.data.frame() |>
    dplyr::mutate(row = rev(1:32)) |>
    tidyr::pivot_longer(V1:V32, names_to = "column", values_to = "gradient") |>
    dplyr::mutate(column = as.integer(gsub("V", "", column))) |>
    dplyr::mutate(column = x_range[1] + column * (x_range[2] - x_range[1])/(32 - 1)) |>
    dplyr::mutate(row = y_range[1] + row * (y_range[2] - y_range[1])/(32 - 1))

  scale_zero_one <- function(x) (x - min(x))/(max(x) - min(x))

  p <- ggplot() +
    ggplot2::geom_raster(data = mutate(smooth_att_map, gradient = scale_zero_one(gradient)), ggplot2::aes(column, row, fill = gradient), alpha = 0.7) +
    scale_fill_gradient(low = "black", high = "white") +
    ggplot2::theme_void() +
    theme(legend.position = "none")

  saveRDS(p, here::here("cached_data/fig1_attention_map.rds"))

}
```

```{r false-check, fig.pos = "!h", fig.cap = 'A summary of the residual plot assessment evaluted on 200 null plots and 200 bootstrapped plots. (A) The true residual plot exhibiting a "left-triangle" shape. (B) The attention map produced by computing the gradient of the output with respect to the greyscale input.  (C) The density plot of estimated distance for null plots and bootstrapped plots. The green area indicates the distribution of estimated distances for bootstrapped plots, while the yellow area represents the distribution of estimated distances for null plots. The fitted model will not be rejected since $\\hat{D} < Q_{null}(0.95)$. (D) plot of first two principal components of features extracted from the global pooling layer of the computer vision model.  ', fig.height = 8}
p1 <- my_vi$plot_resid(theme = theme_light()) + ggtitle(glue::glue("(A) Residual plot (MVI = {format(my_vi$check_result$observed$vss - log(nrow(my_vi$get_fitted_and_resid())) + 10, digits = 3)}, weak)")) 
p2 <- readRDS(here::here("cached_data/fig1_attention_map.rds")) + ggtitle("(B) Attention map")
p3 <- my_vi$summary_plot() +
  ggtitle("(C) Density plot of estimated D", subtitle = glue::glue("Conventional p-value = {format(HETER_MODEL$test(dat)$p_value, digits = 3)}")) +
  theme(legend.position = "bottom", legend.box = "vertical") +
  labs(linetype = "") +
  xlab(expression(hat(D))) +
  scale_linetype(labels = c(expression(Q[null](0.95)), expression(observed~hat(D)))) +
  scale_fill_manual(values = c("#40B0A6", "#E1BE6A")) +
  scale_color_manual(values = c("#40B0A6", "#E1BE6A"))
pca_result <- my_vi$feature_pca()
pc1_per <- (attributes(pca_result)$sdev^2/256)[1]
pc2_per <- (attributes(pca_result)$sdev^2/256)[2]
p4 <- pca_result %>%
  mutate(set = ifelse(set == "boot", "Boot", set)) %>%
  mutate(set = ifelse(set == "null", "Null", set)) %>%
  ggplot() +
  geom_point(data = ~filter(.x, set != "observed"), aes(PC1, PC2, col = set), size = 2, alpha = 0.3) +
  geom_point(data = ~filter(.x, set == "observed"), aes(PC1, PC2), size = 4, col = "black", fill = "black") +
  geom_text(data = ~filter(.x, set == "observed"), aes(PC1, PC2, label = "Observed"), nudge_x = 10) +
  theme_light() +
  scale_color_brewer(palette = "Dark2") +
  xlab(glue::glue("PC1 ({scales::percent(pc1_per)})")) +
  ylab(glue::glue("PC2 ({scales::percent(pc2_per)})")) +
  labs(col = "") +
  theme(legend.position = "bottom") +
  ggtitle("(D) PCA for extracted features") +
  scale_color_manual(values = c("#40B0A6", "#E1BE6A"))

patchwork::wrap_plots(p1, p2, p3, p4, nrow = 2, widths = c(1, 1), heights = c(1, 1))
```

```{r false-lineup, fig.pos = "!h", fig.cap = 'A lineup of residual plots displaying "left-triangle" visual patterns. The true residual plot occupies position 10, yet there are no discernible visual patterns that distinguish it from the other plots.', fig.height = 8}
pos <- 10
result <- tibble()
for (i in c(1:20)[-pos]) {
  result <- bind_rows(result,
                      my_vi$rotate_resid() %>%
                        mutate(k = i))
}
result <- bind_rows(my_vi$get_fitted_and_resid() %>% 
                      mutate(k = pos),
                    result)
result %>%
  VI_MODEL$plot_lineup(remove_grid_line = TRUE,
                       remove_axis = TRUE,
                       theme = theme_light())
```

## Boston Housing

The Boston housing dataset, originally published by @harrison1978hedonic, offers insights into housing in the Boston, Massachusetts area. For illustration purposes, we utilize a reduced version from Kaggle, comprising 489 rows and 4 columns: average number of rooms per dwelling (RM), percentage of lower status of the population (LSTAT), pupil-teacher ratio by town (PTRATIO), and median value of owner-occupied homes in $1000's (MEDV). In our analysis, MEDV will serve as the response variable, while the other columns will function as predictors in a linear regression model. Our primary focus is to detect non-linearity, because the relationships between RM and MEDV or LSTAT and MEDV are non-linear.

Figure \ref{fig:boston-check} displays the residual plot and the assessment conducted by the computer vision model. A clear non-linearity pattern resembling a "U" shape is shown in the plot A. Furthermore, the RESET test yields a very small $p$-value. The estimated distance $\hat{D}$ significantly exceeds $Q_{null}(0.95)$, leading to rejection of $H_0$. The bootstrapped distribution also suggests that almost all the bootstrapped fitted models will be rejected, indicating that the fitted model is unlikely to be correctly specified. The attention map in plot B suggests the center of the image has higher leverage than other areas, and it is the turning point of the "U" shape. The CPA provided in plot D shows two distinct clusters of data points, further underling the visual differences between bootstrapped plots and null plots. This coincides the findings from Figure \ref{fig:boston-lineup}, where the true plot exhibiting a "U" shape is visually distinctive from null plots. If a visual test is conducted by human, $H_0$ will also be rejected.

```{r}
housing <- read_csv(here::here("data/housing.csv"))
mod <- lm(MEDV ~ ., data = housing)

my_vi <- autovi::auto_vi(fitted_model = mod)
```

```{r}
if (!file.exists(here::here("cached_data/boston_check.rds"))) {
  keras_model <- autovi::get_keras_model("vss_phn_32")
  my_vi$check(null_draws = 200L, boot_draws = 200L, keras_model = keras_model, extract_feature_from_layer = "global_max_pooling2d")
  saveRDS(my_vi$check_result, here::here("cached_data/boston_check.rds"))
}

```

```{r}
my_vi$check_result <- readRDS(here::here("cached_data/boston_check.rds"))
```

```{r}
if (!file.exists(here::here("cached_data/boston_attention_map.rds"))) {
  reticulate::py_set_seed(10086, disable_hash_randomization = TRUE)
  define_gradient()
  keras_mod <- autovi::get_keras_model("vss_phn_32")
  plot_path <- my_vi$plot_resid() %>%
    autovi::save_plot()
  input_auxiliary <- my_vi$auxiliary()

  reticulate::py$get_smooth_gradient(keras_mod = keras_mod, 
                                     plot_path = plot_path, 
                                     target_size = c(32L, 32L), 
                                     input_auxiliary = input_auxiliary,
                                     noise_level = 0.5,
                                     n = 1000L) -> smooth_g

  x_range <- ggplot2::ggplot_build(my_vi$plot_resid())$layout$panel_params[[1]]$x.range
  y_range <- ggplot2::ggplot_build(my_vi$plot_resid())$layout$panel_params[[1]]$y.range

  smooth_att_map <- smooth_g[[1]] |> 
    as.data.frame() |>
    dplyr::mutate(row = rev(1:32)) |>
    tidyr::pivot_longer(V1:V32, names_to = "column", values_to = "gradient") |>
    dplyr::mutate(column = as.integer(gsub("V", "", column))) |>
    dplyr::mutate(column = x_range[1] + column * (x_range[2] - x_range[1])/(32 - 1)) |>
    dplyr::mutate(row = y_range[1] + row * (y_range[2] - y_range[1])/(32 - 1))

  scale_zero_one <- function(x) (x - min(x))/(max(x) - min(x))

  p <- ggplot() +
    ggplot2::geom_raster(data = mutate(smooth_att_map, gradient = scale_zero_one(gradient)), ggplot2::aes(column, row, fill = gradient), alpha = 0.7) +
    scale_fill_gradient(low = "black", high = "white") +
    ggplot2::theme_void() +
    theme(legend.position = "none")

  saveRDS(p, here::here("cached_data/boston_attention_map.rds"))

}
```

```{r boston-check, fig.pos = "!h", fig.cap = 'A summary of the residual plot assessment for the Boston housing fitted model evaluted on 200 null plots and 200 bootstrapped plots. (A) The true residual plot exhibiting a "U" shape. (B) The attention map produced by computing the gradient of the output with respect to the greyscale input.  (C) The density plot of estimated distance for null plots and bootstrapped plots. The blue area indicates the distribution of estimated distances for bootstrapped plots, while the yellow area represents the distribution of estimated distances for null plots. The fitted model will be rejected since $\\hat{D} \\geq Q_{null}(0.95)$. (D) plot of first two principal components of features extracted from the global pooling layer of the computer vision model. ', fig.height = 8}
p1 <- my_vi$plot_resid(theme = theme_light()) + ggtitle(glue::glue("(A) Residual plot (MVI = {format(my_vi$check_result$observed$vss - log(nrow(my_vi$get_fitted_and_resid())) + 10, digits = 3)}, strong)")) 
p2 <- readRDS(here::here("cached_data/boston_attention_map.rds")) + ggtitle("(B) Attention map")
p3 <- my_vi$summary_plot() +
  ggtitle("(C) Density plot of estimated D", subtitle = glue::glue("Conventional p-value = {format(lmtest::resettest(mod)$p.value, digits = 3)}")) +
  theme(legend.position = "bottom", legend.box = "vertical") +
  labs(linetype = "") +
  xlab(expression(hat(D))) +
  scale_linetype(labels = c(expression(Q[null](0.95)), expression(observed~hat(D)))) +
  scale_fill_manual(values = c("#40B0A6", "#E1BE6A")) +
  scale_color_manual(values = c("#40B0A6", "#E1BE6A"))
pca_result <- my_vi$feature_pca()
pc1_per <- (attributes(pca_result)$sdev^2/256)[1]
pc2_per <- (attributes(pca_result)$sdev^2/256)[2]
p4 <- pca_result %>%
  mutate(set = ifelse(set == "boot", "Boot", set)) %>%
  mutate(set = ifelse(set == "null", "Null", set)) %>%
  ggplot() +
  geom_point(data = ~filter(.x, set != "observed"), aes(PC1, PC2, col = set), size = 2, alpha = 0.3) +
  geom_point(data = ~filter(.x, set == "observed"), aes(PC1, PC2), size = 4, col = "black", fill = "black") +
  geom_text(data = ~filter(.x, set == "observed"), aes(PC1, PC2, label = "Observed"), nudge_x = 7) +
  theme_light() +
  scale_color_brewer(palette = "Dark2") +
  xlab(glue::glue("PC1 ({scales::percent(pc1_per)})")) +
  ylab(glue::glue("PC2 ({scales::percent(pc2_per)})")) +
  labs(col = "") +
  theme(legend.position = "bottom") +
  ggtitle("(D) PCA for extracted features") +
  scale_color_manual(values = c("#40B0A6", "#E1BE6A"))

patchwork::wrap_plots(p1, p2, p3, p4, nrow = 2, widths = c(1, 1), heights = c(1, 1))
```

```{r boston-lineup, fig.pos = "!h", fig.cap = 'A lineup of residual plots for the Boston housing fitted model. The true residual plot is at position 7. It can be easily identified as the most different plot.', fig.height = 8}
pos <- 7
result <- tibble()
for (i in c(1:20)[-pos]) {
  result <- bind_rows(result,
                      my_vi$rotate_resid() %>%
                        mutate(k = i))
}
result <- bind_rows(my_vi$get_fitted_and_resid() %>% 
                      mutate(k = pos),
                    result)
result %>%
  VI_MODEL$plot_lineup(remove_grid_line = TRUE,
                       remove_axis = TRUE,
                       theme = theme_light())
```

## DatasauRus

The computer vision model possesses the capability to detect not only typical issues like non-linearity, heteroskedasticity, and non-normality but also artifact visual patterns resembling real-world objects, as long as they do not appear in null plots. These visual patterns can be challenging to categorize in terms of model violations. Therefore, we will employ the RESET test, the Breusch-Pagan test, and the Shapiro-Wilk test [@shapiro1965analysis] for comparison.

The "dino" dataset within the `datasauRus` R package exemplifies this scenario. With only two columns, x and y, fitting a regression model to this data yields a residual plot resembling a "dinosaur", as displayed in Figure \ref{fig:dino-check}A. Unsurprisingly, this distinct residual plot stands out in a lineup, as shown in Figure \ref{fig:dino-lineup}. A visual test conducted by humans would undoubtedly reject $H_0$.

According to the residual plot assessment by the computer vision model, $\hat{D}$ exceeds $Q_{null}(0.95)$, warranting a rejection of $H_0$. Additionally, most of the bootstrapped fitted models will be rejected, indicating an misspecified model. However, both the RESET test and the Breusch-Pagan test yield $p$-values greater than 0.3, leading to a non-rejection of $H_0$. Only the Shapiro-Wilk test rejects the normality assumption with a small $p$-value.

More importantly, the attention map in Figure \ref{fig:dino-check}B clearly exhibits a "dinosaur" shape, strongly suggesting the prediction of the distance is based on human perceptible visual patterns. The computer vision model is also capable of capturing the contour or the outline of the embedded shape, just like human being reading residual plots. The PCA in Figure \ref{fig:dino-check}D also shows that the cluster of bootstrapped plots is at the corner of the cluster of null plots.

More importantly, the attention map in Figure \ref{fig:dino-check}B clearly exhibits a "dinosaur" shape, strongly suggesting that the distance prediction is based on human-perceptible visual patterns. The computer vision model effectively captures the contour or outline of the embedded shape, similar to how humans interpret residual plots. Additionally, the PCA in Figure \ref{fig:dino-check}D demonstrates that the cluster of bootstrapped plots is positioned at the corner of the cluster of null plots.

In practice, without accessing the residual plot, it would be challenging to identify the artificial pattern of the residuals. Moreover, conducting a normality test for a fitted regression model is not always standard practice among analysts. Even when performed, violating the normality assumption is sometimes deemed acceptable, especially considering the application of quasi-maximum likelihood estimation in linear regression. This example underscores the importance of evaluating residual plots and highlights how the proposed computer vision model can facilitate this process.

```{r}
dino <- datasauRus::datasaurus_dozen %>% filter(dataset == "dino")
mod <- lm(y ~ ., data = select(dino, -dataset))

my_vi <- autovi::auto_vi(fitted_model = mod)
```

```{r}
if (!file.exists(here::here("cached_data/dino_check.rds"))) {
  keras_model <- autovi::get_keras_model("vss_phn_32")
  my_vi$check(null_draws = 200L, boot_draws = 200L, keras_model = keras_model, extract_feature_from_layer = "global_max_pooling2d")
  saveRDS(my_vi$check_result, here::here("cached_data/dino_check.rds"))
}
```

```{r}
my_vi$check_result <- readRDS(here::here("cached_data/dino_check.rds"))
```

```{r}
if (!file.exists(here::here("cached_data/dino_attention_map.rds"))) {
  reticulate::py_set_seed(10086, disable_hash_randomization = TRUE)
  define_gradient()
  keras_mod <- autovi::get_keras_model("vss_phn_32")
  plot_path <- my_vi$plot_resid() %>%
    autovi::save_plot()
  input_auxiliary <- my_vi$auxiliary()

  reticulate::py$get_smooth_gradient(keras_mod = keras_mod, 
                                     plot_path = plot_path, 
                                     target_size = c(32L, 32L), 
                                     input_auxiliary = input_auxiliary,
                                     noise_level = 0.5,
                                     n = 1000L) -> smooth_g

  x_range <- ggplot2::ggplot_build(my_vi$plot_resid())$layout$panel_params[[1]]$x.range
  y_range <- ggplot2::ggplot_build(my_vi$plot_resid())$layout$panel_params[[1]]$y.range

  smooth_att_map <- smooth_g[[1]] |> 
    as.data.frame() |>
    dplyr::mutate(row = rev(1:32)) |>
    tidyr::pivot_longer(V1:V32, names_to = "column", values_to = "gradient") |>
    dplyr::mutate(column = as.integer(gsub("V", "", column))) |>
    dplyr::mutate(column = x_range[1] + column * (x_range[2] - x_range[1])/(32 - 1)) |>
    dplyr::mutate(row = y_range[1] + row * (y_range[2] - y_range[1])/(32 - 1))

  scale_zero_one <- function(x) (x - min(x))/(max(x) - min(x))

  p <- ggplot() +
    ggplot2::geom_raster(data = mutate(smooth_att_map, gradient = scale_zero_one(gradient)), ggplot2::aes(column, row, fill = gradient), alpha = 0.7) +
    scale_fill_gradient(low = "black", high = "white") +
    ggplot2::theme_void() +
    theme(legend.position = "none")

  saveRDS(p, here::here("cached_data/dino_attention_map.rds"))

}
```

```{r dino-check, fig.pos = "!h", fig.cap = 'A summary of the residual plot assessment for the datasauRus fitted model evaluated on 200 null plots and 200 bootstrapped plots. (A) The residual plot exhibits a "dinosaur" shape. (B) The attention map produced by computing the gradient of the output with respect to the greyscale input.  (C) The density plot of estimated distance for null plots and bootstrapped plots. The blue area indicates the distribution of estimated distances for bootstrapped plots, while the yellow area represents the distribution of estimated distances for null plots. The fitted model will be rejected since $\\hat{D} \\geq Q_{null}(0.95)$. (D) plot of first two principal components of features extracted from the global pooling layer of the computer vision model.', fig.height = 8}

p1 <- my_vi$plot_resid(theme = theme_light()) + ggtitle(glue::glue("(A) Residual plot (MVI = {format(my_vi$check_result$observed$vss - log(nrow(my_vi$get_fitted_and_resid())) + 10, digits = 3)}, strong)")) 
p2 <- readRDS(here::here("cached_data/dino_attention_map.rds")) + ggtitle("(B) Attention map")
p3 <- my_vi$summary_plot() +
  ggtitle("(C) Density plot of estimated D", subtitle = glue::glue("RESET test p-value = {format(lmtest::resettest(mod)$p.value, digits = 3)}\n Breusch-Pagan test p-value = {format(lmtest::bptest(mod)$p.value, digits = 3)} \n Shapiro-Wilk test p-value = {format(shapiro.test(mod$residuals)$p.value, digits = 3)}")) +
  theme(legend.position = "bottom", legend.box = "vertical") +
  labs(linetype = "") +
  xlab(expression(hat(D))) +
  scale_linetype(labels = c(expression(Q[null](0.95)), expression(observed~hat(D)))) +
  scale_fill_manual(values = c("#40B0A6", "#E1BE6A")) +
  scale_color_manual(values = c("#40B0A6", "#E1BE6A"))
pca_result <- my_vi$feature_pca()
pc1_per <- (attributes(pca_result)$sdev^2/256)[1]
pc2_per <- (attributes(pca_result)$sdev^2/256)[2]
p4 <- pca_result %>%
  mutate(set = ifelse(set == "boot", "Boot", set)) %>%
  mutate(set = ifelse(set == "null", "Null", set)) %>%
  ggplot() +
  geom_point(data = ~filter(.x, set != "observed"), aes(PC1, PC2, col = set), size = 2, alpha = 0.3) +
  geom_point(data = ~filter(.x, set == "observed"), aes(PC1, PC2), size = 4, col = "black", fill = "black") +
  geom_text(data = ~filter(.x, set == "observed"), aes(PC1, PC2, label = "Observed"), nudge_x = -10) +
  theme_light() +
  scale_color_brewer(palette = "Dark2") +
  xlab(glue::glue("PC1 ({scales::percent(pc1_per)})")) +
  ylab(glue::glue("PC2 ({scales::percent(pc2_per)})")) +
  labs(col = "", shape = "") +
  theme(legend.position = "bottom", legend.box="vertical") +
  ggtitle("(D) PCA for extracted features") +
  scale_color_manual(values = c("#40B0A6", "#E1BE6A"))

patchwork::wrap_plots(p1, p2, p3, p4, nrow = 2, widths = c(1, 1), heights = c(1, 1))
```

```{r dino-lineup, fig.pos = "!h", fig.cap = 'A lineup of residual plots for the fitted model on the "dinosaur" dataset. The true residual plot is at position 17. It can be easily identified as the most different plot as the visual pattern is extremely artificial.', fig.height = 8}
pos <- 17
result <- tibble()
for (i in c(1:20)[-pos]) {
  result <- bind_rows(result,
                      my_vi$rotate_resid() %>%
                        mutate(k = i))
}
result <- bind_rows(my_vi$get_fitted_and_resid() %>% 
                      mutate(k = pos),
                    result)
result %>%
  VI_MODEL$plot_lineup(remove_grid_line = TRUE,
                       remove_axis = TRUE,
                       theme = theme_light())
```

# Limitations and Future Work

Despite the computer vision model performing well with general cases under the synthetic data generation scheme and the three examples used in this paper, this study has several limitations that could guide future work.

The proposed distance measure assumes that the true model is a classical normal linear regression model, which can be restrictive. Although this paper does not address the relaxation of this assumption, there are potential methods to evaluate other types of regression models. The most comprehensive approach would be to define a distance measure for each different class of regression model and then train the computer vision model following the methodology described in this paper. To accelerate training, one could use the convolutional blocks of our trained model as a feature extractor and perform transfer learning on top of it, as these blocks effectively capture shapes in residual plots. Another approach would be to transform the residuals so they are roughly normally distributed and have constant variance. If only raw residuals are used, the distance-based statistical testing compares the difference in distance to a classical normal linear regression model for the true plot and null plots. This comparison is meaningful only if the difference can be identified by the distance measure proposed in this paper.

There are other types of residual plots commonly used in diagnostics, such as residuals vs. predictor and quantile-quantile plots. In this study, we focused on the most commonly used residual plot as a starting point for exploring the new field of automated visual inference. Similarly, we did not explore other, more sophisticated computer vision model architectures and specifications for the same reason. While the performance of the computer vision model is acceptable, there is still room for improvement to achieve behavior more closely resembling that of humans interpreting residual plots. This may require external survey data or human subject experiment data to understand the fundamental differences between our implementation and human evaluation.

# Conclusions

In this paper, we have introduced a distance measure based on Kullback-Leibler divergence to quantify the disparity between the residual distribution of a fitted classical normal linear regression model and the reference residual distribution assumed under correct model specification. This distance measure effectively captures the magnitude of model violations in misspecified models. We propose a computer vision model to estimate this distance, utilizing the residual plot of the fitted model as input. The resulting estimated distance serves as the foundation for constructing a single Model Violations Index (MVI), facilitating the quantification of various model violations.

Moreover, the estimated distance enables the development of a formal statistical testing procedure by evaluating a large number of null plots generated from the fitted model. Additionally, employing bootstrapping techniques and refitting the regression model allows us to ascertain how frequently the fitted model is considered misspecified if data were repeatedly obtained from the same data generating process.

The trained computer vision model demonstrates strong performance on both the training and test sets, although it exhibits slightly lower performance on residual plots with non-linearity visual patterns compared to other types of violations. The statistical tests relying on the estimated distance predicted by the computer vision model exhibit lower sensitivity compared to conventional tests but higher sensitivity compared to visual tests conducted by humans. While the estimated distance generally mirrors the strength of the visual signal perceived by humans, there remains scope for further improvement in its performance.

Several examples are provided to showcase the effectiveness of the proposed method across different scenarios, emphasizing the similarity between visual tests and distance-based tests. Overall, both visual tests and distance-based tests can be viewed as ensemble of tests, aiming to assess 
any violations of model assumptions collectively. In contrast, individual residual diagnostic tests such as the RESET test and the Breusch-Pagan test only evaluate specific violations of model assumptions. In practice, selecting an appropriate set of statistical tests for regression diagnostics can be challenging, particularly given the necessity of adjusting the significance level for each test.

Our method holds significant value as it helps alleviate a portion of analysts' workload associated with assessing residual plots. While we recommend analysts to continue reading residual plots whenever feasible, as they offer invaluable insights, our approach serves as a valuable tool for automating the diagnostic process or for supplementary purposes when needed.

# Acknowledgement {-}

These `R` packages were used for the work: `tidyverse` [@tidyverse], `lmtest` [@lmtest], `mpoly` [@mpoly], `ggmosaic` [@ggmosaic], `kableExtra` [@kableextra], `patchwork` [@patchwork], `rcartocolor` [@rcartocolor], `glue` [@glue], `ggpcp` [@ggpcp], `here` [@here], `magick` [@magick], `yardstick` [@yardstick] and `reticulate` [@reticulate].

The article was created with R packages `rticles` [@rticles], `knitr` [@knitr] and `rmarkdown` [@rmarkdown]. The project's GitHub repository (https://github.com/TengMCing/auto_residual_reading_paper) contains all materials required to reproduce this article.

### 2024-11-01 paper.Rmd
keep files

@@
 # remotes::install_github("TengMCing/visage")
+options(tinytex.clean = FALSE)
 library(tidyverse)

### 2025-06-05 paper.Rmd
Modify contact

@@
   - name: Weihao Li
-    affil: a
-    email: weihao.li@monash.edu
+    affil: a, b
+    email: patrick.li@anu.edu.au
   - name: Dianne Cook

### 2025-06-05 paper.Rmd
update section 2.1

@@
 A straightforward architecture of the input layer involves feeding a vector of residuals along with a vector of fitted values, essentially providing all the necessary information for creating a residuals vs fitted values plot.
-However, a drawback of this method is the dynamic input size, which changes based on the number of observations.
+However, a drawback of this method is the variable input size, which changes based on the number of observations.
 For modern computer vision models implemented in mainstream software like TensorFlow [@abadi2016tensorflow], the input shape is typically fixed.
-One solution is to pad the input vectors with leading or trailing zeros when the input tensor expects longer vectors, but it may fail if the input vector surpasses the designed length.
-
-Another strategy is to summarize the residuals and fitted values separately using histograms and utilize the counts as the input.
-By controlling the number of bins in the histograms, it becomes possible to provide fixed-length input vectors.
-Still, since histograms only capture the marginal distribution of residuals and fitted values respectively, they can not be used to differentiate visual patterns with same marginal distributions but different joint distributions.
+One strategy is to pad the input vectors with leading or trailing zeros when the input tensor expects longer vectors, but it may fail if the input vector surpasses the designed length.
+Alternatively, sampling a fixed number of residual-fitted value pairs can ensure consistent input size.
+By carefully designing the sampling process, it is possible to maintain the input's representativeness while conforming to a fixed format.
 
@@
 These architectures can effectively capture and summarize spatial information from nearby pixels, which is less straightforward with vector input.
-While encoding data points as pixels might lead to some loss of detail, this approach still provides a more comprehensive representation than the histogram-based strategy because it captures the joint distribution of fitted values and residuals.
-The main considerations are the image resolution and the aesthetics of the residual plot.
-In general, a higher resolution provides more information to the model but comes with a trade-off of increased complexity and greater difficulty in training.
+Encoding data points as pixels introduces certain trade-offs.
+While some detail may be lost during the discretization process, this approach aligns with the sampling-based strategy by focusing on summarizing the overall data distribution.
+Unlike vectors, which struggle to represent spatial relationships, image-based inputs naturally encode the joint distribution of residuals and fitted values in a visually interpretable format.
+This richer representation gives the model a more comprehensive and intuitive understanding of the residual plot's underlying structure.
+Key considerations for using image-based input include image resolution and the aesthetics of the residual plot.
+In general, a higher resolution provides more information to the model but comes at the cost of of increased complexity and greater difficulty in training.
 As for the aesthetics of the residual plot, a practical solution is to consistently present residual plots in the same style to the model.
@@
 
-Taking into account the implementation cost and the need for model interpretability, we used the single residual plot input format in this paper.
+Taking into account the implementation cost and the need for model interpretability, we used the single residual plot input format in this chapter.
 

### 2025-06-05 paper.Rmd
Fix typo
Update section 2.2

@@
 
-Taking into account the implementation cost and the need for model interpretability, we used the single residual plot input format in this chapter.
+Taking into account the implementation cost and the need for model interpretability, we used the single residual plot input format in this paper.
 
@@
 
-Given that the input is a single residual plot represented as a fixed-resolution image, we can choose the output from the computer vision model to be either binary (classification) or numeric (regression).
+Given that the input is a single residual plot represented as a fixed-resolution image, the output of the computer vision model can be designed for various tasks: binary classification, categorical classification with multiple classes, or numeric regression.
 
@@
 
+For a multi-class outcome, the model could classify images into categories such as "contains outliers" vs.
+"does not contain outliers" or "is non-linear" vs.
+"not non-linear," among other diagnostic features.
+While this approach is appealing in theory, it essentially reverts to the traditional testing framework, where tests focus on detecting specific types of model violations in isolation.
+
+One significant advantage of the lineup protocol for residual diagnostics is its ability to detect multiple model violations simultaneously, provided these violations produce visually distinguishable differences in the plots.
+In contrast, training a classification model would require explicitly defining and mapping all potential violations of interest, or at least the most common ones, which can be impractical.
+If the goal is to identify specific cases, conventional statistical tests grounded in regression model assumptions are often more effective.
+Computer vision models may be less suited to such tasks.
+
 Alternatively, the output could be a meaningful and interpretable numerical measure useful for assessing residual plots, such as the strength of suspicious visual patterns reflecting the extent of model violations, or the difficulty index for identifying whether a residual plot has no issues.
@@
 In this study, we chose to define and use a distance between a true residual plot and a theoretically "good" residual plot.
-This is further explained in Section \ref{sec-model-distance-between-residual-plots}.
+This is further explained in @sec-model-distance-between-residual-plots.
 @vo2016localizing have also demonstrated that defining a proper distance between images can enhance the matching accuracy in image search compared to a binary outcome model.

### 2025-06-05 paper.Rmd
Fix table cross ref issue

@@
            range = c("$\\text{MVI} > 8$", "$6 < \\text{MVI} < 8$", "$\\text{MVI} < 6$")) %>%
-  kableExtra::kable(booktabs = TRUE,
+  kableExtra::kable(format = "latex",
+                    booktabs = TRUE,
                     label = "mvi",

### 2025-06-05 paper.Rmd
Fix more table cross ref issues

@@
        Domain = c("\\{2, 3, ..., 18\\}", "[-1, 1]", "[0, 100]", "\\{0, 1\\}", "\\{0, 1\\}", "\\{discrete, uniform, normal, lognormal\\}", "\\{discrete, uniform, normal, lognormal\\}", "\\{discrete, uniform, normal, lognormal\\}", "[0.0625, 9]", "[0.3, 0.6]", "[0.3, 0.6]", "[50, 500]")) %>%
-  kableExtra::kable(escape = FALSE,
+  kableExtra::kable(format = "latex",
+                    escape = FALSE,
                     label = "factor",
@@
            Domain = c("\\{4, 8, 16, 32, 64\\}", "[0.1, 0.6]", "\\{false, true\\}", "\\{max, average\\}", "\\{false, true\\}", "\\{128, 256, 512, 1024, 2048\\}", "\\{false, true\\}", "[0.1, 0.6]", "[$10^{-8}$, $10^{-1}$]")) %>%
-  kableExtra::kable(booktabs = TRUE, 
+  kableExtra::kable(format = "latex",
+                    booktabs = TRUE, 
                     label = "hyperparameter",

### 2025-06-05 paper.Rmd
Add one paragraph about predictors in experimental design
Fix figure cross ref issues

@@
 
+The predictors, $\boldsymbol{x}_1$ and $\boldsymbol{x}_2$, are randomly generated from four distinct distributions, including $U(-1, 1)$ (uniform), $N(0, 0.3^2)$ (normal), $\text{lognormal}(0, 0.6^2)/3$ (skewed) and $U\{-1, 1\}$ (discrete uniform).
+
 ## Balanced Dataset
@@
            `128` = c("64", "0.4", "true", "average", "false", "256", "true", "0.1", "0.0052")) %>%
-  kableExtra::kable(booktabs = TRUE, 
+  kableExtra::kable(format = "latex",
+                    booktabs = TRUE, 
                     label = "best-hyperparameter",
@@
   mutate(across(HUBER, ~format(.x, digits = 2))) %>%
-  kableExtra::kable(booktabs = TRUE, 
+  kableExtra::kable(format = "latex",
+                    booktabs = TRUE, 
                     label = "performance",
@@
   select(-train_n, -train_RMSE) %>%
-  kableExtra::kable(booktabs = TRUE,
+  kableExtra::kable(format = "latex",
+                    booktabs = TRUE,
                     label = "performance-sub",

### 2025-06-05 paper.Rmd
Fix more figure cross ref issues

@@
   mutate(across(RMSE:HUBER, ~format(.x, digits = 3))) %>%
-  kableExtra::kable(booktabs = TRUE, 
+  kableExtra::kable(format = "latex",
+                    booktabs = TRUE, 
                     label = "experiment-performance",
@@
   ) %>%
-  kableExtra::kable(booktabs = TRUE,
+  kableExtra::kable(format = "latex",
+                    booktabs = TRUE,
                     label = "human-conv-table",

### 2025-06-05 paper.Rmd
Fix typo

@@
 The attention map in plot B suggests the center of the image has higher leverage than other areas, and it is the turning point of the "U" shape.
-The CPA provided in plot D shows two distinct clusters of data points, further underling the visual differences between bootstrapped plots and null plots.
+The PCA provided in plot D shows two distinct clusters of data points, further underling the visual differences between bootstrapped plots and null plots.
 This coincides the findings from Figure \ref{fig:boston-lineup}, where the true plot exhibiting a "U" shape is visually distinctive from null plots.
@@
 
-More importantly, the attention map in Figure \ref{fig:dino-check}B clearly exhibits a "dinosaur" shape, strongly suggesting the prediction of the distance is based on human perceptible visual patterns.
-The computer vision model is also capable of capturing the contour or the outline of the embedded shape, just like human being reading residual plots.
-The PCA in Figure \ref{fig:dino-check}D also shows that the cluster of bootstrapped plots is at the corner of the cluster of null plots.
-
 More importantly, the attention map in Figure \ref{fig:dino-check}B clearly exhibits a "dinosaur" shape, strongly suggesting that the distance prediction is based on human-perceptible visual patterns.

### 2025-06-05 paper.tex
Compile

% interactcadsample.tex
% v1.04 - May 2023

\documentclass[]{interact}

\usepackage{epstopdf}% To incorporate .eps illustrations using PDFLaTeX, etc.
\usepackage{subfigure}% Support for small, `sub' figures and tables
%\usepackage[nolists,tablesfirst]{endfloat}% To `separate' figures and tables from text if required

\usepackage{natbib}% Citation support using natbib.sty
\bibpunct[, ]{(}{)}{;}{a}{}{,}% Citation support using natbib.sty
\renewcommand\bibfont{\fontsize{10}{12}\selectfont}% Bibliography support using natbib.sty

\theoremstyle{plain}% Theorem-like structures provided by amsthm.sty
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}

\theoremstyle{remark}
\newtheorem{remark}{Remark}
\newtheorem{notation}{Notation}

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{lscape}
\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\def\tightlist{}
\usepackage{setspace}
\doublespacing

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\begin{document}

\articletype{}

\title{Automated Assessment of Residual Plots with Computer Vision
Models}

\author{\name{Weihao Li$^{a, b}$, Dianne Cook$^{a}$, Emi Tanaka$^{a, b,
c}$, Susan VanderPlas$^{d}$, Klaus Ackermann$^{a}$}
\affil{$^{a}$Department of Econometrics and Business Statistics, Monash
University, Clayton, VIC, Australia; $^{b}$Biological Data Science
Institute, Australian National University, Acton, ACT,
Australia; $^{c}$Research School of Finance, Actuarial Studies and
Statistics, Australian National University, Acton, ACT,
Australia; $^{d}$Department of Statistics, University of Nebraska,
Lincoln, Nebraska, USA}
}

\maketitle

\begin{abstract}
Plotting the residuals is a recommended procedure to diagnose deviations
from linear model assumptions, such as non-linearity,
heteroscedasticity, and non-normality. The presence of structure in
residual plots can be tested using the lineup protocol to do visual
inference. There are a variety of conventional residual tests, but the
lineup protocol, used as a statistical test, performs better for
diagnostic purposes because it is less sensitive and applies more
broadly to different types of departures. However, the lineup protocol
relies on human judgment which limits its scalability. This work
presents a solution by providing a computer vision model to automate the
assessment of residual plots. It is trained to predict a distance
measure that quantifies the disparity between the residual distribution
of a fitted classical normal linear regression model and the reference
distribution, based on Kullback-Leibler divergence. From extensive
simulation studies, the computer vision model exhibits lower sensitivity
than conventional tests but higher sensitivity than human visual tests.
It is slightly less effective on non-linearity patterns. Several
examples from classical papers and contemporary data illustrate the new
procedures, highlighting its usefulness in automating the diagnostic
process and supplementing existing methods.
\end{abstract}

\begin{keywords}
statistical graphics; data visualization; visual inference; computer
vision; machine learning; hypothesis testing; reression analysis;
cognitive perception; simulation; practical significance
\end{keywords}

\section{Introduction}\label{sec-model-introduction}

Plotting residuals is commonly regarded as a standard practice in linear
regression diagnostics \citep{belsley1980regression, cook1982residuals}.
This visual assessment plays a crucial role in identifying whether model
assumptions, such as linearity, homoscedasticity, and normality, are
reasonable. It also helps in understanding the goodness of fit and
various unexpected characteristics of the model.

Generating a residual plot in most statistical software is often as
straightforward as executing a line of code or clicking a button.
However, accurately interpreting a residual plot can be challenging. A
residual plot can exhibit various visual features, but it is crucial to
recognize that some may arise from the characteristics of predictors and
the natural stochastic variation of the observational unit, rather than
indicating a violation of model assumptions \citep{li2024plot}. Consider
Figure \ref{fig:false-finding} as an example, the residual plot displays
a triangular left-pointing shape. The distinct difference in the spread
of the residuals across the fitted values may result in the analyst
suggesting that there may be heteroskedasticity, however, it is
important to avoid over-interpreting this visual pattern. In this case,
the fitted regression model is correctly specified, and the triangular
shape is actually a result of the skewed distribution of the predictors,
rather than indicating a flaw in the model.

The concept of visual inference, as proposed by
\citet{buja2009statistical}, provides an inferential framework to assess
whether residual plots indeed contain visual patterns inconsistent with
the model assumptions. The fundamental idea involves testing whether the
true residual plot visually differs significantly from null plots, where
null plots are plotted with residuals generated from the residual
rotation distribution \citep{langsrud2005rotation}, which is a
distribution consistent with the null hypothesis \(H_0\) that the linear
regression model is correctly specified. Typically, the visual test is
accomplished through the lineup protocol, where the true residual plot
is embedded within a lineup alongside several null plots. If the true
residual plot can be distinguished from the lineup, it provides evidence
for rejecting \(H_0\).

The practice of delivering a residual plot as a lineup is generally
regarded as a valuable approach. Beyond its application in residual
diagnostics, the lineup protocol has been integrated into the analysis
of diverse subjects. For instance, Loy and Hofmann
\citetext{\citeyear{loy2013diagnostic}; \citeyear{loy2014hlmdiag}; \citeyear{loy2015you}}
illustrated its applicability in diagnosing hierarchical linear models.
Additionally, \citet{widen2016graphical} and \citet{fieberg2024using}
demonstrated its utility in geographical and ecology research
respectively, while \citet{krishnan2021hierarchical} explored its
effectiveness in forensic examinations.

A practical limitation of the lineup protocol lies in its reliance on
human judgements \citep[see][ about the practical
limitations]{li2024plot}. Unlike conventional statistical tests that can
be performed computationally in statistical software, the lineup
protocol requires human evaluation of images. This characteristic makes
it less suitable for large-scale applications, given the associated high
labour costs and time requirements. There is a substantial need to
develop an approach to substitute these human judgement with an
automated reading of data plots using machines.

The utilization of computers to interpret data plots has a rich history,
with early efforts such as ``Scagnostics'' by \citet{tukey1985computer},
a set of numerical statistics that summarize features of scatter plots.
\citet{wilkinson2005graph} expanded on this work, introducing
scagnostics based on computable measures applied to planar proximity
graphs. These measures, including, but not limited to, ``Outlying'',
``Skinny'', ``Stringy'', ``Straight'', ``Monotonic'', ``Skewed'',
``Clumpy'', and ``Striated'', aimed to characterize outliers, shape,
density, trend, coherence and other characteristics of the data. While
this approach has been inspiring, there is a recognition
\citep{buja2009statistical} that it may not capture all the necessary
visual features that differentiate true residual plots from null plots.
A more promising alternative entails enabling machines to learn the
function for extracting visual features from residual plots.
Essentially, this means empowering computers to discern the crucial
visual features for residual diagnostics and determining the method to
extract them.

Modern computer vision models are well-suited for addressing this
challenge. They rely on deep neural networks with convolutional layers
\citep{fukushima1982neocognitron}. These layers use small, sliding
windows to scan the image, performing a dot product to extract local
features and patterns. Numerous studies have demonstrated the efficacy
of convolutional layers in addressing various vision tasks, including
image recognition \citep{rawat2017deep}. Despite the widespread use of
computer vision models in fields like computer-aided diagnosis
\citep{lee2015image}, pedestrian detection \citep{brunetti2018computer},
and facial recognition \citep{emami2012facial}, their application in
reading data plots remains limited. While some studies have explored the
use of computer vision models for tasks such as reading recurrence plots
for time series regression \citep{ojeda2020multivariate}, time series
classification
\citep{chu2019automatic, hailesilassie2019financial, hatami2018classification, zhang2020encoding},
anomaly detection \citep{chen2020convolutional}, and pairwise causality
analysis \citep{singh2017deep}, the application of reading residual
plots with computer vision models is a new field of study.

In this paper, we develop computer vision models and integrate them into
the residual plots diagnostics workflow, addressing the need for an
automated visual inference. The paper is structured as follows. Section
\ref{sec-model-specifications} discusses various specifications of the
computer vision models. Section
\ref{sec-model-distance-between-residual-plots} defines the distance
measure used to detect model violations, while Section
\ref{sec-model-distance-estimation} explains how the computer vision
models estimate this distance measure. Section
\ref{sec-model-statistical-testing} covers the statistical tests based
on the estimated distance, and Section \ref{sec-model-violations-index}
introduces a Model Violations Index, which offers a quicker and more
convenient assessment. Sections \ref{sec-model-data-generation},
\ref{sec-model-architecture}, and \ref{sec-model-training} detail the
data preparation, model architecture, and training process,
respectively. The results are presented in Section
\ref{sec-model-results}. Example dataset applications are discussed in
Section \ref{sec-examples}. Finally, we conclude with a discussion of
our findings and propose ideas for future research directions.

\begin{figure}[!h]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/false-finding-1}

}

\caption{An example residual vs fitted values plot (red line indicates 0 corresponds to the x-intercept, i.e. $y=0$). The vertical spread of the data points varies with the fitted values. This often indicates the existence of heteroskedasticity, however, here the result is due to skewed distribution of the predictors rather than heteroskedasticity. The Breusch-Pagan test rejects this residual plot at 95\% significance level ($p\text{-value} = 0.046$).}\label{fig:false-finding}
\end{figure}

\section{Model Specifications}\label{sec-model-specifications}

There are various specifications of the computer vision model that can
be used to assess residual plots. We discuss these specifications below
focusing on two key components of the model formula: the input and the
output format.

\subsection{Input Formats}\label{input-formats}

Deep learning models are in general very sensitive to the input data.
The quality and relevance of the input data greatly influence the
model's capacity to generate insightful and meaningful results. There
are several designs of the input format that can be considered.

A straightforward architecture of the input layer involves feeding a
vector of residuals along with a vector of fitted values, essentially
providing all the necessary information for creating a residuals vs
fitted values plot. However, a drawback of this method is the variable
input size, which changes based on the number of observations. For
modern computer vision models implemented in mainstream software like
TensorFlow \citep{abadi2016tensorflow}, the input shape is typically
fixed. One strategy is to pad the input vectors with leading or trailing
zeros when the input tensor expects longer vectors, but it may fail if
the input vector surpasses the designed length. Alternatively, sampling
a fixed number of residual-fitted value pairs can ensure consistent
input size. By carefully designing the sampling process, it is possible
to maintain the input's representativeness while conforming to a fixed
format.

Another architecture of the input layer involves using an image as
input. The primary advantage of this design, as opposed to the vector
format, is the availability of the existing and sophisticated image
processing architectures developed over the years, such as the VGG16
architecture proposed in \citet{simonyan2014very}. These architectures
can effectively capture and summarize spatial information from nearby
pixels, which is less straightforward with vector input. Encoding data
points as pixels introduces certain trade-offs. While some detail may be
lost during the discretization process, this approach aligns with the
sampling-based strategy by focusing on summarizing the overall data
distribution. Unlike vectors, which struggle to represent spatial
relationships, image-based inputs naturally encode the joint
distribution of residuals and fitted values in a visually interpretable
format. This richer representation gives the model a more comprehensive
and intuitive understanding of the residual plot's underlying structure.
Key considerations for using image-based input include image resolution
and the aesthetics of the residual plot. In general, a higher resolution
provides more information to the model but comes at the cost of of
increased complexity and greater difficulty in training. As for the
aesthetics of the residual plot, a practical solution is to consistently
present residual plots in the same style to the model. This implies that
the model can not accept arbitrary images as input but requires the use
of the same pre-processing pipeline to convert residuals and fitted
values into a standardized-style residual plot.

Providing multiple residual plots to the model, such as a pair of plots,
a triplet or a lineup is also a possible option.
\citet{chopra2005learning} have shown that computer vision models
designed for image comparison can assess whether a pair of images are
similar or dissimilar. Applied to our specific problem, we can define
null plots of a fitted regression model to be similar to each other,
while considering true residual plots to be distinct from null plots of
any fitted regression model. A triplet constitutes a set of three
images, denoted as \(image_1\), \(image_2\) and \(image_3\). It is often
used to predict whether \(image_2\) or \(image_3\) is more similar to
\(image_1\), proving particularly useful for establishing rankings
between samples. For this setup, we can apply the same criteria to
define similarity between images. However, it is important to note that
these two approaches usually require additional considerations regarding
the loss function and, at times, non-standard training processes due to
shared weights between different convolutional blocks.

Presenting a lineup to a model aligns closely with the lineup protocol.
However, as the number of residual plots in a lineup increases, the
resolution of the input image grows rapidly, posing challenges in
training the model. We experimented with this approach in a pilot study,
but the performance of the trained model was sub-optimal.

Taking into account the implementation cost and the need for model
interpretability, we used the single residual plot input format in this
paper.

\subsection{Output Formats}\label{output-formats}

Given that the input is a single residual plot represented as a
fixed-resolution image, the output of the computer vision model can be
designed for various tasks: binary classification, categorical
classification with multiple classes, or numeric regression.

The binary outcome can represent whether the input image is consistent
with a null plot as determined by either (1) the data generating process
or (2) the result of a visual test based on human judgement. Training a
model following the latter option requires data from prior human subject
experiments, presenting difficulties in controlling the quality of data
due to variations in experimental settings across different studies.
Additionally, some visual inference experiments are unrelated to linear
regression models or residual plot diagnostics, resulting in a limited
amount of available training data.

For a multi-class outcome, the model could classify images into
categories such as ``contains outliers'' vs.~``does not contain
outliers'' or ``is non-linear'' vs.~``not non-linear,'' among other
diagnostic features. While this approach is appealing in theory, it
essentially reverts to the traditional testing framework, where tests
focus on detecting specific types of model violations in isolation.

One significant advantage of the lineup protocol for residual
diagnostics is its ability to detect multiple model violations
simultaneously, provided these violations produce visually
distinguishable differences in the plots. In contrast, training a
classification model would require explicitly defining and mapping all
potential violations of interest, or at least the most common ones,
which can be impractical. If the goal is to identify specific cases,
conventional statistical tests grounded in regression model assumptions
are often more effective. Computer vision models may be less suited to
such tasks.

Alternatively, the output could be a meaningful and interpretable
numerical measure useful for assessing residual plots, such as the
strength of suspicious visual patterns reflecting the extent of model
violations, or the difficulty index for identifying whether a residual
plot has no issues. However, these numeric measures are often informally
used in daily communication but are not typically formalized or
rigorously defined. For the purpose of training a model, this numeric
measure has to be quantifiable.

In this study, we chose to define and use a distance between a true
residual plot and a theoretically ``good'' residual plot. This is
further explained in \citet{sec-model-distance-between-residual-plots}.
\citet{vo2016localizing} have also demonstrated that defining a proper
distance between images can enhance the matching accuracy in image
search compared to a binary outcome model.

\subsection{Auxiliary Information with
Scagnostics}\label{auxiliary-information-with-scagnostics}

In Section \ref{sec-model-introduction}, we mention that scagnostics
consist of a set of manually designed visual feature extraction
functions. While our computer vision model will learn its own feature
extraction function during training, leveraging additional information
from scagnostics can enhance the model's predictive accuracy.

For each residual plot used as an input image, we calculated four
scagnostics --- ``Monotonic'', ``Sparse'', ``Splines'', and ``Striped''
-- using the \texttt{cassowaryr} R package \citep{mason2022cassowaryr}.
These computed measures, along with the number of observations from the
fitted model, were provided as the second input for the computer vision
model. We selected these scagnostics due to their reliability and
efficiency, as other scagnostics occasionally caused R process crashes
(approximately 5\% of the time) during training data preparation, due to
a bug in the \texttt{interp} R package \citep{Albrecht2023interp}.
Although the package maintainer later fixed this bug at our request, the
fix came too late to retrain the model, and additionally, their high
computational costs make them unsuitable for rapid inference, which was
a critical factor in our choice.

\section{Distance from a Theoretically ``Good'' Residual
Plot}\label{sec-model-distance-between-residual-plots}

To develop a computer vision model for assessing residual plots within
the visual inference framework, it is important to precisely define a
numerical measure of ``difference'' or ``distance'' between plots. This
distance can take the form of a basic statistical operation on pixels,
such as the sum of square differences, however, a pixel-to-pixel
comparison makes little sense in comparing residual plots where the main
interest would be structural patterns. Alternatively, it could involve
established image similarity metrics like the Structural Similarity
Index Measure \citep{wang2004image} which compares images by integrating
three perception features of an image: contrast, luminance, and
structure (related to average, standard deviation and correlation of
pixel values over a window, respectively). These image similarity
metrics are tailored for image comparison in vastly different tasks to
evaluating data plots, where only essential plot elements require
assessment \citep{chowdhury2018measuring}. We can alternatively define a
notion of distance by integrating key plot elements (instead of key
perception features like luminance, contrast, and structure), such as
those captured by scagnostics mentioned in Section
\ref{sec-model-introduction}, but the functional form still needs to be
carefully refined to accurately reflect the extent of the violations.

In this section, we introduce a distance measure between a true residual
plot and a theoretically `good' residual plot. This measure quantifies
the divergence between the residual distribution of a given fitted
regression model and that of a correctly specified model. The
computation assumes knowledge of the data generating processes for
predictors and response variables. Since these processes are often
unknown in practice, we will discuss a method to estimate this distance
using a computer vision model in Section
\ref{sec-model-distance-estimation}.

\subsection{Residual Distribution}\label{residual-distribution}

For a classical normal linear regression model,
\(\boldsymbol{y} = \boldsymbol{X}\boldsymbol{\beta} + \boldsymbol{e}\),
the residual \(\hat{\boldsymbol{e}}\) are derived as the difference of
the fitted values and observed values \(\boldsymbol{y}\). Suppose the
data generating process is known and the regression model is correctly
specified, by the Frisch-Waugh-Lowell theorem \citep{frisch1933partial},
residuals \(\hat{\boldsymbol{e}}\) can also be treated as random
variables and written as a linear transformation of the error
\(\boldsymbol{e}\) formulated as
\(\hat{\boldsymbol{e}} = \boldsymbol{R}\boldsymbol{e}\), where
\(\boldsymbol{R}=\boldsymbol{I}_n -\boldsymbol{X}(\boldsymbol{X}^\top\boldsymbol{X})^{-1}\boldsymbol{X}^\top\)
is the residual operator, \(\boldsymbol{I}_n\) is a \(n\) by \(n\)
identity matrix, and \(n\) is the number of observations.

One of the assumptions of the classical normal linear regression model
is that the error \(\boldsymbol{e}\) follows a multivariate normal
distribution with zero mean and constant variance, i.e.,
\(\boldsymbol{e} \sim N(\boldsymbol{0}_n,\sigma^2\boldsymbol{I}_n)\). It
follows that the distribution of residuals \(\hat{\boldsymbol{e}}\) can
be characterized by a certain probability distribution, denoted as
\(Q\), which is transformed from the multivariate normal distribution.
This reference distribution \(Q\) summarizes what ``good'' residuals
should follow given the design matrix \(\boldsymbol{X}\) is known and
fixed.

Suppose the design matrix \(\boldsymbol{X}\) has linearly independent
columns, the trace of the hat matrix
\(\boldsymbol{H} = \boldsymbol{X}(\boldsymbol{X}^\top\boldsymbol{X})^{-1}\boldsymbol{X}^\top\)
will equal to the number of columns in \(\boldsymbol{X}\) denoted as
\(k\). As a result, the rank of \(\boldsymbol{R}\) is \(n - k\), and
\(Q\) is a degenerate multivariate distribution. To capture the
characteristics of \(Q\), such as moments, we can simulate a large
numbers of \(\boldsymbol{\varepsilon}\) and transform it to
\(\boldsymbol{e}\) to get the empirical estimates. For simplicity, in
this study, we replaced the variance-covariance matrix of residuals
\(\text{cov}(\boldsymbol{e}, \boldsymbol{e}) = \boldsymbol{R}\sigma^2\boldsymbol{R}^\top = \boldsymbol{R}\sigma^2\)
with a full-rank diagonal matrix
\(\text{diag}(\boldsymbol{R}\sigma^2)\), where \(\text{diag}(.)\) sets
the non-diagonal entries of a matrix to zeros. The resulting
distribution for \(Q\) is
\(N(\boldsymbol{0}_n, \text{diag}(\boldsymbol{R}\sigma^2))\).

Distribution \(Q\) is derived from the correctly specified model.
However, if the model is misspecified, then the actual distribution of
residuals denoted as \(P\), will be different from \(Q\). For example,
if the data generating process contains variables correlated with any
column of \(\boldsymbol{X}\) but missing from \(\boldsymbol{X}\),
causing an omitted variable problem, \(P\) will be different from \(Q\)
because the residual operator obtained from the fitted regression model
will not be the same as \(\boldsymbol{R}\). Besides, if the
\(\boldsymbol{\varepsilon}\) follows a non-normal distribution such as a
multivariate lognormal distribution, \(P\) will usually be skewed and
has a long tail.

\subsection{\texorpdfstring{Distance of \(P\) from
\(Q\)}{Distance of P from Q}}\label{distance-of-p-from-q}

Defining a proper distance between distributions is usually easier than
defining a proper distance between data plots. Given the true residual
distribution \(Q\) and the reference residual distribution \(P\), we
used a distance measure based on Kullback-Leibler divergence
\citep{kullback1951information} to quantify the difference between two
distributions as

\begin{equation} \label{eq:kl-0}
D = \log\left(1 + D_{KL}\right),
\end{equation}

where \(D_{KL}\) is defined as

\begin{equation} \label{eq:kl-1}
D_{KL} = \int_{\mathbb{R}^{n}}\log\frac{p(\boldsymbol{e})}{q(\boldsymbol{e})}p(\boldsymbol{e})d\boldsymbol{e},
\end{equation}

\noindent and \(p(.)\) and \(q(.)\) are the probability density
functions for distribution \(P\) and distribution \(Q\), respectively.

This distance measure was first proposed in \citet{li2024plot}. It was
mainly designed for measuring the effect size of non-linearity and
heteroskedasticity in a residual plot. \citet{li2024plot} have derived
that, for a classical normal linear regression model that omits
necessary higher-order predictors \(\boldsymbol{Z}\) and the
corresponding parameter \(\boldsymbol{\beta}_z\), and incorrectly
assumes
\(\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n,\sigma^2\boldsymbol{I}_n)\)
while in fact
\(\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n, \boldsymbol{V})\)
where \(\boldsymbol{V}\) is an arbitrary symmetric positive
semi-definite matrix, \(Q\) can be represented as
\(N(\boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z, \text{diag}(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}))\).
Note that the variance-covariance matrix is replaced with the diagonal
matrix to ensure it is a full-rank matrix.

Since both \(P\) and \(Q\) are adjusted to be multivariate normal
distributions, Equation \ref{eq:kl-1} can be further expanded to

\begin{equation} \label{eq:kl-2}
D_{KL} = \frac{1}{2}\left(\log\frac{|\boldsymbol{W}|}{|\text{diag}(\boldsymbol{R}\sigma^2)|} - n + \text{tr}(\boldsymbol{W}^{-1}\text{diag}(\boldsymbol{R}\sigma^2)) + \boldsymbol{\mu}_z^\top\boldsymbol{W}^{-1}\boldsymbol{\mu}_z\right),
\end{equation}

\noindent where
\(\boldsymbol{\mu}_z = \boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z\),
and
\(\boldsymbol{W} = \text{diag}(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R})\).
The assumed error variance \(\sigma^2\) is set to be
\(\text{tr}(\boldsymbol{V})/n\), which is the expectation of the
estimated variance.

\subsection{\texorpdfstring{Non-normal
\(P\)}{Non-normal P}}\label{non-normal-p}

For non-normal error \(\boldsymbol{\varepsilon}\), the true residual
distribution \(P\) is unlikely to be a multivariate normal distribution.
Thus, Equation \ref{eq:kl-2} given in \citet{li2024plot} will not be
applicable to models violating the normality assumption.

To evaluate the Kullback-Leibler divergence of non-normal \(P\) from
\(Q\), the fallback is to solve Equation \ref{eq:kl-1} numerically.
However, since \(\boldsymbol{e}\) is a linear transformation of
non-normal random variables, it is very common that the general form of
\(P\) is unknown, meaning that we can not easily compute
\(p(\boldsymbol{e})\) using a well-known probability density function.
Additionally, even if \(p(\boldsymbol{e})\) can be calculated for any
\(\boldsymbol{e} \in \mathbb{R}^n\), it will be very difficult to do
numerical integration over the \(n\)-dimensional space, because \(n\)
could be potentially very large.

In order to approximate \(D_{KL}\) in a practically computable manner,
the elements of \(\boldsymbol{e}\) are assumed to be independent of each
other. This assumption solves both of the issues mentioned above. First,
we no longer need to integrate over \(n\) random variables. The result
of Equation \ref{eq:kl-1} is now the sum of the Kullback-Leibler
divergence evaluated for each individual residual due to the assumption
of independence between observations. Second, it is not required to know
the joint probability density \(p(\boldsymbol{e})\) any more. Instead,
the evaluation of Kullback-Leibler divergence for an individual residual
relies on the knowledge of the marginal density \(p_i(e_i)\), where
\(e_i\) is the \(i\)-th residual for \(i = 1, ..., n\). This is much
easier to approximate through simulation. It is also worth mentioning
that this independence assumption generally will not hold if
\(\text{cov}(e_i, e_j) \neq 0\) for any \(1 \leq i < j \leq n\), but its
existence is essential for reducing the computational cost.

Given \(\boldsymbol{X}\) and \(\boldsymbol{\beta}\), the algorithm for
approximating Equation \ref{eq:kl-1} starts from simulating \(m\) sets
of observed values \(\boldsymbol{y}\) according to the data generating
process. The observed values are stored in a matrix \(\boldsymbol{A}\)
with \(n\) rows and \(m\) columns, where each column of
\(\boldsymbol{A}\) is a set of observed values. Then, we can get \(m\)
sets of realized values of \(\boldsymbol{e}\) stored in the matrix
\(\boldsymbol{B}\) by applying the residual operator
\(\boldsymbol{B} = \boldsymbol{R}\boldsymbol{A}\). Furthermore, kernel
density estimation (KDE) with Gaussian kernel and optimal bandwidth
selected by the Silverman's rule of thumb \citep{silverman2018density}
is applied on each row of \(\boldsymbol{B}\) to estimate \(p_i(e_i)\)
for \(i = 1, ..., n\). The KDE computation can be done by the
\texttt{density} function in R.

Since the Kullback-Leibler divergence can be viewed as the expectation
of the log-likelihood ratio between distribution \(P\) and distribution
\(Q\) evaluated on distribution \(P\), we can reuse the simulated
residuals in matrix \(\boldsymbol{B}\) to estimate the expectation by
the sample mean. With the independence assumption, for non-normal \(P\),
\(D_{KL}\) can be approximated by

\begin{align*} \label{eq:kl-3}
D_{KL} &\approx \sum_{i = 1}^{n} \hat{D}_{KL}^{(i)}, \\
\hat{D}_{KL}^{(i)} &= \frac{1}{m}\sum_{j = 1}^{m} \log\frac{\hat{p}_i(B_{ij})}{q(B_{ij})},
\end{align*}

\noindent where \(\hat{D}_{KL}^{(i)}\) is the estimator of the
Kullback-Leibler divergence for an individual residual \(e_i\),
\(\boldsymbol{B}_{ij}\) is the \(i\)-th row and \(j\)-th column entry of
the matrix \(\boldsymbol{B}\), \(\hat{p}_i(.)\) is the kernel density
estimator of \(p_i(.)\), \(q(.)\) is the normal density function with
mean zero and an assumed variance estimated as
\(\hat{\sigma}^2 = \sum_{b \in vec(\boldsymbol{B})}(b - \sum_{b \in vec(\boldsymbol{B})} b/nm)^2/(nm - 1)\),
and \(vec(.)\) is the vectorization operator which turns a
\(n \times m\) matrix into a \(nm \times 1\) column vector by stacking
the columns of the matrix on top of each other.

\section{Distance Estimation}\label{sec-model-distance-estimation}

In the previous sections, we have defined a distance measure given in
Equation \ref{eq:kl-0} for quantifying the difference between the true
residual distribution \(P\) and an ideal reference distribution \(Q\).
However, this distance measure can only be computed when the data
generating process is known. In reality, we often have no knowledge
about the data generating process, otherwise we do not need to do a
residual diagnostic in the first place.

We use a computer vision model to estimate this distance measure for a
residual plot. Let \(D\) be the result of Equation \ref{eq:kl-0}, and
our estimator \(\hat{D}\) is formulated as

\begin{equation} \label{eq:d-approx}
\hat{D} = f_{CV}(V_{h \times w}(\boldsymbol{e}, \hat{\boldsymbol{y}})),
\end{equation}

\noindent where \(V_{h \times w}(.)\) is a plotting function that saves
a residuals vs fitted values plot with fixed aesthetic as an image with
\(h \times w\) pixels in three channels (RGB), \(f_{CV}(.)\) is a
computer vision model which takes an \(h \times w\) image as input and
predicts the distance in the domain \([0, +\infty)\).

With the estimated distance \(\hat{D}\), we can compare the underlying
distribution of the residuals to a theoretically ``good'' residual
distribution. \(\hat{D}\) can also be used as an index of the model
violations indicating the strength of the visual signal embedded in the
residual plot.

It is not expected that \(\hat{D}\) will be equal to original distance
\(D\). This is largely because information contained in a single
residual plot is limited and it may not be able to summarize all the
important characteristics of the residual distribution. For a given
residual distribution \(P\), many different residual plots can be
simulated, where many will share similar visual patterns, but some of
them could be visually very different from the rest, especially for
regression models with small \(n\). This suggests the error of the
estimation will vary depends on whether the input residual plot is
representative or not.

\section{Statistical Testing}\label{sec-model-statistical-testing}

\subsection{Lineup Evaluation}\label{sec-model-lineup-evaluation}

Theoretically, the distance \(D\) for a correctly specified model is
\(0\), because \(P\) will be the same as \(Q\). However, the computer
vision model may not necessary predict \(0\) for a null plot. Using
Figure \ref{fig:false-finding} as an example, it contains a visual
pattern which is an indication of heteroskedasticity. We would not
expect the model to be able to magically tell if the suspicious pattern
is caused by the skewed distribution of the fitted values or the
existence of heteroskedasticity. Additionally, some null plots could
have outliers or strong visual patterns due to randomness, and a
reasonable model will try to summarize those information into the
prediction, resulting in \(\hat{D} > 0\).

This property is not an issue if \(\hat{D} \gg 0\) for which the visual
signal of the residual plot is very strong, and we usually do not need
any further examination of the significance of the result. However, if
the visual pattern is weak or moderate, having \(\hat{D}\) will not be
sufficient to determine if \(H_0\) should be rejected.

To address this issue we can adhere to the paradigm of visual inference,
by comparing the estimated distance \(\hat{D}\) to the estimated
distances for the null plots in a lineup. Specifically, if a lineup
comprises 20 plots, the null hypothesis \(H_0\) will be rejected if
\(\hat{D}\) exceeds the maximum estimated distance among the \(m - 1\)
null plots, denoted as
\(\max\limits_{1 \leq i \leq m-1} {\hat{D}_{null}^{(i)}}\), where
\(\hat{D}_{null}^{(i)}\) represents the estimated distance for the
\(i\)-th null plot. This approach is equivalent to the typical lineup
protocol requiring a 95\% significance level, where \(H_0\) is rejected
if the data plot is identified as the most distinct plot by the sole
observer. The estimated distance serves as a metric to quantify the
difference between the data plot and the null plots, as intended.

Moreover, if the number of plots in a lineup, denoted by \(m\), is
sufficiently large, the empirical distribution of
\({\hat{D}_{null}^{(i)}}\) can be viewed as an approximation of the null
distribution of the estimated distance. Consequently, quantiles of the
null distribution can be estimated using the sample quantiles, and these
quantiles can be utilized for decision-making purposes. The details of
the sample quantile computation can be found in
\citet{hyndman1996sample}. For instance, if \(\hat{D}\) is greater than
or equal to the 95\% sample quantile, denoted as \(Q_{null}(0.95)\), we
can conclude that the estimated distance for the true residual plot is
significantly different from the estimated distance for null plots with
a 95\% significance level. Based on our experience, to obtain a stable
estimate of the 95\% quantile, the number of null plots, \(n_{null}\),
typically needs to be at least 100. However, if the null distribution
exhibits a long tail, a larger number of null plots may be required.
Alternatively, a \(p\)-value is the probability of observing a distance
equally or greater than \(\hat{D}\) under the null hypothesis \(H_0\),
and it can be estimated by
\(\frac{1}{m} + \frac{1}{m}\sum_{i=1}^{m-1}I\left(\hat{D}_{null}^{(i)} \geq \hat{D}\right)\).

To alleviate computation burden, a lattice of quantiles for \(\hat{D}\)
under \(H_0\) with specified sample sizes can be precomputed. We can
then map the \(\hat{D}\) and sample size to the closet quantile and
sample size in lattice to calculate the corresponding \(p\)-value. This
approach lose precision in \(p\)-value calculation, however,
significantly improves computational efficiency.

\subsection{Bootstrapping}\label{bootstrapping}

Bootstrap is often employed in linear regression when conducting
inference for estimated parameters \citep[see][ and
\citet{efron1994introduction}]{davison1997bootstrap}. It is typically
done by sampling individual cases with replacement and refitting the
regression model. If the observed data accurately reflects the true
distribution of the population, the bootstrapped estimates can be used
to measure the variability of the parameter estimate without making
strong distributional assumptions about the data generating process.

Similarly, bootstrap can be applied on the estimated distance
\(\hat{D}\). For each refitted model \(M_{boot}^{(i)}\), there will be
an associated residual plot \(V_{boot}^{(i)}\) which can be fed into the
computer vision model to obtain \(\hat{D}_{boot}^{(i)}\), where
\(i = 1,...,n_{boot}\), and \(n_{boot}\) is the number of bootstrapped
samples. If we are interested in the variation of \(\hat{D}\), we can
use \(\hat{D}_{boot}^{(i)}\) to estimate a confidence interval.

Alternatively, since each \(M_{boot}^{(i)}\) has a set of estimated
coefficients \(\hat{\boldsymbol{\beta}}_{boot}^{(i)}\) and an estimated
variance \(\hat{\sigma^2}_{boot}^{(i)}\), a new approximated null
distribution can be construed and the corresponding 95\% sample quantile
\(Q_{boot}^{(i)}(0.95)\) can be computed. Then, if
\(\hat{D}_{boot}^{(i)} \geq Q_{boot}^{(i)}(0.95)\), \(H_0\) will be
rejected for \(M_{boot}^{(i)}\). The ratio of rejected
\(M_{boot}^{(i)}\) among all the refitted models provides an indication
of how often the assumed regression model are considered to be incorrect
if the data can be obtained repetitively from the same data generating
process. But this approach is computationally very expensive because it
requires \(n_{boot} \times n_{null}\) times of residual plot assessment.
In practice, \(Q_{null}(0.95)\) can be used to replace
\(Q_{boot}^{(i)}(0.95)\) in the computation.

\section{Model Violations Index}\label{sec-model-violations-index}

While statistical testing is a powerful tool for detecting model
violations, it can become cumbersome and time-consuming when quick
decisions are needed, particularly due to the need to evaluate numerous
null plots. In practice, a more convenient and immediate method for
assessing model performance is often required. This is where an index,
such as the Model Violations Index (MVI), becomes valuable. It offers a
straightforward way to quantify deviations from model assumptions,
enabling rapid assessment and easier comparison across models.

The estimator \(\hat{D}\) measures the difference between the true
residual distribution and the reference residual distribution, a
difference primarily arises from deviations in model assumptions. The
magnitude of \(D\) directly reflects the degree of these deviations,
thus making \(\hat{D}\) instrumental in forming a model violations index
(MVI).

Note that if more observations are used for estimating the linear
regression, the result of Equation \ref{eq:kl-1} will increase, as the
integration will be performed with larger \(n\). For a given data
generating process, \(D\) typically increases logarithmically with the
number of observations. This behavior comes from the relationship
\(D = \text{log}(1 + D_{KL})\), where
\(D_{KL} = \sum_{i=1}^{n}D_{KL}^{(i)}\) under the assumption of
independence.

Since \(\hat{D}\) is an estimate of \(D\), it is expected that a larger
number of observations will also lead to a higher \(\hat{D}\). However,
this does not imply that \(\hat{D}\) fails to accurately represent the
extent of model violations. In fact, when examining residual plots with
more observations, we often observe a stronger visual signal strength,
as the underlying patterns are more likely to be revealed, except in
cases of significant overlapping.

Therefore, the Model Violations Index (MVI) can be proposed as

\begin{equation} \label{eq:mvi}
\text{MVI} = C + \hat{D} - \log(n),
\end{equation}

\noindent where \(C\) is a large enough constant keeping the result
positive and the term \(-\log(n)\) is used to offset the increase in
\(D\) due to sample size.

Figure \ref{fig:poly-heter-index} displays the residual plots for fitted
models exhibiting varying degrees of non-linearity and
heteroskedasticity. Each residual plot's MVI is computed using Equation
\ref{eq:mvi} with \(C = 10\). When \(\text{MVI} > 8\), the visual
patterns are notably strong and easily discernible by humans. In the
range \(6 < \text{MVI} < 8\), the visibility of the visual pattern
diminishes as MVI decreases. Conversely, when \(\text{MVI} < 6\), the
visual pattern tends to become relatively faint and challenging to
observe. Table \ref{tab:mvi} provides a summary of the MVI usage and it
is applicable to other linear regression models.

\begin{table}

\caption{\label{tab:mvi}Degree of model violations or the strength of the visual signals according to the Model Violations Index (MVI). The constant $C$ is set to be 10.}
\centering
\begin{tabular}[t]{lc}
\toprule
Degree of model violations & Range ($C$ = 10)\\
\midrule
Strong & $\text{MVI} > 8$\\
Moderate & $6 < \text{MVI} < 8$\\
Weak & $\text{MVI} < 6$\\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}[!h]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/poly-heter-index-1}

}

\caption{Residual plots generated from fitted models exhibiting varying degrees of (A) non-linearity and (B) heteroskedasticity violations. The model violations index (MVI) is displayed atop each residual plot. The non-linearity patterns are relatively strong for $MVI > 8$, and relatively weak for $MVI < 6$, while the heteroskedasticity patterns are relatively strong for $MVI > 8$, and relatively weak for $MVI < 6$.}\label{fig:poly-heter-index}
\end{figure}

\section{Data Generation}\label{sec-model-data-generation}

\subsection{Simulation Scheme}\label{simulation-scheme}

While observational data is frequently employed in training models for
real-world applications, the data generating process of observational
data often remains unknown, making computation for our target variable
\(D\) unattainable. Consequently, the computer vision models developed
in this study were trained using synthetic data, including 80,000
training images and 8,000 test images. This approach provided us with
precise label annotations. Additionally, it ensured a large and diverse
training dataset, as we had control over the data generating process,
and the simulation of the training data was relatively cost-effective.

We have incorporated three types of residual departures of linear
regression model in the training data, including non-linearity,
heteroskedasticity and non-normality. All three departures can be
summarized by the data generating process formulated as

\begin{align} \label{eq:data-sim}
\boldsymbol{y} &= \boldsymbol{1}_n + \boldsymbol{x}_1 + \beta_1\boldsymbol{x}_2 + \beta_2(\boldsymbol{z} + \beta_1\boldsymbol{w}) + \boldsymbol{k} \odot \boldsymbol{\varepsilon}, \\
\boldsymbol{z} &= \text{He}_j(g(\boldsymbol{x}_1, 2)), \\
\boldsymbol{w} &= \text{He}_j(g(\boldsymbol{x}_2, 2)), \\
\boldsymbol{k} &= \left[\boldsymbol{1}_n + b(2 - |a|)(\boldsymbol{x}_1 + \beta_1\boldsymbol{x}_2 - a\boldsymbol{1}_n)^{\circ2}\right]^{\circ1/2},
\end{align}

\noindent where \(\boldsymbol{y}\), \(\boldsymbol{x}_1\),
\(\boldsymbol{x}_2\), \(\boldsymbol{z}\), \(\boldsymbol{w}\),
\(\boldsymbol{k}\) and \(\boldsymbol{\varepsilon}\) are vectors of size
\(n\), \(\boldsymbol{1}_n\) is a vector of ones of size \(n\),
\(\boldsymbol{x}_1\) and \(\boldsymbol{x}_2\) are two independent
predictors, \(\text{He}_j(.)\) is the \(j\)th-order probabilist's
Hermite polynomials \citep{hermite1864nouveau}, \((.)^{\circ2}\) and
\((.)^{\circ1/2}\) are Hadamard square and square root, \(\odot\) is the
Hadamard product, and \(g(\boldsymbol{x}, k)\) is a scaling function to
enforce the support of the random vector to be \([-k, k]^n\) defined as

\[g(\boldsymbol{x}, k) = 2k \cdot \frac{\boldsymbol{x} - x_{\min}\boldsymbol{1}_n}{x_{\max} - x_{\min}} - k\boldsymbol{1}_n,~for~k > 0,\]
\noindent where \(x_{\min} = \underset{i \in \{ 1,...,n\}}{\min} x_i\),
\(x_{\max} = \underset{i \in \{ 1,...,n\}}{\max} x_i\) and \(x_i\) is
the \(i\)-th entry of \(\boldsymbol{x}\).

\begin{table}

\caption{\label{tab:factor}Factors used in the data generating process for synthetic data simulation. Factor $j$ and $a$ controls the non-linearity shape and the heteroskedasticity shape respectively. Factor $b$, $\sigma_\varepsilon$ and $n$ control the signal strength. Factor $\text{dist}_\varepsilon$, $\text{dist}_{x1}$ and $\text{dist}_{x2}$ specifies the distribution of $\varepsilon$, $X_1$ and $X_2$ respectively.}
\centering
\begin{tabular}[t]{ll}
\toprule
Factor & Domain\\
\midrule
j & \{2, 3, ..., 18\}\\
a & {}[-1, 1]\\
b & {}[0, 100]\\
$\beta_1$ & \{0, 1\}\\
$\beta_2$ & \{0, 1\}\\
\addlinespace
$\text{dist}_\varepsilon$ & \{discrete, uniform, normal, lognormal\}\\
$\text{dist}_{x1}$ & \{discrete, uniform, normal, lognormal\}\\
$\text{dist}_{x2}$ & \{discrete, uniform, normal, lognormal\}\\
$\sigma_{\varepsilon}$ & {}[0.0625, 9]\\
$\sigma_{X1}$ & {}[0.3, 0.6]\\
\addlinespace
$\sigma_{X2}$ & {}[0.3, 0.6]\\
n & {}[50, 500]\\
\bottomrule
\end{tabular}
\end{table}

The residuals and fitted values of the fitted model were obtained by
regressing \(\boldsymbol{y}\) on \(\boldsymbol{x}_1\). If
\(\beta_1 \neq 0\), \(\boldsymbol{x}_2\) was also included in the design
matrix. This data generation process was adapted from
\citet{li2024plot}, where it was utilized to simulate residual plots
exhibiting non-linearity and heteroskedasticity visual patterns for
human subject experiments. A summary of the factors utilized in Equation
\ref{eq:data-sim} is provided in Table \ref{tab:factor}.

In Equation \ref{eq:data-sim}, \(\boldsymbol{z}\) and \(\boldsymbol{w}\)
represent higher-order terms of \(\boldsymbol{x}_1\) and
\(\boldsymbol{x}_2\), respectively. If \(\beta_2 \neq 0\), the
regression model will encounter non-linearity issues. Parameter \(j\)
serves as a shape parameter that controls the number of tuning points in
the non-linear pattern. Typically, higher values of \(j\) lead to an
increase in the number of tuning points, as illustrated in Figure
\ref{fig:different-j}.

\begin{figure}[!h]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/different-j-1}

}

\caption{Non-linearity forms generated for the synthetic data simulation. The 17 shapes are generated by varying the order of polynomial given by $j$ in $He_j(.)$.}\label{fig:different-j}
\end{figure}

Additionally, scaling factor \(\boldsymbol{k}\) directly affects the
error distribution and it is correlated with \(\boldsymbol{x}_1\) and
\(\boldsymbol{x}_2\). If \(b \neq 0\) and
\(\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n, \sigma^2\boldsymbol{I}_n)\),
the constant variance assumption will be violated. Parameter \(a\) is a
shape parameter controlling the location of the smallest variance in a
residual plot as shown in Figure \ref{fig:different-a}.

\begin{figure}[!h]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/different-a-1}

}

\caption{Heteroskedasticity forms generated for the synthetic data simulation. Different shapes are controlled by the continuous factor $a$ between -1 and 1. For $a = -1$, the residual plot exhibits a "left-triangle" shape. And for $a = 1$, the residual plot exhibits a "right-triangle" shape. }\label{fig:different-a}
\end{figure}

\begin{figure}[!h]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/different-e-1}

}

\caption{Non-normality forms generated for the synthetic data simulation. Four different error distributions including discrete, lognormal, normal and uniform are considered.}\label{fig:different-e}
\end{figure}

Non-normality violations arise from specifying a non-normal distribution
for \(\boldsymbol{\varepsilon}\). In the synthetic data simulation, four
distinct error distributions are considered, including discrete,
uniform, normal, and lognormal distributions, as presented in Figure
\ref{fig:different-e}. Each distribution imparts unique characteristics
in the residual plot. The discrete error distribution introduces
discrete clusters in residuals, while the lognormal distribution
typically yields outliers. Uniform error distribution may result in
residuals filling the entire space of the residual plot. All of these
distributions exhibit visual distinctions from the normal error
distribution.

\begin{figure}[!h]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/different-j-x2-1}

}

\caption{Residual plots of multiple linear regression models with non-linearity issues. The 17 shapes are generated by varying the order of polynomial given by $j$ in $He_j(.)$. A second predictor $\boldsymbol{x}_2$ is introduced to the regression model to create complex shapes.}\label{fig:different-j-x2}
\end{figure}

\begin{figure}[!h]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/different-j-heter-1}

}

\caption{Residual plots of models violating both the non-linearity and the heteroskedasticity assumptions. The 17 shapes are generated by varying the order of polynomial given by $j$ in $He_j(.)$, and the "left-triangle" shape is introduced by setting $a = -1$.}\label{fig:different-j-heter}
\end{figure}

\begin{figure}[!h]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/different-e-heter-1}

}

\caption{Residual plots of models violating both the non-normality and the heteroskedasticity assumptions. The four shapes are generated by using four different error distributions including discrete, lognormal, normal and uniform, and the "left-triangle" shape is introduced by setting $a = -1$. }\label{fig:different-e-heter}
\end{figure}

Equation \ref{eq:data-sim} accommodates the incorporation of the second
predictor \(\boldsymbol{x}_2\). Introducing it into the data generation
process by setting \(\beta_1 = 1\) significantly enhances the complexity
of the shapes, as illustrated in Figure \ref{fig:different-j-x2}. In
comparison to Figure \ref{fig:different-j}, Figure
\ref{fig:different-j-x2} demonstrates that the non-linear shape
resembles a surface rather than a single curve. This augmentation can
facilitate the computer vision model in learning visual patterns from
residual plots of the multiple linear regression model.

In real-world analysis, it is not uncommon to encounter instances where
multiple model violations coexist. In such cases, the residual plots
often exhibit a mixed pattern of visual anomalies corresponding to
different types of model violations. Figure \ref{fig:different-j-heter}
and Figure \ref{fig:different-e-heter} show the visual patterns of
models with multiple model violations.

The predictors, \(\boldsymbol{x}_1\) and \(\boldsymbol{x}_2\), are
randomly generated from four distinct distributions, including
\(U(-1, 1)\) (uniform), \(N(0, 0.3^2)\) (normal),
\(\text{lognormal}(0, 0.6^2)/3\) (skewed) and \(U\{-1, 1\}\) (discrete
uniform).

\subsection{Balanced Dataset}\label{balanced-dataset}

To train a robust computer vision model, we deliberately controlled the
distribution of the target variable \(D\) in the training data. We
ensured that it followed a uniform distribution between \(0\) and \(7\).
This was achieved by organizing \(50\) buckets, each exclusively
accepting training samples with \(D\) falling within the range
\([7(i - 1)/49, 7i/49)\) for \(i < 50\), where \(i\) represents the
index of the \(i\)-th bucket. For the \(50\)-th bucket, any training
samples with \(D \geq 7\) were accepted.

With 80,000 training images prepared, each bucket accommodated a maximum
of \(80000/ 50 = 1600\) training samples. The simulator iteratively
sampled parameter values from the parameter space, generated residuals
and fitted values using the data generation process, computed the
distance, and checked if the sample fitted within the corresponding
bucket. This process continued until all buckets were filled.

Similarly, we adopted the same methodology to prepare 8,000 test images
for performance evaluation and model diagnostics.

\section{Model Architecture}\label{sec-model-architecture}

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/cnn-diag-1}

}

\caption{Diagram of the architecture of the optimized computer vision model. Numbers at the bottom of each box show the shape of the output of each layer. The band of each box drawn in a darker color indicates the use of the rectified linear unit activation function.  Yellow boxes are 2D convolutional layers, orange boxes are pooling layers, the grey box is the concatenation layer, and the purple boxes are dense layers.}\label{fig:cnn-diag}
\end{figure}

The architecture of the computer vision model is adapted from the
well-established VGG16 architecture \citep{simonyan2014very}. While more
recent architectures like ResNet \citep{he2016deep} and
DenseNet\citep{huang2017densely}, have achieved even greater
performance, VGG16 remains a solid choice for many applications due to
its simplicity and effectiveness. Our decision to use VGG16 aligns with
our goal of starting with a proven and straightforward model. Figure
\ref{fig:cnn-diag} provides a diagram of the architecture. More details
about the neural network layers used in this study are provided in
Appendix A.

The model begins with an input layer of shape
\(n \times h \times w \times 3\), capable of handling \(n\) RGB images.
This is followed by a grayscale conversion layer utilizing the luma
formula under the Rec. 601 standard \citep{series2011studio}, which
converts the color image to grayscale. Grayscale suffices for our task
since data points are plotted in black. We experiment with three
combinations of \(h\) and \(w\): \(32 \times 32\), \(64 \times 64\), and
\(128 \times 128\), aiming to achieve sufficiently high image resolution
for the problem at hand.

The processed image is used as the input for the first convolutional
block. The model comprises at most five consecutive convolutional
blocks, mirroring the original VGG16 architecture. Within each block,
there are two 2D convolutional layers followed by two activation layers,
respectively. Subsequently, a 2D max-pooling layer follows the second
activation layer. The 2D convolutional layer convolves the input with a
fixed number of \(3 \times 3\) convolution filters, while the 2D
max-pooling layer downsamples the input along its spatial dimensions by
taking the maximum value over a \(2 \times 2\) window for each channel
of the input. The activation layer employs the rectified linear unit
(ReLU) activation function, a standard practice in deep learning, which
introduces a non-linear transformation of the output of the 2D
convolutional layer. Additionally, to regularize training, a batch
normalization layer is added after each 2D convolutional layer and
before the activation layer. Finally, a dropout layer is appended at the
end of each convolutional block to randomly set some inputs to zero
during training, further aiding in regularization.

The output of the last convolutional block is summarized by either a
global max pooling layer or a global average pooling layer, resulting in
a two-dimensional tensor. To leverage the information contained in
scagnostics, this tensor is concatenated with an additional
\(n \times 5\) tensor, which contains the ``Monotonic'', ``Sparse'',
``Splines'', and ``Striped'' measures, along with the number of
observations for \(n\) residual plots.

The concatenated tensor is then fed into the final prediction block.
This block consists of two fully-connected layers. The first layer
contains at least \(128\) units, followed by a dropout layer.
Occasionally, a batch normalization layer is inserted between the
fully-connected layer and the dropout layer for regularization purposes.
The second fully-connected layer consists of only one unit, serving as
the output of the model.

The model weights \(\boldsymbol{\theta}\) were randomly initialized and
they were optimized by the Adam optimizer \citep{kingma2014adam} with
the mean square error loss function

\[\hat{\boldsymbol{\theta}} = \underset{\boldsymbol{\theta}}{\text{arg min}}\frac{1}{n_{\text{train}}}\sum_{i=1}^{n_{\text{train}}}(D_i - f_{\boldsymbol{\theta}}(V_i, S_i))^2,\]

\noindent where \(n_{\text{train}}\) is the number of training samples,
\(V_i\) is the \(i\)-th residual plot and \(S_i\) is the additional
information about the \(i\)-th residual plot including four scagnostics
and the number of observations.

\section{Model Training}\label{sec-model-training}

To achieve a near-optimal deep learning model, hyperparameters like the
learning rate often need to be fine-tuned using a tuner. In our study,
we utilized the Bayesian optimization tuner from the \texttt{KerasTuner}
Python library \citep{omalley2019kerastuner} for this purpose. A
comprehensive list of hyperparameters is provided in Table
\ref{tab:hyperparameter}.

The number of base filters determines the number of filters for the
first 2D convolutional layer. In the VGG16 architecture, the number of
filters for the 2D convolutional layer in a block is typically twice the
number in the previous block, except for the last block, which maintains
the same number of convolution filters as the previous one. This
hyperparameter aids in controlling the complexity of the computer vision
model. A higher number of base filters results in more trainable
parameters. Likewise, the number of units for the fully-connected layer
determines the complexity of the final prediction block. Increasing the
number of units enhances model complexity, resulting in more trainable
parameters.

The dropout rate and batch normalization are flexible hyperparameters
that work in conjunction with other parameters to facilitate smooth
training. A higher dropout rate is necessary when the model tends to
overfit the training dataset by learning too much noise
\citep{srivastava2014dropout}. Conversely, a lower dropout rate is
preferred when the model is complex and challenging to converge. Batch
normalization, on the other hand, addresses the internal covariate shift
problem arising from the randomness in weight initialization and input
data \citep{goodfellow2016deep}. It helps stabilize and accelerate the
training process by normalizing the activations of each layer.

Additionally, incorporating additional inputs such as scagnostics and
the number of observations can potentially enhance prediction accuracy.
Therefore, we allow the tuner to determine whether these inputs were
necessary for optimal model performance.

The learning rate is a crucial hyperparameter, as it dictates the step
size of the optimization algorithm. A high learning rate can help the
model avoid local minima but risks overshooting and missing the global
minimum. Conversely, a low learning rate smoothens the training process
but makes the convergence time longer and increases the likelihood of
getting trapped in local minima.

Our model was trained on the MASSIVE M3 high-performance computing
platform \citep{goscinski2014multi}, using TensorFlow
\citep{abadi2016tensorflow} and Keras \citep{chollet2015keras}. During
training, 80\% of the training data was utilized for actual training,
while the remaining 20\% was used as validation data. The Bayesian
optimization tuner conducted 100 trials to identify the best
hyperparameter values based on validation root mean square error. The
tuner then restored the best epoch of the best model from the trials.
Additionally, we applied early stopping, terminating the training
process if the validation root mean square error fails to improve for 50
epochs. The maximum allowed epochs was set at 2,000, although no models
reached this threshold.

\begin{table}

\caption{\label{tab:hyperparameter}Name of hyperparameters and their correspoding domain for the computer vision model.}
\centering
\begin{tabular}[t]{ll}
\toprule
Hyperparameter & Domain\\
\midrule
Number of base filters & \{4, 8, 16, 32, 64\}\\
Dropout rate for convolutional blocks & {}[0.1, 0.6]\\
Batch normalization for convolutional blocks & \{false, true\}\\
Type of global pooling & \{max, average\}\\
Ignore additional inputs & \{false, true\}\\
\addlinespace
Number of units for the fully-connected layer & \{128, 256, 512, 1024, 2048\}\\
Batch normalization for the fully-connected layer & \{false, true\}\\
Dropout rate for the fully-connected layer & {}[0.1, 0.6]\\
Learning rate & {}[$10^{-8}$, $10^{-1}$]\\
\bottomrule
\end{tabular}
\end{table}

Based on the tuning process described above, the optimized
hyperparameter values are presented in Table
\ref{tab:best-hyperparameter}. It was observable that a minimum of
\(32\) base filters was necessary, with the preferable choice being
\(64\) base filters for both the \(64 \times 64\) and \(128 \times 128\)
models, mirroring the original VGG16 architecture. The optimized dropout
rate for convolutional blocks hovered around \(0.4\), and incorporating
batch normalization for convolutional blocks proved beneficial for
performance.

All optimized models chose to retain the additional inputs, contributing
to the reduction of validation error. The number of units required for
the fully-connected layer was \(256\), a relatively modest number
compared to the VGG16 classifier, suggesting that the problem at hand
was less complex. The optimized learning rates were higher for models
with higher resolution input, likely because models with more parameters
are more prone to getting stuck in local minima, requiring a higher
learning rate.

\begin{table}
\centering
\caption{\label{tab:best-hyperparameter}Hyperparameters values for the optimized computer vision models with different input sizes.}
\centering
\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
\begin{tabular}[t]{llll}
\toprule
Hyperparameter & $32 \times 32$ & $64 \times 64$ & $128 \times 128$\\
\midrule
Number of base filters & 32 & 64 & 64\\
Dropout rate for convolutional blocks & 0.4 & 0.3 & 0.4\\
Batch normalization for convolutional blocks & true & true & true\\
Type of global pooling & max & average & average\\
Ignore additional inputs & false & false & false\\
\addlinespace
Number of units for the fully-connected layer & 256 & 256 & 256\\
Batch normalization for the fully-connected layer & false & true & true\\
Dropout rate for the fully-connected layer & 0.2 & 0.4 & 0.1\\
Learning rate & 0.0003 & 0.0006 & 0.0052\\
\bottomrule
\end{tabular}}
\end{table}

\section{Results}\label{sec-model-results}

\subsection{Model Performance}\label{model-performance}

The test performance for the optimized models with three different input
sizes are summarized in Table \ref{tab:performance}. Among these models,
the \(32 \times 32\) model consistently exhibited the best test
performance. The mean absolute error of the \(32 \times 32\) model
indicated that the difference between \(\hat{D}\) and \(D\) was
approximately \(0.43\) on the test set, a negligible deviation
considering the normal range of \(D\) typically falls between \(0\) and
\(7\). The high \(R^2\) values also suggested that the predictions were
largely linearly correlated with the target.

Figure \ref{fig:model-performance} presents a hexagonal heatmap for
\(D - \hat{D}\) versus \(D\). The brown smoothing curves, fitted by
generalized additive models \citep{hastie2017generalized}, demonstrate
that all the optimized models perform admirably on the test sets when
\(1.5 < D < 6\), where no structural issues are noticeable. However,
over-predictions occurred when \(D < 1.5\), while under-predictions
occurred predominantly when \(\hat{D} > 6\).

For input images representing null plots where \(D = 0\), it was
expected that the models will over-predict the distance, as explained in
Section \ref{sec-model-lineup-evaluation}. However, it can not explain
the under-prediction issue. Therefore, we analysed the relationship
between residuals and all the factors involved in the data generating
process. We found that most issues actually arose from non-linearity
problems and the presence of a second predictor in the regression model
as illustrated in Figure \ref{fig:over-under}. When the variance for the
error distribution was small, the optimized model tended to
under-predict the distance. Conversely, when the error distribution had
a large variance, the model tended to over-predict the distance.

Since most of the deviation stemmed from the presence of non-linearity
violations, to further investigate this, we split the test set based on
violation types and re-evaluated the performance, as detailed in Table
\ref{tab:performance-sub}. It was evident that metrics for null plots
were notably worse compared to other categories. Furthermore, residual
plots solely exhibiting non-normality issues were the easiest to
predict, with very low test root mean square error (RMSE) at around
\(0.3\). Residual plots with non-linearity issues were more challenging
to assess than those with heteroskedasticity or non-normality issues.
When multiple violations were introduced to a residual plot, the
performance metrics typically lay between the metrics for each
individual violation.

Based on the model performance metrics, we chose to use the
best-performing model evaluated on the test set, namely the
\(32 \times 32\) model, for the subsequent analysis.

\begin{table}

\caption{\label{tab:performance}The test performance of three optimized models with different input sizes.}
\centering
\begin{tabular}[t]{lrrrr}
\toprule
 & RMSE & $R^2$ & MAE & Huber loss\\
\midrule
$32 \times 32$ & 0.660 & 0.901 & 0.434 & 0.18\\
$64 \times 64$ & 0.674 & 0.897 & 0.438 & 0.19\\
$128 \times 128$ & 0.692 & 0.892 & 0.460 & 0.20\\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}[!h]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/model-performance-1}

}

\caption{Hexagonal heatmap for difference in $D$ and $\hat{D}$ vs $D$ on test data for three optimized models with different input sizes. The brown lines are smoothing curves produced by fitting generalized additive models. The area over the zero line in light yellow indicates under-prediction, and the area under the zero line in light green indicates over-prediction.}\label{fig:model-performance}
\end{figure}

\begin{figure}[!h]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/over-under-1}

}

\caption{Scatter plots for difference in $D$ and $\hat{D}$ vs $\sigma$ on test data for the $32 \times 32$ optimized model. The data is grouped by whether the regression has only non-linearity violation, and whether it includes a second predictor in the regression formula. The brown lines are smoothing curves produced by fitting generalized additive models. The area over the zero line in light yellow indicates under-prediction, and the area under the zero line in light green indicates over-prediction.}\label{fig:over-under}
\end{figure}

\begin{table}

\caption{\label{tab:performance-sub}The training and test performance of the $32 \times 32$ model presented with different model violations.}
\centering
\begin{tabular}[t]{lrr}
\toprule
Violations & \#samples & RMSE\\
\midrule
no violations & 155 & 1.267\\
non-linearity & 2218 & 0.787\\
heteroskedasticity & 1067 & 0.602\\
non-linearity + heteroskedasticity & 985 & 0.751\\
non-normality & 1111 & 0.320\\
non-linearity + non-normality & 928 & 0.600\\
heteroskedasticity + non-normality & 819 & 0.489\\
non-linearity + heteroskedasticity + non-normality & 717 & 0.620\\
\bottomrule
\end{tabular}
\end{table}

\subsection{Comparison with Human Visual Inference and Conventional
Tests}\label{comparison-with-human-visual-inference-and-conventional-tests}

\subsubsection{Overview of the Human Subject
Experiment}\label{overview-of-the-human-subject-experiment}

In order to check the validity of the proposed computer vision model,
residual plots presented in the human subject experiment conducted by
\citet{li2024plot} will be assessed.

This study has collected 7,974 human responses to 1,152 lineups. Each
lineup contains one randomly placed true residual plot and 19 null
plots. Among the 1,152 lineups, 24 are attention check lineups in which
the visual patterns are designed to be extremely obvious and very
different from the corresponding to null plots, 36 are null lineups
where all the lineups consist of only null plots, 279 are lineups with
uniform predictor distribution evaluated by 11 participants, and the
remaining 813 are lineups with discrete, skewed or normal predictor
distribution evaluated by 5 participants. Attention check lineups and
null lineups will not be assessed in the following analysis.

In \citet{li2024plot}, the residual plots are simulated from a data
generating process which is a special case of Equation
\ref{eq:data-sim}. The main characteristic is the model violations are
introduced separately, meaning non-linearity and heteroskedasticity will
not co-exist in one lineup but assigned uniformly to all lineups.
Additionally, non-normality and multiple predictors are not considered
in the experimental design.

\subsubsection{Model Performance on the Human-evaluated
Data}\label{model-performance-on-the-human-evaluated-data}

\begin{table}

\caption{\label{tab:experiment-performance}The performance of the $32 \times 32$ model on the data used in the human subject experiment.}
\centering
\begin{tabular}[t]{lrrrr}
\toprule
Violation & RMSE & $R^2$ & MAE & Huber loss\\
\midrule
heteroskedasticity & 0.721 & 0.852 & 0.553 & 0.235\\
non-linearity & 0.738 & 0.770 & 0.566 & 0.246\\
\bottomrule
\end{tabular}
\end{table}

For each lineup used in \citet{li2024plot}, there is one true residual
plot and 19 null plots. While the distance \(D\) for the true residual
plot depends on the underlying data generating process, the distance
\(D\) for the null plots is zero. We have used our optimized computer
vision model to estimate distance for both the true residual plots and
the null plots. To have a fair comparison, \(H_0\) will be rejected if
the true residual plot has the greatest estimated distance among all
plots in a lineup. Additionally, the appropriate conventional tests
including the Ramsey Regression Equation Specification Error Test
(RESET) \citep{ramsey1969tests} for non-linearity and the Breusch-Pagan
test \citep{breusch1979simple} for heteroskedasticity were applied on
the same data for comparison.

The performance metrics of \(\hat{D}\) for true residual plots are
outlined in Table \ref{tab:experiment-performance}. It is notable that
all performance metrics are slightly worse than those evaluated on the
test data. Nevertheless, the mean absolute error remains at a low level,
and the linear correlation between the prediction and the true value
remains very high. Consistent with results in Table
\ref{tab:performance-sub}, lineups with non-linearity issues are more
challenging to predict than those with heteroskedasticity issues.

Table \ref{tab:human-conv-table} provides a summary of the agreement
between decisions made by the computer vision model and conventional
tests. The agreement rates between conventional tests and the computer
vision model are 85.95\% and 79.69\% for residual plots containing
heteroskedasticity and non-linearity patterns, respectively. These
figures are higher than those calculated for visual tests conducted by
human, indicating that the computer vision model exhibits behavior more
akin to the best available conventional tests. However, Figure
\ref{fig:conv-mosaic} shows that the computer vision model does not
always reject when the conventional tests reject. And a small number of
plots will be rejected by computer vision model but not by conventional
tests. This suggests that conventional tests are more sensitive than the
computer vision model.

Figure \ref{fig:pcp} further illustrates the decisions made by visual
tests conducted by human, computer vision models, and conventional
tests, using a parallel coordinate plots. It can be observed that all
three tests will agree with each other for around 50\% of the cases.
When visual tests conducted by human do not reject, there are
substantial amount of cases where computer vision model also do not
reject but conventional tests reject. There are much fewer cases that do
not reject by visual tests and conventional tests, but is rejected by
computer vision models. This indicates computer vision model can behave
like visual tests conducted by human better than conventional tests.
Moreover, there are great proportion of cases where visual tests
conducted by human is the only test who does not reject.

When plotting the decision against the distance, as illustrated in
Figure \ref{fig:power}, several notable observations emerge. Firstly,
compared to conventional tests, the computer vision model tends to have
fewer rejected cases when \(D < 2\) and fewer non-rejected cases when
\(2< D < 4\). This suggests tests based on the computer vision model are
less sensitive to small deviations from model assumptions than
conventional tests but more sensitive to moderate deviations.
Additionally, visual tests demonstrate the lowest sensitivity to
residual plots with small distances where not many residual plots are
rejected when \(D < 2\). Similarly, for large distances where \(D > 4\),
almost all residual plots are rejected by the computer vision model and
conventional tests, but for visual tests conducted by humans, the
threshold is higher with \(D > 5\).

In Figure \ref{fig:power}, rejection decisions are fitted by logistic
regression models with no intercept terms and an offset equals to
\(\text{log}(0.05/0.95)\). The fitted curves for the computer vision
model fall between those of conventional tests and visual tests for both
non-linearity and heteroskedasticity, which means there is still
potential to refine the computer vision model to better align its
behavior with visual tests conducted by humans.

In the experiment conducted in \citet{li2024plot}, participants were
allowed to make multiple selections for a lineup. The weighted detection
rate was computed by assigning weights to each detection. If the
participant selected zero plots, a weight of 0.05 was assigned;
otherwise, if the true residual plot was detected, the weight was 1
divided by the number of selections. This weighted detection rate allow
us to assess the quality of the distance measure purposed in this paper,
by using the \(\delta\)-difference statistic. The \(\delta\)-difference
is originally defined by \citet{chowdhury2018measuring}, is given by

\[
\delta = \bar{d}_{\text{true}} - \underset{j}{\text{max}}\left(\bar{d}_{\text{null}}^{(j)}\right) \quad \text{for}~j = 1,...,m-1,
\]

where \(\bar{d}_{\text{null}}^{(j)}\) is the mean distance between the
\(j\)-th null plot and the other null plots, \(\bar{d}_{\text{true}}\)
is the mean distance between the true residual plot and null plots, and
\(m\) is the number of plots in a lineup. These mean distances are used
because, as noted by \citet{chowdhury2018measuring}, the distances can
vary depending on which data plot is used for comparison. For instance,
with three null plots, A, B and C, the distance between A and B may
differ from the distance between A and C. To obtain a consistent
distance for null plot A, averaging is necessary. However, this approach
is not applicable to the distance proposed in this paper, as we only
compare the residual plot against a theoretically good residual plot.
Consequently, the statistic must be adjusted to evaluate our distance
measure effectively.

One important aspect that the \(\delta\)-difference was designed to
capture is the empirical distribution of distances for null plot. If we
were to replace the mean distances \(\bar{d}_{\text{null}}^{(j)}\)
directly with \(D_{\text{null}}^{(j)}\), the distance of the \(j\)-th
null plot, the resulting distribution would be degenerate, since
\(D_{null}\) equals zero by definition. Additionally, \(D\) can not be
derived from an image, meaning it falls outside the scope of the
distances considered by \citet{chowdhury2018measuring}. Instead, the
focus should be on the empirical distribution of \(\hat{D}\), as it
influences decision-making. Therefore, the adjusted \(\delta\)-different
is defined as

\[
\delta_{\text{adj}} = \hat{D} - \underset{j}{\text{max}}\left(\hat{D}_{\text{null}}^{(j)}\right) \quad \text{for}~j = 1,...,m-1,
\]

\noindent where \(\hat{D}_{\text{null}}^{(j)}\) is the estimated
distance for the \(j\)-th null plot, and \(m\) is the number of plots in
a lineup.

Figure \ref{fig:delta} displays the scatter plot of the weighted
detection rate vs the adjusted \(\delta\)-difference. It indicates that
the weighted detection rate increases as the adjusted
\(\delta\)-difference increases, particularly when the adjusted
\(\delta\)-difference is greater than zero. A negative adjusted
\(\delta\)-difference suggests that there is at least one null plot in
the lineup with a stronger visual signal than the true residual plot. In
some instances, the weighted detection rate is close to one, yet the
adjusted \(\delta\)-difference is negative. This discrepancy implies
that the distance measure, or the estimated distance, may not perfectly
reflect actual human behavior.

\begin{table}

\caption{\label{tab:human-conv-table}Summary of the comparison of decisions made by computer vision model with decisions made by conventional tests and visual tests conducted by human.}
\centering
\begin{tabular}[t]{lrrr}
\toprule
Violations & \#Samples & \#Agreements & Agreement rate\\
\midrule
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Compared with conventional tests}}\\
\hspace{1em}heteroskedasticity & 540 & 464 & 0.8593\\
\hspace{1em}non-linearity & 576 & 459 & 0.7969\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Compared with visual tests conducted by human}}\\
\hspace{1em}heteroskedasticity & 540 & 367 & 0.6796\\
\hspace{1em}non-linearity & 576 & 385 & 0.6684\\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}[!h]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/conv-mosaic-1}

}

\caption{Rejection rate ($p$-value $\leq0.05$) of computer vision models conditional on conventional tests on non-linearity (left) and heteroskedasticity (right) lineups displayed using a mosaic plot. When the conventional test fails to reject, the computer vision mostly fails to reject the same plot as well as indicated by the height of the top right yellow rectangle, but there are non negliable amount of plots where the conventional test rejects but the computer vision model fails to reject as indicated by the width of the top left yellow rectangle.}\label{fig:conv-mosaic}
\end{figure}

\begin{figure}[!h]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/power-1}

}

\caption{Comparison of power of visual tests, conventional tests and the computer vision model. Marks along the x-axis at the bottom of the plot represent rejections made by each type of test. Marks at the top of the plot represent acceptances. Power curves are fitted by logistic regression models with no intercept but an offset equals to $\text{log}(0.05/0.95)$.}\label{fig:power}
\end{figure}

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/pcp-1}

}

\caption{Parallel coordinate plots of decisions made by computer vision model, conventional tests and visual tests made by human.}\label{fig:pcp}
\end{figure}

\begin{figure}[!h]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/delta-1}

}

\caption{A weighted detection rate vs adjusted $\delta$-difference plot. The brown line is smoothing curve produced by fitting generalized additive models.}\label{fig:delta}
\end{figure}

\section{Examples}\label{sec-examples}

In this section, we present the performance of trained computer vision
model on three example datasets. These include the dataset associated
with the residual plot displaying a ``left-triangle'' shape, as
displayed in Figure \ref{fig:false-finding}, along with the Boston
housing dataset \citep{harrison1978hedonic}, and the ``dino'' datasets
from the \texttt{datasauRus} R package \citep{datasaurus}.

The first example illustrates a scenario where both the computer vision
model and human visual inspection successfully avoid rejecting \(H_0\)
when \(H_0\) is true, contrary to conventional tests. This underscores
the necessity of visually examining the residual plot.

In the second example, we encounter a more pronounced violation of the
model, resulting in rejection of \(H_0\) by all three tests. This
highlights the practicality of the computer vision model, particularly
for less intricate tasks.

The third example presents a situation where the model deviation is
non-typical. Here, the computer vision model and human visual inspection
reject \(H_0\), whereas some commonly used conventional tests do not.
This emphasizes the benefits of visual inspection and the unique
advantage of the computer vision model, which, like humans, makes
decisions based on visual discoveries.

\subsection{Left-triangle}\label{left-triangle}

In Section \ref{sec-model-introduction}, we presented an example
residual plot showcased in Figure \ref{fig:false-finding}, illustrating
how humans might misinterpret the ``left-triangle'' shape as indicative
of heteroskedasticity. Additionally, the Breusch-Pagan test yielded a
rejection with a \(p\)-value of 0.046, despite the residuals originating
from a correctly specified model. Figure \ref{fig:false-lineup} offers a
lineup for this fitted model, showcasing various degrees of
``left-triangle'' shape across all residual plots. This phenomenon is
evidently caused by the skewed distribution of the fitted values.
Notably, if the residual plot were evaluated through a visual test, it
would not be rejected since the true residual plot positioned at 10 can
not be distinguished from the others.

Figure \ref{fig:false-check} presents the results of the assessment by
the computer vision model. Notably, the observed visual signal strength
is considerably lower than the 95\% sample quantile of the null
distribution. Moreover, the bootstrapped distribution suggests that it
is highly improbable for the fitted model to be misspecified as the
majority of bootstrapped fitted models will not be rejected. Thus, for
this particular fitted model, both the visual test and the computer
vision model will not reject \(H_0\). However, the Breusch-Pagan test
will reject \(H_0\) because it can not effectively utilize information
from null plots.

The attention map at Figure \ref{fig:false-check}B suggests that the
estimation is highly influenced by the top-right and bottom-right part
of the residual plot, as it forms two vertices of the triangular shape.
A principal component analysis (PCA) is also performed on the output of
the global pooling layer of the computer vision model. As mentioned in
\citet{simonyan2014very}, a computer vision model built upon the
convolutional blocks can be viewed as a feature extractor. For the
\(32 \times 32\) model, there are 256 features outputted from the global
pooling layer, which can be further used for different visual tasks not
limited to distance prediction. To see if these features can be
effectively used for distinguishing null plots and true residual plot,
we linearly project them into the first and second principal components
space as shown in Figure \ref{fig:false-check}D. It can be observed that
because the bootstrapped plots are mostly similar to the null plots, the
points drawn in different colors are mixed together. The true residual
plot is also covered by both the cluster of null plots and cluster of
bootstrapped plots. This accurately reflects our understanding of Figure
\ref{fig:false-lineup}.

\begin{figure}[!h]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/false-check-1}

}

\caption{A summary of the residual plot assessment evaluted on 200 null plots and 200 bootstrapped plots. (A) The true residual plot exhibiting a "left-triangle" shape. (B) The attention map produced by computing the gradient of the output with respect to the greyscale input.  (C) The density plot of estimated distance for null plots and bootstrapped plots. The green area indicates the distribution of estimated distances for bootstrapped plots, while the yellow area represents the distribution of estimated distances for null plots. The fitted model will not be rejected since $\hat{D} < Q_{null}(0.95)$. (D) plot of first two principal components of features extracted from the global pooling layer of the computer vision model.  }\label{fig:false-check}
\end{figure}

\begin{figure}[!h]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/false-lineup-1}

}

\caption{A lineup of residual plots displaying "left-triangle" visual patterns. The true residual plot occupies position 10, yet there are no discernible visual patterns that distinguish it from the other plots.}\label{fig:false-lineup}
\end{figure}

\subsection{Boston Housing}\label{boston-housing}

The Boston housing dataset, originally published by
\citet{harrison1978hedonic}, offers insights into housing in the Boston,
Massachusetts area. For illustration purposes, we utilize a reduced
version from Kaggle, comprising 489 rows and 4 columns: average number
of rooms per dwelling (RM), percentage of lower status of the population
(LSTAT), pupil-teacher ratio by town (PTRATIO), and median value of
owner-occupied homes in \$1000's (MEDV). In our analysis, MEDV will
serve as the response variable, while the other columns will function as
predictors in a linear regression model. Our primary focus is to detect
non-linearity, because the relationships between RM and MEDV or LSTAT
and MEDV are non-linear.

Figure \ref{fig:boston-check} displays the residual plot and the
assessment conducted by the computer vision model. A clear non-linearity
pattern resembling a ``U'' shape is shown in the plot A. Furthermore,
the RESET test yields a very small \(p\)-value. The estimated distance
\(\hat{D}\) significantly exceeds \(Q_{null}(0.95)\), leading to
rejection of \(H_0\). The bootstrapped distribution also suggests that
almost all the bootstrapped fitted models will be rejected, indicating
that the fitted model is unlikely to be correctly specified. The
attention map in plot B suggests the center of the image has higher
leverage than other areas, and it is the turning point of the ``U''
shape. The PCA provided in plot D shows two distinct clusters of data
points, further underling the visual differences between bootstrapped
plots and null plots. This coincides the findings from Figure
\ref{fig:boston-lineup}, where the true plot exhibiting a ``U'' shape is
visually distinctive from null plots. If a visual test is conducted by
human, \(H_0\) will also be rejected.

\begin{figure}[!h]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/boston-check-1}

}

\caption{A summary of the residual plot assessment for the Boston housing fitted model evaluted on 200 null plots and 200 bootstrapped plots. (A) The true residual plot exhibiting a "U" shape. (B) The attention map produced by computing the gradient of the output with respect to the greyscale input.  (C) The density plot of estimated distance for null plots and bootstrapped plots. The blue area indicates the distribution of estimated distances for bootstrapped plots, while the yellow area represents the distribution of estimated distances for null plots. The fitted model will be rejected since $\hat{D} \geq Q_{null}(0.95)$. (D) plot of first two principal components of features extracted from the global pooling layer of the computer vision model. }\label{fig:boston-check}
\end{figure}

\begin{figure}[!h]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/boston-lineup-1}

}

\caption{A lineup of residual plots for the Boston housing fitted model. The true residual plot is at position 7. It can be easily identified as the most different plot.}\label{fig:boston-lineup}
\end{figure}

\subsection{DatasauRus}\label{datasaurus}

The computer vision model possesses the capability to detect not only
typical issues like non-linearity, heteroskedasticity, and non-normality
but also artifact visual patterns resembling real-world objects, as long
as they do not appear in null plots. These visual patterns can be
challenging to categorize in terms of model violations. Therefore, we
will employ the RESET test, the Breusch-Pagan test, and the Shapiro-Wilk
test \citep{shapiro1965analysis} for comparison.

The ``dino'' dataset within the \texttt{datasauRus} R package
exemplifies this scenario. With only two columns, x and y, fitting a
regression model to this data yields a residual plot resembling a
``dinosaur'', as displayed in Figure \ref{fig:dino-check}A.
Unsurprisingly, this distinct residual plot stands out in a lineup, as
shown in Figure \ref{fig:dino-lineup}. A visual test conducted by humans
would undoubtedly reject \(H_0\).

According to the residual plot assessment by the computer vision model,
\(\hat{D}\) exceeds \(Q_{null}(0.95)\), warranting a rejection of
\(H_0\). Additionally, most of the bootstrapped fitted models will be
rejected, indicating an misspecified model. However, both the RESET test
and the Breusch-Pagan test yield \(p\)-values greater than 0.3, leading
to a non-rejection of \(H_0\). Only the Shapiro-Wilk test rejects the
normality assumption with a small \(p\)-value.

More importantly, the attention map in Figure \ref{fig:dino-check}B
clearly exhibits a ``dinosaur'' shape, strongly suggesting that the
distance prediction is based on human-perceptible visual patterns. The
computer vision model effectively captures the contour or outline of the
embedded shape, similar to how humans interpret residual plots.
Additionally, the PCA in Figure \ref{fig:dino-check}D demonstrates that
the cluster of bootstrapped plots is positioned at the corner of the
cluster of null plots.

In practice, without accessing the residual plot, it would be
challenging to identify the artificial pattern of the residuals.
Moreover, conducting a normality test for a fitted regression model is
not always standard practice among analysts. Even when performed,
violating the normality assumption is sometimes deemed acceptable,
especially considering the application of quasi-maximum likelihood
estimation in linear regression. This example underscores the importance
of evaluating residual plots and highlights how the proposed computer
vision model can facilitate this process.

\begin{figure}[!h]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/dino-check-1}

}

\caption{A summary of the residual plot assessment for the datasauRus fitted model evaluated on 200 null plots and 200 bootstrapped plots. (A) The residual plot exhibits a "dinosaur" shape. (B) The attention map produced by computing the gradient of the output with respect to the greyscale input.  (C) The density plot of estimated distance for null plots and bootstrapped plots. The blue area indicates the distribution of estimated distances for bootstrapped plots, while the yellow area represents the distribution of estimated distances for null plots. The fitted model will be rejected since $\hat{D} \geq Q_{null}(0.95)$. (D) plot of first two principal components of features extracted from the global pooling layer of the computer vision model.}\label{fig:dino-check}
\end{figure}

\begin{figure}[!h]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/dino-lineup-1}

}

\caption{A lineup of residual plots for the fitted model on the "dinosaur" dataset. The true residual plot is at position 17. It can be easily identified as the most different plot as the visual pattern is extremely artificial.}\label{fig:dino-lineup}
\end{figure}

\section{Limitations and Future Work}\label{limitations-and-future-work}

Despite the computer vision model performing well with general cases
under the synthetic data generation scheme and the three examples used
in this paper, this study has several limitations that could guide
future work.

The proposed distance measure assumes that the true model is a classical
normal linear regression model, which can be restrictive. Although this
paper does not address the relaxation of this assumption, there are
potential methods to evaluate other types of regression models. The most
comprehensive approach would be to define a distance measure for each
different class of regression model and then train the computer vision
model following the methodology described in this paper. To accelerate
training, one could use the convolutional blocks of our trained model as
a feature extractor and perform transfer learning on top of it, as these
blocks effectively capture shapes in residual plots. Another approach
would be to transform the residuals so they are roughly normally
distributed and have constant variance. If only raw residuals are used,
the distance-based statistical testing compares the difference in
distance to a classical normal linear regression model for the true plot
and null plots. This comparison is meaningful only if the difference can
be identified by the distance measure proposed in this paper.

There are other types of residual plots commonly used in diagnostics,
such as residuals vs.~predictor and quantile-quantile plots. In this
study, we focused on the most commonly used residual plot as a starting
point for exploring the new field of automated visual inference.
Similarly, we did not explore other, more sophisticated computer vision
model architectures and specifications for the same reason. While the
performance of the computer vision model is acceptable, there is still
room for improvement to achieve behavior more closely resembling that of
humans interpreting residual plots. This may require external survey
data or human subject experiment data to understand the fundamental
differences between our implementation and human evaluation.

\section{Conclusions}\label{conclusions}

In this paper, we have introduced a distance measure based on
Kullback-Leibler divergence to quantify the disparity between the
residual distribution of a fitted classical normal linear regression
model and the reference residual distribution assumed under correct
model specification. This distance measure effectively captures the
magnitude of model violations in misspecified models. We propose a
computer vision model to estimate this distance, utilizing the residual
plot of the fitted model as input. The resulting estimated distance
serves as the foundation for constructing a single Model Violations
Index (MVI), facilitating the quantification of various model
violations.

Moreover, the estimated distance enables the development of a formal
statistical testing procedure by evaluating a large number of null plots
generated from the fitted model. Additionally, employing bootstrapping
techniques and refitting the regression model allows us to ascertain how
frequently the fitted model is considered misspecified if data were
repeatedly obtained from the same data generating process.

The trained computer vision model demonstrates strong performance on
both the training and test sets, although it exhibits slightly lower
performance on residual plots with non-linearity visual patterns
compared to other types of violations. The statistical tests relying on
the estimated distance predicted by the computer vision model exhibit
lower sensitivity compared to conventional tests but higher sensitivity
compared to visual tests conducted by humans. While the estimated
distance generally mirrors the strength of the visual signal perceived
by humans, there remains scope for further improvement in its
performance.

Several examples are provided to showcase the effectiveness of the
proposed method across different scenarios, emphasizing the similarity
between visual tests and distance-based tests. Overall, both visual
tests and distance-based tests can be viewed as ensemble of tests,
aiming to assess any violations of model assumptions collectively. In
contrast, individual residual diagnostic tests such as the RESET test
and the Breusch-Pagan test only evaluate specific violations of model
assumptions. In practice, selecting an appropriate set of statistical
tests for regression diagnostics can be challenging, particularly given
the necessity of adjusting the significance level for each test.

Our method holds significant value as it helps alleviate a portion of
analysts' workload associated with assessing residual plots. While we
recommend analysts to continue reading residual plots whenever feasible,
as they offer invaluable insights, our approach serves as a valuable
tool for automating the diagnostic process or for supplementary purposes
when needed.

\section*{Acknowledgement}\label{acknowledgement}
\addcontentsline{toc}{section}{Acknowledgement}

These \texttt{R} packages were used for the work: \texttt{tidyverse}
\citep{tidyverse}, \texttt{lmtest} \citep{lmtest}, \texttt{mpoly}
\citep{mpoly}, \texttt{ggmosaic} \citep{ggmosaic}, \texttt{kableExtra}
\citep{kableextra}, \texttt{patchwork} \citep{patchwork},
\texttt{rcartocolor} \citep{rcartocolor}, \texttt{glue} \citep{glue},
\texttt{ggpcp} \citep{ggpcp}, \texttt{here} \citep{here},
\texttt{magick} \citep{magick}, \texttt{yardstick} \citep{yardstick} and
\texttt{reticulate} \citep{reticulate}.

The article was created with R packages \texttt{rticles}
\citep{rticles}, \texttt{knitr} \citep{knitr} and \texttt{rmarkdown}
\citep{rmarkdown}. The project's GitHub repository
(\url{https://github.com/TengMCing/auto_residual_reading_paper})
contains all materials required to reproduce this article.

\bibliographystyle{tfcad}
\bibliography{bibliography.bib}

\end{document}

### 2025-06-12 appendix.Rmd
Init the appendix

--- 2025-06-05 paper.Rmd
@@
 ---
-title: "Automated Assessment of Residual Plots with Computer Vision Models"
+title: Appendix to "Automated Assessment of Residual Plots with Computer Vision Models"
 author:
@@
 ---
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(
-  message = FALSE, 
-  warning = FALSE, 
-  echo = FALSE,
-  fig.width = 8,
-  fig.height = 6,
-  out.width = "100%", 
-  fig.align = "center")
-```
-
-```{r}
-# Visual inference models and p-value calculation
-# remotes::install_github("TengMCing/visage")
-options(tinytex.clean = FALSE)
-library(tidyverse)
-library(visage)
-library(glue)
-
-# To control the simulation in this file
-set.seed(10086)
-```
-
-# Introduction {#sec-model-introduction}
-
-Plotting residuals is commonly regarded as a standard practice in linear regression diagnostics [@belsley1980regression; @cook1982residuals].
-This visual assessment plays a crucial role in identifying whether model assumptions, such as linearity, homoscedasticity, and normality, are reasonable.
-It also helps in understanding the goodness of fit and various unexpected characteristics of the model.
-
-Generating a residual plot in most statistical software is often as straightforward as executing a line of code or clicking a button.
-However, accurately interpreting a residual plot can be challenging.
-A residual plot can exhibit various visual features, but it is crucial to recognize that some may arise from the characteristics of predictors and the natural stochastic variation of the observational unit, rather than indicating a violation of model assumptions [@li2024plot].
-Consider Figure \ref{fig:false-finding} as an example, the residual plot displays a triangular left-pointing shape.
-The distinct difference in the spread of the residuals across the fitted values may result in the analyst suggesting that there may be heteroskedasticity, however, it is important to avoid over-interpreting this visual pattern.
-In this case, the fitted regression model is correctly specified, and the triangular shape is actually a result of the skewed distribution of the predictors, rather than indicating a flaw in the model.
-
-The concept of visual inference, as proposed by @buja2009statistical, provides an inferential framework to assess whether residual plots indeed contain visual patterns inconsistent with the model assumptions.
-The fundamental idea involves testing whether the true residual plot visually differs significantly from null plots, where null plots are plotted with residuals generated from the residual rotation distribution [@langsrud2005rotation], which is a distribution consistent with the null hypothesis $H_0$ that the linear regression model is correctly specified.
-Typically, the visual test is accomplished through the lineup protocol, where the true residual plot is embedded within a lineup alongside several null plots.
-If the true residual plot can be distinguished from the lineup, it provides evidence for rejecting $H_0$.
-
-The practice of delivering a residual plot as a lineup is generally regarded as a valuable approach.
-Beyond its application in residual diagnostics, the lineup protocol has been integrated into the analysis of diverse subjects.
-For instance, Loy and Hofmann [-@loy2013diagnostic; -@loy2014hlmdiag; -@loy2015you] illustrated its applicability in diagnosing hierarchical linear models.
-Additionally, @widen2016graphical and @fieberg2024using demonstrated its utility in geographical and ecology research respectively, while @krishnan2021hierarchical explored its effectiveness in forensic examinations.
-
-A practical limitation of the lineup protocol lies in its reliance on human judgements [see @li2024plot about the practical limitations].
-Unlike conventional statistical tests that can be performed computationally in statistical software, the lineup protocol requires human evaluation of images.
-This characteristic makes it less suitable for large-scale applications, given the associated high labour costs and time requirements.
-There is a substantial need to develop an approach to substitute these human judgement with an automated reading of data plots using machines.
-
-The utilization of computers to interpret data plots has a rich history, with early efforts such as "Scagnostics" by @tukey1985computer, a set of numerical statistics that summarize features of scatter plots.
-@wilkinson2005graph expanded on this work, introducing scagnostics based on computable measures applied to planar proximity graphs.
-These measures, including, but not limited to, "Outlying", "Skinny", "Stringy", "Straight", "Monotonic", "Skewed", "Clumpy", and "Striated",  aimed to characterize outliers, shape, density, trend, coherence and other characteristics of the data.
-While this approach has been inspiring, there is a recognition [@buja2009statistical] that it may not capture all the necessary visual features that differentiate true residual plots from null plots.
-A more promising alternative entails enabling machines to learn the function for extracting visual features from residual plots.
-Essentially, this means empowering computers to discern the crucial visual features for residual diagnostics and determining the method to extract them.
-
-Modern computer vision models are well-suited for addressing this challenge.
-They rely on deep neural networks with convolutional layers [@fukushima1982neocognitron].
-These layers use small, sliding windows to scan the image, performing a dot product to extract local features and patterns.
-Numerous studies have demonstrated the efficacy of convolutional layers in addressing various vision tasks, including image recognition [@rawat2017deep].
-Despite the widespread use of computer vision models in fields like computer-aided diagnosis [@lee2015image], pedestrian detection [@brunetti2018computer], and facial recognition [@emami2012facial], their application in reading data plots remains limited.
-While some studies have explored the use of computer vision models for tasks such as reading recurrence plots for time series regression [@ojeda2020multivariate], time series classification [@chu2019automatic; @hailesilassie2019financial; @hatami2018classification; @zhang2020encoding], anomaly detection [@chen2020convolutional], and pairwise causality analysis [@singh2017deep], the application of reading residual plots with computer vision models is a new field of study.
-
-In this paper, we develop computer vision models and integrate them into the residual plots diagnostics workflow, addressing the need for an automated visual inference.
-The paper is structured as follows.
-Section \ref{sec-model-specifications} discusses various specifications of the computer vision models.
-Section \ref{sec-model-distance-between-residual-plots} defines the distance measure used to detect model violations, while Section \ref{sec-model-distance-estimation} explains how the computer vision models estimate this distance measure.
-Section \ref{sec-model-statistical-testing} covers the statistical tests based on the estimated distance, and Section \ref{sec-model-violations-index} introduces a Model Violations Index, which offers a quicker and more convenient assessment.
-Sections \ref{sec-model-data-generation}, \ref{sec-model-architecture}, and  \ref{sec-model-training} detail the data preparation, model architecture, and training process, respectively.
-The results are presented in Section \ref{sec-model-results}.
-Example dataset applications are discussed in Section \ref{sec-examples}.
-Finally, we conclude with a discussion of our findings and propose ideas for future research directions.
-
-```{r false-finding, fig.pos="!h", fig.width = 4, fig.height = 4*4/5, fig.pos="!h", fig.cap="An example residual vs fitted values plot (red line indicates 0 corresponds to the x-intercept, i.e.
-$y=0$).
-The vertical spread of the data points varies with the fitted values.
-This often indicates the existence of heteroskedasticity, however, here the result is due to skewed distribution of the predictors rather than heteroskedasticity.
-The Breusch-Pagan test rejects this residual plot at 95\\% significance level ($p\\text{-value} = 0.046$)."}
-set.seed(452)
-ori_x <- rand_lognormal()
-mod <- heter_model(b = 0, x = closed_form(~-ori_x))
-ori_dat <- mod$gen(300)
-
-ori_dat %>%
-  VI_MODEL$plot(theme = theme_light(), remove_grid_line = TRUE) +
-  xlab("Fitted values") +
-  ylab("Residuals")
-```
-
-# Model Specifications {#sec-model-specifications}
-
-There are various specifications of the computer vision model that can be used to assess residual plots.
-We discuss these specifications below focusing on two key components of the model formula: the input and the output format.
-
-## Input Formats
-
-Deep learning models are in general very sensitive to the input data.
-The quality and relevance of the input data greatly influence the model's capacity to generate insightful and meaningful results.
-There are several designs of the input format that can be considered.
-
-A straightforward architecture of the input layer involves feeding a vector of residuals along with a vector of fitted values, essentially providing all the necessary information for creating a residuals vs fitted values plot.
-However, a drawback of this method is the variable input size, which changes based on the number of observations.
-For modern computer vision models implemented in mainstream software like TensorFlow [@abadi2016tensorflow], the input shape is typically fixed.
-One strategy is to pad the input vectors with leading or trailing zeros when the input tensor expects longer vectors, but it may fail if the input vector surpasses the designed length.
-Alternatively, sampling a fixed number of residual-fitted value pairs can ensure consistent input size.
-By carefully designing the sampling process, it is possible to maintain the input's representativeness while conforming to a fixed format.
-
-Another architecture of the input layer involves using an image as input.
-The primary advantage of this design, as opposed to the vector format, is the availability of the existing and sophisticated image processing architectures developed over the years, such as the VGG16 architecture proposed in @simonyan2014very.
-These architectures can effectively capture and summarize spatial information from nearby pixels, which is less straightforward with vector input.
-Encoding data points as pixels introduces certain trade-offs.
-While some detail may be lost during the discretization process, this approach aligns with the sampling-based strategy by focusing on summarizing the overall data distribution.
-Unlike vectors, which struggle to represent spatial relationships, image-based inputs naturally encode the joint distribution of residuals and fitted values in a visually interpretable format.
-This richer representation gives the model a more comprehensive and intuitive understanding of the residual plot's underlying structure.
-Key considerations for using image-based input include image resolution and the aesthetics of the residual plot.
-In general, a higher resolution provides more information to the model but comes at the cost of of increased complexity and greater difficulty in training.
-As for the aesthetics of the residual plot, a practical solution is to consistently present residual plots in the same style to the model.
-This implies that the model can not accept arbitrary images as input but requires the use of the same pre-processing pipeline to convert residuals and fitted values into a standardized-style residual plot.
-
-Providing multiple residual plots to the model, such as a pair of plots, a triplet or a lineup is also a possible option.
-@chopra2005learning have shown that computer vision models designed for image comparison can assess whether a pair of images are similar or dissimilar.
-Applied to our specific problem, we can define null plots of a fitted regression model to be similar to each other, while considering true residual plots to be distinct from null plots of any fitted regression model.
-A triplet constitutes a set of three images, denoted as $image_1$, $image_2$ and $image_3$.
-It is often used to predict whether $image_2$ or $image_3$ is more similar to $image_1$, proving particularly useful for establishing rankings between samples.
-For this setup, we can apply the same criteria to define similarity between images.
-However, it is important to note that these two approaches usually require additional considerations regarding the loss function and, at times, non-standard training processes due to shared weights between different convolutional blocks.
-
-Presenting a lineup to a model aligns closely with the lineup protocol.
-However, as the number of residual plots in a lineup increases, the resolution of the input image grows rapidly, posing challenges in training the model.
-We experimented with this approach in a pilot study, but the performance of the trained model was sub-optimal.
-
-Taking into account the implementation cost and the need for model interpretability, we used the single residual plot input format in this paper.
-
-## Output Formats
-
-Given that the input is a single residual plot represented as a fixed-resolution image, the output of the computer vision model can be designed for various tasks: binary classification, categorical classification with multiple classes, or numeric regression.
-
-The binary outcome can represent whether the input image is consistent with a null plot as determined by either (1) the data generating process or (2) the result of a visual test based on human judgement.
-Training a model following the latter option requires data from prior human subject experiments, presenting difficulties in controlling the quality of data due to variations in experimental settings across different studies.
-Additionally, some visual inference experiments are unrelated to linear regression models or residual plot diagnostics, resulting in a limited amount of available training data.
-
-For a multi-class outcome, the model could classify images into categories such as "contains outliers" vs.
-"does not contain outliers" or "is non-linear" vs.
-"not non-linear," among other diagnostic features.
-While this approach is appealing in theory, it essentially reverts to the traditional testing framework, where tests focus on detecting specific types of model violations in isolation.
-
-One significant advantage of the lineup protocol for residual diagnostics is its ability to detect multiple model violations simultaneously, provided these violations produce visually distinguishable differences in the plots.
-In contrast, training a classification model would require explicitly defining and mapping all potential violations of interest, or at least the most common ones, which can be impractical.
-If the goal is to identify specific cases, conventional statistical tests grounded in regression model assumptions are often more effective.
-Computer vision models may be less suited to such tasks.
-
-Alternatively, the output could be a meaningful and interpretable numerical measure useful for assessing residual plots, such as the strength of suspicious visual patterns reflecting the extent of model violations, or the difficulty index for identifying whether a residual plot has no issues.
-However, these numeric measures are often informally used in daily communication but are not typically formalized or rigorously defined.
-For the purpose of training a model, this numeric measure has to be quantifiable.
-
-In this study, we chose to define and use a distance between a true residual plot and a theoretically "good" residual plot.
-This is further explained in @sec-model-distance-between-residual-plots.
-@vo2016localizing have also demonstrated that defining a proper distance between images can enhance the matching accuracy in image search compared to a binary outcome model.
-
-## Auxiliary Information with Scagnostics
-
-In Section \ref{sec-model-introduction}, we mention that scagnostics consist of a set of manually designed visual feature extraction functions.
-While our computer vision model will learn its own feature extraction function during training, leveraging additional information from scagnostics can enhance the model's predictive accuracy.
-
-For each residual plot used as an input image, we calculated four scagnostics — "Monotonic", "Sparse", "Splines", and "Striped" – using the `cassowaryr` R package [@mason2022cassowaryr].
-These computed measures, along with the number of observations from the fitted model, were provided as the second input for the computer vision model.
-We selected these scagnostics due to their reliability and efficiency, as other scagnostics occasionally caused R process crashes (approximately 5% of the time) during training data preparation, due to a bug in the `interp` R package [@Albrecht2023interp].
-Although the package maintainer later fixed this bug at our request, the fix came too late to retrain the model, and additionally, their high computational costs make them unsuitable for rapid inference, which was a critical factor in our choice.
-
-# Distance from a Theoretically "Good" Residual Plot  {#sec-model-distance-between-residual-plots}
-
-To develop a computer vision model for assessing residual plots within the visual inference framework, it is important to precisely define a numerical measure of "difference" or "distance" between plots.
-This distance can take the form of a basic statistical operation on pixels, such as the sum of square differences, however, a pixel-to-pixel comparison makes little sense in comparing residual plots where the main interest would be structural patterns.
-Alternatively, it could involve established image similarity metrics like the Structural Similarity Index Measure [@wang2004image] which compares images by integrating three perception features of an image: contrast, luminance, and structure (related to average, standard deviation and correlation of pixel values over a window, respectively).
-These image similarity metrics are tailored for image comparison in vastly different tasks to evaluating data plots, where only essential plot elements require assessment [@chowdhury2018measuring].
-We can alternatively define a notion of distance by integrating key plot elements (instead of key perception features like luminance, contrast, and structure), such as those captured by scagnostics mentioned in Section \ref{sec-model-introduction}, but the functional form still needs to be carefully refined to accurately reflect the extent of the violations.
-
-In this section, we introduce a distance measure between a true residual plot and a theoretically 'good' residual plot.
-This measure quantifies the divergence between the residual distribution of a given fitted regression model and that of a correctly specified model.
-The computation assumes knowledge of the data generating processes for predictors and response variables.
-Since these processes are often unknown in practice, we will discuss a method to estimate this distance using a computer vision model in Section \ref{sec-model-distance-estimation}.
-
-## Residual Distribution
-
-For a classical normal linear regression model, $\boldsymbol{y} = \boldsymbol{X}\boldsymbol{\beta} + \boldsymbol{e}$, the residual $\hat{\boldsymbol{e}}$ are derived as the difference of the fitted values and observed values $\boldsymbol{y}$.
-Suppose the data generating process is known and the regression model is correctly specified, by the Frisch-Waugh-Lowell theorem [@frisch1933partial], residuals $\hat{\boldsymbol{e}}$ can also be treated as random variables and written as a linear transformation of the error $\boldsymbol{e}$ formulated as $\hat{\boldsymbol{e}} = \boldsymbol{R}\boldsymbol{e}$, where $\boldsymbol{R}=\boldsymbol{I}_n -\boldsymbol{X}(\boldsymbol{X}^\top\boldsymbol{X})^{-1}\boldsymbol{X}^\top$ is the residual operator, $\boldsymbol{I}_n$ is a $n$ by $n$ identity matrix, and $n$ is the number of observations.
-
-One of the assumptions of the classical normal linear regression model is that the error $\boldsymbol{e}$ follows a multivariate normal distribution with zero mean and constant variance, i.e., $\boldsymbol{e} \sim N(\boldsymbol{0}_n,\sigma^2\boldsymbol{I}_n)$.
-It follows that the distribution of residuals $\hat{\boldsymbol{e}}$ can be characterized by a certain probability distribution, denoted as $Q$, which is transformed from the multivariate normal distribution.
-This reference distribution $Q$ summarizes what "good" residuals should follow given the design matrix $\boldsymbol{X}$ is known and fixed.
-
-Suppose the design matrix $\boldsymbol{X}$ has linearly independent columns, the trace of the hat matrix $\boldsymbol{H} = \boldsymbol{X}(\boldsymbol{X}^\top\boldsymbol{X})^{-1}\boldsymbol{X}^\top$ will equal to the number of columns in $\boldsymbol{X}$ denoted as $k$.
-As a result, the rank of $\boldsymbol{R}$ is $n - k$, and $Q$ is a degenerate multivariate distribution.
-To capture the characteristics of $Q$, such as moments, we can simulate a large numbers of $\boldsymbol{\varepsilon}$ and transform it to $\boldsymbol{e}$ to get the empirical estimates.
-For simplicity, in this study, we replaced the variance-covariance matrix of residuals $\text{cov}(\boldsymbol{e}, \boldsymbol{e}) = \boldsymbol{R}\sigma^2\boldsymbol{R}^\top = \boldsymbol{R}\sigma^2$ with a full-rank diagonal matrix $\text{diag}(\boldsymbol{R}\sigma^2)$, where $\text{diag}(.)$ sets the non-diagonal entries of a matrix to zeros.
-The resulting distribution for $Q$ is $N(\boldsymbol{0}_n, \text{diag}(\boldsymbol{R}\sigma^2))$.
-
-Distribution $Q$ is derived from the correctly specified model.
-However, if the model is misspecified, then the actual distribution of residuals denoted as $P$, will be different from $Q$.
-For example, if the data generating process contains variables correlated with any column of $\boldsymbol{X}$ but missing from $\boldsymbol{X}$, causing an omitted variable problem, $P$ will be different from $Q$ because the residual operator obtained from the fitted regression model will not be the same as $\boldsymbol{R}$.
-Besides, if the $\boldsymbol{\varepsilon}$ follows a non-normal distribution such as a multivariate lognormal distribution, $P$ will usually be skewed and has a long tail.
-
-## Distance of $P$ from $Q$
-
-Defining a proper distance between distributions is usually easier than defining a proper distance between data plots.
-Given the true residual distribution $Q$ and the reference residual distribution $P$, we used a distance measure based on Kullback-Leibler divergence [@kullback1951information] to quantify the difference between two distributions as
-
-\begin{equation} \label{eq:kl-0}
-D = \log\left(1 + D_{KL}\right),
-\end{equation}
-
-where $D_{KL}$ is defined as
-
-\begin{equation} \label{eq:kl-1}
-D_{KL} = \int_{\mathbb{R}^{n}}\log\frac{p(\boldsymbol{e})}{q(\boldsymbol{e})}p(\boldsymbol{e})d\boldsymbol{e},
-\end{equation}
-
-\noindent and $p(.)$ and $q(.)$ are the probability density functions for distribution $P$ and distribution $Q$, respectively.
-
-This distance measure was first proposed in @li2024plot.
-It was mainly designed for measuring the effect size of non-linearity and heteroskedasticity in a residual plot.
-@li2024plot have derived that, for a classical normal linear regression model that omits necessary higher-order predictors $\boldsymbol{Z}$ and the corresponding parameter $\boldsymbol{\beta}_z$, and incorrectly assumes $\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n,\sigma^2\boldsymbol{I}_n)$ while in fact $\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n, \boldsymbol{V})$ where $\boldsymbol{V}$ is an arbitrary symmetric positive semi-definite matrix, $Q$ can be represented as $N(\boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z, \text{diag}(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}))$.
-Note that the variance-covariance matrix is replaced with the diagonal matrix to ensure it is a full-rank matrix.
-
-Since both $P$ and $Q$ are adjusted to be multivariate normal distributions, Equation \ref{eq:kl-1} can be further expanded to
-
-\begin{equation} \label{eq:kl-2}
-D_{KL} = \frac{1}{2}\left(\log\frac{|\boldsymbol{W}|}{|\text{diag}(\boldsymbol{R}\sigma^2)|} - n + \text{tr}(\boldsymbol{W}^{-1}\text{diag}(\boldsymbol{R}\sigma^2)) + \boldsymbol{\mu}_z^\top\boldsymbol{W}^{-1}\boldsymbol{\mu}_z\right),
-\end{equation}
-
-\noindent where $\boldsymbol{\mu}_z = \boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z$, and $\boldsymbol{W} = \text{diag}(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R})$.
-The assumed error variance $\sigma^2$ is set to be $\text{tr}(\boldsymbol{V})/n$, which is the expectation of the estimated variance.
-
-## Non-normal $P$
-
-For non-normal error $\boldsymbol{\varepsilon}$, the true residual distribution $P$ is unlikely to be a multivariate normal distribution.
-Thus, Equation \ref{eq:kl-2} given in @li2024plot will not be applicable to models violating the normality assumption.
-
-To evaluate the Kullback-Leibler divergence of non-normal $P$ from $Q$, the fallback is to solve Equation \ref{eq:kl-1} numerically.
-However, since $\boldsymbol{e}$ is a linear transformation of non-normal random variables, it is very common that the general form of $P$ is unknown, meaning that we can not easily compute $p(\boldsymbol{e})$ using a well-known probability density function.
-Additionally, even if $p(\boldsymbol{e})$ can be calculated for any $\boldsymbol{e} \in \mathbb{R}^n$, it will be very difficult to do numerical integration over the $n$-dimensional space, because $n$ could be potentially very large.
-
-In order to approximate $D_{KL}$ in a practically computable manner, the elements of $\boldsymbol{e}$ are assumed to be independent of each other.
-This assumption solves both of the issues mentioned above.
-First, we no longer need to integrate over $n$ random variables.
-The result of Equation \ref{eq:kl-1} is now the sum of the Kullback-Leibler divergence evaluated for each individual residual due to the assumption of independence between observations.
-Second, it is not required to know the joint probability density $p(\boldsymbol{e})$ any more.
-Instead, the evaluation of Kullback-Leibler divergence for an individual residual relies on the knowledge of the marginal density $p_i(e_i)$, where $e_i$ is the $i$-th residual for $i = 1, ..., n$.
-This is much easier to approximate through simulation.
-It is also worth mentioning that this independence assumption generally will not hold if $\text{cov}(e_i, e_j) \neq 0$ for any $1 \leq i < j \leq n$, but its existence is essential for reducing the computational cost.
-
-Given $\boldsymbol{X}$ and $\boldsymbol{\beta}$, the algorithm for approximating Equation \ref{eq:kl-1} starts from simulating $m$ sets of observed values $\boldsymbol{y}$ according to the data generating process.
-The observed values are stored in a matrix $\boldsymbol{A}$ with $n$ rows and $m$ columns, where each column of $\boldsymbol{A}$ is a set of observed values.
-Then, we can get $m$ sets of realized values of $\boldsymbol{e}$ stored in the matrix $\boldsymbol{B}$ by applying the residual operator $\boldsymbol{B} = \boldsymbol{R}\boldsymbol{A}$.
-Furthermore, kernel density estimation (KDE) with Gaussian kernel and optimal bandwidth selected by the Silverman's rule of thumb [@silverman2018density] is applied on each row of $\boldsymbol{B}$ to estimate $p_i(e_i)$ for $i = 1, ..., n$.
-The KDE computation can be done by the `density` function in R.
-
-Since the Kullback-Leibler divergence can be viewed as the expectation of the log-likelihood ratio between distribution $P$ and distribution $Q$ evaluated on distribution $P$, we can reuse the simulated residuals in matrix $\boldsymbol{B}$ to estimate the expectation by the sample mean.
-With the independence assumption, for non-normal $P$, $D_{KL}$ can be approximated by
-
-\begin{align*} \label{eq:kl-3}
-D_{KL} &\approx \sum_{i = 1}^{n} \hat{D}_{KL}^{(i)}, \\
-\hat{D}_{KL}^{(i)} &= \frac{1}{m}\sum_{j = 1}^{m} \log\frac{\hat{p}_i(B_{ij})}{q(B_{ij})},
-\end{align*}
-
-\noindent where $\hat{D}_{KL}^{(i)}$ is the estimator of the Kullback-Leibler divergence for an individual residual $e_i$, $\boldsymbol{B}_{ij}$ is the $i$-th row and $j$-th column entry of the matrix $\boldsymbol{B}$, $\hat{p}_i(.)$ is the kernel density estimator of $p_i(.)$, $q(.)$ is the normal density function with mean zero and an assumed variance estimated as $\hat{\sigma}^2 = \sum_{b \in vec(\boldsymbol{B})}(b - \sum_{b \in vec(\boldsymbol{B})} b/nm)^2/(nm - 1)$, and $vec(.)$ is the vectorization operator which turns a $n \times m$ matrix into a $nm \times 1$ column vector by stacking the columns of the matrix on top of each other.
-
-# Distance Estimation {#sec-model-distance-estimation}
-
-In the previous sections, we have defined a distance measure given in Equation \ref{eq:kl-0} for quantifying the difference between the true residual distribution $P$ and an ideal reference distribution $Q$.
-However, this distance measure can only be computed when the data generating process is known.
-In reality, we often have no knowledge about the data generating process, otherwise we do not need to do a residual diagnostic in the first place.
-
-We use a computer vision model to estimate this distance measure for a residual plot. Let $D$ be the result of Equation \ref{eq:kl-0}, and our estimator $\hat{D}$ is formulated as
-
-\begin{equation} \label{eq:d-approx}
-\hat{D} = f_{CV}(V_{h \times w}(\boldsymbol{e}, \hat{\boldsymbol{y}})),
-\end{equation}
-
-\noindent where $V_{h \times w}(.)$ is a plotting function that saves a residuals vs fitted values plot with fixed aesthetic as an image with $h \times w$ pixels in three channels (RGB), $f_{CV}(.)$ is a computer vision model which takes an $h \times w$ image as input and predicts the distance in the domain $[0, +\infty)$.
-
-With the estimated distance $\hat{D}$, we can compare the underlying distribution of the residuals to a theoretically "good" residual distribution.
-$\hat{D}$ can also be used as an index of the model violations indicating the strength of the visual signal embedded in the residual plot.
-
-It is not expected that $\hat{D}$ will be equal to original distance $D$.
-This is largely because information contained in a single residual plot is limited and it may not be able to summarize all the important characteristics of the residual distribution.
-For a given residual distribution $P$, many different residual plots can be simulated, where many will share similar visual patterns, but some of them could be visually very different from the rest, especially for regression models with small $n$.
-This suggests the error of the estimation will vary depends on whether the input residual plot is representative or not.
-
-# Statistical Testing {#sec-model-statistical-testing}
-
-## Lineup Evaluation {#sec-model-lineup-evaluation}
-
-Theoretically, the distance $D$ for a correctly specified model is $0$, because $P$ will be the same as $Q$.
-However, the computer vision model may not necessary predict $0$ for a null plot.
-Using Figure \ref{fig:false-finding} as an example, it contains a visual pattern which is an indication of heteroskedasticity.
-We would not expect the model to be able to magically tell if the suspicious pattern is caused by the skewed distribution of the fitted values or the existence of heteroskedasticity.
-Additionally, some null plots could have outliers or strong visual patterns due to randomness, and a reasonable model will try to summarize those information into the prediction, resulting in $\hat{D} > 0$.
-
-This property is not an issue if $\hat{D} \gg 0$ for which the visual signal of the residual plot is very strong, and we usually do not need any further examination of the significance of the result.
-However, if the visual pattern is weak or moderate, having $\hat{D}$ will not be sufficient to determine if $H_0$ should be rejected.
-
-To address this issue we can adhere to the paradigm of visual inference, by comparing the estimated distance $\hat{D}$ to the estimated distances for the null plots in a lineup.
-Specifically, if a lineup comprises 20 plots, the null hypothesis $H_0$ will be rejected if $\hat{D}$ exceeds the maximum estimated distance among the $m - 1$ null plots, denoted as $\max\limits_{1 \leq i \leq m-1} {\hat{D}_{null}^{(i)}}$, where $\hat{D}_{null}^{(i)}$ represents the estimated distance for the $i$-th null plot.
-This approach is equivalent to the typical lineup protocol requiring a 95% significance level, where $H_0$ is rejected if the data plot is identified as the most distinct plot by the sole observer.
-The estimated distance serves as a metric to quantify the difference between the data plot and the null plots, as intended.
-
-<!-- For lineups consisting of more than 20 plots, the 95% significance level can be maintained if the number of plots is a multiple of 20.
-Specifically, for lineups comprising $20t$ plots, where $t$ is a positive integer, we reject $H_0$ if $\hat{D}$ exceeds 95% ${\hat{D}_{null}^{(i)}}$ for $i = 1, \ldots, 20t-1$.
-The $p$-value in this case is given by $\frac{1}{20t} + \frac{1}{20t}\sum_{i=1}^{20t-1} I\left(\hat{D}_{null}^{(i)} > \hat{D}\right)$, where $I(\cdot)$ is the indicator function.
--->
-
-Moreover, if the number of plots in a lineup, denoted by $m$, is sufficiently large, the empirical distribution of ${\hat{D}_{null}^{(i)}}$ can be viewed as an approximation of the null distribution of the estimated distance.
-Consequently, quantiles of the null distribution can be estimated using the sample quantiles, and these quantiles can be utilized for decision-making purposes.
-The details of the sample quantile computation can be found in @hyndman1996sample.
-For instance, if $\hat{D}$ is greater than or equal to the 95% sample quantile, denoted as $Q_{null}(0.95)$, we can conclude that the estimated distance for the true residual plot is significantly different from the estimated distance for null plots with a 95% significance level.
-Based on our experience, to obtain a stable estimate of the 95% quantile, the number of null plots, $n_{null}$, typically needs to be at least 100.
-However, if the null distribution exhibits a long tail, a larger number of null plots may be required.
-Alternatively, a $p$-value is the probability of observing a distance equally or greater than $\hat{D}$ under the null hypothesis $H_0$, and it can be estimated by $\frac{1}{m} + \frac{1}{m}\sum_{i=1}^{m-1}I\left(\hat{D}_{null}^{(i)} \geq \hat{D}\right)$.
-
-To alleviate computation burden, a lattice of quantiles for $\hat{D}$ under $H_0$ with specified sample sizes can be precomputed.
-We can then map the $\hat{D}$ and sample size to the closet quantile and sample size in lattice to calculate the corresponding $p$-value.
-This approach lose precision in $p$-value calculation, however, significantly improves computational efficiency.
-
-## Bootstrapping
-
-Bootstrap is often employed in linear regression when conducting inference for estimated parameters [see @davison1997bootstrap and @efron1994introduction].
-It is typically done by sampling individual cases with replacement and refitting the regression model.
-If the observed data accurately reflects the true distribution of the population, the bootstrapped estimates can be used to measure the variability of the parameter estimate without making strong distributional assumptions about the data generating process.
-
-Similarly, bootstrap can be applied on the estimated distance $\hat{D}$.
-For each refitted model $M_{boot}^{(i)}$, there will be an associated residual plot $V_{boot}^{(i)}$ which can be fed into the computer vision model to obtain $\hat{D}_{boot}^{(i)}$, where $i = 1,...,n_{boot}$, and $n_{boot}$ is the number of bootstrapped samples.
-If we are interested in the variation of $\hat{D}$, we can use $\hat{D}_{boot}^{(i)}$ to estimate a confidence interval.
-
-Alternatively, since each $M_{boot}^{(i)}$ has a set of estimated coefficients $\hat{\boldsymbol{\beta}}_{boot}^{(i)}$ and an estimated variance $\hat{\sigma^2}_{boot}^{(i)}$, a new approximated null distribution can be construed and the corresponding 95% sample quantile $Q_{boot}^{(i)}(0.95)$ can be computed.
-Then, if $\hat{D}_{boot}^{(i)} \geq Q_{boot}^{(i)}(0.95)$, $H_0$ will be rejected for $M_{boot}^{(i)}$.
-The ratio of rejected $M_{boot}^{(i)}$ among all the refitted models provides an indication of how often the assumed regression model are considered to be incorrect if the data can be obtained repetitively from the same data generating process.
-But this approach is computationally very expensive because it requires $n_{boot} \times n_{null}$ times of residual plot assessment.
-In practice, $Q_{null}(0.95)$ can be used to replace $Q_{boot}^{(i)}(0.95)$ in the computation.
-
-# Model Violations Index {#sec-model-violations-index}
-
-While statistical testing is a powerful tool for detecting model violations, it can become cumbersome and time-consuming when quick decisions are needed, particularly due to the need to evaluate numerous null plots.
-In practice, a more convenient and immediate method for assessing model performance is often required.
-This is where an index, such as the Model Violations Index (MVI), becomes valuable.
-It offers a straightforward way to quantify deviations from model assumptions, enabling rapid assessment and easier comparison across models.
-
-The estimator $\hat{D}$ measures the difference between the true residual distribution and the reference residual distribution, a difference primarily arises from deviations in model assumptions.
-The magnitude of $D$ directly reflects the degree of these deviations, thus making $\hat{D}$ instrumental in forming a model violations index (MVI).
-
-Note that if more observations are used for estimating the linear regression, the result of Equation \ref{eq:kl-1} will increase, as the integration will be performed with larger $n$.
-For a given data generating process, $D$ typically increases logarithmically with the number of observations.
-This behavior comes from the relationship $D = \text{log}(1 + D_{KL})$, where $D_{KL} = \sum_{i=1}^{n}D_{KL}^{(i)}$ under the assumption of independence.
-
-Since $\hat{D}$ is an estimate of $D$, it is expected that a larger number of observations will also lead to a higher $\hat{D}$.
-However, this does not imply that $\hat{D}$ fails to accurately represent the extent of model violations.
-In fact, when examining residual plots with more observations, we often observe a stronger visual signal strength, as the underlying patterns are more likely to be revealed, except in cases of significant overlapping.
-
-Therefore, the Model Violations Index (MVI) can be proposed as
-
-\begin{equation} \label{eq:mvi}
-\text{MVI} = C + \hat{D} - \log(n),
-\end{equation}
-
-\noindent where $C$ is a large enough constant keeping the result positive and the term $-\log(n)$ is used to offset the increase in $D$ due to sample size.
-
-Figure \ref{fig:poly-heter-index} displays the residual plots for fitted models exhibiting varying degrees of non-linearity and heteroskedasticity.
-Each residual plot's MVI is computed using Equation \ref{eq:mvi} with $C = 10$.
-When $\text{MVI} > 8$, the visual patterns are notably strong and easily discernible by humans.
-In the range $6 < \text{MVI} < 8$, the visibility of the visual pattern diminishes as MVI decreases.
-Conversely, when $\text{MVI} < 6$, the visual pattern tends to become relatively faint and challenging to observe.
-Table \ref{tab:mvi} provides a summary of the MVI usage and it is applicable to other linear regression models.
-
-```{r}
-data.frame(degree = c("Strong", "Moderate", "Weak"),
-           range = c("$\\text{MVI} > 8$", "$6 < \\text{MVI} < 8$", "$\\text{MVI} < 6$")) %>%
-  kableExtra::kable(format = "latex",
-                    booktabs = TRUE,
-                    label = "mvi",
-                    caption = "Degree of model violations or the strength of the visual signals according to the Model Violations Index (MVI). The constant $C$ is set to be 10.",
-                    escape = FALSE,
-                    col.names = c("Degree of model violations", "Range ($C$ = 10)"),
-                    align = "lc")
-```
-
-```{r}
-set.seed(1495)
-if (!file.exists(here::here("cached_data/poly_index.rds"))) {
-  e_sigma_factor <- 3
-  keras_mod <- autovi::get_keras_model("vss_phn_32")
-  poly_dgp <- phn_model(j = 3, sigma = 0.1^e_sigma_factor)
-  ori_dat <- poly_dgp$gen(300)
-
-  vss <- c()
-  actual_ss <- c()
-  dat_combined <- tibble()
-
-  for (e_sigma in seq(0.7, 1.8, length.out = 20)^e_sigma_factor) {
-    poly_dgp <- phn_model(j = 3, sigma = e_sigma)
-    dat <- poly_dgp$gen(300, computed = list(x1 = ori_dat$x1, e = ori_dat$e, k = ori_dat$k * e_sigma / 0.1^e_sigma_factor))
-    mod <- lm(y ~ x1, data = dat)
-    my_vi <- autovi::auto_vi(fitted_mod = mod, keras_mod = keras_mod)
-    vss <- c(vss, my_vi$vss()$vss)
-    actual_ss <- c(actual_ss, log(poly_dgp$sample_effect_size(dat) + 1))
-
-    dat_combined <- bind_rows(dat_combined, mutate(dat, e_sigma = e_sigma))
-  }
-
-  saveRDS(dat_combined, here::here("cached_data/poly_index.rds"))
-  saveRDS(vss, here::here("cached_data/poly_index_vss.rds"))
-}
-
-dat_combined <- readRDS(here::here("cached_data/poly_index.rds"))
-vss <- readRDS(here::here("cached_data/poly_index_vss.rds"))
-
-p1 <- dat_combined %>%
-  autovi::AUTO_VI$plot_resid() +
-  facet_wrap(~e_sigma, scales = "free", 
-             labeller = function(...){ 
-               tibble(nb = map(glue("'MVI' == {format(vss - log(300) + 10, digits = 3)}"), str2expression))
-               },
-             ncol = 5) +
-  theme(strip.text = element_text(size = 10), plot.title = element_text(size = 10)) +
-  ggtitle("(A) Non-linearity")
-```
-
-```{r}
-set.seed(1495)
-if (!file.exists(here::here("cached_data/heter_index.rds"))) {
-  b_factor <- 2
-  keras_mod <- autovi::get_keras_model("vss_phn_32")
-  heter_dgp <- phn_model(include_z = FALSE, b = 0)
-  ori_dat <- poly_dgp$gen(300)
-
-  vss <- c()
-  actual_ss <- c()
-  dat_combined <- tibble()
-
-  for (b in rev(seq(0.2, 4, length.out = 20))^b_factor) {
-    heter_dgp <- phn_model(include_z = FALSE, b = b)
-    dat <- heter_dgp$gen(300, computed = list(x1 = ori_dat$x1, e = ori_dat$e))
-    mod <- lm(y ~ x1, data = dat)
-    my_vi <- autovi::auto_vi(fitted_mod = mod, keras_mod = keras_mod)
-    vss <- c(vss, my_vi$vss()$vss)
-    actual_ss <- c(actual_ss, log(heter_dgp$sample_effect_size(dat) + 1))
-
-    dat_combined <- bind_rows(dat_combined, mutate(dat, nb = -b))
-  }
-
-  saveRDS(dat_combined, here::here("cached_data/heter_index.rds"))
-  saveRDS(vss, here::here("cached_data/heter_index_vss.rds"))
-}
-
-dat_combined <- readRDS(here::here("cached_data/heter_index.rds"))
-vss <- readRDS(here::here("cached_data/heter_index_vss.rds"))
-
-p2 <- dat_combined %>%
-  autovi::AUTO_VI$plot_resid() +
-  facet_wrap(~nb, scales = "free", 
-             labeller = function(...){ 
-               tibble(nb = map(glue("'MVI' == {format(vss - log(300) + 10, digits = 3)}"), str2expression))
-               },
-             ncol = 5) +
-  theme(strip.text = element_text(size = 10), plot.title = element_text(size = 10)) +
-  ggtitle("(B) Heteroskedasticity")
-```
-
-```{r poly-heter-index, fig.pos = "!h", fig.cap = "Residual plots generated from fitted models exhibiting varying degrees of (A) non-linearity and (B) heteroskedasticity violations.
-The model violations index (MVI) is displayed atop each residual plot.
-The non-linearity patterns are relatively strong for $MVI > 8$, and relatively weak for $MVI < 6$, while the heteroskedasticity patterns are relatively strong for $MVI > 8$, and relatively weak for $MVI < 6$.", fig.height = 10}
-patchwork::wrap_plots(p1, p2, ncol = 1)
-```
-
-# Data Generation {#sec-model-data-generation}
-
-## Simulation Scheme
-
-While observational data is frequently employed in training models for real-world applications, the data generating process of observational data often remains unknown, making computation for our target variable $D$ unattainable.
-Consequently, the computer vision models developed in this study were trained using synthetic data, including 80,000 training images and 8,000 test images.
-This approach provided us with precise label annotations.
-Additionally, it ensured a large and diverse training dataset, as we had control over the data generating process, and the simulation of the training data was relatively cost-effective.
-
-We have incorporated three types of residual departures of linear regression model in the training data, including non-linearity, heteroskedasticity and non-normality.
-All three departures can be summarized by the data generating process formulated as
-
-\begin{align} \label{eq:data-sim}
-\boldsymbol{y} &= \boldsymbol{1}_n + \boldsymbol{x}_1 + \beta_1\boldsymbol{x}_2 + \beta_2(\boldsymbol{z} + \beta_1\boldsymbol{w}) + \boldsymbol{k} \odot \boldsymbol{\varepsilon}, \\
-\boldsymbol{z} &= \text{He}_j(g(\boldsymbol{x}_1, 2)), \\
-\boldsymbol{w} &= \text{He}_j(g(\boldsymbol{x}_2, 2)), \\
-\boldsymbol{k} &= \left[\boldsymbol{1}_n + b(2 - |a|)(\boldsymbol{x}_1 + \beta_1\boldsymbol{x}_2 - a\boldsymbol{1}_n)^{\circ2}\right]^{\circ1/2},
-\end{align}
-
-\noindent where $\boldsymbol{y}$, $\boldsymbol{x}_1$, $\boldsymbol{x}_2$, $\boldsymbol{z}$, $\boldsymbol{w}$, $\boldsymbol{k}$ and $\boldsymbol{\varepsilon}$ are vectors of size $n$, $\boldsymbol{1}_n$ is a vector of ones of size $n$, $\boldsymbol{x}_1$ and $\boldsymbol{x}_2$ are two independent predictors, $\text{He}_j(.)$ is the $j$th-order probabilist's Hermite polynomials [@hermite1864nouveau], $(.)^{\circ2}$ and $(.)^{\circ1/2}$ are Hadamard square and square root, $\odot$ is the Hadamard product, and $g(\boldsymbol{x}, k)$ is a scaling function to enforce the support of the random vector to be $[-k, k]^n$ defined as
-
-$$g(\boldsymbol{x}, k) = 2k \cdot \frac{\boldsymbol{x} - x_{\min}\boldsymbol{1}_n}{x_{\max} - x_{\min}} - k\boldsymbol{1}_n,~for~k > 0,$$
-\noindent where $x_{\min} = \underset{i \in \{ 1,...,n\}}{\min} x_i$, $x_{\max} = \underset{i \in \{ 1,...,n\}}{\max} x_i$ and $x_i$ is the $i$-th entry of $\boldsymbol{x}$.
-
-```{r}
-tibble(Factor = c("j", "a", "b", "$\\beta_1$", "$\\beta_2$", " $\\text{dist}_\\varepsilon$", "$\\text{dist}_{x1}$", "$\\text{dist}_{x2}$", "$\\sigma_{\\varepsilon}$", "$\\sigma_{X1}$", "$\\sigma_{X2}$", "n"),
-       Domain = c("\\{2, 3, ..., 18\\}", "[-1, 1]", "[0, 100]", "\\{0, 1\\}", "\\{0, 1\\}", "\\{discrete, uniform, normal, lognormal\\}", "\\{discrete, uniform, normal, lognormal\\}", "\\{discrete, uniform, normal, lognormal\\}", "[0.0625, 9]", "[0.3, 0.6]", "[0.3, 0.6]", "[50, 500]")) %>%
-  kableExtra::kable(format = "latex",
-                    escape = FALSE,
-                    label = "factor",
-                    caption = "Factors used in the data generating process for synthetic data simulation.
-Factor $j$ and $a$ controls the non-linearity shape and the heteroskedasticity shape respectively.
-Factor $b$, $\\sigma_\\varepsilon$ and $n$ control the signal strength.
-Factor $\\text{dist}_\\varepsilon$, $\\text{dist}_{x1}$ and $\\text{dist}_{x2}$ specifies the distribution of $\\varepsilon$, $X_1$ and $X_2$ respectively.",
-                    booktabs = TRUE)
-```
-
-The residuals and fitted values of the fitted model were obtained by regressing $\boldsymbol{y}$ on $\boldsymbol{x}_1$.
-If $\beta_1 \neq 0$, $\boldsymbol{x}_2$ was also included in the design matrix.
-This data generation process was adapted from @li2024plot, where it was utilized to simulate residual plots exhibiting non-linearity and heteroskedasticity visual patterns for human subject experiments.
-A summary of the factors utilized in Equation \ref{eq:data-sim} is provided in Table \ref{tab:factor}.
-
-In Equation \ref{eq:data-sim}, $\boldsymbol{z}$ and $\boldsymbol{w}$ represent higher-order terms of $\boldsymbol{x}_1$ and $\boldsymbol{x}_2$, respectively.
-If $\beta_2 \neq 0$, the regression model will encounter non-linearity issues.
-Parameter $j$ serves as a shape parameter that controls the number of tuning points in the non-linear pattern.
-Typically, higher values of $j$ lead to an increase in the number of tuning points, as illustrated in Figure \ref{fig:different-j}.
-
-```{r different-j, fig.pos="!h", fig.cap = "Non-linearity forms generated for the synthetic data simulation. The 17 shapes are generated by varying the order of polynomial given by $j$ in $He_j(.)$."}
-
-set.seed(10086)
-
-# Data for shape 1
-dat_shape_1 <- phn_model(j = 2, include_x2 = FALSE, sigma = 0.05)$gen(500) %>%
-  mutate(j = 2)
-
-# Generate data for shape 2, 3 and 4. Reuse x and e.
-map_df(3:18, function(j) {
-  phn_model(j = j, include_x2 = FALSE, sigma = 0.05)$
-    gen(500, computed = select(dat_shape_1, x1, e)) %>%
-  mutate(j = j)
-}) %>%
-
-  # Combined with data for shape 1
-  bind_rows(dat_shape_1) %>%
-  mutate(j = factor(j)) %>%
-  VI_MODEL$plot(remove_axis = TRUE, remove_grid_line = TRUE, theme = theme_light()) +
-  facet_wrap(~j, scales = "free", labeller = label_parsed, ncol = 5)
-```
-
-Additionally, scaling factor $\boldsymbol{k}$ directly affects the error distribution and it is correlated with $\boldsymbol{x}_1$ and $\boldsymbol{x}_2$.
-If $b \neq 0$ and $\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n, \sigma^2\boldsymbol{I}_n)$, the constant variance assumption will be violated.
-Parameter $a$ is a shape parameter controlling the location of the smallest variance in a residual plot as shown in Figure \ref{fig:different-a}.
-
-```{r different-a, fig.width=8/5*3*2, fig.height=6/4*3*2, fig.pos="!h", fig.cap = 'Heteroskedasticity forms generated for the synthetic data simulation.
-Different shapes are controlled by the continuous factor $a$ between -1 and 1.
-For $a = -1$, the residual plot exhibits a "left-triangle" shape.
-And for $a = 1$, the residual plot exhibits a "right-triangle" shape.
-'}
-
-set.seed(10085)
-
-# Generate data for a = -1
-dat_a_n1 <- phn_model(include_z = FALSE,
-                      include_x2 = FALSE,
-                      a = -1,
-                      b = 100)$gen(500) %>%
-  mutate(a = -1)
-
-# Generate data for other a
-map(c(-0.75, -0.5, -0.25, 0, 0.25, 0.5, 0.75, 1), function(a) {
-  phn_model(include_z = FALSE,
-            include_x2 = FALSE,
-            a = a,
-            b = 100)$gen(500) %>%
-  mutate(a = a)
-}) %>%
-
-  # Combined with data for a = -1
-  bind_rows(dat_a_n1) %>%
-  mutate(a = factor(a)) %>%
-  VI_MODEL$plot(remove_axis = TRUE, remove_grid_line = TRUE, theme = theme_light()) +
-  facet_wrap(~a, scales = "free", ncol = 3) +
-  xlab("Fitted values") +
-  ylab("Residuals")
-```
-
-```{r different-e, fig.pos="!h", fig.cap = 'Non-normality forms generated for the synthetic data simulation.
-Four different error distributions including discrete, lognormal, normal and uniform are considered.'}
-
-set.seed(10086)
-
-# Data for shape 1
-dat_shape_1 <- phn_model(include_z = FALSE, include_x2 = FALSE, e = rand_uniform(-1.4, 1.4))$gen(500) %>%
-  mutate(e_dist = "uniform")
-
-dat_shape_2 <- phn_model(include_z = FALSE, include_x2 = FALSE, sigma = 0.8)$gen(500) %>%
-  mutate(e_dist = "normal")
-
-dat_shape_3 <- phn_model(include_z = FALSE, include_x2 = FALSE, e = rand_lognormal(sigma = 0.6))$gen(500) %>%
-  mutate(e_dist = "lognormal")
-
-dat_shape_4 <- phn_model(include_z = FALSE, include_x2 = FALSE, e = rand_uniform_d(-1.4, 1.4, even = TRUE))$gen(500) %>%
-  mutate(e_dist = "discrete")
-
-# Generate data for shape 2, 3 and 4. Reuse x and e.
-bind_rows(dat_shape_1, dat_shape_2, dat_shape_3, dat_shape_4) %>%
-  VI_MODEL$plot(remove_axis = TRUE, remove_grid_line = TRUE, theme = theme_light()) +
-  facet_wrap(~e_dist, scales = "free", labeller = label_parsed, ncol = 2)
-```
-
-Non-normality violations arise from specifying a non-normal distribution for $\boldsymbol{\varepsilon}$.
-In the synthetic data simulation, four distinct error distributions are considered, including discrete, uniform, normal, and lognormal distributions, as presented in Figure \ref{fig:different-e}.
-Each distribution imparts unique characteristics in the residual plot.
-The discrete error distribution introduces discrete clusters in residuals, while the lognormal distribution typically yields outliers.
-Uniform error distribution may result in residuals filling the entire space of the residual plot.
-All of these distributions exhibit visual distinctions from the normal error distribution.
-
-```{r different-j-x2, fig.pos="!h", fig.cap = "Residual plots of multiple linear regression models with non-linearity issues.
-The 17 shapes are generated by varying the order of polynomial given by $j$ in $He_j(.)$.
-A second predictor $\\boldsymbol{x}_2$ is introduced to the regression model to create complex shapes."}
-
-set.seed(10086)
-
-# Data for shape 1
-dat_shape_1 <- phn_model(j = 2, include_x2 = TRUE, sigma = 0.05)$gen(500) %>%
-  mutate(j = 2)
-
-# Generate data for shape 2, 3 and 4. Reuse x and e.
-map_df(3:18, function(j) {
-  phn_model(j = j, include_x2 = TRUE, sigma = 0.05)$
-    gen(500, computed = select(dat_shape_1, x1, e)) %>%
-  mutate(j = j)
-}) %>%
-
-  # Combined with data for shape 1
-  bind_rows(dat_shape_1) %>%
-  mutate(j = factor(j)) %>%
-  VI_MODEL$plot(remove_axis = TRUE, remove_grid_line = TRUE, theme = theme_light()) +
-  facet_wrap(~j, scales = "free", labeller = label_parsed, ncol = 5)
-```
-
-```{r different-j-heter, fig.pos="!h", fig.cap = 'Residual plots of models violating both the non-linearity and the heteroskedasticity assumptions.
-The 17 shapes are generated by varying the order of polynomial given by $j$ in $He_j(.)$, and the "left-triangle" shape is introduced by setting $a = -1$.'}
-
-set.seed(10086)
-
-# Data for shape 1
-dat_shape_1 <- phn_model(j = 2, a = -1, b = 100, include_x2 = FALSE, sigma = 0.05)$gen(500) %>%
-  mutate(j = 2)
-
-# Generate data for shape 2, 3 and 4. Reuse x and e.
-map_df(3:18, function(j) {
-  phn_model(j = j, a = -1, b = 100, include_x2 = FALSE, sigma = 0.05)$
-    gen(500, computed = select(dat_shape_1, x1, e)) %>%
-  mutate(j = j)
-}) %>%
-
-  # Combined with data for shape 1
-  bind_rows(dat_shape_1) %>%
-  mutate(j = factor(j)) %>%
-  VI_MODEL$plot(remove_axis = TRUE, remove_grid_line = TRUE, theme = theme_light()) +
-  facet_wrap(~j, scales = "free", labeller = label_parsed, ncol = 5)
-```
-
-```{r different-e-heter, fig.pos="!h", fig.cap = 'Residual plots of models violating both the non-normality and the heteroskedasticity assumptions.
-The four shapes are generated by using four different error distributions including discrete, lognormal, normal and uniform, and the "left-triangle" shape is introduced by setting $a = -1$.
-'}
-
-set.seed(10085)
-
-# Data for shape 1
-dat_shape_1 <- phn_model(a = -1, b = 100, include_z = FALSE, include_x2 = FALSE, e = rand_uniform(-1.4, 1.4))$gen(500) %>%
-  mutate(e_dist = "uniform")
-
-dat_shape_2 <- phn_model(a = -1, b = 100, include_z = FALSE, include_x2 = FALSE, sigma = 0.8)$gen(500) %>%
-  mutate(e_dist = "normal")
-
-dat_shape_3 <- phn_model(a = -1, b = 100, include_z = FALSE, include_x2 = FALSE, e = rand_lognormal(sigma = 0.6))$gen(500) %>%
-  mutate(e_dist = "lognormal")
-
-dat_shape_4 <- phn_model(a = -1, b = 100, include_z = FALSE, include_x2 = FALSE, e = rand_uniform_d(-1.4, 1.4, even = TRUE))$gen(500) %>%
-  mutate(e_dist = "discrete")
-
-# Generate data for shape 2, 3 and 4. Reuse x and e.
-bind_rows(dat_shape_1, dat_shape_2, dat_shape_3, dat_shape_4) %>%
-  VI_MODEL$plot(remove_axis = TRUE, remove_grid_line = TRUE, theme = theme_light()) +
-  facet_wrap(~e_dist, scales = "free", labeller = label_parsed, ncol = 2)
-```
-
-Equation \ref{eq:data-sim} accommodates the incorporation of the second predictor $\boldsymbol{x}_2$.
-Introducing it into the data generation process by setting $\beta_1 = 1$ significantly enhances the complexity of the shapes, as illustrated in Figure \ref{fig:different-j-x2}.
-In comparison to Figure \ref{fig:different-j}, Figure \ref{fig:different-j-x2} demonstrates that the non-linear shape resembles a surface rather than a single curve.
-This augmentation can facilitate the computer vision model in learning visual patterns from residual plots of the multiple linear regression model.
-
-In real-world analysis, it is not uncommon to encounter instances where multiple model violations coexist.
-In such cases, the residual plots often exhibit a mixed pattern of visual anomalies corresponding to different types of model violations.
-Figure \ref{fig:different-j-heter} and Figure \ref{fig:different-e-heter} show the visual patterns of models with multiple model violations.
-
-The predictors, $\boldsymbol{x}_1$ and $\boldsymbol{x}_2$, are randomly generated from four distinct distributions, including $U(-1, 1)$ (uniform), $N(0, 0.3^2)$ (normal), $\text{lognormal}(0, 0.6^2)/3$ (skewed) and $U\{-1, 1\}$ (discrete uniform).
-
-## Balanced Dataset
-
-To train a robust computer vision model, we deliberately controlled the distribution of the target variable $D$ in the training data.
-We ensured that it followed a uniform distribution between $0$ and $7$.
-This was achieved by organizing $50$ buckets, each exclusively accepting training samples with $D$ falling within the range $[7(i - 1)/49, 7i/49)$ for $i < 50$, where $i$ represents the index of the $i$-th bucket.
-For the $50$-th bucket, any training samples with $D \geq 7$ were accepted.
-
-With 80,000 training images prepared, each bucket accommodated a maximum of $80000/ 50 = 1600$ training samples.
-The simulator iteratively sampled parameter values from the parameter space, generated residuals and fitted values using the data generation process, computed the distance, and checked if the sample fitted within the corresponding bucket.
-This process continued until all buckets were filled.
-
-Similarly, we adopted the same methodology to prepare 8,000 test images for performance evaluation and model diagnostics.
-
-# Model Architecture {#sec-model-architecture}
-
-```{r cnn-diag, fig.cap = "Diagram of the architecture of the optimized computer vision model.
-Numbers at the bottom of each box show the shape of the output of each layer.
-The band of each box drawn in a darker color indicates the use of the rectified linear unit activation function.
-Yellow boxes are 2D convolutional layers, orange boxes are pooling layers, the grey box is the concatenation layer, and the purple boxes are dense layers.", out.width = "100%"}
-magick::image_read_pdf("figures/cnn.pdf")
-```
-
-The architecture of the computer vision model is adapted from the well-established VGG16 architecture [@simonyan2014very].
-While more recent architectures like ResNet [@he2016deep] and DenseNet[@huang2017densely], have achieved even greater performance, VGG16 remains a solid choice for many applications due to its simplicity and effectiveness.
-Our decision to use VGG16 aligns with our goal of starting with a proven and straightforward model.
-Figure \ref{fig:cnn-diag} provides a diagram of the architecture.
-More details about the neural network layers used in this study are provided in Appendix A.
-
-The model begins with an input layer of shape $n \times h \times w \times 3$, capable of handling $n$ RGB images.
-This is followed by a grayscale conversion layer utilizing the luma formula under the Rec.
-601 standard [@series2011studio], which converts the color image to grayscale.
-Grayscale suffices for our task since data points are plotted in black.
-We experiment with three combinations of $h$ and $w$: $32 \times 32$, $64 \times 64$, and $128 \times 128$, aiming to achieve sufficiently high image resolution for the problem at hand.
-
-The processed image is used as the input for the first convolutional block.
-The model comprises at most five consecutive convolutional blocks, mirroring the original VGG16 architecture.
-Within each block, there are two 2D convolutional layers followed by two activation layers, respectively.
-Subsequently, a 2D max-pooling layer follows the second activation layer.
-The 2D convolutional layer convolves the input with a fixed number of $3 \times 3$ convolution filters, while the 2D max-pooling layer downsamples the input along its spatial dimensions by taking the maximum value over a $2 \times 2$ window for each channel of the input.
-The activation layer employs the rectified linear unit (ReLU) activation function, a standard practice in deep learning, which introduces a non-linear transformation of the output of the 2D convolutional layer.
-Additionally, to regularize training, a batch normalization layer is added after each 2D convolutional layer and before the activation layer.
-Finally, a dropout layer is appended at the end of each convolutional block to randomly set some inputs to zero during training, further aiding in regularization.
-
-The output of the last convolutional block is summarized by either a global max pooling layer or a global average pooling layer, resulting in a two-dimensional tensor.
-To leverage the information contained in scagnostics, this tensor is concatenated with an additional $n \times 5$ tensor, which contains the "Monotonic", "Sparse", "Splines", and "Striped" measures, along with the number of observations for $n$ residual plots.
-
-The concatenated tensor is then fed into the final prediction block.
-This block consists of two fully-connected layers.
-The first layer contains at least $128$ units, followed by a dropout layer.
-Occasionally, a batch normalization layer is inserted between the fully-connected layer and the dropout layer for regularization purposes.
-The second fully-connected layer consists of only one unit, serving as the output of the model.
-
-The model weights $\boldsymbol{\theta}$ were randomly initialized and they were optimized by the Adam optimizer [@kingma2014adam] with the mean square error loss function
-
-$$\hat{\boldsymbol{\theta}} = \underset{\boldsymbol{\theta}}{\text{arg min}}\frac{1}{n_{\text{train}}}\sum_{i=1}^{n_{\text{train}}}(D_i - f_{\boldsymbol{\theta}}(V_i, S_i))^2,$$
-
-\noindent where $n_{\text{train}}$ is the number of training samples, $V_i$ is the $i$-th residual plot and $S_i$ is the additional information about the $i$-th residual plot including four scagnostics and the number of observations.
-
-# Model Training {#sec-model-training}
-
-To achieve a near-optimal deep learning model, hyperparameters like the learning rate often need to be fine-tuned using a tuner.
-In our study, we utilized the Bayesian optimization tuner from the `KerasTuner` Python library [@omalley2019kerastuner] for this purpose.
-A comprehensive list of hyperparameters is provided in Table \ref{tab:hyperparameter}.
-
-The number of base filters determines the number of filters for the first 2D convolutional layer.
-In the VGG16 architecture, the number of filters for the 2D convolutional layer in a block is typically twice the number in the previous block, except for the last block, which maintains the same number of convolution filters as the previous one.
-This hyperparameter aids in controlling the complexity of the computer vision model.
-A higher number of base filters results in more trainable parameters.
-Likewise, the number of units for the fully-connected layer determines the complexity of the final prediction block.
-Increasing the number of units enhances model complexity, resulting in more trainable parameters.
-
-The dropout rate and batch normalization are flexible hyperparameters that work in conjunction with other parameters to facilitate smooth training.
-A higher dropout rate is necessary when the model tends to overfit the training dataset by learning too much noise [@srivastava2014dropout].
-Conversely, a lower dropout rate is preferred when the model is complex and challenging to converge.
-Batch normalization, on the other hand, addresses the internal covariate shift problem arising from the randomness in weight initialization and input data [@goodfellow2016deep].
-It helps stabilize and accelerate the training process by normalizing the activations of each layer.
-
-Additionally, incorporating additional inputs such as scagnostics and the number of observations can potentially enhance prediction accuracy.
-Therefore, we allow the tuner to determine whether these inputs were necessary for optimal model performance.
-
-The learning rate is a crucial hyperparameter, as it dictates the step size of the optimization algorithm.
-A high learning rate can help the model avoid local minima but risks overshooting and missing the global minimum.
-Conversely, a low learning rate smoothens the training process but makes the convergence time longer and increases the likelihood of getting trapped in local minima.
-
-Our model was trained on the MASSIVE M3 high-performance computing platform [@goscinski2014multi], using TensorFlow [@abadi2016tensorflow] and Keras [@chollet2015keras].
-During training, 80% of the training data was utilized for actual training, while the remaining 20% was used as validation data.
-The Bayesian optimization tuner conducted 100 trials to identify the best hyperparameter values based on validation root mean square error.
-The tuner then restored the best epoch of the best model from the trials.
-Additionally, we applied early stopping, terminating the training process if the validation root mean square error fails to improve for 50 epochs.
-The maximum allowed epochs was set at 2,000, although no models reached this threshold.
-
-```{r}
-data.frame(`Hyperparameter` = c("Number of base filters", "Dropout rate for convolutional blocks", "Batch normalization for convolutional blocks", "Type of global pooling", "Ignore additional inputs", "Number of units for the fully-connected layer", "Batch normalization for the fully-connected layer", "Dropout rate for the fully-connected layer", "Learning rate"),
-           Domain = c("\\{4, 8, 16, 32, 64\\}", "[0.1, 0.6]", "\\{false, true\\}", "\\{max, average\\}", "\\{false, true\\}", "\\{128, 256, 512, 1024, 2048\\}", "\\{false, true\\}", "[0.1, 0.6]", "[$10^{-8}$, $10^{-1}$]")) %>%
-  kableExtra::kable(format = "latex",
-                    booktabs = TRUE, 
-                    label = "hyperparameter",
-                    caption = "Name of hyperparameters and their correspoding domain for the computer vision model.",
-                    escape = FALSE)
-```
-
-Based on the tuning process described above, the optimized hyperparameter values are presented in Table \ref{tab:best-hyperparameter}.
-It was observable that a minimum of $32$ base filters was necessary, with the preferable choice being $64$ base filters for both the $64 \times 64$ and $128 \times 128$ models, mirroring the original VGG16 architecture.
-The optimized dropout rate for convolutional blocks hovered around $0.4$, and incorporating batch normalization for convolutional blocks proved beneficial for performance.
-
-All optimized models chose to retain the additional inputs, contributing to the reduction of validation error.
-The number of units required for the fully-connected layer was $256$, a relatively modest number compared to the VGG16 classifier, suggesting that the problem at hand was less complex.
-The optimized learning rates were higher for models with higher resolution input, likely because models with more parameters are more prone to getting stuck in local minima, requiring a higher learning rate.
-
-```{r}
-data.frame(`Hyperparameter` = c("Number of base filters", "Dropout rate for convolutional blocks", "Batch normalization for convolutional blocks", "Type of global pooling", "Ignore additional inputs", "Number of units for the fully-connected layer", "Batch normalization for the fully-connected layer", "Dropout rate for the fully-connected layer", "Learning rate"),
-           `32` = c("32", "0.4", "true", "max", "false", "256", "false", "0.2", "0.0003"),
-           `64` = c("64", "0.3", "true", "average", "false", "256", "true", "0.4", "0.0006"),
-           `128` = c("64", "0.4", "true", "average", "false", "256", "true", "0.1", "0.0052")) %>%
-  kableExtra::kable(format = "latex",
-                    booktabs = TRUE, 
-                    label = "best-hyperparameter",
-                    caption = "Hyperparameters values for the optimized computer vision models with different input sizes.",
-                    col.names = c("Hyperparameter", "$32 \\times 32$", "$64 \\times 64$", "$128 \\times 128$"),
-                    escape = FALSE) %>%
-  kableExtra::kable_styling(latex_options = c("scale_down"))
-```
-
-# Results {#sec-model-results}
-
-## Model Performance
-
-The test performance for the optimized models with three different input sizes are summarized in Table \ref{tab:performance}.
-Among these models, the $32 \times 32$ model consistently exhibited the best test performance.
-The mean absolute error of the $32 \times 32$ model indicated that the difference between $\hat{D}$ and $D$ was approximately $0.43$ on the test set, a negligible deviation considering the normal range of $D$ typically falls between $0$ and $7$.
-The high $R^2$ values also suggested that the predictions were largely linearly correlated with the target.
-
-Figure \ref{fig:model-performance} presents a hexagonal heatmap for $D - \hat{D}$ versus $D$.
-The brown smoothing curves, fitted by generalized additive models [@hastie2017generalized], demonstrate that all the optimized models perform admirably on the test sets when $1.5 < D < 6$, where no structural issues are noticeable.
-However, over-predictions occurred when $D < 1.5$, while under-predictions occurred predominantly when $\hat{D} > 6$.
-
-For input images representing null plots where $D = 0$, it was expected that the models will over-predict the distance, as explained in Section \ref{sec-model-lineup-evaluation}.
-However, it can not explain the under-prediction issue.
-Therefore, we analysed the relationship between residuals and all the factors involved in the data generating process.
-We found that most issues actually arose from non-linearity problems and the presence of a second predictor in the regression model as illustrated in Figure \ref{fig:over-under}.
-When the variance for the error distribution was small, the optimized model tended to under-predict the distance.
-Conversely, when the error distribution had a large variance, the model tended to over-predict the distance.
-
-Since most of the deviation stemmed from the presence of non-linearity violations, to further investigate this, we split the test set based on violation types and re-evaluated the performance, as detailed in Table \ref{tab:performance-sub}.
-It was evident that metrics for null plots were notably worse compared to other categories.
-Furthermore, residual plots solely exhibiting non-normality issues were the easiest to predict, with very low test root mean square error (RMSE) at around $0.3$.
-Residual plots with non-linearity issues were more challenging to assess than those with heteroskedasticity or non-normality issues.
-When multiple violations were introduced to a residual plot, the performance metrics typically lay between the metrics for each individual violation.
-
-Based on the model performance metrics, we chose to use the best-performing model evaluated on the test set, namely the $32 \times 32$ model, for the subsequent analysis.
-
-```{r}
-test_pred <- readr::read_csv(here::here("data/test_pred.csv"))
-train_pred <- readr::read_csv(here::here("data/train_pred.csv"))
-meta <- readr::read_csv(here::here("data/meta.csv"))
-
-test_summary <- test_pred %>% 
-  left_join(meta) %>%
-  group_by(res) %>%
-  summarise(RMSE = yardstick::rmse_vec(effect_size, vss),
-            R2 = yardstick::rsq_vec(effect_size, vss),
-            MAE = yardstick::mae_vec(effect_size, vss),
-            HUBER = yardstick::huber_loss_vec(effect_size, vss)) %>%
-  mutate(type = "test")
-
-train_summary <- train_pred %>% 
-  left_join(meta) %>%
-  group_by(res) %>%
-  summarise(RMSE = yardstick::rmse_vec(effect_size, vss),
-            R2 = yardstick::rsq_vec(effect_size, vss),
-            MAE = yardstick::mae_vec(effect_size, vss),
-            HUBER = yardstick::huber_loss_vec(effect_size, vss)) %>%
-  mutate(type = "train")
-```
-
-```{r}
-test_summary %>%
-  select(res, RMSE, R2, MAE, HUBER) %>%
-  mutate(res = glue::glue("${res} \\times {res}$")) %>%
-  mutate(across(RMSE:MAE, ~format(.x, digits = 3))) %>%
-  mutate(across(HUBER, ~format(.x, digits = 2))) %>%
-  kableExtra::kable(format = "latex",
-                    booktabs = TRUE, 
-                    label = "performance",
-                    caption = "The test performance of three optimized models with different input sizes.",
-                    escape = FALSE,
-                    align = "lrrrr",
-                    col.names = c("", "RMSE", "$R^2$", "MAE", "Huber loss"))
-```
-
-```{r model-performance, fig.pos="!h", fig.cap = "Hexagonal heatmap for difference in $D$ and $\\hat{D}$ vs $D$ on test data for three optimized models with different input sizes.
-The brown lines are smoothing curves produced by fitting generalized additive models.
-The area over the zero line in light yellow indicates under-prediction, and the area under the zero line in light green indicates over-prediction.", dev = 'png', dpi = 300}
-model_pred <- train_pred %>% 
-  left_join(meta) %>%
-  mutate(type = "train") %>%
-  bind_rows(test_pred %>% 
-              left_join(meta) %>%
-              mutate(type = "test"))
-
-model_pred %>%
-  filter(type == "test") %>%
-  ggplot() +
-  geom_hline(yintercept = 0, alpha = 0.5) +
-  annotate("rect", 
-           ymin = min(model_pred$effect_size - model_pred$vss),
-           ymax = 0,
-           xmin = min(model_pred$effect_size),
-           xmax = max(model_pred$effect_size),
-            fill = "#40B0A6",
-            alpha = 0.3) +
-  geom_point(data = NULL, aes(5, 0, col = "over-prediction"), shape = 15) +
-  annotate("rect", 
-           ymin = 0,
-           ymax = max(model_pred$effect_size - model_pred$vss),
-           xmin = min(model_pred$effect_size),
-           xmax = max(model_pred$effect_size),
-           fill = "#E1BE6A",
-           alpha = 0.3) +
-  geom_point(data = NULL, aes(5, 0, col = "under-prediction"), shape = 15) +
-  geom_hex(aes(effect_size, effect_size - vss), bins = 20) +
-  geom_smooth(aes(effect_size, effect_size - vss), se = FALSE, col = "#994F00") +
-  facet_wrap(~ res) +
-  coord_fixed() +
-  ylab(expression(D - hat(D))) +
-  xlab(expression(D)) +
-  theme_light() +
-  scale_fill_continuous(limits = c(1, 6000), trans = "log10", low = "#56B1F7", high = "#132B43") +
-  scale_color_manual(values = c("under-prediction" = scales::alpha("#E1BE6A", 0.3),
-                                "over-prediction" = scales::alpha("#40B0A6", 0.3)),
-                     breaks = c("under-prediction", "over-prediction")) +
-  guides(col = guide_legend(title = NULL, override.aes = list(size = 8), order = 1))
-```
-
-```{r over-under, fig.pos="!h", fig.cap = "Scatter plots for difference in $D$ and $\\hat{D}$ vs $\\sigma$ on test data for the $32 \\times 32$ optimized model.
-The data is grouped by whether the regression has only non-linearity violation, and whether it includes a second predictor in the regression formula.
-The brown lines are smoothing curves produced by fitting generalized additive models.
-The area over the zero line in light yellow indicates under-prediction, and the area under the zero line in light green indicates over-prediction.", dev = 'png', dpi = 300}
-model_pred %>%
-  filter(res == 32L) %>%
-  filter(type == "test") %>%
-  mutate(only_z = include_z & !include_heter & !include_non_normal) %>%
-  rename(`Only has non-linearity violation` = only_z, `Has second predictor` = include_x2) %>%
-  ggplot() +
-  geom_hline(yintercept = 0, alpha = 0.5) +
-  annotate("rect", 
-           ymin = min(model_pred$effect_size[model_pred$type == "test"] - model_pred$vss[model_pred$type == "test"]),
-                ymax = 0,
-                xmin = min(model_pred$e_sigma),
-                xmax = max(model_pred$e_sigma),
-            fill = "#40B0A6",
-            alpha = 0.3) +
-  annotate("rect", 
-           ymin = 0,
-                ymax = max(model_pred$effect_size[model_pred$type == "test"] - model_pred$vss[model_pred$type == "test"]),
-                xmin = min(model_pred$e_sigma),
-                xmax = max(model_pred$e_sigma),
-            fill = "#E1BE6A",
-            alpha = 0.3) +
-  geom_point(aes(e_sigma, effect_size - vss), alpha = 0.2) +
-  geom_smooth(aes(e_sigma, effect_size - vss), se = FALSE, col = "#994F00") +
-  facet_grid(`Has second predictor` ~ `Only has non-linearity violation`,
-             labeller = label_both) +
-  theme_light() +
-  ylab(expression(D - hat(D))) +
-  xlab(expression(sigma))
-```
-
-```{r}
-model_pred %>%
-  filter(res == 32L) %>%
-  group_by(include_non_normal, include_heter, include_z) %>%
-  summarise(train_n = sum(type == "train"),
-            train_RMSE = yardstick::rmse_vec(effect_size[type == "train"], vss[type == "train"]),
-            test_n = sum(type == "test"),
-            test_RMSE = yardstick::rmse_vec(effect_size[type == "test"], vss[type == "test"])) %>%
-  mutate(violations = ifelse(include_z, "non-linearity", "null")) %>%
-  mutate(violations = ifelse(include_heter, glue("{violations} + heteroskedasticity"), violations)) %>%
-  mutate(violations = ifelse(include_non_normal, glue("{violations} + non-normality"), violations)) %>%
-  mutate(violations = gsub("null \\+ ", "", violations)) %>%
-  ungroup() %>%
-  select(violations, train_n, train_RMSE, test_n, test_RMSE) %>%
-  mutate(train_n = format(train_n)) %>%
-  mutate(test_n = format(test_n)) %>%
-  mutate(across(c(train_RMSE, test_RMSE), ~format(.x, digits = 3))) %>%
-  mutate(violations = ifelse(violations == "null", "no violations", violations)) %>%
-  select(-train_n, -train_RMSE) %>%
-  kableExtra::kable(format = "latex",
-                    booktabs = TRUE,
-                    label = "performance-sub",
-                    caption = "The training and test performance of the $32 \\times 32$ model presented with different model violations.",
-                    escape = FALSE,
-                    linesep = "", 
-                    align = "lrr",
-                    col.names = c("Violations", "\\#samples", "RMSE"))
-```
-
-<!-- - check if j has an impact on residuals more carefully -->
-<!-- - the RMSE has very little to do with j if x2 is not included. And its impact is not as large as sigma. -->
-
-## Comparison with Human Visual Inference and Conventional Tests
-
-### Overview of the Human Subject Experiment
-
-In order to check the validity of the proposed computer vision model, residual plots presented in the human subject experiment conducted by @li2024plot will be assessed.
-
-<!-- The experiment revealed that conventional tests are more sensitive to weak departures from model assumptions than visual test, and they often reject the null hypothesis when departures are not visibly different from null residual plots.
--->
-
-This study has collected 7,974 human responses to 1,152 lineups.
-Each lineup contains one randomly placed true residual plot and 19 null plots.
-Among the 1,152 lineups, 24 are attention check lineups in which the visual patterns are designed to be extremely obvious and very different from the corresponding to null plots, 36 are null lineups where all the lineups consist of only null plots, 279 are lineups with uniform predictor distribution evaluated by 11 participants, and the remaining 813 are lineups with discrete, skewed or normal predictor distribution evaluated by 5 participants.
-Attention check lineups and null lineups will not be assessed in the following analysis.
-
-In @li2024plot, the residual plots are simulated from a data generating process which is a special case of Equation \ref{eq:data-sim}.
-The main characteristic is the model violations are introduced separately, meaning non-linearity and heteroskedasticity will not co-exist in one lineup but assigned uniformly to all lineups.
-Additionally, non-normality and multiple predictors are not considered in the experimental design.
-
-### Model Performance on the Human-evaluated Data
-
-```{r rd-human}
-vss_32 <- readRDS(here::here("data/vss_32.rds"))
-
-experiment <- vi_survey %>%
-  group_by(unique_lineup_id) %>%
-  summarise(across(everything(), first)) %>%
-  select(unique_lineup_id, attention_check, null_lineup, prop_detect,
-         answer, effect_size, conventional_p_value,
-         p_value, type, shape, a, b, x_dist, e_dist,
-         e_sigma, include_z, k, n)
-
-experiment <- vss_32 %>%
-  left_join(experiment) %>%
-  mutate(conventional_reject = conventional_p_value <= 0.05) %>%
-  mutate(reject = p_value <= 0.05) %>%
-  mutate(model_reject = vss_p_value <= 0.05) %>%
-  mutate(model_boot_reject = vss_boot_p_value <= 0.05)
-
-experiment <- experiment %>%
-  left_join(readRDS(here::here("data/actual_ss.rds")))
-```
-
-```{r}
-experiment %>%
-  filter(!attention_check) %>%
-  filter(!null_lineup) %>%
-  mutate(type = ifelse(type == "polynomial", "non-linearity", type)) %>%
-  group_by(type) %>%
-  summarise(RMSE = yardstick::rmse_vec(actual_ss, vss),
-            R2 = yardstick::rsq_vec(actual_ss, vss),
-            MAE = yardstick::mae_vec(actual_ss, vss),
-            HUBER = yardstick::huber_loss_vec(actual_ss, vss)) %>%
-  mutate(across(RMSE:HUBER, ~format(.x, digits = 3))) %>%
-  kableExtra::kable(format = "latex",
-                    booktabs = TRUE, 
-                    label = "experiment-performance",
-                    caption = "The performance of the $32 \\times 32$ model on the data used in the human subject experiment.",
-                    escape = FALSE,
-                    align = "lrrrr",
-                    col.names = c("Violation", "RMSE", "$R^2$", "MAE", "Huber loss"))
-```
-
-For each lineup used in @li2024plot, there is one true residual plot and 19 null plots.
-While the distance $D$ for the true residual plot depends on the underlying data generating process, the distance $D$ for the null plots is zero.
-We have used our optimized computer vision model to estimate distance for both the true residual plots and the null plots.
-To have a fair comparison, $H_0$ will be rejected if the true residual plot has the greatest estimated distance among all plots in a lineup.
-Additionally, the appropriate conventional tests including the Ramsey Regression Equation Specification Error Test (RESET) [@ramsey1969tests] for non-linearity and the Breusch-Pagan test  [@breusch1979simple] for heteroskedasticity were applied on the same data for comparison.
-
-The performance metrics of $\hat{D}$ for true residual plots are outlined in Table \ref{tab:experiment-performance}.
-It is notable that all performance metrics are slightly worse than those evaluated on the test data.
-Nevertheless, the mean absolute error remains at a low level, and the linear correlation between the prediction and the true value remains very high.
-Consistent with results in Table \ref{tab:performance-sub}, lineups with non-linearity issues are more challenging to predict than those with heteroskedasticity issues.
-
-Table \ref{tab:human-conv-table} provides a summary of the agreement between decisions made by the computer vision model and conventional tests.
-The agreement rates between conventional tests and the computer vision model are 85.95% and 79.69% for residual plots containing heteroskedasticity and non-linearity patterns, respectively.
-These figures are higher than those calculated for visual tests conducted by human, indicating that the computer vision model exhibits behavior more akin to the best available conventional tests.
-However, Figure \ref{fig:conv-mosaic} shows that the computer vision model does not always reject when the conventional tests reject.
-And a small number of plots will be rejected by computer vision model but not by conventional tests.
-This suggests that conventional tests are more sensitive than the computer vision model.
-
-Figure \ref{fig:pcp} further illustrates the decisions made by visual tests conducted by human, computer vision models, and conventional tests, using a parallel coordinate plots.
-It can be observed that all three tests will agree with each other for around 50% of the cases.
-When visual tests conducted by human do not reject, there are substantial amount of cases where computer vision model also do not reject but conventional tests reject.
-There are much fewer cases that do not reject by visual tests and conventional tests, but is rejected by computer vision models.
-This indicates computer vision model can behave like visual tests conducted by human better than conventional tests.
-Moreover, there are great proportion of cases where visual tests conducted by human is the only test who does not reject.
-
-When plotting the decision against the distance, as illustrated in Figure \ref{fig:power}, several notable observations emerge.
-Firstly, compared to conventional tests, the computer vision model tends to have fewer rejected cases when $D < 2$ and fewer non-rejected cases when $2< D < 4$.
-This suggests tests based on the computer vision model are less sensitive to small deviations from model assumptions than conventional tests but more sensitive to moderate deviations.
-Additionally, visual tests demonstrate the lowest sensitivity to residual plots with small distances where not many residual plots are rejected when $D < 2$.
-Similarly, for large distances where $D > 4$, almost all residual plots are rejected by the computer vision model and conventional tests, but for visual tests conducted by humans, the threshold is higher with $D > 5$.
-
-In Figure \ref{fig:power}, rejection decisions are fitted by logistic regression models with no intercept terms and an offset equals to $\text{log}(0.05/0.95)$.
-The fitted curves for the computer vision model fall between those of conventional tests and visual tests for both non-linearity and heteroskedasticity, which means there is still potential to refine the computer vision model to better align its behavior with visual tests conducted by humans.
-
-In the experiment conducted in @li2024plot, participants were allowed to make multiple selections for a lineup.
-The weighted detection rate was computed by assigning weights to each detection.
-If the participant selected zero plots, a weight of 0.05 was assigned; otherwise, if the true residual plot was detected, the weight was 1 divided by the number of selections.
-This weighted detection rate allow us to assess the quality of the distance measure purposed in this paper, by using the $\delta$-difference statistic.
-The $\delta$-difference is originally defined by @chowdhury2018measuring, is given by
-
-$$
-\delta = \bar{d}_{\text{true}} - \underset{j}{\text{max}}\left(\bar{d}_{\text{null}}^{(j)}\right) \quad \text{for}~j = 1,...,m-1,
-$$
-
-where $\bar{d}_{\text{null}}^{(j)}$ is the mean distance between the $j$-th null plot and the other null plots, $\bar{d}_{\text{true}}$ is the mean distance between the true residual plot and null plots, and $m$ is the number of plots in a lineup.
-These mean distances are used because, as noted by @chowdhury2018measuring, the distances can vary depending on which data plot is used for comparison.
-For instance, with three null plots, A, B and C, the distance between A and B may differ from the distance between A and C.
-To obtain a consistent distance for null plot A, averaging is necessary.
-However, this approach is not applicable to the distance proposed in this paper, as we only compare the residual plot against a theoretically good residual plot.
-Consequently, the statistic must be adjusted to evaluate our distance measure effectively.
-
-One important aspect that the $\delta$-difference was designed to capture is the empirical distribution of distances for null plot.
-If we were to replace the mean distances $\bar{d}_{\text{null}}^{(j)}$ directly with $D_{\text{null}}^{(j)}$, the distance of the $j$-th null plot, the resulting distribution would be degenerate, since $D_{null}$ equals zero by definition.
-Additionally, $D$ can not be derived from an image, meaning it falls outside the scope of the distances considered by @chowdhury2018measuring.
-Instead, the focus should be on the empirical distribution of $\hat{D}$, as it influences decision-making.
-Therefore, the adjusted $\delta$-different is defined as
-
-$$
-\delta_{\text{adj}} = \hat{D} - \underset{j}{\text{max}}\left(\hat{D}_{\text{null}}^{(j)}\right) \quad \text{for}~j = 1,...,m-1,
-$$
-
-\noindent where $\hat{D}_{\text{null}}^{(j)}$ is the estimated distance for the $j$-th null plot, and $m$ is the number of plots in a lineup.
-
-Figure \ref{fig:delta} displays the scatter plot of the weighted detection rate vs the adjusted $\delta$-difference.
-It indicates that the weighted detection rate increases as the adjusted $\delta$-difference increases, particularly when the adjusted $\delta$-difference is greater than zero.
-A negative adjusted $\delta$-difference suggests that there is at least one null plot in the lineup with a stronger visual signal than the true residual plot.
-In some instances, the weighted detection rate is close to one, yet the adjusted $\delta$-difference is negative.
-This discrepancy implies that the distance measure, or the estimated distance, may not perfectly reflect actual human behavior.
-
-```{r hist-null-human}
-lineup_vss <- readRDS(here::here("data/lineup_vss.rds"))
-lineup_vss <- lineup_vss %>%
-  group_by(unique_lineup_id) %>%
-  mutate(delta_diff = vss[!null] - max(vss[null])) %>%
-  mutate(gamma_diff = sum(vss[null] > vss[!null]))
-
-experiment <- experiment %>%
-  left_join(lineup_vss %>%
-  filter(null == FALSE) %>%
-  mutate(model_reject_20 = rank == 1) %>%
-  select(unique_lineup_id, rank, model_reject_20))
-```
-
-```{r cache = TRUE}
-vi_lineup <- get_vi_lineup()
-```
-
-```{r}
-bind_rows(
-  experiment %>%
-    filter(!attention_check) %>%
-    filter(!null_lineup) %>%
-    mutate(type = ifelse(type == "polynomial", "non-linearity", type)) %>%
-    group_by(type) %>%
-    summarise(n = format(n()), agree = format(sum(model_reject == conventional_reject)), rate = format(mean(conventional_reject == model_reject), digits = 4)),
-  experiment %>%
-    filter(!attention_check) %>%
-    filter(!null_lineup) %>%
-    mutate(type = ifelse(type == "polynomial", "non-linearity", type)) %>%
-    group_by(type) %>%
-    summarise(n = format(n()), agree = format(sum(model_reject == reject)), rate = format(mean(model_reject == reject), digits = 4))
-  ) %>%
-  kableExtra::kable(format = "latex",
-                    booktabs = TRUE,
-                    label = "human-conv-table",
-                    caption = "Summary of the comparison of decisions made by computer vision model with decisions made by conventional tests and visual tests conducted by human.",
-                    escape = FALSE,
-                    align = "lrrr",
-                    col.names = c("Violations", "\\#Samples", "\\#Agreements", "Agreement rate")) %>%
-  kableExtra::pack_rows("Compared with conventional tests", 1, 2) %>%
-  kableExtra::pack_rows("Compared with visual tests conducted by human", 3, 4)
-```
-
-```{r conv-mosaic, fig.pos = "!h", fig.cap = "Rejection rate ($p$-value $\\leq0.05$) of computer vision models conditional on conventional tests on non-linearity (left) and heteroskedasticity (right) lineups displayed using a mosaic plot.
-When the conventional test fails to reject, the computer vision mostly fails to reject the same plot as well as indicated by the height of the top right yellow rectangle, but there are non negliable amount of plots where the conventional test rejects but the computer vision model fails to reject as indicated by the width of the top left yellow rectangle."}
-library(ggmosaic)
-
-experiment %>%
-  filter(!attention_check) %>%
-  filter(!null_lineup) %>%
-  mutate(type = ifelse(type == "polynomial", "non-linearity", type)) %>%
-  mutate(type = ifelse(type == "non-linearity", "Non-linearity", "Heteroskedasticity")) %>%
-  mutate(type = factor(type, levels = c("Non-linearity", "Heteroskedasticity"))) %>%
-  mutate(model_reject_20 = ifelse(model_reject_20, "Reject", "No")) %>%
-  mutate(conventional_reject = ifelse(conventional_reject, "Reject", "No")) %>%
-  mutate(across(c(model_reject_20, conventional_reject), ~factor(.x, levels = c("Reject", "No")))) %>%
-  ggplot() +
-  geom_mosaic(aes(x = ggmosaic::product(model_reject_20, conventional_reject), 
-                  fill = model_reject_20)) +
-  facet_grid(~type) +
-  ylab("Computer vision model rejects") +
-  xlab("Conventional tests reject ") +
-  # labs(fill = "Visual tests reject") +
-  scale_fill_manual(values = c("#40B0A6", "#E1BE6A")) + 
-  theme_bw() +
-  theme(legend.position = "none") +
-  coord_fixed()
-```
-
-```{r power, fig.pos = "!h", fig.cap = "Comparison of power of visual tests, conventional tests and the computer vision model.
-Marks along the x-axis at the bottom of the plot represent rejections made by each type of test.
-Marks at the top of the plot represent acceptances.
-Power curves are fitted by logistic regression models with no intercept but an offset equals to $\\text{log}(0.05/0.95)$."}
-
-max_actual_ss <- max(filter(experiment, !attention_check, !null_lineup)$actual_ss)
-max_poly_ss <- max(filter(experiment, !attention_check, !null_lineup, type == "polynomial")$actual_ss)
-max_heter_ss <- max(filter(experiment, !attention_check, !null_lineup, type != "polynomial")$actual_ss)
-
-model_glm_pred_poly <- glm(model_reject_20 ~ actual_ss - 1, 
-    data = filter(experiment, !attention_check, !null_lineup, type == "polynomial") %>% mutate(offset0 = log(0.05/0.95)), 
-    offset = offset0,
-    family = binomial()) %>%
-  predict(data.frame(actual_ss = seq(0, max_poly_ss, 0.01), offset0 = log(0.05/0.95)),
-          type = "response") %>%
-  data.frame(d = seq(0, max_poly_ss, 0.01), type = "Non-linearity", name = "Computer vision model", value = .) %>%
-  mutate(type = factor(type, levels = c("Non-linearity", "Heteroskedasticity")))
-
-visual_glm_pred_poly <- glm(reject ~ actual_ss - 1, 
-    data = filter(experiment, !attention_check, !null_lineup, type == "polynomial") %>% mutate(offset0 = log(0.05/0.95)), 
-    offset = offset0,
-    family = binomial()) %>%
-  predict(data.frame(actual_ss = seq(0, max_poly_ss, 0.01), offset0 = log(0.05/0.95)),
-          type = "response") %>%
-  data.frame(d = seq(0, max_poly_ss, 0.01), type = "Non-linearity", name = "Visual test", value = .) %>%
-  mutate(type = factor(type, levels = c("Non-linearity", "Heteroskedasticity")))
-
-conv_glm_pred_poly <- glm(conventional_reject ~ actual_ss - 1, 
-    data = filter(experiment, !attention_check, !null_lineup, type == "polynomial") %>% mutate(offset0 = log(0.05/0.95)), 
-    offset = offset0,
-    family = binomial()) %>%
-  predict(data.frame(actual_ss = seq(0, max_poly_ss, 0.01), offset0 = log(0.05/0.95)),
-          type = "response") %>%
-  data.frame(d = seq(0, max_poly_ss, 0.01), type = "Non-linearity", name = "Conventional test", value = .) %>%
-  mutate(type = factor(type, levels = c("Non-linearity", "Heteroskedasticity")))
-
-model_glm_pred_heter <- glm(model_reject_20 ~ actual_ss - 1, 
-    data = filter(experiment, !attention_check, !null_lineup, type != "polynomial") %>% mutate(offset0 = log(0.05/0.95)), 
-    offset = offset0,
-    family = binomial()) %>%
-  predict(data.frame(actual_ss = seq(0, max_heter_ss, 0.01), offset0 = log(0.05/0.95)),
-          type = "response") %>%
-  data.frame(d = seq(0, max_heter_ss, 0.01), type = "Heteroskedasticity", name = "Computer vision model", value = .) %>%
-  mutate(type = factor(type, levels = c("Non-linearity", "Heteroskedasticity")))
-
-visual_glm_pred_heter <- glm(reject ~ actual_ss - 1, 
-    data = filter(experiment, !attention_check, !null_lineup, type != "polynomial") %>% mutate(offset0 = log(0.05/0.95)), 
-    offset = offset0,
-    family = binomial()) %>%
-  predict(data.frame(actual_ss = seq(0, max_heter_ss, 0.01), offset0 = log(0.05/0.95)),
-          type = "response") %>%
-  data.frame(d = seq(0, max_heter_ss, 0.01), type = "Heteroskedasticity", name = "Visual test", value = .) %>%
-  mutate(type = factor(type, levels = c("Non-linearity", "Heteroskedasticity")))
-
-conv_glm_pred_heter <- glm(conventional_reject ~ actual_ss - 1, 
-    data = filter(experiment, !attention_check, !null_lineup, type != "polynomial") %>% mutate(offset0 = log(0.05/0.95)), 
-    offset = offset0,
-    family = binomial()) %>%
-  predict(data.frame(actual_ss = seq(0, max_heter_ss, 0.01), offset0 = log(0.05/0.95)),
-          type = "response") %>%
-  data.frame(d = seq(0, max_heter_ss, 0.01), type = "Heteroskedasticity", name = "Conventional test", value = .) %>%
-  mutate(type = factor(type, levels = c("Non-linearity", "Heteroskedasticity")))
-
-mutate(experiment, type = ifelse(type == "polynomial", "Non-linearity", "Heteroskedasticity")) %>%
-  mutate(type = factor(type, levels = c("Non-linearity", "Heteroskedasticity"))) %>%
-  filter(!null_lineup, !attention_check) %>%
-ggplot() +
-  geom_line(data = model_glm_pred_poly, aes(d, value, col = "Computer vision model"), linewidth = 0.8) +
-  geom_line(data = model_glm_pred_heter, aes(d, value, col = "Computer vision model"), linewidth = 0.8) +
-  geom_line(data = visual_glm_pred_poly, aes(d, value, col = "Visual test"), linewidth = 0.8) +
-  geom_line(data = visual_glm_pred_heter, aes(d, value, col = "Visual test"), linewidth = 0.8) +
-  geom_line(data = conv_glm_pred_poly, aes(d, value, col = "Conventional test"), linewidth = 0.8) +
-  geom_line(data = conv_glm_pred_heter, aes(d, value, col = "Conventional test"), linewidth = 0.8) +
-  geom_segment(data = ~filter(.x, !reject), aes(x = actual_ss, xend = actual_ss, y = -0.1, yend = -0.1 + 0.03, col = "Visual test"), alpha = 0.4) +
-  geom_segment(data = ~filter(.x, reject), aes(x = actual_ss, xend = actual_ss, y = 1.1 - 0.03 * 2, yend = 1.1 - 0.03 * 3, col = "Visual test"), alpha = 0.4) +
-  geom_segment(data = ~filter(.x, !model_reject_20), aes(x = actual_ss, xend = actual_ss, y = -0.1 + 0.03, yend = -0.1 + 0.03 * 2, col = "Computer vision model"), alpha = 0.4) +
-  geom_segment(data = ~filter(.x, model_reject_20), aes(x = actual_ss, xend = actual_ss, y = 1.1 - 0.03, yend = 1.1 - 0.03 * 2, col = "Computer vision model"), alpha = 0.4) +
-  geom_segment(data = ~filter(.x, !conventional_reject), aes(x = actual_ss, xend = actual_ss, y = -0.1 + 0.03 * 2, yend = -0.1 + 0.03 * 3, col = "Conventional test"), alpha = 0.4) +
-  geom_segment(data = ~filter(.x, conventional_reject), aes(x = actual_ss, xend = actual_ss, y = 1.1, yend = 1.1 - 0.03, col = "Conventional test"), alpha = 0.4) +
-  facet_grid(~type, scales = "free_x") +
-  theme_light() +
-  scale_color_brewer("", palette = "Dark2") +
-  theme(legend.position = "bottom") +
-  xlab("D") +
-  ylab("Reject")
-```
-
-```{r pcp, fig.cap = "Parallel coordinate plots of decisions made by computer vision model, conventional tests and visual tests made by human.", eval = TRUE, fig.height = 4}
-library(ggpcp)
-mutate(experiment, type = ifelse(type == "polynomial", "Non-linearity", "Heteroskedasticity")) %>%
-  mutate(type = factor(type, levels = c("Non-linearity", "Heteroskedasticity"))) %>%
-  filter(!null_lineup, !attention_check) %>%
-  select(type, actual_ss, model_reject_20, reject, conventional_reject) %>%
-  mutate(test = rnorm(n())) %>%
-  mutate(across(c(model_reject_20, conventional_reject, reject), ~ifelse(.x, "Reject", "Not reject"))) %>%
-  mutate(agree = model_reject_20 == reject) %>%
-  pcp_select(c(reject, model_reject_20, conventional_reject)) %>%
-  group_by(type) %>%
-  pcp_scale() %>%
-  ungroup() %>%
-  pcp_arrange(method = "from-right") %>%
-  ggplot() +
-  geom_pcp(aes_pcp(), linewidth = 0.1) +
-  geom_pcp_labels(aes_pcp()) +
-  geom_pcp_boxes(aes_pcp(), boxwidth = 0.1) +
-  facet_wrap(~type, scales = "free_y") +
-  theme_light() +
-  scale_x_discrete(labels = c("Visual\ntest", "Computer\nvision\nmodel", "Conventional\ntest")) +
-  labs(col = "D") +
-  ylab("") +
-  xlab("") +
-  theme(axis.text.y = element_blank(), 
-        axis.ticks.y = element_blank(), 
-        axis.title = element_blank())
-```
-
-```{r delta, fig.pos = "!h", fig.cap = "A weighted detection rate vs adjusted $\\delta$-difference plot.
-The brown line is smoothing curve produced by fitting generalized additive models.", fig.height = 4}
-experiment %>%
-  left_join(lineup_vss %>%
-              group_by(unique_lineup_id) %>%
-              summarise(delta_diff = first(delta_diff), gamma_diff = first(gamma_diff))) %>%
-  ggplot() +
-  geom_vline(xintercept = 0, alpha = 0.4, linetype = 2) +
-  geom_point(aes(delta_diff, prop_detect), alpha = 0.3) +
-  geom_smooth(aes(delta_diff, prop_detect), se = FALSE, col = "#994F00") +
-  ylab("Weighted detection rate") +
-  xlab(expression(delta[adj])) +
-  theme_light()
-```
-
-# Examples {#sec-examples}
-
-In this section, we present the performance of trained computer vision model on three example datasets.
-These include the dataset associated with the residual plot displaying a "left-triangle" shape, as displayed in Figure \ref{fig:false-finding}, along with the Boston housing dataset [@harrison1978hedonic], and the "dino" datasets from the `datasauRus` R package [@datasaurus].
-
-The first example illustrates a scenario where both the computer vision model and human visual inspection successfully avoid rejecting $H_0$ when $H_0$ is true, contrary to conventional tests.
-This underscores the necessity of visually examining the residual plot.
-
-In the second example, we encounter a more pronounced violation of the model, resulting in rejection of $H_0$ by all three tests.
-This highlights the practicality of the computer vision model, particularly for less intricate tasks.
-
-The third example presents a situation where the model deviation is non-typical.
-Here, the computer vision model and human visual inspection reject $H_0$, whereas some commonly used conventional tests do not.
-This emphasizes the benefits of visual inspection and the unique advantage of the computer vision model, which, like humans, makes decisions based on visual discoveries.
-
-## Left-triangle
-
-In Section \ref{sec-model-introduction}, we presented an example residual plot showcased in Figure \ref{fig:false-finding}, illustrating how humans might misinterpret the "left-triangle" shape as indicative of heteroskedasticity.
-Additionally, the Breusch-Pagan test yielded a rejection with a $p$-value of 0.046, despite the residuals originating from a correctly specified model.
-Figure \ref{fig:false-lineup} offers a lineup for this fitted model, showcasing various degrees of "left-triangle" shape across all residual plots.
-This phenomenon is evidently caused by the skewed distribution of the fitted values.
-Notably, if the residual plot were evaluated through a visual test, it would not be rejected since the true residual plot positioned at 10 can not be distinguished from the others.
-
-Figure \ref{fig:false-check} presents the results of the assessment by the computer vision model.
-Notably, the observed visual signal strength is considerably lower than the 95% sample quantile of the null distribution.
-Moreover, the bootstrapped distribution suggests that it is highly improbable for the fitted model to be misspecified as the majority of bootstrapped fitted models will not be rejected.
-Thus, for this particular fitted model, both the visual test and the computer vision model will not reject $H_0$.
-However, the Breusch-Pagan test will reject $H_0$ because it can not effectively utilize information from null plots.
-
-The attention map at Figure \ref{fig:false-check}B suggests that the estimation is highly influenced by the top-right and bottom-right part of the residual plot, as it forms two vertices of the triangular shape.
-A principal component analysis (PCA) is also performed on the output of the global pooling layer of the computer vision model.
-As mentioned in @simonyan2014very, a computer vision model built upon the convolutional blocks can be viewed as a feature extractor.
-For the $32 \times 32$ model, there are 256 features outputted from the global pooling layer, which can be further used for different visual tasks not limited to distance prediction.
-To see if these features can be effectively used for distinguishing null plots and true residual plot, we linearly project them into the first and second principal components space as shown in Figure \ref{fig:false-check}D.
-It can be observed that because the bootstrapped plots are mostly similar to the null plots, the points drawn in different colors are mixed together.
-The true residual plot is also covered by both the cluster of null plots and cluster of bootstrapped plots.
-This accurately reflects our understanding of Figure \ref{fig:false-lineup}.
-
-```{r}
-set.seed(452)
-ori_x <- rand_lognormal()
-dat <- heter_model(b = 0, x = closed_form(~-ori_x))$gen(300)
-
-mod <- lm(y ~ x, data = dat)
-my_vi <- autovi::auto_vi(fitted_model = mod)
-```
-
-```{r}
-if (!file.exists(here::here("cached_data/fig1_check.rds"))) {
-  keras_model <- autovi::get_keras_model("vss_phn_32")
-  my_vi$check(null_draws = 200L, boot_draws = 200L, keras_model = keras_model, extract_feature_from_layer = "global_max_pooling2d")
-  saveRDS(my_vi$check_result, here::here("cached_data/fig1_check.rds"))
-}
-
-```
-
-```{r}
-my_vi$check_result <- readRDS(here::here("cached_data/fig1_check.rds"))
-```
-
-```{r}
-define_gradient <- function() {
-  reticulate::py_run_string(glue::glue(
-"
-import tensorflow as tf
-import PIL
-import numpy as np
-
-def get_smooth_gradient(keras_mod, plot_path, target_size, input_auxiliary = None, noise_level = 0.1, n = 50):
-    im = PIL.Image.open(plot_path)
-    im = im.resize(target_size)
-    input_im = tf.keras.utils.img_to_array(im)
-    input_im = np.reshape(input_im, tuple([1]) + tuple(target_size) + tuple([3]))
-    if input_auxiliary is not None:
-        input_auxiliary = np.reshape(np.array(input_auxiliary), (1, 5))
-
-    if input_auxiliary is not None:
-        input_auxiliary = tf.Variable(input_auxiliary)
-
-    im_grad = None
-    sigma = noise_level * (np.max(input_im) - np.min(input_im))
-
-    for i in range(n):    
-        new_input_im = input_im + np.random.normal(0, sigma, input_im.shape)
-        new_input_im = tf.Variable(new_input_im)
-
-        with tf.GradientTape() as tape:
-            if input_auxiliary is not None:
-                pred = keras_mod([new_input_im, input_auxiliary])
-            else:
-                pred = keras_mod(new_input_im)
-        new_im_grad = tape.gradient(pred, new_input_im)
-        new_im_grad = tf.image.rgb_to_grayscale(new_im_grad)[0].numpy()
-        if im_grad is None:
-            im_grad = new_im_grad
-        else:
-            im_grad = im_grad + new_im_grad
-
-    im_grad = im_grad / n
-
-    input_im = tf.Variable(input_im)
-    if input_auxiliary is not None:
-        with tf.GradientTape() as tape:
-            pred = keras_mod([input_im, input_auxiliary])
-        auxiliary_grad = tape.gradient(pred, input_auxiliary)
-    if input_auxiliary is not None:
-        return (im_grad, auxiliary_grad[0].numpy())
-    else:
-        return im_grad
-
-def get_raw_gradient(keras_mod, plot_path, target_size, input_auxiliary = None):
-    im = PIL.Image.open(plot_path)
-    im = im.resize(target_size)
-    input_im = tf.keras.utils.img_to_array(im)
-    input_im = np.reshape(input_im, tuple([1]) + tuple(target_size) + tuple([3]))
-    if input_auxiliary is not None:
-        input_auxiliary = np.reshape(np.array(input_auxiliary), (1, 5))
-
-    input_im = tf.Variable(input_im)
-    if input_auxiliary is not None:
-        input_auxiliary = tf.Variable(input_auxiliary)
-
-    with tf.GradientTape() as tape:
-        if input_auxiliary is not None:
-            pred = keras_mod([input_im, input_auxiliary])
-        else:
-            pred = keras_mod(input_im)
-    im_grad = tape.gradient(pred, input_im)
-
-    if input_auxiliary is not None:
-        with tf.GradientTape() as tape:
-            pred = keras_mod([input_im, input_auxiliary])
-        auxiliary_grad = tape.gradient(pred, input_auxiliary)
-
-    im_grad = tf.image.rgb_to_grayscale(im_grad)
-    if input_auxiliary is not None:
-        return (im_grad[0].numpy(), auxiliary_grad[0].numpy())
-    else:
-        return im_grad[0].numpy()
-"
-))
-}
-
-```
-
-```{r}
-if (!file.exists(here::here("cached_data/fig1_attention_map.rds"))) {
-  reticulate::py_set_seed(10086, disable_hash_randomization = TRUE)
-  define_gradient()
-  keras_mod <- autovi::get_keras_model("vss_phn_32")
-  plot_path <- my_vi$plot_resid() %>%
-    autovi::save_plot()
-  input_auxiliary <- my_vi$auxiliary()
-
-  reticulate::py$get_smooth_gradient(keras_mod = keras_mod, 
-                                     plot_path = plot_path, 
-                                     target_size = c(32L, 32L), 
-                                     input_auxiliary = input_auxiliary,
-                                     noise_level = 0.5,
-                                     n = 1000L) -> smooth_g
-
-  x_range <- ggplot2::ggplot_build(my_vi$plot_resid())$layout$panel_params[[1]]$x.range
-  y_range <- ggplot2::ggplot_build(my_vi$plot_resid())$layout$panel_params[[1]]$y.range
-
-  smooth_att_map <- smooth_g[[1]] |> 
-    as.data.frame() |>
-    dplyr::mutate(row = rev(1:32)) |>
-    tidyr::pivot_longer(V1:V32, names_to = "column", values_to = "gradient") |>
-    dplyr::mutate(column = as.integer(gsub("V", "", column))) |>
-    dplyr::mutate(column = x_range[1] + column * (x_range[2] - x_range[1])/(32 - 1)) |>
-    dplyr::mutate(row = y_range[1] + row * (y_range[2] - y_range[1])/(32 - 1))
-
-  scale_zero_one <- function(x) (x - min(x))/(max(x) - min(x))
-
-  p <- ggplot() +
-    ggplot2::geom_raster(data = mutate(smooth_att_map, gradient = scale_zero_one(gradient)), ggplot2::aes(column, row, fill = gradient), alpha = 0.7) +
-    scale_fill_gradient(low = "black", high = "white") +
-    ggplot2::theme_void() +
-    theme(legend.position = "none")
-
-  saveRDS(p, here::here("cached_data/fig1_attention_map.rds"))
-
-}
-```
-
-```{r false-check, fig.pos = "!h", fig.cap = 'A summary of the residual plot assessment evaluted on 200 null plots and 200 bootstrapped plots.
-(A) The true residual plot exhibiting a "left-triangle" shape.
-(B) The attention map produced by computing the gradient of the output with respect to the greyscale input.
-(C) The density plot of estimated distance for null plots and bootstrapped plots.
-The green area indicates the distribution of estimated distances for bootstrapped plots, while the yellow area represents the distribution of estimated distances for null plots.
-The fitted model will not be rejected since $\\hat{D} < Q_{null}(0.95)$.
-(D) plot of first two principal components of features extracted from the global pooling layer of the computer vision model.
-', fig.height = 8}
-p1 <- my_vi$plot_resid(theme = theme_light()) + ggtitle(glue::glue("(A) Residual plot (MVI = {format(my_vi$check_result$observed$vss - log(nrow(my_vi$get_fitted_and_resid())) + 10, digits = 3)}, weak)"))
-p2 <- readRDS(here::here("cached_data/fig1_attention_map.rds")) + ggtitle("(B) Attention map")
-p3 <- my_vi$summary_plot() +
-  ggtitle("(C) Density plot of estimated D", subtitle = glue::glue("Conventional p-value = {format(HETER_MODEL$test(dat)$p_value, digits = 3)}")) +
-  theme(legend.position = "bottom", legend.box = "vertical") +
-  labs(linetype = "") +
-  xlab(expression(hat(D))) +
-  scale_linetype(labels = c(expression(Q[null](0.95)), expression(observed~hat(D)))) +
-  scale_fill_manual(values = c("#40B0A6", "#E1BE6A")) +
-  scale_color_manual(values = c("#40B0A6", "#E1BE6A"))
-pca_result <- my_vi$feature_pca()
-pc1_per <- (attributes(pca_result)$sdev^2/256)[1]
-pc2_per <- (attributes(pca_result)$sdev^2/256)[2]
-p4 <- pca_result %>%
-  mutate(set = ifelse(set == "boot", "Boot", set)) %>%
-  mutate(set = ifelse(set == "null", "Null", set)) %>%
-  ggplot() +
-  geom_point(data = ~filter(.x, set != "observed"), aes(PC1, PC2, col = set), size = 2, alpha = 0.3) +
-  geom_point(data = ~filter(.x, set == "observed"), aes(PC1, PC2), size = 4, col = "black", fill = "black") +
-  geom_text(data = ~filter(.x, set == "observed"), aes(PC1, PC2, label = "Observed"), nudge_x = 10) +
-  theme_light() +
-  scale_color_brewer(palette = "Dark2") +
-  xlab(glue::glue("PC1 ({scales::percent(pc1_per)})")) +
-  ylab(glue::glue("PC2 ({scales::percent(pc2_per)})")) +
-  labs(col = "") +
-  theme(legend.position = "bottom") +
-  ggtitle("(D) PCA for extracted features") +
-  scale_color_manual(values = c("#40B0A6", "#E1BE6A"))
-
-patchwork::wrap_plots(p1, p2, p3, p4, nrow = 2, widths = c(1, 1), heights = c(1, 1))
-```
-
-```{r false-lineup, fig.pos = "!h", fig.cap = 'A lineup of residual plots displaying "left-triangle" visual patterns.
-The true residual plot occupies position 10, yet there are no discernible visual patterns that distinguish it from the other plots.', fig.height = 8}
-pos <- 10
-result <- tibble()
-for (i in c(1:20)[-pos]) {
-  result <- bind_rows(result,
-                      my_vi$rotate_resid() %>%
-                        mutate(k = i))
-}
-result <- bind_rows(my_vi$get_fitted_and_resid() %>% 
-                      mutate(k = pos),
-                    result)
-result %>%
-  VI_MODEL$plot_lineup(remove_grid_line = TRUE,
-                       remove_axis = TRUE,
-                       theme = theme_light())
-```
-
-## Boston Housing
-
-The Boston housing dataset, originally published by @harrison1978hedonic, offers insights into housing in the Boston, Massachusetts area.
-For illustration purposes, we utilize a reduced version from Kaggle, comprising 489 rows and 4 columns: average number of rooms per dwelling (RM), percentage of lower status of the population (LSTAT), pupil-teacher ratio by town (PTRATIO), and median value of owner-occupied homes in $1000's (MEDV).
-In our analysis, MEDV will serve as the response variable, while the other columns will function as predictors in a linear regression model.
-Our primary focus is to detect non-linearity, because the relationships between RM and MEDV or LSTAT and MEDV are non-linear.
-
-Figure \ref{fig:boston-check} displays the residual plot and the assessment conducted by the computer vision model.
-A clear non-linearity pattern resembling a "U" shape is shown in the plot A.
-Furthermore, the RESET test yields a very small $p$-value.
-The estimated distance $\hat{D}$ significantly exceeds $Q_{null}(0.95)$, leading to rejection of $H_0$.
-The bootstrapped distribution also suggests that almost all the bootstrapped fitted models will be rejected, indicating that the fitted model is unlikely to be correctly specified.
-The attention map in plot B suggests the center of the image has higher leverage than other areas, and it is the turning point of the "U" shape.
-The PCA provided in plot D shows two distinct clusters of data points, further underling the visual differences between bootstrapped plots and null plots.
-This coincides the findings from Figure \ref{fig:boston-lineup}, where the true plot exhibiting a "U" shape is visually distinctive from null plots.
-If a visual test is conducted by human, $H_0$ will also be rejected.
-
-```{r}
-housing <- read_csv(here::here("data/housing.csv"))
-mod <- lm(MEDV ~ ., data = housing)
-
-my_vi <- autovi::auto_vi(fitted_model = mod)
-```
-
-```{r}
-if (!file.exists(here::here("cached_data/boston_check.rds"))) {
-  keras_model <- autovi::get_keras_model("vss_phn_32")
-  my_vi$check(null_draws = 200L, boot_draws = 200L, keras_model = keras_model, extract_feature_from_layer = "global_max_pooling2d")
-  saveRDS(my_vi$check_result, here::here("cached_data/boston_check.rds"))
-}
-
-```
-
-```{r}
-my_vi$check_result <- readRDS(here::here("cached_data/boston_check.rds"))
-```
-
-```{r}
-if (!file.exists(here::here("cached_data/boston_attention_map.rds"))) {
-  reticulate::py_set_seed(10086, disable_hash_randomization = TRUE)
-  define_gradient()
-  keras_mod <- autovi::get_keras_model("vss_phn_32")
-  plot_path <- my_vi$plot_resid() %>%
-    autovi::save_plot()
-  input_auxiliary <- my_vi$auxiliary()
-
-  reticulate::py$get_smooth_gradient(keras_mod = keras_mod, 
-                                     plot_path = plot_path, 
-                                     target_size = c(32L, 32L), 
-                                     input_auxiliary = input_auxiliary,
-                                     noise_level = 0.5,
-                                     n = 1000L) -> smooth_g
-
-  x_range <- ggplot2::ggplot_build(my_vi$plot_resid())$layout$panel_params[[1]]$x.range
-  y_range <- ggplot2::ggplot_build(my_vi$plot_resid())$layout$panel_params[[1]]$y.range
-
-  smooth_att_map <- smooth_g[[1]] |> 
-    as.data.frame() |>
-    dplyr::mutate(row = rev(1:32)) |>
-    tidyr::pivot_longer(V1:V32, names_to = "column", values_to = "gradient") |>
-    dplyr::mutate(column = as.integer(gsub("V", "", column))) |>
-    dplyr::mutate(column = x_range[1] + column * (x_range[2] - x_range[1])/(32 - 1)) |>
-    dplyr::mutate(row = y_range[1] + row * (y_range[2] - y_range[1])/(32 - 1))
-
-  scale_zero_one <- function(x) (x - min(x))/(max(x) - min(x))
-
-  p <- ggplot() +
-    ggplot2::geom_raster(data = mutate(smooth_att_map, gradient = scale_zero_one(gradient)), ggplot2::aes(column, row, fill = gradient), alpha = 0.7) +
-    scale_fill_gradient(low = "black", high = "white") +
-    ggplot2::theme_void() +
-    theme(legend.position = "none")
-
-  saveRDS(p, here::here("cached_data/boston_attention_map.rds"))
-
-}
-```
-
-```{r boston-check, fig.pos = "!h", fig.cap = 'A summary of the residual plot assessment for the Boston housing fitted model evaluted on 200 null plots and 200 bootstrapped plots.
-(A) The true residual plot exhibiting a "U" shape.
-(B) The attention map produced by computing the gradient of the output with respect to the greyscale input.
-(C) The density plot of estimated distance for null plots and bootstrapped plots.
-The blue area indicates the distribution of estimated distances for bootstrapped plots, while the yellow area represents the distribution of estimated distances for null plots.
-The fitted model will be rejected since $\\hat{D} \\geq Q_{null}(0.95)$.
-(D) plot of first two principal components of features extracted from the global pooling layer of the computer vision model.
-', fig.height = 8}
-p1 <- my_vi$plot_resid(theme = theme_light()) + ggtitle(glue::glue("(A) Residual plot (MVI = {format(my_vi$check_result$observed$vss - log(nrow(my_vi$get_fitted_and_resid())) + 10, digits = 3)}, strong)"))
-p2 <- readRDS(here::here("cached_data/boston_attention_map.rds")) + ggtitle("(B) Attention map")
-p3 <- my_vi$summary_plot() +
-  ggtitle("(C) Density plot of estimated D", subtitle = glue::glue("Conventional p-value = {format(lmtest::resettest(mod)$p.value, digits = 3)}")) +
-  theme(legend.position = "bottom", legend.box = "vertical") +
-  labs(linetype = "") +
-  xlab(expression(hat(D))) +
-  scale_linetype(labels = c(expression(Q[null](0.95)), expression(observed~hat(D)))) +
-  scale_fill_manual(values = c("#40B0A6", "#E1BE6A")) +
-  scale_color_manual(values = c("#40B0A6", "#E1BE6A"))
-pca_result <- my_vi$feature_pca()
-pc1_per <- (attributes(pca_result)$sdev^2/256)[1]
-pc2_per <- (attributes(pca_result)$sdev^2/256)[2]
-p4 <- pca_result %>%
-  mutate(set = ifelse(set == "boot", "Boot", set)) %>%
-  mutate(set = ifelse(set == "null", "Null", set)) %>%
-  ggplot() +
-  geom_point(data = ~filter(.x, set != "observed"), aes(PC1, PC2, col = set), size = 2, alpha = 0.3) +
-  geom_point(data = ~filter(.x, set == "observed"), aes(PC1, PC2), size = 4, col = "black", fill = "black") +
-  geom_text(data = ~filter(.x, set == "observed"), aes(PC1, PC2, label = "Observed"), nudge_x = 7) +
-  theme_light() +
-  scale_color_brewer(palette = "Dark2") +
-  xlab(glue::glue("PC1 ({scales::percent(pc1_per)})")) +
-  ylab(glue::glue("PC2 ({scales::percent(pc2_per)})")) +
-  labs(col = "") +
-  theme(legend.position = "bottom") +
-  ggtitle("(D) PCA for extracted features") +
-  scale_color_manual(values = c("#40B0A6", "#E1BE6A"))
-
-patchwork::wrap_plots(p1, p2, p3, p4, nrow = 2, widths = c(1, 1), heights = c(1, 1))
-```
-
-```{r boston-lineup, fig.pos = "!h", fig.cap = 'A lineup of residual plots for the Boston housing fitted model.
-The true residual plot is at position 7.
-It can be easily identified as the most different plot.', fig.height = 8}
-pos <- 7
-result <- tibble()
-for (i in c(1:20)[-pos]) {
-  result <- bind_rows(result,
-                      my_vi$rotate_resid() %>%
-                        mutate(k = i))
-}
-result <- bind_rows(my_vi$get_fitted_and_resid() %>% 
-                      mutate(k = pos),
-                    result)
-result %>%
-  VI_MODEL$plot_lineup(remove_grid_line = TRUE,
-                       remove_axis = TRUE,
-                       theme = theme_light())
-```
-
-## DatasauRus
-
-The computer vision model possesses the capability to detect not only typical issues like non-linearity, heteroskedasticity, and non-normality but also artifact visual patterns resembling real-world objects, as long as they do not appear in null plots.
-These visual patterns can be challenging to categorize in terms of model violations.
-Therefore, we will employ the RESET test, the Breusch-Pagan test, and the Shapiro-Wilk test [@shapiro1965analysis] for comparison.
-
-The "dino" dataset within the `datasauRus` R package exemplifies this scenario.
-With only two columns, x and y, fitting a regression model to this data yields a residual plot resembling a "dinosaur", as displayed in Figure \ref{fig:dino-check}A.
-Unsurprisingly, this distinct residual plot stands out in a lineup, as shown in Figure \ref{fig:dino-lineup}.
-A visual test conducted by humans would undoubtedly reject $H_0$.
-
-According to the residual plot assessment by the computer vision model, $\hat{D}$ exceeds $Q_{null}(0.95)$, warranting a rejection of $H_0$.
-Additionally, most of the bootstrapped fitted models will be rejected, indicating an misspecified model.
-However, both the RESET test and the Breusch-Pagan test yield $p$-values greater than 0.3, leading to a non-rejection of $H_0$.
-Only the Shapiro-Wilk test rejects the normality assumption with a small $p$-value.
-
-More importantly, the attention map in Figure \ref{fig:dino-check}B clearly exhibits a "dinosaur" shape, strongly suggesting that the distance prediction is based on human-perceptible visual patterns.
-The computer vision model effectively captures the contour or outline of the embedded shape, similar to how humans interpret residual plots.
-Additionally, the PCA in Figure \ref{fig:dino-check}D demonstrates that the cluster of bootstrapped plots is positioned at the corner of the cluster of null plots.
-
-In practice, without accessing the residual plot, it would be challenging to identify the artificial pattern of the residuals.
-Moreover, conducting a normality test for a fitted regression model is not always standard practice among analysts.
-Even when performed, violating the normality assumption is sometimes deemed acceptable, especially considering the application of quasi-maximum likelihood estimation in linear regression.
-This example underscores the importance of evaluating residual plots and highlights how the proposed computer vision model can facilitate this process.
-
-```{r}
-dino <- datasauRus::datasaurus_dozen %>% filter(dataset == "dino")
-mod <- lm(y ~ ., data = select(dino, -dataset))
-
-my_vi <- autovi::auto_vi(fitted_model = mod)
-```
-
-```{r}
-if (!file.exists(here::here("cached_data/dino_check.rds"))) {
-  keras_model <- autovi::get_keras_model("vss_phn_32")
-  my_vi$check(null_draws = 200L, boot_draws = 200L, keras_model = keras_model, extract_feature_from_layer = "global_max_pooling2d")
-  saveRDS(my_vi$check_result, here::here("cached_data/dino_check.rds"))
-}
-```
-
-```{r}
-my_vi$check_result <- readRDS(here::here("cached_data/dino_check.rds"))
-```
-
-```{r}
-if (!file.exists(here::here("cached_data/dino_attention_map.rds"))) {
-  reticulate::py_set_seed(10086, disable_hash_randomization = TRUE)
-  define_gradient()
-  keras_mod <- autovi::get_keras_model("vss_phn_32")
-  plot_path <- my_vi$plot_resid() %>%
-    autovi::save_plot()
-  input_auxiliary <- my_vi$auxiliary()
-
-  reticulate::py$get_smooth_gradient(keras_mod = keras_mod, 
-                                     plot_path = plot_path, 
-                                     target_size = c(32L, 32L), 
-                                     input_auxiliary = input_auxiliary,
-                                     noise_level = 0.5,
-                                     n = 1000L) -> smooth_g
-
-  x_range <- ggplot2::ggplot_build(my_vi$plot_resid())$layout$panel_params[[1]]$x.range
-  y_range <- ggplot2::ggplot_build(my_vi$plot_resid())$layout$panel_params[[1]]$y.range
-
-  smooth_att_map <- smooth_g[[1]] |> 
-    as.data.frame() |>
-    dplyr::mutate(row = rev(1:32)) |>
-    tidyr::pivot_longer(V1:V32, names_to = "column", values_to = "gradient") |>
-    dplyr::mutate(column = as.integer(gsub("V", "", column))) |>
-    dplyr::mutate(column = x_range[1] + column * (x_range[2] - x_range[1])/(32 - 1)) |>
-    dplyr::mutate(row = y_range[1] + row * (y_range[2] - y_range[1])/(32 - 1))
-
-  scale_zero_one <- function(x) (x - min(x))/(max(x) - min(x))
-
-  p <- ggplot() +
-    ggplot2::geom_raster(data = mutate(smooth_att_map, gradient = scale_zero_one(gradient)), ggplot2::aes(column, row, fill = gradient), alpha = 0.7) +
-    scale_fill_gradient(low = "black", high = "white") +
-    ggplot2::theme_void() +
-    theme(legend.position = "none")
-
-  saveRDS(p, here::here("cached_data/dino_attention_map.rds"))
-
-}
-```
-
-```{r dino-check, fig.pos = "!h", fig.cap = 'A summary of the residual plot assessment for the datasauRus fitted model evaluated on 200 null plots and 200 bootstrapped plots.
-(A) The residual plot exhibits a "dinosaur" shape.
-(B) The attention map produced by computing the gradient of the output with respect to the greyscale input.
-(C) The density plot of estimated distance for null plots and bootstrapped plots.
-The blue area indicates the distribution of estimated distances for bootstrapped plots, while the yellow area represents the distribution of estimated distances for null plots.
-The fitted model will be rejected since $\\hat{D} \\geq Q_{null}(0.95)$.
-(D) plot of first two principal components of features extracted from the global pooling layer of the computer vision model.', fig.height = 8}
-
-p1 <- my_vi$plot_resid(theme = theme_light()) + ggtitle(glue::glue("(A) Residual plot (MVI = {format(my_vi$check_result$observed$vss - log(nrow(my_vi$get_fitted_and_resid())) + 10, digits = 3)}, strong)"))
-p2 <- readRDS(here::here("cached_data/dino_attention_map.rds")) + ggtitle("(B) Attention map")
-p3 <- my_vi$summary_plot() +
-  ggtitle("(C) Density plot of estimated D", subtitle = glue::glue("RESET test p-value = {format(lmtest::resettest(mod)$p.value, digits = 3)}\n Breusch-Pagan test p-value = {format(lmtest::bptest(mod)$p.value, digits = 3)} \n Shapiro-Wilk test p-value = {format(shapiro.test(mod$residuals)$p.value, digits = 3)}")) +
-  theme(legend.position = "bottom", legend.box = "vertical") +
-  labs(linetype = "") +
-  xlab(expression(hat(D))) +
-  scale_linetype(labels = c(expression(Q[null](0.95)), expression(observed~hat(D)))) +
-  scale_fill_manual(values = c("#40B0A6", "#E1BE6A")) +
-  scale_color_manual(values = c("#40B0A6", "#E1BE6A"))
-pca_result <- my_vi$feature_pca()
-pc1_per <- (attributes(pca_result)$sdev^2/256)[1]
-pc2_per <- (attributes(pca_result)$sdev^2/256)[2]
-p4 <- pca_result %>%
-  mutate(set = ifelse(set == "boot", "Boot", set)) %>%
-  mutate(set = ifelse(set == "null", "Null", set)) %>%
-  ggplot() +
-  geom_point(data = ~filter(.x, set != "observed"), aes(PC1, PC2, col = set), size = 2, alpha = 0.3) +
-  geom_point(data = ~filter(.x, set == "observed"), aes(PC1, PC2), size = 4, col = "black", fill = "black") +
-  geom_text(data = ~filter(.x, set == "observed"), aes(PC1, PC2, label = "Observed"), nudge_x = -10) +
-  theme_light() +
-  scale_color_brewer(palette = "Dark2") +
-  xlab(glue::glue("PC1 ({scales::percent(pc1_per)})")) +
-  ylab(glue::glue("PC2 ({scales::percent(pc2_per)})")) +
-  labs(col = "", shape = "") +
-  theme(legend.position = "bottom", legend.box="vertical") +
-  ggtitle("(D) PCA for extracted features") +
-  scale_color_manual(values = c("#40B0A6", "#E1BE6A"))
-
-patchwork::wrap_plots(p1, p2, p3, p4, nrow = 2, widths = c(1, 1), heights = c(1, 1))
-```
-
-```{r dino-lineup, fig.pos = "!h", fig.cap = 'A lineup of residual plots for the fitted model on the "dinosaur" dataset.
-The true residual plot is at position 17.
-It can be easily identified as the most different plot as the visual pattern is extremely artificial.', fig.height = 8}
-pos <- 17
-result <- tibble()
-for (i in c(1:20)[-pos]) {
-  result <- bind_rows(result,
-                      my_vi$rotate_resid() %>%
-                        mutate(k = i))
-}
-result <- bind_rows(my_vi$get_fitted_and_resid() %>% 
-                      mutate(k = pos),
-                    result)
-result %>%
-  VI_MODEL$plot_lineup(remove_grid_line = TRUE,
-                       remove_axis = TRUE,
-                       theme = theme_light())
-```
-
-# Limitations and Future Work
-
-Despite the computer vision model performing well with general cases under the synthetic data generation scheme and the three examples used in this paper, this study has several limitations that could guide future work.
-
-The proposed distance measure assumes that the true model is a classical normal linear regression model, which can be restrictive.
-Although this paper does not address the relaxation of this assumption, there are potential methods to evaluate other types of regression models.
-The most comprehensive approach would be to define a distance measure for each different class of regression model and then train the computer vision model following the methodology described in this paper.
-To accelerate training, one could use the convolutional blocks of our trained model as a feature extractor and perform transfer learning on top of it, as these blocks effectively capture shapes in residual plots.
-Another approach would be to transform the residuals so they are roughly normally distributed and have constant variance.
-If only raw residuals are used, the distance-based statistical testing compares the difference in distance to a classical normal linear regression model for the true plot and null plots.
-This comparison is meaningful only if the difference can be identified by the distance measure proposed in this paper.
-
-There are other types of residual plots commonly used in diagnostics, such as residuals vs.
-predictor and quantile-quantile plots.
-In this study, we focused on the most commonly used residual plot as a starting point for exploring the new field of automated visual inference.
-Similarly, we did not explore other, more sophisticated computer vision model architectures and specifications for the same reason.
-While the performance of the computer vision model is acceptable, there is still room for improvement to achieve behavior more closely resembling that of humans interpreting residual plots.
-This may require external survey data or human subject experiment data to understand the fundamental differences between our implementation and human evaluation.
-
-# Conclusions
-
-In this paper, we have introduced a distance measure based on Kullback-Leibler divergence to quantify the disparity between the residual distribution of a fitted classical normal linear regression model and the reference residual distribution assumed under correct model specification.
-This distance measure effectively captures the magnitude of model violations in misspecified models.
-We propose a computer vision model to estimate this distance, utilizing the residual plot of the fitted model as input.
-The resulting estimated distance serves as the foundation for constructing a single Model Violations Index (MVI), facilitating the quantification of various model violations.
-
-Moreover, the estimated distance enables the development of a formal statistical testing procedure by evaluating a large number of null plots generated from the fitted model.
-Additionally, employing bootstrapping techniques and refitting the regression model allows us to ascertain how frequently the fitted model is considered misspecified if data were repeatedly obtained from the same data generating process.
-
-The trained computer vision model demonstrates strong performance on both the training and test sets, although it exhibits slightly lower performance on residual plots with non-linearity visual patterns compared to other types of violations.
-The statistical tests relying on the estimated distance predicted by the computer vision model exhibit lower sensitivity compared to conventional tests but higher sensitivity compared to visual tests conducted by humans.
-While the estimated distance generally mirrors the strength of the visual signal perceived by humans, there remains scope for further improvement in its performance.
-
-Several examples are provided to showcase the effectiveness of the proposed method across different scenarios, emphasizing the similarity between visual tests and distance-based tests.
-Overall, both visual tests and distance-based tests can be viewed as ensemble of tests, aiming to assess
-any violations of model assumptions collectively.
-In contrast, individual residual diagnostic tests such as the RESET test and the Breusch-Pagan test only evaluate specific violations of model assumptions.
-In practice, selecting an appropriate set of statistical tests for regression diagnostics can be challenging, particularly given the necessity of adjusting the significance level for each test.
-
-Our method holds significant value as it helps alleviate a portion of analysts' workload associated with assessing residual plots.
-While we recommend analysts to continue reading residual plots whenever feasible, as they offer invaluable insights, our approach serves as a valuable tool for automating the diagnostic process or for supplementary purposes when needed.
-
-# Acknowledgement {-}
-
-These `R` packages were used for the work: `tidyverse` [@tidyverse], `lmtest` [@lmtest], `mpoly` [@mpoly], `ggmosaic` [@ggmosaic], `kableExtra` [@kableextra], `patchwork` [@patchwork], `rcartocolor` [@rcartocolor], `glue` [@glue], `ggpcp` [@ggpcp], `here` [@here], `magick` [@magick], `yardstick` [@yardstick] and `reticulate` [@reticulate].
-
-The article was created with R packages `rticles` [@rticles], `knitr` [@knitr] and `rmarkdown` [@rmarkdown].
-The project's GitHub repository (https://github.com/TengMCing/auto_residual_reading_paper) contains all materials required to reproduce this article.

### 2025-06-12 appendix.tex
Init the appendix

--- 2025-06-05 paper.tex
@@
 
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{wrapfig}
-\usepackage{float}
-\usepackage{colortbl}
-\usepackage{pdflscape}
-\usepackage{tabu}
-\usepackage{threeparttable}
-\usepackage{threeparttablex}
-\usepackage[normalem]{ulem}
-\usepackage{makecell}
-\usepackage{xcolor}
-
 \begin{document}
@@
 
-\title{Automated Assessment of Residual Plots with Computer Vision
-Models}
+\title{Appendix to ``Automated Assessment of Residual Plots with
+Computer Vision Models''}
 
@@
 
-\section{Introduction}\label{sec-model-introduction}
-
-Plotting residuals is commonly regarded as a standard practice in linear
-regression diagnostics \citep{belsley1980regression, cook1982residuals}.
-This visual assessment plays a crucial role in identifying whether model
-assumptions, such as linearity, homoscedasticity, and normality, are
-reasonable. It also helps in understanding the goodness of fit and
-various unexpected characteristics of the model.
-
-Generating a residual plot in most statistical software is often as
-straightforward as executing a line of code or clicking a button.
-However, accurately interpreting a residual plot can be challenging. A
-residual plot can exhibit various visual features, but it is crucial to
-recognize that some may arise from the characteristics of predictors and
-the natural stochastic variation of the observational unit, rather than
-indicating a violation of model assumptions \citep{li2024plot}. Consider
-Figure \ref{fig:false-finding} as an example, the residual plot displays
-a triangular left-pointing shape. The distinct difference in the spread
-of the residuals across the fitted values may result in the analyst
-suggesting that there may be heteroskedasticity, however, it is
-important to avoid over-interpreting this visual pattern. In this case,
-the fitted regression model is correctly specified, and the triangular
-shape is actually a result of the skewed distribution of the predictors,
-rather than indicating a flaw in the model.
-
-The concept of visual inference, as proposed by
-\citet{buja2009statistical}, provides an inferential framework to assess
-whether residual plots indeed contain visual patterns inconsistent with
-the model assumptions. The fundamental idea involves testing whether the
-true residual plot visually differs significantly from null plots, where
-null plots are plotted with residuals generated from the residual
-rotation distribution \citep{langsrud2005rotation}, which is a
-distribution consistent with the null hypothesis \(H_0\) that the linear
-regression model is correctly specified. Typically, the visual test is
-accomplished through the lineup protocol, where the true residual plot
-is embedded within a lineup alongside several null plots. If the true
-residual plot can be distinguished from the lineup, it provides evidence
-for rejecting \(H_0\).
-
-The practice of delivering a residual plot as a lineup is generally
-regarded as a valuable approach. Beyond its application in residual
-diagnostics, the lineup protocol has been integrated into the analysis
-of diverse subjects. For instance, Loy and Hofmann
-\citetext{\citeyear{loy2013diagnostic}; \citeyear{loy2014hlmdiag}; \citeyear{loy2015you}}
-illustrated its applicability in diagnosing hierarchical linear models.
-Additionally, \citet{widen2016graphical} and \citet{fieberg2024using}
-demonstrated its utility in geographical and ecology research
-respectively, while \citet{krishnan2021hierarchical} explored its
-effectiveness in forensic examinations.
-
-A practical limitation of the lineup protocol lies in its reliance on
-human judgements \citep[see][ about the practical
-limitations]{li2024plot}. Unlike conventional statistical tests that can
-be performed computationally in statistical software, the lineup
-protocol requires human evaluation of images. This characteristic makes
-it less suitable for large-scale applications, given the associated high
-labour costs and time requirements. There is a substantial need to
-develop an approach to substitute these human judgement with an
-automated reading of data plots using machines.
-
-The utilization of computers to interpret data plots has a rich history,
-with early efforts such as ``Scagnostics'' by \citet{tukey1985computer},
-a set of numerical statistics that summarize features of scatter plots.
-\citet{wilkinson2005graph} expanded on this work, introducing
-scagnostics based on computable measures applied to planar proximity
-graphs. These measures, including, but not limited to, ``Outlying'',
-``Skinny'', ``Stringy'', ``Straight'', ``Monotonic'', ``Skewed'',
-``Clumpy'', and ``Striated'', aimed to characterize outliers, shape,
-density, trend, coherence and other characteristics of the data. While
-this approach has been inspiring, there is a recognition
-\citep{buja2009statistical} that it may not capture all the necessary
-visual features that differentiate true residual plots from null plots.
-A more promising alternative entails enabling machines to learn the
-function for extracting visual features from residual plots.
-Essentially, this means empowering computers to discern the crucial
-visual features for residual diagnostics and determining the method to
-extract them.
-
-Modern computer vision models are well-suited for addressing this
-challenge. They rely on deep neural networks with convolutional layers
-\citep{fukushima1982neocognitron}. These layers use small, sliding
-windows to scan the image, performing a dot product to extract local
-features and patterns. Numerous studies have demonstrated the efficacy
-of convolutional layers in addressing various vision tasks, including
-image recognition \citep{rawat2017deep}. Despite the widespread use of
-computer vision models in fields like computer-aided diagnosis
-\citep{lee2015image}, pedestrian detection \citep{brunetti2018computer},
-and facial recognition \citep{emami2012facial}, their application in
-reading data plots remains limited. While some studies have explored the
-use of computer vision models for tasks such as reading recurrence plots
-for time series regression \citep{ojeda2020multivariate}, time series
-classification
-\citep{chu2019automatic, hailesilassie2019financial, hatami2018classification, zhang2020encoding},
-anomaly detection \citep{chen2020convolutional}, and pairwise causality
-analysis \citep{singh2017deep}, the application of reading residual
-plots with computer vision models is a new field of study.
-
-In this paper, we develop computer vision models and integrate them into
-the residual plots diagnostics workflow, addressing the need for an
-automated visual inference. The paper is structured as follows. Section
-\ref{sec-model-specifications} discusses various specifications of the
-computer vision models. Section
-\ref{sec-model-distance-between-residual-plots} defines the distance
-measure used to detect model violations, while Section
-\ref{sec-model-distance-estimation} explains how the computer vision
-models estimate this distance measure. Section
-\ref{sec-model-statistical-testing} covers the statistical tests based
-on the estimated distance, and Section \ref{sec-model-violations-index}
-introduces a Model Violations Index, which offers a quicker and more
-convenient assessment. Sections \ref{sec-model-data-generation},
-\ref{sec-model-architecture}, and \ref{sec-model-training} detail the
-data preparation, model architecture, and training process,
-respectively. The results are presented in Section
-\ref{sec-model-results}. Example dataset applications are discussed in
-Section \ref{sec-examples}. Finally, we conclude with a discussion of
-our findings and propose ideas for future research directions.
-
-\begin{figure}[!h]
-
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/false-finding-1}
-
-}
-
-\caption{An example residual vs fitted values plot (red line indicates 0 corresponds to the x-intercept, i.e.
-$y=0$).
-The vertical spread of the data points varies with the fitted values.
-This often indicates the existence of heteroskedasticity, however, here the result is due to skewed distribution of the predictors rather than heteroskedasticity.
-The Breusch-Pagan test rejects this residual plot at 95\% significance level ($p\text{-value} = 0.046$).}\label{fig:false-finding}
-\end{figure}
-
-\section{Model Specifications}\label{sec-model-specifications}
-
-There are various specifications of the computer vision model that can
-be used to assess residual plots. We discuss these specifications below
-focusing on two key components of the model formula: the input and the
-output format.
-
-\subsection{Input Formats}\label{input-formats}
-
-Deep learning models are in general very sensitive to the input data.
-The quality and relevance of the input data greatly influence the
-model's capacity to generate insightful and meaningful results. There
-are several designs of the input format that can be considered.
-
-A straightforward architecture of the input layer involves feeding a
-vector of residuals along with a vector of fitted values, essentially
-providing all the necessary information for creating a residuals vs
-fitted values plot. However, a drawback of this method is the variable
-input size, which changes based on the number of observations. For
-modern computer vision models implemented in mainstream software like
-TensorFlow \citep{abadi2016tensorflow}, the input shape is typically
-fixed. One strategy is to pad the input vectors with leading or trailing
-zeros when the input tensor expects longer vectors, but it may fail if
-the input vector surpasses the designed length. Alternatively, sampling
-a fixed number of residual-fitted value pairs can ensure consistent
-input size. By carefully designing the sampling process, it is possible
-to maintain the input's representativeness while conforming to a fixed
-format.
-
-Another architecture of the input layer involves using an image as
-input. The primary advantage of this design, as opposed to the vector
-format, is the availability of the existing and sophisticated image
-processing architectures developed over the years, such as the VGG16
-architecture proposed in \citet{simonyan2014very}. These architectures
-can effectively capture and summarize spatial information from nearby
-pixels, which is less straightforward with vector input. Encoding data
-points as pixels introduces certain trade-offs. While some detail may be
-lost during the discretization process, this approach aligns with the
-sampling-based strategy by focusing on summarizing the overall data
-distribution. Unlike vectors, which struggle to represent spatial
-relationships, image-based inputs naturally encode the joint
-distribution of residuals and fitted values in a visually interpretable
-format. This richer representation gives the model a more comprehensive
-and intuitive understanding of the residual plot's underlying structure.
-Key considerations for using image-based input include image resolution
-and the aesthetics of the residual plot. In general, a higher resolution
-provides more information to the model but comes at the cost of of
-increased complexity and greater difficulty in training. As for the
-aesthetics of the residual plot, a practical solution is to consistently
-present residual plots in the same style to the model. This implies that
-the model can not accept arbitrary images as input but requires the use
-of the same pre-processing pipeline to convert residuals and fitted
-values into a standardized-style residual plot.
-
-Providing multiple residual plots to the model, such as a pair of plots,
-a triplet or a lineup is also a possible option.
-\citet{chopra2005learning} have shown that computer vision models
-designed for image comparison can assess whether a pair of images are
-similar or dissimilar. Applied to our specific problem, we can define
-null plots of a fitted regression model to be similar to each other,
-while considering true residual plots to be distinct from null plots of
-any fitted regression model. A triplet constitutes a set of three
-images, denoted as \(image_1\), \(image_2\) and \(image_3\). It is often
-used to predict whether \(image_2\) or \(image_3\) is more similar to
-\(image_1\), proving particularly useful for establishing rankings
-between samples. For this setup, we can apply the same criteria to
-define similarity between images. However, it is important to note that
-these two approaches usually require additional considerations regarding
-the loss function and, at times, non-standard training processes due to
-shared weights between different convolutional blocks.
-
-Presenting a lineup to a model aligns closely with the lineup protocol.
-However, as the number of residual plots in a lineup increases, the
-resolution of the input image grows rapidly, posing challenges in
-training the model. We experimented with this approach in a pilot study,
-but the performance of the trained model was sub-optimal.
-
-Taking into account the implementation cost and the need for model
-interpretability, we used the single residual plot input format in this
-paper.
-
-\subsection{Output Formats}\label{output-formats}
-
-Given that the input is a single residual plot represented as a
-fixed-resolution image, the output of the computer vision model can be
-designed for various tasks: binary classification, categorical
-classification with multiple classes, or numeric regression.
-
-The binary outcome can represent whether the input image is consistent
-with a null plot as determined by either (1) the data generating process
-or (2) the result of a visual test based on human judgement. Training a
-model following the latter option requires data from prior human subject
-experiments, presenting difficulties in controlling the quality of data
-due to variations in experimental settings across different studies.
-Additionally, some visual inference experiments are unrelated to linear
-regression models or residual plot diagnostics, resulting in a limited
-amount of available training data.
-
-For a multi-class outcome, the model could classify images into
-categories such as ``contains outliers'' vs.~``does not contain
-outliers'' or ``is non-linear'' vs.~``not non-linear,'' among other
-diagnostic features. While this approach is appealing in theory, it
-essentially reverts to the traditional testing framework, where tests
-focus on detecting specific types of model violations in isolation.
-
-One significant advantage of the lineup protocol for residual
-diagnostics is its ability to detect multiple model violations
-simultaneously, provided these violations produce visually
-distinguishable differences in the plots. In contrast, training a
-classification model would require explicitly defining and mapping all
-potential violations of interest, or at least the most common ones,
-which can be impractical. If the goal is to identify specific cases,
-conventional statistical tests grounded in regression model assumptions
-are often more effective. Computer vision models may be less suited to
-such tasks.
-
-Alternatively, the output could be a meaningful and interpretable
-numerical measure useful for assessing residual plots, such as the
-strength of suspicious visual patterns reflecting the extent of model
-violations, or the difficulty index for identifying whether a residual
-plot has no issues. However, these numeric measures are often informally
-used in daily communication but are not typically formalized or
-rigorously defined. For the purpose of training a model, this numeric
-measure has to be quantifiable.
-
-In this study, we chose to define and use a distance between a true
-residual plot and a theoretically ``good'' residual plot. This is
-further explained in \citet{sec-model-distance-between-residual-plots}.
-\citet{vo2016localizing} have also demonstrated that defining a proper
-distance between images can enhance the matching accuracy in image
-search compared to a binary outcome model.
-
-\subsection{Auxiliary Information with
-Scagnostics}\label{auxiliary-information-with-scagnostics}
-
-In Section \ref{sec-model-introduction}, we mention that scagnostics
-consist of a set of manually designed visual feature extraction
-functions. While our computer vision model will learn its own feature
-extraction function during training, leveraging additional information
-from scagnostics can enhance the model's predictive accuracy.
-
-For each residual plot used as an input image, we calculated four
-scagnostics --- ``Monotonic'', ``Sparse'', ``Splines'', and ``Striped''
--- using the \texttt{cassowaryr} R package \citep{mason2022cassowaryr}.
-These computed measures, along with the number of observations from the
-fitted model, were provided as the second input for the computer vision
-model. We selected these scagnostics due to their reliability and
-efficiency, as other scagnostics occasionally caused R process crashes
-(approximately 5\% of the time) during training data preparation, due to
-a bug in the \texttt{interp} R package \citep{Albrecht2023interp}.
-Although the package maintainer later fixed this bug at our request, the
-fix came too late to retrain the model, and additionally, their high
-computational costs make them unsuitable for rapid inference, which was
-a critical factor in our choice.
-
-\section{Distance from a Theoretically ``Good'' Residual
-Plot}\label{sec-model-distance-between-residual-plots}
-
-To develop a computer vision model for assessing residual plots within
-the visual inference framework, it is important to precisely define a
-numerical measure of ``difference'' or ``distance'' between plots. This
-distance can take the form of a basic statistical operation on pixels,
-such as the sum of square differences, however, a pixel-to-pixel
-comparison makes little sense in comparing residual plots where the main
-interest would be structural patterns. Alternatively, it could involve
-established image similarity metrics like the Structural Similarity
-Index Measure \citep{wang2004image} which compares images by integrating
-three perception features of an image: contrast, luminance, and
-structure (related to average, standard deviation and correlation of
-pixel values over a window, respectively). These image similarity
-metrics are tailored for image comparison in vastly different tasks to
-evaluating data plots, where only essential plot elements require
-assessment \citep{chowdhury2018measuring}. We can alternatively define a
-notion of distance by integrating key plot elements (instead of key
-perception features like luminance, contrast, and structure), such as
-those captured by scagnostics mentioned in Section
-\ref{sec-model-introduction}, but the functional form still needs to be
-carefully refined to accurately reflect the extent of the violations.
-
-In this section, we introduce a distance measure between a true residual
-plot and a theoretically `good' residual plot. This measure quantifies
-the divergence between the residual distribution of a given fitted
-regression model and that of a correctly specified model. The
-computation assumes knowledge of the data generating processes for
-predictors and response variables. Since these processes are often
-unknown in practice, we will discuss a method to estimate this distance
-using a computer vision model in Section
-\ref{sec-model-distance-estimation}.
-
-\subsection{Residual Distribution}\label{residual-distribution}
-
-For a classical normal linear regression model,
-\(\boldsymbol{y} = \boldsymbol{X}\boldsymbol{\beta} + \boldsymbol{e}\),
-the residual \(\hat{\boldsymbol{e}}\) are derived as the difference of
-the fitted values and observed values \(\boldsymbol{y}\). Suppose the
-data generating process is known and the regression model is correctly
-specified, by the Frisch-Waugh-Lowell theorem \citep{frisch1933partial},
-residuals \(\hat{\boldsymbol{e}}\) can also be treated as random
-variables and written as a linear transformation of the error
-\(\boldsymbol{e}\) formulated as
-\(\hat{\boldsymbol{e}} = \boldsymbol{R}\boldsymbol{e}\), where
-\(\boldsymbol{R}=\boldsymbol{I}_n -\boldsymbol{X}(\boldsymbol{X}^\top\boldsymbol{X})^{-1}\boldsymbol{X}^\top\)
-is the residual operator, \(\boldsymbol{I}_n\) is a \(n\) by \(n\)
-identity matrix, and \(n\) is the number of observations.
-
-One of the assumptions of the classical normal linear regression model
-is that the error \(\boldsymbol{e}\) follows a multivariate normal
-distribution with zero mean and constant variance, i.e.,
-\(\boldsymbol{e} \sim N(\boldsymbol{0}_n,\sigma^2\boldsymbol{I}_n)\). It
-follows that the distribution of residuals \(\hat{\boldsymbol{e}}\) can
-be characterized by a certain probability distribution, denoted as
-\(Q\), which is transformed from the multivariate normal distribution.
-This reference distribution \(Q\) summarizes what ``good'' residuals
-should follow given the design matrix \(\boldsymbol{X}\) is known and
-fixed.
-
-Suppose the design matrix \(\boldsymbol{X}\) has linearly independent
-columns, the trace of the hat matrix
-\(\boldsymbol{H} = \boldsymbol{X}(\boldsymbol{X}^\top\boldsymbol{X})^{-1}\boldsymbol{X}^\top\)
-will equal to the number of columns in \(\boldsymbol{X}\) denoted as
-\(k\). As a result, the rank of \(\boldsymbol{R}\) is \(n - k\), and
-\(Q\) is a degenerate multivariate distribution. To capture the
-characteristics of \(Q\), such as moments, we can simulate a large
-numbers of \(\boldsymbol{\varepsilon}\) and transform it to
-\(\boldsymbol{e}\) to get the empirical estimates. For simplicity, in
-this study, we replaced the variance-covariance matrix of residuals
-\(\text{cov}(\boldsymbol{e}, \boldsymbol{e}) = \boldsymbol{R}\sigma^2\boldsymbol{R}^\top = \boldsymbol{R}\sigma^2\)
-with a full-rank diagonal matrix
-\(\text{diag}(\boldsymbol{R}\sigma^2)\), where \(\text{diag}(.)\) sets
-the non-diagonal entries of a matrix to zeros. The resulting
-distribution for \(Q\) is
-\(N(\boldsymbol{0}_n, \text{diag}(\boldsymbol{R}\sigma^2))\).
-
-Distribution \(Q\) is derived from the correctly specified model.
-However, if the model is misspecified, then the actual distribution of
-residuals denoted as \(P\), will be different from \(Q\). For example,
-if the data generating process contains variables correlated with any
-column of \(\boldsymbol{X}\) but missing from \(\boldsymbol{X}\),
-causing an omitted variable problem, \(P\) will be different from \(Q\)
-because the residual operator obtained from the fitted regression model
-will not be the same as \(\boldsymbol{R}\). Besides, if the
-\(\boldsymbol{\varepsilon}\) follows a non-normal distribution such as a
-multivariate lognormal distribution, \(P\) will usually be skewed and
-has a long tail.
-
-\subsection{\texorpdfstring{Distance of \(P\) from
-\(Q\)}{Distance of P from Q}}\label{distance-of-p-from-q}
-
-Defining a proper distance between distributions is usually easier than
-defining a proper distance between data plots. Given the true residual
-distribution \(Q\) and the reference residual distribution \(P\), we
-used a distance measure based on Kullback-Leibler divergence
-\citep{kullback1951information} to quantify the difference between two
-distributions as
-
-\begin{equation} \label{eq:kl-0}
-D = \log\left(1 + D_{KL}\right),
-\end{equation}
-
-where \(D_{KL}\) is defined as
-
-\begin{equation} \label{eq:kl-1}
-D_{KL} = \int_{\mathbb{R}^{n}}\log\frac{p(\boldsymbol{e})}{q(\boldsymbol{e})}p(\boldsymbol{e})d\boldsymbol{e},
-\end{equation}
-
-\noindent and \(p(.)\) and \(q(.)\) are the probability density
-functions for distribution \(P\) and distribution \(Q\), respectively.
-
-This distance measure was first proposed in \citet{li2024plot}. It was
-mainly designed for measuring the effect size of non-linearity and
-heteroskedasticity in a residual plot. \citet{li2024plot} have derived
-that, for a classical normal linear regression model that omits
-necessary higher-order predictors \(\boldsymbol{Z}\) and the
-corresponding parameter \(\boldsymbol{\beta}_z\), and incorrectly
-assumes
-\(\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n,\sigma^2\boldsymbol{I}_n)\)
-while in fact
-\(\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n, \boldsymbol{V})\)
-where \(\boldsymbol{V}\) is an arbitrary symmetric positive
-semi-definite matrix, \(Q\) can be represented as
-\(N(\boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z, \text{diag}(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}))\).
-Note that the variance-covariance matrix is replaced with the diagonal
-matrix to ensure it is a full-rank matrix.
-
-Since both \(P\) and \(Q\) are adjusted to be multivariate normal
-distributions, Equation \ref{eq:kl-1} can be further expanded to
-
-\begin{equation} \label{eq:kl-2}
-D_{KL} = \frac{1}{2}\left(\log\frac{|\boldsymbol{W}|}{|\text{diag}(\boldsymbol{R}\sigma^2)|} - n + \text{tr}(\boldsymbol{W}^{-1}\text{diag}(\boldsymbol{R}\sigma^2)) + \boldsymbol{\mu}_z^\top\boldsymbol{W}^{-1}\boldsymbol{\mu}_z\right),
-\end{equation}
-
-\noindent where
-\(\boldsymbol{\mu}_z = \boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z\),
-and
-\(\boldsymbol{W} = \text{diag}(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R})\).
-The assumed error variance \(\sigma^2\) is set to be
-\(\text{tr}(\boldsymbol{V})/n\), which is the expectation of the
-estimated variance.
-
-\subsection{\texorpdfstring{Non-normal
-\(P\)}{Non-normal P}}\label{non-normal-p}
-
-For non-normal error \(\boldsymbol{\varepsilon}\), the true residual
-distribution \(P\) is unlikely to be a multivariate normal distribution.
-Thus, Equation \ref{eq:kl-2} given in \citet{li2024plot} will not be
-applicable to models violating the normality assumption.
-
-To evaluate the Kullback-Leibler divergence of non-normal \(P\) from
-\(Q\), the fallback is to solve Equation \ref{eq:kl-1} numerically.
-However, since \(\boldsymbol{e}\) is a linear transformation of
-non-normal random variables, it is very common that the general form of
-\(P\) is unknown, meaning that we can not easily compute
-\(p(\boldsymbol{e})\) using a well-known probability density function.
-Additionally, even if \(p(\boldsymbol{e})\) can be calculated for any
-\(\boldsymbol{e} \in \mathbb{R}^n\), it will be very difficult to do
-numerical integration over the \(n\)-dimensional space, because \(n\)
-could be potentially very large.
-
-In order to approximate \(D_{KL}\) in a practically computable manner,
-the elements of \(\boldsymbol{e}\) are assumed to be independent of each
-other. This assumption solves both of the issues mentioned above. First,
-we no longer need to integrate over \(n\) random variables. The result
-of Equation \ref{eq:kl-1} is now the sum of the Kullback-Leibler
-divergence evaluated for each individual residual due to the assumption
-of independence between observations. Second, it is not required to know
-the joint probability density \(p(\boldsymbol{e})\) any more. Instead,
-the evaluation of Kullback-Leibler divergence for an individual residual
-relies on the knowledge of the marginal density \(p_i(e_i)\), where
-\(e_i\) is the \(i\)-th residual for \(i = 1, ..., n\). This is much
-easier to approximate through simulation. It is also worth mentioning
-that this independence assumption generally will not hold if
-\(\text{cov}(e_i, e_j) \neq 0\) for any \(1 \leq i < j \leq n\), but its
-existence is essential for reducing the computational cost.
-
-Given \(\boldsymbol{X}\) and \(\boldsymbol{\beta}\), the algorithm for
-approximating Equation \ref{eq:kl-1} starts from simulating \(m\) sets
-of observed values \(\boldsymbol{y}\) according to the data generating
-process. The observed values are stored in a matrix \(\boldsymbol{A}\)
-with \(n\) rows and \(m\) columns, where each column of
-\(\boldsymbol{A}\) is a set of observed values. Then, we can get \(m\)
-sets of realized values of \(\boldsymbol{e}\) stored in the matrix
-\(\boldsymbol{B}\) by applying the residual operator
-\(\boldsymbol{B} = \boldsymbol{R}\boldsymbol{A}\). Furthermore, kernel
-density estimation (KDE) with Gaussian kernel and optimal bandwidth
-selected by the Silverman's rule of thumb \citep{silverman2018density}
-is applied on each row of \(\boldsymbol{B}\) to estimate \(p_i(e_i)\)
-for \(i = 1, ..., n\). The KDE computation can be done by the
-\texttt{density} function in R.
-
-Since the Kullback-Leibler divergence can be viewed as the expectation
-of the log-likelihood ratio between distribution \(P\) and distribution
-\(Q\) evaluated on distribution \(P\), we can reuse the simulated
-residuals in matrix \(\boldsymbol{B}\) to estimate the expectation by
-the sample mean. With the independence assumption, for non-normal \(P\),
-\(D_{KL}\) can be approximated by
-
-\begin{align*} \label{eq:kl-3}
-D_{KL} &\approx \sum_{i = 1}^{n} \hat{D}_{KL}^{(i)}, \\
-\hat{D}_{KL}^{(i)} &= \frac{1}{m}\sum_{j = 1}^{m} \log\frac{\hat{p}_i(B_{ij})}{q(B_{ij})},
-\end{align*}
-
-\noindent where \(\hat{D}_{KL}^{(i)}\) is the estimator of the
-Kullback-Leibler divergence for an individual residual \(e_i\),
-\(\boldsymbol{B}_{ij}\) is the \(i\)-th row and \(j\)-th column entry of
-the matrix \(\boldsymbol{B}\), \(\hat{p}_i(.)\) is the kernel density
-estimator of \(p_i(.)\), \(q(.)\) is the normal density function with
-mean zero and an assumed variance estimated as
-\(\hat{\sigma}^2 = \sum_{b \in vec(\boldsymbol{B})}(b - \sum_{b \in vec(\boldsymbol{B})} b/nm)^2/(nm - 1)\),
-and \(vec(.)\) is the vectorization operator which turns a
-\(n \times m\) matrix into a \(nm \times 1\) column vector by stacking
-the columns of the matrix on top of each other.
-
-\section{Distance Estimation}\label{sec-model-distance-estimation}
-
-In the previous sections, we have defined a distance measure given in
-Equation \ref{eq:kl-0} for quantifying the difference between the true
-residual distribution \(P\) and an ideal reference distribution \(Q\).
-However, this distance measure can only be computed when the data
-generating process is known. In reality, we often have no knowledge
-about the data generating process, otherwise we do not need to do a
-residual diagnostic in the first place.
-
-We use a computer vision model to estimate this distance measure for a
-residual plot. Let \(D\) be the result of Equation \ref{eq:kl-0}, and
-our estimator \(\hat{D}\) is formulated as
-
-\begin{equation} \label{eq:d-approx}
-\hat{D} = f_{CV}(V_{h \times w}(\boldsymbol{e}, \hat{\boldsymbol{y}})),
-\end{equation}
-
-\noindent where \(V_{h \times w}(.)\) is a plotting function that saves
-a residuals vs fitted values plot with fixed aesthetic as an image with
-\(h \times w\) pixels in three channels (RGB), \(f_{CV}(.)\) is a
-computer vision model which takes an \(h \times w\) image as input and
-predicts the distance in the domain \([0, +\infty)\).
-
-With the estimated distance \(\hat{D}\), we can compare the underlying
-distribution of the residuals to a theoretically ``good'' residual
-distribution. \(\hat{D}\) can also be used as an index of the model
-violations indicating the strength of the visual signal embedded in the
-residual plot.
-
-It is not expected that \(\hat{D}\) will be equal to original distance
-\(D\). This is largely because information contained in a single
-residual plot is limited and it may not be able to summarize all the
-important characteristics of the residual distribution. For a given
-residual distribution \(P\), many different residual plots can be
-simulated, where many will share similar visual patterns, but some of
-them could be visually very different from the rest, especially for
-regression models with small \(n\). This suggests the error of the
-estimation will vary depends on whether the input residual plot is
-representative or not.
-
-\section{Statistical Testing}\label{sec-model-statistical-testing}
-
-\subsection{Lineup Evaluation}\label{sec-model-lineup-evaluation}
-
-Theoretically, the distance \(D\) for a correctly specified model is
-\(0\), because \(P\) will be the same as \(Q\). However, the computer
-vision model may not necessary predict \(0\) for a null plot. Using
-Figure \ref{fig:false-finding} as an example, it contains a visual
-pattern which is an indication of heteroskedasticity. We would not
-expect the model to be able to magically tell if the suspicious pattern
-is caused by the skewed distribution of the fitted values or the
-existence of heteroskedasticity. Additionally, some null plots could
-have outliers or strong visual patterns due to randomness, and a
-reasonable model will try to summarize those information into the
-prediction, resulting in \(\hat{D} > 0\).
-
-This property is not an issue if \(\hat{D} \gg 0\) for which the visual
-signal of the residual plot is very strong, and we usually do not need
-any further examination of the significance of the result. However, if
-the visual pattern is weak or moderate, having \(\hat{D}\) will not be
-sufficient to determine if \(H_0\) should be rejected.
-
-To address this issue we can adhere to the paradigm of visual inference,
-by comparing the estimated distance \(\hat{D}\) to the estimated
-distances for the null plots in a lineup. Specifically, if a lineup
-comprises 20 plots, the null hypothesis \(H_0\) will be rejected if
-\(\hat{D}\) exceeds the maximum estimated distance among the \(m - 1\)
-null plots, denoted as
-\(\max\limits_{1 \leq i \leq m-1} {\hat{D}_{null}^{(i)}}\), where
-\(\hat{D}_{null}^{(i)}\) represents the estimated distance for the
-\(i\)-th null plot. This approach is equivalent to the typical lineup
-protocol requiring a 95\% significance level, where \(H_0\) is rejected
-if the data plot is identified as the most distinct plot by the sole
-observer. The estimated distance serves as a metric to quantify the
-difference between the data plot and the null plots, as intended.
-
-Moreover, if the number of plots in a lineup, denoted by \(m\), is
-sufficiently large, the empirical distribution of
-\({\hat{D}_{null}^{(i)}}\) can be viewed as an approximation of the null
-distribution of the estimated distance. Consequently, quantiles of the
-null distribution can be estimated using the sample quantiles, and these
-quantiles can be utilized for decision-making purposes. The details of
-the sample quantile computation can be found in
-\citet{hyndman1996sample}. For instance, if \(\hat{D}\) is greater than
-or equal to the 95\% sample quantile, denoted as \(Q_{null}(0.95)\), we
-can conclude that the estimated distance for the true residual plot is
-significantly different from the estimated distance for null plots with
-a 95\% significance level. Based on our experience, to obtain a stable
-estimate of the 95\% quantile, the number of null plots, \(n_{null}\),
-typically needs to be at least 100. However, if the null distribution
-exhibits a long tail, a larger number of null plots may be required.
-Alternatively, a \(p\)-value is the probability of observing a distance
-equally or greater than \(\hat{D}\) under the null hypothesis \(H_0\),
-and it can be estimated by
-\(\frac{1}{m} + \frac{1}{m}\sum_{i=1}^{m-1}I\left(\hat{D}_{null}^{(i)} \geq \hat{D}\right)\).
-
-To alleviate computation burden, a lattice of quantiles for \(\hat{D}\)
-under \(H_0\) with specified sample sizes can be precomputed. We can
-then map the \(\hat{D}\) and sample size to the closet quantile and
-sample size in lattice to calculate the corresponding \(p\)-value. This
-approach lose precision in \(p\)-value calculation, however,
-significantly improves computational efficiency.
-
-\subsection{Bootstrapping}\label{bootstrapping}
-
-Bootstrap is often employed in linear regression when conducting
-inference for estimated parameters \citep[see][ and
-\citet{efron1994introduction}]{davison1997bootstrap}. It is typically
-done by sampling individual cases with replacement and refitting the
-regression model. If the observed data accurately reflects the true
-distribution of the population, the bootstrapped estimates can be used
-to measure the variability of the parameter estimate without making
-strong distributional assumptions about the data generating process.
-
-Similarly, bootstrap can be applied on the estimated distance
-\(\hat{D}\). For each refitted model \(M_{boot}^{(i)}\), there will be
-an associated residual plot \(V_{boot}^{(i)}\) which can be fed into the
-computer vision model to obtain \(\hat{D}_{boot}^{(i)}\), where
-\(i = 1,...,n_{boot}\), and \(n_{boot}\) is the number of bootstrapped
-samples. If we are interested in the variation of \(\hat{D}\), we can
-use \(\hat{D}_{boot}^{(i)}\) to estimate a confidence interval.
-
-Alternatively, since each \(M_{boot}^{(i)}\) has a set of estimated
-coefficients \(\hat{\boldsymbol{\beta}}_{boot}^{(i)}\) and an estimated
-variance \(\hat{\sigma^2}_{boot}^{(i)}\), a new approximated null
-distribution can be construed and the corresponding 95\% sample quantile
-\(Q_{boot}^{(i)}(0.95)\) can be computed. Then, if
-\(\hat{D}_{boot}^{(i)} \geq Q_{boot}^{(i)}(0.95)\), \(H_0\) will be
-rejected for \(M_{boot}^{(i)}\). The ratio of rejected
-\(M_{boot}^{(i)}\) among all the refitted models provides an indication
-of how often the assumed regression model are considered to be incorrect
-if the data can be obtained repetitively from the same data generating
-process. But this approach is computationally very expensive because it
-requires \(n_{boot} \times n_{null}\) times of residual plot assessment.
-In practice, \(Q_{null}(0.95)\) can be used to replace
-\(Q_{boot}^{(i)}(0.95)\) in the computation.
-
-\section{Model Violations Index}\label{sec-model-violations-index}
-
-While statistical testing is a powerful tool for detecting model
-violations, it can become cumbersome and time-consuming when quick
-decisions are needed, particularly due to the need to evaluate numerous
-null plots. In practice, a more convenient and immediate method for
-assessing model performance is often required. This is where an index,
-such as the Model Violations Index (MVI), becomes valuable. It offers a
-straightforward way to quantify deviations from model assumptions,
-enabling rapid assessment and easier comparison across models.
-
-The estimator \(\hat{D}\) measures the difference between the true
-residual distribution and the reference residual distribution, a
-difference primarily arises from deviations in model assumptions. The
-magnitude of \(D\) directly reflects the degree of these deviations,
-thus making \(\hat{D}\) instrumental in forming a model violations index
-(MVI).
-
-Note that if more observations are used for estimating the linear
-regression, the result of Equation \ref{eq:kl-1} will increase, as the
-integration will be performed with larger \(n\). For a given data
-generating process, \(D\) typically increases logarithmically with the
-number of observations. This behavior comes from the relationship
-\(D = \text{log}(1 + D_{KL})\), where
-\(D_{KL} = \sum_{i=1}^{n}D_{KL}^{(i)}\) under the assumption of
-independence.
-
-Since \(\hat{D}\) is an estimate of \(D\), it is expected that a larger
-number of observations will also lead to a higher \(\hat{D}\). However,
-this does not imply that \(\hat{D}\) fails to accurately represent the
-extent of model violations. In fact, when examining residual plots with
-more observations, we often observe a stronger visual signal strength,
-as the underlying patterns are more likely to be revealed, except in
-cases of significant overlapping.
-
-Therefore, the Model Violations Index (MVI) can be proposed as
-
-\begin{equation} \label{eq:mvi}
-\text{MVI} = C + \hat{D} - \log(n),
-\end{equation}
-
-\noindent where \(C\) is a large enough constant keeping the result
-positive and the term \(-\log(n)\) is used to offset the increase in
-\(D\) due to sample size.
-
-Figure \ref{fig:poly-heter-index} displays the residual plots for fitted
-models exhibiting varying degrees of non-linearity and
-heteroskedasticity. Each residual plot's MVI is computed using Equation
-\ref{eq:mvi} with \(C = 10\). When \(\text{MVI} > 8\), the visual
-patterns are notably strong and easily discernible by humans. In the
-range \(6 < \text{MVI} < 8\), the visibility of the visual pattern
-diminishes as MVI decreases. Conversely, when \(\text{MVI} < 6\), the
-visual pattern tends to become relatively faint and challenging to
-observe. Table \ref{tab:mvi} provides a summary of the MVI usage and it
-is applicable to other linear regression models.
-
-\begin{table}
-
-\caption{\label{tab:mvi}Degree of model violations or the strength of the visual signals according to the Model Violations Index (MVI). The constant $C$ is set to be 10.}
-\centering
-\begin{tabular}[t]{lc}
-\toprule
-Degree of model violations & Range ($C$ = 10)\\
-\midrule
-Strong & $\text{MVI} > 8$\\
-Moderate & $6 < \text{MVI} < 8$\\
-Weak & $\text{MVI} < 6$\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\begin{figure}[!h]
-
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/poly-heter-index-1}
-
-}
-
-\caption{Residual plots generated from fitted models exhibiting varying degrees of (A) non-linearity and (B) heteroskedasticity violations.
-The model violations index (MVI) is displayed atop each residual plot.
-The non-linearity patterns are relatively strong for $MVI > 8$, and relatively weak for $MVI < 6$, while the heteroskedasticity patterns are relatively strong for $MVI > 8$, and relatively weak for $MVI < 6$.}\label{fig:poly-heter-index}
-\end{figure}
-
-\section{Data Generation}\label{sec-model-data-generation}
-
-\subsection{Simulation Scheme}\label{simulation-scheme}
-
-While observational data is frequently employed in training models for
-real-world applications, the data generating process of observational
-data often remains unknown, making computation for our target variable
-\(D\) unattainable. Consequently, the computer vision models developed
-in this study were trained using synthetic data, including 80,000
-training images and 8,000 test images. This approach provided us with
-precise label annotations. Additionally, it ensured a large and diverse
-training dataset, as we had control over the data generating process,
-and the simulation of the training data was relatively cost-effective.
-
-We have incorporated three types of residual departures of linear
-regression model in the training data, including non-linearity,
-heteroskedasticity and non-normality. All three departures can be
-summarized by the data generating process formulated as
-
-\begin{align} \label{eq:data-sim}
-\boldsymbol{y} &= \boldsymbol{1}_n + \boldsymbol{x}_1 + \beta_1\boldsymbol{x}_2 + \beta_2(\boldsymbol{z} + \beta_1\boldsymbol{w}) + \boldsymbol{k} \odot \boldsymbol{\varepsilon}, \\
-\boldsymbol{z} &= \text{He}_j(g(\boldsymbol{x}_1, 2)), \\
-\boldsymbol{w} &= \text{He}_j(g(\boldsymbol{x}_2, 2)), \\
-\boldsymbol{k} &= \left[\boldsymbol{1}_n + b(2 - |a|)(\boldsymbol{x}_1 + \beta_1\boldsymbol{x}_2 - a\boldsymbol{1}_n)^{\circ2}\right]^{\circ1/2},
-\end{align}
-
-\noindent where \(\boldsymbol{y}\), \(\boldsymbol{x}_1\),
-\(\boldsymbol{x}_2\), \(\boldsymbol{z}\), \(\boldsymbol{w}\),
-\(\boldsymbol{k}\) and \(\boldsymbol{\varepsilon}\) are vectors of size
-\(n\), \(\boldsymbol{1}_n\) is a vector of ones of size \(n\),
-\(\boldsymbol{x}_1\) and \(\boldsymbol{x}_2\) are two independent
-predictors, \(\text{He}_j(.)\) is the \(j\)th-order probabilist's
-Hermite polynomials \citep{hermite1864nouveau}, \((.)^{\circ2}\) and
-\((.)^{\circ1/2}\) are Hadamard square and square root, \(\odot\) is the
-Hadamard product, and \(g(\boldsymbol{x}, k)\) is a scaling function to
-enforce the support of the random vector to be \([-k, k]^n\) defined as
-
-\[g(\boldsymbol{x}, k) = 2k \cdot \frac{\boldsymbol{x} - x_{\min}\boldsymbol{1}_n}{x_{\max} - x_{\min}} - k\boldsymbol{1}_n,~for~k > 0,\]
-\noindent where \(x_{\min} = \underset{i \in \{ 1,...,n\}}{\min} x_i\),
-\(x_{\max} = \underset{i \in \{ 1,...,n\}}{\max} x_i\) and \(x_i\) is
-the \(i\)-th entry of \(\boldsymbol{x}\).
-
-\begin{table}
-
-\caption{\label{tab:factor}Factors used in the data generating process for synthetic data simulation.
-Factor $j$ and $a$ controls the non-linearity shape and the heteroskedasticity shape respectively.
-Factor $b$, $\sigma_\varepsilon$ and $n$ control the signal strength.
-Factor $\text{dist}_\varepsilon$, $\text{dist}_{x1}$ and $\text{dist}_{x2}$ specifies the distribution of $\varepsilon$, $X_1$ and $X_2$ respectively.}
-\centering
-\begin{tabular}[t]{ll}
-\toprule
-Factor & Domain\\
-\midrule
-j & \{2, 3, ..., 18\}\\
-a & {}[-1, 1]\\
-b & {}[0, 100]\\
-$\beta_1$ & \{0, 1\}\\
-$\beta_2$ & \{0, 1\}\\
-\addlinespace
-$\text{dist}_\varepsilon$ & \{discrete, uniform, normal, lognormal\}\\
-$\text{dist}_{x1}$ & \{discrete, uniform, normal, lognormal\}\\
-$\text{dist}_{x2}$ & \{discrete, uniform, normal, lognormal\}\\
-$\sigma_{\varepsilon}$ & {}[0.0625, 9]\\
-$\sigma_{X1}$ & {}[0.3, 0.6]\\
-\addlinespace
-$\sigma_{X2}$ & {}[0.3, 0.6]\\
-n & {}[50, 500]\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-The residuals and fitted values of the fitted model were obtained by
-regressing \(\boldsymbol{y}\) on \(\boldsymbol{x}_1\). If
-\(\beta_1 \neq 0\), \(\boldsymbol{x}_2\) was also included in the design
-matrix. This data generation process was adapted from
-\citet{li2024plot}, where it was utilized to simulate residual plots
-exhibiting non-linearity and heteroskedasticity visual patterns for
-human subject experiments. A summary of the factors utilized in Equation
-\ref{eq:data-sim} is provided in Table \ref{tab:factor}.
-
-In Equation \ref{eq:data-sim}, \(\boldsymbol{z}\) and \(\boldsymbol{w}\)
-represent higher-order terms of \(\boldsymbol{x}_1\) and
-\(\boldsymbol{x}_2\), respectively. If \(\beta_2 \neq 0\), the
-regression model will encounter non-linearity issues. Parameter \(j\)
-serves as a shape parameter that controls the number of tuning points in
-the non-linear pattern. Typically, higher values of \(j\) lead to an
-increase in the number of tuning points, as illustrated in Figure
-\ref{fig:different-j}.
-
-\begin{figure}[!h]
-
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/different-j-1}
-
-}
-
-\caption{Non-linearity forms generated for the synthetic data simulation. The 17 shapes are generated by varying the order of polynomial given by $j$ in $He_j(.)$.}\label{fig:different-j}
-\end{figure}
-
-Additionally, scaling factor \(\boldsymbol{k}\) directly affects the
-error distribution and it is correlated with \(\boldsymbol{x}_1\) and
-\(\boldsymbol{x}_2\). If \(b \neq 0\) and
-\(\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n, \sigma^2\boldsymbol{I}_n)\),
-the constant variance assumption will be violated. Parameter \(a\) is a
-shape parameter controlling the location of the smallest variance in a
-residual plot as shown in Figure \ref{fig:different-a}.
-
-\begin{figure}[!h]
-
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/different-a-1}
-
-}
-
-\caption{Heteroskedasticity forms generated for the synthetic data simulation.
-Different shapes are controlled by the continuous factor $a$ between -1 and 1.
-For $a = -1$, the residual plot exhibits a "left-triangle" shape.
-And for $a = 1$, the residual plot exhibits a "right-triangle" shape.
-}\label{fig:different-a}
-\end{figure}
-
-\begin{figure}[!h]
-
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/different-e-1}
-
-}
-
-\caption{Non-normality forms generated for the synthetic data simulation. Four different error distributions including discrete, lognormal, normal and uniform are considered.}\label{fig:different-e}
-\end{figure}
-
-Non-normality violations arise from specifying a non-normal distribution
-for \(\boldsymbol{\varepsilon}\). In the synthetic data simulation, four
-distinct error distributions are considered, including discrete,
-uniform, normal, and lognormal distributions, as presented in Figure
-\ref{fig:different-e}. Each distribution imparts unique characteristics
-in the residual plot. The discrete error distribution introduces
-discrete clusters in residuals, while the lognormal distribution
-typically yields outliers. Uniform error distribution may result in
-residuals filling the entire space of the residual plot. All of these
-distributions exhibit visual distinctions from the normal error
-distribution.
-
-\begin{figure}[!h]
-
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/different-j-x2-1}
-
-}
-
-\caption{Residual plots of multiple linear regression models with non-linearity issues.
-The 17 shapes are generated by varying the order of polynomial given by $j$ in $He_j(.)$.
-A second predictor $\boldsymbol{x}_2$ is introduced to the regression model to create complex shapes.}\label{fig:different-j-x2}
-\end{figure}
-
-\begin{figure}[!h]
-
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/different-j-heter-1}
-
-}
-
-\caption{Residual plots of models violating both the non-linearity and the heteroskedasticity assumptions.
-The 17 shapes are generated by varying the order of polynomial given by $j$ in $He_j(.)$, and the "left-triangle" shape is introduced by setting $a = -1$.}\label{fig:different-j-heter}
-\end{figure}
-
-\begin{figure}[!h]
-
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/different-e-heter-1}
-
-}
-
-\caption{Residual plots of models violating both the non-normality and the heteroskedasticity assumptions.
-The four shapes are generated by using four different error distributions including discrete, lognormal, normal and uniform, and the "left-triangle" shape is introduced by setting $a = -1$.
-}\label{fig:different-e-heter}
-\end{figure}
-
-Equation \ref{eq:data-sim} accommodates the incorporation of the second
-predictor \(\boldsymbol{x}_2\). Introducing it into the data generation
-process by setting \(\beta_1 = 1\) significantly enhances the complexity
-of the shapes, as illustrated in Figure \ref{fig:different-j-x2}. In
-comparison to Figure \ref{fig:different-j}, Figure
-\ref{fig:different-j-x2} demonstrates that the non-linear shape
-resembles a surface rather than a single curve. This augmentation can
-facilitate the computer vision model in learning visual patterns from
-residual plots of the multiple linear regression model.
-
-In real-world analysis, it is not uncommon to encounter instances where
-multiple model violations coexist. In such cases, the residual plots
-often exhibit a mixed pattern of visual anomalies corresponding to
-different types of model violations. Figure \ref{fig:different-j-heter}
-and Figure \ref{fig:different-e-heter} show the visual patterns of
-models with multiple model violations.
-
-The predictors, \(\boldsymbol{x}_1\) and \(\boldsymbol{x}_2\), are
-randomly generated from four distinct distributions, including
-\(U(-1, 1)\) (uniform), \(N(0, 0.3^2)\) (normal),
-\(\text{lognormal}(0, 0.6^2)/3\) (skewed) and \(U\{-1, 1\}\) (discrete
-uniform).
-
-\subsection{Balanced Dataset}\label{balanced-dataset}
-
-To train a robust computer vision model, we deliberately controlled the
-distribution of the target variable \(D\) in the training data. We
-ensured that it followed a uniform distribution between \(0\) and \(7\).
-This was achieved by organizing \(50\) buckets, each exclusively
-accepting training samples with \(D\) falling within the range
-\([7(i - 1)/49, 7i/49)\) for \(i < 50\), where \(i\) represents the
-index of the \(i\)-th bucket. For the \(50\)-th bucket, any training
-samples with \(D \geq 7\) were accepted.
-
-With 80,000 training images prepared, each bucket accommodated a maximum
-of \(80000/ 50 = 1600\) training samples. The simulator iteratively
-sampled parameter values from the parameter space, generated residuals
-and fitted values using the data generation process, computed the
-distance, and checked if the sample fitted within the corresponding
-bucket. This process continued until all buckets were filled.
-
-Similarly, we adopted the same methodology to prepare 8,000 test images
-for performance evaluation and model diagnostics.
-
-\section{Model Architecture}\label{sec-model-architecture}
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/cnn-diag-1}
-
-}
-
-\caption{Diagram of the architecture of the optimized computer vision model.
-Numbers at the bottom of each box show the shape of the output of each layer.
-The band of each box drawn in a darker color indicates the use of the rectified linear unit activation function.
-Yellow boxes are 2D convolutional layers, orange boxes are pooling layers, the grey box is the concatenation layer, and the purple boxes are dense layers.}\label{fig:cnn-diag}
-\end{figure}
-
-The architecture of the computer vision model is adapted from the
-well-established VGG16 architecture \citep{simonyan2014very}. While more
-recent architectures like ResNet \citep{he2016deep} and
-DenseNet\citep{huang2017densely}, have achieved even greater
-performance, VGG16 remains a solid choice for many applications due to
-its simplicity and effectiveness. Our decision to use VGG16 aligns with
-our goal of starting with a proven and straightforward model. Figure
-\ref{fig:cnn-diag} provides a diagram of the architecture. More details
-about the neural network layers used in this study are provided in
-Appendix A.
-
-The model begins with an input layer of shape
-\(n \times h \times w \times 3\), capable of handling \(n\) RGB images.
-This is followed by a grayscale conversion layer utilizing the luma
-formula under the Rec. 601 standard \citep{series2011studio}, which
-converts the color image to grayscale. Grayscale suffices for our task
-since data points are plotted in black. We experiment with three
-combinations of \(h\) and \(w\): \(32 \times 32\), \(64 \times 64\), and
-\(128 \times 128\), aiming to achieve sufficiently high image resolution
-for the problem at hand.
-
-The processed image is used as the input for the first convolutional
-block. The model comprises at most five consecutive convolutional
-blocks, mirroring the original VGG16 architecture. Within each block,
-there are two 2D convolutional layers followed by two activation layers,
-respectively. Subsequently, a 2D max-pooling layer follows the second
-activation layer. The 2D convolutional layer convolves the input with a
-fixed number of \(3 \times 3\) convolution filters, while the 2D
-max-pooling layer downsamples the input along its spatial dimensions by
-taking the maximum value over a \(2 \times 2\) window for each channel
-of the input. The activation layer employs the rectified linear unit
-(ReLU) activation function, a standard practice in deep learning, which
-introduces a non-linear transformation of the output of the 2D
-convolutional layer. Additionally, to regularize training, a batch
-normalization layer is added after each 2D convolutional layer and
-before the activation layer. Finally, a dropout layer is appended at the
-end of each convolutional block to randomly set some inputs to zero
-during training, further aiding in regularization.
-
-The output of the last convolutional block is summarized by either a
-global max pooling layer or a global average pooling layer, resulting in
-a two-dimensional tensor. To leverage the information contained in
-scagnostics, this tensor is concatenated with an additional
-\(n \times 5\) tensor, which contains the ``Monotonic'', ``Sparse'',
-``Splines'', and ``Striped'' measures, along with the number of
-observations for \(n\) residual plots.
-
-The concatenated tensor is then fed into the final prediction block.
-This block consists of two fully-connected layers. The first layer
-contains at least \(128\) units, followed by a dropout layer.
-Occasionally, a batch normalization layer is inserted between the
-fully-connected layer and the dropout layer for regularization purposes.
-The second fully-connected layer consists of only one unit, serving as
-the output of the model.
-
-The model weights \(\boldsymbol{\theta}\) were randomly initialized and
-they were optimized by the Adam optimizer \citep{kingma2014adam} with
-the mean square error loss function
-
-\[\hat{\boldsymbol{\theta}} = \underset{\boldsymbol{\theta}}{\text{arg min}}\frac{1}{n_{\text{train}}}\sum_{i=1}^{n_{\text{train}}}(D_i - f_{\boldsymbol{\theta}}(V_i, S_i))^2,\]
-
-\noindent where \(n_{\text{train}}\) is the number of training samples,
-\(V_i\) is the \(i\)-th residual plot and \(S_i\) is the additional
-information about the \(i\)-th residual plot including four scagnostics
-and the number of observations.
-
-\section{Model Training}\label{sec-model-training}
-
-To achieve a near-optimal deep learning model, hyperparameters like the
-learning rate often need to be fine-tuned using a tuner. In our study,
-we utilized the Bayesian optimization tuner from the \texttt{KerasTuner}
-Python library \citep{omalley2019kerastuner} for this purpose. A
-comprehensive list of hyperparameters is provided in Table
-\ref{tab:hyperparameter}.
-
-The number of base filters determines the number of filters for the
-first 2D convolutional layer. In the VGG16 architecture, the number of
-filters for the 2D convolutional layer in a block is typically twice the
-number in the previous block, except for the last block, which maintains
-the same number of convolution filters as the previous one. This
-hyperparameter aids in controlling the complexity of the computer vision
-model. A higher number of base filters results in more trainable
-parameters. Likewise, the number of units for the fully-connected layer
-determines the complexity of the final prediction block. Increasing the
-number of units enhances model complexity, resulting in more trainable
-parameters.
-
-The dropout rate and batch normalization are flexible hyperparameters
-that work in conjunction with other parameters to facilitate smooth
-training. A higher dropout rate is necessary when the model tends to
-overfit the training dataset by learning too much noise
-\citep{srivastava2014dropout}. Conversely, a lower dropout rate is
-preferred when the model is complex and challenging to converge. Batch
-normalization, on the other hand, addresses the internal covariate shift
-problem arising from the randomness in weight initialization and input
-data \citep{goodfellow2016deep}. It helps stabilize and accelerate the
-training process by normalizing the activations of each layer.
-
-Additionally, incorporating additional inputs such as scagnostics and
-the number of observations can potentially enhance prediction accuracy.
-Therefore, we allow the tuner to determine whether these inputs were
-necessary for optimal model performance.
-
-The learning rate is a crucial hyperparameter, as it dictates the step
-size of the optimization algorithm. A high learning rate can help the
-model avoid local minima but risks overshooting and missing the global
-minimum. Conversely, a low learning rate smoothens the training process
-but makes the convergence time longer and increases the likelihood of
-getting trapped in local minima.
-
-Our model was trained on the MASSIVE M3 high-performance computing
-platform \citep{goscinski2014multi}, using TensorFlow
-\citep{abadi2016tensorflow} and Keras \citep{chollet2015keras}. During
-training, 80\% of the training data was utilized for actual training,
-while the remaining 20\% was used as validation data. The Bayesian
-optimization tuner conducted 100 trials to identify the best
-hyperparameter values based on validation root mean square error. The
-tuner then restored the best epoch of the best model from the trials.
-Additionally, we applied early stopping, terminating the training
-process if the validation root mean square error fails to improve for 50
-epochs. The maximum allowed epochs was set at 2,000, although no models
-reached this threshold.
-
-\begin{table}
-
-\caption{\label{tab:hyperparameter}Name of hyperparameters and their correspoding domain for the computer vision model.}
-\centering
-\begin{tabular}[t]{ll}
-\toprule
-Hyperparameter & Domain\\
-\midrule
-Number of base filters & \{4, 8, 16, 32, 64\}\\
-Dropout rate for convolutional blocks & {}[0.1, 0.6]\\
-Batch normalization for convolutional blocks & \{false, true\}\\
-Type of global pooling & \{max, average\}\\
-Ignore additional inputs & \{false, true\}\\
-\addlinespace
-Number of units for the fully-connected layer & \{128, 256, 512, 1024, 2048\}\\
-Batch normalization for the fully-connected layer & \{false, true\}\\
-Dropout rate for the fully-connected layer & {}[0.1, 0.6]\\
-Learning rate & {}[$10^{-8}$, $10^{-1}$]\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-Based on the tuning process described above, the optimized
-hyperparameter values are presented in Table
-\ref{tab:best-hyperparameter}. It was observable that a minimum of
-\(32\) base filters was necessary, with the preferable choice being
-\(64\) base filters for both the \(64 \times 64\) and \(128 \times 128\)
-models, mirroring the original VGG16 architecture. The optimized dropout
-rate for convolutional blocks hovered around \(0.4\), and incorporating
-batch normalization for convolutional blocks proved beneficial for
-performance.
-
-All optimized models chose to retain the additional inputs, contributing
-to the reduction of validation error. The number of units required for
-the fully-connected layer was \(256\), a relatively modest number
-compared to the VGG16 classifier, suggesting that the problem at hand
-was less complex. The optimized learning rates were higher for models
-with higher resolution input, likely because models with more parameters
-are more prone to getting stuck in local minima, requiring a higher
-learning rate.
-
-\begin{table}
-\centering
-\caption{\label{tab:best-hyperparameter}Hyperparameters values for the optimized computer vision models with different input sizes.}
-\centering
-\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
-\begin{tabular}[t]{llll}
-\toprule
-Hyperparameter & $32 \times 32$ & $64 \times 64$ & $128 \times 128$\\
-\midrule
-Number of base filters & 32 & 64 & 64\\
-Dropout rate for convolutional blocks & 0.4 & 0.3 & 0.4\\
-Batch normalization for convolutional blocks & true & true & true\\
-Type of global pooling & max & average & average\\
-Ignore additional inputs & false & false & false\\
-\addlinespace
-Number of units for the fully-connected layer & 256 & 256 & 256\\
-Batch normalization for the fully-connected layer & false & true & true\\
-Dropout rate for the fully-connected layer & 0.2 & 0.4 & 0.1\\
-Learning rate & 0.0003 & 0.0006 & 0.0052\\
-\bottomrule
-\end{tabular}}
-\end{table}
-
-\section{Results}\label{sec-model-results}
-
-\subsection{Model Performance}\label{model-performance}
-
-The test performance for the optimized models with three different input
-sizes are summarized in Table \ref{tab:performance}. Among these models,
-the \(32 \times 32\) model consistently exhibited the best test
-performance. The mean absolute error of the \(32 \times 32\) model
-indicated that the difference between \(\hat{D}\) and \(D\) was
-approximately \(0.43\) on the test set, a negligible deviation
-considering the normal range of \(D\) typically falls between \(0\) and
-\(7\). The high \(R^2\) values also suggested that the predictions were
-largely linearly correlated with the target.
-
-Figure \ref{fig:model-performance} presents a hexagonal heatmap for
-\(D - \hat{D}\) versus \(D\). The brown smoothing curves, fitted by
-generalized additive models \citep{hastie2017generalized}, demonstrate
-that all the optimized models perform admirably on the test sets when
-\(1.5 < D < 6\), where no structural issues are noticeable. However,
-over-predictions occurred when \(D < 1.5\), while under-predictions
-occurred predominantly when \(\hat{D} > 6\).
-
-For input images representing null plots where \(D = 0\), it was
-expected that the models will over-predict the distance, as explained in
-Section \ref{sec-model-lineup-evaluation}. However, it can not explain
-the under-prediction issue. Therefore, we analysed the relationship
-between residuals and all the factors involved in the data generating
-process. We found that most issues actually arose from non-linearity
-problems and the presence of a second predictor in the regression model
-as illustrated in Figure \ref{fig:over-under}. When the variance for the
-error distribution was small, the optimized model tended to
-under-predict the distance. Conversely, when the error distribution had
-a large variance, the model tended to over-predict the distance.
-
-Since most of the deviation stemmed from the presence of non-linearity
-violations, to further investigate this, we split the test set based on
-violation types and re-evaluated the performance, as detailed in Table
-\ref{tab:performance-sub}. It was evident that metrics for null plots
-were notably worse compared to other categories. Furthermore, residual
-plots solely exhibiting non-normality issues were the easiest to
-predict, with very low test root mean square error (RMSE) at around
-\(0.3\). Residual plots with non-linearity issues were more challenging
-to assess than those with heteroskedasticity or non-normality issues.
-When multiple violations were introduced to a residual plot, the
-performance metrics typically lay between the metrics for each
-individual violation.
-
-Based on the model performance metrics, we chose to use the
-best-performing model evaluated on the test set, namely the
-\(32 \times 32\) model, for the subsequent analysis.
-
-\begin{table}
-
-\caption{\label{tab:performance}The test performance of three optimized models with different input sizes.}
-\centering
-\begin{tabular}[t]{lrrrr}
-\toprule
- & RMSE & $R^2$ & MAE & Huber loss\\
-\midrule
-$32 \times 32$ & 0.660 & 0.901 & 0.434 & 0.18\\
-$64 \times 64$ & 0.674 & 0.897 & 0.438 & 0.19\\
-$128 \times 128$ & 0.692 & 0.892 & 0.460 & 0.20\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\begin{figure}[!h]
-
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/model-performance-1}
-
-}
-
-\caption{Hexagonal heatmap for difference in $D$ and $\hat{D}$ vs $D$ on test data for three optimized models with different input sizes.
-The brown lines are smoothing curves produced by fitting generalized additive models.
-The area over the zero line in light yellow indicates under-prediction, and the area under the zero line in light green indicates over-prediction.}\label{fig:model-performance}
-\end{figure}
-
-\begin{figure}[!h]
-
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/over-under-1}
-
-}
-
-\caption{Scatter plots for difference in $D$ and $\hat{D}$ vs $\sigma$ on test data for the $32 \times 32$ optimized model.
-The data is grouped by whether the regression has only non-linearity violation, and whether it includes a second predictor in the regression formula.
-The brown lines are smoothing curves produced by fitting generalized additive models.
-The area over the zero line in light yellow indicates under-prediction, and the area under the zero line in light green indicates over-prediction.}\label{fig:over-under}
-\end{figure}
-
-\begin{table}
-
-\caption{\label{tab:performance-sub}The training and test performance of the $32 \times 32$ model presented with different model violations.}
-\centering
-\begin{tabular}[t]{lrr}
-\toprule
-Violations & \#samples & RMSE\\
-\midrule
-no violations & 155 & 1.267\\
-non-linearity & 2218 & 0.787\\
-heteroskedasticity & 1067 & 0.602\\
-non-linearity + heteroskedasticity & 985 & 0.751\\
-non-normality & 1111 & 0.320\\
-non-linearity + non-normality & 928 & 0.600\\
-heteroskedasticity + non-normality & 819 & 0.489\\
-non-linearity + heteroskedasticity + non-normality & 717 & 0.620\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\subsection{Comparison with Human Visual Inference and Conventional
-Tests}\label{comparison-with-human-visual-inference-and-conventional-tests}
-
-\subsubsection{Overview of the Human Subject
-Experiment}\label{overview-of-the-human-subject-experiment}
-
-In order to check the validity of the proposed computer vision model,
-residual plots presented in the human subject experiment conducted by
-\citet{li2024plot} will be assessed.
-
-This study has collected 7,974 human responses to 1,152 lineups. Each
-lineup contains one randomly placed true residual plot and 19 null
-plots. Among the 1,152 lineups, 24 are attention check lineups in which
-the visual patterns are designed to be extremely obvious and very
-different from the corresponding to null plots, 36 are null lineups
-where all the lineups consist of only null plots, 279 are lineups with
-uniform predictor distribution evaluated by 11 participants, and the
-remaining 813 are lineups with discrete, skewed or normal predictor
-distribution evaluated by 5 participants. Attention check lineups and
-null lineups will not be assessed in the following analysis.
-
-In \citet{li2024plot}, the residual plots are simulated from a data
-generating process which is a special case of Equation
-\ref{eq:data-sim}. The main characteristic is the model violations are
-introduced separately, meaning non-linearity and heteroskedasticity will
-not co-exist in one lineup but assigned uniformly to all lineups.
-Additionally, non-normality and multiple predictors are not considered
-in the experimental design.
-
-\subsubsection{Model Performance on the Human-evaluated
-Data}\label{model-performance-on-the-human-evaluated-data}
-
-\begin{table}
-
-\caption{\label{tab:experiment-performance}The performance of the $32 \times 32$ model on the data used in the human subject experiment.}
-\centering
-\begin{tabular}[t]{lrrrr}
-\toprule
-Violation & RMSE & $R^2$ & MAE & Huber loss\\
-\midrule
-heteroskedasticity & 0.721 & 0.852 & 0.553 & 0.235\\
-non-linearity & 0.738 & 0.770 & 0.566 & 0.246\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-For each lineup used in \citet{li2024plot}, there is one true residual
-plot and 19 null plots. While the distance \(D\) for the true residual
-plot depends on the underlying data generating process, the distance
-\(D\) for the null plots is zero. We have used our optimized computer
-vision model to estimate distance for both the true residual plots and
-the null plots. To have a fair comparison, \(H_0\) will be rejected if
-the true residual plot has the greatest estimated distance among all
-plots in a lineup. Additionally, the appropriate conventional tests
-including the Ramsey Regression Equation Specification Error Test
-(RESET) \citep{ramsey1969tests} for non-linearity and the Breusch-Pagan
-test \citep{breusch1979simple} for heteroskedasticity were applied on
-the same data for comparison.
-
-The performance metrics of \(\hat{D}\) for true residual plots are
-outlined in Table \ref{tab:experiment-performance}. It is notable that
-all performance metrics are slightly worse than those evaluated on the
-test data. Nevertheless, the mean absolute error remains at a low level,
-and the linear correlation between the prediction and the true value
-remains very high. Consistent with results in Table
-\ref{tab:performance-sub}, lineups with non-linearity issues are more
-challenging to predict than those with heteroskedasticity issues.
-
-Table \ref{tab:human-conv-table} provides a summary of the agreement
-between decisions made by the computer vision model and conventional
-tests. The agreement rates between conventional tests and the computer
-vision model are 85.95\% and 79.69\% for residual plots containing
-heteroskedasticity and non-linearity patterns, respectively. These
-figures are higher than those calculated for visual tests conducted by
-human, indicating that the computer vision model exhibits behavior more
-akin to the best available conventional tests. However, Figure
-\ref{fig:conv-mosaic} shows that the computer vision model does not
-always reject when the conventional tests reject. And a small number of
-plots will be rejected by computer vision model but not by conventional
-tests. This suggests that conventional tests are more sensitive than the
-computer vision model.
-
-Figure \ref{fig:pcp} further illustrates the decisions made by visual
-tests conducted by human, computer vision models, and conventional
-tests, using a parallel coordinate plots. It can be observed that all
-three tests will agree with each other for around 50\% of the cases.
-When visual tests conducted by human do not reject, there are
-substantial amount of cases where computer vision model also do not
-reject but conventional tests reject. There are much fewer cases that do
-not reject by visual tests and conventional tests, but is rejected by
-computer vision models. This indicates computer vision model can behave
-like visual tests conducted by human better than conventional tests.
-Moreover, there are great proportion of cases where visual tests
-conducted by human is the only test who does not reject.
-
-When plotting the decision against the distance, as illustrated in
-Figure \ref{fig:power}, several notable observations emerge. Firstly,
-compared to conventional tests, the computer vision model tends to have
-fewer rejected cases when \(D < 2\) and fewer non-rejected cases when
-\(2< D < 4\). This suggests tests based on the computer vision model are
-less sensitive to small deviations from model assumptions than
-conventional tests but more sensitive to moderate deviations.
-Additionally, visual tests demonstrate the lowest sensitivity to
-residual plots with small distances where not many residual plots are
-rejected when \(D < 2\). Similarly, for large distances where \(D > 4\),
-almost all residual plots are rejected by the computer vision model and
-conventional tests, but for visual tests conducted by humans, the
-threshold is higher with \(D > 5\).
-
-In Figure \ref{fig:power}, rejection decisions are fitted by logistic
-regression models with no intercept terms and an offset equals to
-\(\text{log}(0.05/0.95)\). The fitted curves for the computer vision
-model fall between those of conventional tests and visual tests for both
-non-linearity and heteroskedasticity, which means there is still
-potential to refine the computer vision model to better align its
-behavior with visual tests conducted by humans.
-
-In the experiment conducted in \citet{li2024plot}, participants were
-allowed to make multiple selections for a lineup. The weighted detection
-rate was computed by assigning weights to each detection. If the
-participant selected zero plots, a weight of 0.05 was assigned;
-otherwise, if the true residual plot was detected, the weight was 1
-divided by the number of selections. This weighted detection rate allow
-us to assess the quality of the distance measure purposed in this paper,
-by using the \(\delta\)-difference statistic. The \(\delta\)-difference
-is originally defined by \citet{chowdhury2018measuring}, is given by
-
-\[
-\delta = \bar{d}_{\text{true}} - \underset{j}{\text{max}}\left(\bar{d}_{\text{null}}^{(j)}\right) \quad \text{for}~j = 1,...,m-1,
-\]
-
-where \(\bar{d}_{\text{null}}^{(j)}\) is the mean distance between the
-\(j\)-th null plot and the other null plots, \(\bar{d}_{\text{true}}\)
-is the mean distance between the true residual plot and null plots, and
-\(m\) is the number of plots in a lineup. These mean distances are used
-because, as noted by \citet{chowdhury2018measuring}, the distances can
-vary depending on which data plot is used for comparison. For instance,
-with three null plots, A, B and C, the distance between A and B may
-differ from the distance between A and C. To obtain a consistent
-distance for null plot A, averaging is necessary. However, this approach
-is not applicable to the distance proposed in this paper, as we only
-compare the residual plot against a theoretically good residual plot.
-Consequently, the statistic must be adjusted to evaluate our distance
-measure effectively.
-
-One important aspect that the \(\delta\)-difference was designed to
-capture is the empirical distribution of distances for null plot. If we
-were to replace the mean distances \(\bar{d}_{\text{null}}^{(j)}\)
-directly with \(D_{\text{null}}^{(j)}\), the distance of the \(j\)-th
-null plot, the resulting distribution would be degenerate, since
-\(D_{null}\) equals zero by definition. Additionally, \(D\) can not be
-derived from an image, meaning it falls outside the scope of the
-distances considered by \citet{chowdhury2018measuring}. Instead, the
-focus should be on the empirical distribution of \(\hat{D}\), as it
-influences decision-making. Therefore, the adjusted \(\delta\)-different
-is defined as
-
-\[
-\delta_{\text{adj}} = \hat{D} - \underset{j}{\text{max}}\left(\hat{D}_{\text{null}}^{(j)}\right) \quad \text{for}~j = 1,...,m-1,
-\]
-
-\noindent where \(\hat{D}_{\text{null}}^{(j)}\) is the estimated
-distance for the \(j\)-th null plot, and \(m\) is the number of plots in
-a lineup.
-
-Figure \ref{fig:delta} displays the scatter plot of the weighted
-detection rate vs the adjusted \(\delta\)-difference. It indicates that
-the weighted detection rate increases as the adjusted
-\(\delta\)-difference increases, particularly when the adjusted
-\(\delta\)-difference is greater than zero. A negative adjusted
-\(\delta\)-difference suggests that there is at least one null plot in
-the lineup with a stronger visual signal than the true residual plot. In
-some instances, the weighted detection rate is close to one, yet the
-adjusted \(\delta\)-difference is negative. This discrepancy implies
-that the distance measure, or the estimated distance, may not perfectly
-reflect actual human behavior.
-
-\begin{table}
-
-\caption{\label{tab:human-conv-table}Summary of the comparison of decisions made by computer vision model with decisions made by conventional tests and visual tests conducted by human.}
-\centering
-\begin{tabular}[t]{lrrr}
-\toprule
-Violations & \#Samples & \#Agreements & Agreement rate\\
-\midrule
-\addlinespace[0.3em]
-\multicolumn{4}{l}{\textbf{Compared with conventional tests}}\\
-\hspace{1em}heteroskedasticity & 540 & 464 & 0.8593\\
-\hspace{1em}non-linearity & 576 & 459 & 0.7969\\
-\addlinespace[0.3em]
-\multicolumn{4}{l}{\textbf{Compared with visual tests conducted by human}}\\
-\hspace{1em}heteroskedasticity & 540 & 367 & 0.6796\\
-\hspace{1em}non-linearity & 576 & 385 & 0.6684\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\begin{figure}[!h]
-
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/conv-mosaic-1}
-
-}
-
-\caption{Rejection rate ($p$-value $\leq0.05$) of computer vision models conditional on conventional tests on non-linearity (left) and heteroskedasticity (right) lineups displayed using a mosaic plot.
-When the conventional test fails to reject, the computer vision mostly fails to reject the same plot as well as indicated by the height of the top right yellow rectangle, but there are non negliable amount of plots where the conventional test rejects but the computer vision model fails to reject as indicated by the width of the top left yellow rectangle.}\label{fig:conv-mosaic}
-\end{figure}
-
-\begin{figure}[!h]
-
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/power-1}
-
-}
-
-\caption{Comparison of power of visual tests, conventional tests and the computer vision model.
-Marks along the x-axis at the bottom of the plot represent rejections made by each type of test.
-Marks at the top of the plot represent acceptances.
-Power curves are fitted by logistic regression models with no intercept but an offset equals to $\text{log}(0.05/0.95)$.}\label{fig:power}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/pcp-1}
-
-}
-
-\caption{Parallel coordinate plots of decisions made by computer vision model, conventional tests and visual tests made by human.}\label{fig:pcp}
-\end{figure}
-
-\begin{figure}[!h]
-
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/delta-1}
-
-}
-
-\caption{A weighted detection rate vs adjusted $\delta$-difference plot. The brown line is smoothing curve produced by fitting generalized additive models.}\label{fig:delta}
-\end{figure}
-
-\section{Examples}\label{sec-examples}
-
-In this section, we present the performance of trained computer vision
-model on three example datasets. These include the dataset associated
-with the residual plot displaying a ``left-triangle'' shape, as
-displayed in Figure \ref{fig:false-finding}, along with the Boston
-housing dataset \citep{harrison1978hedonic}, and the ``dino'' datasets
-from the \texttt{datasauRus} R package \citep{datasaurus}.
-
-The first example illustrates a scenario where both the computer vision
-model and human visual inspection successfully avoid rejecting \(H_0\)
-when \(H_0\) is true, contrary to conventional tests. This underscores
-the necessity of visually examining the residual plot.
-
-In the second example, we encounter a more pronounced violation of the
-model, resulting in rejection of \(H_0\) by all three tests. This
-highlights the practicality of the computer vision model, particularly
-for less intricate tasks.
-
-The third example presents a situation where the model deviation is
-non-typical. Here, the computer vision model and human visual inspection
-reject \(H_0\), whereas some commonly used conventional tests do not.
-This emphasizes the benefits of visual inspection and the unique
-advantage of the computer vision model, which, like humans, makes
-decisions based on visual discoveries.
-
-\subsection{Left-triangle}\label{left-triangle}
-
-In Section \ref{sec-model-introduction}, we presented an example
-residual plot showcased in Figure \ref{fig:false-finding}, illustrating
-how humans might misinterpret the ``left-triangle'' shape as indicative
-of heteroskedasticity. Additionally, the Breusch-Pagan test yielded a
-rejection with a \(p\)-value of 0.046, despite the residuals originating
-from a correctly specified model. Figure \ref{fig:false-lineup} offers a
-lineup for this fitted model, showcasing various degrees of
-``left-triangle'' shape across all residual plots. This phenomenon is
-evidently caused by the skewed distribution of the fitted values.
-Notably, if the residual plot were evaluated through a visual test, it
-would not be rejected since the true residual plot positioned at 10 can
-not be distinguished from the others.
-
-Figure \ref{fig:false-check} presents the results of the assessment by
-the computer vision model. Notably, the observed visual signal strength
-is considerably lower than the 95\% sample quantile of the null
-distribution. Moreover, the bootstrapped distribution suggests that it
-is highly improbable for the fitted model to be misspecified as the
-majority of bootstrapped fitted models will not be rejected. Thus, for
-this particular fitted model, both the visual test and the computer
-vision model will not reject \(H_0\). However, the Breusch-Pagan test
-will reject \(H_0\) because it can not effectively utilize information
-from null plots.
-
-The attention map at Figure \ref{fig:false-check}B suggests that the
-estimation is highly influenced by the top-right and bottom-right part
-of the residual plot, as it forms two vertices of the triangular shape.
-A principal component analysis (PCA) is also performed on the output of
-the global pooling layer of the computer vision model. As mentioned in
-\citet{simonyan2014very}, a computer vision model built upon the
-convolutional blocks can be viewed as a feature extractor. For the
-\(32 \times 32\) model, there are 256 features outputted from the global
-pooling layer, which can be further used for different visual tasks not
-limited to distance prediction. To see if these features can be
-effectively used for distinguishing null plots and true residual plot,
-we linearly project them into the first and second principal components
-space as shown in Figure \ref{fig:false-check}D. It can be observed that
-because the bootstrapped plots are mostly similar to the null plots, the
-points drawn in different colors are mixed together. The true residual
-plot is also covered by both the cluster of null plots and cluster of
-bootstrapped plots. This accurately reflects our understanding of Figure
-\ref{fig:false-lineup}.
-
-\begin{figure}[!h]
-
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/false-check-1}
-
-}
-
-\caption{A summary of the residual plot assessment evaluted on 200 null plots and 200 bootstrapped plots.
-(A) The true residual plot exhibiting a "left-triangle" shape.
-(B) The attention map produced by computing the gradient of the output with respect to the greyscale input.
-(C) The density plot of estimated distance for null plots and bootstrapped plots.
-The green area indicates the distribution of estimated distances for bootstrapped plots, while the yellow area represents the distribution of estimated distances for null plots.
-The fitted model will not be rejected since $\hat{D} < Q_{null}(0.95)$.
-(D) plot of first two principal components of features extracted from the global pooling layer of the computer vision model.
-}\label{fig:false-check}
-\end{figure}
-
-\begin{figure}[!h]
-
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/false-lineup-1}
-
-}
-
-\caption{A lineup of residual plots displaying "left-triangle" visual patterns.
-The true residual plot occupies position 10, yet there are no discernible visual patterns that distinguish it from the other plots.}\label{fig:false-lineup}
-\end{figure}
-
-\subsection{Boston Housing}\label{boston-housing}
-
-The Boston housing dataset, originally published by
-\citet{harrison1978hedonic}, offers insights into housing in the Boston,
-Massachusetts area. For illustration purposes, we utilize a reduced
-version from Kaggle, comprising 489 rows and 4 columns: average number
-of rooms per dwelling (RM), percentage of lower status of the population
-(LSTAT), pupil-teacher ratio by town (PTRATIO), and median value of
-owner-occupied homes in \$1000's (MEDV). In our analysis, MEDV will
-serve as the response variable, while the other columns will function as
-predictors in a linear regression model. Our primary focus is to detect
-non-linearity, because the relationships between RM and MEDV or LSTAT
-and MEDV are non-linear.
-
-Figure \ref{fig:boston-check} displays the residual plot and the
-assessment conducted by the computer vision model. A clear non-linearity
-pattern resembling a ``U'' shape is shown in the plot A. Furthermore,
-the RESET test yields a very small \(p\)-value. The estimated distance
-\(\hat{D}\) significantly exceeds \(Q_{null}(0.95)\), leading to
-rejection of \(H_0\). The bootstrapped distribution also suggests that
-almost all the bootstrapped fitted models will be rejected, indicating
-that the fitted model is unlikely to be correctly specified. The
-attention map in plot B suggests the center of the image has higher
-leverage than other areas, and it is the turning point of the ``U''
-shape. The PCA provided in plot D shows two distinct clusters of data
-points, further underling the visual differences between bootstrapped
-plots and null plots. This coincides the findings from Figure
-\ref{fig:boston-lineup}, where the true plot exhibiting a ``U'' shape is
-visually distinctive from null plots. If a visual test is conducted by
-human, \(H_0\) will also be rejected.
-
-\begin{figure}[!h]
-
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/boston-check-1}
-
-}
-
-\caption{A summary of the residual plot assessment for the Boston housing fitted model evaluted on 200 null plots and 200 bootstrapped plots.
-(A) The true residual plot exhibiting a "U" shape.
-(B) The attention map produced by computing the gradient of the output with respect to the greyscale input.
-(C) The density plot of estimated distance for null plots and bootstrapped plots.
-The blue area indicates the distribution of estimated distances for bootstrapped plots, while the yellow area represents the distribution of estimated distances for null plots.
-The fitted model will be rejected since $\hat{D} \geq Q_{null}(0.95)$.
-(D) plot of first two principal components of features extracted from the global pooling layer of the computer vision model.
-}\label{fig:boston-check}
-\end{figure}
-
-\begin{figure}[!h]
-
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/boston-lineup-1}
-
-}
-
-\caption{A lineup of residual plots for the Boston housing fitted model. The true residual plot is at position 7. It can be easily identified as the most different plot.}\label{fig:boston-lineup}
-\end{figure}
-
-\subsection{DatasauRus}\label{datasaurus}
-
-The computer vision model possesses the capability to detect not only
-typical issues like non-linearity, heteroskedasticity, and non-normality
-but also artifact visual patterns resembling real-world objects, as long
-as they do not appear in null plots. These visual patterns can be
-challenging to categorize in terms of model violations. Therefore, we
-will employ the RESET test, the Breusch-Pagan test, and the Shapiro-Wilk
-test \citep{shapiro1965analysis} for comparison.
-
-The ``dino'' dataset within the \texttt{datasauRus} R package
-exemplifies this scenario. With only two columns, x and y, fitting a
-regression model to this data yields a residual plot resembling a
-``dinosaur'', as displayed in Figure \ref{fig:dino-check}A.
-Unsurprisingly, this distinct residual plot stands out in a lineup, as
-shown in Figure \ref{fig:dino-lineup}. A visual test conducted by humans
-would undoubtedly reject \(H_0\).
-
-According to the residual plot assessment by the computer vision model,
-\(\hat{D}\) exceeds \(Q_{null}(0.95)\), warranting a rejection of
-\(H_0\). Additionally, most of the bootstrapped fitted models will be
-rejected, indicating an misspecified model. However, both the RESET test
-and the Breusch-Pagan test yield \(p\)-values greater than 0.3, leading
-to a non-rejection of \(H_0\). Only the Shapiro-Wilk test rejects the
-normality assumption with a small \(p\)-value.
-
-More importantly, the attention map in Figure \ref{fig:dino-check}B
-clearly exhibits a ``dinosaur'' shape, strongly suggesting that the
-distance prediction is based on human-perceptible visual patterns. The
-computer vision model effectively captures the contour or outline of the
-embedded shape, similar to how humans interpret residual plots.
-Additionally, the PCA in Figure \ref{fig:dino-check}D demonstrates that
-the cluster of bootstrapped plots is positioned at the corner of the
-cluster of null plots.
-
-In practice, without accessing the residual plot, it would be
-challenging to identify the artificial pattern of the residuals.
-Moreover, conducting a normality test for a fitted regression model is
-not always standard practice among analysts. Even when performed,
-violating the normality assumption is sometimes deemed acceptable,
-especially considering the application of quasi-maximum likelihood
-estimation in linear regression. This example underscores the importance
-of evaluating residual plots and highlights how the proposed computer
-vision model can facilitate this process.
-
-\begin{figure}[!h]
-
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/dino-check-1}
-
-}
-
-\caption{A summary of the residual plot assessment for the datasauRus fitted model evaluated on 200 null plots and 200 bootstrapped plots.
-(A) The residual plot exhibits a "dinosaur" shape.
-(B) The attention map produced by computing the gradient of the output with respect to the greyscale input.
-(C) The density plot of estimated distance for null plots and bootstrapped plots.
-The blue area indicates the distribution of estimated distances for bootstrapped plots, while the yellow area represents the distribution of estimated distances for null plots.
-The fitted model will be rejected since $\hat{D} \geq Q_{null}(0.95)$.
-(D) plot of first two principal components of features extracted from the global pooling layer of the computer vision model.}\label{fig:dino-check}
-\end{figure}
-
-\begin{figure}[!h]
-
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/dino-lineup-1}
-
-}
-
-\caption{A lineup of residual plots for the fitted model on the "dinosaur" dataset.
-The true residual plot is at position 17.
-It can be easily identified as the most different plot as the visual pattern is extremely artificial.}\label{fig:dino-lineup}
-\end{figure}
-
-\section{Limitations and Future Work}\label{limitations-and-future-work}
-
-Despite the computer vision model performing well with general cases
-under the synthetic data generation scheme and the three examples used
-in this paper, this study has several limitations that could guide
-future work.
-
-The proposed distance measure assumes that the true model is a classical
-normal linear regression model, which can be restrictive. Although this
-paper does not address the relaxation of this assumption, there are
-potential methods to evaluate other types of regression models. The most
-comprehensive approach would be to define a distance measure for each
-different class of regression model and then train the computer vision
-model following the methodology described in this paper. To accelerate
-training, one could use the convolutional blocks of our trained model as
-a feature extractor and perform transfer learning on top of it, as these
-blocks effectively capture shapes in residual plots. Another approach
-would be to transform the residuals so they are roughly normally
-distributed and have constant variance. If only raw residuals are used,
-the distance-based statistical testing compares the difference in
-distance to a classical normal linear regression model for the true plot
-and null plots. This comparison is meaningful only if the difference can
-be identified by the distance measure proposed in this paper.
-
-There are other types of residual plots commonly used in diagnostics,
-such as residuals vs.~predictor and quantile-quantile plots. In this
-study, we focused on the most commonly used residual plot as a starting
-point for exploring the new field of automated visual inference.
-Similarly, we did not explore other, more sophisticated computer vision
-model architectures and specifications for the same reason. While the
-performance of the computer vision model is acceptable, there is still
-room for improvement to achieve behavior more closely resembling that of
-humans interpreting residual plots. This may require external survey
-data or human subject experiment data to understand the fundamental
-differences between our implementation and human evaluation.
-
-\section{Conclusions}\label{conclusions}
-
-In this paper, we have introduced a distance measure based on
-Kullback-Leibler divergence to quantify the disparity between the
-residual distribution of a fitted classical normal linear regression
-model and the reference residual distribution assumed under correct
-model specification. This distance measure effectively captures the
-magnitude of model violations in misspecified models. We propose a
-computer vision model to estimate this distance, utilizing the residual
-plot of the fitted model as input. The resulting estimated distance
-serves as the foundation for constructing a single Model Violations
-Index (MVI), facilitating the quantification of various model
-violations.
-
-Moreover, the estimated distance enables the development of a formal
-statistical testing procedure by evaluating a large number of null plots
-generated from the fitted model. Additionally, employing bootstrapping
-techniques and refitting the regression model allows us to ascertain how
-frequently the fitted model is considered misspecified if data were
-repeatedly obtained from the same data generating process.
-
-The trained computer vision model demonstrates strong performance on
-both the training and test sets, although it exhibits slightly lower
-performance on residual plots with non-linearity visual patterns
-compared to other types of violations. The statistical tests relying on
-the estimated distance predicted by the computer vision model exhibit
-lower sensitivity compared to conventional tests but higher sensitivity
-compared to visual tests conducted by humans. While the estimated
-distance generally mirrors the strength of the visual signal perceived
-by humans, there remains scope for further improvement in its
-performance.
-
-Several examples are provided to showcase the effectiveness of the
-proposed method across different scenarios, emphasizing the similarity
-between visual tests and distance-based tests. Overall, both visual
-tests and distance-based tests can be viewed as ensemble of tests,
-aiming to assess any violations of model assumptions collectively. In
-contrast, individual residual diagnostic tests such as the RESET test
-and the Breusch-Pagan test only evaluate specific violations of model
-assumptions. In practice, selecting an appropriate set of statistical
-tests for regression diagnostics can be challenging, particularly given
-the necessity of adjusting the significance level for each test.
-
-Our method holds significant value as it helps alleviate a portion of
-analysts' workload associated with assessing residual plots. While we
-recommend analysts to continue reading residual plots whenever feasible,
-as they offer invaluable insights, our approach serves as a valuable
-tool for automating the diagnostic process or for supplementary purposes
-when needed.
-
-\section*{Acknowledgement}\label{acknowledgement}
-\addcontentsline{toc}{section}{Acknowledgement}
-
-These \texttt{R} packages were used for the work: \texttt{tidyverse}
-\citep{tidyverse}, \texttt{lmtest} \citep{lmtest}, \texttt{mpoly}
-\citep{mpoly}, \texttt{ggmosaic} \citep{ggmosaic}, \texttt{kableExtra}
-\citep{kableextra}, \texttt{patchwork} \citep{patchwork},
-\texttt{rcartocolor} \citep{rcartocolor}, \texttt{glue} \citep{glue},
-\texttt{ggpcp} \citep{ggpcp}, \texttt{here} \citep{here},
-\texttt{magick} \citep{magick}, \texttt{yardstick} \citep{yardstick} and
-\texttt{reticulate} \citep{reticulate}.
-
-The article was created with R packages \texttt{rticles}
-\citep{rticles}, \texttt{knitr} \citep{knitr} and \texttt{rmarkdown}
-\citep{rmarkdown}. The project's GitHub repository
-(\url{https://github.com/TengMCing/auto_residual_reading_paper})
-contains all materials required to reproduce this article.
-
 \bibliographystyle{tfcad}

### 2025-06-12 appendix.Rmd
Make a copy of section 7 to Appendix

@@
 ---
+
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(
+  message = FALSE, 
+  warning = FALSE, 
+  echo = FALSE,
+  fig.width = 8,
+  fig.height = 6,
+  out.width = "100%", 
+  fig.align = "center")
+```
+
+```{r}
+# Visual inference models and p-value calculation
+# remotes::install_github("TengMCing/visage")
+options(tinytex.clean = FALSE)
+library(tidyverse)
+library(visage)
+library(glue)
+
+# To control the simulation in this file
+set.seed(10086)
+```
+
+# Data Generation {#sec-model-data-generation}
+
+## Simulation Scheme
+
+While observational data is frequently employed in training models for real-world applications, the data generating process of observational data often remains unknown, making computation for our target variable $D$ unattainable.
+Consequently, the computer vision models developed in this study were trained using synthetic data, including 80,000 training images and 8,000 test images.
+This approach provided us with precise label annotations.
+Additionally, it ensured a large and diverse training dataset, as we had control over the data generating process, and the simulation of the training data was relatively cost-effective.
+
+We have incorporated three types of residual departures of linear regression model in the training data, including non-linearity, heteroskedasticity and non-normality.
+All three departures can be summarized by the data generating process formulated as
+
+\begin{align} \label{eq:data-sim}
+\boldsymbol{y} &= \boldsymbol{1}_n + \boldsymbol{x}_1 + \beta_1\boldsymbol{x}_2 + \beta_2(\boldsymbol{z} + \beta_1\boldsymbol{w}) + \boldsymbol{k} \odot \boldsymbol{\varepsilon}, \\
+\boldsymbol{z} &= \text{He}_j(g(\boldsymbol{x}_1, 2)), \\
+\boldsymbol{w} &= \text{He}_j(g(\boldsymbol{x}_2, 2)), \\
+\boldsymbol{k} &= \left[\boldsymbol{1}_n + b(2 - |a|)(\boldsymbol{x}_1 + \beta_1\boldsymbol{x}_2 - a\boldsymbol{1}_n)^{\circ2}\right]^{\circ1/2},
+\end{align}
+
+\noindent where $\boldsymbol{y}$, $\boldsymbol{x}_1$, $\boldsymbol{x}_2$, $\boldsymbol{z}$, $\boldsymbol{w}$, $\boldsymbol{k}$ and $\boldsymbol{\varepsilon}$ are vectors of size $n$, $\boldsymbol{1}_n$ is a vector of ones of size $n$, $\boldsymbol{x}_1$ and $\boldsymbol{x}_2$ are two independent predictors, $\text{He}_j(.)$ is the $j$th-order probabilist's Hermite polynomials [@hermite1864nouveau], $(.)^{\circ2}$ and $(.)^{\circ1/2}$ are Hadamard square and square root, $\odot$ is the Hadamard product, and $g(\boldsymbol{x}, k)$ is a scaling function to enforce the support of the random vector to be $[-k, k]^n$ defined as
+
+$$g(\boldsymbol{x}, k) = 2k \cdot \frac{\boldsymbol{x} - x_{\min}\boldsymbol{1}_n}{x_{\max} - x_{\min}} - k\boldsymbol{1}_n,~for~k > 0,$$
+\noindent where $x_{\min} = \underset{i \in \{ 1,...,n\}}{\min} x_i$, $x_{\max} = \underset{i \in \{ 1,...,n\}}{\max} x_i$ and $x_i$ is the $i$-th entry of $\boldsymbol{x}$.
+
+```{r}
+tibble(Factor = c("j", "a", "b", "$\\beta_1$", "$\\beta_2$", " $\\text{dist}_\\varepsilon$", "$\\text{dist}_{x1}$", "$\\text{dist}_{x2}$", "$\\sigma_{\\varepsilon}$", "$\\sigma_{X1}$", "$\\sigma_{X2}$", "n"),
+       Domain = c("\\{2, 3, ..., 18\\}", "[-1, 1]", "[0, 100]", "\\{0, 1\\}", "\\{0, 1\\}", "\\{discrete, uniform, normal, lognormal\\}", "\\{discrete, uniform, normal, lognormal\\}", "\\{discrete, uniform, normal, lognormal\\}", "[0.0625, 9]", "[0.3, 0.6]", "[0.3, 0.6]", "[50, 500]")) %>%
+  kableExtra::kable(format = "latex",
+                    escape = FALSE,
+                    label = "factor",
+                    caption = "Factors used in the data generating process for synthetic data simulation.
+Factor $j$ and $a$ controls the non-linearity shape and the heteroskedasticity shape respectively.
+Factor $b$, $\\sigma_\\varepsilon$ and $n$ control the signal strength.
+Factor $\\text{dist}_\\varepsilon$, $\\text{dist}_{x1}$ and $\\text{dist}_{x2}$ specifies the distribution of $\\varepsilon$, $X_1$ and $X_2$ respectively.",
+                    booktabs = TRUE)
+```
+
+The residuals and fitted values of the fitted model were obtained by regressing $\boldsymbol{y}$ on $\boldsymbol{x}_1$.
+If $\beta_1 \neq 0$, $\boldsymbol{x}_2$ was also included in the design matrix.
+This data generation process was adapted from @li2024plot, where it was utilized to simulate residual plots exhibiting non-linearity and heteroskedasticity visual patterns for human subject experiments.
+A summary of the factors utilized in Equation \ref{eq:data-sim} is provided in Table \ref{tab:factor}.
+
+In Equation \ref{eq:data-sim}, $\boldsymbol{z}$ and $\boldsymbol{w}$ represent higher-order terms of $\boldsymbol{x}_1$ and $\boldsymbol{x}_2$, respectively.
+If $\beta_2 \neq 0$, the regression model will encounter non-linearity issues.
+Parameter $j$ serves as a shape parameter that controls the number of tuning points in the non-linear pattern.
+Typically, higher values of $j$ lead to an increase in the number of tuning points, as illustrated in Figure \ref{fig:different-j}.
+
+```{r different-j, fig.pos="!h", fig.cap = "Non-linearity forms generated for the synthetic data simulation. The 17 shapes are generated by varying the order of polynomial given by $j$ in $He_j(.)$."}
+
+set.seed(10086)
+
+# Data for shape 1
+dat_shape_1 <- phn_model(j = 2, include_x2 = FALSE, sigma = 0.05)$gen(500) %>%
+  mutate(j = 2)
+
+# Generate data for shape 2, 3 and 4. Reuse x and e.
+map_df(3:18, function(j) {
+  phn_model(j = j, include_x2 = FALSE, sigma = 0.05)$
+    gen(500, computed = select(dat_shape_1, x1, e)) %>%
+  mutate(j = j)
+}) %>%
+
+  # Combined with data for shape 1
+  bind_rows(dat_shape_1) %>%
+  mutate(j = factor(j)) %>%
+  VI_MODEL$plot(remove_axis = TRUE, remove_grid_line = TRUE, theme = theme_light()) +
+  facet_wrap(~j, scales = "free", labeller = label_parsed, ncol = 5)
+```
+
+Additionally, scaling factor $\boldsymbol{k}$ directly affects the error distribution and it is correlated with $\boldsymbol{x}_1$ and $\boldsymbol{x}_2$.
+If $b \neq 0$ and $\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n, \sigma^2\boldsymbol{I}_n)$, the constant variance assumption will be violated.
+Parameter $a$ is a shape parameter controlling the location of the smallest variance in a residual plot as shown in Figure \ref{fig:different-a}.
+
+```{r different-a, fig.width=8/5*3*2, fig.height=6/4*3*2, fig.pos="!h", fig.cap = 'Heteroskedasticity forms generated for the synthetic data simulation.
+Different shapes are controlled by the continuous factor $a$ between -1 and 1.
+For $a = -1$, the residual plot exhibits a "left-triangle" shape.
+And for $a = 1$, the residual plot exhibits a "right-triangle" shape.
+'}
+
+set.seed(10085)
+
+# Generate data for a = -1
+dat_a_n1 <- phn_model(include_z = FALSE,
+                      include_x2 = FALSE,
+                      a = -1,
+                      b = 100)$gen(500) %>%
+  mutate(a = -1)
+
+# Generate data for other a
+map(c(-0.75, -0.5, -0.25, 0, 0.25, 0.5, 0.75, 1), function(a) {
+  phn_model(include_z = FALSE,
+            include_x2 = FALSE,
+            a = a,
+            b = 100)$gen(500) %>%
+  mutate(a = a)
+}) %>%
+
+  # Combined with data for a = -1
+  bind_rows(dat_a_n1) %>%
+  mutate(a = factor(a)) %>%
+  VI_MODEL$plot(remove_axis = TRUE, remove_grid_line = TRUE, theme = theme_light()) +
+  facet_wrap(~a, scales = "free", ncol = 3) +
+  xlab("Fitted values") +
+  ylab("Residuals")
+```
+
+```{r different-e, fig.pos="!h", fig.cap = 'Non-normality forms generated for the synthetic data simulation.
+Four different error distributions including discrete, lognormal, normal and uniform are considered.'}
+
+set.seed(10086)
+
+# Data for shape 1
+dat_shape_1 <- phn_model(include_z = FALSE, include_x2 = FALSE, e = rand_uniform(-1.4, 1.4))$gen(500) %>%
+  mutate(e_dist = "uniform")
+
+dat_shape_2 <- phn_model(include_z = FALSE, include_x2 = FALSE, sigma = 0.8)$gen(500) %>%
+  mutate(e_dist = "normal")
+
+dat_shape_3 <- phn_model(include_z = FALSE, include_x2 = FALSE, e = rand_lognormal(sigma = 0.6))$gen(500) %>%
+  mutate(e_dist = "lognormal")
+
+dat_shape_4 <- phn_model(include_z = FALSE, include_x2 = FALSE, e = rand_uniform_d(-1.4, 1.4, even = TRUE))$gen(500) %>%
+  mutate(e_dist = "discrete")
+
+# Generate data for shape 2, 3 and 4. Reuse x and e.
+bind_rows(dat_shape_1, dat_shape_2, dat_shape_3, dat_shape_4) %>%
+  VI_MODEL$plot(remove_axis = TRUE, remove_grid_line = TRUE, theme = theme_light()) +
+  facet_wrap(~e_dist, scales = "free", labeller = label_parsed, ncol = 2)
+```
+
+Non-normality violations arise from specifying a non-normal distribution for $\boldsymbol{\varepsilon}$.
+In the synthetic data simulation, four distinct error distributions are considered, including discrete, uniform, normal, and lognormal distributions, as presented in Figure \ref{fig:different-e}.
+Each distribution imparts unique characteristics in the residual plot.
+The discrete error distribution introduces discrete clusters in residuals, while the lognormal distribution typically yields outliers.
+Uniform error distribution may result in residuals filling the entire space of the residual plot.
+All of these distributions exhibit visual distinctions from the normal error distribution.
+
+```{r different-j-x2, fig.pos="!h", fig.cap = "Residual plots of multiple linear regression models with non-linearity issues.
+The 17 shapes are generated by varying the order of polynomial given by $j$ in $He_j(.)$.
+A second predictor $\\boldsymbol{x}_2$ is introduced to the regression model to create complex shapes."}
+
+set.seed(10086)
+
+# Data for shape 1
+dat_shape_1 <- phn_model(j = 2, include_x2 = TRUE, sigma = 0.05)$gen(500) %>%
+  mutate(j = 2)
+
+# Generate data for shape 2, 3 and 4. Reuse x and e.
+map_df(3:18, function(j) {
+  phn_model(j = j, include_x2 = TRUE, sigma = 0.05)$
+    gen(500, computed = select(dat_shape_1, x1, e)) %>%
+  mutate(j = j)
+}) %>%
+
+  # Combined with data for shape 1
+  bind_rows(dat_shape_1) %>%
+  mutate(j = factor(j)) %>%
+  VI_MODEL$plot(remove_axis = TRUE, remove_grid_line = TRUE, theme = theme_light()) +
+  facet_wrap(~j, scales = "free", labeller = label_parsed, ncol = 5)
+```
+
+```{r different-j-heter, fig.pos="!h", fig.cap = 'Residual plots of models violating both the non-linearity and the heteroskedasticity assumptions.
+The 17 shapes are generated by varying the order of polynomial given by $j$ in $He_j(.)$, and the "left-triangle" shape is introduced by setting $a = -1$.'}
+
+set.seed(10086)
+
+# Data for shape 1
+dat_shape_1 <- phn_model(j = 2, a = -1, b = 100, include_x2 = FALSE, sigma = 0.05)$gen(500) %>%
+  mutate(j = 2)
+
+# Generate data for shape 2, 3 and 4. Reuse x and e.
+map_df(3:18, function(j) {
+  phn_model(j = j, a = -1, b = 100, include_x2 = FALSE, sigma = 0.05)$
+    gen(500, computed = select(dat_shape_1, x1, e)) %>%
+  mutate(j = j)
+}) %>%
+
+  # Combined with data for shape 1
+  bind_rows(dat_shape_1) %>%
+  mutate(j = factor(j)) %>%
+  VI_MODEL$plot(remove_axis = TRUE, remove_grid_line = TRUE, theme = theme_light()) +
+  facet_wrap(~j, scales = "free", labeller = label_parsed, ncol = 5)
+```
+
+```{r different-e-heter, fig.pos="!h", fig.cap = 'Residual plots of models violating both the non-normality and the heteroskedasticity assumptions.
+The four shapes are generated by using four different error distributions including discrete, lognormal, normal and uniform, and the "left-triangle" shape is introduced by setting $a = -1$.
+'}
+
+set.seed(10085)
+
+# Data for shape 1
+dat_shape_1 <- phn_model(a = -1, b = 100, include_z = FALSE, include_x2 = FALSE, e = rand_uniform(-1.4, 1.4))$gen(500) %>%
+  mutate(e_dist = "uniform")
+
+dat_shape_2 <- phn_model(a = -1, b = 100, include_z = FALSE, include_x2 = FALSE, sigma = 0.8)$gen(500) %>%
+  mutate(e_dist = "normal")
+
+dat_shape_3 <- phn_model(a = -1, b = 100, include_z = FALSE, include_x2 = FALSE, e = rand_lognormal(sigma = 0.6))$gen(500) %>%
+  mutate(e_dist = "lognormal")
+
+dat_shape_4 <- phn_model(a = -1, b = 100, include_z = FALSE, include_x2 = FALSE, e = rand_uniform_d(-1.4, 1.4, even = TRUE))$gen(500) %>%
+  mutate(e_dist = "discrete")
+
+# Generate data for shape 2, 3 and 4. Reuse x and e.
+bind_rows(dat_shape_1, dat_shape_2, dat_shape_3, dat_shape_4) %>%
+  VI_MODEL$plot(remove_axis = TRUE, remove_grid_line = TRUE, theme = theme_light()) +
+  facet_wrap(~e_dist, scales = "free", labeller = label_parsed, ncol = 2)
+```
+
+Equation \ref{eq:data-sim} accommodates the incorporation of the second predictor $\boldsymbol{x}_2$.
+Introducing it into the data generation process by setting $\beta_1 = 1$ significantly enhances the complexity of the shapes, as illustrated in Figure \ref{fig:different-j-x2}.
+In comparison to Figure \ref{fig:different-j}, Figure \ref{fig:different-j-x2} demonstrates that the non-linear shape resembles a surface rather than a single curve.
+This augmentation can facilitate the computer vision model in learning visual patterns from residual plots of the multiple linear regression model.
+
+In real-world analysis, it is not uncommon to encounter instances where multiple model violations coexist.
+In such cases, the residual plots often exhibit a mixed pattern of visual anomalies corresponding to different types of model violations.
+Figure \ref{fig:different-j-heter} and Figure \ref{fig:different-e-heter} show the visual patterns of models with multiple model violations.
+
+The predictors, $\boldsymbol{x}_1$ and $\boldsymbol{x}_2$, are randomly generated from four distinct distributions, including $U(-1, 1)$ (uniform), $N(0, 0.3^2)$ (normal), $\text{lognormal}(0, 0.6^2)/3$ (skewed) and $U\{-1, 1\}$ (discrete uniform).
+
+## Balanced Dataset
+
+To train a robust computer vision model, we deliberately controlled the distribution of the target variable $D$ in the training data.
+We ensured that it followed a uniform distribution between $0$ and $7$.
+This was achieved by organizing $50$ buckets, each exclusively accepting training samples with $D$ falling within the range $[7(i - 1)/49, 7i/49)$ for $i < 50$, where $i$ represents the index of the $i$-th bucket.
+For the $50$-th bucket, any training samples with $D \geq 7$ were accepted.
+
+With 80,000 training images prepared, each bucket accommodated a maximum of $80000/ 50 = 1600$ training samples.
+The simulator iteratively sampled parameter values from the parameter space, generated residuals and fitted values using the data generation process, computed the distance, and checked if the sample fitted within the corresponding bucket.
+This process continued until all buckets were filled.
+
+Similarly, we adopted the same methodology to prepare 8,000 test images for performance evaluation and model diagnostics.

### 2025-06-12 appendix.tex
Make a copy of section 7 to Appendix

@@
 \doublespacing
+
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{wrapfig}
+\usepackage{float}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
+\usepackage{xcolor}
 
@@
 
+\section{Data Generation}\label{sec-model-data-generation}
+
+\subsection{Simulation Scheme}\label{simulation-scheme}
+
+While observational data is frequently employed in training models for
+real-world applications, the data generating process of observational
+data often remains unknown, making computation for our target variable
+\(D\) unattainable. Consequently, the computer vision models developed
+in this study were trained using synthetic data, including 80,000
+training images and 8,000 test images. This approach provided us with
+precise label annotations. Additionally, it ensured a large and diverse
+training dataset, as we had control over the data generating process,
+and the simulation of the training data was relatively cost-effective.
+
+We have incorporated three types of residual departures of linear
+regression model in the training data, including non-linearity,
+heteroskedasticity and non-normality. All three departures can be
+summarized by the data generating process formulated as
+
+\begin{align} \label{eq:data-sim}
+\boldsymbol{y} &= \boldsymbol{1}_n + \boldsymbol{x}_1 + \beta_1\boldsymbol{x}_2 + \beta_2(\boldsymbol{z} + \beta_1\boldsymbol{w}) + \boldsymbol{k} \odot \boldsymbol{\varepsilon}, \\
+\boldsymbol{z} &= \text{He}_j(g(\boldsymbol{x}_1, 2)), \\
+\boldsymbol{w} &= \text{He}_j(g(\boldsymbol{x}_2, 2)), \\
+\boldsymbol{k} &= \left[\boldsymbol{1}_n + b(2 - |a|)(\boldsymbol{x}_1 + \beta_1\boldsymbol{x}_2 - a\boldsymbol{1}_n)^{\circ2}\right]^{\circ1/2},
+\end{align}
+
+\noindent where \(\boldsymbol{y}\), \(\boldsymbol{x}_1\),
+\(\boldsymbol{x}_2\), \(\boldsymbol{z}\), \(\boldsymbol{w}\),
+\(\boldsymbol{k}\) and \(\boldsymbol{\varepsilon}\) are vectors of size
+\(n\), \(\boldsymbol{1}_n\) is a vector of ones of size \(n\),
+\(\boldsymbol{x}_1\) and \(\boldsymbol{x}_2\) are two independent
+predictors, \(\text{He}_j(.)\) is the \(j\)th-order probabilist's
+Hermite polynomials \citep{hermite1864nouveau}, \((.)^{\circ2}\) and
+\((.)^{\circ1/2}\) are Hadamard square and square root, \(\odot\) is the
+Hadamard product, and \(g(\boldsymbol{x}, k)\) is a scaling function to
+enforce the support of the random vector to be \([-k, k]^n\) defined as
+
+\[g(\boldsymbol{x}, k) = 2k \cdot \frac{\boldsymbol{x} - x_{\min}\boldsymbol{1}_n}{x_{\max} - x_{\min}} - k\boldsymbol{1}_n,~for~k > 0,\]
+\noindent where \(x_{\min} = \underset{i \in \{ 1,...,n\}}{\min} x_i\),
+\(x_{\max} = \underset{i \in \{ 1,...,n\}}{\max} x_i\) and \(x_i\) is
+the \(i\)-th entry of \(\boldsymbol{x}\).
+
+\begin{table}
+
+\caption{\label{tab:factor}Factors used in the data generating process for synthetic data simulation.
+Factor $j$ and $a$ controls the non-linearity shape and the heteroskedasticity shape respectively.
+Factor $b$, $\sigma_\varepsilon$ and $n$ control the signal strength.
+Factor $\text{dist}_\varepsilon$, $\text{dist}_{x1}$ and $\text{dist}_{x2}$ specifies the distribution of $\varepsilon$, $X_1$ and $X_2$ respectively.}
+\centering
+\begin{tabular}[t]{ll}
+\toprule
+Factor & Domain\\
+\midrule
+j & \{2, 3, ..., 18\}\\
+a & {}[-1, 1]\\
+b & {}[0, 100]\\
+$\beta_1$ & \{0, 1\}\\
+$\beta_2$ & \{0, 1\}\\
+\addlinespace
+$\text{dist}_\varepsilon$ & \{discrete, uniform, normal, lognormal\}\\
+$\text{dist}_{x1}$ & \{discrete, uniform, normal, lognormal\}\\
+$\text{dist}_{x2}$ & \{discrete, uniform, normal, lognormal\}\\
+$\sigma_{\varepsilon}$ & {}[0.0625, 9]\\
+$\sigma_{X1}$ & {}[0.3, 0.6]\\
+\addlinespace
+$\sigma_{X2}$ & {}[0.3, 0.6]\\
+n & {}[50, 500]\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+The residuals and fitted values of the fitted model were obtained by
+regressing \(\boldsymbol{y}\) on \(\boldsymbol{x}_1\). If
+\(\beta_1 \neq 0\), \(\boldsymbol{x}_2\) was also included in the design
+matrix. This data generation process was adapted from
+\citet{li2024plot}, where it was utilized to simulate residual plots
+exhibiting non-linearity and heteroskedasticity visual patterns for
+human subject experiments. A summary of the factors utilized in Equation
+\ref{eq:data-sim} is provided in Table \ref{tab:factor}.
+
+In Equation \ref{eq:data-sim}, \(\boldsymbol{z}\) and \(\boldsymbol{w}\)
+represent higher-order terms of \(\boldsymbol{x}_1\) and
+\(\boldsymbol{x}_2\), respectively. If \(\beta_2 \neq 0\), the
+regression model will encounter non-linearity issues. Parameter \(j\)
+serves as a shape parameter that controls the number of tuning points in
+the non-linear pattern. Typically, higher values of \(j\) lead to an
+increase in the number of tuning points, as illustrated in Figure
+\ref{fig:different-j}.
+
+\begin{figure}[!h]
+
+{\centering \includegraphics[width=1\linewidth]{appendix_files/figure-latex/different-j-1}
+
+}
+
+\caption{Non-linearity forms generated for the synthetic data simulation. The 17 shapes are generated by varying the order of polynomial given by $j$ in $He_j(.)$.}\label{fig:different-j}
+\end{figure}
+
+Additionally, scaling factor \(\boldsymbol{k}\) directly affects the
+error distribution and it is correlated with \(\boldsymbol{x}_1\) and
+\(\boldsymbol{x}_2\). If \(b \neq 0\) and
+\(\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n, \sigma^2\boldsymbol{I}_n)\),
+the constant variance assumption will be violated. Parameter \(a\) is a
+shape parameter controlling the location of the smallest variance in a
+residual plot as shown in Figure \ref{fig:different-a}.
+
+\begin{figure}[!h]
+
+{\centering \includegraphics[width=1\linewidth]{appendix_files/figure-latex/different-a-1}
+
+}
+
+\caption{Heteroskedasticity forms generated for the synthetic data simulation.
+Different shapes are controlled by the continuous factor $a$ between -1 and 1.
+For $a = -1$, the residual plot exhibits a "left-triangle" shape.
+And for $a = 1$, the residual plot exhibits a "right-triangle" shape.
+}\label{fig:different-a}
+\end{figure}
+
+\begin{figure}[!h]
+
+{\centering \includegraphics[width=1\linewidth]{appendix_files/figure-latex/different-e-1}
+
+}
+
+\caption{Non-normality forms generated for the synthetic data simulation. Four different error distributions including discrete, lognormal, normal and uniform are considered.}\label{fig:different-e}
+\end{figure}
+
+Non-normality violations arise from specifying a non-normal distribution
+for \(\boldsymbol{\varepsilon}\). In the synthetic data simulation, four
+distinct error distributions are considered, including discrete,
+uniform, normal, and lognormal distributions, as presented in Figure
+\ref{fig:different-e}. Each distribution imparts unique characteristics
+in the residual plot. The discrete error distribution introduces
+discrete clusters in residuals, while the lognormal distribution
+typically yields outliers. Uniform error distribution may result in
+residuals filling the entire space of the residual plot. All of these
+distributions exhibit visual distinctions from the normal error
+distribution.
+
+\begin{figure}[!h]
+
+{\centering \includegraphics[width=1\linewidth]{appendix_files/figure-latex/different-j-x2-1}
+
+}
+
+\caption{Residual plots of multiple linear regression models with non-linearity issues.
+The 17 shapes are generated by varying the order of polynomial given by $j$ in $He_j(.)$.
+A second predictor $\boldsymbol{x}_2$ is introduced to the regression model to create complex shapes.}\label{fig:different-j-x2}
+\end{figure}
+
+\begin{figure}[!h]
+
+{\centering \includegraphics[width=1\linewidth]{appendix_files/figure-latex/different-j-heter-1}
+
+}
+
+\caption{Residual plots of models violating both the non-linearity and the heteroskedasticity assumptions.
+The 17 shapes are generated by varying the order of polynomial given by $j$ in $He_j(.)$, and the "left-triangle" shape is introduced by setting $a = -1$.}\label{fig:different-j-heter}
+\end{figure}
+
+\begin{figure}[!h]
+
+{\centering \includegraphics[width=1\linewidth]{appendix_files/figure-latex/different-e-heter-1}
+
+}
+
+\caption{Residual plots of models violating both the non-normality and the heteroskedasticity assumptions.
+The four shapes are generated by using four different error distributions including discrete, lognormal, normal and uniform, and the "left-triangle" shape is introduced by setting $a = -1$.
+}\label{fig:different-e-heter}
+\end{figure}
+
+Equation \ref{eq:data-sim} accommodates the incorporation of the second
+predictor \(\boldsymbol{x}_2\). Introducing it into the data generation
+process by setting \(\beta_1 = 1\) significantly enhances the complexity
+of the shapes, as illustrated in Figure \ref{fig:different-j-x2}. In
+comparison to Figure \ref{fig:different-j}, Figure
+\ref{fig:different-j-x2} demonstrates that the non-linear shape
+resembles a surface rather than a single curve. This augmentation can
+facilitate the computer vision model in learning visual patterns from
+residual plots of the multiple linear regression model.
+
+In real-world analysis, it is not uncommon to encounter instances where
+multiple model violations coexist. In such cases, the residual plots
+often exhibit a mixed pattern of visual anomalies corresponding to
+different types of model violations. Figure \ref{fig:different-j-heter}
+and Figure \ref{fig:different-e-heter} show the visual patterns of
+models with multiple model violations.
+
+The predictors, \(\boldsymbol{x}_1\) and \(\boldsymbol{x}_2\), are
+randomly generated from four distinct distributions, including
+\(U(-1, 1)\) (uniform), \(N(0, 0.3^2)\) (normal),
+\(\text{lognormal}(0, 0.6^2)/3\) (skewed) and \(U\{-1, 1\}\) (discrete
+uniform).
+
+\subsection{Balanced Dataset}\label{balanced-dataset}
+
+To train a robust computer vision model, we deliberately controlled the
+distribution of the target variable \(D\) in the training data. We
+ensured that it followed a uniform distribution between \(0\) and \(7\).
+This was achieved by organizing \(50\) buckets, each exclusively
+accepting training samples with \(D\) falling within the range
+\([7(i - 1)/49, 7i/49)\) for \(i < 50\), where \(i\) represents the
+index of the \(i\)-th bucket. For the \(50\)-th bucket, any training
+samples with \(D \geq 7\) were accepted.
+
+With 80,000 training images prepared, each bucket accommodated a maximum
+of \(80000/ 50 = 1600\) training samples. The simulator iteratively
+sampled parameter values from the parameter space, generated residuals
+and fitted values using the data generation process, computed the
+distance, and checked if the sample fitted within the corresponding
+bucket. This process continued until all buckets were filled.
+
+Similarly, we adopted the same methodology to prepare 8,000 test images
+for performance evaluation and model diagnostics.
+
 \bibliographystyle{tfcad}

### 2025-06-12 paper.Rmd
Remove all section 7 figures

@@
 
-```{r different-j, fig.pos="!h", fig.cap = "Non-linearity forms generated for the synthetic data simulation. The 17 shapes are generated by varying the order of polynomial given by $j$ in $He_j(.)$."}
-
-set.seed(10086)
-
-# Data for shape 1
-dat_shape_1 <- phn_model(j = 2, include_x2 = FALSE, sigma = 0.05)$gen(500) %>%
-  mutate(j = 2)
-
-# Generate data for shape 2, 3 and 4. Reuse x and e.
-map_df(3:18, function(j) {
-  phn_model(j = j, include_x2 = FALSE, sigma = 0.05)$
-    gen(500, computed = select(dat_shape_1, x1, e)) %>%
-  mutate(j = j)
-}) %>%
-
-  # Combined with data for shape 1
-  bind_rows(dat_shape_1) %>%
-  mutate(j = factor(j)) %>%
-  VI_MODEL$plot(remove_axis = TRUE, remove_grid_line = TRUE, theme = theme_light()) +
-  facet_wrap(~j, scales = "free", labeller = label_parsed, ncol = 5)
-```
-
 Additionally, scaling factor $\boldsymbol{k}$ directly affects the error distribution and it is correlated with $\boldsymbol{x}_1$ and $\boldsymbol{x}_2$.
@@
 Parameter $a$ is a shape parameter controlling the location of the smallest variance in a residual plot as shown in Figure \ref{fig:different-a}.
-
-```{r different-a, fig.width=8/5*3*2, fig.height=6/4*3*2, fig.pos="!h", fig.cap = 'Heteroskedasticity forms generated for the synthetic data simulation.
-Different shapes are controlled by the continuous factor $a$ between -1 and 1.
-For $a = -1$, the residual plot exhibits a "left-triangle" shape.
-And for $a = 1$, the residual plot exhibits a "right-triangle" shape.
-'}
-
-set.seed(10085)
-
-# Generate data for a = -1
-dat_a_n1 <- phn_model(include_z = FALSE,
-                      include_x2 = FALSE,
-                      a = -1,
-                      b = 100)$gen(500) %>%
-  mutate(a = -1)
-
-# Generate data for other a
-map(c(-0.75, -0.5, -0.25, 0, 0.25, 0.5, 0.75, 1), function(a) {
-  phn_model(include_z = FALSE,
-            include_x2 = FALSE,
-            a = a,
-            b = 100)$gen(500) %>%
-  mutate(a = a)
-}) %>%
-
-  # Combined with data for a = -1
-  bind_rows(dat_a_n1) %>%
-  mutate(a = factor(a)) %>%
-  VI_MODEL$plot(remove_axis = TRUE, remove_grid_line = TRUE, theme = theme_light()) +
-  facet_wrap(~a, scales = "free", ncol = 3) +
-  xlab("Fitted values") +
-  ylab("Residuals")
-```
-
-```{r different-e, fig.pos="!h", fig.cap = 'Non-normality forms generated for the synthetic data simulation.
-Four different error distributions including discrete, lognormal, normal and uniform are considered.'}
-
-set.seed(10086)
-
-# Data for shape 1
-dat_shape_1 <- phn_model(include_z = FALSE, include_x2 = FALSE, e = rand_uniform(-1.4, 1.4))$gen(500) %>%
-  mutate(e_dist = "uniform")
-
-dat_shape_2 <- phn_model(include_z = FALSE, include_x2 = FALSE, sigma = 0.8)$gen(500) %>%
-  mutate(e_dist = "normal")
-
-dat_shape_3 <- phn_model(include_z = FALSE, include_x2 = FALSE, e = rand_lognormal(sigma = 0.6))$gen(500) %>%
-  mutate(e_dist = "lognormal")
-
-dat_shape_4 <- phn_model(include_z = FALSE, include_x2 = FALSE, e = rand_uniform_d(-1.4, 1.4, even = TRUE))$gen(500) %>%
-  mutate(e_dist = "discrete")
-
-# Generate data for shape 2, 3 and 4. Reuse x and e.
-bind_rows(dat_shape_1, dat_shape_2, dat_shape_3, dat_shape_4) %>%
-  VI_MODEL$plot(remove_axis = TRUE, remove_grid_line = TRUE, theme = theme_light()) +
-  facet_wrap(~e_dist, scales = "free", labeller = label_parsed, ncol = 2)
-```
 
@@
 All of these distributions exhibit visual distinctions from the normal error distribution.
-
-```{r different-j-x2, fig.pos="!h", fig.cap = "Residual plots of multiple linear regression models with non-linearity issues.
-The 17 shapes are generated by varying the order of polynomial given by $j$ in $He_j(.)$.
-A second predictor $\\boldsymbol{x}_2$ is introduced to the regression model to create complex shapes."}
-
-set.seed(10086)
-
-# Data for shape 1
-dat_shape_1 <- phn_model(j = 2, include_x2 = TRUE, sigma = 0.05)$gen(500) %>%
-  mutate(j = 2)
-
-# Generate data for shape 2, 3 and 4. Reuse x and e.
-map_df(3:18, function(j) {
-  phn_model(j = j, include_x2 = TRUE, sigma = 0.05)$
-    gen(500, computed = select(dat_shape_1, x1, e)) %>%
-  mutate(j = j)
-}) %>%
-
-  # Combined with data for shape 1
-  bind_rows(dat_shape_1) %>%
-  mutate(j = factor(j)) %>%
-  VI_MODEL$plot(remove_axis = TRUE, remove_grid_line = TRUE, theme = theme_light()) +
-  facet_wrap(~j, scales = "free", labeller = label_parsed, ncol = 5)
-```
-
-```{r different-j-heter, fig.pos="!h", fig.cap = 'Residual plots of models violating both the non-linearity and the heteroskedasticity assumptions.
-The 17 shapes are generated by varying the order of polynomial given by $j$ in $He_j(.)$, and the "left-triangle" shape is introduced by setting $a = -1$.'}
-
-set.seed(10086)
-
-# Data for shape 1
-dat_shape_1 <- phn_model(j = 2, a = -1, b = 100, include_x2 = FALSE, sigma = 0.05)$gen(500) %>%
-  mutate(j = 2)
-
-# Generate data for shape 2, 3 and 4. Reuse x and e.
-map_df(3:18, function(j) {
-  phn_model(j = j, a = -1, b = 100, include_x2 = FALSE, sigma = 0.05)$
-    gen(500, computed = select(dat_shape_1, x1, e)) %>%
-  mutate(j = j)
-}) %>%
-
-  # Combined with data for shape 1
-  bind_rows(dat_shape_1) %>%
-  mutate(j = factor(j)) %>%
-  VI_MODEL$plot(remove_axis = TRUE, remove_grid_line = TRUE, theme = theme_light()) +
-  facet_wrap(~j, scales = "free", labeller = label_parsed, ncol = 5)
-```
-
-```{r different-e-heter, fig.pos="!h", fig.cap = 'Residual plots of models violating both the non-normality and the heteroskedasticity assumptions.
-The four shapes are generated by using four different error distributions including discrete, lognormal, normal and uniform, and the "left-triangle" shape is introduced by setting $a = -1$.
-'}
-
-set.seed(10085)
-
-# Data for shape 1
-dat_shape_1 <- phn_model(a = -1, b = 100, include_z = FALSE, include_x2 = FALSE, e = rand_uniform(-1.4, 1.4))$gen(500) %>%
-  mutate(e_dist = "uniform")
-
-dat_shape_2 <- phn_model(a = -1, b = 100, include_z = FALSE, include_x2 = FALSE, sigma = 0.8)$gen(500) %>%
-  mutate(e_dist = "normal")
-
-dat_shape_3 <- phn_model(a = -1, b = 100, include_z = FALSE, include_x2 = FALSE, e = rand_lognormal(sigma = 0.6))$gen(500) %>%
-  mutate(e_dist = "lognormal")
-
-dat_shape_4 <- phn_model(a = -1, b = 100, include_z = FALSE, include_x2 = FALSE, e = rand_uniform_d(-1.4, 1.4, even = TRUE))$gen(500) %>%
-  mutate(e_dist = "discrete")
-
-# Generate data for shape 2, 3 and 4. Reuse x and e.
-bind_rows(dat_shape_1, dat_shape_2, dat_shape_3, dat_shape_4) %>%
-  VI_MODEL$plot(remove_axis = TRUE, remove_grid_line = TRUE, theme = theme_light()) +
-  facet_wrap(~e_dist, scales = "free", labeller = label_parsed, ncol = 2)
-```
 
@@
 The brown lines are smoothing curves produced by fitting generalized additive models.
-The area over the zero line in light yellow indicates under-prediction, and the area under the zero line in light green indicates over-prediction.", dev = 'png', dpi = 300}
+The area over the zero line in light yellow indicates under-prediction, and the area under the zero line in light green indicates over-prediction.", dev = 'png', dpi = 300, fig.height = 3}
 model_pred <- train_pred %>% 

### 2025-06-12 paper.tex
Remove all section 7 figures

@@
 
-\begin{figure}[!h]
-
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/different-j-1}
-
-}
-
-\caption{Non-linearity forms generated for the synthetic data simulation. The 17 shapes are generated by varying the order of polynomial given by $j$ in $He_j(.)$.}\label{fig:different-j}
-\end{figure}
-
 Additionally, scaling factor \(\boldsymbol{k}\) directly affects the
@@
 residual plot as shown in Figure \ref{fig:different-a}.
-
-\begin{figure}[!h]
-
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/different-a-1}
-
-}
-
-\caption{Heteroskedasticity forms generated for the synthetic data simulation.
-Different shapes are controlled by the continuous factor $a$ between -1 and 1.
-For $a = -1$, the residual plot exhibits a "left-triangle" shape.
-And for $a = 1$, the residual plot exhibits a "right-triangle" shape.
-}\label{fig:different-a}
-\end{figure}
-
-\begin{figure}[!h]
-
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/different-e-1}
-
-}
-
-\caption{Non-normality forms generated for the synthetic data simulation. Four different error distributions including discrete, lognormal, normal and uniform are considered.}\label{fig:different-e}
-\end{figure}
 
@@
 distribution.
-
-\begin{figure}[!h]
-
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/different-j-x2-1}
-
-}
-
-\caption{Residual plots of multiple linear regression models with non-linearity issues.
-The 17 shapes are generated by varying the order of polynomial given by $j$ in $He_j(.)$.
-A second predictor $\boldsymbol{x}_2$ is introduced to the regression model to create complex shapes.}\label{fig:different-j-x2}
-\end{figure}
-
-\begin{figure}[!h]
-
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/different-j-heter-1}
-
-}
-
-\caption{Residual plots of models violating both the non-linearity and the heteroskedasticity assumptions.
-The 17 shapes are generated by varying the order of polynomial given by $j$ in $He_j(.)$, and the "left-triangle" shape is introduced by setting $a = -1$.}\label{fig:different-j-heter}
-\end{figure}
-
-\begin{figure}[!h]
-
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/different-e-heter-1}
-
-}
-
-\caption{Residual plots of models violating both the non-normality and the heteroskedasticity assumptions.
-The four shapes are generated by using four different error distributions including discrete, lognormal, normal and uniform, and the "left-triangle" shape is introduced by setting $a = -1$.
-}\label{fig:different-e-heter}
-\end{figure}
 

### 2025-06-12 paper.Rmd
Attempt to replace section 7 with a summary

@@
 
-## Simulation Scheme
-
-While observational data is frequently employed in training models for real-world applications, the data generating process of observational data often remains unknown, making computation for our target variable $D$ unattainable.
-Consequently, the computer vision models developed in this study were trained using synthetic data, including 80,000 training images and 8,000 test images.
-This approach provided us with precise label annotations.
-Additionally, it ensured a large and diverse training dataset, as we had control over the data generating process, and the simulation of the training data was relatively cost-effective.
-
-We have incorporated three types of residual departures of linear regression model in the training data, including non-linearity, heteroskedasticity and non-normality.
-All three departures can be summarized by the data generating process formulated as
-
-\begin{align} \label{eq:data-sim}
-\boldsymbol{y} &= \boldsymbol{1}_n + \boldsymbol{x}_1 + \beta_1\boldsymbol{x}_2 + \beta_2(\boldsymbol{z} + \beta_1\boldsymbol{w}) + \boldsymbol{k} \odot \boldsymbol{\varepsilon}, \\
-\boldsymbol{z} &= \text{He}_j(g(\boldsymbol{x}_1, 2)), \\
-\boldsymbol{w} &= \text{He}_j(g(\boldsymbol{x}_2, 2)), \\
-\boldsymbol{k} &= \left[\boldsymbol{1}_n + b(2 - |a|)(\boldsymbol{x}_1 + \beta_1\boldsymbol{x}_2 - a\boldsymbol{1}_n)^{\circ2}\right]^{\circ1/2},
-\end{align}
-
-\noindent where $\boldsymbol{y}$, $\boldsymbol{x}_1$, $\boldsymbol{x}_2$, $\boldsymbol{z}$, $\boldsymbol{w}$, $\boldsymbol{k}$ and $\boldsymbol{\varepsilon}$ are vectors of size $n$, $\boldsymbol{1}_n$ is a vector of ones of size $n$, $\boldsymbol{x}_1$ and $\boldsymbol{x}_2$ are two independent predictors, $\text{He}_j(.)$ is the $j$th-order probabilist's Hermite polynomials [@hermite1864nouveau], $(.)^{\circ2}$ and $(.)^{\circ1/2}$ are Hadamard square and square root, $\odot$ is the Hadamard product, and $g(\boldsymbol{x}, k)$ is a scaling function to enforce the support of the random vector to be $[-k, k]^n$ defined as
-
-$$g(\boldsymbol{x}, k) = 2k \cdot \frac{\boldsymbol{x} - x_{\min}\boldsymbol{1}_n}{x_{\max} - x_{\min}} - k\boldsymbol{1}_n,~for~k > 0,$$
-\noindent where $x_{\min} = \underset{i \in \{ 1,...,n\}}{\min} x_i$, $x_{\max} = \underset{i \in \{ 1,...,n\}}{\max} x_i$ and $x_i$ is the $i$-th entry of $\boldsymbol{x}$.
-
-```{r}
-tibble(Factor = c("j", "a", "b", "$\\beta_1$", "$\\beta_2$", " $\\text{dist}_\\varepsilon$", "$\\text{dist}_{x1}$", "$\\text{dist}_{x2}$", "$\\sigma_{\\varepsilon}$", "$\\sigma_{X1}$", "$\\sigma_{X2}$", "n"),
-       Domain = c("\\{2, 3, ..., 18\\}", "[-1, 1]", "[0, 100]", "\\{0, 1\\}", "\\{0, 1\\}", "\\{discrete, uniform, normal, lognormal\\}", "\\{discrete, uniform, normal, lognormal\\}", "\\{discrete, uniform, normal, lognormal\\}", "[0.0625, 9]", "[0.3, 0.6]", "[0.3, 0.6]", "[50, 500]")) %>%
-  kableExtra::kable(format = "latex",
-                    escape = FALSE,
-                    label = "factor",
-                    caption = "Factors used in the data generating process for synthetic data simulation.
-Factor $j$ and $a$ controls the non-linearity shape and the heteroskedasticity shape respectively.
-Factor $b$, $\\sigma_\\varepsilon$ and $n$ control the signal strength.
-Factor $\\text{dist}_\\varepsilon$, $\\text{dist}_{x1}$ and $\\text{dist}_{x2}$ specifies the distribution of $\\varepsilon$, $X_1$ and $X_2$ respectively.",
-                    booktabs = TRUE)
-```
-
-The residuals and fitted values of the fitted model were obtained by regressing $\boldsymbol{y}$ on $\boldsymbol{x}_1$.
-If $\beta_1 \neq 0$, $\boldsymbol{x}_2$ was also included in the design matrix.
-This data generation process was adapted from @li2024plot, where it was utilized to simulate residual plots exhibiting non-linearity and heteroskedasticity visual patterns for human subject experiments.
-A summary of the factors utilized in Equation \ref{eq:data-sim} is provided in Table \ref{tab:factor}.
-
-In Equation \ref{eq:data-sim}, $\boldsymbol{z}$ and $\boldsymbol{w}$ represent higher-order terms of $\boldsymbol{x}_1$ and $\boldsymbol{x}_2$, respectively.
-If $\beta_2 \neq 0$, the regression model will encounter non-linearity issues.
-Parameter $j$ serves as a shape parameter that controls the number of tuning points in the non-linear pattern.
-Typically, higher values of $j$ lead to an increase in the number of tuning points, as illustrated in Figure \ref{fig:different-j}.
-
-Additionally, scaling factor $\boldsymbol{k}$ directly affects the error distribution and it is correlated with $\boldsymbol{x}_1$ and $\boldsymbol{x}_2$.
-If $b \neq 0$ and $\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n, \sigma^2\boldsymbol{I}_n)$, the constant variance assumption will be violated.
-Parameter $a$ is a shape parameter controlling the location of the smallest variance in a residual plot as shown in Figure \ref{fig:different-a}.
-
-Non-normality violations arise from specifying a non-normal distribution for $\boldsymbol{\varepsilon}$.
-In the synthetic data simulation, four distinct error distributions are considered, including discrete, uniform, normal, and lognormal distributions, as presented in Figure \ref{fig:different-e}.
-Each distribution imparts unique characteristics in the residual plot.
-The discrete error distribution introduces discrete clusters in residuals, while the lognormal distribution typically yields outliers.
-Uniform error distribution may result in residuals filling the entire space of the residual plot.
-All of these distributions exhibit visual distinctions from the normal error distribution.
-
-Equation \ref{eq:data-sim} accommodates the incorporation of the second predictor $\boldsymbol{x}_2$.
-Introducing it into the data generation process by setting $\beta_1 = 1$ significantly enhances the complexity of the shapes, as illustrated in Figure \ref{fig:different-j-x2}.
-In comparison to Figure \ref{fig:different-j}, Figure \ref{fig:different-j-x2} demonstrates that the non-linear shape resembles a surface rather than a single curve.
-This augmentation can facilitate the computer vision model in learning visual patterns from residual plots of the multiple linear regression model.
-
-In real-world analysis, it is not uncommon to encounter instances where multiple model violations coexist.
-In such cases, the residual plots often exhibit a mixed pattern of visual anomalies corresponding to different types of model violations.
-Figure \ref{fig:different-j-heter} and Figure \ref{fig:different-e-heter} show the visual patterns of models with multiple model violations.
-
-The predictors, $\boldsymbol{x}_1$ and $\boldsymbol{x}_2$, are randomly generated from four distinct distributions, including $U(-1, 1)$ (uniform), $N(0, 0.3^2)$ (normal), $\text{lognormal}(0, 0.6^2)/3$ (skewed) and $U\{-1, 1\}$ (discrete uniform).
-
-## Balanced Dataset
-
-To train a robust computer vision model, we deliberately controlled the distribution of the target variable $D$ in the training data.
-We ensured that it followed a uniform distribution between $0$ and $7$.
-This was achieved by organizing $50$ buckets, each exclusively accepting training samples with $D$ falling within the range $[7(i - 1)/49, 7i/49)$ for $i < 50$, where $i$ represents the index of the $i$-th bucket.
-For the $50$-th bucket, any training samples with $D \geq 7$ were accepted.
-
-With 80,000 training images prepared, each bucket accommodated a maximum of $80000/ 50 = 1600$ training samples.
-The simulator iteratively sampled parameter values from the parameter space, generated residuals and fitted values using the data generation process, computed the distance, and checked if the sample fitted within the corresponding bucket.
-This process continued until all buckets were filled.
-
-Similarly, we adopted the same methodology to prepare 8,000 test images for performance evaluation and model diagnostics.
+To enable supervised training of computer vision models for detecting violations in linear regression, we generated a synthetic dataset of 80,000 training and 8,000 test residual plots.
+This simulation-based approach provided full control over the data-generating process, allowing precise computation of $D$, cost-effective data scaling, and the ability to model diverse visual patterns of model violations.
+
+The simulated data incorporated three common types of residual departures: non-linearity, heteroskedasticity, and non-normality.
+These were introduced by fitting a standard simple linear regression model to data generated from a more flexible model based on Hermite polynomials, multiple predictors, and varying distributions in both predictors and error terms.
+A comprehensive grid of parameter combinations was explored to generate a wide range of residual patterns.
+Importantly, the simulation included scenarios with multiple violations occurring simultaneously.
+
+To ensure uniform coverage across the difficulty scale of the target variable $D$, a bucket sampling scheme was used to create a balanced dataset.
+Samples were iteratively simulated and accepted into one of 50 buckets, each representing a distinct range of $D$.
+
+All mathematical formulations, parameter specifications, and supporting figures are detailed in the Appendix.
 

### 2025-06-12 paper.tex
Attempt to replace section 7 with a summary

@@
 
-\subsection{Simulation Scheme}\label{simulation-scheme}
-
-While observational data is frequently employed in training models for
-real-world applications, the data generating process of observational
-data often remains unknown, making computation for our target variable
-\(D\) unattainable. Consequently, the computer vision models developed
-in this study were trained using synthetic data, including 80,000
-training images and 8,000 test images. This approach provided us with
-precise label annotations. Additionally, it ensured a large and diverse
-training dataset, as we had control over the data generating process,
-and the simulation of the training data was relatively cost-effective.
-
-We have incorporated three types of residual departures of linear
-regression model in the training data, including non-linearity,
-heteroskedasticity and non-normality. All three departures can be
-summarized by the data generating process formulated as
-
-\begin{align} \label{eq:data-sim}
-\boldsymbol{y} &= \boldsymbol{1}_n + \boldsymbol{x}_1 + \beta_1\boldsymbol{x}_2 + \beta_2(\boldsymbol{z} + \beta_1\boldsymbol{w}) + \boldsymbol{k} \odot \boldsymbol{\varepsilon}, \\
-\boldsymbol{z} &= \text{He}_j(g(\boldsymbol{x}_1, 2)), \\
-\boldsymbol{w} &= \text{He}_j(g(\boldsymbol{x}_2, 2)), \\
-\boldsymbol{k} &= \left[\boldsymbol{1}_n + b(2 - |a|)(\boldsymbol{x}_1 + \beta_1\boldsymbol{x}_2 - a\boldsymbol{1}_n)^{\circ2}\right]^{\circ1/2},
-\end{align}
-
-\noindent where \(\boldsymbol{y}\), \(\boldsymbol{x}_1\),
-\(\boldsymbol{x}_2\), \(\boldsymbol{z}\), \(\boldsymbol{w}\),
-\(\boldsymbol{k}\) and \(\boldsymbol{\varepsilon}\) are vectors of size
-\(n\), \(\boldsymbol{1}_n\) is a vector of ones of size \(n\),
-\(\boldsymbol{x}_1\) and \(\boldsymbol{x}_2\) are two independent
-predictors, \(\text{He}_j(.)\) is the \(j\)th-order probabilist's
-Hermite polynomials \citep{hermite1864nouveau}, \((.)^{\circ2}\) and
-\((.)^{\circ1/2}\) are Hadamard square and square root, \(\odot\) is the
-Hadamard product, and \(g(\boldsymbol{x}, k)\) is a scaling function to
-enforce the support of the random vector to be \([-k, k]^n\) defined as
-
-\[g(\boldsymbol{x}, k) = 2k \cdot \frac{\boldsymbol{x} - x_{\min}\boldsymbol{1}_n}{x_{\max} - x_{\min}} - k\boldsymbol{1}_n,~for~k > 0,\]
-\noindent where \(x_{\min} = \underset{i \in \{ 1,...,n\}}{\min} x_i\),
-\(x_{\max} = \underset{i \in \{ 1,...,n\}}{\max} x_i\) and \(x_i\) is
-the \(i\)-th entry of \(\boldsymbol{x}\).
-
-\begin{table}
-
-\caption{\label{tab:factor}Factors used in the data generating process for synthetic data simulation.
-Factor $j$ and $a$ controls the non-linearity shape and the heteroskedasticity shape respectively.
-Factor $b$, $\sigma_\varepsilon$ and $n$ control the signal strength.
-Factor $\text{dist}_\varepsilon$, $\text{dist}_{x1}$ and $\text{dist}_{x2}$ specifies the distribution of $\varepsilon$, $X_1$ and $X_2$ respectively.}
-\centering
-\begin{tabular}[t]{ll}
-\toprule
-Factor & Domain\\
-\midrule
-j & \{2, 3, ..., 18\}\\
-a & {}[-1, 1]\\
-b & {}[0, 100]\\
-$\beta_1$ & \{0, 1\}\\
-$\beta_2$ & \{0, 1\}\\
-\addlinespace
-$\text{dist}_\varepsilon$ & \{discrete, uniform, normal, lognormal\}\\
-$\text{dist}_{x1}$ & \{discrete, uniform, normal, lognormal\}\\
-$\text{dist}_{x2}$ & \{discrete, uniform, normal, lognormal\}\\
-$\sigma_{\varepsilon}$ & {}[0.0625, 9]\\
-$\sigma_{X1}$ & {}[0.3, 0.6]\\
-\addlinespace
-$\sigma_{X2}$ & {}[0.3, 0.6]\\
-n & {}[50, 500]\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-The residuals and fitted values of the fitted model were obtained by
-regressing \(\boldsymbol{y}\) on \(\boldsymbol{x}_1\). If
-\(\beta_1 \neq 0\), \(\boldsymbol{x}_2\) was also included in the design
-matrix. This data generation process was adapted from
-\citet{li2024plot}, where it was utilized to simulate residual plots
-exhibiting non-linearity and heteroskedasticity visual patterns for
-human subject experiments. A summary of the factors utilized in Equation
-\ref{eq:data-sim} is provided in Table \ref{tab:factor}.
-
-In Equation \ref{eq:data-sim}, \(\boldsymbol{z}\) and \(\boldsymbol{w}\)
-represent higher-order terms of \(\boldsymbol{x}_1\) and
-\(\boldsymbol{x}_2\), respectively. If \(\beta_2 \neq 0\), the
-regression model will encounter non-linearity issues. Parameter \(j\)
-serves as a shape parameter that controls the number of tuning points in
-the non-linear pattern. Typically, higher values of \(j\) lead to an
-increase in the number of tuning points, as illustrated in Figure
-\ref{fig:different-j}.
-
-Additionally, scaling factor \(\boldsymbol{k}\) directly affects the
-error distribution and it is correlated with \(\boldsymbol{x}_1\) and
-\(\boldsymbol{x}_2\). If \(b \neq 0\) and
-\(\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n, \sigma^2\boldsymbol{I}_n)\),
-the constant variance assumption will be violated. Parameter \(a\) is a
-shape parameter controlling the location of the smallest variance in a
-residual plot as shown in Figure \ref{fig:different-a}.
-
-Non-normality violations arise from specifying a non-normal distribution
-for \(\boldsymbol{\varepsilon}\). In the synthetic data simulation, four
-distinct error distributions are considered, including discrete,
-uniform, normal, and lognormal distributions, as presented in Figure
-\ref{fig:different-e}. Each distribution imparts unique characteristics
-in the residual plot. The discrete error distribution introduces
-discrete clusters in residuals, while the lognormal distribution
-typically yields outliers. Uniform error distribution may result in
-residuals filling the entire space of the residual plot. All of these
-distributions exhibit visual distinctions from the normal error
-distribution.
-
-Equation \ref{eq:data-sim} accommodates the incorporation of the second
-predictor \(\boldsymbol{x}_2\). Introducing it into the data generation
-process by setting \(\beta_1 = 1\) significantly enhances the complexity
-of the shapes, as illustrated in Figure \ref{fig:different-j-x2}. In
-comparison to Figure \ref{fig:different-j}, Figure
-\ref{fig:different-j-x2} demonstrates that the non-linear shape
-resembles a surface rather than a single curve. This augmentation can
-facilitate the computer vision model in learning visual patterns from
-residual plots of the multiple linear regression model.
-
-In real-world analysis, it is not uncommon to encounter instances where
-multiple model violations coexist. In such cases, the residual plots
-often exhibit a mixed pattern of visual anomalies corresponding to
-different types of model violations. Figure \ref{fig:different-j-heter}
-and Figure \ref{fig:different-e-heter} show the visual patterns of
-models with multiple model violations.
-
-The predictors, \(\boldsymbol{x}_1\) and \(\boldsymbol{x}_2\), are
-randomly generated from four distinct distributions, including
-\(U(-1, 1)\) (uniform), \(N(0, 0.3^2)\) (normal),
-\(\text{lognormal}(0, 0.6^2)/3\) (skewed) and \(U\{-1, 1\}\) (discrete
-uniform).
-
-\subsection{Balanced Dataset}\label{balanced-dataset}
-
-To train a robust computer vision model, we deliberately controlled the
-distribution of the target variable \(D\) in the training data. We
-ensured that it followed a uniform distribution between \(0\) and \(7\).
-This was achieved by organizing \(50\) buckets, each exclusively
-accepting training samples with \(D\) falling within the range
-\([7(i - 1)/49, 7i/49)\) for \(i < 50\), where \(i\) represents the
-index of the \(i\)-th bucket. For the \(50\)-th bucket, any training
-samples with \(D \geq 7\) were accepted.
-
-With 80,000 training images prepared, each bucket accommodated a maximum
-of \(80000/ 50 = 1600\) training samples. The simulator iteratively
-sampled parameter values from the parameter space, generated residuals
-and fitted values using the data generation process, computed the
-distance, and checked if the sample fitted within the corresponding
-bucket. This process continued until all buckets were filled.
-
-Similarly, we adopted the same methodology to prepare 8,000 test images
-for performance evaluation and model diagnostics.
+To enable supervised training of computer vision models for detecting
+violations in linear regression, we generated a synthetic dataset of
+80,000 training and 8,000 test residual plots. This simulation-based
+approach provided full control over the data-generating process,
+allowing precise computation of \(D\), cost-effective data scaling, and
+the ability to model diverse visual patterns of model violations.
+
+The simulated data incorporated three common types of residual
+departures: non-linearity, heteroskedasticity, and non-normality. These
+were introduced by fitting a standard simple linear regression model to
+data generated from a more flexible model based on Hermite polynomials,
+multiple predictors, and varying distributions in both predictors and
+error terms. A comprehensive grid of parameter combinations was explored
+to generate a wide range of residual patterns. Importantly, the
+simulation included scenarios with multiple violations occurring
+simultaneously.
+
+To ensure uniform coverage across the difficulty scale of the target
+variable \(D\), a bucket sampling scheme was used to create a balanced
+dataset. Samples were iteratively simulated and accepted into one of 50
+buckets, each representing a distinct range of \(D\).
+
+All mathematical formulations, parameter specifications, and supporting
+figures are detailed in the Appendix.
 

### 2025-06-12 appendix.Rmd
Make a copy of section 9 to Appendix

@@
 Similarly, we adopted the same methodology to prepare 8,000 test images for performance evaluation and model diagnostics.
+
+# Model Training {#sec-model-training}
+
+To achieve a near-optimal deep learning model, hyperparameters like the learning rate often need to be fine-tuned using a tuner.
+In our study, we utilized the Bayesian optimization tuner from the `KerasTuner` Python library [@omalley2019kerastuner] for this purpose.
+A comprehensive list of hyperparameters is provided in Table \ref{tab:hyperparameter}.
+
+The number of base filters determines the number of filters for the first 2D convolutional layer.
+In the VGG16 architecture, the number of filters for the 2D convolutional layer in a block is typically twice the number in the previous block, except for the last block, which maintains the same number of convolution filters as the previous one.
+This hyperparameter aids in controlling the complexity of the computer vision model.
+A higher number of base filters results in more trainable parameters.
+Likewise, the number of units for the fully-connected layer determines the complexity of the final prediction block.
+Increasing the number of units enhances model complexity, resulting in more trainable parameters.
+
+The dropout rate and batch normalization are flexible hyperparameters that work in conjunction with other parameters to facilitate smooth training.
+A higher dropout rate is necessary when the model tends to overfit the training dataset by learning too much noise [@srivastava2014dropout].
+Conversely, a lower dropout rate is preferred when the model is complex and challenging to converge.
+Batch normalization, on the other hand, addresses the internal covariate shift problem arising from the randomness in weight initialization and input data [@goodfellow2016deep].
+It helps stabilize and accelerate the training process by normalizing the activations of each layer.
+
+Additionally, incorporating additional inputs such as scagnostics and the number of observations can potentially enhance prediction accuracy.
+Therefore, we allow the tuner to determine whether these inputs were necessary for optimal model performance.
+
+The learning rate is a crucial hyperparameter, as it dictates the step size of the optimization algorithm.
+A high learning rate can help the model avoid local minima but risks overshooting and missing the global minimum.
+Conversely, a low learning rate smoothens the training process but makes the convergence time longer and increases the likelihood of getting trapped in local minima.
+
+Our model was trained on the MASSIVE M3 high-performance computing platform [@goscinski2014multi], using TensorFlow [@abadi2016tensorflow] and Keras [@chollet2015keras].
+During training, 80% of the training data was utilized for actual training, while the remaining 20% was used as validation data.
+The Bayesian optimization tuner conducted 100 trials to identify the best hyperparameter values based on validation root mean square error.
+The tuner then restored the best epoch of the best model from the trials.
+Additionally, we applied early stopping, terminating the training process if the validation root mean square error fails to improve for 50 epochs.
+The maximum allowed epochs was set at 2,000, although no models reached this threshold.
+
+```{r}
+data.frame(`Hyperparameter` = c("Number of base filters", "Dropout rate for convolutional blocks", "Batch normalization for convolutional blocks", "Type of global pooling", "Ignore additional inputs", "Number of units for the fully-connected layer", "Batch normalization for the fully-connected layer", "Dropout rate for the fully-connected layer", "Learning rate"),
+           Domain = c("\\{4, 8, 16, 32, 64\\}", "[0.1, 0.6]", "\\{false, true\\}", "\\{max, average\\}", "\\{false, true\\}", "\\{128, 256, 512, 1024, 2048\\}", "\\{false, true\\}", "[0.1, 0.6]", "[$10^{-8}$, $10^{-1}$]")) %>%
+  kableExtra::kable(format = "latex",
+                    booktabs = TRUE, 
+                    label = "hyperparameter",
+                    caption = "Name of hyperparameters and their correspoding domain for the computer vision model.",
+                    escape = FALSE)
+```
+
+Based on the tuning process described above, the optimized hyperparameter values are presented in Table \ref{tab:best-hyperparameter}.
+It was observable that a minimum of $32$ base filters was necessary, with the preferable choice being $64$ base filters for both the $64 \times 64$ and $128 \times 128$ models, mirroring the original VGG16 architecture.
+The optimized dropout rate for convolutional blocks hovered around $0.4$, and incorporating batch normalization for convolutional blocks proved beneficial for performance.
+
+All optimized models chose to retain the additional inputs, contributing to the reduction of validation error.
+The number of units required for the fully-connected layer was $256$, a relatively modest number compared to the VGG16 classifier, suggesting that the problem at hand was less complex.
+The optimized learning rates were higher for models with higher resolution input, likely because models with more parameters are more prone to getting stuck in local minima, requiring a higher learning rate.
+
+```{r}
+data.frame(`Hyperparameter` = c("Number of base filters", "Dropout rate for convolutional blocks", "Batch normalization for convolutional blocks", "Type of global pooling", "Ignore additional inputs", "Number of units for the fully-connected layer", "Batch normalization for the fully-connected layer", "Dropout rate for the fully-connected layer", "Learning rate"),
+           `32` = c("32", "0.4", "true", "max", "false", "256", "false", "0.2", "0.0003"),
+           `64` = c("64", "0.3", "true", "average", "false", "256", "true", "0.4", "0.0006"),
+           `128` = c("64", "0.4", "true", "average", "false", "256", "true", "0.1", "0.0052")) %>%
+  kableExtra::kable(format = "latex",
+                    booktabs = TRUE, 
+                    label = "best-hyperparameter",
+                    caption = "Hyperparameters values for the optimized computer vision models with different input sizes.",
+                    col.names = c("Hyperparameter", "$32 \\times 32$", "$64 \\times 64$", "$128 \\times 128$"),
+                    escape = FALSE) %>%
+  kableExtra::kable_styling(latex_options = c("scale_down"))
+```

### 2025-06-12 appendix.tex
Make a copy of section 9 to Appendix

@@
 
+\section{Model Training}\label{sec-model-training}
+
+To achieve a near-optimal deep learning model, hyperparameters like the
+learning rate often need to be fine-tuned using a tuner. In our study,
+we utilized the Bayesian optimization tuner from the \texttt{KerasTuner}
+Python library \citep{omalley2019kerastuner} for this purpose. A
+comprehensive list of hyperparameters is provided in Table
+\ref{tab:hyperparameter}.
+
+The number of base filters determines the number of filters for the
+first 2D convolutional layer. In the VGG16 architecture, the number of
+filters for the 2D convolutional layer in a block is typically twice the
+number in the previous block, except for the last block, which maintains
+the same number of convolution filters as the previous one. This
+hyperparameter aids in controlling the complexity of the computer vision
+model. A higher number of base filters results in more trainable
+parameters. Likewise, the number of units for the fully-connected layer
+determines the complexity of the final prediction block. Increasing the
+number of units enhances model complexity, resulting in more trainable
+parameters.
+
+The dropout rate and batch normalization are flexible hyperparameters
+that work in conjunction with other parameters to facilitate smooth
+training. A higher dropout rate is necessary when the model tends to
+overfit the training dataset by learning too much noise
+\citep{srivastava2014dropout}. Conversely, a lower dropout rate is
+preferred when the model is complex and challenging to converge. Batch
+normalization, on the other hand, addresses the internal covariate shift
+problem arising from the randomness in weight initialization and input
+data \citep{goodfellow2016deep}. It helps stabilize and accelerate the
+training process by normalizing the activations of each layer.
+
+Additionally, incorporating additional inputs such as scagnostics and
+the number of observations can potentially enhance prediction accuracy.
+Therefore, we allow the tuner to determine whether these inputs were
+necessary for optimal model performance.
+
+The learning rate is a crucial hyperparameter, as it dictates the step
+size of the optimization algorithm. A high learning rate can help the
+model avoid local minima but risks overshooting and missing the global
+minimum. Conversely, a low learning rate smoothens the training process
+but makes the convergence time longer and increases the likelihood of
+getting trapped in local minima.
+
+Our model was trained on the MASSIVE M3 high-performance computing
+platform \citep{goscinski2014multi}, using TensorFlow
+\citep{abadi2016tensorflow} and Keras \citep{chollet2015keras}. During
+training, 80\% of the training data was utilized for actual training,
+while the remaining 20\% was used as validation data. The Bayesian
+optimization tuner conducted 100 trials to identify the best
+hyperparameter values based on validation root mean square error. The
+tuner then restored the best epoch of the best model from the trials.
+Additionally, we applied early stopping, terminating the training
+process if the validation root mean square error fails to improve for 50
+epochs. The maximum allowed epochs was set at 2,000, although no models
+reached this threshold.
+
+\begin{table}
+
+\caption{\label{tab:hyperparameter}Name of hyperparameters and their correspoding domain for the computer vision model.}
+\centering
+\begin{tabular}[t]{ll}
+\toprule
+Hyperparameter & Domain\\
+\midrule
+Number of base filters & \{4, 8, 16, 32, 64\}\\
+Dropout rate for convolutional blocks & {}[0.1, 0.6]\\
+Batch normalization for convolutional blocks & \{false, true\}\\
+Type of global pooling & \{max, average\}\\
+Ignore additional inputs & \{false, true\}\\
+\addlinespace
+Number of units for the fully-connected layer & \{128, 256, 512, 1024, 2048\}\\
+Batch normalization for the fully-connected layer & \{false, true\}\\
+Dropout rate for the fully-connected layer & {}[0.1, 0.6]\\
+Learning rate & {}[$10^{-8}$, $10^{-1}$]\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+Based on the tuning process described above, the optimized
+hyperparameter values are presented in Table
+\ref{tab:best-hyperparameter}. It was observable that a minimum of
+\(32\) base filters was necessary, with the preferable choice being
+\(64\) base filters for both the \(64 \times 64\) and \(128 \times 128\)
+models, mirroring the original VGG16 architecture. The optimized dropout
+rate for convolutional blocks hovered around \(0.4\), and incorporating
+batch normalization for convolutional blocks proved beneficial for
+performance.
+
+All optimized models chose to retain the additional inputs, contributing
+to the reduction of validation error. The number of units required for
+the fully-connected layer was \(256\), a relatively modest number
+compared to the VGG16 classifier, suggesting that the problem at hand
+was less complex. The optimized learning rates were higher for models
+with higher resolution input, likely because models with more parameters
+are more prone to getting stuck in local minima, requiring a higher
+learning rate.
+
+\begin{table}
+\centering
+\caption{\label{tab:best-hyperparameter}Hyperparameters values for the optimized computer vision models with different input sizes.}
+\centering
+\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
+\begin{tabular}[t]{llll}
+\toprule
+Hyperparameter & $32 \times 32$ & $64 \times 64$ & $128 \times 128$\\
+\midrule
+Number of base filters & 32 & 64 & 64\\
+Dropout rate for convolutional blocks & 0.4 & 0.3 & 0.4\\
+Batch normalization for convolutional blocks & true & true & true\\
+Type of global pooling & max & average & average\\
+Ignore additional inputs & false & false & false\\
+\addlinespace
+Number of units for the fully-connected layer & 256 & 256 & 256\\
+Batch normalization for the fully-connected layer & false & true & true\\
+Dropout rate for the fully-connected layer & 0.2 & 0.4 & 0.1\\
+Learning rate & 0.0003 & 0.0006 & 0.0052\\
+\bottomrule
+\end{tabular}}
+\end{table}
+
 \bibliographystyle{tfcad}

### 2025-06-12 paper.Rmd
Attempt to replace section 9 with a summary

@@
 
-All mathematical formulations, parameter specifications, and supporting figures are detailed in the Appendix.
+The mathematical formulations of the synthetic data model, parameter specifications, supporting tables and example figures of visual patterns of simulated residuals are provided in the Appendix.
 
@@
 
-To achieve a near-optimal deep learning model, hyperparameters like the learning rate often need to be fine-tuned using a tuner.
-In our study, we utilized the Bayesian optimization tuner from the `KerasTuner` Python library [@omalley2019kerastuner] for this purpose.
-A comprehensive list of hyperparameters is provided in Table \ref{tab:hyperparameter}.
-
-The number of base filters determines the number of filters for the first 2D convolutional layer.
-In the VGG16 architecture, the number of filters for the 2D convolutional layer in a block is typically twice the number in the previous block, except for the last block, which maintains the same number of convolution filters as the previous one.
-This hyperparameter aids in controlling the complexity of the computer vision model.
-A higher number of base filters results in more trainable parameters.
-Likewise, the number of units for the fully-connected layer determines the complexity of the final prediction block.
-Increasing the number of units enhances model complexity, resulting in more trainable parameters.
-
-The dropout rate and batch normalization are flexible hyperparameters that work in conjunction with other parameters to facilitate smooth training.
-A higher dropout rate is necessary when the model tends to overfit the training dataset by learning too much noise [@srivastava2014dropout].
-Conversely, a lower dropout rate is preferred when the model is complex and challenging to converge.
-Batch normalization, on the other hand, addresses the internal covariate shift problem arising from the randomness in weight initialization and input data [@goodfellow2016deep].
-It helps stabilize and accelerate the training process by normalizing the activations of each layer.
-
-Additionally, incorporating additional inputs such as scagnostics and the number of observations can potentially enhance prediction accuracy.
-Therefore, we allow the tuner to determine whether these inputs were necessary for optimal model performance.
-
-The learning rate is a crucial hyperparameter, as it dictates the step size of the optimization algorithm.
-A high learning rate can help the model avoid local minima but risks overshooting and missing the global minimum.
-Conversely, a low learning rate smoothens the training process but makes the convergence time longer and increases the likelihood of getting trapped in local minima.
-
-Our model was trained on the MASSIVE M3 high-performance computing platform [@goscinski2014multi], using TensorFlow [@abadi2016tensorflow] and Keras [@chollet2015keras].
-During training, 80% of the training data was utilized for actual training, while the remaining 20% was used as validation data.
-The Bayesian optimization tuner conducted 100 trials to identify the best hyperparameter values based on validation root mean square error.
-The tuner then restored the best epoch of the best model from the trials.
-Additionally, we applied early stopping, terminating the training process if the validation root mean square error fails to improve for 50 epochs.
-The maximum allowed epochs was set at 2,000, although no models reached this threshold.
-
-```{r}
-data.frame(`Hyperparameter` = c("Number of base filters", "Dropout rate for convolutional blocks", "Batch normalization for convolutional blocks", "Type of global pooling", "Ignore additional inputs", "Number of units for the fully-connected layer", "Batch normalization for the fully-connected layer", "Dropout rate for the fully-connected layer", "Learning rate"),
-           Domain = c("\\{4, 8, 16, 32, 64\\}", "[0.1, 0.6]", "\\{false, true\\}", "\\{max, average\\}", "\\{false, true\\}", "\\{128, 256, 512, 1024, 2048\\}", "\\{false, true\\}", "[0.1, 0.6]", "[$10^{-8}$, $10^{-1}$]")) %>%
-  kableExtra::kable(format = "latex",
-                    booktabs = TRUE, 
-                    label = "hyperparameter",
-                    caption = "Name of hyperparameters and their correspoding domain for the computer vision model.",
-                    escape = FALSE)
-```
-
-Based on the tuning process described above, the optimized hyperparameter values are presented in Table \ref{tab:best-hyperparameter}.
-It was observable that a minimum of $32$ base filters was necessary, with the preferable choice being $64$ base filters for both the $64 \times 64$ and $128 \times 128$ models, mirroring the original VGG16 architecture.
-The optimized dropout rate for convolutional blocks hovered around $0.4$, and incorporating batch normalization for convolutional blocks proved beneficial for performance.
-
-All optimized models chose to retain the additional inputs, contributing to the reduction of validation error.
-The number of units required for the fully-connected layer was $256$, a relatively modest number compared to the VGG16 classifier, suggesting that the problem at hand was less complex.
-The optimized learning rates were higher for models with higher resolution input, likely because models with more parameters are more prone to getting stuck in local minima, requiring a higher learning rate.
-
-```{r}
-data.frame(`Hyperparameter` = c("Number of base filters", "Dropout rate for convolutional blocks", "Batch normalization for convolutional blocks", "Type of global pooling", "Ignore additional inputs", "Number of units for the fully-connected layer", "Batch normalization for the fully-connected layer", "Dropout rate for the fully-connected layer", "Learning rate"),
-           `32` = c("32", "0.4", "true", "max", "false", "256", "false", "0.2", "0.0003"),
-           `64` = c("64", "0.3", "true", "average", "false", "256", "true", "0.4", "0.0006"),
-           `128` = c("64", "0.4", "true", "average", "false", "256", "true", "0.1", "0.0052")) %>%
-  kableExtra::kable(format = "latex",
-                    booktabs = TRUE, 
-                    label = "best-hyperparameter",
-                    caption = "Hyperparameters values for the optimized computer vision models with different input sizes.",
-                    col.names = c("Hyperparameter", "$32 \\times 32$", "$64 \\times 64$", "$128 \\times 128$"),
-                    escape = FALSE) %>%
-  kableExtra::kable_styling(latex_options = c("scale_down"))
-```
+The computer vision models were trained on the MASSIVE M3 high-performance computing platform [@goscinski2014multi], using TensorFlow [@abadi2016tensorflow] and Keras [@chollet2015keras], with hyperparameters optimized via Bayesian tuning in `KerasTuner` [@omalley2019kerastuner] to minimize validation RMSE.
+The tuning process involved 100 trials, early stopping, and consideration of factors such as dropout, batch normalization, input resolution, and auxiliary inputs.
+Details of the hyperparameter domains, optimization strategy, and final selected values are provided in Appendix.
 

### 2025-06-12 paper.tex
Attempt to replace section 9 with a summary

@@
 
-All mathematical formulations, parameter specifications, and supporting
-figures are detailed in the Appendix.
+The mathematical formulations of the synthetic data model, parameter
+specifications, supporting tables and example figures of visual patterns
+of simulated residuals are provided in the Appendix.
 
@@
 
-To achieve a near-optimal deep learning model, hyperparameters like the
-learning rate often need to be fine-tuned using a tuner. In our study,
-we utilized the Bayesian optimization tuner from the \texttt{KerasTuner}
-Python library \citep{omalley2019kerastuner} for this purpose. A
-comprehensive list of hyperparameters is provided in Table
-\ref{tab:hyperparameter}.
-
-The number of base filters determines the number of filters for the
-first 2D convolutional layer. In the VGG16 architecture, the number of
-filters for the 2D convolutional layer in a block is typically twice the
-number in the previous block, except for the last block, which maintains
-the same number of convolution filters as the previous one. This
-hyperparameter aids in controlling the complexity of the computer vision
-model. A higher number of base filters results in more trainable
-parameters. Likewise, the number of units for the fully-connected layer
-determines the complexity of the final prediction block. Increasing the
-number of units enhances model complexity, resulting in more trainable
-parameters.
-
-The dropout rate and batch normalization are flexible hyperparameters
-that work in conjunction with other parameters to facilitate smooth
-training. A higher dropout rate is necessary when the model tends to
-overfit the training dataset by learning too much noise
-\citep{srivastava2014dropout}. Conversely, a lower dropout rate is
-preferred when the model is complex and challenging to converge. Batch
-normalization, on the other hand, addresses the internal covariate shift
-problem arising from the randomness in weight initialization and input
-data \citep{goodfellow2016deep}. It helps stabilize and accelerate the
-training process by normalizing the activations of each layer.
-
-Additionally, incorporating additional inputs such as scagnostics and
-the number of observations can potentially enhance prediction accuracy.
-Therefore, we allow the tuner to determine whether these inputs were
-necessary for optimal model performance.
-
-The learning rate is a crucial hyperparameter, as it dictates the step
-size of the optimization algorithm. A high learning rate can help the
-model avoid local minima but risks overshooting and missing the global
-minimum. Conversely, a low learning rate smoothens the training process
-but makes the convergence time longer and increases the likelihood of
-getting trapped in local minima.
-
-Our model was trained on the MASSIVE M3 high-performance computing
-platform \citep{goscinski2014multi}, using TensorFlow
-\citep{abadi2016tensorflow} and Keras \citep{chollet2015keras}. During
-training, 80\% of the training data was utilized for actual training,
-while the remaining 20\% was used as validation data. The Bayesian
-optimization tuner conducted 100 trials to identify the best
-hyperparameter values based on validation root mean square error. The
-tuner then restored the best epoch of the best model from the trials.
-Additionally, we applied early stopping, terminating the training
-process if the validation root mean square error fails to improve for 50
-epochs. The maximum allowed epochs was set at 2,000, although no models
-reached this threshold.
-
-\begin{table}
-
-\caption{\label{tab:hyperparameter}Name of hyperparameters and their correspoding domain for the computer vision model.}
-\centering
-\begin{tabular}[t]{ll}
-\toprule
-Hyperparameter & Domain\\
-\midrule
-Number of base filters & \{4, 8, 16, 32, 64\}\\
-Dropout rate for convolutional blocks & {}[0.1, 0.6]\\
-Batch normalization for convolutional blocks & \{false, true\}\\
-Type of global pooling & \{max, average\}\\
-Ignore additional inputs & \{false, true\}\\
-\addlinespace
-Number of units for the fully-connected layer & \{128, 256, 512, 1024, 2048\}\\
-Batch normalization for the fully-connected layer & \{false, true\}\\
-Dropout rate for the fully-connected layer & {}[0.1, 0.6]\\
-Learning rate & {}[$10^{-8}$, $10^{-1}$]\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-Based on the tuning process described above, the optimized
-hyperparameter values are presented in Table
-\ref{tab:best-hyperparameter}. It was observable that a minimum of
-\(32\) base filters was necessary, with the preferable choice being
-\(64\) base filters for both the \(64 \times 64\) and \(128 \times 128\)
-models, mirroring the original VGG16 architecture. The optimized dropout
-rate for convolutional blocks hovered around \(0.4\), and incorporating
-batch normalization for convolutional blocks proved beneficial for
-performance.
-
-All optimized models chose to retain the additional inputs, contributing
-to the reduction of validation error. The number of units required for
-the fully-connected layer was \(256\), a relatively modest number
-compared to the VGG16 classifier, suggesting that the problem at hand
-was less complex. The optimized learning rates were higher for models
-with higher resolution input, likely because models with more parameters
-are more prone to getting stuck in local minima, requiring a higher
-learning rate.
-
-\begin{table}
-\centering
-\caption{\label{tab:best-hyperparameter}Hyperparameters values for the optimized computer vision models with different input sizes.}
-\centering
-\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
-\begin{tabular}[t]{llll}
-\toprule
-Hyperparameter & $32 \times 32$ & $64 \times 64$ & $128 \times 128$\\
-\midrule
-Number of base filters & 32 & 64 & 64\\
-Dropout rate for convolutional blocks & 0.4 & 0.3 & 0.4\\
-Batch normalization for convolutional blocks & true & true & true\\
-Type of global pooling & max & average & average\\
-Ignore additional inputs & false & false & false\\
-\addlinespace
-Number of units for the fully-connected layer & 256 & 256 & 256\\
-Batch normalization for the fully-connected layer & false & true & true\\
-Dropout rate for the fully-connected layer & 0.2 & 0.4 & 0.1\\
-Learning rate & 0.0003 & 0.0006 & 0.0052\\
-\bottomrule
-\end{tabular}}
-\end{table}
+The computer vision models were trained on the MASSIVE M3
+high-performance computing platform \citep{goscinski2014multi}, using
+TensorFlow \citep{abadi2016tensorflow} and Keras
+\citep{chollet2015keras}, with hyperparameters optimized via Bayesian
+tuning in \texttt{KerasTuner} \citep{omalley2019kerastuner} to minimize
+validation RMSE. The tuning process involved 100 trials, early stopping,
+and consideration of factors such as dropout, batch normalization, input
+resolution, and auxiliary inputs. Details of the hyperparameter domains,
+optimization strategy, and final selected values are provided in
+Appendix.
 

### 2025-06-13 paper.Rmd
Combine model training with data generation into one section

@@
 Section \ref{sec-model-statistical-testing} covers the statistical tests based on the estimated distance, and Section \ref{sec-model-violations-index} introduces a Model Violations Index, which offers a quicker and more convenient assessment.
-Sections \ref{sec-model-data-generation}, \ref{sec-model-architecture}, and  \ref{sec-model-training} detail the data preparation, model architecture, and training process, respectively.
+Sections \ref{sec-model-architecture} and \ref{sec-model-data-generation} describe the model architecture and the corresponding data generation and training process, respectively.
 The results are presented in Section \ref{sec-model-results}.
@@
 In this study, we chose to define and use a distance between a true residual plot and a theoretically "good" residual plot.
-This is further explained in @sec-model-distance-between-residual-plots.
+This is further explained in \ref{sec-model-distance-estimation}.
 @vo2016localizing have also demonstrated that defining a proper distance between images can enhance the matching accuracy in image search compared to a binary outcome model.
@@
 
-# Data Generation {#sec-model-data-generation}
-
-To enable supervised training of computer vision models for detecting violations in linear regression, we generated a synthetic dataset of 80,000 training and 8,000 test residual plots.
-This simulation-based approach provided full control over the data-generating process, allowing precise computation of $D$, cost-effective data scaling, and the ability to model diverse visual patterns of model violations.
-
-The simulated data incorporated three common types of residual departures: non-linearity, heteroskedasticity, and non-normality.
-These were introduced by fitting a standard simple linear regression model to data generated from a more flexible model based on Hermite polynomials, multiple predictors, and varying distributions in both predictors and error terms.
-A comprehensive grid of parameter combinations was explored to generate a wide range of residual patterns.
-Importantly, the simulation included scenarios with multiple violations occurring simultaneously.
-
-To ensure uniform coverage across the difficulty scale of the target variable $D$, a bucket sampling scheme was used to create a balanced dataset.
-Samples were iteratively simulated and accepted into one of 50 buckets, each representing a distinct range of $D$.
-
-The mathematical formulations of the synthetic data model, parameter specifications, supporting tables and example figures of visual patterns of simulated residuals are provided in the Appendix.
-
 # Model Architecture {#sec-model-architecture}
@@
 
-# Model Training {#sec-model-training}
-
-The computer vision models were trained on the MASSIVE M3 high-performance computing platform [@goscinski2014multi], using TensorFlow [@abadi2016tensorflow] and Keras [@chollet2015keras], with hyperparameters optimized via Bayesian tuning in `KerasTuner` [@omalley2019kerastuner] to minimize validation RMSE.
-The tuning process involved 100 trials, early stopping, and consideration of factors such as dropout, batch normalization, input resolution, and auxiliary inputs.
-Details of the hyperparameter domains, optimization strategy, and final selected values are provided in Appendix.
+# Data Generation and Model Training {#sec-model-data-generation}
+
+To enable supervised training of computer vision models for detecting violations in linear regression, we generated a synthetic dataset of 80,000 training and 8,000 test residual plots.
+This simulation-based approach provided full control over the data-generating process, allowing precise computation of $D$, cost-effective data scaling, and the ability to model diverse visual patterns of model violations.
+
+The simulated data incorporated three common types of residual departures: non-linearity, heteroskedasticity, and non-normality.
+These were introduced by fitting a standard simple linear regression model to data generated from a more flexible model based on Hermite polynomials [@hermite1864nouveau], multiple predictors, and varying distributions in both predictors and error terms.
+A comprehensive grid of parameter combinations was explored to generate a wide range of residual patterns.
+Importantly, the simulation included scenarios with multiple violations occurring simultaneously.
+To ensure uniform coverage across the difficulty scale of the target variable $D$, a bucket sampling scheme was used to create a balanced dataset.
+Samples were iteratively simulated and accepted into one of 50 buckets, each representing a distinct range of $D$.
+
+Model training was conducted on the MASSIVE M3 high-performance computing platform [@goscinski2014multi] using TensorFlow [@abadi2016tensorflow] and Keras [@chollet2015keras].
+Hyperparameters were optimized via Bayesian tuning with KerasTuner [@omalley2019kerastuner], minimizing validation RMSE across 100 trials.
+The tuning process included early stopping and considered dropout, batch normalization, input resolution, and auxiliary inputs.
+
+Further details, including mathematical formulations of the simulation model, parameter specifications, sampling scheme, example residual plots and hyperparameter tuning configuration, are provided in the Appendix.
 
@@
 
-In @li2024plot, the residual plots are simulated from a data generating process which is a special case of Equation \ref{eq:data-sim}.
-The main characteristic is the model violations are introduced separately, meaning non-linearity and heteroskedasticity will not co-exist in one lineup but assigned uniformly to all lineups.
-Additionally, non-normality and multiple predictors are not considered in the experimental design.
+In @li2024plot, the residual plots are simulated from a data generating process that corresponds to a special case of the synthetic data model used in this study.
+A key feature of the design is that model violations are introduced independently, meaning non-linearity and heteroskedasticity do not co-exist within a single lineup but are instead assigned uniformly across different lineups.
+Moreover, the experimental design does not account for non-normality or multiple predictors.
 

### 2025-06-13 paper.tex
Combine model training with data generation into one section

@@
 introduces a Model Violations Index, which offers a quicker and more
-convenient assessment. Sections \ref{sec-model-data-generation},
-\ref{sec-model-architecture}, and \ref{sec-model-training} detail the
-data preparation, model architecture, and training process,
-respectively. The results are presented in Section
-\ref{sec-model-results}. Example dataset applications are discussed in
-Section \ref{sec-examples}. Finally, we conclude with a discussion of
-our findings and propose ideas for future research directions.
+convenient assessment. Sections \ref{sec-model-architecture} and
+\ref{sec-model-data-generation} describe the model architecture and the
+corresponding data generation and training process, respectively. The
+results are presented in Section \ref{sec-model-results}. Example
+dataset applications are discussed in Section \ref{sec-examples}.
+Finally, we conclude with a discussion of our findings and propose ideas
+for future research directions.
 
@@
 residual plot and a theoretically ``good'' residual plot. This is
-further explained in \citet{sec-model-distance-between-residual-plots}.
+further explained in \ref{sec-model-distance-estimation}.
 \citet{vo2016localizing} have also demonstrated that defining a proper
@@
 
-\section{Data Generation}\label{sec-model-data-generation}
-
-To enable supervised training of computer vision models for detecting
-violations in linear regression, we generated a synthetic dataset of
-80,000 training and 8,000 test residual plots. This simulation-based
-approach provided full control over the data-generating process,
-allowing precise computation of \(D\), cost-effective data scaling, and
-the ability to model diverse visual patterns of model violations.
-
-The simulated data incorporated three common types of residual
-departures: non-linearity, heteroskedasticity, and non-normality. These
-were introduced by fitting a standard simple linear regression model to
-data generated from a more flexible model based on Hermite polynomials,
-multiple predictors, and varying distributions in both predictors and
-error terms. A comprehensive grid of parameter combinations was explored
-to generate a wide range of residual patterns. Importantly, the
-simulation included scenarios with multiple violations occurring
-simultaneously.
-
-To ensure uniform coverage across the difficulty scale of the target
-variable \(D\), a bucket sampling scheme was used to create a balanced
-dataset. Samples were iteratively simulated and accepted into one of 50
-buckets, each representing a distinct range of \(D\).
-
-The mathematical formulations of the synthetic data model, parameter
-specifications, supporting tables and example figures of visual patterns
-of simulated residuals are provided in the Appendix.
-
 \section{Model Architecture}\label{sec-model-architecture}
@@
 
-\section{Model Training}\label{sec-model-training}
-
-The computer vision models were trained on the MASSIVE M3
-high-performance computing platform \citep{goscinski2014multi}, using
-TensorFlow \citep{abadi2016tensorflow} and Keras
-\citep{chollet2015keras}, with hyperparameters optimized via Bayesian
-tuning in \texttt{KerasTuner} \citep{omalley2019kerastuner} to minimize
-validation RMSE. The tuning process involved 100 trials, early stopping,
-and consideration of factors such as dropout, batch normalization, input
-resolution, and auxiliary inputs. Details of the hyperparameter domains,
-optimization strategy, and final selected values are provided in
-Appendix.
+\section{Data Generation and Model
+Training}\label{sec-model-data-generation}
+
+To enable supervised training of computer vision models for detecting
+violations in linear regression, we generated a synthetic dataset of
+80,000 training and 8,000 test residual plots. This simulation-based
+approach provided full control over the data-generating process,
+allowing precise computation of \(D\), cost-effective data scaling, and
+the ability to model diverse visual patterns of model violations.
+
+The simulated data incorporated three common types of residual
+departures: non-linearity, heteroskedasticity, and non-normality. These
+were introduced by fitting a standard simple linear regression model to
+data generated from a more flexible model based on Hermite polynomials
+\citep{hermite1864nouveau}, multiple predictors, and varying
+distributions in both predictors and error terms. A comprehensive grid
+of parameter combinations was explored to generate a wide range of
+residual patterns. Importantly, the simulation included scenarios with
+multiple violations occurring simultaneously. To ensure uniform coverage
+across the difficulty scale of the target variable \(D\), a bucket
+sampling scheme was used to create a balanced dataset. Samples were
+iteratively simulated and accepted into one of 50 buckets, each
+representing a distinct range of \(D\).
+
+Model training was conducted on the MASSIVE M3 high-performance
+computing platform \citep{goscinski2014multi} using TensorFlow
+\citep{abadi2016tensorflow} and Keras \citep{chollet2015keras}.
+Hyperparameters were optimized via Bayesian tuning with KerasTuner
+\citep{omalley2019kerastuner}, minimizing validation RMSE across 100
+trials. The tuning process included early stopping and considered
+dropout, batch normalization, input resolution, and auxiliary inputs.
+
+Further details, including mathematical formulations of the simulation
+model, parameter specifications, sampling scheme, example residual plots
+and hyperparameter tuning configuration, are provided in the Appendix.
 
@@
 In \citet{li2024plot}, the residual plots are simulated from a data
-generating process which is a special case of Equation
-\ref{eq:data-sim}. The main characteristic is the model violations are
-introduced separately, meaning non-linearity and heteroskedasticity will
-not co-exist in one lineup but assigned uniformly to all lineups.
-Additionally, non-normality and multiple predictors are not considered
-in the experimental design.
+generating process that corresponds to a special case of the synthetic
+data model used in this study. A key feature of the design is that model
+violations are introduced independently, meaning non-linearity and
+heteroskedasticity do not co-exist within a single lineup but are
+instead assigned uniformly across different lineups. Moreover, the
+experimental design does not account for non-normality or multiple
+predictors.
 

### 2025-06-13 paper.Rmd
Put auxiliary information into model arch

@@
 
-## Auxiliary Information with Scagnostics
-
-In Section \ref{sec-model-introduction}, we mention that scagnostics consist of a set of manually designed visual feature extraction functions.
-While our computer vision model will learn its own feature extraction function during training, leveraging additional information from scagnostics can enhance the model's predictive accuracy.
-
-For each residual plot used as an input image, we calculated four scagnostics — "Monotonic", "Sparse", "Splines", and "Striped" – using the `cassowaryr` R package [@mason2022cassowaryr].
-These computed measures, along with the number of observations from the fitted model, were provided as the second input for the computer vision model.
-We selected these scagnostics due to their reliability and efficiency, as other scagnostics occasionally caused R process crashes (approximately 5% of the time) during training data preparation, due to a bug in the `interp` R package [@Albrecht2023interp].
-Although the package maintainer later fixed this bug at our request, the fix came too late to retrain the model, and additionally, their high computational costs make them unsuitable for rapid inference, which was a critical factor in our choice.
-
 # Distance from a Theoretically "Good" Residual Plot  {#sec-model-distance-between-residual-plots}
@@
 Figure \ref{fig:cnn-diag} provides a diagram of the architecture.
-More details about the neural network layers used in this study are provided in Appendix A.
+More details about the neural network layers used in this study are provided in the Appendix.
 
@@
 
-The output of the last convolutional block is summarized by either a global max pooling layer or a global average pooling layer, resulting in a two-dimensional tensor.
-To leverage the information contained in scagnostics, this tensor is concatenated with an additional $n \times 5$ tensor, which contains the "Monotonic", "Sparse", "Splines", and "Striped" measures, along with the number of observations for $n$ residual plots.
+The output of the last convolutional block is summarized by either a global max-pooling layer or a global average-pooling layer, resulting in a two-dimensional tensor.
+To enrich predictions with additional information beyond visual features, this tensor is concatenated with an additional $n \times 5$ tensor, which contains the "Monotonic", "Sparse", "Splines", and "Striped" measures computed using the `cassowaryr` R package [@mason2022cassowaryr] , along with the number of observations for $n$ residual plots.
+These measures were selected for their reliability and efficiency, as other scagnostics occasionally caused R process crashes ($\sim 5\%$) during data preparation due to a bug in the `interp` package [@Albrecht2023interp].
+Although this bug was later fixed at our request, the fix came too late to re-train the model.
+Moreover, their high computational cost makes them unsuitable for fast inference.
 
@@
 The first layer contains at least $128$ units, followed by a dropout layer.
-Occasionally, a batch normalization layer is inserted between the fully-connected layer and the dropout layer for regularization purposes.
+A batch normalization layer is inserted between the fully-connected layer and the dropout layer for regularization purposes.
 The second fully-connected layer consists of only one unit, serving as the output of the model.
 
-The model weights $\boldsymbol{\theta}$ were randomly initialized and they were optimized by the Adam optimizer [@kingma2014adam] with the mean square error loss function
+The model weights $\boldsymbol{\theta}$ were randomly initialized using the Glorot Uniform method [@glorot2010understanding] and they were optimized by the Adam optimizer [@kingma2014adam] with the mean square error loss function
 
@@
 
-\noindent where $n_{\text{train}}$ is the number of training samples, $V_i$ is the $i$-th residual plot and $S_i$ is the additional information about the $i$-th residual plot including four scagnostics and the number of observations.
+\noindent where $n_{\text{train}}$ is the number of training samples, $V_i$ is the $i$-th residual plot and $S_i$ is the auxiliary information about the $i$-th residual plot including four scagnostics and the number of observations.
 
@@
 Hyperparameters were optimized via Bayesian tuning with KerasTuner [@omalley2019kerastuner], minimizing validation RMSE across 100 trials.
-The tuning process included early stopping and considered dropout, batch normalization, input resolution, and auxiliary inputs.
+The tuning process included early stopping and considered dropout rate, batch normalization, input resolution, and auxiliary inputs.
 

### 2025-06-13 paper.tex
Put auxiliary information into model arch

@@
 
-\subsection{Auxiliary Information with
-Scagnostics}\label{auxiliary-information-with-scagnostics}
-
-In Section \ref{sec-model-introduction}, we mention that scagnostics
-consist of a set of manually designed visual feature extraction
-functions. While our computer vision model will learn its own feature
-extraction function during training, leveraging additional information
-from scagnostics can enhance the model's predictive accuracy.
-
-For each residual plot used as an input image, we calculated four
-scagnostics --- ``Monotonic'', ``Sparse'', ``Splines'', and ``Striped''
--- using the \texttt{cassowaryr} R package \citep{mason2022cassowaryr}.
-These computed measures, along with the number of observations from the
-fitted model, were provided as the second input for the computer vision
-model. We selected these scagnostics due to their reliability and
-efficiency, as other scagnostics occasionally caused R process crashes
-(approximately 5\% of the time) during training data preparation, due to
-a bug in the \texttt{interp} R package \citep{Albrecht2023interp}.
-Although the package maintainer later fixed this bug at our request, the
-fix came too late to retrain the model, and additionally, their high
-computational costs make them unsuitable for rapid inference, which was
-a critical factor in our choice.
-
 \section{Distance from a Theoretically ``Good'' Residual
@@
 \ref{fig:cnn-diag} provides a diagram of the architecture. More details
-about the neural network layers used in this study are provided in
-Appendix A.
+about the neural network layers used in this study are provided in the
+Appendix.
 
@@
 The output of the last convolutional block is summarized by either a
-global max pooling layer or a global average pooling layer, resulting in
-a two-dimensional tensor. To leverage the information contained in
-scagnostics, this tensor is concatenated with an additional
-\(n \times 5\) tensor, which contains the ``Monotonic'', ``Sparse'',
-``Splines'', and ``Striped'' measures, along with the number of
-observations for \(n\) residual plots.
+global max-pooling layer or a global average-pooling layer, resulting in
+a two-dimensional tensor. To enrich predictions with additional
+information beyond visual features, this tensor is concatenated with an
+additional \(n \times 5\) tensor, which contains the ``Monotonic'',
+``Sparse'', ``Splines'', and ``Striped'' measures computed using the
+\texttt{cassowaryr} R package \citep{mason2022cassowaryr} , along with
+the number of observations for \(n\) residual plots. These measures were
+selected for their reliability and efficiency, as other scagnostics
+occasionally caused R process crashes (\(\sim 5\%\)) during data
+preparation due to a bug in the \texttt{interp} package
+\citep{Albrecht2023interp}. Although this bug was later fixed at our
+request, the fix came too late to re-train the model. Moreover, their
+high computational cost makes them unsuitable for fast inference.
 
@@
 This block consists of two fully-connected layers. The first layer
-contains at least \(128\) units, followed by a dropout layer.
-Occasionally, a batch normalization layer is inserted between the
-fully-connected layer and the dropout layer for regularization purposes.
-The second fully-connected layer consists of only one unit, serving as
-the output of the model.
-
-The model weights \(\boldsymbol{\theta}\) were randomly initialized and
-they were optimized by the Adam optimizer \citep{kingma2014adam} with
-the mean square error loss function
+contains at least \(128\) units, followed by a dropout layer. A batch
+normalization layer is inserted between the fully-connected layer and
+the dropout layer for regularization purposes. The second
+fully-connected layer consists of only one unit, serving as the output
+of the model.
+
+The model weights \(\boldsymbol{\theta}\) were randomly initialized
+using the Glorot Uniform method \citep{glorot2010understanding} and they
+were optimized by the Adam optimizer \citep{kingma2014adam} with the
+mean square error loss function
 
@@
 \noindent where \(n_{\text{train}}\) is the number of training samples,
-\(V_i\) is the \(i\)-th residual plot and \(S_i\) is the additional
+\(V_i\) is the \(i\)-th residual plot and \(S_i\) is the auxiliary
 information about the \(i\)-th residual plot including four scagnostics
@@
 trials. The tuning process included early stopping and considered
-dropout, batch normalization, input resolution, and auxiliary inputs.
+dropout rate, batch normalization, input resolution, and auxiliary
+inputs.
 

### 2025-06-13 paper.Rmd
Reduce the Model Specification section

@@
 
-Deep learning models are in general very sensitive to the input data.
-The quality and relevance of the input data greatly influence the model's capacity to generate insightful and meaningful results.
-There are several designs of the input format that can be considered.
-
-A straightforward architecture of the input layer involves feeding a vector of residuals along with a vector of fitted values, essentially providing all the necessary information for creating a residuals vs fitted values plot.
-However, a drawback of this method is the variable input size, which changes based on the number of observations.
-For modern computer vision models implemented in mainstream software like TensorFlow [@abadi2016tensorflow], the input shape is typically fixed.
-One strategy is to pad the input vectors with leading or trailing zeros when the input tensor expects longer vectors, but it may fail if the input vector surpasses the designed length.
-Alternatively, sampling a fixed number of residual-fitted value pairs can ensure consistent input size.
-By carefully designing the sampling process, it is possible to maintain the input's representativeness while conforming to a fixed format.
-
-Another architecture of the input layer involves using an image as input.
-The primary advantage of this design, as opposed to the vector format, is the availability of the existing and sophisticated image processing architectures developed over the years, such as the VGG16 architecture proposed in @simonyan2014very.
-These architectures can effectively capture and summarize spatial information from nearby pixels, which is less straightforward with vector input.
-Encoding data points as pixels introduces certain trade-offs.
-While some detail may be lost during the discretization process, this approach aligns with the sampling-based strategy by focusing on summarizing the overall data distribution.
-Unlike vectors, which struggle to represent spatial relationships, image-based inputs naturally encode the joint distribution of residuals and fitted values in a visually interpretable format.
-This richer representation gives the model a more comprehensive and intuitive understanding of the residual plot's underlying structure.
-Key considerations for using image-based input include image resolution and the aesthetics of the residual plot.
-In general, a higher resolution provides more information to the model but comes at the cost of of increased complexity and greater difficulty in training.
-As for the aesthetics of the residual plot, a practical solution is to consistently present residual plots in the same style to the model.
-This implies that the model can not accept arbitrary images as input but requires the use of the same pre-processing pipeline to convert residuals and fitted values into a standardized-style residual plot.
-
-Providing multiple residual plots to the model, such as a pair of plots, a triplet or a lineup is also a possible option.
-@chopra2005learning have shown that computer vision models designed for image comparison can assess whether a pair of images are similar or dissimilar.
-Applied to our specific problem, we can define null plots of a fitted regression model to be similar to each other, while considering true residual plots to be distinct from null plots of any fitted regression model.
-A triplet constitutes a set of three images, denoted as $image_1$, $image_2$ and $image_3$.
-It is often used to predict whether $image_2$ or $image_3$ is more similar to $image_1$, proving particularly useful for establishing rankings between samples.
-For this setup, we can apply the same criteria to define similarity between images.
-However, it is important to note that these two approaches usually require additional considerations regarding the loss function and, at times, non-standard training processes due to shared weights between different convolutional blocks.
-
-Presenting a lineup to a model aligns closely with the lineup protocol.
-However, as the number of residual plots in a lineup increases, the resolution of the input image grows rapidly, posing challenges in training the model.
-We experimented with this approach in a pilot study, but the performance of the trained model was sub-optimal.
-
-Taking into account the implementation cost and the need for model interpretability, we used the single residual plot input format in this paper.
+Deep learning models are sensitive to input design, and several alternatives are available for encoding residual plots.
+
+A simple approach involves feeding vectors of residuals and fitted values into the model.
+While this format contains all relevant information, the input length varies with sample size, which conflicts with the fixed-shape requirements of modern vision models such as those implemented in TensorFlow [@abadi2016tensorflow].
+Padding can be used to enforce uniform length but risks truncation if inputs exceed the fixed size.
+Alternatively, fixed-size sampling of residual–fitted value pairs can ensure consistency, though at the cost of some information loss.
+
+Another strategy is to convert residual plots into images.
+This enables the use of standard convolutional architectures such as VGG16 [@simonyan2014very], which can effectively capture spatial patterns.
+Although discretization introduces some losses in detail, image-based inputs offer a richer representation by preserving the joint distribution of residuals and fitted values in a visually interpretable format.
+To ensure model generalization, it is essential to use residual plots in a consistent style, i.e., same aesthetics, layout, scaling and color schemes.
+
+Using multiple residual plots as input, such as pairs, triplets, or full lineups is also a possible option.
+For instance, triplet networks [@chopra2005learning] can assess visual similarity if we assume null plots are more similar from each other than from the true residual plot, which may help distinguish null plots from true residual plots.
+However, these architectures introduce complexity in training due to weight sharing and specialized loss functions.
+We experimented with this setting in a pilot study, and found multiple residual plots led to high input resolution, high computational cost and suboptimal model performance.
+
+Balancing accuracy, interpretability, and implementation cost, we adopt a single residual plot in image format as the model input.
 
@@
 
-Given that the input is a single residual plot represented as a fixed-resolution image, the output of the computer vision model can be designed for various tasks: binary classification, categorical classification with multiple classes, or numeric regression.
-
-The binary outcome can represent whether the input image is consistent with a null plot as determined by either (1) the data generating process or (2) the result of a visual test based on human judgement.
-Training a model following the latter option requires data from prior human subject experiments, presenting difficulties in controlling the quality of data due to variations in experimental settings across different studies.
-Additionally, some visual inference experiments are unrelated to linear regression models or residual plot diagnostics, resulting in a limited amount of available training data.
-
-For a multi-class outcome, the model could classify images into categories such as "contains outliers" vs.
+Given the input is a single fixed-resolution image, the model output can be defined for various tasks, including binary classification, multiclass classification, and numeric regression.
+
+In the binary case, the outcome may indicate whether the input image is consistent with a null plot, as determined either by (1) the data-generating process or (2) the outcome of a visual test based on human judgment.
+The first approach models the probability that a residual plot is not a null plot and is a viable option.
+In contrast, the second relies on experimental data that are often scarce or inconsistent across studies.
+
+Alternatively, a multiclass outcome may classify images into categories such as "contains outliers" vs.
 "does not contain outliers" or "is non-linear" vs.
@@
 
-One significant advantage of the lineup protocol for residual diagnostics is its ability to detect multiple model violations simultaneously, provided these violations produce visually distinguishable differences in the plots.
-In contrast, training a classification model would require explicitly defining and mapping all potential violations of interest, or at least the most common ones, which can be impractical.
-If the goal is to identify specific cases, conventional statistical tests grounded in regression model assumptions are often more effective.
-Computer vision models may be less suited to such tasks.
-
-Alternatively, the output could be a meaningful and interpretable numerical measure useful for assessing residual plots, such as the strength of suspicious visual patterns reflecting the extent of model violations, or the difficulty index for identifying whether a residual plot has no issues.
-However, these numeric measures are often informally used in daily communication but are not typically formalized or rigorously defined.
-For the purpose of training a model, this numeric measure has to be quantifiable.
-
-In this study, we chose to define and use a distance between a true residual plot and a theoretically "good" residual plot.
-This is further explained in \ref{sec-model-distance-estimation}.
-@vo2016localizing have also demonstrated that defining a proper distance between images can enhance the matching accuracy in image search compared to a binary outcome model.
+A third option is to produce a meaninful and interpretable numerical measure that quantifies the strength of suspicious visual patterns reflecting the extent of model violations, or the difficulty index for identifying whether a residual plot has no issues.
+However, such measures are often used informally in daily communications but rarely formalized.
+For supervised learning, they must be clearly defined and reliably estimable.
+
+In this study, we defined and used a continuous distance between a true residual plot and a theoretically "good" residual plot.
+This approach captures the degree to which a plot deviates from model assumptions.
+In contrast, a probability output based on binary labels, where true residual plots are labeled as 1 and null plots as 0, cannot express the severity of violation.
+It forces the model to infer this information solely from the data distribution, which may be inefficient and discards useful prior knowledge.
 

### 2025-06-13 paper.tex
Reduce the Model Specification section

@@
 
-Deep learning models are in general very sensitive to the input data.
-The quality and relevance of the input data greatly influence the
-model's capacity to generate insightful and meaningful results. There
-are several designs of the input format that can be considered.
-
-A straightforward architecture of the input layer involves feeding a
-vector of residuals along with a vector of fitted values, essentially
-providing all the necessary information for creating a residuals vs
-fitted values plot. However, a drawback of this method is the variable
-input size, which changes based on the number of observations. For
-modern computer vision models implemented in mainstream software like
-TensorFlow \citep{abadi2016tensorflow}, the input shape is typically
-fixed. One strategy is to pad the input vectors with leading or trailing
-zeros when the input tensor expects longer vectors, but it may fail if
-the input vector surpasses the designed length. Alternatively, sampling
-a fixed number of residual-fitted value pairs can ensure consistent
-input size. By carefully designing the sampling process, it is possible
-to maintain the input's representativeness while conforming to a fixed
-format.
-
-Another architecture of the input layer involves using an image as
-input. The primary advantage of this design, as opposed to the vector
-format, is the availability of the existing and sophisticated image
-processing architectures developed over the years, such as the VGG16
-architecture proposed in \citet{simonyan2014very}. These architectures
-can effectively capture and summarize spatial information from nearby
-pixels, which is less straightforward with vector input. Encoding data
-points as pixels introduces certain trade-offs. While some detail may be
-lost during the discretization process, this approach aligns with the
-sampling-based strategy by focusing on summarizing the overall data
-distribution. Unlike vectors, which struggle to represent spatial
-relationships, image-based inputs naturally encode the joint
+Deep learning models are sensitive to input design, and several
+alternatives are available for encoding residual plots.
+
+A simple approach involves feeding vectors of residuals and fitted
+values into the model. While this format contains all relevant
+information, the input length varies with sample size, which conflicts
+with the fixed-shape requirements of modern vision models such as those
+implemented in TensorFlow \citep{abadi2016tensorflow}. Padding can be
+used to enforce uniform length but risks truncation if inputs exceed the
+fixed size. Alternatively, fixed-size sampling of residual--fitted value
+pairs can ensure consistency, though at the cost of some information
+loss.
+
+Another strategy is to convert residual plots into images. This enables
+the use of standard convolutional architectures such as VGG16
+\citep{simonyan2014very}, which can effectively capture spatial
+patterns. Although discretization introduces some losses in detail,
+image-based inputs offer a richer representation by preserving the joint
 distribution of residuals and fitted values in a visually interpretable
-format. This richer representation gives the model a more comprehensive
-and intuitive understanding of the residual plot's underlying structure.
-Key considerations for using image-based input include image resolution
-and the aesthetics of the residual plot. In general, a higher resolution
-provides more information to the model but comes at the cost of of
-increased complexity and greater difficulty in training. As for the
-aesthetics of the residual plot, a practical solution is to consistently
-present residual plots in the same style to the model. This implies that
-the model can not accept arbitrary images as input but requires the use
-of the same pre-processing pipeline to convert residuals and fitted
-values into a standardized-style residual plot.
-
-Providing multiple residual plots to the model, such as a pair of plots,
-a triplet or a lineup is also a possible option.
-\citet{chopra2005learning} have shown that computer vision models
-designed for image comparison can assess whether a pair of images are
-similar or dissimilar. Applied to our specific problem, we can define
-null plots of a fitted regression model to be similar to each other,
-while considering true residual plots to be distinct from null plots of
-any fitted regression model. A triplet constitutes a set of three
-images, denoted as \(image_1\), \(image_2\) and \(image_3\). It is often
-used to predict whether \(image_2\) or \(image_3\) is more similar to
-\(image_1\), proving particularly useful for establishing rankings
-between samples. For this setup, we can apply the same criteria to
-define similarity between images. However, it is important to note that
-these two approaches usually require additional considerations regarding
-the loss function and, at times, non-standard training processes due to
-shared weights between different convolutional blocks.
-
-Presenting a lineup to a model aligns closely with the lineup protocol.
-However, as the number of residual plots in a lineup increases, the
-resolution of the input image grows rapidly, posing challenges in
-training the model. We experimented with this approach in a pilot study,
-but the performance of the trained model was sub-optimal.
-
-Taking into account the implementation cost and the need for model
-interpretability, we used the single residual plot input format in this
-paper.
+format. To ensure model generalization, it is essential to use residual
+plots in a consistent style, i.e., same aesthetics, layout, scaling and
+color schemes.
+
+Using multiple residual plots as input, such as pairs, triplets, or full
+lineups is also a possible option. For instance, triplet networks
+\citep{chopra2005learning} can assess visual similarity if we assume
+null plots are more similar from each other than from the true residual
+plot, which may help distinguish null plots from true residual plots.
+However, these architectures introduce complexity in training due to
+weight sharing and specialized loss functions. We experimented with this
+setting in a pilot study, and found multiple residual plots led to high
+input resolution, high computational cost and suboptimal model
+performance.
+
+Balancing accuracy, interpretability, and implementation cost, we adopt
+a single residual plot in image format as the model input.
 
@@
 
-Given that the input is a single residual plot represented as a
-fixed-resolution image, the output of the computer vision model can be
-designed for various tasks: binary classification, categorical
-classification with multiple classes, or numeric regression.
-
-The binary outcome can represent whether the input image is consistent
-with a null plot as determined by either (1) the data generating process
-or (2) the result of a visual test based on human judgement. Training a
-model following the latter option requires data from prior human subject
-experiments, presenting difficulties in controlling the quality of data
-due to variations in experimental settings across different studies.
-Additionally, some visual inference experiments are unrelated to linear
-regression models or residual plot diagnostics, resulting in a limited
-amount of available training data.
-
-For a multi-class outcome, the model could classify images into
-categories such as ``contains outliers'' vs.~``does not contain
-outliers'' or ``is non-linear'' vs.~``not non-linear,'' among other
-diagnostic features. While this approach is appealing in theory, it
-essentially reverts to the traditional testing framework, where tests
-focus on detecting specific types of model violations in isolation.
-
-One significant advantage of the lineup protocol for residual
-diagnostics is its ability to detect multiple model violations
-simultaneously, provided these violations produce visually
-distinguishable differences in the plots. In contrast, training a
-classification model would require explicitly defining and mapping all
-potential violations of interest, or at least the most common ones,
-which can be impractical. If the goal is to identify specific cases,
-conventional statistical tests grounded in regression model assumptions
-are often more effective. Computer vision models may be less suited to
-such tasks.
-
-Alternatively, the output could be a meaningful and interpretable
-numerical measure useful for assessing residual plots, such as the
-strength of suspicious visual patterns reflecting the extent of model
-violations, or the difficulty index for identifying whether a residual
-plot has no issues. However, these numeric measures are often informally
-used in daily communication but are not typically formalized or
-rigorously defined. For the purpose of training a model, this numeric
-measure has to be quantifiable.
-
-In this study, we chose to define and use a distance between a true
-residual plot and a theoretically ``good'' residual plot. This is
-further explained in \ref{sec-model-distance-estimation}.
-\citet{vo2016localizing} have also demonstrated that defining a proper
-distance between images can enhance the matching accuracy in image
-search compared to a binary outcome model.
+Given the input is a single fixed-resolution image, the model output can
+be defined for various tasks, including binary classification,
+multiclass classification, or numeric regression.
+
+In the binary case, the outcome may indicate whether the input image is
+consistent with a null plot, as determined either by (1) the
+data-generating process or (2) the outcome of a visual test based on
+human judgment. The first approach models the probability that a
+residual plot is not a null plot and is a viable option. In contrast,
+the second relies on experimental data that are often scarce or
+inconsistent across studies.
+
+Alternatively, a multiclass outcome may classify images into categories
+such as ``contains outliers'' vs.~``does not contain outliers'' or ``is
+non-linear'' vs.~``not non-linear,'' among other diagnostic features.
+While this approach is appealing in theory, it essentially reverts to
+the traditional testing framework, where tests focus on detecting
+specific types of model violations in isolation.
+
+A third option is to produce a meaninful and interpretable numerical
+measure that quantifies the strength of suspicious visual patterns
+reflecting the extent of model violations, or the difficulty index for
+identifying whether a residual plot has no issues. However, such
+measures are often used informally in daily communications but rarely
+formalized. For supervised learning, they must be clearly defined and
+reliably estimable.
+
+In this study, we defined and used a continuous distance between a true
+residual plot and a theoretically ``good'' residual plot. This approach
+captures the degree to which a plot deviates from model assumptions. In
+contrast, a probability output based on binary labels, where true
+residual plots are labeled as 1 and null plots as 0, cannot express the
+severity of violation. It forces the model to infer this information
+solely from the data distribution, which may be inefficient and discards
+useful prior knowledge.
 

### 2025-06-13 paper.Rmd
Reduce Distance Estimation section

@@
 
-In the previous sections, we have defined a distance measure given in Equation \ref{eq:kl-0} for quantifying the difference between the true residual distribution $P$ and an ideal reference distribution $Q$.
+We previously defined a distance measure (Equation \ref{eq:kl-0}) to quantify the difference between the true residual distribution $P$ and an ideal reference distribution $Q$.
 However, this distance measure can only be computed when the data generating process is known.
@@
 
-We use a computer vision model to estimate this distance measure for a residual plot. Let $D$ be the result of Equation \ref{eq:kl-0}, and our estimator $\hat{D}$ is formulated as
+To approximate this distance from a residual plot, we proposed a computer vision estimator $\hat{D}$ formulated as
 
@@
 
-\noindent where $V_{h \times w}(.)$ is a plotting function that saves a residuals vs fitted values plot with fixed aesthetic as an image with $h \times w$ pixels in three channels (RGB), $f_{CV}(.)$ is a computer vision model which takes an $h \times w$ image as input and predicts the distance in the domain $[0, +\infty)$.
-
-With the estimated distance $\hat{D}$, we can compare the underlying distribution of the residuals to a theoretically "good" residual distribution.
-$\hat{D}$ can also be used as an index of the model violations indicating the strength of the visual signal embedded in the residual plot.
-
-It is not expected that $\hat{D}$ will be equal to original distance $D$.
-This is largely because information contained in a single residual plot is limited and it may not be able to summarize all the important characteristics of the residual distribution.
-For a given residual distribution $P$, many different residual plots can be simulated, where many will share similar visual patterns, but some of them could be visually very different from the rest, especially for regression models with small $n$.
-This suggests the error of the estimation will vary depends on whether the input residual plot is representative or not.
+\noindent where $V_{h \times w}(.)$ renders the residuals vs fitted values plot as an $h \times w$ RGB image, and $f_{CV}(.)$ is a model that maps the image to an estimated distance $\hat{D} \in [0, +\infty)$.
+
+The distance estimator $\hat{D}$ allows us to assess how closely the residuals resemble an ideal distribution and use as an index of model violation severity.
+However, $\hat{D}$ is not expected to equal the true distance $D$, as a single residual plot may not capture all characteristics of the residual distribution.
+For the same distribution $P$, simulated plots can vary visually, especially with small $n$, leading to estimation error that depends on how representative the input plot is.
 

### 2025-06-13 paper.tex
Reduce Distance Estimation section

@@
 be defined for various tasks, including binary classification,
-multiclass classification, or numeric regression.
+multiclass classification, and numeric regression.
 
@@
 
-In the previous sections, we have defined a distance measure given in
-Equation \ref{eq:kl-0} for quantifying the difference between the true
-residual distribution \(P\) and an ideal reference distribution \(Q\).
-However, this distance measure can only be computed when the data
-generating process is known. In reality, we often have no knowledge
-about the data generating process, otherwise we do not need to do a
-residual diagnostic in the first place.
-
-We use a computer vision model to estimate this distance measure for a
-residual plot. Let \(D\) be the result of Equation \ref{eq:kl-0}, and
-our estimator \(\hat{D}\) is formulated as
+We previously defined a distance measure (Equation \ref{eq:kl-0}) to
+quantify the difference between the true residual distribution \(P\) and
+an ideal reference distribution \(Q\). However, this distance measure
+can only be computed when the data generating process is known. In
+reality, we often have no knowledge about the data generating process,
+otherwise we do not need to do a residual diagnostic in the first place.
+
+To approximate this distance from a residual plot, we proposed a
+computer vision estimator \(\hat{D}\) formulated as
 
@@
 
-\noindent where \(V_{h \times w}(.)\) is a plotting function that saves
-a residuals vs fitted values plot with fixed aesthetic as an image with
-\(h \times w\) pixels in three channels (RGB), \(f_{CV}(.)\) is a
-computer vision model which takes an \(h \times w\) image as input and
-predicts the distance in the domain \([0, +\infty)\).
-
-With the estimated distance \(\hat{D}\), we can compare the underlying
-distribution of the residuals to a theoretically ``good'' residual
-distribution. \(\hat{D}\) can also be used as an index of the model
-violations indicating the strength of the visual signal embedded in the
-residual plot.
-
-It is not expected that \(\hat{D}\) will be equal to original distance
-\(D\). This is largely because information contained in a single
-residual plot is limited and it may not be able to summarize all the
-important characteristics of the residual distribution. For a given
-residual distribution \(P\), many different residual plots can be
-simulated, where many will share similar visual patterns, but some of
-them could be visually very different from the rest, especially for
-regression models with small \(n\). This suggests the error of the
-estimation will vary depends on whether the input residual plot is
-representative or not.
+\noindent where \(V_{h \times w}(.)\) renders the residuals vs fitted
+values plot as an \(h \times w\) RGB image, and \(f_{CV}(.)\) is a model
+that maps the image to an estimated distance
+\(\hat{D} \in [0, +\infty)\).
+
+The distance estimator \(\hat{D}\) allows us to assess how closely the
+residuals resemble an ideal distribution and use as an index of model
+violation severity. However, \(\hat{D}\) is not expected to equal the
+true distance \(D\), as a single residual plot may not capture all
+characteristics of the residual distribution. For the same distribution
+\(P\), simulated plots can vary visually, especially with small \(n\),
+leading to estimation error that depends on how representative the input
+plot is.
 

### 2025-06-13 paper.Rmd
Reduce Lineup Evaluation section

@@
 
-Theoretically, the distance $D$ for a correctly specified model is $0$, because $P$ will be the same as $Q$.
-However, the computer vision model may not necessary predict $0$ for a null plot.
-Using Figure \ref{fig:false-finding} as an example, it contains a visual pattern which is an indication of heteroskedasticity.
-We would not expect the model to be able to magically tell if the suspicious pattern is caused by the skewed distribution of the fitted values or the existence of heteroskedasticity.
+Theoretically, the distance $D$ for a correctly specified model is $0$, as $P = Q$.
+However, a computer vision model may not predict $\hat{D} = 0$ for a null plot.
+For example, Figure \ref{fig:false-finding} shows a null plot with a pattern suggestive of heteroskedasticity.
+The model can not discern whether such patterns arise from heteroskedasticity or from skewed fitted values.
 Additionally, some null plots could have outliers or strong visual patterns due to randomness, and a reasonable model will try to summarize those information into the prediction, resulting in $\hat{D} > 0$.
-
-This property is not an issue if $\hat{D} \gg 0$ for which the visual signal of the residual plot is very strong, and we usually do not need any further examination of the significance of the result.
-However, if the visual pattern is weak or moderate, having $\hat{D}$ will not be sufficient to determine if $H_0$ should be rejected.
+This is not problematic when $\hat{D}$ is large, as strong visual signals typically justify rejection of $H_0$.
+But when $\hat{D}$ is moderate, it is insufficient alone for decision-making.
 
 To address this issue we can adhere to the paradigm of visual inference, by comparing the estimated distance $\hat{D}$ to the estimated distances for the null plots in a lineup.
-Specifically, if a lineup comprises 20 plots, the null hypothesis $H_0$ will be rejected if $\hat{D}$ exceeds the maximum estimated distance among the $m - 1$ null plots, denoted as $\max\limits_{1 \leq i \leq m-1} {\hat{D}_{null}^{(i)}}$, where $\hat{D}_{null}^{(i)}$ represents the estimated distance for the $i$-th null plot.
-This approach is equivalent to the typical lineup protocol requiring a 95% significance level, where $H_0$ is rejected if the data plot is identified as the most distinct plot by the sole observer.
-The estimated distance serves as a metric to quantify the difference between the data plot and the null plots, as intended.
-
-<!-- For lineups consisting of more than 20 plots, the 95% significance level can be maintained if the number of plots is a multiple of 20.
-Specifically, for lineups comprising $20t$ plots, where $t$ is a positive integer, we reject $H_0$ if $\hat{D}$ exceeds 95% ${\hat{D}_{null}^{(i)}}$ for $i = 1, \ldots, 20t-1$.
-The $p$-value in this case is given by $\frac{1}{20t} + \frac{1}{20t}\sum_{i=1}^{20t-1} I\left(\hat{D}_{null}^{(i)} > \hat{D}\right)$, where $I(\cdot)$ is the indicator function.
--->
+Specifically, if a lineup comprises 20 plots, $H_0$ will be rejected if $\hat{D}$ exceeds the maximum estimated distance among the $m - 1$ null plots, denoted as $\max\limits_{1 \leq i \leq m-1} {\hat{D}_{null}^{(i)}}$, where $\hat{D}_{null}^{(i)}$ represents the estimated distance for the $i$-th null plot.
+This approach is conceptually equivalent to the typical lineup protocol requiring a 95% significance level, where $H_0$ is rejected if the data plot is identified as the most distinct plot by the sole observer.
 
 Moreover, if the number of plots in a lineup, denoted by $m$, is sufficiently large, the empirical distribution of ${\hat{D}_{null}^{(i)}}$ can be viewed as an approximation of the null distribution of the estimated distance.
-Consequently, quantiles of the null distribution can be estimated using the sample quantiles, and these quantiles can be utilized for decision-making purposes.
+Consequently, quantiles of the null distribution can be estimated using the sample quantiles and used for decision-making purposes.
 The details of the sample quantile computation can be found in @hyndman1996sample.
 For instance, if $\hat{D}$ is greater than or equal to the 95% sample quantile, denoted as $Q_{null}(0.95)$, we can conclude that the estimated distance for the true residual plot is significantly different from the estimated distance for null plots with a 95% significance level.
-Based on our experience, to obtain a stable estimate of the 95% quantile, the number of null plots, $n_{null}$, typically needs to be at least 100.
-However, if the null distribution exhibits a long tail, a larger number of null plots may be required.
-Alternatively, a $p$-value is the probability of observing a distance equally or greater than $\hat{D}$ under the null hypothesis $H_0$, and it can be estimated by $\frac{1}{m} + \frac{1}{m}\sum_{i=1}^{m-1}I\left(\hat{D}_{null}^{(i)} \geq \hat{D}\right)$.
-
-To alleviate computation burden, a lattice of quantiles for $\hat{D}$ under $H_0$ with specified sample sizes can be precomputed.
-We can then map the $\hat{D}$ and sample size to the closet quantile and sample size in lattice to calculate the corresponding $p$-value.
-This approach lose precision in $p$-value calculation, however, significantly improves computational efficiency.
+Based on our experience, at least 100 null plots are typically needed for stable estimation, but more may be required if the null distribution has heavy tails.
+Alternatively, a $p$-value is the probability of observing a distance equally or greater than $\hat{D}$ under $H_0$, and it can be estimated as $\frac{1}{m} + \frac{1}{m}\sum_{i=1}^{m-1}I\left(\hat{D}_{null}^{(i)} \geq \hat{D}\right)$.
+
+To reduce computational cost, a precomputed lattice of $\hat{D}$ quantiles under $H_0$ for various sample sizes can be used.
+By matching observed values to the closest entries in this lattice, approximate $p$-values can be obtained efficiently, with minor loss of precision.
 
@@
 
-Bootstrap is often employed in linear regression when conducting inference for estimated parameters [see @davison1997bootstrap and @efron1994introduction].
-It is typically done by sampling individual cases with replacement and refitting the regression model.
-If the observed data accurately reflects the true distribution of the population, the bootstrapped estimates can be used to measure the variability of the parameter estimate without making strong distributional assumptions about the data generating process.
-
-Similarly, bootstrap can be applied on the estimated distance $\hat{D}$.
-For each refitted model $M_{boot}^{(i)}$, there will be an associated residual plot $V_{boot}^{(i)}$ which can be fed into the computer vision model to obtain $\hat{D}_{boot}^{(i)}$, where $i = 1,...,n_{boot}$, and $n_{boot}$ is the number of bootstrapped samples.
-If we are interested in the variation of $\hat{D}$, we can use $\hat{D}_{boot}^{(i)}$ to estimate a confidence interval.
-
-Alternatively, since each $M_{boot}^{(i)}$ has a set of estimated coefficients $\hat{\boldsymbol{\beta}}_{boot}^{(i)}$ and an estimated variance $\hat{\sigma^2}_{boot}^{(i)}$, a new approximated null distribution can be construed and the corresponding 95% sample quantile $Q_{boot}^{(i)}(0.95)$ can be computed.
+Bootstrap methods are commonly used in linear regression to estimate parameter variability without strong distributional assumptions [@davison1997bootstrap; @efron1994introduction].
+This involves resampling observations with replacement and refitting the model.
+
+Similarly, we can apply bootstrapping to the estimated distance $\hat{D}$.
+For each bootstrap sample $i = 1, \dots, n_{boot}$, we obtain a refitted model $M^{(i)}_{boot}$, a corresponding residual plot $V^{(i)}_{boot}$, and a predicted distance $\hat{D}^{(i)}_{boot}$.
+If we are interested in the variation of $\hat{D}$, the distribution of $\hat{D}^{(i)}_{boot}$ can be used to construct confidence intervals.
+
+Alternatively, since each $M_{boot}^{(i)}$ has its own residual distribution, a new approximated null distribution can be construed and the corresponding 95% sample quantile $Q_{boot}^{(i)}(0.95)$ can be computed.
 Then, if $\hat{D}_{boot}^{(i)} \geq Q_{boot}^{(i)}(0.95)$, $H_0$ will be rejected for $M_{boot}^{(i)}$.
-The ratio of rejected $M_{boot}^{(i)}$ among all the refitted models provides an indication of how often the assumed regression model are considered to be incorrect if the data can be obtained repetitively from the same data generating process.
-But this approach is computationally very expensive because it requires $n_{boot} \times n_{null}$ times of residual plot assessment.
+The ratio of rejected $M_{boot}^{(i)}$ among all the refitted models reflects how often the assumed regression model are considered to be misspecified if data were repeatedly drawn from the same process.
+But this approach is computationally very expensive because it requires $n_{boot} \times (n_{null} + 1)$ times of residual plot assessment.
 In practice, $Q_{null}(0.95)$ can be used to replace $Q_{boot}^{(i)}(0.95)$ in the computation.

### 2025-06-13 paper.tex
Reduce Lineup Evaluation section

@@
 Theoretically, the distance \(D\) for a correctly specified model is
-\(0\), because \(P\) will be the same as \(Q\). However, the computer
-vision model may not necessary predict \(0\) for a null plot. Using
-Figure \ref{fig:false-finding} as an example, it contains a visual
-pattern which is an indication of heteroskedasticity. We would not
-expect the model to be able to magically tell if the suspicious pattern
-is caused by the skewed distribution of the fitted values or the
-existence of heteroskedasticity. Additionally, some null plots could
-have outliers or strong visual patterns due to randomness, and a
-reasonable model will try to summarize those information into the
-prediction, resulting in \(\hat{D} > 0\).
-
-This property is not an issue if \(\hat{D} \gg 0\) for which the visual
-signal of the residual plot is very strong, and we usually do not need
-any further examination of the significance of the result. However, if
-the visual pattern is weak or moderate, having \(\hat{D}\) will not be
-sufficient to determine if \(H_0\) should be rejected.
+\(0\), as \(P = Q\). However, a computer vision model may not predict
+\(\hat{D} = 0\) for a null plot. For example, Figure
+\ref{fig:false-finding} shows a null plot with a pattern suggestive of
+heteroskedasticity. The model can not discern whether such patterns
+arise from heteroskedasticity or from skewed fitted values.
+Additionally, some null plots could have outliers or strong visual
+patterns due to randomness, and a reasonable model will try to summarize
+those information into the prediction, resulting in \(\hat{D} > 0\).
+This is not problematic when \(\hat{D}\) is large, as strong visual
+signals typically justify rejection of \(H_0\). But when \(\hat{D}\) is
+moderate, it is insufficient alone for decision-making.
 
@@
 distances for the null plots in a lineup. Specifically, if a lineup
-comprises 20 plots, the null hypothesis \(H_0\) will be rejected if
-\(\hat{D}\) exceeds the maximum estimated distance among the \(m - 1\)
-null plots, denoted as
+comprises 20 plots, \(H_0\) will be rejected if \(\hat{D}\) exceeds the
+maximum estimated distance among the \(m - 1\) null plots, denoted as
 \(\max\limits_{1 \leq i \leq m-1} {\hat{D}_{null}^{(i)}}\), where
 \(\hat{D}_{null}^{(i)}\) represents the estimated distance for the
-\(i\)-th null plot. This approach is equivalent to the typical lineup
-protocol requiring a 95\% significance level, where \(H_0\) is rejected
-if the data plot is identified as the most distinct plot by the sole
-observer. The estimated distance serves as a metric to quantify the
-difference between the data plot and the null plots, as intended.
+\(i\)-th null plot. This approach is conceptually equivalent to the
+typical lineup protocol requiring a 95\% significance level, where
+\(H_0\) is rejected if the data plot is identified as the most distinct
+plot by the sole observer.
 
@@
 distribution of the estimated distance. Consequently, quantiles of the
-null distribution can be estimated using the sample quantiles, and these
-quantiles can be utilized for decision-making purposes. The details of
-the sample quantile computation can be found in
-\citet{hyndman1996sample}. For instance, if \(\hat{D}\) is greater than
-or equal to the 95\% sample quantile, denoted as \(Q_{null}(0.95)\), we
-can conclude that the estimated distance for the true residual plot is
-significantly different from the estimated distance for null plots with
-a 95\% significance level. Based on our experience, to obtain a stable
-estimate of the 95\% quantile, the number of null plots, \(n_{null}\),
-typically needs to be at least 100. However, if the null distribution
-exhibits a long tail, a larger number of null plots may be required.
-Alternatively, a \(p\)-value is the probability of observing a distance
-equally or greater than \(\hat{D}\) under the null hypothesis \(H_0\),
-and it can be estimated by
+null distribution can be estimated using the sample quantiles and used
+for decision-making purposes. The details of the sample quantile
+computation can be found in \citet{hyndman1996sample}. For instance, if
+\(\hat{D}\) is greater than or equal to the 95\% sample quantile,
+denoted as \(Q_{null}(0.95)\), we can conclude that the estimated
+distance for the true residual plot is significantly different from the
+estimated distance for null plots with a 95\% significance level. Based
+on our experience, at least 100 null plots are typically needed for
+stable estimation, but more may be required if the null distribution has
+heavy tails. Alternatively, a \(p\)-value is the probability of
+observing a distance equally or greater than \(\hat{D}\) under \(H_0\),
+and it can be estimated as
 \(\frac{1}{m} + \frac{1}{m}\sum_{i=1}^{m-1}I\left(\hat{D}_{null}^{(i)} \geq \hat{D}\right)\).
 
-To alleviate computation burden, a lattice of quantiles for \(\hat{D}\)
-under \(H_0\) with specified sample sizes can be precomputed. We can
-then map the \(\hat{D}\) and sample size to the closet quantile and
-sample size in lattice to calculate the corresponding \(p\)-value. This
-approach lose precision in \(p\)-value calculation, however,
-significantly improves computational efficiency.
+To reduce computational cost, a precomputed lattice of \(\hat{D}\)
+quantiles under \(H_0\) for various sample sizes can be used. By
+matching observed values to the closest entries in this lattice,
+approximate \(p\)-values can be obtained efficiently, with minor loss of
+precision.
 
@@
 
-Bootstrap is often employed in linear regression when conducting
-inference for estimated parameters \citep[see][ and
-\citet{efron1994introduction}]{davison1997bootstrap}. It is typically
-done by sampling individual cases with replacement and refitting the
-regression model. If the observed data accurately reflects the true
-distribution of the population, the bootstrapped estimates can be used
-to measure the variability of the parameter estimate without making
-strong distributional assumptions about the data generating process.
-
-Similarly, bootstrap can be applied on the estimated distance
-\(\hat{D}\). For each refitted model \(M_{boot}^{(i)}\), there will be
-an associated residual plot \(V_{boot}^{(i)}\) which can be fed into the
-computer vision model to obtain \(\hat{D}_{boot}^{(i)}\), where
-\(i = 1,...,n_{boot}\), and \(n_{boot}\) is the number of bootstrapped
-samples. If we are interested in the variation of \(\hat{D}\), we can
-use \(\hat{D}_{boot}^{(i)}\) to estimate a confidence interval.
-
-Alternatively, since each \(M_{boot}^{(i)}\) has a set of estimated
-coefficients \(\hat{\boldsymbol{\beta}}_{boot}^{(i)}\) and an estimated
-variance \(\hat{\sigma^2}_{boot}^{(i)}\), a new approximated null
-distribution can be construed and the corresponding 95\% sample quantile
-\(Q_{boot}^{(i)}(0.95)\) can be computed. Then, if
-\(\hat{D}_{boot}^{(i)} \geq Q_{boot}^{(i)}(0.95)\), \(H_0\) will be
-rejected for \(M_{boot}^{(i)}\). The ratio of rejected
-\(M_{boot}^{(i)}\) among all the refitted models provides an indication
-of how often the assumed regression model are considered to be incorrect
-if the data can be obtained repetitively from the same data generating
-process. But this approach is computationally very expensive because it
-requires \(n_{boot} \times n_{null}\) times of residual plot assessment.
-In practice, \(Q_{null}(0.95)\) can be used to replace
+Bootstrap methods are commonly used in linear regression to estimate
+parameter variability without strong distributional assumptions
+\citep{davison1997bootstrap, efron1994introduction}. This involves
+resampling observations with replacement and refitting the model.
+
+Similarly, we can apply bootstrapping to the estimated distance
+\(\hat{D}\). For each bootstrap sample \(i = 1, \dots, n_{boot}\), we
+obtain a refitted model \(M^{(i)}_{boot}\), a corresponding residual
+plot \(V^{(i)}_{boot}\), and a predicted distance
+\(\hat{D}^{(i)}_{boot}\). If we are interested in the variation of
+\(\hat{D}\), the distribution of \(\hat{D}^{(i)}_{boot}\) can be used to
+construct confidence intervals.
+
+Alternatively, since each \(M_{boot}^{(i)}\) has its own residual
+distribution, a new approximated null distribution can be construed and
+the corresponding 95\% sample quantile \(Q_{boot}^{(i)}(0.95)\) can be
+computed. Then, if \(\hat{D}_{boot}^{(i)} \geq Q_{boot}^{(i)}(0.95)\),
+\(H_0\) will be rejected for \(M_{boot}^{(i)}\). The ratio of rejected
+\(M_{boot}^{(i)}\) among all the refitted models reflects how often the
+assumed regression model are considered to be misspecified if data were
+repeatedly drawn from the same process. But this approach is
+computationally very expensive because it requires
+\(n_{boot} \times (n_{null} + 1)\) times of residual plot assessment. In
+practice, \(Q_{null}(0.95)\) can be used to replace
 \(Q_{boot}^{(i)}(0.95)\) in the computation.

### 2025-06-13 paper.Rmd
Reduce MVI section

@@
 
-To reduce computational cost, a precomputed lattice of $\hat{D}$ quantiles under $H_0$ for various sample sizes can be used.
+To reduce computational cost, a pre-computed lattice of $\hat{D}$ quantiles under $H_0$ for various sample sizes can be used.
 By matching observed values to the closest entries in this lattice, approximate $p$-values can be obtained efficiently, with minor loss of precision.
@@
 
-While statistical testing is a powerful tool for detecting model violations, it can become cumbersome and time-consuming when quick decisions are needed, particularly due to the need to evaluate numerous null plots.
-In practice, a more convenient and immediate method for assessing model performance is often required.
-This is where an index, such as the Model Violations Index (MVI), becomes valuable.
-It offers a straightforward way to quantify deviations from model assumptions, enabling rapid assessment and easier comparison across models.
-
-The estimator $\hat{D}$ measures the difference between the true residual distribution and the reference residual distribution, a difference primarily arises from deviations in model assumptions.
-The magnitude of $D$ directly reflects the degree of these deviations, thus making $\hat{D}$ instrumental in forming a model violations index (MVI).
-
-Note that if more observations are used for estimating the linear regression, the result of Equation \ref{eq:kl-1} will increase, as the integration will be performed with larger $n$.
-For a given data generating process, $D$ typically increases logarithmically with the number of observations.
-This behavior comes from the relationship $D = \text{log}(1 + D_{KL})$, where $D_{KL} = \sum_{i=1}^{n}D_{KL}^{(i)}$ under the assumption of independence.
-
-Since $\hat{D}$ is an estimate of $D$, it is expected that a larger number of observations will also lead to a higher $\hat{D}$.
-However, this does not imply that $\hat{D}$ fails to accurately represent the extent of model violations.
-In fact, when examining residual plots with more observations, we often observe a stronger visual signal strength, as the underlying patterns are more likely to be revealed, except in cases of significant overlapping.
-
-Therefore, the Model Violations Index (MVI) can be proposed as
+In \ref{sec-model-lineup-evaluation}, we noted that a pre-computed lattice of $\hat{D}$ quantiles can reduce the computation cost of lineup tests.
+Another practical approach is to assess model performance directly using the value of $\hat{D}$.
+
+The estimator $\hat{D}$ captures the difference between the true and reference residual distributions, which reflects the extent of model violations, making it instrumental in forming a model violations index (MVI).
+However, when more observations are used in regression, the value of $\hat{D}$ tends to increase logarithmically.
+This is because $D = \log(1 + D_{KL})$, and under the assumption of independence, $D_{KL}$ is the sum of $D_{KL}^{(i)}$ across all observations.
+This does not mean that $\hat{D}$ becomes less reliable.
+In fact, larger samples often make model violations more visible in residual plots, unless strong overlapping masks the patterns.
+
+However, to create a standardized and generalizable index, it is important to adjust for the effect of sample size. Therefore, the Model Violations Index (MVI) is proposed as
 
@@
 
-\noindent where $C$ is a large enough constant keeping the result positive and the term $-\log(n)$ is used to offset the increase in $D$ due to sample size.
+\noindent where $C$ is a sufficiently large constant to ensure the result remains positive, and the $\log(n)$ term offset the increase in $D$ with larger sample sizes.
 
@@
 
-The architecture of the computer vision model is adapted from the well-established VGG16 architecture [@simonyan2014very].
+The architecture of the computer vision model $f_{CV}$ is adapted from the well-established VGG16 architecture [@simonyan2014very].
 While more recent architectures like ResNet [@he2016deep] and DenseNet[@huang2017densely], have achieved even greater performance, VGG16 remains a solid choice for many applications due to its simplicity and effectiveness.

### 2025-06-13 paper.tex
Reduce MVI section

@@
 
-To reduce computational cost, a precomputed lattice of \(\hat{D}\)
+To reduce computational cost, a pre-computed lattice of \(\hat{D}\)
 quantiles under \(H_0\) for various sample sizes can be used. By
@@
 
-While statistical testing is a powerful tool for detecting model
-violations, it can become cumbersome and time-consuming when quick
-decisions are needed, particularly due to the need to evaluate numerous
-null plots. In practice, a more convenient and immediate method for
-assessing model performance is often required. This is where an index,
-such as the Model Violations Index (MVI), becomes valuable. It offers a
-straightforward way to quantify deviations from model assumptions,
-enabling rapid assessment and easier comparison across models.
-
-The estimator \(\hat{D}\) measures the difference between the true
-residual distribution and the reference residual distribution, a
-difference primarily arises from deviations in model assumptions. The
-magnitude of \(D\) directly reflects the degree of these deviations,
-thus making \(\hat{D}\) instrumental in forming a model violations index
-(MVI).
-
-Note that if more observations are used for estimating the linear
-regression, the result of Equation \ref{eq:kl-1} will increase, as the
-integration will be performed with larger \(n\). For a given data
-generating process, \(D\) typically increases logarithmically with the
-number of observations. This behavior comes from the relationship
-\(D = \text{log}(1 + D_{KL})\), where
-\(D_{KL} = \sum_{i=1}^{n}D_{KL}^{(i)}\) under the assumption of
-independence.
-
-Since \(\hat{D}\) is an estimate of \(D\), it is expected that a larger
-number of observations will also lead to a higher \(\hat{D}\). However,
-this does not imply that \(\hat{D}\) fails to accurately represent the
-extent of model violations. In fact, when examining residual plots with
-more observations, we often observe a stronger visual signal strength,
-as the underlying patterns are more likely to be revealed, except in
-cases of significant overlapping.
-
-Therefore, the Model Violations Index (MVI) can be proposed as
+In \ref{sec-model-lineup-evaluation}, we noted that a pre-computed
+lattice of \(\hat{D}\) quantiles can reduce the computation cost of
+lineup tests. Another practical approach is to assess model performance
+directly using the value of \(\hat{D}\).
+
+The estimator \(\hat{D}\) captures the difference between the true and
+reference residual distributions, which reflects the extent of model
+violations, making it instrumental in forming a model violations index
+(MVI). However, when more observations are used in regression, the value
+of \(\hat{D}\) tends to increase logarithmically. This is because
+\(D = \log(1 + D_{KL})\), and under the assumption of independence,
+\(D_{KL}\) is the sum of \(D_{KL}^{(i)}\) across all observations. This
+does not mean that \(\hat{D}\) becomes less reliable. In fact, larger
+samples often make model violations more visible in residual plots,
+unless strong overlapping masks the patterns.
+
+However, to create a standardized and generalizable index, it is
+important to adjust for the effect of sample size. Therefore, the Model
+Violations Index (MVI) is proposed as
 
@@
 
-\noindent where \(C\) is a large enough constant keeping the result
-positive and the term \(-\log(n)\) is used to offset the increase in
-\(D\) due to sample size.
+\noindent where \(C\) is a sufficiently large constant to ensure the
+result remains positive, and the \(\log(n)\) term offset the increase in
+\(D\) with larger sample sizes.
 
@@
 
-The architecture of the computer vision model is adapted from the
-well-established VGG16 architecture \citep{simonyan2014very}. While more
-recent architectures like ResNet \citep{he2016deep} and
+The architecture of the computer vision model \(f_{CV}\) is adapted from
+the well-established VGG16 architecture \citep{simonyan2014very}. While
+more recent architectures like ResNet \citep{he2016deep} and
 DenseNet\citep{huang2017densely}, have achieved even greater

### 2025-06-14 paper.Rmd
Reduce Model Performance section

@@
 
-The test performance for the optimized models with three different input sizes are summarized in Table \ref{tab:performance}.
-Among these models, the $32 \times 32$ model consistently exhibited the best test performance.
-The mean absolute error of the $32 \times 32$ model indicated that the difference between $\hat{D}$ and $D$ was approximately $0.43$ on the test set, a negligible deviation considering the normal range of $D$ typically falls between $0$ and $7$.
-The high $R^2$ values also suggested that the predictions were largely linearly correlated with the target.
-
-Figure \ref{fig:model-performance} presents a hexagonal heatmap for $D - \hat{D}$ versus $D$.
-The brown smoothing curves, fitted by generalized additive models [@hastie2017generalized], demonstrate that all the optimized models perform admirably on the test sets when $1.5 < D < 6$, where no structural issues are noticeable.
-However, over-predictions occurred when $D < 1.5$, while under-predictions occurred predominantly when $\hat{D} > 6$.
-
-For input images representing null plots where $D = 0$, it was expected that the models will over-predict the distance, as explained in Section \ref{sec-model-lineup-evaluation}.
-However, it can not explain the under-prediction issue.
+Table \ref{tab:performance} summarizes the test performance of optimized models using three input sizes.
+The $32 \times 32$ model consistently outperformed others, with a mean absolute error of approximately $0.43$, a negligible deviation given the typical $D$ range of $0$ to $7$.
+High $R^2$ values also indicated strong linear correlation between predictions and targets.
+Based on these results, we used the best-performing $32 \times 32$ model for subsequent analyses.
+
+Figure \ref{fig:model-performance} shows a hexagonal heatmap of $D - \hat{D}$ versus $D$.
+Smoothing curves (fitted by generalized additive models [@hastie2017generalized]) reveal that all models performed well for $1.5 < D < 6$, where no structural issues were observed.
+Over-prediction occurred when $D < 1.5$, and under-prediction primarily when $\hat{D} > 6$.
+
+For null plots ($D = 0$), over-prediction was expected (see Section \ref{sec-model-lineup-evaluation}), but not under-prediction.
 Therefore, we analysed the relationship between residuals and all the factors involved in the data generating process.
-We found that most issues actually arose from non-linearity problems and the presence of a second predictor in the regression model as illustrated in Figure \ref{fig:over-under}.
-When the variance for the error distribution was small, the optimized model tended to under-predict the distance.
-Conversely, when the error distribution had a large variance, the model tended to over-predict the distance.
-
-Since most of the deviation stemmed from the presence of non-linearity violations, to further investigate this, we split the test set based on violation types and re-evaluated the performance, as detailed in Table \ref{tab:performance-sub}.
-It was evident that metrics for null plots were notably worse compared to other categories.
-Furthermore, residual plots solely exhibiting non-normality issues were the easiest to predict, with very low test root mean square error (RMSE) at around $0.3$.
-Residual plots with non-linearity issues were more challenging to assess than those with heteroskedasticity or non-normality issues.
-When multiple violations were introduced to a residual plot, the performance metrics typically lay between the metrics for each individual violation.
-
-Based on the model performance metrics, we chose to use the best-performing model evaluated on the test set, namely the $32 \times 32$ model, for the subsequent analysis.
+We found that most deviations stemmed from non-linearity and the presence of a second predictor in the simulation model, where small $\varepsilon$ variances caused under-predictions and large $\varepsilon$ variances caused over-prediction, as illustrated in Figure \ref{fig:over-under}.
+
+To investigate further, we stratified the test set by violation type and re-evaluated performance (Table \ref{tab:performance-sub}).
+It was found that null plots had the poorest metrics, while non-normality cases were easiest to predict, with RMSE around $0.3$.
+Non-linearity was harder to assess than heteroskedasticity or non-normality.
+Plots with multiple violations showed intermediate performance.
 
@@
                     label = "performance-sub",
-                    caption = "The training and test performance of the $32 \\times 32$ model presented with different model violations.",
+                    caption = "The test performance of the $32 \\times 32$ model presented with different model violations.",
                     escape = FALSE,

### 2025-06-14 paper.tex
Reduce Model Performance section

@@
 
-The test performance for the optimized models with three different input
-sizes are summarized in Table \ref{tab:performance}. Among these models,
-the \(32 \times 32\) model consistently exhibited the best test
-performance. The mean absolute error of the \(32 \times 32\) model
-indicated that the difference between \(\hat{D}\) and \(D\) was
-approximately \(0.43\) on the test set, a negligible deviation
-considering the normal range of \(D\) typically falls between \(0\) and
-\(7\). The high \(R^2\) values also suggested that the predictions were
-largely linearly correlated with the target.
-
-Figure \ref{fig:model-performance} presents a hexagonal heatmap for
-\(D - \hat{D}\) versus \(D\). The brown smoothing curves, fitted by
-generalized additive models \citep{hastie2017generalized}, demonstrate
-that all the optimized models perform admirably on the test sets when
-\(1.5 < D < 6\), where no structural issues are noticeable. However,
-over-predictions occurred when \(D < 1.5\), while under-predictions
-occurred predominantly when \(\hat{D} > 6\).
-
-For input images representing null plots where \(D = 0\), it was
-expected that the models will over-predict the distance, as explained in
-Section \ref{sec-model-lineup-evaluation}. However, it can not explain
-the under-prediction issue. Therefore, we analysed the relationship
-between residuals and all the factors involved in the data generating
-process. We found that most issues actually arose from non-linearity
-problems and the presence of a second predictor in the regression model
-as illustrated in Figure \ref{fig:over-under}. When the variance for the
-error distribution was small, the optimized model tended to
-under-predict the distance. Conversely, when the error distribution had
-a large variance, the model tended to over-predict the distance.
-
-Since most of the deviation stemmed from the presence of non-linearity
-violations, to further investigate this, we split the test set based on
-violation types and re-evaluated the performance, as detailed in Table
-\ref{tab:performance-sub}. It was evident that metrics for null plots
-were notably worse compared to other categories. Furthermore, residual
-plots solely exhibiting non-normality issues were the easiest to
-predict, with very low test root mean square error (RMSE) at around
-\(0.3\). Residual plots with non-linearity issues were more challenging
-to assess than those with heteroskedasticity or non-normality issues.
-When multiple violations were introduced to a residual plot, the
-performance metrics typically lay between the metrics for each
-individual violation.
-
-Based on the model performance metrics, we chose to use the
-best-performing model evaluated on the test set, namely the
-\(32 \times 32\) model, for the subsequent analysis.
+Table \ref{tab:performance} summarizes the test performance of optimized
+models using three input sizes. The \(32 \times 32\) model consistently
+outperformed others, with a mean absolute error of approximately
+\(0.43\), a negligible deviation given the typical \(D\) range of \(0\)
+to \(7\). High \(R^2\) values also indicated strong linear correlation
+between predictions and targets. Based on these results, we used the
+best-performing \(32 \times 32\) model for subsequent analyses.
+
+Figure \ref{fig:model-performance} shows a hexagonal heatmap of
+\(D - \hat{D}\) versus \(D\). Smoothing curves (fitted by generalized
+additive models \citep{hastie2017generalized}) reveal that all models
+performed well for \(1.5 < D < 6\), where no structural issues were
+observed. Over-prediction occurred when \(D < 1.5\), and
+under-prediction primarily when \(\hat{D} > 6\).
+
+For null plots (\(D = 0\)), over-prediction was expected (see Section
+\ref{sec-model-lineup-evaluation}), but not under-prediction. Therefore,
+we analysed the relationship between residuals and all the factors
+involved in the data generating process. We found that most deviations
+stemmed from non-linearity and the presence of a second predictor in the
+simulation model, where small \(\varepsilon\) variances caused
+under-predictions and large \(\varepsilon\) variances caused
+over-prediction, as illustrated in Figure \ref{fig:over-under}.
+
+To investigate further, we stratified the test set by violation type and
+re-evaluated performance (Table \ref{tab:performance-sub}). It was found
+that null plots had the poorest metrics, while non-normality cases were
+easiest to predict, with RMSE around \(0.3\). Non-linearity was harder
+to assess than heteroskedasticity or non-normality. Plots with multiple
+violations showed intermediate performance.
 
@@
 
-\caption{\label{tab:performance-sub}The training and test performance of the $32 \times 32$ model presented with different model violations.}
+\caption{\label{tab:performance-sub}The test performance of the $32 \times 32$ model presented with different model violations.}
 \centering

### 2025-06-16 appendix.Rmd
Make a copy of MVI to appendix

@@
 bibliography: bibliography.bib
-abstract: |
-  Plotting the residuals is a recommended procedure to diagnose deviations from linear model assumptions, such as non-linearity, heteroscedasticity, and non-normality.
-The presence of structure in residual plots can be tested using the lineup protocol to do visual inference.
-There are a variety of conventional residual tests, but the lineup protocol, used as a statistical test, performs better for diagnostic purposes because it is less sensitive and applies more broadly to different types of departures.
-However, the lineup protocol relies on human judgment which limits its scalability.
-This work presents a solution by providing a computer vision model to automate the assessment of residual plots.
-It is trained to predict a distance measure that quantifies the disparity between the residual distribution of a fitted classical normal linear regression model and the reference distribution, based on Kullback-Leibler divergence.
-From extensive simulation studies, the computer vision model exhibits lower sensitivity than conventional tests but higher sensitivity than human visual tests.
-It is slightly less effective on non-linearity patterns.
-Several examples from classical papers and contemporary data illustrate the new procedures, highlighting its usefulness in automating the diagnostic process and supplementing existing methods.
-keywords: |
-  statistical graphics; data visualization; visual inference; computer vision; machine learning; hypothesis testing; reression analysis; cognitive perception; simulation; practical significance
 header-includes: |
@@
 ```
+
+\appendix
 
@@
 ```
+
+# Model Violations Index {#sec-model-violations-index}
+
+In Section 5.1, we noted that a pre-computed lattice of $\hat{D}$ quantiles can reduce the computation cost of lineup tests.
+Another practical approach is to assess model performance directly using the value of $\hat{D}$.
+
+The estimator $\hat{D}$ captures the difference between the true and reference residual distributions, which reflects the extent of model violations, making it instrumental in forming a model violations index (MVI).
+However, when more observations are used in regression, the value of $\hat{D}$ tends to increase logarithmically.
+This is because $D = \log(1 + D_{KL})$, and under the assumption of independence, $D_{KL}$ is the sum of $D_{KL}^{(i)}$ across all observations.
+This does not mean that $\hat{D}$ becomes less reliable.
+In fact, larger samples often make model violations more visible in residual plots, unless strong overlapping masks the patterns.
+
+However, to create a standardized and generalizable index, it is important to adjust for the effect of sample size. Therefore, the Model Violations Index (MVI) is proposed as
+
+\begin{equation} \label{eq:mvi}
+\text{MVI} = C + \hat{D} - \log(n),
+\end{equation}
+
+\noindent where $C$ is a sufficiently large constant to ensure the result remains positive, and the $\log(n)$ term offset the increase in $D$ with larger sample sizes.
+
+Figure \ref{fig:poly-heter-index} displays the residual plots for fitted models exhibiting varying degrees of non-linearity and heteroskedasticity.
+Each residual plot's MVI is computed using Equation \ref{eq:mvi} with $C = 10$.
+When $\text{MVI} > 8$, the visual patterns are notably strong and easily discernible by humans.
+In the range $6 < \text{MVI} < 8$, the visibility of the visual pattern diminishes as MVI decreases.
+Conversely, when $\text{MVI} < 6$, the visual pattern tends to become relatively faint and challenging to observe.
+Table \ref{tab:mvi} provides a summary of the MVI usage and it is applicable to other linear regression models.
+
+```{r}
+data.frame(degree = c("Strong", "Moderate", "Weak"),
+           range = c("$\\text{MVI} > 8$", "$6 < \\text{MVI} < 8$", "$\\text{MVI} < 6$")) %>%
+  kableExtra::kable(format = "latex",
+                    booktabs = TRUE,
+                    label = "mvi",
+                    caption = "Degree of model violations or the strength of the visual signals according to the Model Violations Index (MVI). The constant $C$ is set to be 10.",
+                    escape = FALSE,
+                    col.names = c("Degree of model violations", "Range ($C$ = 10)"),
+                    align = "lc")
+```
+
+```{r}
+set.seed(1495)
+if (!file.exists(here::here("cached_data/poly_index.rds"))) {
+  e_sigma_factor <- 3
+  keras_mod <- autovi::get_keras_model("vss_phn_32")
+  poly_dgp <- phn_model(j = 3, sigma = 0.1^e_sigma_factor)
+  ori_dat <- poly_dgp$gen(300)
+
+  vss <- c()
+  actual_ss <- c()
+  dat_combined <- tibble()
+
+  for (e_sigma in seq(0.7, 1.8, length.out = 20)^e_sigma_factor) {
+    poly_dgp <- phn_model(j = 3, sigma = e_sigma)
+    dat <- poly_dgp$gen(300, computed = list(x1 = ori_dat$x1, e = ori_dat$e, k = ori_dat$k * e_sigma / 0.1^e_sigma_factor))
+    mod <- lm(y ~ x1, data = dat)
+    my_vi <- autovi::auto_vi(fitted_mod = mod, keras_mod = keras_mod)
+    vss <- c(vss, my_vi$vss()$vss)
+    actual_ss <- c(actual_ss, log(poly_dgp$sample_effect_size(dat) + 1))
+
+    dat_combined <- bind_rows(dat_combined, mutate(dat, e_sigma = e_sigma))
+  }
+
+  saveRDS(dat_combined, here::here("cached_data/poly_index.rds"))
+  saveRDS(vss, here::here("cached_data/poly_index_vss.rds"))
+}
+
+dat_combined <- readRDS(here::here("cached_data/poly_index.rds"))
+vss <- readRDS(here::here("cached_data/poly_index_vss.rds"))
+
+p1 <- dat_combined %>%
+  autovi::AUTO_VI$plot_resid() +
+  facet_wrap(~e_sigma, scales = "free", 
+             labeller = function(...){ 
+               tibble(nb = map(glue("'MVI' == {format(vss - log(300) + 10, digits = 3)}"), str2expression))
+               },
+             ncol = 5) +
+  theme(strip.text = element_text(size = 10), plot.title = element_text(size = 10)) +
+  ggtitle("(A) Non-linearity")
+```
+
+```{r}
+set.seed(1495)
+if (!file.exists(here::here("cached_data/heter_index.rds"))) {
+  b_factor <- 2
+  keras_mod <- autovi::get_keras_model("vss_phn_32")
+  heter_dgp <- phn_model(include_z = FALSE, b = 0)
+  ori_dat <- poly_dgp$gen(300)
+
+  vss <- c()
+  actual_ss <- c()
+  dat_combined <- tibble()
+
+  for (b in rev(seq(0.2, 4, length.out = 20))^b_factor) {
+    heter_dgp <- phn_model(include_z = FALSE, b = b)
+    dat <- heter_dgp$gen(300, computed = list(x1 = ori_dat$x1, e = ori_dat$e))
+    mod <- lm(y ~ x1, data = dat)
+    my_vi <- autovi::auto_vi(fitted_mod = mod, keras_mod = keras_mod)
+    vss <- c(vss, my_vi$vss()$vss)
+    actual_ss <- c(actual_ss, log(heter_dgp$sample_effect_size(dat) + 1))
+
+    dat_combined <- bind_rows(dat_combined, mutate(dat, nb = -b))
+  }
+
+  saveRDS(dat_combined, here::here("cached_data/heter_index.rds"))
+  saveRDS(vss, here::here("cached_data/heter_index_vss.rds"))
+}
+
+dat_combined <- readRDS(here::here("cached_data/heter_index.rds"))
+vss <- readRDS(here::here("cached_data/heter_index_vss.rds"))
+
+p2 <- dat_combined %>%
+  autovi::AUTO_VI$plot_resid() +
+  facet_wrap(~nb, scales = "free", 
+             labeller = function(...){ 
+               tibble(nb = map(glue("'MVI' == {format(vss - log(300) + 10, digits = 3)}"), str2expression))
+               },
+             ncol = 5) +
+  theme(strip.text = element_text(size = 10), plot.title = element_text(size = 10)) +
+  ggtitle("(B) Heteroskedasticity")
+```
+
+```{r poly-heter-index, fig.pos = "!h", fig.cap = "Residual plots generated from fitted models exhibiting varying degrees of (A) non-linearity and (B) heteroskedasticity violations.
+The model violations index (MVI) is displayed atop each residual plot.
+The non-linearity patterns are relatively strong for $MVI > 8$, and relatively weak for $MVI < 6$, while the heteroskedasticity patterns are relatively strong for $MVI > 8$, and relatively weak for $MVI < 6$.", fig.height = 10}
+patchwork::wrap_plots(p1, p2, ncol = 1)
+```
+
+\clearpage

### 2025-06-16 appendix.tex
Make a copy of MVI to appendix

@@
 
-\begin{abstract}
-Plotting the residuals is a recommended procedure to diagnose deviations
-from linear model assumptions, such as non-linearity,
-heteroscedasticity, and non-normality. The presence of structure in
-residual plots can be tested using the lineup protocol to do visual
-inference. There are a variety of conventional residual tests, but the
-lineup protocol, used as a statistical test, performs better for
-diagnostic purposes because it is less sensitive and applies more
-broadly to different types of departures. However, the lineup protocol
-relies on human judgment which limits its scalability. This work
-presents a solution by providing a computer vision model to automate the
-assessment of residual plots. It is trained to predict a distance
-measure that quantifies the disparity between the residual distribution
-of a fitted classical normal linear regression model and the reference
-distribution, based on Kullback-Leibler divergence. From extensive
-simulation studies, the computer vision model exhibits lower sensitivity
-than conventional tests but higher sensitivity than human visual tests.
-It is slightly less effective on non-linearity patterns. Several
-examples from classical papers and contemporary data illustrate the new
-procedures, highlighting its usefulness in automating the diagnostic
-process and supplementing existing methods.
-\end{abstract}
-
-\begin{keywords}
-statistical graphics; data visualization; visual inference; computer
-vision; machine learning; hypothesis testing; reression analysis;
-cognitive perception; simulation; practical significance
-\end{keywords}
+\appendix
 
@@
 
+\section{Model Violations Index}\label{sec-model-violations-index}
+
+In Section 5.1, we noted that a pre-computed lattice of \(\hat{D}\)
+quantiles can reduce the computation cost of lineup tests. Another
+practical approach is to assess model performance directly using the
+value of \(\hat{D}\).
+
+The estimator \(\hat{D}\) captures the difference between the true and
+reference residual distributions, which reflects the extent of model
+violations, making it instrumental in forming a model violations index
+(MVI). However, when more observations are used in regression, the value
+of \(\hat{D}\) tends to increase logarithmically. This is because
+\(D = \log(1 + D_{KL})\), and under the assumption of independence,
+\(D_{KL}\) is the sum of \(D_{KL}^{(i)}\) across all observations. This
+does not mean that \(\hat{D}\) becomes less reliable. In fact, larger
+samples often make model violations more visible in residual plots,
+unless strong overlapping masks the patterns.
+
+However, to create a standardized and generalizable index, it is
+important to adjust for the effect of sample size. Therefore, the Model
+Violations Index (MVI) is proposed as
+
+\begin{equation} \label{eq:mvi}
+\text{MVI} = C + \hat{D} - \log(n),
+\end{equation}
+
+\noindent where \(C\) is a sufficiently large constant to ensure the
+result remains positive, and the \(\log(n)\) term offset the increase in
+\(D\) with larger sample sizes.
+
+Figure \ref{fig:poly-heter-index} displays the residual plots for fitted
+models exhibiting varying degrees of non-linearity and
+heteroskedasticity. Each residual plot's MVI is computed using Equation
+\ref{eq:mvi} with \(C = 10\). When \(\text{MVI} > 8\), the visual
+patterns are notably strong and easily discernible by humans. In the
+range \(6 < \text{MVI} < 8\), the visibility of the visual pattern
+diminishes as MVI decreases. Conversely, when \(\text{MVI} < 6\), the
+visual pattern tends to become relatively faint and challenging to
+observe. Table \ref{tab:mvi} provides a summary of the MVI usage and it
+is applicable to other linear regression models.
+
+\begin{table}
+
+\caption{\label{tab:mvi}Degree of model violations or the strength of the visual signals according to the Model Violations Index (MVI). The constant $C$ is set to be 10.}
+\centering
+\begin{tabular}[t]{lc}
+\toprule
+Degree of model violations & Range ($C$ = 10)\\
+\midrule
+Strong & $\text{MVI} > 8$\\
+Moderate & $6 < \text{MVI} < 8$\\
+Weak & $\text{MVI} < 6$\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\begin{figure}[!h]
+
+{\centering \includegraphics[width=1\linewidth]{appendix_files/figure-latex/poly-heter-index-1}
+
+}
+
+\caption{Residual plots generated from fitted models exhibiting varying degrees of (A) non-linearity and (B) heteroskedasticity violations.
+The model violations index (MVI) is displayed atop each residual plot.
+The non-linearity patterns are relatively strong for $MVI > 8$, and relatively weak for $MVI < 6$, while the heteroskedasticity patterns are relatively strong for $MVI > 8$, and relatively weak for $MVI < 6$.}\label{fig:poly-heter-index}
+\end{figure}
+
+\clearpage
+
 \bibliographystyle{tfcad}

### 2025-06-16 paper.Rmd
Remove MVI from the main paper

@@
 Section \ref{sec-model-distance-between-residual-plots} defines the distance measure used to detect model violations, while Section \ref{sec-model-distance-estimation} explains how the computer vision models estimate this distance measure.
-Section \ref{sec-model-statistical-testing} covers the statistical tests based on the estimated distance, and Section \ref{sec-model-violations-index} introduces a Model Violations Index, which offers a quicker and more convenient assessment.
+Section \ref{sec-model-statistical-testing} covers the statistical tests based on the estimated distance.
 Sections \ref{sec-model-architecture} and \ref{sec-model-data-generation} describe the model architecture and the corresponding data generation and training process, respectively.
@@
 
-The distance estimator $\hat{D}$ allows us to assess how closely the residuals resemble an ideal distribution and use as an index of model violation severity.
+The distance estimator $\hat{D}$ allows us to assess how closely the residuals resemble an ideal distribution and use as an index of model violation severity (see Appendix for details).
 However, $\hat{D}$ is not expected to equal the true distance $D$, as a single residual plot may not capture all characteristics of the residual distribution.
@@
 In practice, $Q_{null}(0.95)$ can be used to replace $Q_{boot}^{(i)}(0.95)$ in the computation.
-
-# Model Violations Index {#sec-model-violations-index}
-
-In \ref{sec-model-lineup-evaluation}, we noted that a pre-computed lattice of $\hat{D}$ quantiles can reduce the computation cost of lineup tests.
-Another practical approach is to assess model performance directly using the value of $\hat{D}$.
-
-The estimator $\hat{D}$ captures the difference between the true and reference residual distributions, which reflects the extent of model violations, making it instrumental in forming a model violations index (MVI).
-However, when more observations are used in regression, the value of $\hat{D}$ tends to increase logarithmically.
-This is because $D = \log(1 + D_{KL})$, and under the assumption of independence, $D_{KL}$ is the sum of $D_{KL}^{(i)}$ across all observations.
-This does not mean that $\hat{D}$ becomes less reliable.
-In fact, larger samples often make model violations more visible in residual plots, unless strong overlapping masks the patterns.
-
-However, to create a standardized and generalizable index, it is important to adjust for the effect of sample size. Therefore, the Model Violations Index (MVI) is proposed as
-
-\begin{equation} \label{eq:mvi}
-\text{MVI} = C + \hat{D} - \log(n),
-\end{equation}
-
-\noindent where $C$ is a sufficiently large constant to ensure the result remains positive, and the $\log(n)$ term offset the increase in $D$ with larger sample sizes.
-
-Figure \ref{fig:poly-heter-index} displays the residual plots for fitted models exhibiting varying degrees of non-linearity and heteroskedasticity.
-Each residual plot's MVI is computed using Equation \ref{eq:mvi} with $C = 10$.
-When $\text{MVI} > 8$, the visual patterns are notably strong and easily discernible by humans.
-In the range $6 < \text{MVI} < 8$, the visibility of the visual pattern diminishes as MVI decreases.
-Conversely, when $\text{MVI} < 6$, the visual pattern tends to become relatively faint and challenging to observe.
-Table \ref{tab:mvi} provides a summary of the MVI usage and it is applicable to other linear regression models.
-
-```{r}
-data.frame(degree = c("Strong", "Moderate", "Weak"),
-           range = c("$\\text{MVI} > 8$", "$6 < \\text{MVI} < 8$", "$\\text{MVI} < 6$")) %>%
-  kableExtra::kable(format = "latex",
-                    booktabs = TRUE,
-                    label = "mvi",
-                    caption = "Degree of model violations or the strength of the visual signals according to the Model Violations Index (MVI). The constant $C$ is set to be 10.",
-                    escape = FALSE,
-                    col.names = c("Degree of model violations", "Range ($C$ = 10)"),
-                    align = "lc")
-```
-
-```{r}
-set.seed(1495)
-if (!file.exists(here::here("cached_data/poly_index.rds"))) {
-  e_sigma_factor <- 3
-  keras_mod <- autovi::get_keras_model("vss_phn_32")
-  poly_dgp <- phn_model(j = 3, sigma = 0.1^e_sigma_factor)
-  ori_dat <- poly_dgp$gen(300)
-
-  vss <- c()
-  actual_ss <- c()
-  dat_combined <- tibble()
-
-  for (e_sigma in seq(0.7, 1.8, length.out = 20)^e_sigma_factor) {
-    poly_dgp <- phn_model(j = 3, sigma = e_sigma)
-    dat <- poly_dgp$gen(300, computed = list(x1 = ori_dat$x1, e = ori_dat$e, k = ori_dat$k * e_sigma / 0.1^e_sigma_factor))
-    mod <- lm(y ~ x1, data = dat)
-    my_vi <- autovi::auto_vi(fitted_mod = mod, keras_mod = keras_mod)
-    vss <- c(vss, my_vi$vss()$vss)
-    actual_ss <- c(actual_ss, log(poly_dgp$sample_effect_size(dat) + 1))
-
-    dat_combined <- bind_rows(dat_combined, mutate(dat, e_sigma = e_sigma))
-  }
-
-  saveRDS(dat_combined, here::here("cached_data/poly_index.rds"))
-  saveRDS(vss, here::here("cached_data/poly_index_vss.rds"))
-}
-
-dat_combined <- readRDS(here::here("cached_data/poly_index.rds"))
-vss <- readRDS(here::here("cached_data/poly_index_vss.rds"))
-
-p1 <- dat_combined %>%
-  autovi::AUTO_VI$plot_resid() +
-  facet_wrap(~e_sigma, scales = "free", 
-             labeller = function(...){ 
-               tibble(nb = map(glue("'MVI' == {format(vss - log(300) + 10, digits = 3)}"), str2expression))
-               },
-             ncol = 5) +
-  theme(strip.text = element_text(size = 10), plot.title = element_text(size = 10)) +
-  ggtitle("(A) Non-linearity")
-```
-
-```{r}
-set.seed(1495)
-if (!file.exists(here::here("cached_data/heter_index.rds"))) {
-  b_factor <- 2
-  keras_mod <- autovi::get_keras_model("vss_phn_32")
-  heter_dgp <- phn_model(include_z = FALSE, b = 0)
-  ori_dat <- poly_dgp$gen(300)
-
-  vss <- c()
-  actual_ss <- c()
-  dat_combined <- tibble()
-
-  for (b in rev(seq(0.2, 4, length.out = 20))^b_factor) {
-    heter_dgp <- phn_model(include_z = FALSE, b = b)
-    dat <- heter_dgp$gen(300, computed = list(x1 = ori_dat$x1, e = ori_dat$e))
-    mod <- lm(y ~ x1, data = dat)
-    my_vi <- autovi::auto_vi(fitted_mod = mod, keras_mod = keras_mod)
-    vss <- c(vss, my_vi$vss()$vss)
-    actual_ss <- c(actual_ss, log(heter_dgp$sample_effect_size(dat) + 1))
-
-    dat_combined <- bind_rows(dat_combined, mutate(dat, nb = -b))
-  }
-
-  saveRDS(dat_combined, here::here("cached_data/heter_index.rds"))
-  saveRDS(vss, here::here("cached_data/heter_index_vss.rds"))
-}
-
-dat_combined <- readRDS(here::here("cached_data/heter_index.rds"))
-vss <- readRDS(here::here("cached_data/heter_index_vss.rds"))
-
-p2 <- dat_combined %>%
-  autovi::AUTO_VI$plot_resid() +
-  facet_wrap(~nb, scales = "free", 
-             labeller = function(...){ 
-               tibble(nb = map(glue("'MVI' == {format(vss - log(300) + 10, digits = 3)}"), str2expression))
-               },
-             ncol = 5) +
-  theme(strip.text = element_text(size = 10), plot.title = element_text(size = 10)) +
-  ggtitle("(B) Heteroskedasticity")
-```
-
-```{r poly-heter-index, fig.pos = "!h", fig.cap = "Residual plots generated from fitted models exhibiting varying degrees of (A) non-linearity and (B) heteroskedasticity violations.
-The model violations index (MVI) is displayed atop each residual plot.
-The non-linearity patterns are relatively strong for $MVI > 8$, and relatively weak for $MVI < 6$, while the heteroskedasticity patterns are relatively strong for $MVI > 8$, and relatively weak for $MVI < 6$.", fig.height = 10}
-patchwork::wrap_plots(p1, p2, ncol = 1)
-```
 
@@
 
-Further details, including mathematical formulations of the simulation model, parameter specifications, sampling scheme, example residual plots and hyperparameter tuning configuration, are provided in the Appendix.
+Further details, including mathematical formulations of the  synthetic data model, parameter specifications, sampling scheme, example residual plots and hyperparameter tuning configuration, are provided in the Appendix.
 
@@
 Therefore, we analysed the relationship between residuals and all the factors involved in the data generating process.
-We found that most deviations stemmed from non-linearity and the presence of a second predictor in the simulation model, where small $\varepsilon$ variances caused under-predictions and large $\varepsilon$ variances caused over-prediction, as illustrated in Figure \ref{fig:over-under}.
+We found that most deviations stemmed from non-linearity and the presence of a second predictor in the synthetic data model, where small $\varepsilon$ variances caused under-predictions and large $\varepsilon$ variances caused over-prediction, as illustrated in Figure \ref{fig:over-under}.
 
@@
 It was found that null plots had the poorest metrics, while non-normality cases were easiest to predict, with RMSE around $0.3$.
-Non-linearity was harder to assess than heteroskedasticity or non-normality.
+Non-linearity ($\text{RMSE} \approx  0.8$) was harder to assess than heteroskedasticity ($\text{RMSE} \approx  0.6$) or non-normality ($\text{RMSE} \approx  0.3$).
 Plots with multiple violations showed intermediate performance.
@@
 
-<!-- The experiment revealed that conventional tests are more sensitive to weak departures from model assumptions than visual test, and they often reject the null hypothesis when departures are not visibly different from null residual plots.
--->
-
 This study has collected 7,974 human responses to 1,152 lineups.
@@
 ', fig.height = 8}
-p1 <- my_vi$plot_resid(theme = theme_light()) + ggtitle(glue::glue("(A) Residual plot (MVI = {format(my_vi$check_result$observed$vss - log(nrow(my_vi$get_fitted_and_resid())) + 10, digits = 3)}, weak)"))
+p1 <- my_vi$plot_resid(theme = theme_light()) + ggtitle(glue::glue("(A) Residual plot")) 
 p2 <- readRDS(here::here("cached_data/fig1_attention_map.rds")) + ggtitle("(B) Attention map")
@@
 ', fig.height = 8}
-p1 <- my_vi$plot_resid(theme = theme_light()) + ggtitle(glue::glue("(A) Residual plot (MVI = {format(my_vi$check_result$observed$vss - log(nrow(my_vi$get_fitted_and_resid())) + 10, digits = 3)}, strong)"))
+p1 <- my_vi$plot_resid(theme = theme_light()) + ggtitle(glue::glue("(A) Residual plot")) 
 p2 <- readRDS(here::here("cached_data/boston_attention_map.rds")) + ggtitle("(B) Attention map")
@@
 
-p1 <- my_vi$plot_resid(theme = theme_light()) + ggtitle(glue::glue("(A) Residual plot (MVI = {format(my_vi$check_result$observed$vss - log(nrow(my_vi$get_fitted_and_resid())) + 10, digits = 3)}, strong)"))
+p1 <- my_vi$plot_resid(theme = theme_light()) + ggtitle(glue::glue("(A) Residual plot")) 
 p2 <- readRDS(here::here("cached_data/dino_attention_map.rds")) + ggtitle("(B) Attention map")
@@
 We propose a computer vision model to estimate this distance, utilizing the residual plot of the fitted model as input.
-The resulting estimated distance serves as the foundation for constructing a single Model Violations Index (MVI), facilitating the quantification of various model violations.
 

### 2025-06-16 paper.tex
Remove MVI from the main paper

@@
 \ref{sec-model-statistical-testing} covers the statistical tests based
-on the estimated distance, and Section \ref{sec-model-violations-index}
-introduces a Model Violations Index, which offers a quicker and more
-convenient assessment. Sections \ref{sec-model-architecture} and
+on the estimated distance. Sections \ref{sec-model-architecture} and
 \ref{sec-model-data-generation} describe the model architecture and the
@@
 residuals resemble an ideal distribution and use as an index of model
-violation severity. However, \(\hat{D}\) is not expected to equal the
-true distance \(D\), as a single residual plot may not capture all
-characteristics of the residual distribution. For the same distribution
-\(P\), simulated plots can vary visually, especially with small \(n\),
-leading to estimation error that depends on how representative the input
-plot is.
+violation severity (see Appendix for details). However, \(\hat{D}\) is
+not expected to equal the true distance \(D\), as a single residual plot
+may not capture all characteristics of the residual distribution. For
+the same distribution \(P\), simulated plots can vary visually,
+especially with small \(n\), leading to estimation error that depends on
+how representative the input plot is.
 
@@
 
-\section{Model Violations Index}\label{sec-model-violations-index}
-
-In \ref{sec-model-lineup-evaluation}, we noted that a pre-computed
-lattice of \(\hat{D}\) quantiles can reduce the computation cost of
-lineup tests. Another practical approach is to assess model performance
-directly using the value of \(\hat{D}\).
-
-The estimator \(\hat{D}\) captures the difference between the true and
-reference residual distributions, which reflects the extent of model
-violations, making it instrumental in forming a model violations index
-(MVI). However, when more observations are used in regression, the value
-of \(\hat{D}\) tends to increase logarithmically. This is because
-\(D = \log(1 + D_{KL})\), and under the assumption of independence,
-\(D_{KL}\) is the sum of \(D_{KL}^{(i)}\) across all observations. This
-does not mean that \(\hat{D}\) becomes less reliable. In fact, larger
-samples often make model violations more visible in residual plots,
-unless strong overlapping masks the patterns.
-
-However, to create a standardized and generalizable index, it is
-important to adjust for the effect of sample size. Therefore, the Model
-Violations Index (MVI) is proposed as
-
-\begin{equation} \label{eq:mvi}
-\text{MVI} = C + \hat{D} - \log(n),
-\end{equation}
-
-\noindent where \(C\) is a sufficiently large constant to ensure the
-result remains positive, and the \(\log(n)\) term offset the increase in
-\(D\) with larger sample sizes.
-
-Figure \ref{fig:poly-heter-index} displays the residual plots for fitted
-models exhibiting varying degrees of non-linearity and
-heteroskedasticity. Each residual plot's MVI is computed using Equation
-\ref{eq:mvi} with \(C = 10\). When \(\text{MVI} > 8\), the visual
-patterns are notably strong and easily discernible by humans. In the
-range \(6 < \text{MVI} < 8\), the visibility of the visual pattern
-diminishes as MVI decreases. Conversely, when \(\text{MVI} < 6\), the
-visual pattern tends to become relatively faint and challenging to
-observe. Table \ref{tab:mvi} provides a summary of the MVI usage and it
-is applicable to other linear regression models.
-
-\begin{table}
-
-\caption{\label{tab:mvi}Degree of model violations or the strength of the visual signals according to the Model Violations Index (MVI). The constant $C$ is set to be 10.}
-\centering
-\begin{tabular}[t]{lc}
-\toprule
-Degree of model violations & Range ($C$ = 10)\\
-\midrule
-Strong & $\text{MVI} > 8$\\
-Moderate & $6 < \text{MVI} < 8$\\
-Weak & $\text{MVI} < 6$\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\begin{figure}[!h]
-
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/poly-heter-index-1}
-
-}
-
-\caption{Residual plots generated from fitted models exhibiting varying degrees of (A) non-linearity and (B) heteroskedasticity violations.
-The model violations index (MVI) is displayed atop each residual plot.
-The non-linearity patterns are relatively strong for $MVI > 8$, and relatively weak for $MVI < 6$, while the heteroskedasticity patterns are relatively strong for $MVI > 8$, and relatively weak for $MVI < 6$.}\label{fig:poly-heter-index}
-\end{figure}
-
 \section{Model Architecture}\label{sec-model-architecture}
@@
 
-Further details, including mathematical formulations of the simulation
-model, parameter specifications, sampling scheme, example residual plots
-and hyperparameter tuning configuration, are provided in the Appendix.
+Further details, including mathematical formulations of the synthetic
+data model, parameter specifications, sampling scheme, example residual
+plots and hyperparameter tuning configuration, are provided in the
+Appendix.
 
@@
 stemmed from non-linearity and the presence of a second predictor in the
-simulation model, where small \(\varepsilon\) variances caused
+synthetic data model, where small \(\varepsilon\) variances caused
 under-predictions and large \(\varepsilon\) variances caused
@@
 that null plots had the poorest metrics, while non-normality cases were
-easiest to predict, with RMSE around \(0.3\). Non-linearity was harder
-to assess than heteroskedasticity or non-normality. Plots with multiple
-violations showed intermediate performance.
+easiest to predict, with RMSE around \(0.3\). Non-linearity
+(\(\text{RMSE} \approx  0.8\)) was harder to assess than
+heteroskedasticity (\(\text{RMSE} \approx  0.6\)) or non-normality
+(\(\text{RMSE} \approx  0.3\)). Plots with multiple violations showed
+intermediate performance.
 
@@
 computer vision model to estimate this distance, utilizing the residual
-plot of the fitted model as input. The resulting estimated distance
-serves as the foundation for constructing a single Model Violations
-Index (MVI), facilitating the quantification of various model
-violations.
+plot of the fitted model as input.
 

### 2025-06-16 paper.Rmd
Reduce Model Performance on the Human-evaluated Data section

@@
 
-For each lineup used in @li2024plot, there is one true residual plot and 19 null plots.
-While the distance $D$ for the true residual plot depends on the underlying data generating process, the distance $D$ for the null plots is zero.
-We have used our optimized computer vision model to estimate distance for both the true residual plots and the null plots.
-To have a fair comparison, $H_0$ will be rejected if the true residual plot has the greatest estimated distance among all plots in a lineup.
-Additionally, the appropriate conventional tests including the Ramsey Regression Equation Specification Error Test (RESET) [@ramsey1969tests] for non-linearity and the Breusch-Pagan test  [@breusch1979simple] for heteroskedasticity were applied on the same data for comparison.
-
-The performance metrics of $\hat{D}$ for true residual plots are outlined in Table \ref{tab:experiment-performance}.
-It is notable that all performance metrics are slightly worse than those evaluated on the test data.
-Nevertheless, the mean absolute error remains at a low level, and the linear correlation between the prediction and the true value remains very high.
-Consistent with results in Table \ref{tab:performance-sub}, lineups with non-linearity issues are more challenging to predict than those with heteroskedasticity issues.
-
-Table \ref{tab:human-conv-table} provides a summary of the agreement between decisions made by the computer vision model and conventional tests.
-The agreement rates between conventional tests and the computer vision model are 85.95% and 79.69% for residual plots containing heteroskedasticity and non-linearity patterns, respectively.
-These figures are higher than those calculated for visual tests conducted by human, indicating that the computer vision model exhibits behavior more akin to the best available conventional tests.
-However, Figure \ref{fig:conv-mosaic} shows that the computer vision model does not always reject when the conventional tests reject.
-And a small number of plots will be rejected by computer vision model but not by conventional tests.
-This suggests that conventional tests are more sensitive than the computer vision model.
-
-Figure \ref{fig:pcp} further illustrates the decisions made by visual tests conducted by human, computer vision models, and conventional tests, using a parallel coordinate plots.
-It can be observed that all three tests will agree with each other for around 50% of the cases.
-When visual tests conducted by human do not reject, there are substantial amount of cases where computer vision model also do not reject but conventional tests reject.
-There are much fewer cases that do not reject by visual tests and conventional tests, but is rejected by computer vision models.
-This indicates computer vision model can behave like visual tests conducted by human better than conventional tests.
-Moreover, there are great proportion of cases where visual tests conducted by human is the only test who does not reject.
-
-When plotting the decision against the distance, as illustrated in Figure \ref{fig:power}, several notable observations emerge.
-Firstly, compared to conventional tests, the computer vision model tends to have fewer rejected cases when $D < 2$ and fewer non-rejected cases when $2< D < 4$.
-This suggests tests based on the computer vision model are less sensitive to small deviations from model assumptions than conventional tests but more sensitive to moderate deviations.
-Additionally, visual tests demonstrate the lowest sensitivity to residual plots with small distances where not many residual plots are rejected when $D < 2$.
-Similarly, for large distances where $D > 4$, almost all residual plots are rejected by the computer vision model and conventional tests, but for visual tests conducted by humans, the threshold is higher with $D > 5$.
-
-In Figure \ref{fig:power}, rejection decisions are fitted by logistic regression models with no intercept terms and an offset equals to $\text{log}(0.05/0.95)$.
-The fitted curves for the computer vision model fall between those of conventional tests and visual tests for both non-linearity and heteroskedasticity, which means there is still potential to refine the computer vision model to better align its behavior with visual tests conducted by humans.
-
-In the experiment conducted in @li2024plot, participants were allowed to make multiple selections for a lineup.
-The weighted detection rate was computed by assigning weights to each detection.
-If the participant selected zero plots, a weight of 0.05 was assigned; otherwise, if the true residual plot was detected, the weight was 1 divided by the number of selections.
-This weighted detection rate allow us to assess the quality of the distance measure purposed in this paper, by using the $\delta$-difference statistic.
-The $\delta$-difference is originally defined by @chowdhury2018measuring, is given by
+Each lineup in @li2024plot contains one true residual plot and 19 null plots.
+While the true plot’s distance $D$ depends on the data-generating process, null plots have $D = 0$.
+We used our optimized computer vision model to estimate $D$ for all plots.
+To ensure fair comparison, we reject $H_0$ if the true plot has the highest estimated distance.
+Conventional tests, including the Ramsey RESET [@ramsey1969tests] for non-linearity and the Breusch-Pagan test [@breusch1979simple] for heteroskedasticity, were also applied to the same data.
+
+Table \ref{tab:experiment-performance} reports performance metrics for $\hat{D}$ on true plots.
+While the performance is slightly worse than on the test data provided in Table \ref{tab:performance}, the mean absolute error remains low and the correlation between predictions and true values remains high.
+Consistent with results in Table \ref{tab:performance-sub}, non-linearity cases are harder to predict than heteroskedasticity cases.
+
+To assess how well the model’s decisions align with other methods, Table \ref{tab:human-conv-table} compares its rejection decisions with those of conventional tests and human visual tests.
+Agreement rates are 85.95% for heteroskedasticity cases and 79.69% for non-linearity cases, higher than those observed for human visual tests (67.96% and 66.84%, respectively).
+This suggests that the model aligns more closely with the best available conventional tests.
+However, as Figure \ref{fig:conv-mosaic} shows, the model does not always reject when conventional tests do, indicating a lower sensitivity to model violations.
+
+To further illustrate these patterns, Figure \ref{fig:pcp} presents parallel coordinate plots comparing decisions from human visual tests, the computer vision model, and conventional tests.
+All three agree in about 50% of cases.
+When humans do not reject, conventional tests reject in about half of those cases, whereas the model rejects in only 25%.
+Notably, the model rarely rejects when both humans and conventional tests do not, suggesting that while it is more sensitive than humans, its rejection behavior is more aligned with human judgment than with conventional tests.
+
+Figure \ref{fig:power} plots rejection decisions against the true distance $D$.
+Compared to conventional tests, the computer vision model rejects fewer cases when $D < 4$, but a similar number of cases when $D > 4$.
+This indicates that the model is less sensitive to minor deviations from model assumptions, while remaining comparably sensitive to more substantial violations.
+Visual tests are the least sensitive overall, showing few rejections for $D < 2$, and requiring a higher threshold ($D > 5$) for near-universal rejection, compared to $D > 4$ for both the model and conventional tests.
+
+Finally, the power curves in Figure \ref{fig:power} are fitted using logistic regression without intercepts and with an offset of $\log(0.05/0.95)$ to ensure a 5% rejection rate under $H_0$.
+The model’s rejection curve consistently lies between those of conventional and human visual tests, indicating potential to further align its decision boundary with human visual assessment.
+
+### Adjusted $\delta$-difference
+
+In the experiment conducted by @li2024plot, participants could select multiple plots per lineup.
+To reflect detection confidence, a weighted detection rate was computed for each lineup: selecting zero plots yielded a weight of 0.05; otherwise, if the true residual plot was selected, its weight was $1 / \text{(number of selections)}$.
+
+To evaluate the proposed distance measure, we could examine the relationship between the weighted detection rate and the $\delta$-difference statistic from @chowdhury2018measuring:
 
@@
 
-where $\bar{d}_{\text{null}}^{(j)}$ is the mean distance between the $j$-th null plot and the other null plots, $\bar{d}_{\text{true}}$ is the mean distance between the true residual plot and null plots, and $m$ is the number of plots in a lineup.
-These mean distances are used because, as noted by @chowdhury2018measuring, the distances can vary depending on which data plot is used for comparison.
-For instance, with three null plots, A, B and C, the distance between A and B may differ from the distance between A and C.
-To obtain a consistent distance for null plot A, averaging is necessary.
-However, this approach is not applicable to the distance proposed in this paper, as we only compare the residual plot against a theoretically good residual plot.
-Consequently, the statistic must be adjusted to evaluate our distance measure effectively.
-
-One important aspect that the $\delta$-difference was designed to capture is the empirical distribution of distances for null plot.
-If we were to replace the mean distances $\bar{d}_{\text{null}}^{(j)}$ directly with $D_{\text{null}}^{(j)}$, the distance of the $j$-th null plot, the resulting distribution would be degenerate, since $D_{null}$ equals zero by definition.
-Additionally, $D$ can not be derived from an image, meaning it falls outside the scope of the distances considered by @chowdhury2018measuring.
-Instead, the focus should be on the empirical distribution of $\hat{D}$, as it influences decision-making.
-Therefore, the adjusted $\delta$-different is defined as
+\noindent where $\bar{d}_{\text{true}}$ is the mean distance between the true plot and all null plots, and $\bar{d}_{\text{null}}^{(j)}$ is the mean distance from null plot $j$ to other nulls.
+These averages help address asymmetries in pairwise distances.
+
+However, our method only compares each plot to a theoretically "good" residual plot, making the original formulation inapplicable.
+Moreover, since $D_{\text{null}} = 0$ by definition and $D$ can not be derived from an image, we instead focus on $\hat{D}$, the model-estimated distance.
+This leads to the adjusted $\delta$-difference:
 
@@
 
-Figure \ref{fig:delta} displays the scatter plot of the weighted detection rate vs the adjusted $\delta$-difference.
-It indicates that the weighted detection rate increases as the adjusted $\delta$-difference increases, particularly when the adjusted $\delta$-difference is greater than zero.
-A negative adjusted $\delta$-difference suggests that there is at least one null plot in the lineup with a stronger visual signal than the true residual plot.
-In some instances, the weighted detection rate is close to one, yet the adjusted $\delta$-difference is negative.
+Figure \ref{fig:delta} displays the scatter plot of the weighted detection rate against $\delta_{\text{adj}}$.
+It shows that weighted detection generally increases with $\delta_{\text{adj}}$, especially when it is positive.
+A negative $\delta_{\text{adj}}$ suggests some null plots may appear more visually distinctive to humans than the true plot.
+In some instances, the weighted detection rate is close to one, yet the $\delta_{\text{adj}}$ is negative.
 This discrepancy implies that the distance measure, or the estimated distance, may not perfectly reflect actual human behavior.
@@
 ```{r conv-mosaic, fig.pos = "!h", fig.cap = "Rejection rate ($p$-value $\\leq0.05$) of computer vision models conditional on conventional tests on non-linearity (left) and heteroskedasticity (right) lineups displayed using a mosaic plot.
-When the conventional test fails to reject, the computer vision mostly fails to reject the same plot as well as indicated by the height of the top right yellow rectangle, but there are non negliable amount of plots where the conventional test rejects but the computer vision model fails to reject as indicated by the width of the top left yellow rectangle."}
+When the conventional test fails to reject, the computer vision mostly fails to reject the same plot as well as indicated by the height of the top right yellow rectangle, but there are non negliable amount of plots where the conventional test rejects but the computer vision model fails to reject as indicated by the width of the top left yellow rectangle.", fig.height = 4}
 library(ggmosaic)
@@
   coord_fixed()
-```
-
-```{r power, fig.pos = "!h", fig.cap = "Comparison of power of visual tests, conventional tests and the computer vision model.
-Marks along the x-axis at the bottom of the plot represent rejections made by each type of test.
-Marks at the top of the plot represent acceptances.
-Power curves are fitted by logistic regression models with no intercept but an offset equals to $\\text{log}(0.05/0.95)$."}
-
-max_actual_ss <- max(filter(experiment, !attention_check, !null_lineup)$actual_ss)
-max_poly_ss <- max(filter(experiment, !attention_check, !null_lineup, type == "polynomial")$actual_ss)
-max_heter_ss <- max(filter(experiment, !attention_check, !null_lineup, type != "polynomial")$actual_ss)
-
-model_glm_pred_poly <- glm(model_reject_20 ~ actual_ss - 1, 
-    data = filter(experiment, !attention_check, !null_lineup, type == "polynomial") %>% mutate(offset0 = log(0.05/0.95)), 
-    offset = offset0,
-    family = binomial()) %>%
-  predict(data.frame(actual_ss = seq(0, max_poly_ss, 0.01), offset0 = log(0.05/0.95)),
-          type = "response") %>%
-  data.frame(d = seq(0, max_poly_ss, 0.01), type = "Non-linearity", name = "Computer vision model", value = .) %>%
-  mutate(type = factor(type, levels = c("Non-linearity", "Heteroskedasticity")))
-
-visual_glm_pred_poly <- glm(reject ~ actual_ss - 1, 
-    data = filter(experiment, !attention_check, !null_lineup, type == "polynomial") %>% mutate(offset0 = log(0.05/0.95)), 
-    offset = offset0,
-    family = binomial()) %>%
-  predict(data.frame(actual_ss = seq(0, max_poly_ss, 0.01), offset0 = log(0.05/0.95)),
-          type = "response") %>%
-  data.frame(d = seq(0, max_poly_ss, 0.01), type = "Non-linearity", name = "Visual test", value = .) %>%
-  mutate(type = factor(type, levels = c("Non-linearity", "Heteroskedasticity")))
-
-conv_glm_pred_poly <- glm(conventional_reject ~ actual_ss - 1, 
-    data = filter(experiment, !attention_check, !null_lineup, type == "polynomial") %>% mutate(offset0 = log(0.05/0.95)), 
-    offset = offset0,
-    family = binomial()) %>%
-  predict(data.frame(actual_ss = seq(0, max_poly_ss, 0.01), offset0 = log(0.05/0.95)),
-          type = "response") %>%
-  data.frame(d = seq(0, max_poly_ss, 0.01), type = "Non-linearity", name = "Conventional test", value = .) %>%
-  mutate(type = factor(type, levels = c("Non-linearity", "Heteroskedasticity")))
-
-model_glm_pred_heter <- glm(model_reject_20 ~ actual_ss - 1, 
-    data = filter(experiment, !attention_check, !null_lineup, type != "polynomial") %>% mutate(offset0 = log(0.05/0.95)), 
-    offset = offset0,
-    family = binomial()) %>%
-  predict(data.frame(actual_ss = seq(0, max_heter_ss, 0.01), offset0 = log(0.05/0.95)),
-          type = "response") %>%
-  data.frame(d = seq(0, max_heter_ss, 0.01), type = "Heteroskedasticity", name = "Computer vision model", value = .) %>%
-  mutate(type = factor(type, levels = c("Non-linearity", "Heteroskedasticity")))
-
-visual_glm_pred_heter <- glm(reject ~ actual_ss - 1, 
-    data = filter(experiment, !attention_check, !null_lineup, type != "polynomial") %>% mutate(offset0 = log(0.05/0.95)), 
-    offset = offset0,
-    family = binomial()) %>%
-  predict(data.frame(actual_ss = seq(0, max_heter_ss, 0.01), offset0 = log(0.05/0.95)),
-          type = "response") %>%
-  data.frame(d = seq(0, max_heter_ss, 0.01), type = "Heteroskedasticity", name = "Visual test", value = .) %>%
-  mutate(type = factor(type, levels = c("Non-linearity", "Heteroskedasticity")))
-
-conv_glm_pred_heter <- glm(conventional_reject ~ actual_ss - 1, 
-    data = filter(experiment, !attention_check, !null_lineup, type != "polynomial") %>% mutate(offset0 = log(0.05/0.95)), 
-    offset = offset0,
-    family = binomial()) %>%
-  predict(data.frame(actual_ss = seq(0, max_heter_ss, 0.01), offset0 = log(0.05/0.95)),
-          type = "response") %>%
-  data.frame(d = seq(0, max_heter_ss, 0.01), type = "Heteroskedasticity", name = "Conventional test", value = .) %>%
-  mutate(type = factor(type, levels = c("Non-linearity", "Heteroskedasticity")))
-
-mutate(experiment, type = ifelse(type == "polynomial", "Non-linearity", "Heteroskedasticity")) %>%
-  mutate(type = factor(type, levels = c("Non-linearity", "Heteroskedasticity"))) %>%
-  filter(!null_lineup, !attention_check) %>%
-ggplot() +
-  geom_line(data = model_glm_pred_poly, aes(d, value, col = "Computer vision model"), linewidth = 0.8) +
-  geom_line(data = model_glm_pred_heter, aes(d, value, col = "Computer vision model"), linewidth = 0.8) +
-  geom_line(data = visual_glm_pred_poly, aes(d, value, col = "Visual test"), linewidth = 0.8) +
-  geom_line(data = visual_glm_pred_heter, aes(d, value, col = "Visual test"), linewidth = 0.8) +
-  geom_line(data = conv_glm_pred_poly, aes(d, value, col = "Conventional test"), linewidth = 0.8) +
-  geom_line(data = conv_glm_pred_heter, aes(d, value, col = "Conventional test"), linewidth = 0.8) +
-  geom_segment(data = ~filter(.x, !reject), aes(x = actual_ss, xend = actual_ss, y = -0.1, yend = -0.1 + 0.03, col = "Visual test"), alpha = 0.4) +
-  geom_segment(data = ~filter(.x, reject), aes(x = actual_ss, xend = actual_ss, y = 1.1 - 0.03 * 2, yend = 1.1 - 0.03 * 3, col = "Visual test"), alpha = 0.4) +
-  geom_segment(data = ~filter(.x, !model_reject_20), aes(x = actual_ss, xend = actual_ss, y = -0.1 + 0.03, yend = -0.1 + 0.03 * 2, col = "Computer vision model"), alpha = 0.4) +
-  geom_segment(data = ~filter(.x, model_reject_20), aes(x = actual_ss, xend = actual_ss, y = 1.1 - 0.03, yend = 1.1 - 0.03 * 2, col = "Computer vision model"), alpha = 0.4) +
-  geom_segment(data = ~filter(.x, !conventional_reject), aes(x = actual_ss, xend = actual_ss, y = -0.1 + 0.03 * 2, yend = -0.1 + 0.03 * 3, col = "Conventional test"), alpha = 0.4) +
-  geom_segment(data = ~filter(.x, conventional_reject), aes(x = actual_ss, xend = actual_ss, y = 1.1, yend = 1.1 - 0.03, col = "Conventional test"), alpha = 0.4) +
-  facet_grid(~type, scales = "free_x") +
-  theme_light() +
-  scale_color_brewer("", palette = "Dark2") +
-  theme(legend.position = "bottom") +
-  xlab("D") +
-  ylab("Reject")
 ```
@@
         axis.title = element_blank())
+```
+
+```{r power, fig.pos = "!h", fig.cap = "Comparison of power of visual tests, conventional tests and the computer vision model.
+Marks along the x-axis at the bottom of the plot represent rejections made by each type of test.
+Marks at the top of the plot represent acceptances.
+Power curves are fitted by logistic regression models with no intercept but an offset equals to $\\text{log}(0.05/0.95)$.", fig.height = 4}
+
+max_actual_ss <- max(filter(experiment, !attention_check, !null_lineup)$actual_ss)
+max_poly_ss <- max(filter(experiment, !attention_check, !null_lineup, type == "polynomial")$actual_ss)
+max_heter_ss <- max(filter(experiment, !attention_check, !null_lineup, type != "polynomial")$actual_ss)
+
+model_glm_pred_poly <- glm(model_reject_20 ~ actual_ss - 1, 
+    data = filter(experiment, !attention_check, !null_lineup, type == "polynomial") %>% mutate(offset0 = log(0.05/0.95)), 
+    offset = offset0,
+    family = binomial()) %>%
+  predict(data.frame(actual_ss = seq(0, max_poly_ss, 0.01), offset0 = log(0.05/0.95)),
+          type = "response") %>%
+  data.frame(d = seq(0, max_poly_ss, 0.01), type = "Non-linearity", name = "Computer vision model", value = .) %>%
+  mutate(type = factor(type, levels = c("Non-linearity", "Heteroskedasticity")))
+
+visual_glm_pred_poly <- glm(reject ~ actual_ss - 1, 
+    data = filter(experiment, !attention_check, !null_lineup, type == "polynomial") %>% mutate(offset0 = log(0.05/0.95)), 
+    offset = offset0,
+    family = binomial()) %>%
+  predict(data.frame(actual_ss = seq(0, max_poly_ss, 0.01), offset0 = log(0.05/0.95)),
+          type = "response") %>%
+  data.frame(d = seq(0, max_poly_ss, 0.01), type = "Non-linearity", name = "Visual test", value = .) %>%
+  mutate(type = factor(type, levels = c("Non-linearity", "Heteroskedasticity")))
+
+conv_glm_pred_poly <- glm(conventional_reject ~ actual_ss - 1, 
+    data = filter(experiment, !attention_check, !null_lineup, type == "polynomial") %>% mutate(offset0 = log(0.05/0.95)), 
+    offset = offset0,
+    family = binomial()) %>%
+  predict(data.frame(actual_ss = seq(0, max_poly_ss, 0.01), offset0 = log(0.05/0.95)),
+          type = "response") %>%
+  data.frame(d = seq(0, max_poly_ss, 0.01), type = "Non-linearity", name = "Conventional test", value = .) %>%
+  mutate(type = factor(type, levels = c("Non-linearity", "Heteroskedasticity")))
+
+model_glm_pred_heter <- glm(model_reject_20 ~ actual_ss - 1, 
+    data = filter(experiment, !attention_check, !null_lineup, type != "polynomial") %>% mutate(offset0 = log(0.05/0.95)), 
+    offset = offset0,
+    family = binomial()) %>%
+  predict(data.frame(actual_ss = seq(0, max_heter_ss, 0.01), offset0 = log(0.05/0.95)),
+          type = "response") %>%
+  data.frame(d = seq(0, max_heter_ss, 0.01), type = "Heteroskedasticity", name = "Computer vision model", value = .) %>%
+  mutate(type = factor(type, levels = c("Non-linearity", "Heteroskedasticity")))
+
+visual_glm_pred_heter <- glm(reject ~ actual_ss - 1, 
+    data = filter(experiment, !attention_check, !null_lineup, type != "polynomial") %>% mutate(offset0 = log(0.05/0.95)), 
+    offset = offset0,
+    family = binomial()) %>%
+  predict(data.frame(actual_ss = seq(0, max_heter_ss, 0.01), offset0 = log(0.05/0.95)),
+          type = "response") %>%
+  data.frame(d = seq(0, max_heter_ss, 0.01), type = "Heteroskedasticity", name = "Visual test", value = .) %>%
+  mutate(type = factor(type, levels = c("Non-linearity", "Heteroskedasticity")))
+
+conv_glm_pred_heter <- glm(conventional_reject ~ actual_ss - 1, 
+    data = filter(experiment, !attention_check, !null_lineup, type != "polynomial") %>% mutate(offset0 = log(0.05/0.95)), 
+    offset = offset0,
+    family = binomial()) %>%
+  predict(data.frame(actual_ss = seq(0, max_heter_ss, 0.01), offset0 = log(0.05/0.95)),
+          type = "response") %>%
+  data.frame(d = seq(0, max_heter_ss, 0.01), type = "Heteroskedasticity", name = "Conventional test", value = .) %>%
+  mutate(type = factor(type, levels = c("Non-linearity", "Heteroskedasticity")))
+
+mutate(experiment, type = ifelse(type == "polynomial", "Non-linearity", "Heteroskedasticity")) %>%
+  mutate(type = factor(type, levels = c("Non-linearity", "Heteroskedasticity"))) %>%
+  filter(!null_lineup, !attention_check) %>%
+ggplot() +
+  geom_line(data = model_glm_pred_poly, aes(d, value, col = "Computer vision model"), linewidth = 0.8) +
+  geom_line(data = model_glm_pred_heter, aes(d, value, col = "Computer vision model"), linewidth = 0.8) +
+  geom_line(data = visual_glm_pred_poly, aes(d, value, col = "Visual test"), linewidth = 0.8) +
+  geom_line(data = visual_glm_pred_heter, aes(d, value, col = "Visual test"), linewidth = 0.8) +
+  geom_line(data = conv_glm_pred_poly, aes(d, value, col = "Conventional test"), linewidth = 0.8) +
+  geom_line(data = conv_glm_pred_heter, aes(d, value, col = "Conventional test"), linewidth = 0.8) +
+  geom_segment(data = ~filter(.x, !reject), aes(x = actual_ss, xend = actual_ss, y = -0.1, yend = -0.1 + 0.03, col = "Visual test"), alpha = 0.4) +
+  geom_segment(data = ~filter(.x, reject), aes(x = actual_ss, xend = actual_ss, y = 1.1 - 0.03 * 2, yend = 1.1 - 0.03 * 3, col = "Visual test"), alpha = 0.4) +
+  geom_segment(data = ~filter(.x, !model_reject_20), aes(x = actual_ss, xend = actual_ss, y = -0.1 + 0.03, yend = -0.1 + 0.03 * 2, col = "Computer vision model"), alpha = 0.4) +
+  geom_segment(data = ~filter(.x, model_reject_20), aes(x = actual_ss, xend = actual_ss, y = 1.1 - 0.03, yend = 1.1 - 0.03 * 2, col = "Computer vision model"), alpha = 0.4) +
+  geom_segment(data = ~filter(.x, !conventional_reject), aes(x = actual_ss, xend = actual_ss, y = -0.1 + 0.03 * 2, yend = -0.1 + 0.03 * 3, col = "Conventional test"), alpha = 0.4) +
+  geom_segment(data = ~filter(.x, conventional_reject), aes(x = actual_ss, xend = actual_ss, y = 1.1, yend = 1.1 - 0.03, col = "Conventional test"), alpha = 0.4) +
+  facet_grid(~type, scales = "free_x") +
+  theme_light() +
+  scale_color_brewer("", palette = "Dark2") +
+  theme(legend.position = "bottom") +
+  xlab("D") +
+  ylab("Reject")
 ```

### 2025-06-16 paper.tex
Reduce Model Performance on the Human-evaluated Data section

@@
 
-For each lineup used in \citet{li2024plot}, there is one true residual
-plot and 19 null plots. While the distance \(D\) for the true residual
-plot depends on the underlying data generating process, the distance
-\(D\) for the null plots is zero. We have used our optimized computer
-vision model to estimate distance for both the true residual plots and
-the null plots. To have a fair comparison, \(H_0\) will be rejected if
-the true residual plot has the greatest estimated distance among all
-plots in a lineup. Additionally, the appropriate conventional tests
-including the Ramsey Regression Equation Specification Error Test
-(RESET) \citep{ramsey1969tests} for non-linearity and the Breusch-Pagan
-test \citep{breusch1979simple} for heteroskedasticity were applied on
-the same data for comparison.
-
-The performance metrics of \(\hat{D}\) for true residual plots are
-outlined in Table \ref{tab:experiment-performance}. It is notable that
-all performance metrics are slightly worse than those evaluated on the
-test data. Nevertheless, the mean absolute error remains at a low level,
-and the linear correlation between the prediction and the true value
-remains very high. Consistent with results in Table
-\ref{tab:performance-sub}, lineups with non-linearity issues are more
-challenging to predict than those with heteroskedasticity issues.
-
-Table \ref{tab:human-conv-table} provides a summary of the agreement
-between decisions made by the computer vision model and conventional
-tests. The agreement rates between conventional tests and the computer
-vision model are 85.95\% and 79.69\% for residual plots containing
-heteroskedasticity and non-linearity patterns, respectively. These
-figures are higher than those calculated for visual tests conducted by
-human, indicating that the computer vision model exhibits behavior more
-akin to the best available conventional tests. However, Figure
-\ref{fig:conv-mosaic} shows that the computer vision model does not
-always reject when the conventional tests reject. And a small number of
-plots will be rejected by computer vision model but not by conventional
-tests. This suggests that conventional tests are more sensitive than the
-computer vision model.
-
-Figure \ref{fig:pcp} further illustrates the decisions made by visual
-tests conducted by human, computer vision models, and conventional
-tests, using a parallel coordinate plots. It can be observed that all
-three tests will agree with each other for around 50\% of the cases.
-When visual tests conducted by human do not reject, there are
-substantial amount of cases where computer vision model also do not
-reject but conventional tests reject. There are much fewer cases that do
-not reject by visual tests and conventional tests, but is rejected by
-computer vision models. This indicates computer vision model can behave
-like visual tests conducted by human better than conventional tests.
-Moreover, there are great proportion of cases where visual tests
-conducted by human is the only test who does not reject.
-
-When plotting the decision against the distance, as illustrated in
-Figure \ref{fig:power}, several notable observations emerge. Firstly,
-compared to conventional tests, the computer vision model tends to have
-fewer rejected cases when \(D < 2\) and fewer non-rejected cases when
-\(2< D < 4\). This suggests tests based on the computer vision model are
-less sensitive to small deviations from model assumptions than
-conventional tests but more sensitive to moderate deviations.
-Additionally, visual tests demonstrate the lowest sensitivity to
-residual plots with small distances where not many residual plots are
-rejected when \(D < 2\). Similarly, for large distances where \(D > 4\),
-almost all residual plots are rejected by the computer vision model and
-conventional tests, but for visual tests conducted by humans, the
-threshold is higher with \(D > 5\).
-
-In Figure \ref{fig:power}, rejection decisions are fitted by logistic
-regression models with no intercept terms and an offset equals to
-\(\text{log}(0.05/0.95)\). The fitted curves for the computer vision
-model fall between those of conventional tests and visual tests for both
-non-linearity and heteroskedasticity, which means there is still
-potential to refine the computer vision model to better align its
-behavior with visual tests conducted by humans.
-
-In the experiment conducted in \citet{li2024plot}, participants were
-allowed to make multiple selections for a lineup. The weighted detection
-rate was computed by assigning weights to each detection. If the
-participant selected zero plots, a weight of 0.05 was assigned;
-otherwise, if the true residual plot was detected, the weight was 1
-divided by the number of selections. This weighted detection rate allow
-us to assess the quality of the distance measure purposed in this paper,
-by using the \(\delta\)-difference statistic. The \(\delta\)-difference
-is originally defined by \citet{chowdhury2018measuring}, is given by
+Each lineup in \citet{li2024plot} contains one true residual plot and 19
+null plots. While the true plot's distance \(D\) depends on the
+data-generating process, null plots have \(D = 0\). We used our
+optimized computer vision model to estimate \(D\) for all plots. To
+ensure fair comparison, we reject \(H_0\) if the true plot has the
+highest estimated distance. Conventional tests, including the Ramsey
+RESET \citep{ramsey1969tests} for non-linearity and the Breusch-Pagan
+test \citep{breusch1979simple} for heteroskedasticity, were also applied
+to the same data.
+
+Table \ref{tab:experiment-performance} reports performance metrics for
+\(\hat{D}\) on true plots. While the performance is slightly worse than
+on the test data provided in Table \ref{tab:performance}, the mean
+absolute error remains low and the correlation between predictions and
+true values remains high. Consistent with results in Table
+\ref{tab:performance-sub}, non-linearity cases are harder to predict
+than heteroskedasticity cases.
+
+To assess how well the model's decisions align with other methods, Table
+\ref{tab:human-conv-table} compares its rejection decisions with those
+of conventional tests and human visual tests. Agreement rates are
+85.95\% for heteroskedasticity cases and 79.69\% for non-linearity
+cases, higher than those observed for human visual tests (67.96\% and
+66.84\%, respectively). This suggests that the model aligns more closely
+with the best available conventional tests. However, as Figure
+\ref{fig:conv-mosaic} shows, the model does not always reject when
+conventional tests do, indicating a lower sensitivity to model
+violations.
+
+To further illustrate these patterns, Figure \ref{fig:pcp} presents
+parallel coordinate plots comparing decisions from human visual tests,
+the computer vision model, and conventional tests. All three agree in
+about 50\% of cases. When humans do not reject, conventional tests
+reject in about half of those cases, whereas the model rejects in only
+25\%. Notably, the model rarely rejects when both humans and
+conventional tests do not, suggesting that while it is more sensitive
+than humans, its rejection behavior is more aligned with human judgment
+than with conventional tests.
+
+Figure \ref{fig:power} plots rejection decisions against the true
+distance \(D\). Compared to conventional tests, the computer vision
+model rejects fewer cases when \(D < 4\), but a similar number of cases
+when \(D > 4\). This indicates that the model is less sensitive to minor
+deviations from model assumptions, while remaining comparably sensitive
+to more substantial violations. Visual tests are the least sensitive
+overall, showing few rejections for \(D < 2\), and requiring a higher
+threshold (\(D > 5\)) for near-universal rejection, compared to
+\(D > 4\) for both the model and conventional tests.
+
+Finally, the power curves in Figure \ref{fig:power} are fitted using
+logistic regression without intercepts and with an offset of
+\(\log(0.05/0.95)\) to ensure a 5\% rejection rate under \(H_0\). The
+model's rejection curve consistently lies between those of conventional
+and human visual tests, indicating potential to further align its
+decision boundary with human visual assessment.
+
+\subsubsection{\texorpdfstring{Adjusted
+\(\delta\)-difference}{Adjusted \textbackslash delta-difference}}\label{adjusted-delta-difference}
+
+In the experiment conducted by \citet{li2024plot}, participants could
+select multiple plots per lineup. To reflect detection confidence, a
+weighted detection rate was computed for each lineup: selecting zero
+plots yielded a weight of 0.05; otherwise, if the true residual plot was
+selected, its weight was \(1 / \text{(number of selections)}\).
+
+To evaluate the proposed distance measure, we could examine the
+relationship between the weighted detection rate and the
+\(\delta\)-difference statistic from \citet{chowdhury2018measuring}:
 
@@
 
-where \(\bar{d}_{\text{null}}^{(j)}\) is the mean distance between the
-\(j\)-th null plot and the other null plots, \(\bar{d}_{\text{true}}\)
-is the mean distance between the true residual plot and null plots, and
-\(m\) is the number of plots in a lineup. These mean distances are used
-because, as noted by \citet{chowdhury2018measuring}, the distances can
-vary depending on which data plot is used for comparison. For instance,
-with three null plots, A, B and C, the distance between A and B may
-differ from the distance between A and C. To obtain a consistent
-distance for null plot A, averaging is necessary. However, this approach
-is not applicable to the distance proposed in this paper, as we only
-compare the residual plot against a theoretically good residual plot.
-Consequently, the statistic must be adjusted to evaluate our distance
-measure effectively.
-
-One important aspect that the \(\delta\)-difference was designed to
-capture is the empirical distribution of distances for null plot. If we
-were to replace the mean distances \(\bar{d}_{\text{null}}^{(j)}\)
-directly with \(D_{\text{null}}^{(j)}\), the distance of the \(j\)-th
-null plot, the resulting distribution would be degenerate, since
-\(D_{null}\) equals zero by definition. Additionally, \(D\) can not be
-derived from an image, meaning it falls outside the scope of the
-distances considered by \citet{chowdhury2018measuring}. Instead, the
-focus should be on the empirical distribution of \(\hat{D}\), as it
-influences decision-making. Therefore, the adjusted \(\delta\)-different
-is defined as
+\noindent where \(\bar{d}_{\text{true}}\) is the mean distance between
+the true plot and all null plots, and \(\bar{d}_{\text{null}}^{(j)}\) is
+the mean distance from null plot \(j\) to other nulls. These averages
+help address asymmetries in pairwise distances.
+
+However, our method only compares each plot to a theoretically ``good''
+residual plot, making the original formulation inapplicable. Moreover,
+since \(D_{\text{null}} = 0\) by definition and \(D\) can not be derived
+from an image, we instead focus on \(\hat{D}\), the model-estimated
+distance. This leads to the adjusted \(\delta\)-difference:
 
@@
 Figure \ref{fig:delta} displays the scatter plot of the weighted
-detection rate vs the adjusted \(\delta\)-difference. It indicates that
-the weighted detection rate increases as the adjusted
-\(\delta\)-difference increases, particularly when the adjusted
-\(\delta\)-difference is greater than zero. A negative adjusted
-\(\delta\)-difference suggests that there is at least one null plot in
-the lineup with a stronger visual signal than the true residual plot. In
-some instances, the weighted detection rate is close to one, yet the
-adjusted \(\delta\)-difference is negative. This discrepancy implies
+detection rate against \(\delta_{\text{adj}}\). It shows that weighted
+detection generally increases with \(\delta_{\text{adj}}\), especially
+when it is positive. A negative \(\delta_{\text{adj}}\) suggests some
+null plots may appear more visually distinctive to humans than the true
+plot. In some instances, the weighted detection rate is close to one,
+yet the \(\delta_{\text{adj}}\) is negative. This discrepancy implies
 that the distance measure, or the estimated distance, may not perfectly
@@
 
+\begin{figure}
+
+{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/pcp-1}
+
+}
+
+\caption{Parallel coordinate plots of decisions made by computer vision model, conventional tests and visual tests made by human.}\label{fig:pcp}
+\end{figure}
+
 \begin{figure}[!h]
@@
 Power curves are fitted by logistic regression models with no intercept but an offset equals to $\text{log}(0.05/0.95)$.}\label{fig:power}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/pcp-1}
-
-}
-
-\caption{Parallel coordinate plots of decisions made by computer vision model, conventional tests and visual tests made by human.}\label{fig:pcp}
 \end{figure}

### 2025-06-16 paper.Rmd
Adjust plot size

@@
 This often indicates the existence of heteroskedasticity, however, here the result is due to skewed distribution of the predictors rather than heteroskedasticity.
-The Breusch-Pagan test rejects this residual plot at 95\\% significance level ($p\\text{-value} = 0.046$)."}
+The Breusch-Pagan test rejects this residual plot at 95\\% significance level ($p\\text{-value} = 0.046$).", out.width = "50%"}
 set.seed(452)
@@
 The brown lines are smoothing curves produced by fitting generalized additive models.
-The area over the zero line in light yellow indicates under-prediction, and the area under the zero line in light green indicates over-prediction.", dev = 'png', dpi = 300}
+The area over the zero line in light yellow indicates under-prediction, and the area under the zero line in light green indicates over-prediction.", dev = 'png', dpi = 300, fig.height = 4.5}
 model_pred %>%
@@
 Each lineup in @li2024plot contains one true residual plot and 19 null plots.
-While the true plot’s distance $D$ depends on the data-generating process, null plots have $D = 0$.
+While the true residual plot’s distance $D$ depends on the data-generating process, null plots have $D = 0$.
 We used our optimized computer vision model to estimate $D$ for all plots.
-To ensure fair comparison, we reject $H_0$ if the true plot has the highest estimated distance.
+To ensure fair comparison, we reject $H_0$ if the true residual plot has the highest estimated distance.
 Conventional tests, including the Ramsey RESET [@ramsey1969tests] for non-linearity and the Breusch-Pagan test [@breusch1979simple] for heteroskedasticity, were also applied to the same data.
 
-Table \ref{tab:experiment-performance} reports performance metrics for $\hat{D}$ on true plots.
+Table \ref{tab:experiment-performance} reports performance metrics for $\hat{D}$ on true residual plots.
 While the performance is slightly worse than on the test data provided in Table \ref{tab:performance}, the mean absolute error remains low and the correlation between predictions and true values remains high.
@@
 
-\noindent where $\bar{d}_{\text{true}}$ is the mean distance between the true plot and all null plots, and $\bar{d}_{\text{null}}^{(j)}$ is the mean distance from null plot $j$ to other nulls.
+\noindent where $\bar{d}_{\text{true}}$ is the mean distance between the true residual plot and all null plots, and $\bar{d}_{\text{null}}^{(j)}$ is the mean distance from null plot $j$ to other nulls.
 These averages help address asymmetries in pairwise distances.
@@
 It shows that weighted detection generally increases with $\delta_{\text{adj}}$, especially when it is positive.
-A negative $\delta_{\text{adj}}$ suggests some null plots may appear more visually distinctive to humans than the true plot.
-In some instances, the weighted detection rate is close to one, yet the $\delta_{\text{adj}}$ is negative.
-This discrepancy implies that the distance measure, or the estimated distance, may not perfectly reflect actual human behavior.
+This suggests that the differences in $\hat{D}$ capture some aspect of the visual separability between the true and null plots as perceived by humans, with larger positive differences generally indicating greater distinctiveness.
+However, the considerable variability in detection rates for similar $\delta_{\text{adj}}$ values implies that $\hat{D}$ does not fully account for all factors influencing human perception, and the alignment remains imperfect.
+A negative $\delta_{\text{adj}}$ suggests some null plots may appear more visually distinctive to humans than the true residual plot.
+In some instances, the weighted detection rate is close to one despite a negative $\delta_{\text{adj}}$.
+This discrepancy implies that the distance measure may not perfectly reflect actual human behavior.
 
@@
 ```{r conv-mosaic, fig.pos = "!h", fig.cap = "Rejection rate ($p$-value $\\leq0.05$) of computer vision models conditional on conventional tests on non-linearity (left) and heteroskedasticity (right) lineups displayed using a mosaic plot.
-When the conventional test fails to reject, the computer vision mostly fails to reject the same plot as well as indicated by the height of the top right yellow rectangle, but there are non negliable amount of plots where the conventional test rejects but the computer vision model fails to reject as indicated by the width of the top left yellow rectangle.", fig.height = 4}
+When the conventional test fails to reject, the computer vision mostly fails to reject the same plot as well as indicated by the height of the top right yellow rectangle, but there are non negliable amount of plots where the conventional test rejects but the computer vision model fails to reject as indicated by the width of the top left yellow rectangle.", fig.height = 3}
 library(ggmosaic)
@@
 
-```{r pcp, fig.cap = "Parallel coordinate plots of decisions made by computer vision model, conventional tests and visual tests made by human.", eval = TRUE, fig.height = 4}
+```{r pcp, fig.cap = "Parallel coordinate plots of decisions made by computer vision model, conventional tests and visual tests made by human.", eval = TRUE, fig.height = 3.5}
 library(ggpcp)
@@
 Marks at the top of the plot represent acceptances.
-Power curves are fitted by logistic regression models with no intercept but an offset equals to $\\text{log}(0.05/0.95)$.", fig.height = 4}
+Power curves are fitted by logistic regression models with no intercept but an offset equals to $\\text{log}(0.05/0.95)$.", fig.width = 7, fig.height = 3.5}
 
@@
 ```{r delta, fig.pos = "!h", fig.cap = "A weighted detection rate vs adjusted $\\delta$-difference plot.
-The brown line is smoothing curve produced by fitting generalized additive models.", fig.height = 4}
+The brown line is smoothing curve produced by fitting generalized additive models.", fig.width = 6, fig.height = 3, out.width = "80%"}
 experiment %>%
@@
 
-In this section, we present the performance of trained computer vision model on three example datasets.
-These include the dataset associated with the residual plot displaying a "left-triangle" shape, as displayed in Figure \ref{fig:false-finding}, along with the Boston housing dataset [@harrison1978hedonic], and the "dino" datasets from the `datasauRus` R package [@datasaurus].
-
-The first example illustrates a scenario where both the computer vision model and human visual inspection successfully avoid rejecting $H_0$ when $H_0$ is true, contrary to conventional tests.
-This underscores the necessity of visually examining the residual plot.
-
-In the second example, we encounter a more pronounced violation of the model, resulting in rejection of $H_0$ by all three tests.
-This highlights the practicality of the computer vision model, particularly for less intricate tasks.
-
-The third example presents a situation where the model deviation is non-typical.
-Here, the computer vision model and human visual inspection reject $H_0$, whereas some commonly used conventional tests do not.
-This emphasizes the benefits of visual inspection and the unique advantage of the computer vision model, which, like humans, makes decisions based on visual discoveries.
+In this section, we present the performance of the trained computer vision model on three example datasets, including the dataset associated with the residual plot displaying a "left-triangle" shape (Figure \ref{fig:false-finding}), the Boston housing dataset [@harrison1978hedonic], and the "dino" datasets from the from the `datasauRus` package [@datasaurus].
+
+In the first case, both the model and human inspection correctly avoid rejecting $H_0$ when it is true, contrary to conventional tests, highlighting the necessity of visual examination.
+The second case shows a clear model violation detected by all tests, demonstrating the model’s practical use for less intricate tasks.
+The third case involves an unusual deviation where the model and humans reject $H_0$, but some conventional tests do not, illustrating the benefits of visual-based decision-making.
 
@@
 ```{r false-lineup, fig.pos = "!h", fig.cap = 'A lineup of residual plots displaying "left-triangle" visual patterns.
-The true residual plot occupies position 10, yet there are no discernible visual patterns that distinguish it from the other plots.', fig.height = 8}
+The true residual plot occupies position 10, yet there are no discernible visual patterns that distinguish it from the other plots.', fig.height = 8, out.width = "70%"}
 pos <- 10
@@
 The PCA provided in plot D shows two distinct clusters of data points, further underling the visual differences between bootstrapped plots and null plots.
-This coincides the findings from Figure \ref{fig:boston-lineup}, where the true plot exhibiting a "U" shape is visually distinctive from null plots.
+This coincides the findings from Figure \ref{fig:boston-lineup}, where the true residual plot exhibiting a "U" shape is visually distinctive from null plots.
 If a visual test is conducted by human, $H_0$ will also be rejected.
@@
 The true residual plot is at position 7.
-It can be easily identified as the most different plot.', fig.height = 8}
+It can be easily identified as the most different plot.', fig.height = 8, out.width = "70%"}
 pos <- 7
@@
 The true residual plot is at position 17.
-It can be easily identified as the most different plot as the visual pattern is extremely artificial.', fig.height = 8}
+It can be easily identified as the most different plot as the visual pattern is extremely artificial.', fig.height = 8, out.width = "70%"}
 pos <- 17
@@
 Another approach would be to transform the residuals so they are roughly normally distributed and have constant variance.
-If only raw residuals are used, the distance-based statistical testing compares the difference in distance to a classical normal linear regression model for the true plot and null plots.
+If only raw residuals are used, the distance-based statistical testing compares the difference in distance to a classical normal linear regression model for the true residual plot and null plots.
 This comparison is meaningful only if the difference can be identified by the distance measure proposed in this paper.

### 2025-06-16 paper.tex
Adjust plot size

@@
 
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/false-finding-1}
+{\centering \includegraphics[width=0.5\linewidth]{paper_files/figure-latex/false-finding-1}
 
@@
 Each lineup in \citet{li2024plot} contains one true residual plot and 19
-null plots. While the true plot's distance \(D\) depends on the
+null plots. While the true residual plot's distance \(D\) depends on the
 data-generating process, null plots have \(D = 0\). We used our
 optimized computer vision model to estimate \(D\) for all plots. To
-ensure fair comparison, we reject \(H_0\) if the true plot has the
-highest estimated distance. Conventional tests, including the Ramsey
+ensure fair comparison, we reject \(H_0\) if the true residual plot has
+the highest estimated distance. Conventional tests, including the Ramsey
 RESET \citep{ramsey1969tests} for non-linearity and the Breusch-Pagan
@@
 Table \ref{tab:experiment-performance} reports performance metrics for
-\(\hat{D}\) on true plots. While the performance is slightly worse than
-on the test data provided in Table \ref{tab:performance}, the mean
-absolute error remains low and the correlation between predictions and
-true values remains high. Consistent with results in Table
+\(\hat{D}\) on true residual plots. While the performance is slightly
+worse than on the test data provided in Table \ref{tab:performance}, the
+mean absolute error remains low and the correlation between predictions
+and true values remains high. Consistent with results in Table
 \ref{tab:performance-sub}, non-linearity cases are harder to predict
@@
 \noindent where \(\bar{d}_{\text{true}}\) is the mean distance between
-the true plot and all null plots, and \(\bar{d}_{\text{null}}^{(j)}\) is
-the mean distance from null plot \(j\) to other nulls. These averages
-help address asymmetries in pairwise distances.
+the true residual plot and all null plots, and
+\(\bar{d}_{\text{null}}^{(j)}\) is the mean distance from null plot
+\(j\) to other nulls. These averages help address asymmetries in
+pairwise distances.
 
@@
 detection generally increases with \(\delta_{\text{adj}}\), especially
-when it is positive. A negative \(\delta_{\text{adj}}\) suggests some
-null plots may appear more visually distinctive to humans than the true
-plot. In some instances, the weighted detection rate is close to one,
-yet the \(\delta_{\text{adj}}\) is negative. This discrepancy implies
-that the distance measure, or the estimated distance, may not perfectly
-reflect actual human behavior.
+when it is positive. This suggests that the differences in \(\hat{D}\)
+capture some aspect of the visual separability between the true and null
+plots as perceived by humans, with larger positive differences generally
+indicating greater distinctiveness. However, the considerable
+variability in detection rates for similar \(\delta_{\text{adj}}\)
+values implies that \(\hat{D}\) does not fully account for all factors
+influencing human perception, and the alignment remains imperfect. A
+negative \(\delta_{\text{adj}}\) suggests some null plots may appear
+more visually distinctive to humans than the true residual plot. In some
+instances, the weighted detection rate is close to one despite a
+negative \(\delta_{\text{adj}}\). This discrepancy implies that the
+distance measure may not perfectly reflect actual human behavior.
 
@@
 
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/delta-1}
+{\centering \includegraphics[width=0.8\linewidth]{paper_files/figure-latex/delta-1}
 
@@
 
-In this section, we present the performance of trained computer vision
-model on three example datasets. These include the dataset associated
-with the residual plot displaying a ``left-triangle'' shape, as
-displayed in Figure \ref{fig:false-finding}, along with the Boston
-housing dataset \citep{harrison1978hedonic}, and the ``dino'' datasets
-from the \texttt{datasauRus} R package \citep{datasaurus}.
-
-The first example illustrates a scenario where both the computer vision
-model and human visual inspection successfully avoid rejecting \(H_0\)
-when \(H_0\) is true, contrary to conventional tests. This underscores
-the necessity of visually examining the residual plot.
-
-In the second example, we encounter a more pronounced violation of the
-model, resulting in rejection of \(H_0\) by all three tests. This
-highlights the practicality of the computer vision model, particularly
-for less intricate tasks.
-
-The third example presents a situation where the model deviation is
-non-typical. Here, the computer vision model and human visual inspection
-reject \(H_0\), whereas some commonly used conventional tests do not.
-This emphasizes the benefits of visual inspection and the unique
-advantage of the computer vision model, which, like humans, makes
-decisions based on visual discoveries.
+In this section, we present the performance of the trained computer
+vision model on three example datasets, including the dataset associated
+with the residual plot displaying a ``left-triangle'' shape (Figure
+\ref{fig:false-finding}), the Boston housing dataset
+\citep{harrison1978hedonic}, and the ``dino'' datasets from the from the
+\texttt{datasauRus} package \citep{datasaurus}.
+
+In the first case, both the model and human inspection correctly avoid
+rejecting \(H_0\) when it is true, contrary to conventional tests,
+highlighting the necessity of visual examination. The second case shows
+a clear model violation detected by all tests, demonstrating the model's
+practical use for less intricate tasks. The third case involves an
+unusual deviation where the model and humans reject \(H_0\), but some
+conventional tests do not, illustrating the benefits of visual-based
+decision-making.
 
@@
 
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/false-lineup-1}
+{\centering \includegraphics[width=0.7\linewidth]{paper_files/figure-latex/false-lineup-1}
 
@@
 plots and null plots. This coincides the findings from Figure
-\ref{fig:boston-lineup}, where the true plot exhibiting a ``U'' shape is
-visually distinctive from null plots. If a visual test is conducted by
-human, \(H_0\) will also be rejected.
+\ref{fig:boston-lineup}, where the true residual plot exhibiting a ``U''
+shape is visually distinctive from null plots. If a visual test is
+conducted by human, \(H_0\) will also be rejected.
 
@@
 
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/boston-lineup-1}
+{\centering \includegraphics[width=0.7\linewidth]{paper_files/figure-latex/boston-lineup-1}
 
@@
 
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/dino-lineup-1}
+{\centering \includegraphics[width=0.7\linewidth]{paper_files/figure-latex/dino-lineup-1}
 
@@
 the distance-based statistical testing compares the difference in
-distance to a classical normal linear regression model for the true plot
-and null plots. This comparison is meaningful only if the difference can
-be identified by the distance measure proposed in this paper.
+distance to a classical normal linear regression model for the true
+residual plot and null plots. This comparison is meaningful only if the
+difference can be identified by the distance measure proposed in this
+paper.
 

### 2025-06-16 paper.Rmd
Adjust Examples, Limitation and Future Works and Conclusions sections

@@
 
-In this section, we present the performance of the trained computer vision model on three example datasets, including the dataset associated with the residual plot displaying a "left-triangle" shape (Figure \ref{fig:false-finding}), the Boston housing dataset [@harrison1978hedonic], and the "dino" datasets from the from the `datasauRus` package [@datasaurus].
+In this section, we present the performance of the trained computer vision model on three example datasets, including the dataset associated with the residual plot displaying a "left-triangle" shape, the Boston housing dataset [@harrison1978hedonic], and the "dino" datasets from the from the `datasauRus` package [@datasaurus].
 
@@
 
-In Section \ref{sec-model-introduction}, we presented an example residual plot showcased in Figure \ref{fig:false-finding}, illustrating how humans might misinterpret the "left-triangle" shape as indicative of heteroskedasticity.
-Additionally, the Breusch-Pagan test yielded a rejection with a $p$-value of 0.046, despite the residuals originating from a correctly specified model.
-Figure \ref{fig:false-lineup} offers a lineup for this fitted model, showcasing various degrees of "left-triangle" shape across all residual plots.
-This phenomenon is evidently caused by the skewed distribution of the fitted values.
-Notably, if the residual plot were evaluated through a visual test, it would not be rejected since the true residual plot positioned at 10 can not be distinguished from the others.
-
-Figure \ref{fig:false-check} presents the results of the assessment by the computer vision model.
-Notably, the observed visual signal strength is considerably lower than the 95% sample quantile of the null distribution.
-Moreover, the bootstrapped distribution suggests that it is highly improbable for the fitted model to be misspecified as the majority of bootstrapped fitted models will not be rejected.
-Thus, for this particular fitted model, both the visual test and the computer vision model will not reject $H_0$.
-However, the Breusch-Pagan test will reject $H_0$ because it can not effectively utilize information from null plots.
-
-The attention map at Figure \ref{fig:false-check}B suggests that the estimation is highly influenced by the top-right and bottom-right part of the residual plot, as it forms two vertices of the triangular shape.
-A principal component analysis (PCA) is also performed on the output of the global pooling layer of the computer vision model.
-As mentioned in @simonyan2014very, a computer vision model built upon the convolutional blocks can be viewed as a feature extractor.
-For the $32 \times 32$ model, there are 256 features outputted from the global pooling layer, which can be further used for different visual tasks not limited to distance prediction.
-To see if these features can be effectively used for distinguishing null plots and true residual plot, we linearly project them into the first and second principal components space as shown in Figure \ref{fig:false-check}D.
-It can be observed that because the bootstrapped plots are mostly similar to the null plots, the points drawn in different colors are mixed together.
-The true residual plot is also covered by both the cluster of null plots and cluster of bootstrapped plots.
-This accurately reflects our understanding of Figure \ref{fig:false-lineup}.
+In Section \ref{sec-model-introduction}, we introduced a residual plot (Figure \ref{fig:false-finding}) where the "left-triangle" shape could be misinterpreted as evidence of heteroskedasticity.
+Although the residuals are drawn from a correctly specified model, the Breusch-Pagan test rejects $H_0$ ($p = 0.046$).
+This visual artifact stems from the skewed distribution of fitted values, as illustrated in the lineup in Figure \ref{fig:false-lineup}, where similar "left-triangle" patterns appear across all residual plots.
+Since the true plot (at position 10) is visually indistinguishable from the null plots, a visual test would not reject $H_0$.
+
+The computer vision model yields a consistent conclusion.
+As shown in Figure \ref{fig:false-check}, the estimated distance $\hat{D}$ falls well below the 95% quantile of the null distribution, with a $p$-value of $0.33$.
+The bootstrapped distribution further suggests that rejection of the fitted model is highly unlikely.
+Together, the visual and model-based tests do not reject $H_0$, in contrast to the Breusch-Pagan test, which lacks the context provided by null plots and may be misled by visual artifacts.
+
+To further interpret the model’s output, the attention map in Figure \ref{fig:false-check}B, computed as the gradient of the output with respect to the greyscale input, shows that the top-right and bottom-right corners of the residual plot have the most influence, corresponding to the triangle’s vertices.
+In addition, a principal component analysis (PCA) of the 256 features extracted by the model’s global pooling layer (Figure \ref{fig:false-check}D) shows that null and bootstrapped plots cluster together in the space defined by the first two principal components, and the true residual plot is covered by the cluster.
+This overlap reinforces the conclusion that the true plot is visually indistinct, aligning with our interpretation of the lineup in Figure \ref{fig:false-lineup}.
 
@@
 (D) plot of first two principal components of features extracted from the global pooling layer of the computer vision model.
-', fig.height = 8}
+', fig.height = 8, out.width = "80%"}
 p1 <- my_vi$plot_resid(theme = theme_light()) + ggtitle(glue::glue("(A) Residual plot")) 
@@
 p3 <- my_vi$summary_plot() +
+  annotate("text", x = 2.8, y = 1.25, label = glue::glue("p-value = {format(my_vi$p_value(), digits = 3)}")) +
   ggtitle("(C) Density plot of estimated D", subtitle = glue::glue("Conventional p-value = {format(HETER_MODEL$test(dat)$p_value, digits = 3)}")) +
-  theme(legend.position = "bottom", legend.box = "vertical") +
+  theme(legend.position = "bottom") +
   labs(linetype = "") +
@@
 
-The Boston housing dataset, originally published by @harrison1978hedonic, offers insights into housing in the Boston, Massachusetts area.
-For illustration purposes, we utilize a reduced version from Kaggle, comprising 489 rows and 4 columns: average number of rooms per dwelling (RM), percentage of lower status of the population (LSTAT), pupil-teacher ratio by town (PTRATIO), and median value of owner-occupied homes in $1000's (MEDV).
-In our analysis, MEDV will serve as the response variable, while the other columns will function as predictors in a linear regression model.
-Our primary focus is to detect non-linearity, because the relationships between RM and MEDV or LSTAT and MEDV are non-linear.
-
-Figure \ref{fig:boston-check} displays the residual plot and the assessment conducted by the computer vision model.
-A clear non-linearity pattern resembling a "U" shape is shown in the plot A.
-Furthermore, the RESET test yields a very small $p$-value.
-The estimated distance $\hat{D}$ significantly exceeds $Q_{null}(0.95)$, leading to rejection of $H_0$.
-The bootstrapped distribution also suggests that almost all the bootstrapped fitted models will be rejected, indicating that the fitted model is unlikely to be correctly specified.
-The attention map in plot B suggests the center of the image has higher leverage than other areas, and it is the turning point of the "U" shape.
-The PCA provided in plot D shows two distinct clusters of data points, further underling the visual differences between bootstrapped plots and null plots.
-This coincides the findings from Figure \ref{fig:boston-lineup}, where the true residual plot exhibiting a "U" shape is visually distinctive from null plots.
-If a visual test is conducted by human, $H_0$ will also be rejected.
+The Boston housing dataset, originally published by @harrison1978hedonic, provides insights into housing in the Boston, Massachusetts area.
+For illustration, we use a reduced Kaggle version containing 489 observations and four variables: average number of rooms per dwelling (RM), percentage of lower status population (LSTAT), pupil-teacher ratio by town (PTRATIO), and median home value (MEDV).
+We fit a linear regression model with MEDV as the response and the remaining variables as predictors, focusing on detecting non-linearity, because the relationships between RM and MEDV or LSTAT and MEDV are non-linear.
+
+Figure \ref{fig:boston-check} summarizes the model evaluation.
+The residual plot (Figure \ref{fig:boston-check}A) reveals a distinct "U"-shaped pattern, suggesting non-linearity.
+The RESET test strongly rejects $H_0$ with a very small $p$-value, and the computer vision model yields a large estimated distance $\hat{D}$, well above the 95% quantile of the null distribution ($p\text{-value} = 0.00498$).
+The bootstrapped distribution also shows widespread rejection, indicating the model is likely misspecified.
+The attention map (Figure \ref{fig:boston-check}B) highlights the central region of the residual plot, corresponding to the turning point of the "U", as most influential.
+In the PCA projection (Figure \ref{fig:boston-check}D), bootstrapped and null plots form two distinct clusters, emphasizing the visual separability.
+This aligns with the lineup in Figure \ref{fig:boston-lineup}, where the true plot stands out clearly.
+A human visual test would also likely reject $H_0$ in this case.
 
@@
 p3 <- my_vi$summary_plot() +
+    annotate("text", x = 6, y = 2.25, label = glue::glue("p-value = {format(my_vi$p_value(), digits = 3)}")) +
   ggtitle("(C) Density plot of estimated D", subtitle = glue::glue("Conventional p-value = {format(lmtest::resettest(mod)$p.value, digits = 3)}")) +
-  theme(legend.position = "bottom", legend.box = "vertical") +
+  theme(legend.position = "bottom") +
   labs(linetype = "") +
@@
 
-The computer vision model possesses the capability to detect not only typical issues like non-linearity, heteroskedasticity, and non-normality but also artifact visual patterns resembling real-world objects, as long as they do not appear in null plots.
-These visual patterns can be challenging to categorize in terms of model violations.
-Therefore, we will employ the RESET test, the Breusch-Pagan test, and the Shapiro-Wilk test [@shapiro1965analysis] for comparison.
-
-The "dino" dataset within the `datasauRus` R package exemplifies this scenario.
-With only two columns, x and y, fitting a regression model to this data yields a residual plot resembling a "dinosaur", as displayed in Figure \ref{fig:dino-check}A.
-Unsurprisingly, this distinct residual plot stands out in a lineup, as shown in Figure \ref{fig:dino-lineup}.
-A visual test conducted by humans would undoubtedly reject $H_0$.
-
-According to the residual plot assessment by the computer vision model, $\hat{D}$ exceeds $Q_{null}(0.95)$, warranting a rejection of $H_0$.
-Additionally, most of the bootstrapped fitted models will be rejected, indicating an misspecified model.
-However, both the RESET test and the Breusch-Pagan test yield $p$-values greater than 0.3, leading to a non-rejection of $H_0$.
-Only the Shapiro-Wilk test rejects the normality assumption with a small $p$-value.
-
-More importantly, the attention map in Figure \ref{fig:dino-check}B clearly exhibits a "dinosaur" shape, strongly suggesting that the distance prediction is based on human-perceptible visual patterns.
-The computer vision model effectively captures the contour or outline of the embedded shape, similar to how humans interpret residual plots.
-Additionally, the PCA in Figure \ref{fig:dino-check}D demonstrates that the cluster of bootstrapped plots is positioned at the corner of the cluster of null plots.
-
-In practice, without accessing the residual plot, it would be challenging to identify the artificial pattern of the residuals.
-Moreover, conducting a normality test for a fitted regression model is not always standard practice among analysts.
-Even when performed, violating the normality assumption is sometimes deemed acceptable, especially considering the application of quasi-maximum likelihood estimation in linear regression.
-This example underscores the importance of evaluating residual plots and highlights how the proposed computer vision model can facilitate this process.
+Beyond detecting common issues like non-linearity, heteroskedasticity, and non-normality, the computer vision model can also identify unusual visual artifacts, such as shapes resembling real-world objects, as long as such patterns do not appear in null plots.
+These cases are challenging to categorize using conventional model diagnostics, so we compare the model's output with results from the RESET test, Breusch-Pagan test, and Shapiro-Wilk test [@shapiro1965analysis].
+
+The "dino" dataset from the `datasauRus` R package illustrates this situation.
+It contains only two variables, $x$ and $y$, but fitting a regression model produces a residual plot that clearly resembles a dinosaur (Figure \ref{fig:dino-check}A).
+This pattern is visually distinctive in the lineup shown in Figure \ref{fig:dino-lineup}, where a human visual test would undoubtedly reject $H_0$.
+
+Consistent with this, the computer vision model assigns a distance $\hat{D}$ above the 95% quantile of the null distribution ($p-\text{value} = 0.00498$), leading to rejection of $H_0$.
+The bootstrapped distribution further supports this, with most fitted models also being rejected.
+In contrast, the RESET and Breusch-Pagan tests yield $p$-values above $0.3$, suggesting no violation.
+Only the Shapiro-Wilk test rejects normality with a small $p$-value.
+
+Crucially, the attention map in Figure \ref{fig:dino-check}B highlights the dinosaur shape, indicating that the model’s decision is driven by human-perceptible features.
+The model appears to capture contours and outlines in a manner similar to human visual interpretation.
+In the PCA plot (Figure \ref{fig:dino-check}D), the cluster of bootstrapped plots is positioned at the corner of the cluster of null plots, isolated from the cluster of null plots.
+
+In practice, such visual artifacts would be difficult to detect without inspecting the residual plot directly.
+Furthermore, analysts do not always test for normality, and even when violations are found, they are often deemed acceptable due to the robustness of linear regression under quasi-likelihood theory.
+This example highlights the value of visual diagnostics and the unique role of the proposed computer vision model in supporting them.
 
@@
 The fitted model will be rejected since $\\hat{D} \\geq Q_{null}(0.95)$.
-(D) plot of first two principal components of features extracted from the global pooling layer of the computer vision model.', fig.height = 8}
+(D) plot of first two principal components of features extracted from the global pooling layer of the computer vision model.', fig.height = 8, out.width = "80%"}
 
@@
 p3 <- my_vi$summary_plot() +
+    annotate("text", x = 6, y = 1.25, label = glue::glue("p-value = {format(my_vi$p_value(), digits = 3)}")) +
   ggtitle("(C) Density plot of estimated D", subtitle = glue::glue("RESET test p-value = {format(lmtest::resettest(mod)$p.value, digits = 3)}\n Breusch-Pagan test p-value = {format(lmtest::bptest(mod)$p.value, digits = 3)} \n Shapiro-Wilk test p-value = {format(shapiro.test(mod$residuals)$p.value, digits = 3)}")) +
-  theme(legend.position = "bottom", legend.box = "vertical") +
+  theme(legend.position = "bottom") +
   labs(linetype = "") +
@@
 
-Despite the computer vision model performing well with general cases under the synthetic data generation scheme and the three examples used in this paper, this study has several limitations that could guide future work.
-
-The proposed distance measure assumes that the true model is a classical normal linear regression model, which can be restrictive.
-Although this paper does not address the relaxation of this assumption, there are potential methods to evaluate other types of regression models.
-The most comprehensive approach would be to define a distance measure for each different class of regression model and then train the computer vision model following the methodology described in this paper.
-To accelerate training, one could use the convolutional blocks of our trained model as a feature extractor and perform transfer learning on top of it, as these blocks effectively capture shapes in residual plots.
-Another approach would be to transform the residuals so they are roughly normally distributed and have constant variance.
-If only raw residuals are used, the distance-based statistical testing compares the difference in distance to a classical normal linear regression model for the true residual plot and null plots.
-This comparison is meaningful only if the difference can be identified by the distance measure proposed in this paper.
-
-There are other types of residual plots commonly used in diagnostics, such as residuals vs.
-predictor and quantile-quantile plots.
-In this study, we focused on the most commonly used residual plot as a starting point for exploring the new field of automated visual inference.
-Similarly, we did not explore other, more sophisticated computer vision model architectures and specifications for the same reason.
-While the performance of the computer vision model is acceptable, there is still room for improvement to achieve behavior more closely resembling that of humans interpreting residual plots.
-This may require external survey data or human subject experiment data to understand the fundamental differences between our implementation and human evaluation.
+While the computer vision model performs well under the synthetic data generation scheme and across the three illustrative examples, this study has several limitations that point to directions for future work.
+
+First, the proposed distance measure assumes the true model follows a classical normal linear regression framework, which may be restrictive.
+Although we do not explore relaxation of this assumption here, future work could extend the approach to other types of regression models.
+One possibility is to define a distance measure tailored to each model class and re-train the computer vision model accordingly.
+To reduce computational cost, the convolutional blocks from our current model could be reused via transfer learning, as it is already effective at capturing visual features.
+Alternatively, transforming residuals to approximate normality and constant variance could allow for meaningful comparisons under our current distance measure.
+However, if raw residuals are used, the interpretation remains valid only when the visual differences between the true and null plots are detectable by the proposed distance measure.
+
+Second, this study focused solely on the standard residuals vs.
+fitted values plot, leaving out other common diagnostic plots such as residuals vs.
+predictors and Q-Q plots.
+These alternative plots may offer additional information and could be incorporated into a broader visual testing framework in future research.
+
+Finally, we chose a relatively simple computer vision architecture to establish a baseline for automated visual inference.
+While the current model shows promise, there is room to improve its alignment with human judgment.
+Achieving this may require external data from surveys or controlled experiments to better understand how human observers interpret residual plots and how these judgments differ from the model’s outputs.
 
@@
 
-In this paper, we have introduced a distance measure based on Kullback-Leibler divergence to quantify the disparity between the residual distribution of a fitted classical normal linear regression model and the reference residual distribution assumed under correct model specification.
-This distance measure effectively captures the magnitude of model violations in misspecified models.
-We propose a computer vision model to estimate this distance, utilizing the residual plot of the fitted model as input.
-
-Moreover, the estimated distance enables the development of a formal statistical testing procedure by evaluating a large number of null plots generated from the fitted model.
-Additionally, employing bootstrapping techniques and refitting the regression model allows us to ascertain how frequently the fitted model is considered misspecified if data were repeatedly obtained from the same data generating process.
-
-The trained computer vision model demonstrates strong performance on both the training and test sets, although it exhibits slightly lower performance on residual plots with non-linearity visual patterns compared to other types of violations.
-The statistical tests relying on the estimated distance predicted by the computer vision model exhibit lower sensitivity compared to conventional tests but higher sensitivity compared to visual tests conducted by humans.
-While the estimated distance generally mirrors the strength of the visual signal perceived by humans, there remains scope for further improvement in its performance.
-
-Several examples are provided to showcase the effectiveness of the proposed method across different scenarios, emphasizing the similarity between visual tests and distance-based tests.
-Overall, both visual tests and distance-based tests can be viewed as ensemble of tests, aiming to assess
-any violations of model assumptions collectively.
-In contrast, individual residual diagnostic tests such as the RESET test and the Breusch-Pagan test only evaluate specific violations of model assumptions.
-In practice, selecting an appropriate set of statistical tests for regression diagnostics can be challenging, particularly given the necessity of adjusting the significance level for each test.
-
-Our method holds significant value as it helps alleviate a portion of analysts' workload associated with assessing residual plots.
-While we recommend analysts to continue reading residual plots whenever feasible, as they offer invaluable insights, our approach serves as a valuable tool for automating the diagnostic process or for supplementary purposes when needed.
+In this paper, we have introduced a distance measure based on the Kullback-Leibler divergence to quantify disparity between the residual distribution of a fitted classical normal linear regression model and the reference distribution under correct specification.
+This measure captures the extent of model violations and forms the basis of a computer vision model that estimates the distance using residual plots as input.
+
+The estimated distance supports a formal statistical testing procedure by comparing the true residual plot against null plots generated from the fitted model.
+Through bootstrapping and model refitting, we also evaluate how often a correctly specified model would be flagged as misspecified under repeated sampling.
+
+The trained computer vision model performs well on both training and test data, though its performance is somewhat lower for non-linear patterns.
+Statistical tests based on the model’s predicted distance show lower sensitivity than conventional tests but higher sensitivity than human visual tests.
+Importantly, the predicted distance broadly aligns with the strength of visual signals perceived by humans, though further improvements are possible.
+
+We demonstrate the method across several scenarios, highlighting its effectiveness and its conceptual alignment with visual inference.
+Both visual and distance-based tests act as omnibus procedures, detecting a wide range of model violations collectively.
+In contrast, conventional tests such as RESET and Breusch-Pagan target specific assumptions, making them less flexible and potentially harder to combine without adjusting for multiple testing.
+
+Overall, our approach offers a practical tool to support regression diagnostics by reducing the manual effort required for residual plot inspection.
+While we recommend analysts to continue reading residual plots whenever feasible, our method can augment the process, either as an automated screening tool or a supplement to human judgment.
 

### 2025-06-16 paper.tex
Adjust Examples, Limitation and Future Works and Conclusions sections

@@
 vision model on three example datasets, including the dataset associated
-with the residual plot displaying a ``left-triangle'' shape (Figure
-\ref{fig:false-finding}), the Boston housing dataset
-\citep{harrison1978hedonic}, and the ``dino'' datasets from the from the
-\texttt{datasauRus} package \citep{datasaurus}.
+with the residual plot displaying a ``left-triangle'' shape, the Boston
+housing dataset \citep{harrison1978hedonic}, and the ``dino'' datasets
+from the from the \texttt{datasauRus} package \citep{datasaurus}.
 
@@
 
-In Section \ref{sec-model-introduction}, we presented an example
-residual plot showcased in Figure \ref{fig:false-finding}, illustrating
-how humans might misinterpret the ``left-triangle'' shape as indicative
-of heteroskedasticity. Additionally, the Breusch-Pagan test yielded a
-rejection with a \(p\)-value of 0.046, despite the residuals originating
-from a correctly specified model. Figure \ref{fig:false-lineup} offers a
-lineup for this fitted model, showcasing various degrees of
-``left-triangle'' shape across all residual plots. This phenomenon is
-evidently caused by the skewed distribution of the fitted values.
-Notably, if the residual plot were evaluated through a visual test, it
-would not be rejected since the true residual plot positioned at 10 can
-not be distinguished from the others.
-
-Figure \ref{fig:false-check} presents the results of the assessment by
-the computer vision model. Notably, the observed visual signal strength
-is considerably lower than the 95\% sample quantile of the null
-distribution. Moreover, the bootstrapped distribution suggests that it
-is highly improbable for the fitted model to be misspecified as the
-majority of bootstrapped fitted models will not be rejected. Thus, for
-this particular fitted model, both the visual test and the computer
-vision model will not reject \(H_0\). However, the Breusch-Pagan test
-will reject \(H_0\) because it can not effectively utilize information
-from null plots.
-
-The attention map at Figure \ref{fig:false-check}B suggests that the
-estimation is highly influenced by the top-right and bottom-right part
-of the residual plot, as it forms two vertices of the triangular shape.
-A principal component analysis (PCA) is also performed on the output of
-the global pooling layer of the computer vision model. As mentioned in
-\citet{simonyan2014very}, a computer vision model built upon the
-convolutional blocks can be viewed as a feature extractor. For the
-\(32 \times 32\) model, there are 256 features outputted from the global
-pooling layer, which can be further used for different visual tasks not
-limited to distance prediction. To see if these features can be
-effectively used for distinguishing null plots and true residual plot,
-we linearly project them into the first and second principal components
-space as shown in Figure \ref{fig:false-check}D. It can be observed that
-because the bootstrapped plots are mostly similar to the null plots, the
-points drawn in different colors are mixed together. The true residual
-plot is also covered by both the cluster of null plots and cluster of
-bootstrapped plots. This accurately reflects our understanding of Figure
-\ref{fig:false-lineup}.
+In Section \ref{sec-model-introduction}, we introduced a residual plot
+(Figure \ref{fig:false-finding}) where the ``left-triangle'' shape could
+be misinterpreted as evidence of heteroskedasticity. Although the
+residuals are drawn from a correctly specified model, the Breusch-Pagan
+test rejects \(H_0\) (\(p = 0.046\)). This visual artifact stems from
+the skewed distribution of fitted values, as illustrated in the lineup
+in Figure \ref{fig:false-lineup}, where similar ``left-triangle''
+patterns appear across all residual plots. Since the true plot (at
+position 10) is visually indistinguishable from the null plots, a visual
+test would not reject \(H_0\).
+
+The computer vision model yields a consistent conclusion. As shown in
+Figure \ref{fig:false-check}, the estimated distance \(\hat{D}\) falls
+well below the 95\% quantile of the null distribution, with a
+\(p\)-value of \(0.33\). The bootstrapped distribution further suggests
+that rejection of the fitted model is highly unlikely. Together, the
+visual and model-based tests do not reject \(H_0\), in contrast to the
+Breusch-Pagan test, which lacks the context provided by null plots and
+may be misled by visual artifacts.
+
+To further interpret the model's output, the attention map in Figure
+\ref{fig:false-check}B, computed as the gradient of the output with
+respect to the greyscale input, shows that the top-right and
+bottom-right corners of the residual plot have the most influence,
+corresponding to the triangle's vertices. In addition, a principal
+component analysis (PCA) of the 256 features extracted by the model's
+global pooling layer (Figure \ref{fig:false-check}D) shows that null and
+bootstrapped plots cluster together in the space defined by the first
+two principal components, and the true residual plot is covered by the
+cluster. This overlap reinforces the conclusion that the true plot is
+visually indistinct, aligning with our interpretation of the lineup in
+Figure \ref{fig:false-lineup}.
 
@@
 
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/false-check-1}
+{\centering \includegraphics[width=0.8\linewidth]{paper_files/figure-latex/false-check-1}
 
@@
 The Boston housing dataset, originally published by
-\citet{harrison1978hedonic}, offers insights into housing in the Boston,
-Massachusetts area. For illustration purposes, we utilize a reduced
-version from Kaggle, comprising 489 rows and 4 columns: average number
-of rooms per dwelling (RM), percentage of lower status of the population
-(LSTAT), pupil-teacher ratio by town (PTRATIO), and median value of
-owner-occupied homes in \$1000's (MEDV). In our analysis, MEDV will
-serve as the response variable, while the other columns will function as
-predictors in a linear regression model. Our primary focus is to detect
+\citet{harrison1978hedonic}, provides insights into housing in the
+Boston, Massachusetts area. For illustration, we use a reduced Kaggle
+version containing 489 observations and four variables: average number
+of rooms per dwelling (RM), percentage of lower status population
+(LSTAT), pupil-teacher ratio by town (PTRATIO), and median home value
+(MEDV). We fit a linear regression model with MEDV as the response and
+the remaining variables as predictors, focusing on detecting
 non-linearity, because the relationships between RM and MEDV or LSTAT
@@
 
-Figure \ref{fig:boston-check} displays the residual plot and the
-assessment conducted by the computer vision model. A clear non-linearity
-pattern resembling a ``U'' shape is shown in the plot A. Furthermore,
-the RESET test yields a very small \(p\)-value. The estimated distance
-\(\hat{D}\) significantly exceeds \(Q_{null}(0.95)\), leading to
-rejection of \(H_0\). The bootstrapped distribution also suggests that
-almost all the bootstrapped fitted models will be rejected, indicating
-that the fitted model is unlikely to be correctly specified. The
-attention map in plot B suggests the center of the image has higher
-leverage than other areas, and it is the turning point of the ``U''
-shape. The PCA provided in plot D shows two distinct clusters of data
-points, further underling the visual differences between bootstrapped
-plots and null plots. This coincides the findings from Figure
-\ref{fig:boston-lineup}, where the true residual plot exhibiting a ``U''
-shape is visually distinctive from null plots. If a visual test is
-conducted by human, \(H_0\) will also be rejected.
+Figure \ref{fig:boston-check} summarizes the model evaluation. The
+residual plot (Figure \ref{fig:boston-check}A) reveals a distinct
+``U''-shaped pattern, suggesting non-linearity. The RESET test strongly
+rejects \(H_0\) with a very small \(p\)-value, and the computer vision
+model yields a large estimated distance \(\hat{D}\), well above the 95\%
+quantile of the null distribution (\(p\text{-value} = 0.00498\)). The
+bootstrapped distribution also shows widespread rejection, indicating
+the model is likely misspecified. The attention map (Figure
+\ref{fig:boston-check}B) highlights the central region of the residual
+plot, corresponding to the turning point of the ``U'', as most
+influential. In the PCA projection (Figure \ref{fig:boston-check}D),
+bootstrapped and null plots form two distinct clusters, emphasizing the
+visual separability. This aligns with the lineup in Figure
+\ref{fig:boston-lineup}, where the true plot stands out clearly. A human
+visual test would also likely reject \(H_0\) in this case.
 
@@
 
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/boston-check-1}
+{\centering \includegraphics[width=0.8\linewidth]{paper_files/figure-latex/boston-check-1}
 
@@
 
-The computer vision model possesses the capability to detect not only
-typical issues like non-linearity, heteroskedasticity, and non-normality
-but also artifact visual patterns resembling real-world objects, as long
-as they do not appear in null plots. These visual patterns can be
-challenging to categorize in terms of model violations. Therefore, we
-will employ the RESET test, the Breusch-Pagan test, and the Shapiro-Wilk
-test \citep{shapiro1965analysis} for comparison.
-
-The ``dino'' dataset within the \texttt{datasauRus} R package
-exemplifies this scenario. With only two columns, x and y, fitting a
-regression model to this data yields a residual plot resembling a
-``dinosaur'', as displayed in Figure \ref{fig:dino-check}A.
-Unsurprisingly, this distinct residual plot stands out in a lineup, as
-shown in Figure \ref{fig:dino-lineup}. A visual test conducted by humans
-would undoubtedly reject \(H_0\).
-
-According to the residual plot assessment by the computer vision model,
-\(\hat{D}\) exceeds \(Q_{null}(0.95)\), warranting a rejection of
-\(H_0\). Additionally, most of the bootstrapped fitted models will be
-rejected, indicating an misspecified model. However, both the RESET test
-and the Breusch-Pagan test yield \(p\)-values greater than 0.3, leading
-to a non-rejection of \(H_0\). Only the Shapiro-Wilk test rejects the
-normality assumption with a small \(p\)-value.
-
-More importantly, the attention map in Figure \ref{fig:dino-check}B
-clearly exhibits a ``dinosaur'' shape, strongly suggesting that the
-distance prediction is based on human-perceptible visual patterns. The
-computer vision model effectively captures the contour or outline of the
-embedded shape, similar to how humans interpret residual plots.
-Additionally, the PCA in Figure \ref{fig:dino-check}D demonstrates that
-the cluster of bootstrapped plots is positioned at the corner of the
-cluster of null plots.
-
-In practice, without accessing the residual plot, it would be
-challenging to identify the artificial pattern of the residuals.
-Moreover, conducting a normality test for a fitted regression model is
-not always standard practice among analysts. Even when performed,
-violating the normality assumption is sometimes deemed acceptable,
-especially considering the application of quasi-maximum likelihood
-estimation in linear regression. This example underscores the importance
-of evaluating residual plots and highlights how the proposed computer
-vision model can facilitate this process.
+Beyond detecting common issues like non-linearity, heteroskedasticity,
+and non-normality, the computer vision model can also identify unusual
+visual artifacts, such as shapes resembling real-world objects, as long
+as such patterns do not appear in null plots. These cases are
+challenging to categorize using conventional model diagnostics, so we
+compare the model's output with results from the RESET test,
+Breusch-Pagan test, and Shapiro-Wilk test \citep{shapiro1965analysis}.
+
+The ``dino'' dataset from the \texttt{datasauRus} R package illustrates
+this situation. It contains only two variables, \(x\) and \(y\), but
+fitting a regression model produces a residual plot that clearly
+resembles a dinosaur (Figure \ref{fig:dino-check}A). This pattern is
+visually distinctive in the lineup shown in Figure
+\ref{fig:dino-lineup}, where a human visual test would undoubtedly
+reject \(H_0\).
+
+Consistent with this, the computer vision model assigns a distance
+\(\hat{D}\) above the 95\% quantile of the null distribution
+(\(p-\text{value} = 0.00498\)), leading to rejection of \(H_0\). The
+bootstrapped distribution further supports this, with most fitted models
+also being rejected. In contrast, the RESET and Breusch-Pagan tests
+yield \(p\)-values above \(0.3\), suggesting no violation. Only the
+Shapiro-Wilk test rejects normality with a small \(p\)-value.
+
+Crucially, the attention map in Figure \ref{fig:dino-check}B highlights
+the dinosaur shape, indicating that the model's decision is driven by
+human-perceptible features. The model appears to capture contours and
+outlines in a manner similar to human visual interpretation. In the PCA
+plot (Figure \ref{fig:dino-check}D), the cluster of bootstrapped plots
+is positioned at the corner of the cluster of null plots, isolated from
+the cluster of null plots.
+
+In practice, such visual artifacts would be difficult to detect without
+inspecting the residual plot directly. Furthermore, analysts do not
+always test for normality, and even when violations are found, they are
+often deemed acceptable due to the robustness of linear regression under
+quasi-likelihood theory. This example highlights the value of visual
+diagnostics and the unique role of the proposed computer vision model in
+supporting them.
 
@@
 
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/dino-check-1}
+{\centering \includegraphics[width=0.8\linewidth]{paper_files/figure-latex/dino-check-1}
 
@@
 
-Despite the computer vision model performing well with general cases
-under the synthetic data generation scheme and the three examples used
-in this paper, this study has several limitations that could guide
-future work.
-
-The proposed distance measure assumes that the true model is a classical
-normal linear regression model, which can be restrictive. Although this
-paper does not address the relaxation of this assumption, there are
-potential methods to evaluate other types of regression models. The most
-comprehensive approach would be to define a distance measure for each
-different class of regression model and then train the computer vision
-model following the methodology described in this paper. To accelerate
-training, one could use the convolutional blocks of our trained model as
-a feature extractor and perform transfer learning on top of it, as these
-blocks effectively capture shapes in residual plots. Another approach
-would be to transform the residuals so they are roughly normally
-distributed and have constant variance. If only raw residuals are used,
-the distance-based statistical testing compares the difference in
-distance to a classical normal linear regression model for the true
-residual plot and null plots. This comparison is meaningful only if the
-difference can be identified by the distance measure proposed in this
-paper.
-
-There are other types of residual plots commonly used in diagnostics,
-such as residuals vs.~predictor and quantile-quantile plots. In this
-study, we focused on the most commonly used residual plot as a starting
-point for exploring the new field of automated visual inference.
-Similarly, we did not explore other, more sophisticated computer vision
-model architectures and specifications for the same reason. While the
-performance of the computer vision model is acceptable, there is still
-room for improvement to achieve behavior more closely resembling that of
-humans interpreting residual plots. This may require external survey
-data or human subject experiment data to understand the fundamental
-differences between our implementation and human evaluation.
+While the computer vision model performs well under the synthetic data
+generation scheme and across the three illustrative examples, this study
+has several limitations that point to directions for future work.
+
+First, the proposed distance measure assumes the true model follows a
+classical normal linear regression framework, which may be restrictive.
+Although we do not explore relaxation of this assumption here, future
+work could extend the approach to other types of regression models. One
+possibility is to define a distance measure tailored to each model class
+and re-train the computer vision model accordingly. To reduce
+computational cost, the convolutional blocks from our current model
+could be reused via transfer learning, as it is already effective at
+capturing visual features. Alternatively, transforming residuals to
+approximate normality and constant variance could allow for meaningful
+comparisons under our current distance measure. However, if raw
+residuals are used, the interpretation remains valid only when the
+visual differences between the true and null plots are detectable by the
+proposed distance measure.
+
+Second, this study focused solely on the standard residuals vs.~fitted
+values plot, leaving out other common diagnostic plots such as residuals
+vs.~predictors and Q-Q plots. These alternative plots may offer
+additional information and could be incorporated into a broader visual
+testing framework in future research.
+
+Finally, we chose a relatively simple computer vision architecture to
+establish a baseline for automated visual inference. While the current
+model shows promise, there is room to improve its alignment with human
+judgment. Achieving this may require external data from surveys or
+controlled experiments to better understand how human observers
+interpret residual plots and how these judgments differ from the model's
+outputs.
 
@@
 
-In this paper, we have introduced a distance measure based on
-Kullback-Leibler divergence to quantify the disparity between the
-residual distribution of a fitted classical normal linear regression
-model and the reference residual distribution assumed under correct
-model specification. This distance measure effectively captures the
-magnitude of model violations in misspecified models. We propose a
-computer vision model to estimate this distance, utilizing the residual
-plot of the fitted model as input.
-
-Moreover, the estimated distance enables the development of a formal
-statistical testing procedure by evaluating a large number of null plots
-generated from the fitted model. Additionally, employing bootstrapping
-techniques and refitting the regression model allows us to ascertain how
-frequently the fitted model is considered misspecified if data were
-repeatedly obtained from the same data generating process.
-
-The trained computer vision model demonstrates strong performance on
-both the training and test sets, although it exhibits slightly lower
-performance on residual plots with non-linearity visual patterns
-compared to other types of violations. The statistical tests relying on
-the estimated distance predicted by the computer vision model exhibit
-lower sensitivity compared to conventional tests but higher sensitivity
-compared to visual tests conducted by humans. While the estimated
-distance generally mirrors the strength of the visual signal perceived
-by humans, there remains scope for further improvement in its
-performance.
-
-Several examples are provided to showcase the effectiveness of the
-proposed method across different scenarios, emphasizing the similarity
-between visual tests and distance-based tests. Overall, both visual
-tests and distance-based tests can be viewed as ensemble of tests,
-aiming to assess any violations of model assumptions collectively. In
-contrast, individual residual diagnostic tests such as the RESET test
-and the Breusch-Pagan test only evaluate specific violations of model
-assumptions. In practice, selecting an appropriate set of statistical
-tests for regression diagnostics can be challenging, particularly given
-the necessity of adjusting the significance level for each test.
-
-Our method holds significant value as it helps alleviate a portion of
-analysts' workload associated with assessing residual plots. While we
-recommend analysts to continue reading residual plots whenever feasible,
-as they offer invaluable insights, our approach serves as a valuable
-tool for automating the diagnostic process or for supplementary purposes
-when needed.
+In this paper, we have introduced a distance measure based on the
+Kullback-Leibler divergence to quantify disparity between the residual
+distribution of a fitted classical normal linear regression model and
+the reference distribution under correct specification. This measure
+captures the extent of model violations and forms the basis of a
+computer vision model that estimates the distance using residual plots
+as input.
+
+The estimated distance supports a formal statistical testing procedure
+by comparing the true residual plot against null plots generated from
+the fitted model. Through bootstrapping and model refitting, we also
+evaluate how often a correctly specified model would be flagged as
+misspecified under repeated sampling.
+
+The trained computer vision model performs well on both training and
+test data, though its performance is somewhat lower for non-linear
+patterns. Statistical tests based on the model's predicted distance show
+lower sensitivity than conventional tests but higher sensitivity than
+human visual tests. Importantly, the predicted distance broadly aligns
+with the strength of visual signals perceived by humans, though further
+improvements are possible.
+
+We demonstrate the method across several scenarios, highlighting its
+effectiveness and its conceptual alignment with visual inference. Both
+visual and distance-based tests act as omnibus procedures, detecting a
+wide range of model violations collectively. In contrast, conventional
+tests such as RESET and Breusch-Pagan target specific assumptions,
+making them less flexible and potentially harder to combine without
+adjusting for multiple testing.
+
+Overall, our approach offers a practical tool to support regression
+diagnostics by reducing the manual effort required for residual plot
+inspection. While we recommend analysts to continue reading residual
+plots whenever feasible, our method can augment the process, either as
+an automated screening tool or a supplement to human judgment.
 

### 2025-06-17 appendix.Rmd
Add layer info to appendix

@@
 
+## Neural Network Layers Used in the Study
+
+This study used seven types of neural network layers, all of which are standard components frequently found in modern deep learning models.
+These layers are well-documented in textbooks like @goodfellow2016deep and @chollet2021deep, which offer thorough explanations and mathematical insights.
+In this section, we will offer a concise overview of these layers, drawing primarily from the insights provided in @goodfellow2016deep.
+
+### Dense Layer
+
+The Dense layer, also known as the fully-connected layer, is the fundamental unit in neural networks.
+It conducts a matrix multiplication operation between the input matrix $\boldsymbol{I}$ and a weight matrix $\boldsymbol{W}$ to generate the output matrix $\boldsymbol{O}$, which can be written as
+
+$$\boldsymbol{O} = \boldsymbol{I}\boldsymbol{W} + b,$$
+
+where $b$ is the intercept.
+
+### ReLu Layer
+
+The ReLU layer, short for rectified linear unit, is an element-wise non-linear function introduced by @nair2010rectified.
+It sets the output elements to zero if the corresponding input element is negative; otherwise, it retains the original input.
+Mathematically, it can be expressed as:
+
+$$\boldsymbol{O}(i,j) = max\{0, \boldsymbol{I}(i,j)\},$$
+
+where $\boldsymbol{O}(i,j)$ is the $i$th row and $j$th column entry of matrix $\boldsymbol{O}$, and $\boldsymbol{I}(i,j)$ is the $i$th row and $j$th column entry of matrix $\boldsymbol{I}$.
+
+### Convolutional Layer
+
+In Dense layers, matrix multiplication leads to each output unit interacting with every input unit, whereas convolutional layers operate differently with sparse interactions.
+An output unit in a convolutional layer is connected solely to a subset of input units, and the weight is shared across all input units.
+Achieving this involves using a kernel, typically a small square matrix, to conduct matrix multiplication across all input units.
+Precisely, this concept can be formulated as:
+
+$$\boldsymbol{O}(i, j) = \sum_m\sum_n\boldsymbol{I}(i - m, j - n)K(m, n),$$
+
+where $m$ and $n$ are the row and columns indices of the kernel $K$.
+
+If there are multiple kernels used in one covolutional layer, then each kernel will have its own weights and the output will be a three-dimensional tensor, where the length of the third channel is the number of kernels.
+
+### Pooling Layer
+
+A pooling layer substitutes the input at a specific location with a summary statistic derived from nearby input units.
+Typically, there are two types of pooling layers: max pooling and average pooling.
+Max pooling computes the maximum value within a rectangular neighborhood, while average pooling calculates their average.
+Pooling layers helps making the representation approximately invariant to minor translations of the input.
+The output matrix of a pooling layer is approximately $s$ times smaller than the input matrix, where $s$ represents the length of the rectangular area.
+A max pooling layer can be formulated as:
+
+$$\boldsymbol{O}(i, j) = \underset{m,n}{\max} \boldsymbol{I}(si + m,sj+n).$$
+
+### Global Pooling Layer
+
+A global pooling layer condenses an input matrix into a scalar value by either extracting the maximum or computing the average across all elements.
+This layer acts as a crucial link between the convolutional structure and the subsequent dense layers in a neural network architecture.
+When convolutional layers uses multiple kernels, the output becomes a three-dimensional tensor with numerous channels.
+In this scenario, the global pooling layer treats each channel individually, much like distinct features in a conventional classifier.
+This approach facilitates the extraction of essential features from complex data representations, enhancing the network's ability to learn meaningful patterns.
+A global max pooling layer can be formulated as
+
+$$O(i, j) = \underset{m,n,k}{\max} I(si + m,sj+n,k),$$
+where $k$ is the kernel index.
+
+### Batch Normalization Layer
+
+Batch normalization is a method of adaptive reparametrization.
+One of the issues it adjusts is the simultaneous update of parameters in different layers, especially for network with a large number layers.
+At training time, the batch normalization layer normalizes the input matrix $I$ using the formula
+
+$$\boldsymbol{O} = \frac{\boldsymbol{I} - \boldsymbol{\mu}_I}{\boldsymbol{\sigma}_I},$$
+
+where $\boldsymbol{\mu}_I$ and $\boldsymbol{\sigma}_I$ are the mean and the standard deviation of each unit respectively.
+
+It reparametrizes the model to make some units always be standardized by definition, such that the model training is stabilized.
+At inference time, $\boldsymbol{\mu}_I$ and $\boldsymbol{\sigma}_I$ are usually replaced with the running mean and running average obtained during training.
+
+### Dropout Layer
+
+Dropout is a computationally inexpensive way to apply regularization on neural network.
+For each input unit, it randomly sets to be zero during training, effectively training a large number of subnetworks simultaneously, but these subnetworks share weights and each will only be trained for a single steps in a large network.
+It is essentially a different implementation of the bagging algorithm.
+Mathematically, it is formulated as
+
+$$\boldsymbol{O}(i,j) = \boldsymbol{D}(i,j)\boldsymbol{I}(i,j),$$
+
+where $\boldsymbol{D}(i,j) \sim B(1, p)$ and $p$ is a hyperparameter that can be tuned.
+
 # Model Training {#sec-model-training}

### 2025-06-17 appendix.tex
Add layer info to appendix

@@
 
+\subsection{Neural Network Layers Used in the
+Study}\label{neural-network-layers-used-in-the-study}
+
+This study used seven types of neural network layers, all of which are
+standard components frequently found in modern deep learning models.
+These layers are well-documented in textbooks like
+\citet{goodfellow2016deep} and \citet{chollet2021deep}, which offer
+thorough explanations and mathematical insights. In this section, we
+will offer a concise overview of these layers, drawing primarily from
+the insights provided in \citet{goodfellow2016deep}.
+
+\subsubsection{Dense Layer}\label{dense-layer}
+
+The Dense layer, also known as the fully-connected layer, is the
+fundamental unit in neural networks. It conducts a matrix multiplication
+operation between the input matrix \(\boldsymbol{I}\) and a weight
+matrix \(\boldsymbol{W}\) to generate the output matrix
+\(\boldsymbol{O}\), which can be written as
+
+\[\boldsymbol{O} = \boldsymbol{I}\boldsymbol{W} + b,\]
+
+where \(b\) is the intercept.
+
+\subsubsection{ReLu Layer}\label{relu-layer}
+
+The ReLU layer, short for rectified linear unit, is an element-wise
+non-linear function introduced by \citet{nair2010rectified}. It sets the
+output elements to zero if the corresponding input element is negative;
+otherwise, it retains the original input. Mathematically, it can be
+expressed as:
+
+\[\boldsymbol{O}(i,j) = max\{0, \boldsymbol{I}(i,j)\},\]
+
+where \(\boldsymbol{O}(i,j)\) is the \(i\)th row and \(j\)th column
+entry of matrix \(\boldsymbol{O}\), and \(\boldsymbol{I}(i,j)\) is the
+\(i\)th row and \(j\)th column entry of matrix \(\boldsymbol{I}\).
+
+\subsubsection{Convolutional Layer}\label{convolutional-layer}
+
+In Dense layers, matrix multiplication leads to each output unit
+interacting with every input unit, whereas convolutional layers operate
+differently with sparse interactions. An output unit in a convolutional
+layer is connected solely to a subset of input units, and the weight is
+shared across all input units. Achieving this involves using a kernel,
+typically a small square matrix, to conduct matrix multiplication across
+all input units. Precisely, this concept can be formulated as:
+
+\[\boldsymbol{O}(i, j) = \sum_m\sum_n\boldsymbol{I}(i - m, j - n)K(m, n),\]
+
+where \(m\) and \(n\) are the row and columns indices of the kernel
+\(K\).
+
+If there are multiple kernels used in one covolutional layer, then each
+kernel will have its own weights and the output will be a
+three-dimensional tensor, where the length of the third channel is the
+number of kernels.
+
+\subsubsection{Pooling Layer}\label{pooling-layer}
+
+A pooling layer substitutes the input at a specific location with a
+summary statistic derived from nearby input units. Typically, there are
+two types of pooling layers: max pooling and average pooling. Max
+pooling computes the maximum value within a rectangular neighborhood,
+while average pooling calculates their average. Pooling layers helps
+making the representation approximately invariant to minor translations
+of the input. The output matrix of a pooling layer is approximately
+\(s\) times smaller than the input matrix, where \(s\) represents the
+length of the rectangular area. A max pooling layer can be formulated
+as:
+
+\[\boldsymbol{O}(i, j) = \underset{m,n}{\max} \boldsymbol{I}(si + m,sj+n).\]
+
+\subsubsection{Global Pooling Layer}\label{global-pooling-layer}
+
+A global pooling layer condenses an input matrix into a scalar value by
+either extracting the maximum or computing the average across all
+elements. This layer acts as a crucial link between the convolutional
+structure and the subsequent dense layers in a neural network
+architecture. When convolutional layers uses multiple kernels, the
+output becomes a three-dimensional tensor with numerous channels. In
+this scenario, the global pooling layer treats each channel
+individually, much like distinct features in a conventional classifier.
+This approach facilitates the extraction of essential features from
+complex data representations, enhancing the network's ability to learn
+meaningful patterns. A global max pooling layer can be formulated as
+
+\[O(i, j) = \underset{m,n,k}{\max} I(si + m,sj+n,k),\] where \(k\) is
+the kernel index.
+
+\subsubsection{Batch Normalization
+Layer}\label{batch-normalization-layer}
+
+Batch normalization is a method of adaptive reparametrization. One of
+the issues it adjusts is the simultaneous update of parameters in
+different layers, especially for network with a large number layers. At
+training time, the batch normalization layer normalizes the input matrix
+\(I\) using the formula
+
+\[\boldsymbol{O} = \frac{\boldsymbol{I} - \boldsymbol{\mu}_I}{\boldsymbol{\sigma}_I},\]
+
+where \(\boldsymbol{\mu}_I\) and \(\boldsymbol{\sigma}_I\) are the mean
+and the standard deviation of each unit respectively.
+
+It reparametrizes the model to make some units always be standardized by
+definition, such that the model training is stabilized. At inference
+time, \(\boldsymbol{\mu}_I\) and \(\boldsymbol{\sigma}_I\) are usually
+replaced with the running mean and running average obtained during
+training.
+
+\subsubsection{Dropout Layer}\label{dropout-layer}
+
+Dropout is a computationally inexpensive way to apply regularization on
+neural network. For each input unit, it randomly sets to be zero during
+training, effectively training a large number of subnetworks
+simultaneously, but these subnetworks share weights and each will only
+be trained for a single steps in a large network. It is essentially a
+different implementation of the bagging algorithm. Mathematically, it is
+formulated as
+
+\[\boldsymbol{O}(i,j) = \boldsymbol{D}(i,j)\boldsymbol{I}(i,j),\]
+
+where \(\boldsymbol{D}(i,j) \sim B(1, p)\) and \(p\) is a hyperparameter
+that can be tuned.
+
 \section{Model Training}\label{sec-model-training}

### 2025-06-17 paper.Rmd
Combine three lineups into one

@@
 It is trained to predict a distance measure that quantifies the disparity between the residual distribution of a fitted classical normal linear regression model and the reference distribution, based on Kullback-Leibler divergence.
-From extensive simulation studies, the computer vision model exhibits lower sensitivity than conventional tests but higher sensitivity than human visual tests.
-It is slightly less effective on non-linearity patterns.
-Several examples from classical papers and contemporary data illustrate the new procedures, highlighting its usefulness in automating the diagnostic process and supplementing existing methods.
+Evaluation using data from a prior human-subject experiment shows the model exhibits lower sensitivity than conventional tests but higher sensitivity than human visual tests.
+Several examples from classical papers and contemporary data demonstrate the method's paractical utility in automating residual diagnostics.
 keywords: |
@@
 Plotting residuals is commonly regarded as a standard practice in linear regression diagnostics [@belsley1980regression; @cook1982residuals].
-This visual assessment plays a crucial role in identifying whether model assumptions, such as linearity, homoscedasticity, and normality, are reasonable.
+This visual assessment plays a crucial role in identifying whether key model assumptions, such as linearity, homoscedasticity, and normality, are reasonable.
 It also helps in understanding the goodness of fit and various unexpected characteristics of the model.
@@
 
-The utilization of computers to interpret data plots has a rich history, with early efforts such as "Scagnostics" by @tukey1985computer, a set of numerical statistics that summarize features of scatter plots.
+Efforts to automate plot interpretation date back to early work like Tukey’s "Scagnostics" [@tukey1985computer], which is a set of numerical statistics that summarize features of scatter plots.
 @wilkinson2005graph expanded on this work, introducing scagnostics based on computable measures applied to planar proximity graphs.
 These measures, including, but not limited to, "Outlying", "Skinny", "Stringy", "Straight", "Monotonic", "Skewed", "Clumpy", and "Striated", aimed to characterize outliers, shape, density, trend, coherence and other characteristics of the data.
-While this approach has been inspiring, there is a recognition [@buja2009statistical] that it may not capture all the necessary visual features that differentiate true residual plots from null plots.
-A more promising alternative entails enabling machines to learn the function for extracting visual features from residual plots.
-Essentially, this means empowering computers to discern the crucial visual features for residual diagnostics and determining the method to extract them.
+However, as noted by @buja2009statistical, such measures may not capture the full range of visual patterns necessary for residual diagnostics.
+A more promising direction lies in enabling computers to learn the relevant visual features directly mirroring how humans intuitively interpret plots.
 
@@
 These image similarity metrics are tailored for image comparison in vastly different tasks to evaluating data plots, where only essential plot elements require assessment [@chowdhury2018measuring].
-We can alternatively define a notion of distance by integrating key plot elements (instead of key perception features like luminance, contrast, and structure), such as those captured by scagnostics mentioned in Section \ref{sec-model-introduction}, but the functional form still needs to be carefully refined to accurately reflect the extent of the violations.
-
-In this section, we introduce a distance measure between a true residual plot and a theoretically 'good' residual plot.
+A different direction is to define a notion of distance by integrating key plot elements (instead of key perception features like luminance, contrast, and structure), such as those captured by scagnostics (see Section \ref{sec-model-introduction}), but the functional form must be carefully tailored to reflect model violations meaningfully.
+
+In this section, we introduce a distance measure between a true residual plot and a theoretically "good" residual plot.
 This measure quantifies the divergence between the residual distribution of a given fitted regression model and that of a correctly specified model.
@@
 The model begins with an input layer of shape $n \times h \times w \times 3$, capable of handling $n$ RGB images.
-This is followed by a grayscale conversion layer utilizing the luma formula under the Rec.
+This is followed by a grayscale conversion layer using the luma formula under the Rec.
 601 standard [@series2011studio], which converts the color image to grayscale.
@@
 Although the residuals are drawn from a correctly specified model, the Breusch-Pagan test rejects $H_0$ ($p = 0.046$).
-This visual artifact stems from the skewed distribution of fitted values, as illustrated in the lineup in Figure \ref{fig:false-lineup}, where similar "left-triangle" patterns appear across all residual plots.
+This visual artifact stems from the skewed distribution of fitted values, as illustrated in the lineup in Figure \ref{fig:lineup}A, where similar "left-triangle" patterns appear across all residual plots.
 Since the true plot (at position 10) is visually indistinguishable from the null plots, a visual test would not reject $H_0$.
@@
 In addition, a principal component analysis (PCA) of the 256 features extracted by the model’s global pooling layer (Figure \ref{fig:false-check}D) shows that null and bootstrapped plots cluster together in the space defined by the first two principal components, and the true residual plot is covered by the cluster.
-This overlap reinforces the conclusion that the true plot is visually indistinct, aligning with our interpretation of the lineup in Figure \ref{fig:false-lineup}.
+This overlap reinforces the conclusion that the true plot is visually indistinct, aligning with our interpretation of the lineup.
 
@@
                     result)
-result %>%
+l1 <- result %>%
   VI_MODEL$plot_lineup(remove_grid_line = TRUE,
                        remove_axis = TRUE,
-                       theme = theme_light())
+                       theme = theme_light()) +
+  ggtitle("(B) Lineup for left-triangle")
 ```
@@
 In the PCA projection (Figure \ref{fig:boston-check}D), bootstrapped and null plots form two distinct clusters, emphasizing the visual separability.
-This aligns with the lineup in Figure \ref{fig:boston-lineup}, where the true plot stands out clearly.
+This aligns with the lineup in Figure \ref{fig:lineup}B, where the true plot stands out clearly.
 A human visual test would also likely reject $H_0$ in this case.
@@
                     result)
-result %>%
+l2 <- result %>%
   VI_MODEL$plot_lineup(remove_grid_line = TRUE,
                        remove_axis = TRUE,
-                       theme = theme_light())
+                       theme = theme_light()) +
+  ggtitle("(B) Lineup for Boston housing")
 ```
@@
 It contains only two variables, $x$ and $y$, but fitting a regression model produces a residual plot that clearly resembles a dinosaur (Figure \ref{fig:dino-check}A).
-This pattern is visually distinctive in the lineup shown in Figure \ref{fig:dino-lineup}, where a human visual test would undoubtedly reject $H_0$.
+This pattern is visually distinctive in the lineup shown in Figure \ref{fig:lineup}C, where a human visual test would undoubtedly reject $H_0$.
 
@@
                     result)
-result %>%
+l3 <- result %>%
   VI_MODEL$plot_lineup(remove_grid_line = TRUE,
                        remove_axis = TRUE,
-                       theme = theme_light())
+                       theme = theme_light()) +
+  ggtitle("(C) Lineup for datasauRus")
+```
+
+```{r lineup, fig.pos = "!h", fig.cap = 'Lineups of residual plots for the "left-triangle", "Boston housing", and "datasauRus" datasets.
+(A) True plot at position 10; no clear visual difference.
+(B) True plot at position 7; clearly stands out.
+(C) True plot at position 17; distinctly artificial and easily identifiable.', fig.width = 10, fig.height = 10}
+patchwork::wrap_plots(l1, l2, l3, ncol = 2, widths = c(1, 1), heights = c(1, 1))
 ```

### 2025-06-17 paper.tex
Combine three lineups into one

@@
 of a fitted classical normal linear regression model and the reference
-distribution, based on Kullback-Leibler divergence. From extensive
-simulation studies, the computer vision model exhibits lower sensitivity
-than conventional tests but higher sensitivity than human visual tests.
-It is slightly less effective on non-linearity patterns. Several
-examples from classical papers and contemporary data illustrate the new
-procedures, highlighting its usefulness in automating the diagnostic
-process and supplementing existing methods.
+distribution, based on Kullback-Leibler divergence. Evaluation using
+data from a prior human-subject experiment shows the model exhibits
+lower sensitivity than conventional tests but higher sensitivity than
+human visual tests. Several examples from classical papers and
+contemporary data demonstrate the method's paractical utility in
+automating residual diagnostics.
 \end{abstract}
@@
 regression diagnostics \citep{belsley1980regression, cook1982residuals}.
-This visual assessment plays a crucial role in identifying whether model
-assumptions, such as linearity, homoscedasticity, and normality, are
-reasonable. It also helps in understanding the goodness of fit and
+This visual assessment plays a crucial role in identifying whether key
+model assumptions, such as linearity, homoscedasticity, and normality,
+are reasonable. It also helps in understanding the goodness of fit and
 various unexpected characteristics of the model.
@@
 
-The utilization of computers to interpret data plots has a rich history,
-with early efforts such as ``Scagnostics'' by \citet{tukey1985computer},
-a set of numerical statistics that summarize features of scatter plots.
+Efforts to automate plot interpretation date back to early work like
+Tukey's ``Scagnostics'' \citep{tukey1985computer}, which is a set of
+numerical statistics that summarize features of scatter plots.
 \citet{wilkinson2005graph} expanded on this work, introducing
@@
 ``Clumpy'', and ``Striated'', aimed to characterize outliers, shape,
-density, trend, coherence and other characteristics of the data. While
-this approach has been inspiring, there is a recognition
-\citep{buja2009statistical} that it may not capture all the necessary
-visual features that differentiate true residual plots from null plots.
-A more promising alternative entails enabling machines to learn the
-function for extracting visual features from residual plots.
-Essentially, this means empowering computers to discern the crucial
-visual features for residual diagnostics and determining the method to
-extract them.
+density, trend, coherence and other characteristics of the data.
+However, as noted by \citet{buja2009statistical}, such measures may not
+capture the full range of visual patterns necessary for residual
+diagnostics. A more promising direction lies in enabling computers to
+learn the relevant visual features directly mirroring how humans
+intuitively interpret plots.
 
@@
 evaluating data plots, where only essential plot elements require
-assessment \citep{chowdhury2018measuring}. We can alternatively define a
-notion of distance by integrating key plot elements (instead of key
-perception features like luminance, contrast, and structure), such as
-those captured by scagnostics mentioned in Section
-\ref{sec-model-introduction}, but the functional form still needs to be
-carefully refined to accurately reflect the extent of the violations.
+assessment \citep{chowdhury2018measuring}. A different direction is to
+define a notion of distance by integrating key plot elements (instead of
+key perception features like luminance, contrast, and structure), such
+as those captured by scagnostics (see Section
+\ref{sec-model-introduction}), but the functional form must be carefully
+tailored to reflect model violations meaningfully.
 
 In this section, we introduce a distance measure between a true residual
-plot and a theoretically `good' residual plot. This measure quantifies
+plot and a theoretically ``good'' residual plot. This measure quantifies
 the divergence between the residual distribution of a given fitted
@@
 \(n \times h \times w \times 3\), capable of handling \(n\) RGB images.
-This is followed by a grayscale conversion layer utilizing the luma
-formula under the Rec. 601 standard \citep{series2011studio}, which
-converts the color image to grayscale. Grayscale suffices for our task
-since data points are plotted in black. We experiment with three
-combinations of \(h\) and \(w\): \(32 \times 32\), \(64 \times 64\), and
+This is followed by a grayscale conversion layer using the luma formula
+under the Rec. 601 standard \citep{series2011studio}, which converts the
+color image to grayscale. Grayscale suffices for our task since data
+points are plotted in black. We experiment with three combinations of
+\(h\) and \(w\): \(32 \times 32\), \(64 \times 64\), and
 \(128 \times 128\), aiming to achieve sufficiently high image resolution
@@
 the skewed distribution of fitted values, as illustrated in the lineup
-in Figure \ref{fig:false-lineup}, where similar ``left-triangle''
-patterns appear across all residual plots. Since the true plot (at
-position 10) is visually indistinguishable from the null plots, a visual
-test would not reject \(H_0\).
+in Figure \ref{fig:lineup}A, where similar ``left-triangle'' patterns
+appear across all residual plots. Since the true plot (at position 10)
+is visually indistinguishable from the null plots, a visual test would
+not reject \(H_0\).
 
@@
 cluster. This overlap reinforces the conclusion that the true plot is
-visually indistinct, aligning with our interpretation of the lineup in
-Figure \ref{fig:false-lineup}.
+visually indistinct, aligning with our interpretation of the lineup.
 
@@
 }\label{fig:false-check}
-\end{figure}
-
-\begin{figure}[!h]
-
-{\centering \includegraphics[width=0.7\linewidth]{paper_files/figure-latex/false-lineup-1}
-
-}
-
-\caption{A lineup of residual plots displaying "left-triangle" visual patterns.
-The true residual plot occupies position 10, yet there are no discernible visual patterns that distinguish it from the other plots.}\label{fig:false-lineup}
 \end{figure}
@@
 visual separability. This aligns with the lineup in Figure
-\ref{fig:boston-lineup}, where the true plot stands out clearly. A human
+\ref{fig:lineup}B, where the true plot stands out clearly. A human
 visual test would also likely reject \(H_0\) in this case.
@@
 
-\begin{figure}[!h]
-
-{\centering \includegraphics[width=0.7\linewidth]{paper_files/figure-latex/boston-lineup-1}
-
-}
-
-\caption{A lineup of residual plots for the Boston housing fitted model. The true residual plot is at position 7. It can be easily identified as the most different plot.}\label{fig:boston-lineup}
-\end{figure}
-
 \subsection{DatasauRus}\label{datasaurus}
@@
 resembles a dinosaur (Figure \ref{fig:dino-check}A). This pattern is
-visually distinctive in the lineup shown in Figure
-\ref{fig:dino-lineup}, where a human visual test would undoubtedly
-reject \(H_0\).
+visually distinctive in the lineup shown in Figure \ref{fig:lineup}C,
+where a human visual test would undoubtedly reject \(H_0\).
 
@@
 
-{\centering \includegraphics[width=0.7\linewidth]{paper_files/figure-latex/dino-lineup-1}
+{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/lineup-1}
 
@@
 
-\caption{A lineup of residual plots for the fitted model on the "dinosaur" dataset.
-The true residual plot is at position 17.
-It can be easily identified as the most different plot as the visual pattern is extremely artificial.}\label{fig:dino-lineup}
+\caption{Lineups of residual plots for the "left-triangle", "Boston housing", and "datasauRus" datasets.
+(A) True plot at position 10; no clear visual difference.
+(B) True plot at position 7; clearly stands out.
+(C) True plot at position 17; distinctly artificial and easily identifiable.}\label{fig:lineup}
 \end{figure}

### 2025-06-17 appendix.tex
Fix heading issue in appendix

@@
 
-\subsection{Neural Network Layers Used in the
+\section{Neural Network Layers Used in the
 Study}\label{neural-network-layers-used-in-the-study}
@@
 
-\subsubsection{Dense Layer}\label{dense-layer}
+\subsection{Dense Layer}\label{dense-layer}
 
@@
 
-\subsubsection{ReLu Layer}\label{relu-layer}
+\subsection{ReLu Layer}\label{relu-layer}
 
@@
 
-\subsubsection{Convolutional Layer}\label{convolutional-layer}
+\subsection{Convolutional Layer}\label{convolutional-layer}
 
@@
 
-\subsubsection{Pooling Layer}\label{pooling-layer}
+\subsection{Pooling Layer}\label{pooling-layer}
 
@@
 
-\subsubsection{Global Pooling Layer}\label{global-pooling-layer}
+\subsection{Global Pooling Layer}\label{global-pooling-layer}
 
@@
 
-\subsubsection{Batch Normalization
-Layer}\label{batch-normalization-layer}
+\subsection{Batch Normalization Layer}\label{batch-normalization-layer}
 
@@
 
-\subsubsection{Dropout Layer}\label{dropout-layer}
+\subsection{Dropout Layer}\label{dropout-layer}
 

### 2025-06-17 paper.Rmd
Many adjustments to main paper

@@
 
-```{r false-finding, fig.pos="!h", fig.width = 4, fig.height = 4*4/5, fig.pos="!h", fig.cap="An example residual vs fitted values plot (red line indicates 0 corresponds to the x-intercept, i.e.
+```{r false-finding, fig.pos = "!h", fig.width = 4, fig.height = 4*4/5, fig.cap="An example residual vs fitted values plot (red line indicates 0 corresponds to the x-intercept, i.e.
 $y=0$).
@@
 
-The distance estimator $\hat{D}$ allows us to assess how closely the residuals resemble an ideal distribution and use as an index of model violation severity (see Appendix for details).
+The distance estimator $\hat{D}$ allows us to assess how closely the residuals resemble an ideal distribution and use as an index of model violation severity (see Appendix D for details).
 However, $\hat{D}$ is not expected to equal the true distance $D$, as a single residual plot may not capture all characteristics of the residual distribution.
@@
 
-```{r cnn-diag, fig.cap = "Diagram of the architecture of the optimized computer vision model.
+```{r cnn-diag, fig.pos = "!h", fig.cap = "Diagram of the architecture of the optimized computer vision model.
 Numbers at the bottom of each box show the shape of the output of each layer.
@@
 Figure \ref{fig:cnn-diag} provides a diagram of the architecture.
-More details about the neural network layers used in this study are provided in the Appendix.
+More details about the neural network layers used in this study are provided in the Appendix B.
 
@@
 
-Further details, including mathematical formulations of the  synthetic data model, parameter specifications, sampling scheme, example residual plots and hyperparameter tuning configuration, are provided in the Appendix.
+Further details, including mathematical formulations of the  synthetic data model, parameter specifications, sampling scheme, example residual plots and hyperparameter tuning configuration, are provided in the Appendix A and C.
 
@@
 
-```{r model-performance, fig.pos="!h", fig.cap = "Hexagonal heatmap for difference in $D$ and $\\hat{D}$ vs $D$ on test data for three optimized models with different input sizes.
+```{r model-performance, fig.pos = "!h", fig.cap = "Hexagonal heatmap for difference in $D$ and $\\hat{D}$ vs $D$ on test data for three optimized models with different input sizes.
 The brown lines are smoothing curves produced by fitting generalized additive models.
@@
   xlab(expression(D)) +
-  theme_light() +
+  theme_bw() +
   scale_fill_continuous(limits = c(1, 6000), trans = "log10", low = "#56B1F7", high = "#132B43") +
@@
 
-```{r over-under, fig.pos="!h", fig.cap = "Scatter plots for difference in $D$ and $\\hat{D}$ vs $\\sigma$ on test data for the $32 \\times 32$ optimized model.
+```{r over-under, fig.pos = "!h", fig.cap = "Scatter plots for difference in $D$ and $\\hat{D}$ vs $\\sigma$ on test data for the $32 \\times 32$ optimized model.
 The data is grouped by whether the regression has only non-linearity violation, and whether it includes a second predictor in the regression formula.
 The brown lines are smoothing curves produced by fitting generalized additive models.
-The area over the zero line in light yellow indicates under-prediction, and the area under the zero line in light green indicates over-prediction.", dev = 'png', dpi = 300, fig.height = 4.5}
+The area over the zero line in light yellow indicates under-prediction, and the area under the zero line in light green indicates over-prediction.", dev = 'png', dpi = 300, fig.height = 4}
 model_pred %>%
@@
   mutate(only_z = include_z & !include_heter & !include_non_normal) %>%
+  mutate(only_z = ifelse(only_z, "Non-linearity only", "Multiple violations")) %>%
+  mutate(include_x2 = ifelse(include_x2, "With second predictor", "With single predictor")) %>%
   rename(`Only has non-linearity violation` = only_z, `Has second predictor` = include_x2) %>%
@@
   geom_smooth(aes(e_sigma, effect_size - vss), se = FALSE, col = "#994F00") +
-  facet_grid(`Has second predictor` ~ `Only has non-linearity violation`,
-             labeller = label_both) +
-  theme_light() +
+  facet_grid(`Has second predictor` ~ `Only has non-linearity violation`) +
+  theme_bw() +
   ylab(expression(D - hat(D))) +
@@
 
-```{r pcp, fig.cap = "Parallel coordinate plots of decisions made by computer vision model, conventional tests and visual tests made by human.", eval = TRUE, fig.height = 3.5}
+```{r pcp, fig.pos = "!h", fig.cap = "Parallel coordinate plots of decisions made by computer vision model, conventional tests and visual tests made by human.", eval = TRUE, fig.height = 3.5}
 library(ggpcp)
@@
 The fitted model will not be rejected since $\\hat{D} < Q_{null}(0.95)$.
-(D) plot of first two principal components of features extracted from the global pooling layer of the computer vision model.
+(D) The scatter plot of first two principal components of features extracted from the global pooling layer of the computer vision model.
 ', fig.height = 8, out.width = "80%"}
@@
 The fitted model will be rejected since $\\hat{D} \\geq Q_{null}(0.95)$.
-(D) plot of first two principal components of features extracted from the global pooling layer of the computer vision model.
+(D) The scatter plot of first two principal components of features extracted from the global pooling layer of the computer vision model.
 ', fig.height = 8, out.width = "80%"}
@@
 The fitted model will be rejected since $\\hat{D} \\geq Q_{null}(0.95)$.
-(D) plot of first two principal components of features extracted from the global pooling layer of the computer vision model.', fig.height = 8, out.width = "80%"}
+(D) The scatter plot of first two principal components of features extracted from the global pooling layer of the computer vision model.', fig.height = 8, out.width = "80%"}
 
@@
 
+# Supplementary Materials {-}
+
+\begin{description}
+\item{Appendix:} The appendix includes more details about the synthetic data generation process, neural network layers used in the work, model training, and model violation index.
+(appendix.pdf, Portable Document Format file)
+\end{description}
+
 # Acknowledgement {-}
@@
 The project's GitHub repository (https://github.com/TengMCing/auto_residual_reading_paper) contains all materials required to reproduce this article.
+
+# Disclosure Statement {-}
+
+No potential conflict of interest was reported by the author(s).

### 2025-06-17 paper.tex
Many adjustments to main paper

@@
 residuals resemble an ideal distribution and use as an index of model
-violation severity (see Appendix for details). However, \(\hat{D}\) is
+violation severity (see Appendix D for details). However, \(\hat{D}\) is
 not expected to equal the true distance \(D\), as a single residual plot
@@
 
-\begin{figure}
+\begin{figure}[!h]
 
@@
 about the neural network layers used in this study are provided in the
-Appendix.
+Appendix B.
 
@@
 plots and hyperparameter tuning configuration, are provided in the
-Appendix.
+Appendix A and C.
 
@@
 The fitted model will not be rejected since $\hat{D} < Q_{null}(0.95)$.
-(D) plot of first two principal components of features extracted from the global pooling layer of the computer vision model.
+(D) The scatter plot of first two principal components of features extracted from the global pooling layer of the computer vision model.
 }\label{fig:false-check}
@@
 The fitted model will be rejected since $\hat{D} \geq Q_{null}(0.95)$.
-(D) plot of first two principal components of features extracted from the global pooling layer of the computer vision model.
+(D) The scatter plot of first two principal components of features extracted from the global pooling layer of the computer vision model.
 }\label{fig:boston-check}
@@
 The fitted model will be rejected since $\hat{D} \geq Q_{null}(0.95)$.
-(D) plot of first two principal components of features extracted from the global pooling layer of the computer vision model.}\label{fig:dino-check}
+(D) The scatter plot of first two principal components of features extracted from the global pooling layer of the computer vision model.}\label{fig:dino-check}
 \end{figure}
@@
 
+\section*{Supplementary Materials}\label{supplementary-materials}
+\addcontentsline{toc}{section}{Supplementary Materials}
+
+\begin{description}
+\item{Appendix:} The appendix includes more details about the synthetic data generation process, neural network layers used in the work, model training, and model violation index.
+(appendix.pdf, Portable Document Format file)
+\end{description}
+
 \section*{Acknowledgement}\label{acknowledgement}
@@
 
+\section*{Disclosure Statement}\label{disclosure-statement}
+\addcontentsline{toc}{section}{Disclosure Statement}
+
+No potential conflict of interest was reported by the author(s).
+
 \bibliographystyle{tfcad}

### 2025-06-17 paper.Rmd
Minor changes

@@
 Consider Figure \ref{fig:false-finding} as an example, the residual plot displays a triangular left-pointing shape.
-The distinct difference in the spread of the residuals across the fitted values may result in the analyst suggesting that there may be heteroskedasticity, however, it is important to avoid over-interpreting this visual pattern.
+The distinct difference in the spread of the residuals across the fitted values might lead analysts to suspect heteroskedasticity, however, it is important to avoid over-interpreting this visual pattern.
 In this case, the fitted regression model is correctly specified, and the triangular shape is actually a result of the skewed distribution of the predictors, rather than indicating a flaw in the model.
@@
 
-A third option is to produce a meaninful and interpretable numerical measure that quantifies the strength of suspicious visual patterns reflecting the extent of model violations, or the difficulty index for identifying whether a residual plot has no issues.
-However, such measures are often used informally in daily communications but rarely formalized.
+A third option is to produce a meaningful and interpretable numerical measure that quantifies the strength of suspicious visual patterns reflecting the extent of model violations, or the difficulty index for identifying whether a residual plot has no issues.
+However, such measures are often used informally in daily communications but are rarely formalized.
 For supervised learning, they must be clearly defined and reliably estimable.
@@
 This distance can take the form of a basic statistical operation on pixels, such as the sum of square differences, however, a pixel-to-pixel comparison makes little sense in comparing residual plots where the main interest would be structural patterns.
-Alternatively, it could involve established image similarity metrics like the Structural Similarity Index Measure [@wang2004image] which compares images by integrating three perception features of an image: contrast, luminance, and structure (related to average, standard deviation and correlation of pixel values over a window, respectively).
-These image similarity metrics are tailored for image comparison in vastly different tasks to evaluating data plots, where only essential plot elements require assessment [@chowdhury2018measuring].
-A different direction is to define a notion of distance by integrating key plot elements (instead of key perception features like luminance, contrast, and structure), such as those captured by scagnostics (see Section \ref{sec-model-introduction}), but the functional form must be carefully tailored to reflect model violations meaningfully.
+Alternatively, it
+can involve established image similarity metrics like the Structural Similarity Index Measure [@wang2004image], which compares images by integrating three perception features of an image: contrast, luminance, and structure (related to average, standard deviation and correlation of pixel values over a window, respectively).
+These image similarity metrics are tailored for image comparison in tasks that are vastly different from evaluating data plots, where only essential plot elements require assessment [@chowdhury2018measuring].
+A different direction is to define a notion of distance by integrating key plot elements (instead of key perception features like luminance, contrast, and structure), such as those captured by scagnostics (see Section \ref{sec-model-introduction}), but the functional form needs to be carefully tailored to reflect model violations meaningfully.
 
@@
 
-For a classical normal linear regression model, $\boldsymbol{y} = \boldsymbol{X}\boldsymbol{\beta} + \boldsymbol{e}$, the residual $\hat{\boldsymbol{e}}$ are derived as the difference of the fitted values and observed values $\boldsymbol{y}$.
+For a classical normal linear regression model, $\boldsymbol{y} = \boldsymbol{X}\boldsymbol{\beta} + \boldsymbol{e}$, the residuals $\hat{\boldsymbol{e}}$ are derived as the difference of the fitted values and observed values $\boldsymbol{y}$.
 Suppose the data generating process is known and the regression model is correctly specified, by the Frisch-Waugh-Lowell theorem [@frisch1933partial], residuals $\hat{\boldsymbol{e}}$ can also be treated as random variables and written as a linear transformation of the error $\boldsymbol{e}$ formulated as $\hat{\boldsymbol{e}} = \boldsymbol{R}\boldsymbol{e}$, where $\boldsymbol{R}=\boldsymbol{I}_n -\boldsymbol{X}(\boldsymbol{X}^\top\boldsymbol{X})^{-1}\boldsymbol{X}^\top$ is the residual operator, $\boldsymbol{I}_n$ is a $n$ by $n$ identity matrix, and $n$ is the number of observations.
@@
 
-Suppose the design matrix $\boldsymbol{X}$ has linearly independent columns, the trace of the hat matrix $\boldsymbol{H} = \boldsymbol{X}(\boldsymbol{X}^\top\boldsymbol{X})^{-1}\boldsymbol{X}^\top$ will equal to the number of columns in $\boldsymbol{X}$ denoted as $k$.
+Suppose the design matrix $\boldsymbol{X}$ has linearly independent columns, the trace of the hat matrix $\boldsymbol{H} = \boldsymbol{X}(\boldsymbol{X}^\top\boldsymbol{X})^{-1}\boldsymbol{X}^\top$ equals the number of columns in $\boldsymbol{X}$ denoted as $k$.
 As a result, the rank of $\boldsymbol{R}$ is $n - k$, and $Q$ is a degenerate multivariate distribution.
-To capture the characteristics of $Q$, such as moments, we can simulate a large numbers of $\boldsymbol{\varepsilon}$ and transform it to $\boldsymbol{e}$ to get the empirical estimates.
+To capture the characteristics of $Q$, such as moments, we can simulate a large number of $\boldsymbol{\varepsilon}$ and transform it to $\boldsymbol{e}$ to get the empirical estimates.
 For simplicity, in this study, we replaced the variance-covariance matrix of residuals $\text{cov}(\boldsymbol{e}, \boldsymbol{e}) = \boldsymbol{R}\sigma^2\boldsymbol{R}^\top = \boldsymbol{R}\sigma^2$ with a full-rank diagonal matrix $\text{diag}(\boldsymbol{R}\sigma^2)$, where $\text{diag}(.)$ sets the non-diagonal entries of a matrix to zeros.
-The resulting distribution for $Q$ is $N(\boldsymbol{0}_n, \text{diag}(\boldsymbol{R}\sigma^2))$.
+The resulting distribution for $Q$ was $N(\boldsymbol{0}_n, \text{diag}(\boldsymbol{R}\sigma^2))$.
 
@@
 However, if the model is misspecified, then the actual distribution of residuals denoted as $P$, will be different from $Q$.
-For example, if the data generating process contains variables correlated with any column of $\boldsymbol{X}$ but missing from $\boldsymbol{X}$, causing an omitted variable problem, $P$ will be different from $Q$ because the residual operator obtained from the fitted regression model will not be the same as $\boldsymbol{R}$.
+For example, if the data generating process contains variables correlated with any column of $\boldsymbol{X}$ but are missing from $\boldsymbol{X}$, causing an omitted variable problem, $P$ will be different from $Q$ because the residual operator obtained from the fitted regression model will not be the same as $\boldsymbol{R}$.
 Besides, if the $\boldsymbol{\varepsilon}$ follows a non-normal distribution such as a multivariate lognormal distribution, $P$ will usually be skewed and has a long tail.
@@
 Defining a proper distance between distributions is usually easier than defining a proper distance between data plots.
-Given the true residual distribution $Q$ and the reference residual distribution $P$, we used a distance measure based on Kullback-Leibler divergence [@kullback1951information] to quantify the difference between two distributions as
+Given the true residual distribution $P$ and the reference residual distribution $Q$, we used a distance measure based on Kullback-Leibler divergence [@kullback1951information] to quantify the difference between two distributions as
 
@@
 
-This distance measure was first proposed in @li2024plot.
-It was mainly designed for measuring the effect size of non-linearity and heteroskedasticity in a residual plot.
-@li2024plot have derived that, for a classical normal linear regression model that omits necessary higher-order predictors $\boldsymbol{Z}$ and the corresponding parameter $\boldsymbol{\beta}_z$, and incorrectly assumes $\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n,\sigma^2\boldsymbol{I}_n)$ while in fact $\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n, \boldsymbol{V})$ where $\boldsymbol{V}$ is an arbitrary symmetric positive semi-definite matrix, $Q$ can be represented as $N(\boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z, \text{diag}(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}))$.
-Note that the variance-covariance matrix is replaced with the diagonal matrix to ensure it is a full-rank matrix.
+This distance measure was first proposed in @li2024plot, which was mainly designed for measuring the effect size of non-linearity and heteroskedasticity in a residual plot.
+@li2024plot has derived that, for a classical normal linear regression model that omits necessary higher-order predictors $\boldsymbol{Z}$ and the corresponding parameter $\boldsymbol{\beta}_z$, while incorrectly assuming $\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n,\sigma^2\boldsymbol{I}_n)$, but in fact $\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n, \boldsymbol{V})$ where $\boldsymbol{V}$ is an arbitrary symmetric positive semi-definite matrix, $Q$ can be represented as $N(\boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z, \text{diag}(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}))$.
+Note that the variance-covariance matrix was replaced with the diagonal matrix to ensure it is a full-rank matrix.
 

### 2025-06-17 paper.tex
Minor changes

@@
 a triangular left-pointing shape. The distinct difference in the spread
-of the residuals across the fitted values may result in the analyst
-suggesting that there may be heteroskedasticity, however, it is
-important to avoid over-interpreting this visual pattern. In this case,
-the fitted regression model is correctly specified, and the triangular
-shape is actually a result of the skewed distribution of the predictors,
-rather than indicating a flaw in the model.
+of the residuals across the fitted values might lead analysts to suspect
+heteroskedasticity, however, it is important to avoid over-interpreting
+this visual pattern. In this case, the fitted regression model is
+correctly specified, and the triangular shape is actually a result of
+the skewed distribution of the predictors, rather than indicating a flaw
+in the model.
 
@@
 
-A third option is to produce a meaninful and interpretable numerical
+A third option is to produce a meaningful and interpretable numerical
 measure that quantifies the strength of suspicious visual patterns
@@
 identifying whether a residual plot has no issues. However, such
-measures are often used informally in daily communications but rarely
-formalized. For supervised learning, they must be clearly defined and
-reliably estimable.
+measures are often used informally in daily communications but are
+rarely formalized. For supervised learning, they must be clearly defined
+and reliably estimable.
 
@@
 comparison makes little sense in comparing residual plots where the main
-interest would be structural patterns. Alternatively, it could involve
+interest would be structural patterns. Alternatively, it can involve
 established image similarity metrics like the Structural Similarity
-Index Measure \citep{wang2004image} which compares images by integrating
-three perception features of an image: contrast, luminance, and
-structure (related to average, standard deviation and correlation of
+Index Measure \citep{wang2004image}, which compares images by
+integrating three perception features of an image: contrast, luminance,
+and structure (related to average, standard deviation and correlation of
 pixel values over a window, respectively). These image similarity
-metrics are tailored for image comparison in vastly different tasks to
-evaluating data plots, where only essential plot elements require
-assessment \citep{chowdhury2018measuring}. A different direction is to
-define a notion of distance by integrating key plot elements (instead of
-key perception features like luminance, contrast, and structure), such
-as those captured by scagnostics (see Section
-\ref{sec-model-introduction}), but the functional form must be carefully
-tailored to reflect model violations meaningfully.
+metrics are tailored for image comparison in tasks that are vastly
+different from evaluating data plots, where only essential plot elements
+require assessment \citep{chowdhury2018measuring}. A different direction
+is to define a notion of distance by integrating key plot elements
+(instead of key perception features like luminance, contrast, and
+structure), such as those captured by scagnostics (see Section
+\ref{sec-model-introduction}), but the functional form needs to be
+carefully tailored to reflect model violations meaningfully.
 
@@
 \(\boldsymbol{y} = \boldsymbol{X}\boldsymbol{\beta} + \boldsymbol{e}\),
-the residual \(\hat{\boldsymbol{e}}\) are derived as the difference of
+the residuals \(\hat{\boldsymbol{e}}\) are derived as the difference of
 the fitted values and observed values \(\boldsymbol{y}\). Suppose the
@@
 \(\boldsymbol{H} = \boldsymbol{X}(\boldsymbol{X}^\top\boldsymbol{X})^{-1}\boldsymbol{X}^\top\)
-will equal to the number of columns in \(\boldsymbol{X}\) denoted as
-\(k\). As a result, the rank of \(\boldsymbol{R}\) is \(n - k\), and
-\(Q\) is a degenerate multivariate distribution. To capture the
-characteristics of \(Q\), such as moments, we can simulate a large
-numbers of \(\boldsymbol{\varepsilon}\) and transform it to
-\(\boldsymbol{e}\) to get the empirical estimates. For simplicity, in
-this study, we replaced the variance-covariance matrix of residuals
+equals the number of columns in \(\boldsymbol{X}\) denoted as \(k\). As
+a result, the rank of \(\boldsymbol{R}\) is \(n - k\), and \(Q\) is a
+degenerate multivariate distribution. To capture the characteristics of
+\(Q\), such as moments, we can simulate a large number of
+\(\boldsymbol{\varepsilon}\) and transform it to \(\boldsymbol{e}\) to
+get the empirical estimates. For simplicity, in this study, we replaced
+the variance-covariance matrix of residuals
 \(\text{cov}(\boldsymbol{e}, \boldsymbol{e}) = \boldsymbol{R}\sigma^2\boldsymbol{R}^\top = \boldsymbol{R}\sigma^2\)
@@
 the non-diagonal entries of a matrix to zeros. The resulting
-distribution for \(Q\) is
+distribution for \(Q\) was
 \(N(\boldsymbol{0}_n, \text{diag}(\boldsymbol{R}\sigma^2))\).
@@
 if the data generating process contains variables correlated with any
-column of \(\boldsymbol{X}\) but missing from \(\boldsymbol{X}\),
+column of \(\boldsymbol{X}\) but are missing from \(\boldsymbol{X}\),
 causing an omitted variable problem, \(P\) will be different from \(Q\)
@@
 defining a proper distance between data plots. Given the true residual
-distribution \(Q\) and the reference residual distribution \(P\), we
+distribution \(P\) and the reference residual distribution \(Q\), we
 used a distance measure based on Kullback-Leibler divergence
@@
 
-This distance measure was first proposed in \citet{li2024plot}. It was
-mainly designed for measuring the effect size of non-linearity and
-heteroskedasticity in a residual plot. \citet{li2024plot} have derived
+This distance measure was first proposed in \citet{li2024plot}, which
+was mainly designed for measuring the effect size of non-linearity and
+heteroskedasticity in a residual plot. \citet{li2024plot} has derived
 that, for a classical normal linear regression model that omits
 necessary higher-order predictors \(\boldsymbol{Z}\) and the
-corresponding parameter \(\boldsymbol{\beta}_z\), and incorrectly
-assumes
-\(\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n,\sigma^2\boldsymbol{I}_n)\)
-while in fact
+corresponding parameter \(\boldsymbol{\beta}_z\), while incorrectly
+assuming
+\(\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n,\sigma^2\boldsymbol{I}_n)\),
+but in fact
 \(\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n, \boldsymbol{V})\)
@@
 \(N(\boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z, \text{diag}(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}))\).
-Note that the variance-covariance matrix is replaced with the diagonal
+Note that the variance-covariance matrix was replaced with the diagonal
 matrix to ensure it is a full-rank matrix.

### 2025-06-17 paper.Rmd
Fix lineup stroke

@@
   facet_wrap(~type, scales = "free_y") +
-  theme_light() +
+  theme_bw() +
   scale_x_discrete(labels = c("Visual\ntest", "Computer\nvision\nmodel", "Conventional\ntest")) +
@@
   facet_grid(~type, scales = "free_x") +
-  theme_light() +
+  theme_bw() +
   scale_color_brewer("", palette = "Dark2") +
@@
                        remove_axis = TRUE,
-                       theme = theme_light()) +
+                       theme = theme_bw(),
+                       alpha = 0.6,
+                       stroke = 0.3) +
   ggtitle("(B) Lineup for left-triangle")
@@
                        remove_axis = TRUE,
-                       theme = theme_light()) +
+                       theme = theme_bw(),
+                       alpha = 0.6,
+                       stroke = 0.3) +
   ggtitle("(B) Lineup for Boston housing")
@@
                        remove_axis = TRUE,
-                       theme = theme_light()) +
+                       theme = theme_bw(),
+                       alpha = 0.6,
+                       stroke = 0.3) +
   ggtitle("(C) Lineup for datasauRus")

### 2025-06-17 paper.Rmd
fixed some typos

@@
 
+```{r false-finding, fig.pos = "!h", fig.width = 4, fig.height = 4*4/5, fig.cap="An example residual vs fitted values plot (red line indicates 0 corresponds to the x-intercept, i.e.
+$y=0$).
+The vertical spread of the data points varies with the fitted values.
+This often indicates the existence of heteroskedasticity, however, here the result is due to skewed distribution of the predictors rather than heteroskedasticity.
+The Breusch-Pagan test rejects this residual plot at 95\\% significance level ($p\\text{-value} = 0.046$).", out.width = "50%"}
+set.seed(452)
+ori_x <- rand_lognormal()
+mod <- heter_model(b = 0, x = closed_form(~-ori_x))
+ori_dat <- mod$gen(300)
+
+ori_dat %>%
+  VI_MODEL$plot(theme = theme_light(), remove_grid_line = TRUE) +
+  xlab("Fitted values") +
+  ylab("Residuals")
+```
+
 The concept of visual inference, as proposed by @buja2009statistical, provides an inferential framework to assess whether residual plots indeed contain visual patterns inconsistent with the model assumptions.
@@
 This characteristic makes it less suitable for large-scale applications, given the associated high labour costs and time requirements.
-There is a substantial need to develop an approach to substitute these human judgement with an automated reading of data plots using machines.
+There is a substantial need to develop an approach to substitute these human judgements with an automated reading of data plots using machines.
 
@@
 Finally, we conclude with a discussion of our findings and propose ideas for future research directions.
-
-```{r false-finding, fig.pos = "!h", fig.width = 4, fig.height = 4*4/5, fig.cap="An example residual vs fitted values plot (red line indicates 0 corresponds to the x-intercept, i.e.
-$y=0$).
-The vertical spread of the data points varies with the fitted values.
-This often indicates the existence of heteroskedasticity, however, here the result is due to skewed distribution of the predictors rather than heteroskedasticity.
-The Breusch-Pagan test rejects this residual plot at 95\\% significance level ($p\\text{-value} = 0.046$).", out.width = "50%"}
-set.seed(452)
-ori_x <- rand_lognormal()
-mod <- heter_model(b = 0, x = closed_form(~-ori_x))
-ori_dat <- mod$gen(300)
-
-ori_dat %>%
-  VI_MODEL$plot(theme = theme_light(), remove_grid_line = TRUE) +
-  xlab("Fitted values") +
-  ylab("Residuals")
-```
 
@@
 The brown lines are smoothing curves produced by fitting generalized additive models.
-The area over the zero line in light yellow indicates under-prediction, and the area under the zero line in light green indicates over-prediction.", dev = 'png', dpi = 300, fig.height = 3}
+The area above the zero line in light yellow indicates under-prediction, and the area under the zero line in light green indicates over-prediction.", dev = 'png', dpi = 300, fig.height = 3}
 model_pred <- train_pred %>% 
@@
 The brown lines are smoothing curves produced by fitting generalized additive models.
-The area over the zero line in light yellow indicates under-prediction, and the area under the zero line in light green indicates over-prediction.", dev = 'png', dpi = 300, fig.height = 4}
+The area above the zero line in light yellow indicates under-prediction, and the area under the zero line in light green indicates over-prediction.", dev = 'png', dpi = 300, fig.height = 4}
 model_pred %>%
@@
 \item{Appendix:} The appendix includes more details about the synthetic data generation process, neural network layers used in the work, model training, and model violation index. 
-(appendix.pdf, Portable Document Format file)
 \end{description}

### 2025-06-17 paper.tex
fixed some typos

@@
 
+\begin{figure}[!h]
+
+{\centering \includegraphics[width=0.5\linewidth]{paper_files/figure-latex/false-finding-1}
+
+}
+
+\caption{An example residual vs fitted values plot (red line indicates 0 corresponds to the x-intercept, i.e.
+$y=0$).
+The vertical spread of the data points varies with the fitted values.
+This often indicates the existence of heteroskedasticity, however, here the result is due to skewed distribution of the predictors rather than heteroskedasticity.
+The Breusch-Pagan test rejects this residual plot at 95\% significance level ($p\text{-value} = 0.046$).}\label{fig:false-finding}
+\end{figure}
+
 The concept of visual inference, as proposed by
@@
 labour costs and time requirements. There is a substantial need to
-develop an approach to substitute these human judgement with an
+develop an approach to substitute these human judgements with an
 automated reading of data plots using machines.
@@
 for future research directions.
-
-\begin{figure}[!h]
-
-{\centering \includegraphics[width=0.5\linewidth]{paper_files/figure-latex/false-finding-1}
-
-}
-
-\caption{An example residual vs fitted values plot (red line indicates 0 corresponds to the x-intercept, i.e.
-$y=0$).
-The vertical spread of the data points varies with the fitted values.
-This often indicates the existence of heteroskedasticity, however, here the result is due to skewed distribution of the predictors rather than heteroskedasticity.
-The Breusch-Pagan test rejects this residual plot at 95\% significance level ($p\text{-value} = 0.046$).}\label{fig:false-finding}
-\end{figure}
 
@@
 The brown lines are smoothing curves produced by fitting generalized additive models.
-The area over the zero line in light yellow indicates under-prediction, and the area under the zero line in light green indicates over-prediction.}\label{fig:model-performance}
+The area above the zero line in light yellow indicates under-prediction, and the area under the zero line in light green indicates over-prediction.}\label{fig:model-performance}
 \end{figure}
@@
 The brown lines are smoothing curves produced by fitting generalized additive models.
-The area over the zero line in light yellow indicates under-prediction, and the area under the zero line in light green indicates over-prediction.}\label{fig:over-under}
+The area above the zero line in light yellow indicates under-prediction, and the area under the zero line in light green indicates over-prediction.}\label{fig:over-under}
 \end{figure}

### 2025-06-28 paper.Rmd
update affiliations

--- 2025-06-17 paper.Rmd
@@
   - name: Emi Tanaka
-    affil: a, b, c
+    affil: b
     email: emi.tanaka@anu.edu.au
   - name: Susan VanderPlas
-    affil: d
+    affil: c
     email: susan.vanderplas@unl.edu
@@
     address: |
-      Biological Data Science Institute, Australian National University, Acton, ACT, Australia
+      Research School of Finance, Actuarial Studies and Statistics, Australian National University, Acton, ACT, Australia
   - num: c
-    address: |
-      Research School of Finance, Actuarial Studies and Statistics, Australian National University, Acton, ACT, Australia
-  - num: d
     address: |

### 2025-07-10 paper.Rmd
Reduce to 36 pages

--- 2025-06-17 paper.Rmd
@@
 Evaluation using data from a prior human-subject experiment shows the model exhibits lower sensitivity than conventional tests but higher sensitivity than human visual tests.
-Several examples from classical papers and contemporary data demonstrate the method's paractical utility in automating residual diagnostics.
+Several examples from classical papers and contemporary data demonstrate the method's practical utility in automating residual diagnostics.
 keywords: |
@@
 Generating a residual plot in most statistical software is often as straightforward as executing a line of code or clicking a button.
-However, accurately interpreting a residual plot can be challenging.
-A residual plot can exhibit various visual features, but it is crucial to recognize that some may arise from the characteristics of predictors and the natural stochastic variation of the observational unit, rather than indicating a violation of model assumptions [@li2024plot].
-Consider Figure \ref{fig:false-finding} as an example, the residual plot displays a triangular left-pointing shape.
-The distinct difference in the spread of the residuals across the fitted values might lead analysts to suspect heteroskedasticity, however, it is important to avoid over-interpreting this visual pattern.
-In this case, the fitted regression model is correctly specified, and the triangular shape is actually a result of the skewed distribution of the predictors, rather than indicating a flaw in the model.
+However, accurately interpreting a residual plot can be challenging, as some features are a result of predictor characteristics and natural variation, while others indicate violations of model assumptions [@li2024plot].
+For example, in Figure \ref{fig:false-finding}, the residual plot displays a triangular left-pointing shape.
+The distinct difference in the spread of the residuals across the fitted values might lead analysts to suspect heteroskedasticity.
+In this case, however, the fitted regression model is correctly specified, and the triangular shape is actually a result of the skewed distribution of the predictors, rather than indicating a flaw in the model.
+
+```{r false-finding, fig.pos = "!h", fig.width = 4, fig.height = 4*4/5, fig.cap="An example residual vs fitted values plot (red line indicates 0 corresponds to the x-intercept, i.e.
+$y=0$).
+The vertical spread of the data points varies with the fitted values.
+This often indicates the existence of heteroskedasticity, however, here the result is due to skewed distribution of the predictors rather than heteroskedasticity.
+The Breusch-Pagan test rejects this residual plot at 95\\% significance level ($p\\text{-value} = 0.046$).", out.width = "50%"}
+set.seed(452)
+ori_x <- rand_lognormal()
+mod <- heter_model(b = 0, x = closed_form(~-ori_x))
+ori_dat <- mod$gen(300)
+
+ori_dat %>%
+  VI_MODEL$plot(theme = theme_light(), remove_grid_line = TRUE) +
+  xlab("Fitted values") +
+  ylab("Residuals")
+```
 
@@
 
-A practical limitation of the lineup protocol lies in its reliance on human judgements [see @li2024plot about the practical limitations].
-Unlike conventional statistical tests that can be performed computationally in statistical software, the lineup protocol requires human evaluation of images.
-This characteristic makes it less suitable for large-scale applications, given the associated high labour costs and time requirements.
-There is a substantial need to develop an approach to substitute these human judgement with an automated reading of data plots using machines.
+A practical limitation of the lineup protocol lies in its reliance on human judgements [see @li2024plot].
+Unlike conventional statistical tests that can be performed using in statistical software, the lineup protocol requires human evaluation of images.
+This characteristic makes it less suitable for large-scale applications, given the associated high labor costs and time requirements.
+There is a substantial need to develop an approach to substitute these human judgement with an automated reading of data plots using computer vision.
 
@@
 These measures, including, but not limited to, "Outlying", "Skinny", "Stringy", "Straight", "Monotonic", "Skewed", "Clumpy", and "Striated", aimed to characterize outliers, shape, density, trend, coherence and other characteristics of the data.
-However, as noted by @buja2009statistical, such measures may not capture the full range of visual patterns necessary for residual diagnostics.
+However, such measures may not capture the full range of visual patterns necessary for residual diagnostics [@buja2009statistical].
 A more promising direction lies in enabling computers to learn the relevant visual features directly mirroring how humans intuitively interpret plots.
 
-Modern computer vision models are well-suited for addressing this challenge.
-They rely on deep neural networks with convolutional layers [@fukushima1982neocognitron].
+Modern computer vision models rely on deep neural networks with convolutional layers [@fukushima1982neocognitron].
 These layers use small, sliding windows to scan the image, performing a dot product to extract local features and patterns.
@@
 
-```{r false-finding, fig.pos = "!h", fig.width = 4, fig.height = 4*4/5, fig.cap="An example residual vs fitted values plot (red line indicates 0 corresponds to the x-intercept, i.e.
-$y=0$).
-The vertical spread of the data points varies with the fitted values.
-This often indicates the existence of heteroskedasticity, however, here the result is due to skewed distribution of the predictors rather than heteroskedasticity.
-The Breusch-Pagan test rejects this residual plot at 95\\% significance level ($p\\text{-value} = 0.046$).", out.width = "50%"}
-set.seed(452)
-ori_x <- rand_lognormal()
-mod <- heter_model(b = 0, x = closed_form(~-ori_x))
-ori_dat <- mod$gen(300)
-
-ori_dat %>%
-  VI_MODEL$plot(theme = theme_light(), remove_grid_line = TRUE) +
-  xlab("Fitted values") +
-  ylab("Residuals")
-```
-
 # Model Specifications {#sec-model-specifications}
 
-There are various specifications of the computer vision model that can be used to assess residual plots.
-We discuss these specifications below focusing on two key components of the model formula: the input and the output format.
+There are various specifications of the computer vision model that can be used to assess residual plots. We discuss these specifications below focusing on the input and the output formats.
 
@@
 Another strategy is to convert residual plots into images.
-This enables the use of standard convolutional architectures such as VGG16 [@simonyan2014very], which can effectively capture spatial patterns.
-Although discretization introduces some losses in detail, image-based inputs offer a richer representation by preserving the joint distribution of residuals and fitted values in a visually interpretable format.
+This enables the use of standard convolutional architectures in computer vision such as VGG16 [@simonyan2014very].
+Image-based inputs offer a richer representation by preserving the joint distribution of residuals and fitted values in a visually interpretable format, helping models more effectively capture spatial structure and local correlations than vector-based formats.
 To ensure model generalization, it is essential to use residual plots in a consistent style, i.e., same aesthetics, layout, scaling and color schemes.
@@
 Using multiple residual plots as input, such as pairs, triplets, or full lineups is also a possible option.
-For instance, triplet networks [@chopra2005learning] can assess visual similarity if we assume null plots are more similar from each other than from the true residual plot, which may help distinguish null plots from true residual plots.
+For instance, triplet networks [@chopra2005learning] can assess visual similarity if we assume null plots are more similar to each other than to the true residual plot, which may help distinguish null plots from true residual plots.
 However, these architectures introduce complexity in training due to weight sharing and specialized loss functions.
@@
 
-In this study, we defined and used a continuous distance between a true residual plot and a theoretically "good" residual plot.
+In this study, we define and use a continuous distance between a true residual plot and a theoretically "good" residual plot.
 This approach captures the degree to which a plot deviates from model assumptions.
@@
 \begin{align*} \label{eq:kl-3}
-D_{KL} &\approx \sum_{i = 1}^{n} \hat{D}_{KL}^{(i)}, \\
-\hat{D}_{KL}^{(i)} &= \frac{1}{m}\sum_{j = 1}^{m} \log\frac{\hat{p}_i(B_{ij})}{q(B_{ij})},
+D_{KL} &\approx \sum_{i = 1}^{n} \hat{D}_{KL}^{(i)}, \quad
+\hat{D}_{KL}^{(i)} = \frac{1}{m}\sum_{j = 1}^{m} \log\frac{\hat{p}_i(B_{ij})}{q(B_{ij})},
 \end{align*}
 
-\noindent where $\hat{D}_{KL}^{(i)}$ is the estimator of the Kullback-Leibler divergence for an individual residual $e_i$, $\boldsymbol{B}_{ij}$ is the $i$-th row and $j$-th column entry of the matrix $\boldsymbol{B}$, $\hat{p}_i(.)$ is the kernel density estimator of $p_i(.)$, $q(.)$ is the normal density function with mean zero and an assumed variance estimated as $\hat{\sigma}^2 = \sum_{b \in vec(\boldsymbol{B})}(b - \sum_{b \in vec(\boldsymbol{B})} b/nm)^2/(nm - 1)$, and $vec(.)$ is the vectorization operator which turns a $n \times m$ matrix into a $nm \times 1$ column vector by stacking the columns of the matrix on top of each other.
+\noindent where $\hat{D}_{KL}^{(i)}$ is the estimator of the Kullback-Leibler divergence for an individual residual $e_i$, $B_{ij}$ is the $i$-th row and $j$-th column entry of the matrix $\boldsymbol{B}$, $\hat{p}_i(.)$ is the kernel density estimator of $p_i(.)$, $q(.)$ is the normal density function with mean zero and an assumed variance estimated as $\hat{\sigma}^2 = \sum_{b \in vec(\boldsymbol{B})}(b - \sum_{b \in vec(\boldsymbol{B})} b/nm)^2/(nm - 1)$, and $vec(.)$ is the vectorization operator which turns a $n \times m$ matrix into a $nm \times 1$ column vector by stacking the columns of the matrix on top of each other.
 
@@
 
-\begin{equation} \label{eq:d-approx}
+\begin{equation*} \label{eq:d-approx}
 \hat{D} = f_{CV}(V_{h \times w}(\boldsymbol{e}, \hat{\boldsymbol{y}})),
-\end{equation}
-
-\noindent where $V_{h \times w}(.)$ renders the residuals vs fitted values plot as an $h \times w$ RGB image, and $f_{CV}(.)$ is a model that maps the image to an estimated distance $\hat{D} \in [0, +\infty)$.
+\end{equation*}
+
+\noindent where $V_{h \times w}(.)$ renders the residuals vs.
+fitted values plot as an $h \times w$ RGB image, and $f_{CV}(.)$ is a model that maps the image to an estimated distance $\hat{D} \in [0, +\infty)$.
 
@@
 
+The architecture of the computer vision model $f_{CV}$ is adapted from the well-established VGG16 architecture [@simonyan2014very].
+While more recent architectures like ResNet [@he2016deep] and DenseNet[@huang2017densely], have achieved even greater performance, VGG16 remains a solid choice for many applications due to its simplicity and effectiveness.
+Our decision to use VGG16 aligns with our goal of starting with a proven and straightforward model.
+Figure \ref{fig:cnn-diag} provides a diagram of the architecture.
+More details about the neural network layers used in this study are provided in the Appendix B.
+
 ```{r cnn-diag, fig.pos = "!h", fig.cap = "Diagram of the architecture of the optimized computer vision model.
@@
 ```
-
-The architecture of the computer vision model $f_{CV}$ is adapted from the well-established VGG16 architecture [@simonyan2014very].
-While more recent architectures like ResNet [@he2016deep] and DenseNet[@huang2017densely], have achieved even greater performance, VGG16 remains a solid choice for many applications due to its simplicity and effectiveness.
-Our decision to use VGG16 aligns with our goal of starting with a proven and straightforward model.
-Figure \ref{fig:cnn-diag} provides a diagram of the architecture.
-More details about the neural network layers used in this study are provided in the Appendix B.
 
@@
 
-The model weights $\boldsymbol{\theta}$ were randomly initialized using the Glorot Uniform method [@glorot2010understanding] and they were optimized by the Adam optimizer [@kingma2014adam] with the mean square error loss function
-
-$$\hat{\boldsymbol{\theta}} = \underset{\boldsymbol{\theta}}{\text{arg min}}\frac{1}{n_{\text{train}}}\sum_{i=1}^{n_{\text{train}}}(D_i - f_{\boldsymbol{\theta}}(V_i, S_i))^2,$$
-
-\noindent where $n_{\text{train}}$ is the number of training samples, $V_i$ is the $i$-th residual plot and $S_i$ is the auxiliary information about the $i$-th residual plot including four scagnostics and the number of observations.
+The model weights $\boldsymbol{\theta}$ were randomly initialized using the Glorot Uniform method [@glorot2010understanding] and they were optimized by the Adam optimizer [@kingma2014adam] with the MSE loss function $\hat{\boldsymbol{\theta}} = \underset{\boldsymbol{\theta}}{\text{arg min}}\frac{1}{n_{\text{train}}}\sum_{i=1}^{n_{\text{train}}}(D_i - f_{\boldsymbol{\theta}}(V_i, S_i))^2$, where $n_{\text{train}}$ is the number of training samples, $V_i$ is the $i$-th residual plot and $S_i$ is the auxiliary information about the $i$-th residual plot including four scagnostics and the number of observations.
 
@@
 The brown lines are smoothing curves produced by fitting generalized additive models.
-The area over the zero line in light yellow indicates under-prediction, and the area under the zero line in light green indicates over-prediction.", dev = 'png', dpi = 300, fig.height = 3}
+Over-prediction and under-prediction can be observed for small $D$ and large $D$ respectively.", dev = 'png', dpi = 300, fig.height = 3}
 model_pred <- train_pred %>% 
@@
 The brown lines are smoothing curves produced by fitting generalized additive models.
-The area over the zero line in light yellow indicates under-prediction, and the area under the zero line in light green indicates over-prediction.", dev = 'png', dpi = 300, fig.height = 4}
+Under-prediction mainly occurs when the data-generating process has small $\\sigma$, a second predictor, and only non-linearity as the model violation.", dev = 'png', dpi = 300, fig.height = 4}
 model_pred %>%
@@
 
+Each lineup in @li2024plot contains one true residual plot and 19 null plots.
+While the true residual plot’s distance $D$ depends on the data-generating process, null plots have $D = 0$.
+We used our optimized computer vision model to estimate $D$ for all plots.
+To ensure fair comparison, we reject $H_0$ if the true residual plot has the highest estimated distance.
+Conventional tests, including the Ramsey RESET [@ramsey1969tests] for non-linearity and the Breusch-Pagan test [@breusch1979simple] for heteroskedasticity, were also applied to the same data.
+
+Table \ref{tab:experiment-performance} reports performance metrics for $\hat{D}$ on true residual plots.
+While the performance is slightly worse than on the test data provided in Table \ref{tab:performance}, the mean absolute error remains low and the correlation between predictions and true values remains high.
+Consistent with results in Table \ref{tab:performance-sub}, non-linearity cases are harder to predict than heteroskedasticity cases.
+
 ```{r}
@@
 ```
-
-Each lineup in @li2024plot contains one true residual plot and 19 null plots.
-While the true residual plot’s distance $D$ depends on the data-generating process, null plots have $D = 0$.
-We used our optimized computer vision model to estimate $D$ for all plots.
-To ensure fair comparison, we reject $H_0$ if the true residual plot has the highest estimated distance.
-Conventional tests, including the Ramsey RESET [@ramsey1969tests] for non-linearity and the Breusch-Pagan test [@breusch1979simple] for heteroskedasticity, were also applied to the same data.
-
-Table \ref{tab:experiment-performance} reports performance metrics for $\hat{D}$ on true residual plots.
-While the performance is slightly worse than on the test data provided in Table \ref{tab:performance}, the mean absolute error remains low and the correlation between predictions and true values remains high.
-Consistent with results in Table \ref{tab:performance-sub}, non-linearity cases are harder to predict than heteroskedasticity cases.
-
-To assess how well the model’s decisions align with other methods, Table \ref{tab:human-conv-table} compares its rejection decisions with those of conventional tests and human visual tests.
-Agreement rates are 85.95% for heteroskedasticity cases and 79.69% for non-linearity cases, higher than those observed for human visual tests (67.96% and 66.84%, respectively).
-This suggests that the model aligns more closely with the best available conventional tests.
-However, as Figure \ref{fig:conv-mosaic} shows, the model does not always reject when conventional tests do, indicating a lower sensitivity to model violations.
-
-To further illustrate these patterns, Figure \ref{fig:pcp} presents parallel coordinate plots comparing decisions from human visual tests, the computer vision model, and conventional tests.
-All three agree in about 50% of cases.
-When humans do not reject, conventional tests reject in about half of those cases, whereas the model rejects in only 25%.
-Notably, the model rarely rejects when both humans and conventional tests do not, suggesting that while it is more sensitive than humans, its rejection behavior is more aligned with human judgment than with conventional tests.
-
-Figure \ref{fig:power} plots rejection decisions against the true distance $D$.
-Compared to conventional tests, the computer vision model rejects fewer cases when $D < 4$, but a similar number of cases when $D > 4$.
-This indicates that the model is less sensitive to minor deviations from model assumptions, while remaining comparably sensitive to more substantial violations.
-Visual tests are the least sensitive overall, showing few rejections for $D < 2$, and requiring a higher threshold ($D > 5$) for near-universal rejection, compared to $D > 4$ for both the model and conventional tests.
-
-Finally, the power curves in Figure \ref{fig:power} are fitted using logistic regression without intercepts and with an offset of $\log(0.05/0.95)$ to ensure a 5% rejection rate under $H_0$.
-The model’s rejection curve consistently lies between those of conventional and human visual tests, indicating potential to further align its decision boundary with human visual assessment.
-
-### Adjusted $\delta$-difference
-
-In the experiment conducted by @li2024plot, participants could select multiple plots per lineup.
-To reflect detection confidence, a weighted detection rate was computed for each lineup: selecting zero plots yielded a weight of 0.05; otherwise, if the true residual plot was selected, its weight was $1 / \text{(number of selections)}$.
-
-To evaluate the proposed distance measure, we could examine the relationship between the weighted detection rate and the $\delta$-difference statistic from @chowdhury2018measuring:
-
-$$
-\delta = \bar{d}_{\text{true}} - \underset{j}{\text{max}}\left(\bar{d}_{\text{null}}^{(j)}\right) \quad \text{for}~j = 1,...,m-1,
-$$
-
-\noindent where $\bar{d}_{\text{true}}$ is the mean distance between the true residual plot and all null plots, and $\bar{d}_{\text{null}}^{(j)}$ is the mean distance from null plot $j$ to other nulls.
-These averages help address asymmetries in pairwise distances.
-
-However, our method only compares each plot to a theoretically "good" residual plot, making the original formulation inapplicable.
-Moreover, since $D_{\text{null}} = 0$ by definition and $D$ can not be derived from an image, we instead focus on $\hat{D}$, the model-estimated distance.
-This leads to the adjusted $\delta$-difference:
-
-$$
-\delta_{\text{adj}} = \hat{D} - \underset{j}{\text{max}}\left(\hat{D}_{\text{null}}^{(j)}\right) \quad \text{for}~j = 1,...,m-1,
-$$
-
-\noindent where $\hat{D}_{\text{null}}^{(j)}$ is the estimated distance for the $j$-th null plot, and $m$ is the number of plots in a lineup.
-
-Figure \ref{fig:delta} displays the scatter plot of the weighted detection rate against $\delta_{\text{adj}}$.
-It shows that weighted detection generally increases with $\delta_{\text{adj}}$, especially when it is positive.
-This suggests that the differences in $\hat{D}$ capture some aspect of the visual separability between the true and null plots as perceived by humans, with larger positive differences generally indicating greater distinctiveness.
-However, the considerable variability in detection rates for similar $\delta_{\text{adj}}$ values implies that $\hat{D}$ does not fully account for all factors influencing human perception, and the alignment remains imperfect.
-A negative $\delta_{\text{adj}}$ suggests some null plots may appear more visually distinctive to humans than the true residual plot.
-In some instances, the weighted detection rate is close to one despite a negative $\delta_{\text{adj}}$.
-This discrepancy implies that the distance measure may not perfectly reflect actual human behavior.
 
@@
 
+Table \ref{tab:human-conv-table} compares the model's rejection decisions with those of conventional tests and human visual tests.
+Agreement rates are 85.95% for heteroskedasticity cases and 79.69% for non-linearity cases, higher than those observed for human visual tests (67.96% and 66.84%, respectively).
+This suggests that the model aligns more closely with the best available conventional tests.
+However, as Figure \ref{fig:conv-mosaic} shows, the model does not always reject when conventional tests do, indicating a lower sensitivity to model violations.
+
 ```{r conv-mosaic, fig.pos = "!h", fig.cap = "Rejection rate ($p$-value $\\leq0.05$) of computer vision models conditional on conventional tests on non-linearity (left) and heteroskedasticity (right) lineups displayed using a mosaic plot.
@@
 
-```{r pcp, fig.pos = "!h", fig.cap = "Parallel coordinate plots of decisions made by computer vision model, conventional tests and visual tests made by human.", eval = TRUE, fig.height = 3.5}
+To further illustrate these patterns, Figure \ref{fig:pcp} presents parallel coordinate plots comparing decisions from human visual tests, the computer vision model, and conventional tests.
+All three agree in about 50% of cases.
+When humans do not reject, conventional tests reject in about half of those cases, whereas the model rejects in only 25%.
+Notably, the model rarely rejects when both humans and conventional tests do not, suggesting that while it is more sensitive than humans, its rejection behavior is more aligned with human judgment than with conventional tests.
+
+```{r pcp, fig.pos = "!h", fig.cap = "Parallel coordinate plots of decisions made by computer vision model, conventional tests and visual tests made by human.
+All three agree in around 50% of cases.
+The model rejects less often than conventional tests when humans do not, and rarely rejects when both others do not, indicating closer alignment with human judgment.", eval = TRUE, fig.height = 3.5}
 library(ggpcp)
@@
 
+Figure \ref{fig:power} plots rejection decisions against the true distance $D$.
+Compared to conventional tests, the computer vision model rejects fewer cases when $D < 4$, but a similar number of cases when $D > 4$.
+This indicates that the model is less sensitive to minor deviations from model assumptions, while remaining comparably sensitive to more substantial violations.
+Visual tests are the least sensitive overall, showing few rejections for $D < 2$, and requiring a higher threshold ($D > 5$) for near-universal rejection, compared to $D > 4$ for both the model and conventional tests.
+
 ```{r power, fig.pos = "!h", fig.cap = "Comparison of power of visual tests, conventional tests and the computer vision model.
@@
 Marks at the top of the plot represent acceptances.
-Power curves are fitted by logistic regression models with no intercept but an offset equals to $\\text{log}(0.05/0.95)$.", fig.width = 7, fig.height = 3.5}
+Power curves are fitted by logistic regression models with no intercept but an offset equals to $\\text{log}(0.05/0.95)$.
+The model is less sensitive than conventional tests for small $D$ but similarly sensitive for large $D$.
+Visual tests are least sensitive overall.
+The model’s curve lies between those of conventional and visual tests.", fig.width = 7, fig.height = 3.5}
 
@@
 
+Finally, the power curves in Figure \ref{fig:power} are fitted using logistic regression without intercepts and with an offset of $\log(0.05/0.95)$ to ensure a 5% rejection rate under $H_0$.
+The model’s rejection curve consistently lies between those of conventional and human visual tests, indicating potential to further align its decision boundary with human visual assessment.
+
+### Adjusted $\delta$-difference
+
+In the experiment conducted by @li2024plot, participants could select multiple plots per lineup.
+To reflect detection confidence, a weighted detection rate was computed for each lineup: selecting zero plots yielded a weight of 0.05; otherwise, if the true residual plot was selected, its weight was $1 / \text{(number of selections)}$.
+
+To evaluate the proposed distance measure, we could examine the relationship between the weighted detection rate and the $\delta$-difference statistic from @chowdhury2018measuring:
+
+$$
+\delta = \bar{d}_{\text{true}} - \underset{j}{\text{max}}\left(\bar{d}_{\text{null}}^{(j)}\right) \quad \text{for}~j = 1,...,m-1,
+$$
+
+\noindent where $\bar{d}_{\text{true}}$ is the mean distance between the true residual plot and all null plots, and $\bar{d}_{\text{null}}^{(j)}$ is the mean distance from null plot $j$ to other nulls.
+These averages help address asymmetries in pairwise distances.
+
+However, our method only compares each plot to a theoretically "good" residual plot, making the original formulation inapplicable.
+Moreover, since $D_{\text{null}} = 0$ by definition and $D$ can not be derived from an image, we instead focus on $\hat{D}$, the model-estimated distance.
+This leads to the adjusted $\delta$-difference:
+
+$$
+\delta_{\text{adj}} = \hat{D} - \underset{j}{\text{max}}\left(\hat{D}_{\text{null}}^{(j)}\right) \quad \text{for}~j = 1,...,m-1,
+$$
+
+\noindent where $\hat{D}_{\text{null}}^{(j)}$ is the estimated distance for the $j$-th null plot, and $m$ is the number of plots in a lineup.
+
 ```{r delta, fig.pos = "!h", fig.cap = "A weighted detection rate vs adjusted $\\delta$-difference plot.
-The brown line is smoothing curve produced by fitting generalized additive models.", fig.width = 6, fig.height = 3, out.width = "80%"}
+The brown line is smoothing curve produced by fitting generalized additive models.
+Detection generally increases with positive $\\delta_{\\text{adj}}$, but variability and exceptions highlight the distance measure's imperfect alignment with human perception.", fig.width = 6, fig.height = 3, out.width = "80%"}
 experiment %>%
@@
 
+Figure \ref{fig:delta} displays the scatter plot of the weighted detection rate against $\delta_{\text{adj}}$.
+It shows that weighted detection generally increases with $\delta_{\text{adj}}$, especially when it is positive.
+This suggests that the differences in $\hat{D}$ capture some aspect of the visual separability between the true and null plots as perceived by humans, with larger positive differences generally indicating greater distinctiveness.
+However, the considerable variability in detection rates for similar $\delta_{\text{adj}}$ values implies that $\hat{D}$ does not fully account for all factors influencing human perception, and the alignment remains imperfect.
+A negative $\delta_{\text{adj}}$ suggests some null plots may appear more visually distinctive to humans than the true residual plot.
+In some instances, the weighted detection rate is close to one despite a negative $\delta_{\text{adj}}$.
+This discrepancy implies that the distance measure may not perfectly reflect actual human behavior.
+
 # Examples {#sec-examples}
@@
 
-```{r false-check, fig.pos = "!h", fig.cap = 'A summary of the residual plot assessment evaluted on 200 null plots and 200 bootstrapped plots.
+```{r false-check, fig.pos = "!h", fig.cap = 'A summary of the residual plot assessment evaluated on 200 null plots and 200 bootstrapped plots.
 (A) The true residual plot exhibiting a "left-triangle" shape.
-(B) The attention map produced by computing the gradient of the output with respect to the greyscale input.
-(C) The density plot of estimated distance for null plots and bootstrapped plots.
-The green area indicates the distribution of estimated distances for bootstrapped plots, while the yellow area represents the distribution of estimated distances for null plots.
-The fitted model will not be rejected since $\\hat{D} < Q_{null}(0.95)$.
-(D) The scatter plot of first two principal components of features extracted from the global pooling layer of the computer vision model.
-', fig.height = 8, out.width = "80%"}
+(B) The attention map highlights the top-right and bottom-right corners of the residual plot as the most influential.
+(C) The density plot shows estimated distances for null (yellow) and bootstrapped (green) plots.
+The fitted model is not rejected because $\\hat{D} < Q_{null}(0.95)$.
+(D) Null and bootstrapped plots cluster together in the space defined by the first two principal components of the global pooling layer.
+The cluster also covers the true residual plot.', fig.height = 8, out.width = "80%"}
 p1 <- my_vi$plot_resid(theme = theme_light()) + ggtitle(glue::glue("(A) Residual plot")) 
@@
                        theme = theme_bw(),
-                       alpha = 0.6,
+                       alpha = 0.3,
                        stroke = 0.3) +
-  ggtitle("(B) Lineup for left-triangle")
+  ggtitle("(A) Lineup for left-triangle")
 ```
@@
 
-```{r boston-check, fig.pos = "!h", fig.cap = 'A summary of the residual plot assessment for the Boston housing fitted model evaluted on 200 null plots and 200 bootstrapped plots.
+```{r boston-check, fig.pos = "!h", fig.cap = 'A summary of the residual plot assessment for the Boston housing fitted model evaluated on 200 null plots and 200 bootstrapped plots.
 (A) The true residual plot exhibiting a "U" shape.
-(B) The attention map produced by computing the gradient of the output with respect to the greyscale input.
-(C) The density plot of estimated distance for null plots and bootstrapped plots.
-The blue area indicates the distribution of estimated distances for bootstrapped plots, while the yellow area represents the distribution of estimated distances for null plots.
-The fitted model will be rejected since $\\hat{D} \\geq Q_{null}(0.95)$.
-(D) The scatter plot of first two principal components of features extracted from the global pooling layer of the computer vision model.
-', fig.height = 8, out.width = "80%"}
+(B) The attention map highlights the central region of the "U" shape as the most influential part of the residual plot.
+(C) The density plot shows estimated distances for null (yellow) and bootstrapped (green) plots.
+The fitted model is rejected because $\\hat{D} \\geq Q_{null}(0.95)$.
+(D) Bootstrapped and null plots form distinct clusters in the space of the first two principal components from the global pooling layer, highlighting their visual separability.', fig.height = 8, out.width = "80%"}
 p1 <- my_vi$plot_resid(theme = theme_light()) + ggtitle(glue::glue("(A) Residual plot")) 
@@
                        theme = theme_bw(),
-                       alpha = 0.6,
+                       alpha = 0.3,
                        stroke = 0.3) +
@@
 (A) The residual plot exhibits a "dinosaur" shape.
-(B) The attention map produced by computing the gradient of the output with respect to the greyscale input.
-(C) The density plot of estimated distance for null plots and bootstrapped plots.
-The blue area indicates the distribution of estimated distances for bootstrapped plots, while the yellow area represents the distribution of estimated distances for null plots.
-The fitted model will be rejected since $\\hat{D} \\geq Q_{null}(0.95)$.
-(D) The scatter plot of first two principal components of features extracted from the global pooling layer of the computer vision model.', fig.height = 8, out.width = "80%"}
+(B) The attention map highlights the dinosaur shape, indicating the model\'s decision is driven by human-recognizable features.(C) The density plot shows estimated distances for null (yellow) and bootstrapped (green) plots.
+The fitted model is rejected because $\\hat{D} \\geq Q_{null}(0.95)$.
+(D) The bootstrapped plots cluster at the corner of the null plot cluster, yet remain isolated in the space defined by the first two principal components of the global pooling layer.', fig.height = 8, out.width = "80%"}
 
@@
 (A) True plot at position 10; no clear visual difference.
-(B) True plot at position 7; clearly stands out.
-(C) True plot at position 17; distinctly artificial and easily identifiable.', fig.width = 10, fig.height = 10}
-patchwork::wrap_plots(l1, l2, l3, ncol = 2, widths = c(1, 1), heights = c(1, 1))
+(B) True plot at position 7; "U"-shape clearly stands out.
+(C) True plot at position 17; distinctly artificial and easily identifiable.', fig.width = 10, fig.height = 4}
+patchwork::wrap_plots(l1, l2, l3, ncol = 3, widths = c(1, 1, 1))
 ```
@@
 
-These `R` packages were used for the work: `tidyverse` [@tidyverse], `lmtest` [@lmtest], `mpoly` [@mpoly], `ggmosaic` [@ggmosaic], `kableExtra` [@kableextra], `patchwork` [@patchwork], `rcartocolor` [@rcartocolor], `glue` [@glue], `ggpcp` [@ggpcp], `here` [@here], `magick` [@magick], `yardstick` [@yardstick] and `reticulate` [@reticulate].
+These `R` packages were used for the work: `tidyverse` [@tidyverse], `lmtest` [@lmtest], `mpoly` [@mpoly], `ggmosaic` [@ggmosaic], `kableExtra` [@kableextra], `patchwork` [@patchwork], `rcartocolor` [@rcartocolor], `glue` [@glue], `ggpcp` [@ggpcp], `here` [@here], `magick` [@magick], and `reticulate` [@reticulate].
 

### 2025-07-10 paper.tex
Reduce to 36 pages

--- 2025-06-17 paper.tex
@@
 human visual tests. Several examples from classical papers and
-contemporary data demonstrate the method's paractical utility in
+contemporary data demonstrate the method's practical utility in
 automating residual diagnostics.
@@
 straightforward as executing a line of code or clicking a button.
-However, accurately interpreting a residual plot can be challenging. A
-residual plot can exhibit various visual features, but it is crucial to
-recognize that some may arise from the characteristics of predictors and
-the natural stochastic variation of the observational unit, rather than
-indicating a violation of model assumptions \citep{li2024plot}. Consider
-Figure \ref{fig:false-finding} as an example, the residual plot displays
-a triangular left-pointing shape. The distinct difference in the spread
-of the residuals across the fitted values might lead analysts to suspect
-heteroskedasticity, however, it is important to avoid over-interpreting
-this visual pattern. In this case, the fitted regression model is
-correctly specified, and the triangular shape is actually a result of
-the skewed distribution of the predictors, rather than indicating a flaw
-in the model.
+However, accurately interpreting a residual plot can be challenging, as
+some features are a result of predictor characteristics and natural
+variation, while others indicate violations of model assumptions
+\citep{li2024plot}. For example, in Figure \ref{fig:false-finding}, the
+residual plot displays a triangular left-pointing shape. The distinct
+difference in the spread of the residuals across the fitted values might
+lead analysts to suspect heteroskedasticity. In this case, however, the
+fitted regression model is correctly specified, and the triangular shape
+is actually a result of the skewed distribution of the predictors,
+rather than indicating a flaw in the model.
+
+\begin{figure}[!h]
+
+{\centering \includegraphics[width=0.5\linewidth]{paper_files/figure-latex/false-finding-1}
+
+}
+
+\caption{An example residual vs fitted values plot (red line indicates 0 corresponds to the x-intercept, i.e.
+$y=0$).
+The vertical spread of the data points varies with the fitted values.
+This often indicates the existence of heteroskedasticity, however, here the result is due to skewed distribution of the predictors rather than heteroskedasticity.
+The Breusch-Pagan test rejects this residual plot at 95\% significance level ($p\text{-value} = 0.046$).}\label{fig:false-finding}
+\end{figure}
 
@@
 A practical limitation of the lineup protocol lies in its reliance on
-human judgements \citep[see][ about the practical
-limitations]{li2024plot}. Unlike conventional statistical tests that can
-be performed computationally in statistical software, the lineup
-protocol requires human evaluation of images. This characteristic makes
-it less suitable for large-scale applications, given the associated high
-labour costs and time requirements. There is a substantial need to
-develop an approach to substitute these human judgement with an
-automated reading of data plots using machines.
+human judgements \citep[see][]{li2024plot}. Unlike conventional
+statistical tests that can be performed using in statistical software,
+the lineup protocol requires human evaluation of images. This
+characteristic makes it less suitable for large-scale applications,
+given the associated high labor costs and time requirements. There is a
+substantial need to develop an approach to substitute these human
+judgement with an automated reading of data plots using computer vision.
 
@@
 density, trend, coherence and other characteristics of the data.
-However, as noted by \citet{buja2009statistical}, such measures may not
-capture the full range of visual patterns necessary for residual
-diagnostics. A more promising direction lies in enabling computers to
-learn the relevant visual features directly mirroring how humans
-intuitively interpret plots.
-
-Modern computer vision models are well-suited for addressing this
-challenge. They rely on deep neural networks with convolutional layers
-\citep{fukushima1982neocognitron}. These layers use small, sliding
-windows to scan the image, performing a dot product to extract local
-features and patterns. Numerous studies have demonstrated the efficacy
-of convolutional layers in addressing various vision tasks, including
-image recognition \citep{rawat2017deep}. Despite the widespread use of
-computer vision models in fields like computer-aided diagnosis
-\citep{lee2015image}, pedestrian detection \citep{brunetti2018computer},
-and facial recognition \citep{emami2012facial}, their application in
-reading data plots remains limited. While some studies have explored the
-use of computer vision models for tasks such as reading recurrence plots
-for time series regression \citep{ojeda2020multivariate}, time series
-classification
+However, such measures may not capture the full range of visual patterns
+necessary for residual diagnostics \citep{buja2009statistical}. A more
+promising direction lies in enabling computers to learn the relevant
+visual features directly mirroring how humans intuitively interpret
+plots.
+
+Modern computer vision models rely on deep neural networks with
+convolutional layers \citep{fukushima1982neocognitron}. These layers use
+small, sliding windows to scan the image, performing a dot product to
+extract local features and patterns. Numerous studies have demonstrated
+the efficacy of convolutional layers in addressing various vision tasks,
+including image recognition \citep{rawat2017deep}. Despite the
+widespread use of computer vision models in fields like computer-aided
+diagnosis \citep{lee2015image}, pedestrian detection
+\citep{brunetti2018computer}, and facial recognition
+\citep{emami2012facial}, their application in reading data plots remains
+limited. While some studies have explored the use of computer vision
+models for tasks such as reading recurrence plots for time series
+regression \citep{ojeda2020multivariate}, time series classification
 \citep{chu2019automatic, hailesilassie2019financial, hatami2018classification, zhang2020encoding},
@@
 
-\begin{figure}[!h]
-
-{\centering \includegraphics[width=0.5\linewidth]{paper_files/figure-latex/false-finding-1}
-
-}
-
-\caption{An example residual vs fitted values plot (red line indicates 0 corresponds to the x-intercept, i.e.
-$y=0$).
-The vertical spread of the data points varies with the fitted values.
-This often indicates the existence of heteroskedasticity, however, here the result is due to skewed distribution of the predictors rather than heteroskedasticity.
-The Breusch-Pagan test rejects this residual plot at 95\% significance level ($p\text{-value} = 0.046$).}\label{fig:false-finding}
-\end{figure}
-
 \section{Model Specifications}\label{sec-model-specifications}
@@
 be used to assess residual plots. We discuss these specifications below
-focusing on two key components of the model formula: the input and the
-output format.
+focusing on the input and the output formats.
 
@@
 Another strategy is to convert residual plots into images. This enables
-the use of standard convolutional architectures such as VGG16
-\citep{simonyan2014very}, which can effectively capture spatial
-patterns. Although discretization introduces some losses in detail,
-image-based inputs offer a richer representation by preserving the joint
-distribution of residuals and fitted values in a visually interpretable
-format. To ensure model generalization, it is essential to use residual
-plots in a consistent style, i.e., same aesthetics, layout, scaling and
-color schemes.
+the use of standard convolutional architectures in computer vision such
+as VGG16 \citep{simonyan2014very}. Image-based inputs offer a richer
+representation by preserving the joint distribution of residuals and
+fitted values in a visually interpretable format, helping models more
+effectively capture spatial structure and local correlations than
+vector-based formats. To ensure model generalization, it is essential to
+use residual plots in a consistent style, i.e., same aesthetics, layout,
+scaling and color schemes.
 
@@
 \citep{chopra2005learning} can assess visual similarity if we assume
-null plots are more similar from each other than from the true residual
+null plots are more similar to each other than to the true residual
 plot, which may help distinguish null plots from true residual plots.
@@
 
-In this study, we defined and used a continuous distance between a true
+In this study, we define and use a continuous distance between a true
 residual plot and a theoretically ``good'' residual plot. This approach
@@
 \begin{align*} \label{eq:kl-3}
-D_{KL} &\approx \sum_{i = 1}^{n} \hat{D}_{KL}^{(i)}, \\
-\hat{D}_{KL}^{(i)} &= \frac{1}{m}\sum_{j = 1}^{m} \log\frac{\hat{p}_i(B_{ij})}{q(B_{ij})},
+D_{KL} &\approx \sum_{i = 1}^{n} \hat{D}_{KL}^{(i)}, \quad
+\hat{D}_{KL}^{(i)} = \frac{1}{m}\sum_{j = 1}^{m} \log\frac{\hat{p}_i(B_{ij})}{q(B_{ij})},
 \end{align*}
@@
 Kullback-Leibler divergence for an individual residual \(e_i\),
-\(\boldsymbol{B}_{ij}\) is the \(i\)-th row and \(j\)-th column entry of
-the matrix \(\boldsymbol{B}\), \(\hat{p}_i(.)\) is the kernel density
-estimator of \(p_i(.)\), \(q(.)\) is the normal density function with
-mean zero and an assumed variance estimated as
+\(B_{ij}\) is the \(i\)-th row and \(j\)-th column entry of the matrix
+\(\boldsymbol{B}\), \(\hat{p}_i(.)\) is the kernel density estimator of
+\(p_i(.)\), \(q(.)\) is the normal density function with mean zero and
+an assumed variance estimated as
 \(\hat{\sigma}^2 = \sum_{b \in vec(\boldsymbol{B})}(b - \sum_{b \in vec(\boldsymbol{B})} b/nm)^2/(nm - 1)\),
@@
 
-\begin{equation} \label{eq:d-approx}
+\begin{equation*} \label{eq:d-approx}
 \hat{D} = f_{CV}(V_{h \times w}(\boldsymbol{e}, \hat{\boldsymbol{y}})),
-\end{equation}
-
-\noindent where \(V_{h \times w}(.)\) renders the residuals vs fitted
+\end{equation*}
+
+\noindent where \(V_{h \times w}(.)\) renders the residuals vs.~fitted
 values plot as an \(h \times w\) RGB image, and \(f_{CV}(.)\) is a model
@@
 
-\begin{figure}[!h]
-
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/cnn-diag-1}
-
-}
-
-\caption{Diagram of the architecture of the optimized computer vision model.
-Numbers at the bottom of each box show the shape of the output of each layer.
-The band of each box drawn in a darker color indicates the use of the rectified linear unit activation function.
-Yellow boxes are 2D convolutional layers, orange boxes are pooling layers, the grey box is the concatenation layer, and the purple boxes are dense layers.}\label{fig:cnn-diag}
-\end{figure}
-
 The architecture of the computer vision model \(f_{CV}\) is adapted from
@@
 Appendix B.
+
+\begin{figure}[!h]
+
+{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/cnn-diag-1}
+
+}
+
+\caption{Diagram of the architecture of the optimized computer vision model.
+Numbers at the bottom of each box show the shape of the output of each layer.
+The band of each box drawn in a darker color indicates the use of the rectified linear unit activation function.
+Yellow boxes are 2D convolutional layers, orange boxes are pooling layers, the grey box is the concatenation layer, and the purple boxes are dense layers.}\label{fig:cnn-diag}
+\end{figure}
 
@@
 using the Glorot Uniform method \citep{glorot2010understanding} and they
-were optimized by the Adam optimizer \citep{kingma2014adam} with the
-mean square error loss function
-
-\[\hat{\boldsymbol{\theta}} = \underset{\boldsymbol{\theta}}{\text{arg min}}\frac{1}{n_{\text{train}}}\sum_{i=1}^{n_{\text{train}}}(D_i - f_{\boldsymbol{\theta}}(V_i, S_i))^2,\]
-
-\noindent where \(n_{\text{train}}\) is the number of training samples,
-\(V_i\) is the \(i\)-th residual plot and \(S_i\) is the auxiliary
-information about the \(i\)-th residual plot including four scagnostics
-and the number of observations.
+were optimized by the Adam optimizer \citep{kingma2014adam} with the MSE
+loss function
+\(\hat{\boldsymbol{\theta}} = \underset{\boldsymbol{\theta}}{\text{arg min}}\frac{1}{n_{\text{train}}}\sum_{i=1}^{n_{\text{train}}}(D_i - f_{\boldsymbol{\theta}}(V_i, S_i))^2\),
+where \(n_{\text{train}}\) is the number of training samples, \(V_i\) is
+the \(i\)-th residual plot and \(S_i\) is the auxiliary information
+about the \(i\)-th residual plot including four scagnostics and the
+number of observations.
 
@@
 The brown lines are smoothing curves produced by fitting generalized additive models.
-The area over the zero line in light yellow indicates under-prediction, and the area under the zero line in light green indicates over-prediction.}\label{fig:model-performance}
+Over-prediction and under-prediction can be observed for small $D$ and large $D$ respectively.}\label{fig:model-performance}
 \end{figure}
@@
 The brown lines are smoothing curves produced by fitting generalized additive models.
-The area over the zero line in light yellow indicates under-prediction, and the area under the zero line in light green indicates over-prediction.}\label{fig:over-under}
+Under-prediction mainly occurs when the data-generating process has small $\sigma$, a second predictor, and only non-linearity as the model violation.}\label{fig:over-under}
 \end{figure}
@@
 
+Each lineup in \citet{li2024plot} contains one true residual plot and 19
+null plots. While the true residual plot's distance \(D\) depends on the
+data-generating process, null plots have \(D = 0\). We used our
+optimized computer vision model to estimate \(D\) for all plots. To
+ensure fair comparison, we reject \(H_0\) if the true residual plot has
+the highest estimated distance. Conventional tests, including the Ramsey
+RESET \citep{ramsey1969tests} for non-linearity and the Breusch-Pagan
+test \citep{breusch1979simple} for heteroskedasticity, were also applied
+to the same data.
+
+Table \ref{tab:experiment-performance} reports performance metrics for
+\(\hat{D}\) on true residual plots. While the performance is slightly
+worse than on the test data provided in Table \ref{tab:performance}, the
+mean absolute error remains low and the correlation between predictions
+and true values remains high. Consistent with results in Table
+\ref{tab:performance-sub}, non-linearity cases are harder to predict
+than heteroskedasticity cases.
+
 \begin{table}
@@
 \end{table}
-
-Each lineup in \citet{li2024plot} contains one true residual plot and 19
-null plots. While the true residual plot's distance \(D\) depends on the
-data-generating process, null plots have \(D = 0\). We used our
-optimized computer vision model to estimate \(D\) for all plots. To
-ensure fair comparison, we reject \(H_0\) if the true residual plot has
-the highest estimated distance. Conventional tests, including the Ramsey
-RESET \citep{ramsey1969tests} for non-linearity and the Breusch-Pagan
-test \citep{breusch1979simple} for heteroskedasticity, were also applied
-to the same data.
-
-Table \ref{tab:experiment-performance} reports performance metrics for
-\(\hat{D}\) on true residual plots. While the performance is slightly
-worse than on the test data provided in Table \ref{tab:performance}, the
-mean absolute error remains low and the correlation between predictions
-and true values remains high. Consistent with results in Table
-\ref{tab:performance-sub}, non-linearity cases are harder to predict
-than heteroskedasticity cases.
-
-To assess how well the model's decisions align with other methods, Table
-\ref{tab:human-conv-table} compares its rejection decisions with those
-of conventional tests and human visual tests. Agreement rates are
-85.95\% for heteroskedasticity cases and 79.69\% for non-linearity
-cases, higher than those observed for human visual tests (67.96\% and
-66.84\%, respectively). This suggests that the model aligns more closely
-with the best available conventional tests. However, as Figure
-\ref{fig:conv-mosaic} shows, the model does not always reject when
-conventional tests do, indicating a lower sensitivity to model
-violations.
-
-To further illustrate these patterns, Figure \ref{fig:pcp} presents
-parallel coordinate plots comparing decisions from human visual tests,
-the computer vision model, and conventional tests. All three agree in
-about 50\% of cases. When humans do not reject, conventional tests
-reject in about half of those cases, whereas the model rejects in only
-25\%. Notably, the model rarely rejects when both humans and
-conventional tests do not, suggesting that while it is more sensitive
-than humans, its rejection behavior is more aligned with human judgment
-than with conventional tests.
-
-Figure \ref{fig:power} plots rejection decisions against the true
-distance \(D\). Compared to conventional tests, the computer vision
-model rejects fewer cases when \(D < 4\), but a similar number of cases
-when \(D > 4\). This indicates that the model is less sensitive to minor
-deviations from model assumptions, while remaining comparably sensitive
-to more substantial violations. Visual tests are the least sensitive
-overall, showing few rejections for \(D < 2\), and requiring a higher
-threshold (\(D > 5\)) for near-universal rejection, compared to
-\(D > 4\) for both the model and conventional tests.
-
-Finally, the power curves in Figure \ref{fig:power} are fitted using
-logistic regression without intercepts and with an offset of
-\(\log(0.05/0.95)\) to ensure a 5\% rejection rate under \(H_0\). The
-model's rejection curve consistently lies between those of conventional
-and human visual tests, indicating potential to further align its
-decision boundary with human visual assessment.
-
-\subsubsection{\texorpdfstring{Adjusted
-\(\delta\)-difference}{Adjusted \textbackslash delta-difference}}\label{adjusted-delta-difference}
-
-In the experiment conducted by \citet{li2024plot}, participants could
-select multiple plots per lineup. To reflect detection confidence, a
-weighted detection rate was computed for each lineup: selecting zero
-plots yielded a weight of 0.05; otherwise, if the true residual plot was
-selected, its weight was \(1 / \text{(number of selections)}\).
-
-To evaluate the proposed distance measure, we could examine the
-relationship between the weighted detection rate and the
-\(\delta\)-difference statistic from \citet{chowdhury2018measuring}:
-
-\[
-\delta = \bar{d}_{\text{true}} - \underset{j}{\text{max}}\left(\bar{d}_{\text{null}}^{(j)}\right) \quad \text{for}~j = 1,...,m-1,
-\]
-
-\noindent where \(\bar{d}_{\text{true}}\) is the mean distance between
-the true residual plot and all null plots, and
-\(\bar{d}_{\text{null}}^{(j)}\) is the mean distance from null plot
-\(j\) to other nulls. These averages help address asymmetries in
-pairwise distances.
-
-However, our method only compares each plot to a theoretically ``good''
-residual plot, making the original formulation inapplicable. Moreover,
-since \(D_{\text{null}} = 0\) by definition and \(D\) can not be derived
-from an image, we instead focus on \(\hat{D}\), the model-estimated
-distance. This leads to the adjusted \(\delta\)-difference:
-
-\[
-\delta_{\text{adj}} = \hat{D} - \underset{j}{\text{max}}\left(\hat{D}_{\text{null}}^{(j)}\right) \quad \text{for}~j = 1,...,m-1,
-\]
-
-\noindent where \(\hat{D}_{\text{null}}^{(j)}\) is the estimated
-distance for the \(j\)-th null plot, and \(m\) is the number of plots in
-a lineup.
-
-Figure \ref{fig:delta} displays the scatter plot of the weighted
-detection rate against \(\delta_{\text{adj}}\). It shows that weighted
-detection generally increases with \(\delta_{\text{adj}}\), especially
-when it is positive. This suggests that the differences in \(\hat{D}\)
-capture some aspect of the visual separability between the true and null
-plots as perceived by humans, with larger positive differences generally
-indicating greater distinctiveness. However, the considerable
-variability in detection rates for similar \(\delta_{\text{adj}}\)
-values implies that \(\hat{D}\) does not fully account for all factors
-influencing human perception, and the alignment remains imperfect. A
-negative \(\delta_{\text{adj}}\) suggests some null plots may appear
-more visually distinctive to humans than the true residual plot. In some
-instances, the weighted detection rate is close to one despite a
-negative \(\delta_{\text{adj}}\). This discrepancy implies that the
-distance measure may not perfectly reflect actual human behavior.
 
@@
 
+Table \ref{tab:human-conv-table} compares the model's rejection
+decisions with those of conventional tests and human visual tests.
+Agreement rates are 85.95\% for heteroskedasticity cases and 79.69\% for
+non-linearity cases, higher than those observed for human visual tests
+(67.96\% and 66.84\%, respectively). This suggests that the model aligns
+more closely with the best available conventional tests. However, as
+Figure \ref{fig:conv-mosaic} shows, the model does not always reject
+when conventional tests do, indicating a lower sensitivity to model
+violations.
+
 \begin{figure}[!h]
@@
 
+To further illustrate these patterns, Figure \ref{fig:pcp} presents
+parallel coordinate plots comparing decisions from human visual tests,
+the computer vision model, and conventional tests. All three agree in
+about 50\% of cases. When humans do not reject, conventional tests
+reject in about half of those cases, whereas the model rejects in only
+25\%. Notably, the model rarely rejects when both humans and
+conventional tests do not, suggesting that while it is more sensitive
+than humans, its rejection behavior is more aligned with human judgment
+than with conventional tests.
+
 \begin{figure}[!h]
@@
 
-\caption{Parallel coordinate plots of decisions made by computer vision model, conventional tests and visual tests made by human.}\label{fig:pcp}
+\caption{Parallel coordinate plots of decisions made by computer vision model, conventional tests and visual tests made by human.
+All three agree in around 50\% of cases.
+The model rejects less often than conventional tests when humans do not, and rarely rejects when both others do not, indicating closer alignment with human judgment.}\label{fig:pcp}
 \end{figure}
+
+Figure \ref{fig:power} plots rejection decisions against the true
+distance \(D\). Compared to conventional tests, the computer vision
+model rejects fewer cases when \(D < 4\), but a similar number of cases
+when \(D > 4\). This indicates that the model is less sensitive to minor
+deviations from model assumptions, while remaining comparably sensitive
+to more substantial violations. Visual tests are the least sensitive
+overall, showing few rejections for \(D < 2\), and requiring a higher
+threshold (\(D > 5\)) for near-universal rejection, compared to
+\(D > 4\) for both the model and conventional tests.
 
@@
 Marks at the top of the plot represent acceptances.
-Power curves are fitted by logistic regression models with no intercept but an offset equals to $\text{log}(0.05/0.95)$.}\label{fig:power}
+Power curves are fitted by logistic regression models with no intercept but an offset equals to $\text{log}(0.05/0.95)$.
+The model is less sensitive than conventional tests for small $D$ but similarly sensitive for large $D$.
+Visual tests are least sensitive overall.
+The model’s curve lies between those of conventional and visual tests.}\label{fig:power}
 \end{figure}
 
+Finally, the power curves in Figure \ref{fig:power} are fitted using
+logistic regression without intercepts and with an offset of
+\(\log(0.05/0.95)\) to ensure a 5\% rejection rate under \(H_0\). The
+model's rejection curve consistently lies between those of conventional
+and human visual tests, indicating potential to further align its
+decision boundary with human visual assessment.
+
+\subsubsection{\texorpdfstring{Adjusted
+\(\delta\)-difference}{Adjusted \textbackslash delta-difference}}\label{adjusted-delta-difference}
+
+In the experiment conducted by \citet{li2024plot}, participants could
+select multiple plots per lineup. To reflect detection confidence, a
+weighted detection rate was computed for each lineup: selecting zero
+plots yielded a weight of 0.05; otherwise, if the true residual plot was
+selected, its weight was \(1 / \text{(number of selections)}\).
+
+To evaluate the proposed distance measure, we could examine the
+relationship between the weighted detection rate and the
+\(\delta\)-difference statistic from \citet{chowdhury2018measuring}:
+
+\[
+\delta = \bar{d}_{\text{true}} - \underset{j}{\text{max}}\left(\bar{d}_{\text{null}}^{(j)}\right) \quad \text{for}~j = 1,...,m-1,
+\]
+
+\noindent where \(\bar{d}_{\text{true}}\) is the mean distance between
+the true residual plot and all null plots, and
+\(\bar{d}_{\text{null}}^{(j)}\) is the mean distance from null plot
+\(j\) to other nulls. These averages help address asymmetries in
+pairwise distances.
+
+However, our method only compares each plot to a theoretically ``good''
+residual plot, making the original formulation inapplicable. Moreover,
+since \(D_{\text{null}} = 0\) by definition and \(D\) can not be derived
+from an image, we instead focus on \(\hat{D}\), the model-estimated
+distance. This leads to the adjusted \(\delta\)-difference:
+
+\[
+\delta_{\text{adj}} = \hat{D} - \underset{j}{\text{max}}\left(\hat{D}_{\text{null}}^{(j)}\right) \quad \text{for}~j = 1,...,m-1,
+\]
+
+\noindent where \(\hat{D}_{\text{null}}^{(j)}\) is the estimated
+distance for the \(j\)-th null plot, and \(m\) is the number of plots in
+a lineup.
+
 \begin{figure}[!h]
@@
 
-\caption{A weighted detection rate vs adjusted $\delta$-difference plot. The brown line is smoothing curve produced by fitting generalized additive models.}\label{fig:delta}
+\caption{A weighted detection rate vs adjusted $\delta$-difference plot.
+The brown line is smoothing curve produced by fitting generalized additive models.
+Detection generally increases with positive $\delta_{\text{adj}}$, but variability and exceptions highlight the distance measure's imperfect alignment with human perception.}\label{fig:delta}
 \end{figure}
+
+Figure \ref{fig:delta} displays the scatter plot of the weighted
+detection rate against \(\delta_{\text{adj}}\). It shows that weighted
+detection generally increases with \(\delta_{\text{adj}}\), especially
+when it is positive. This suggests that the differences in \(\hat{D}\)
+capture some aspect of the visual separability between the true and null
+plots as perceived by humans, with larger positive differences generally
+indicating greater distinctiveness. However, the considerable
+variability in detection rates for similar \(\delta_{\text{adj}}\)
+values implies that \(\hat{D}\) does not fully account for all factors
+influencing human perception, and the alignment remains imperfect. A
+negative \(\delta_{\text{adj}}\) suggests some null plots may appear
+more visually distinctive to humans than the true residual plot. In some
+instances, the weighted detection rate is close to one despite a
+negative \(\delta_{\text{adj}}\). This discrepancy implies that the
+distance measure may not perfectly reflect actual human behavior.
 
@@
 
-\caption{A summary of the residual plot assessment evaluted on 200 null plots and 200 bootstrapped plots.
+\caption{A summary of the residual plot assessment evaluated on 200 null plots and 200 bootstrapped plots.
 (A) The true residual plot exhibiting a "left-triangle" shape.
-(B) The attention map produced by computing the gradient of the output with respect to the greyscale input.
-(C) The density plot of estimated distance for null plots and bootstrapped plots.
-The green area indicates the distribution of estimated distances for bootstrapped plots, while the yellow area represents the distribution of estimated distances for null plots.
-The fitted model will not be rejected since $\hat{D} < Q_{null}(0.95)$.
-(D) The scatter plot of first two principal components of features extracted from the global pooling layer of the computer vision model.
-}\label{fig:false-check}
+(B) The attention map highlights the top-right and bottom-right corners of the residual plot as the most influential.
+(C) The density plot shows estimated distances for null (yellow) and bootstrapped (green) plots.
+The fitted model is not rejected because $\hat{D} < Q_{null}(0.95)$.
+(D) Null and bootstrapped plots cluster together in the space defined by the first two principal components of the global pooling layer.
+The cluster also covers the true residual plot.}\label{fig:false-check}
 \end{figure}
@@
 
-\caption{A summary of the residual plot assessment for the Boston housing fitted model evaluted on 200 null plots and 200 bootstrapped plots.
+\caption{A summary of the residual plot assessment for the Boston housing fitted model evaluated on 200 null plots and 200 bootstrapped plots.
 (A) The true residual plot exhibiting a "U" shape.
-(B) The attention map produced by computing the gradient of the output with respect to the greyscale input.
-(C) The density plot of estimated distance for null plots and bootstrapped plots.
-The blue area indicates the distribution of estimated distances for bootstrapped plots, while the yellow area represents the distribution of estimated distances for null plots.
-The fitted model will be rejected since $\hat{D} \geq Q_{null}(0.95)$.
-(D) The scatter plot of first two principal components of features extracted from the global pooling layer of the computer vision model.
-}\label{fig:boston-check}
+(B) The attention map highlights the central region of the "U" shape as the most influential part of the residual plot.
+(C) The density plot shows estimated distances for null (yellow) and bootstrapped (green) plots.
+The fitted model is rejected because $\hat{D} \geq Q_{null}(0.95)$.
+(D) Bootstrapped and null plots form distinct clusters in the space of the first two principal components from the global pooling layer, highlighting their visual separability.}\label{fig:boston-check}
 \end{figure}
@@
 (A) The residual plot exhibits a "dinosaur" shape.
-(B) The attention map produced by computing the gradient of the output with respect to the greyscale input.
-(C) The density plot of estimated distance for null plots and bootstrapped plots.
-The blue area indicates the distribution of estimated distances for bootstrapped plots, while the yellow area represents the distribution of estimated distances for null plots.
-The fitted model will be rejected since $\hat{D} \geq Q_{null}(0.95)$.
-(D) The scatter plot of first two principal components of features extracted from the global pooling layer of the computer vision model.}\label{fig:dino-check}
+(B) The attention map highlights the dinosaur shape, indicating the model's decision is driven by human-recognizable features.(C) The density plot shows estimated distances for null (yellow) and bootstrapped (green) plots.
+The fitted model is rejected because $\hat{D} \geq Q_{null}(0.95)$.
+(D) The bootstrapped plots cluster at the corner of the null plot cluster, yet remain isolated in the space defined by the first two principal components of the global pooling layer.}\label{fig:dino-check}
 \end{figure}
@@
 (A) True plot at position 10; no clear visual difference.
-(B) True plot at position 7; clearly stands out.
+(B) True plot at position 7; "U"-shape clearly stands out.
 (C) True plot at position 17; distinctly artificial and easily identifiable.}\label{fig:lineup}
@@
 \texttt{ggpcp} \citep{ggpcp}, \texttt{here} \citep{here},
-\texttt{magick} \citep{magick}, \texttt{yardstick} \citep{yardstick} and
-\texttt{reticulate} \citep{reticulate}.
+\texttt{magick} \citep{magick}, and \texttt{reticulate}
+\citep{reticulate}.
 

### 2025-07-11 paper.Rmd
fix typo in keyword

@@
 keywords: |
-  statistical graphics; data visualization; visual inference; computer vision; machine learning; hypothesis testing; reression analysis; cognitive perception; simulation; practical significance
+  statistical graphics; data visualization; visual inference; computer vision; machine learning; hypothesis testing; regression analysis; cognitive perception; simulation; practical significance
 header-includes: |

### 2025-07-11 paper.Rmd
fix small things in main paper

@@
 The concept of visual inference, as proposed by @buja2009statistical, provides an inferential framework to assess whether residual plots indeed contain visual patterns inconsistent with the model assumptions.
-The fundamental idea involves testing whether the true residual plot visually differs significantly from null plots, where null plots are plotted with residuals generated from the residual rotation distribution [@langsrud2005rotation], which is a distribution consistent with the null hypothesis $H_0$ that the linear regression model is correctly specified.
+The fundamental idea involves testing whether the true residual plot visually differs significantly from the null plots, where null plots are plotted with residuals generated from the residual rotation distribution [@langsrud2005rotation], which is a distribution consistent with the null hypothesis $H_0$ that the linear regression model is correctly specified.
 Typically, the visual test is accomplished through the lineup protocol, where the true residual plot is embedded within a lineup alongside several null plots.
@@
 Deep learning models are sensitive to input design, and several alternatives are available for encoding residual plots.
-
 A simple approach involves feeding vectors of residuals and fitted values into the model.
@@
 We experimented with this setting in a pilot study, and found multiple residual plots led to high input resolution, high computational cost and suboptimal model performance.
-
 Balancing accuracy, interpretability, and implementation cost, we adopt a single residual plot in image format as the model input.
@@
 Figure \ref{fig:model-performance} shows a hexagonal heatmap of $D - \hat{D}$ versus $D$.
-Smoothing curves (fitted by generalized additive models [@hastie2017generalized]) reveal that all models performed well for $1.5 < D < 6$, where no structural issues were observed.
+Smoothing curves [fitted by generalized additive models, @hastie2017generalized] reveal that all models performed well for $1.5 < D < 6$, where no structural issues were observed.
 Over-prediction occurred when $D < 1.5$, and under-prediction primarily when $\hat{D} > 6$.
@@
 In the experiment conducted by @li2024plot, participants could select multiple plots per lineup.
-To reflect detection confidence, a weighted detection rate was computed for each lineup: selecting zero plots yielded a weight of 0.05; otherwise, if the true residual plot was selected, its weight was $1 / \text{(number of selections)}$.
+To reflect detection confidence, a weighted detection rate was computed for each lineup: selecting zero plots yielded a weight of 0.05; otherwise, if the true residual plot was selected, its weight was the inverse of the number of selections.
 

### 2025-07-11 appendix.tex
recompile paper + appendix

@@
 
-\author{\name{Weihao Li$^{a, b}$, Dianne Cook$^{a}$, Emi Tanaka$^{a, b,
-c}$, Susan VanderPlas$^{d}$, Klaus Ackermann$^{a}$}
+\author{\name{Weihao Li$^{a, b}$, Dianne Cook$^{a}$, Emi
+Tanaka$^{b}$, Susan VanderPlas$^{c}$, Klaus Ackermann$^{a}$}
 \affil{$^{a}$Department of Econometrics and Business Statistics, Monash
-University, Clayton, VIC, Australia; $^{b}$Biological Data Science
-Institute, Australian National University, Acton, ACT,
-Australia; $^{c}$Research School of Finance, Actuarial Studies and
-Statistics, Australian National University, Acton, ACT,
-Australia; $^{d}$Department of Statistics, University of Nebraska,
+University, Clayton, VIC, Australia; $^{b}$Research School of Finance,
+Actuarial Studies and Statistics, Australian National University, Acton,
+ACT, Australia; $^{c}$Department of Statistics, University of Nebraska,
 Lincoln, Nebraska, USA}

### 2025-07-11 paper.tex
recompile paper + appendix

@@
 statistical graphics; data visualization; visual inference; computer
-vision; machine learning; hypothesis testing; reression analysis;
+vision; machine learning; hypothesis testing; regression analysis;
 cognitive perception; simulation; practical significance
@@
 the model assumptions. The fundamental idea involves testing whether the
-true residual plot visually differs significantly from null plots, where
-null plots are plotted with residuals generated from the residual
+true residual plot visually differs significantly from the null plots,
+where null plots are plotted with residuals generated from the residual
 rotation distribution \citep{langsrud2005rotation}, which is a
@@
 Deep learning models are sensitive to input design, and several
-alternatives are available for encoding residual plots.
-
-A simple approach involves feeding vectors of residuals and fitted
-values into the model. While this format contains all relevant
-information, the input length varies with sample size, which conflicts
-with the fixed-shape requirements of modern vision models such as those
+alternatives are available for encoding residual plots. A simple
+approach involves feeding vectors of residuals and fitted values into
+the model. While this format contains all relevant information, the
+input length varies with sample size, which conflicts with the
+fixed-shape requirements of modern vision models such as those
 implemented in TensorFlow \citep{abadi2016tensorflow}. Padding can be
@@
 input resolution, high computational cost and suboptimal model
-performance.
-
-Balancing accuracy, interpretability, and implementation cost, we adopt
-a single residual plot in image format as the model input.
+performance. Balancing accuracy, interpretability, and implementation
+cost, we adopt a single residual plot in image format as the model
+input.
 
@@
 Figure \ref{fig:model-performance} shows a hexagonal heatmap of
-\(D - \hat{D}\) versus \(D\). Smoothing curves (fitted by generalized
-additive models \citep{hastie2017generalized}) reveal that all models
-performed well for \(1.5 < D < 6\), where no structural issues were
-observed. Over-prediction occurred when \(D < 1.5\), and
+\(D - \hat{D}\) versus \(D\). Smoothing curves \citep[fitted by
+generalized additive models,][]{hastie2017generalized} reveal that all
+models performed well for \(1.5 < D < 6\), where no structural issues
+were observed. Over-prediction occurred when \(D < 1.5\), and
 under-prediction primarily when \(\hat{D} > 6\).
@@
 plots yielded a weight of 0.05; otherwise, if the true residual plot was
-selected, its weight was \(1 / \text{(number of selections)}\).
+selected, its weight was the inverse of the number of selections.
 

### 2025-07-25 cover_letter/template.qmd
Upload cover letter

---
format: quarto-anu-letter-pdf
author:
  - name: "\u200B" # I need to sign this
    position: Postdoctoral Research Fellow 
    affiliation:
      - name: Australian National University
        department: Research School of Finance, Actuarial Studies and Statistics
        address: "26C Kingsley Street"  
        city: Canberra
        country: Australia
        postal-code: 2600
address:
  - Editors
  - Journal of Computational and Graphical Statistics
subject: "Submission of Manuscript: Automated Assessment of Residual Plots with Computer Vision Models"
# subject-title: "Editors"
title: Letter from ANU
opening: "Dear Prof. Chen and Prof. Sangalli,"
closing: "Sincerely,"
keep-tex: true
---

Please consider our manuscript "Automated Assessment of Residual Plots with Computer Vision Models" for publication in the Journal of Computational and Graphical Statistics.

This paper builds on our previous work, "A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol" (JCGS), which provided empirical evidence for the effectiveness of visual evaluation of residual plots through crowd-sourced lineup experiments. While that study highlighted the cost and effort of human evaluations, this work advances automation through computer vision. We introduce a model trained to predict a custom distance measure that quantifies the discrepancy between residuals and theoretically "good" distributions. Building on these predictions, we construct a statistical testing framework aligned with the original lineup protocol, ensuring a valid visual test. When evaluated on the earlier experiment data, the model performs slightly worse than humans but surpasses conventional statistical tests due to its reduced sensitivity to minor deviations from model assumptions. We further demonstrate that the model, like human observers, relies on visual patterns and shapes, making it a practical tool for reducing manual effort in residual plot diagnostics. This approach should be of particular interest to readers concerned with model diagnostics and the automation of visual data analysis.

Thank you for the consideration of this manuscript. We believe that it is a good fit for the Journal of Computational and Graphical Statistics.

### 2025-07-25 paper.Rmd
Upload cover letter

@@
 
-Consistent with this, the computer vision model assigns a distance $\hat{D}$ above the 95% quantile of the null distribution ($p-\text{value} = 0.00498$), leading to rejection of $H_0$.
+Consistent with this, the computer vision model assigns a distance $\hat{D}$ above the 95% quantile of the null distribution ($p\text{-value} = 0.00498$), leading to rejection of $H_0$.
 The bootstrapped distribution further supports this, with most fitted models also being rejected.

### 2025-07-25 cover_letter/template.qmd
update letter

@@
 This paper builds on our previous work, "A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol" (JCGS), which provided empirical evidence for the effectiveness of visual evaluation of residual plots through crowd-sourced lineup experiments.
-While that study highlighted the cost and effort of human evaluations, this work advances automation through computer vision.
-We introduce a model trained to predict a custom distance measure that quantifies the discrepancy between residuals and theoretically "good" distributions.
+While that study highlighted the cost and effort of human evaluations, this work advances automation of residual plot diagnostics through an application of a computer vision model.
+We specifically trained a model to predict a custom distance measure that quantifies the discrepancy between residuals and theoretically "good" distributions.
 Building on these predictions, we construct a statistical testing framework aligned with the original lineup protocol, ensuring a valid visual test.
-When evaluated on the earlier experiment data, the model performs slightly worse than humans but surpasses conventional statistical tests due to its reduced sensitivity to minor deviations from model assumptions.
+When evaluated on the earlier experimenal data, the model performs slightly worse than humans but surpasses conventional statistical tests due to its reduced sensitivity to minor deviations from model assumptions.
 We further demonstrate that the model, like human observers, relies on visual patterns and shapes, making it a practical tool for reducing manual effort in residual plot diagnostics.
-This approach should be of particular interest to readers concerned with model diagnostics and the automation of visual data analysis.
+This approach would be of particular interest to readers concerned with model diagnostics and the automation of visual data analysis.
 

### 2025-07-25 cover_letter/template.tex
update letter

% Options for packages loaded elsewhere
% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  12pt,
  letterpaper]{letter}
\usepackage{xcolor}
\usepackage[margin=1in,bottom=1in,left=1in,right=1in]{geometry}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother

\setlength{\emergencystretch}{3em} % prevent overfull lines

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{eso-pic} % for background image
\usepackage{graphicx} % necessary for images
\usepackage{xifthen}
\usepackage{parskip} % for spacing
\usepackage{fancyhdr} % for fancy headers
\usepackage{lastpage} 
\usepackage{bera} % for font
\pagestyle{fancy}
\renewcommand{\headrulewidth}{0pt}
\fancyhead{} % footer for pages 2 on
\lfoot{} % Left footer
\cfoot{} % Change for center footer
\rfoot{\footnotesize Page \thepage\ of \pageref{LastPage}} % Right footer page #s 
\renewcommand{\footrulewidth}{0pt}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Letter from ANU},
  pdfauthor={\hspace{0pt}},
  pdfsubject={Submission of Manuscript: Automated Assessment of Residual
Plots with Computer Vision Models},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Letter from ANU}
\author{\hspace{0pt}}
\date{July 25, 2025}
\begin{document}
\AddToShipoutPicture{\ifthenelse{\value{page}<2}{\includegraphics[width=\paperwidth,height=\paperheight]{\_extensions/anuopensci/quarto-anu-letter/assets/images/coverpage.jpg}}{\includegraphics[width=\paperwidth,height=\paperheight]{\_extensions/anuopensci/quarto-anu-letter/assets/images/page-background.jpg}}}

\signature{
\hspace{0pt}}
\address{Australian National University\\Research School of Finance,
Actuarial Studies and Statistics\\26C Kingsley
Street\\Canberra 2600\\Australia\\[2mm]}
\begin{letter}{Editors\\Journal of Computational and Graphical
Statistics\\ ~ \\Subject: Submission of Manuscript: Automated Assessment
of Residual Plots with Computer Vision Models}
\opening{Dear Prof.~Chen and Prof.~Sangalli,}

Please consider our manuscript ``Automated Assessment of Residual Plots
with Computer Vision Models'' for publication in the Journal of
Computational and Graphical Statistics.

This paper builds on our previous work, ``A Plot is Worth a Thousand
Tests: Assessing Residual Diagnostics with the Lineup Protocol'' (JCGS),
which provided empirical evidence for the effectiveness of visual
evaluation of residual plots through crowd-sourced lineup experiments.
While that study highlighted the cost and effort of human evaluations,
this work advances automation of residual plot diagnostics through an
application of a computer vision model. We specifically trained a model
to predict a custom distance measure that quantifies the discrepancy
between residuals and theoretically ``good'' distributions. Building on
these predictions, we construct a statistical testing framework aligned
with the original lineup protocol, ensuring a valid visual test. When
evaluated on the earlier experimenal data, the model performs slightly
worse than humans but surpasses conventional statistical tests due to
its reduced sensitivity to minor deviations from model assumptions. We
further demonstrate that the model, like human observers, relies on
visual patterns and shapes, making it a practical tool for reducing
manual effort in residual plot diagnostics. This approach would be of
particular interest to readers concerned with model diagnostics and the
automation of visual data analysis.

Thank you for the consideration of this manuscript. We believe that it
is a good fit for the Journal of Computational and Graphical Statistics.

\closing{Sincerely,}
\vfill
\end{letter}
\end{document}

### 2025-07-28 arxiv_version_2024_11_2/paper.tex

% interactcadsample.tex
% v1.03 - April 2017

\documentclass[]{interact}

\usepackage{epstopdf}% To incorporate .eps illustrations using PDFLaTeX, etc.
\usepackage{subfigure}% Support for small, `sub' figures and tables
%\usepackage[nolists,tablesfirst]{endfloat}% To `separate' figures and tables from text if required

\usepackage{natbib}% Citation support using natbib.sty
\bibpunct[, ]{(}{)}{;}{a}{}{,}% Citation support using natbib.sty
\renewcommand\bibfont{\fontsize{10}{12}\selectfont}% Bibliography support using natbib.sty

\theoremstyle{plain}% Theorem-like structures provided by amsthm.sty
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}

\theoremstyle{remark}
\newtheorem{remark}{Remark}
\newtheorem{notation}{Notation}

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{lscape}
\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\def\tightlist{}
\usepackage{setspace}
\doublespacing

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\begin{document}

\articletype{}

\title{Automated Assessment of Residual Plots with Computer Vision
Models}

\author{\name{Weihao Li$^{a}$, Dianne Cook$^{a}$, Emi Tanaka$^{a, b,
c}$, Susan VanderPlas$^{d}$, Klaus Ackermann$^{a}$}
\affil{$^{a}$Department of Econometrics and Business Statistics, Monash
University, Clayton, VIC, Australia; $^{b}$Biological Data Science
Institute, Australian National University, Acton, ACT,
Australia; $^{c}$Research School of Finance, Actuarial Studies and
Statistics, Australian National University, Acton, ACT,
Australia; $^{d}$Department of Statistics, University of Nebraska,
Lincoln, Nebraska, USA}
}

\thanks{CONTACT Weihao
Li. Email: \href{mailto:weihao.li@monash.edu}{\nolinkurl{weihao.li@monash.edu}}, Dianne
Cook. Email: \href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}, Emi
Tanaka. Email: \href{mailto:emi.tanaka@anu.edu.au}{\nolinkurl{emi.tanaka@anu.edu.au}}, Susan
VanderPlas. Email: \href{mailto:susan.vanderplas@unl.edu}{\nolinkurl{susan.vanderplas@unl.edu}}, Klaus
Ackermann. Email: \href{mailto:klaus.ackermann@monash.edu}{\nolinkurl{klaus.ackermann@monash.edu}}}

\maketitle

\begin{abstract}
Plotting the residuals is a recommended procedure to diagnose deviations
from linear model assumptions, such as non-linearity,
heteroscedasticity, and non-normality. The presence of structure in
residual plots can be tested using the lineup protocol to do visual
inference. There are a variety of conventional residual tests, but the
lineup protocol, used as a statistical test, performs better for
diagnostic purposes because it is less sensitive and applies more
broadly to different types of departures. However, the lineup protocol
relies on human judgment which limits its scalability. This work
presents a solution by providing a computer vision model to automate the
assessment of residual plots. It is trained to predict a distance
measure that quantifies the disparity between the residual distribution
of a fitted classical normal linear regression model and the reference
distribution, based on Kullback-Leibler divergence. From extensive
simulation studies, the computer vision model exhibits lower sensitivity
than conventional tests but higher sensitivity than human visual tests.
It is slightly less effective on non-linearity patterns. Several
examples from classical papers and contemporary data illustrate the new
procedures, highlighting its usefulness in automating the diagnostic
process and supplementing existing methods.
\end{abstract}

\begin{keywords}
statistical graphics; data visualization; visual inference; computer
vision; machine learning; hypothesis testing; reression analysis;
cognitive perception; simulation; practical significance
\end{keywords}

\section{Introduction}\label{sec-model-introduction}

Plotting residuals is commonly regarded as a standard practice in linear
regression diagnostics \citep{belsley1980regression, cook1982residuals}.
This visual assessment plays a crucial role in identifying whether model
assumptions, such as linearity, homoscedasticity, and normality, are
reasonable. It also helps in understanding the goodness of fit and
various unexpected characteristics of the model.

Generating a residual plot in most statistical software is often as
straightforward as executing a line of code or clicking a button.
However, accurately interpreting a residual plot can be challenging. A
residual plot can exhibit various visual features, but it is crucial to
recognize that some may arise from the characteristics of predictors and
the natural stochastic variation of the observational unit, rather than
indicating a violation of model assumptions \citep{li2024plot}. Consider
Figure \ref{fig:false-finding} as an example, the residual plot displays
a triangular left-pointing shape. The distinct difference in the spread
of the residuals across the fitted values may result in the analyst
suggesting that there may be heteroskedasticity, however, it is
important to avoid over-interpreting this visual pattern. In this case,
the fitted regression model is correctly specified, and the triangular
shape is actually a result of the skewed distribution of the predictors,
rather than indicating a flaw in the model.

The concept of visual inference, as proposed by
\citet{buja2009statistical}, provides an inferential framework to assess
whether residual plots indeed contain visual patterns inconsistent with
the model assumptions. The fundamental idea involves testing whether the
true residual plot visually differs significantly from null plots, where
null plots are plotted with residuals generated from the residual
rotation distribution \citep{langsrud2005rotation}, which is a
distribution consistent with the null hypothesis \(H_0\) that the linear
regression model is correctly specified. Typically, the visual test is
accomplished through the lineup protocol, where the true residual plot
is embedded within a lineup alongside several null plots. If the true
residual plot can be distinguished from the lineup, it provides evidence
for rejecting \(H_0\).

The practice of delivering a residual plot as a lineup is generally
regarded as a valuable approach. Beyond its application in residual
diagnostics, the lineup protocol has been integrated into the analysis
of diverse subjects. For instance, Loy and Hofmann
\citetext{\citeyear{loy2013diagnostic}; \citeyear{loy2014hlmdiag}; \citeyear{loy2015you}}
illustrated its applicability in diagnosing hierarchical linear models.
Additionally, \citet{widen2016graphical} and \citet{fieberg2024using}
demonstrated its utility in geographical and ecology research
respectively, while \citet{krishnan2021hierarchical} explored its
effectiveness in forensic examinations.

A practical limitation of the lineup protocol lies in its reliance on
human judgements \citep[see][ about the practical
limitations]{li2024plot}. Unlike conventional statistical tests that can
be performed computationally in statistical software, the lineup
protocol requires human evaluation of images. This characteristic makes
it less suitable for large-scale applications, given the associated high
labour costs and time requirements. There is a substantial need to
develop an approach to substitute these human judgement with an
automated reading of data plots using machines.

The utilization of computers to interpret data plots has a rich history,
with early efforts such as ``Scagnostics'' by \citet{tukey1985computer},
a set of numerical statistics that summarize features of scatter plots.
\citet{wilkinson2005graph} expanded on this work, introducing
scagnostics based on computable measures applied to planar proximity
graphs. These measures, including, but not limited to, ``Outlying'',
``Skinny'', ``Stringy'', ``Straight'', ``Monotonic'', ``Skewed'',
``Clumpy'', and ``Striated'', aimed to characterize outliers, shape,
density, trend, coherence and other characteristics of the data. While
this approach has been inspiring, there is a recognition
\citep{buja2009statistical} that it may not capture all the necessary
visual features that differentiate true residual plots from null plots.
A more promising alternative entails enabling machines to learn the
function for extracting visual features from residual plots.
Essentially, this means empowering computers to discern the crucial
visual features for residual diagnostics and determining the method to
extract them.

Modern computer vision models are well-suited for addressing this
challenge. They rely on deep neural networks with convolutional layers
\citep{fukushima1982neocognitron}. These layers use small, sliding
windows to scan the image, performing a dot product to extract local
features and patterns. Numerous studies have demonstrated the efficacy
of convolutional layers in addressing various vision tasks, including
image recognition \citep{rawat2017deep}. Despite the widespread use of
computer vision models in fields like computer-aided diagnosis
\citep{lee2015image}, pedestrian detection \citep{brunetti2018computer},
and facial recognition \citep{emami2012facial}, their application in
reading data plots remains limited. While some studies have explored the
use of computer vision models for tasks such as reading recurrence plots
for time series regression \citep{ojeda2020multivariate}, time series
classification
\citep{chu2019automatic, hailesilassie2019financial, hatami2018classification, zhang2020encoding},
anomaly detection \citep{chen2020convolutional}, and pairwise causality
analysis \citep{singh2017deep}, the application of reading residual
plots with computer vision models is a new field of study.

In this paper, we develop computer vision models and integrate them into
the residual plots diagnostics workflow, addressing the need for an
automated visual inference. The paper is structured as follows. Section
\ref{sec-model-specifications} discusses various specifications of the
computer vision models. Section
\ref{sec-model-distance-between-residual-plots} defines the distance
measure used to detect model violations, while Section
\ref{sec-model-distance-estimation} explains how the computer vision
models estimate this distance measure. Section
\ref{sec-model-statistical-testing} covers the statistical tests based
on the estimated distance, and Section \ref{sec-model-violations-index}
introduces a Model Violations Index, which offers a quicker and more
convenient assessment. Sections \ref{sec-model-data-generation},
\ref{sec-model-architecture}, and \ref{sec-model-training} detail the
data preparation, model architecture, and training process,
respectively. The results are presented in Section
\ref{sec-model-results}. Example dataset applications are discussed in
Section \ref{sec-examples}. Finally, we conclude with a discussion of
our findings and propose ideas for future research directions.

\begin{figure}[!h]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/false-finding-1}

}

\caption{An example residual vs fitted values plot (red line indicates 0 corresponds to the x-intercept, i.e. $y=0$). The vertical spread of the data points varies with the fitted values. This often indicates the existence of heteroskedasticity, however, here the result is due to skewed distribution of the predictors rather than heteroskedasticity. The Breusch-Pagan test rejects this residual plot at 95\% significance level ($p\text{-value} = 0.046$).}\label{fig:false-finding}
\end{figure}

\section{Model Specifications}\label{sec-model-specifications}

There are various specifications of the computer vision model that can
be used to assess residual plots. We discuss these specifications below
focusing on two key components of the model formula: the input and the
output format.

\subsection{Input Formats}\label{input-formats}

Deep learning models are in general very sensitive to the input data.
The quality and relevance of the input data greatly influence the
model's capacity to generate insightful and meaningful results. There
are several designs of the input format that can be considered.

A straightforward architecture of the input layer involves feeding a
vector of residuals along with a vector of fitted values, essentially
providing all the necessary information for creating a residuals vs
fitted values plot. However, a drawback of this method is the dynamic
input size, which changes based on the number of observations. For
modern computer vision models implemented in mainstream software like
TensorFlow \citep{abadi2016tensorflow}, the input shape is typically
fixed. One solution is to pad the input vectors with leading or trailing
zeros when the input tensor expects longer vectors, but it may fail if
the input vector surpasses the designed length.

Another strategy is to summarize the residuals and fitted values
separately using histograms and utilize the counts as the input. By
controlling the number of bins in the histograms, it becomes possible to
provide fixed-length input vectors. Still, since histograms only capture
the marginal distribution of residuals and fitted values respectively,
they can not be used to differentiate visual patterns with same marginal
distributions but different joint distributions.

Another architecture of the input layer involves using an image as
input. The primary advantage of this design, as opposed to the vector
format, is the availability of the existing and sophisticated image
processing architectures developed over the years, such as the VGG16
architecture proposed in \citet{simonyan2014very}. These architectures
can effectively capture and summarize spatial information from nearby
pixels, which is less straightforward with vector input. While encoding
data points as pixels might lead to some loss of detail, this approach
still provides a more comprehensive representation than the
histogram-based strategy because it captures the joint distribution of
fitted values and residuals. The main considerations are the image
resolution and the aesthetics of the residual plot. In general, a higher
resolution provides more information to the model but comes with a
trade-off of increased complexity and greater difficulty in training. As
for the aesthetics of the residual plot, a practical solution is to
consistently present residual plots in the same style to the model. This
implies that the model can not accept arbitrary images as input but
requires the use of the same pre-processing pipeline to convert
residuals and fitted values into a standardized-style residual plot.

Providing multiple residual plots to the model, such as a pair of plots,
a triplet or a lineup is also a possible option.
\citet{chopra2005learning} have shown that computer vision models
designed for image comparison can assess whether a pair of images are
similar or dissimilar. Applied to our specific problem, we can define
null plots of a fitted regression model to be similar to each other,
while considering true residual plots to be distinct from null plots of
any fitted regression model. A triplet constitutes a set of three
images, denoted as \(image_1\), \(image_2\) and \(image_3\). It is often
used to predict whether \(image_2\) or \(image_3\) is more similar to
\(image_1\), proving particularly useful for establishing rankings
between samples. For this setup, we can apply the same criteria to
define similarity between images. However, it is important to note that
these two approaches usually require additional considerations regarding
the loss function and, at times, non-standard training processes due to
shared weights between different convolutional blocks.

Presenting a lineup to a model aligns closely with the lineup protocol.
However, as the number of residual plots in a lineup increases, the
resolution of the input image grows rapidly, posing challenges in
training the model. We experimented with this approach in a pilot study,
but the performance of the trained model was sub-optimal.

Taking into account the implementation cost and the need for model
interpretability, we used the single residual plot input format in this
paper.

\subsection{Output Formats}\label{output-formats}

Given that the input is a single residual plot represented as a
fixed-resolution image, we can choose the output from the computer
vision model to be either binary (classification) or numeric
(regression).

The binary outcome can represent whether the input image is consistent
with a null plot as determined by either (1) the data generating process
or (2) the result of a visual test based on human judgement. Training a
model following the latter option requires data from prior human subject
experiments, presenting difficulties in controlling the quality of data
due to variations in experimental settings across different studies.
Additionally, some visual inference experiments are unrelated to linear
regression models or residual plot diagnostics, resulting in a limited
amount of available training data.

Alternatively, the output could be a meaningful and interpretable
numerical measure useful for assessing residual plots, such as the
strength of suspicious visual patterns reflecting the extent of model
violations, or the difficulty index for identifying whether a residual
plot has no issues. However, these numeric measures are often informally
used in daily communication but are not typically formalized or
rigorously defined. For the purpose of training a model, this numeric
measure has to be quantifiable.

In this study, we chose to define and use a distance between a true
residual plot and a theoretically ``good'' residual plot. This is
further explained in Section
\ref{sec-model-distance-between-residual-plots}.
\citet{vo2016localizing} have also demonstrated that defining a proper
distance between images can enhance the matching accuracy in image
search compared to a binary outcome model.

\subsection{Auxiliary Information with
Scagnostics}\label{auxiliary-information-with-scagnostics}

In Section \ref{sec-model-introduction}, we mention that scagnostics
consist of a set of manually designed visual feature extraction
functions. While our computer vision model will learn its own feature
extraction function during training, leveraging additional information
from scagnostics can enhance the model's predictive accuracy.

For each residual plot used as an input image, we calculated four
scagnostics --- ``Monotonic'', ``Sparse'', ``Splines'', and ``Striped''
-- using the \texttt{cassowaryr} R package \citep{mason2022cassowaryr}.
These computed measures, along with the number of observations from the
fitted model, were provided as the second input for the computer vision
model. We selected these scagnostics due to their reliability and
efficiency, as other scagnostics occasionally caused R process crashes
(approximately 5\% of the time) during training data preparation, due to
a bug in the \texttt{interp} R package \citep{Albrecht2023interp}.
Although the package maintainer later fixed this bug at our request, the
fix came too late to retrain the model, and additionally, their high
computational costs make them unsuitable for rapid inference, which was
a critical factor in our choice.

\section{Distance from a Theoretically ``Good'' Residual
Plot}\label{sec-model-distance-between-residual-plots}

To develop a computer vision model for assessing residual plots within
the visual inference framework, it is important to precisely define a
numerical measure of ``difference'' or ``distance'' between plots. This
distance can take the form of a basic statistical operation on pixels,
such as the sum of square differences, however, a pixel-to-pixel
comparison makes little sense in comparing residual plots where the main
interest would be structural patterns. Alternatively, it could involve
established image similarity metrics like the Structural Similarity
Index Measure \citep{wang2004image} which compares images by integrating
three perception features of an image: contrast, luminance, and
structure (related to average, standard deviation and correlation of
pixel values over a window, respectively). These image similarity
metrics are tailored for image comparison in vastly different tasks to
evaluating data plots, where only essential plot elements require
assessment \citep{chowdhury2018measuring}. We can alternatively define a
notion of distance by integrating key plot elements (instead of key
perception features like luminance, contrast, and structure), such as
those captured by scagnostics mentioned in Section
\ref{sec-model-introduction}, but the functional form still needs to be
carefully refined to accurately reflect the extent of the violations.

In this section, we introduce a distance measure between a true residual
plot and a theoretically `good' residual plot. This measure quantifies
the divergence between the residual distribution of a given fitted
regression model and that of a correctly specified model. The
computation assumes knowledge of the data generating processes for
predictors and response variables. Since these processes are often
unknown in practice, we will discuss a method to estimate this distance
using a computer vision model in Section
\ref{sec-model-distance-estimation}.

\subsection{Residual Distribution}\label{residual-distribution}

For a classical normal linear regression model,
\(\boldsymbol{y} = \boldsymbol{X}\boldsymbol{\beta} + \boldsymbol{e}\),
the residual \(\hat{\boldsymbol{e}}\) are derived as the difference of
the fitted values and observed values \(\boldsymbol{y}\). Suppose the
data generating process is known and the regression model is correctly
specified, by the Frisch-Waugh-Lowell theorem \citep{frisch1933partial},
residuals \(\hat{\boldsymbol{e}}\) can also be treated as random
variables and written as a linear transformation of the error
\(\boldsymbol{e}\) formulated as
\(\hat{\boldsymbol{e}} = \boldsymbol{R}\boldsymbol{e}\), where
\(\boldsymbol{R}=\boldsymbol{I}_n -\boldsymbol{X}(\boldsymbol{X}^\top\boldsymbol{X})^{-1}\boldsymbol{X}^\top\)
is the residual operator, \(\boldsymbol{I}_n\) is a \(n\) by \(n\)
identity matrix, and \(n\) is the number of observations.

One of the assumptions of the classical normal linear regression model
is that the error \(\boldsymbol{e}\) follows a multivariate normal
distribution with zero mean and constant variance, i.e.,
\(\boldsymbol{e} \sim N(\boldsymbol{0}_n,\sigma^2\boldsymbol{I}_n)\). It
follows that the distribution of residuals \(\hat{\boldsymbol{e}}\) can
be characterized by a certain probability distribution, denoted as
\(Q\), which is transformed from the multivariate normal distribution.
This reference distribution \(Q\) summarizes what ``good'' residuals
should follow given the design matrix \(\boldsymbol{X}\) is known and
fixed.

Suppose the design matrix \(\boldsymbol{X}\) has linearly independent
columns, the trace of the hat matrix
\(\boldsymbol{H} = \boldsymbol{X}(\boldsymbol{X}^\top\boldsymbol{X})^{-1}\boldsymbol{X}^\top\)
will equal to the number of columns in \(\boldsymbol{X}\) denoted as
\(k\). As a result, the rank of \(\boldsymbol{R}\) is \(n - k\), and
\(Q\) is a degenerate multivariate distribution. To capture the
characteristics of \(Q\), such as moments, we can simulate a large
numbers of \(\boldsymbol{\varepsilon}\) and transform it to
\(\boldsymbol{e}\) to get the empirical estimates. For simplicity, in
this study, we replaced the variance-covariance matrix of residuals
\(\text{cov}(\boldsymbol{e}, \boldsymbol{e}) = \boldsymbol{R}\sigma^2\boldsymbol{R}^\top = \boldsymbol{R}\sigma^2\)
with a full-rank diagonal matrix
\(\text{diag}(\boldsymbol{R}\sigma^2)\), where \(\text{diag}(.)\) sets
the non-diagonal entries of a matrix to zeros. The resulting
distribution for \(Q\) is
\(N(\boldsymbol{0}_n, \text{diag}(\boldsymbol{R}\sigma^2))\).

Distribution \(Q\) is derived from the correctly specified model.
However, if the model is misspecified, then the actual distribution of
residuals denoted as \(P\), will be different from \(Q\). For example,
if the data generating process contains variables correlated with any
column of \(\boldsymbol{X}\) but missing from \(\boldsymbol{X}\),
causing an omitted variable problem, \(P\) will be different from \(Q\)
because the residual operator obtained from the fitted regression model
will not be the same as \(\boldsymbol{R}\). Besides, if the
\(\boldsymbol{\varepsilon}\) follows a non-normal distribution such as a
multivariate lognormal distribution, \(P\) will usually be skewed and
has a long tail.

\subsection{\texorpdfstring{Distance of \(P\) from
\(Q\)}{Distance of P from Q}}\label{distance-of-p-from-q}

Defining a proper distance between distributions is usually easier than
defining a proper distance between data plots. Given the true residual
distribution \(Q\) and the reference residual distribution \(P\), we
used a distance measure based on Kullback-Leibler divergence
\citep{kullback1951information} to quantify the difference between two
distributions as

\begin{equation} \label{eq:kl-0}
D = \log\left(1 + D_{KL}\right),
\end{equation}

where \(D_{KL}\) is defined as

\begin{equation} \label{eq:kl-1}
D_{KL} = \int_{\mathbb{R}^{n}}\log\frac{p(\boldsymbol{e})}{q(\boldsymbol{e})}p(\boldsymbol{e})d\boldsymbol{e},
\end{equation}

\noindent and \(p(.)\) and \(q(.)\) are the probability density
functions for distribution \(P\) and distribution \(Q\), respectively.

This distance measure was first proposed in \citet{li2024plot}. It was
mainly designed for measuring the effect size of non-linearity and
heteroskedasticity in a residual plot. \citet{li2024plot} have derived
that, for a classical normal linear regression model that omits
necessary higher-order predictors \(\boldsymbol{Z}\) and the
corresponding parameter \(\boldsymbol{\beta}_z\), and incorrectly
assumes
\(\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n,\sigma^2\boldsymbol{I}_n)\)
while in fact
\(\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n, \boldsymbol{V})\)
where \(\boldsymbol{V}\) is an arbitrary symmetric positive
semi-definite matrix, \(Q\) can be represented as
\(N(\boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z, \text{diag}(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}))\).
Note that the variance-covariance matrix is replaced with the diagonal
matrix to ensure it is a full-rank matrix.

Since both \(P\) and \(Q\) are adjusted to be multivariate normal
distributions, Equation \ref{eq:kl-1} can be further expanded to

\begin{equation} \label{eq:kl-2}
D_{KL} = \frac{1}{2}\left(\log\frac{|\boldsymbol{W}|}{|\text{diag}(\boldsymbol{R}\sigma^2)|} - n + \text{tr}(\boldsymbol{W}^{-1}\text{diag}(\boldsymbol{R}\sigma^2)) + \boldsymbol{\mu}_z^\top\boldsymbol{W}^{-1}\boldsymbol{\mu}_z\right),
\end{equation}

\noindent where
\(\boldsymbol{\mu}_z = \boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z\),
and
\(\boldsymbol{W} = \text{diag}(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R})\).
The assumed error variance \(\sigma^2\) is set to be
\(\text{tr}(\boldsymbol{V})/n\), which is the expectation of the
estimated variance.

\subsection{\texorpdfstring{Non-normal
\(P\)}{Non-normal P}}\label{non-normal-p}

For non-normal error \(\boldsymbol{\varepsilon}\), the true residual
distribution \(P\) is unlikely to be a multivariate normal distribution.
Thus, Equation \ref{eq:kl-2} given in \citet{li2024plot} will not be
applicable to models violating the normality assumption.

To evaluate the Kullback-Leibler divergence of non-normal \(P\) from
\(Q\), the fallback is to solve Equation \ref{eq:kl-1} numerically.
However, since \(\boldsymbol{e}\) is a linear transformation of
non-normal random variables, it is very common that the general form of
\(P\) is unknown, meaning that we can not easily compute
\(p(\boldsymbol{e})\) using a well-known probability density function.
Additionally, even if \(p(\boldsymbol{e})\) can be calculated for any
\(\boldsymbol{e} \in \mathbb{R}^n\), it will be very difficult to do
numerical integration over the \(n\)-dimensional space, because \(n\)
could be potentially very large.

In order to approximate \(D_{KL}\) in a practically computable manner,
the elements of \(\boldsymbol{e}\) are assumed to be independent of each
other. This assumption solves both of the issues mentioned above. First,
we no longer need to integrate over \(n\) random variables. The result
of Equation \ref{eq:kl-1} is now the sum of the Kullback-Leibler
divergence evaluated for each individual residual due to the assumption
of independence between observations. Second, it is not required to know
the joint probability density \(p(\boldsymbol{e})\) any more. Instead,
the evaluation of Kullback-Leibler divergence for an individual residual
relies on the knowledge of the marginal density \(p_i(e_i)\), where
\(e_i\) is the \(i\)-th residual for \(i = 1, ..., n\). This is much
easier to approximate through simulation. It is also worth mentioning
that this independence assumption generally will not hold if
\(\text{cov}(e_i, e_j) \neq 0\) for any \(1 \leq i < j \leq n\), but its
existence is essential for reducing the computational cost.

Given \(\boldsymbol{X}\) and \(\boldsymbol{\beta}\), the algorithm for
approximating Equation \ref{eq:kl-1} starts from simulating \(m\) sets
of observed values \(\boldsymbol{y}\) according to the data generating
process. The observed values are stored in a matrix \(\boldsymbol{A}\)
with \(n\) rows and \(m\) columns, where each column of
\(\boldsymbol{A}\) is a set of observed values. Then, we can get \(m\)
sets of realized values of \(\boldsymbol{e}\) stored in the matrix
\(\boldsymbol{B}\) by applying the residual operator
\(\boldsymbol{B} = \boldsymbol{R}\boldsymbol{A}\). Furthermore, kernel
density estimation (KDE) with Gaussian kernel and optimal bandwidth
selected by the Silverman's rule of thumb \citep{silverman2018density}
is applied on each row of \(\boldsymbol{B}\) to estimate \(p_i(e_i)\)
for \(i = 1, ..., n\). The KDE computation can be done by the
\texttt{density} function in R.

Since the Kullback-Leibler divergence can be viewed as the expectation
of the log-likelihood ratio between distribution \(P\) and distribution
\(Q\) evaluated on distribution \(P\), we can reuse the simulated
residuals in matrix \(\boldsymbol{B}\) to estimate the expectation by
the sample mean. With the independence assumption, for non-normal \(P\),
\(D_{KL}\) can be approximated by

\begin{align*} \label{eq:kl-3}
D_{KL} &\approx \sum_{i = 1}^{n} \hat{D}_{KL}^{(i)}, \\
\hat{D}_{KL}^{(i)} &= \frac{1}{m}\sum_{j = 1}^{m} \log\frac{\hat{p}_i(B_{ij})}{q(B_{ij})},
\end{align*}

\noindent where \(\hat{D}_{KL}^{(i)}\) is the estimator of the
Kullback-Leibler divergence for an individual residual \(e_i\),
\(\boldsymbol{B}_{ij}\) is the \(i\)-th row and \(j\)-th column entry of
the matrix \(\boldsymbol{B}\), \(\hat{p}_i(.)\) is the kernel density
estimator of \(p_i(.)\), \(q(.)\) is the normal density function with
mean zero and an assumed variance estimated as
\(\hat{\sigma}^2 = \sum_{b \in vec(\boldsymbol{B})}(b - \sum_{b \in vec(\boldsymbol{B})} b/nm)^2/(nm - 1)\),
and \(vec(.)\) is the vectorization operator which turns a
\(n \times m\) matrix into a \(nm \times 1\) column vector by stacking
the columns of the matrix on top of each other.

\section{Distance Estimation}\label{sec-model-distance-estimation}

In the previous sections, we have defined a distance measure given in
Equation \ref{eq:kl-0} for quantifying the difference between the true
residual distribution \(P\) and an ideal reference distribution \(Q\).
However, this distance measure can only be computed when the data
generating process is known. In reality, we often have no knowledge
about the data generating process, otherwise we do not need to do a
residual diagnostic in the first place.

We use a computer vision model to estimate this distance measure for a
residual plot. Let \(D\) be the result of Equation \ref{eq:kl-0}, and
our estimator \(\hat{D}\) is formulated as

\begin{equation} \label{eq:d-approx}
\hat{D} = f_{CV}(V_{h \times w}(\boldsymbol{e}, \hat{\boldsymbol{y}})),
\end{equation}

\noindent where \(V_{h \times w}(.)\) is a plotting function that saves
a residuals vs fitted values plot with fixed aesthetic as an image with
\(h \times w\) pixels in three channels (RGB), \(f_{CV}(.)\) is a
computer vision model which takes an \(h \times w\) image as input and
predicts the distance in the domain \([0, +\infty)\).

With the estimated distance \(\hat{D}\), we can compare the underlying
distribution of the residuals to a theoretically ``good'' residual
distribution. \(\hat{D}\) can also be used as an index of the model
violations indicating the strength of the visual signal embedded in the
residual plot.

It is not expected that \(\hat{D}\) will be equal to original distance
\(D\). This is largely because information contained in a single
residual plot is limited and it may not be able to summarize all the
important characteristics of the residual distribution. For a given
residual distribution \(P\), many different residual plots can be
simulated, where many will share similar visual patterns, but some of
them could be visually very different from the rest, especially for
regression models with small \(n\). This suggests the error of the
estimation will vary depends on whether the input residual plot is
representative or not.

\section{Statistical Testing}\label{sec-model-statistical-testing}

\subsection{Lineup Evaluation}\label{sec-model-lineup-evaluation}

Theoretically, the distance \(D\) for a correctly specified model is
\(0\), because \(P\) will be the same as \(Q\). However, the computer
vision model may not necessary predict \(0\) for a null plot. Using
Figure \ref{fig:false-finding} as an example, it contains a visual
pattern which is an indication of heteroskedasticity. We would not
expect the model to be able to magically tell if the suspicious pattern
is caused by the skewed distribution of the fitted values or the
existence of heteroskedasticity. Additionally, some null plots could
have outliers or strong visual patterns due to randomness, and a
reasonable model will try to summarize those information into the
prediction, resulting in \(\hat{D} > 0\).

This property is not an issue if \(\hat{D} \gg 0\) for which the visual
signal of the residual plot is very strong, and we usually do not need
any further examination of the significance of the result. However, if
the visual pattern is weak or moderate, having \(\hat{D}\) will not be
sufficient to determine if \(H_0\) should be rejected.

To address this issue we can adhere to the paradigm of visual inference,
by comparing the estimated distance \(\hat{D}\) to the estimated
distances for the null plots in a lineup. Specifically, if a lineup
comprises 20 plots, the null hypothesis \(H_0\) will be rejected if
\(\hat{D}\) exceeds the maximum estimated distance among the \(m - 1\)
null plots, denoted as
\(\max\limits_{1 \leq i \leq m-1} {\hat{D}_{null}^{(i)}}\), where
\(\hat{D}_{null}^{(i)}\) represents the estimated distance for the
\(i\)-th null plot. This approach is equivalent to the typical lineup
protocol requiring a 95\% significance level, where \(H_0\) is rejected
if the data plot is identified as the most distinct plot by the sole
observer. The estimated distance serves as a metric to quantify the
difference between the data plot and the null plots, as intended.

Moreover, if the number of plots in a lineup, denoted by \(m\), is
sufficiently large, the empirical distribution of
\({\hat{D}_{null}^{(i)}}\) can be viewed as an approximation of the null
distribution of the estimated distance. Consequently, quantiles of the
null distribution can be estimated using the sample quantiles, and these
quantiles can be utilized for decision-making purposes. The details of
the sample quantile computation can be found in
\citet{hyndman1996sample}. For instance, if \(\hat{D}\) is greater than
or equal to the 95\% sample quantile, denoted as \(Q_{null}(0.95)\), we
can conclude that the estimated distance for the true residual plot is
significantly different from the estimated distance for null plots with
a 95\% significance level. Based on our experience, to obtain a stable
estimate of the 95\% quantile, the number of null plots, \(n_{null}\),
typically needs to be at least 100. However, if the null distribution
exhibits a long tail, a larger number of null plots may be required.
Alternatively, a \(p\)-value is the probability of observing a distance
equally or greater than \(\hat{D}\) under the null hypothesis \(H_0\),
and it can be estimated by
\(\frac{1}{m} + \frac{1}{m}\sum_{i=1}^{m-1}I\left(\hat{D}_{null}^{(i)} \geq \hat{D}\right)\).

To alleviate computation burden, a lattice of quantiles for \(\hat{D}\)
under \(H_0\) with specified sample sizes can be precomputed. We can
then map the \(\hat{D}\) and sample size to the closet quantile and
sample size in lattice to calculate the corresponding \(p\)-value. This
approach lose precision in \(p\)-value calculation, however,
significantly improves computational efficiency.

\subsection{Bootstrapping}\label{bootstrapping}

Bootstrap is often employed in linear regression when conducting
inference for estimated parameters \citep[see][ and
\citet{efron1994introduction}]{davison1997bootstrap}. It is typically
done by sampling individual cases with replacement and refitting the
regression model. If the observed data accurately reflects the true
distribution of the population, the bootstrapped estimates can be used
to measure the variability of the parameter estimate without making
strong distributional assumptions about the data generating process.

Similarly, bootstrap can be applied on the estimated distance
\(\hat{D}\). For each refitted model \(M_{boot}^{(i)}\), there will be
an associated residual plot \(V_{boot}^{(i)}\) which can be fed into the
computer vision model to obtain \(\hat{D}_{boot}^{(i)}\), where
\(i = 1,...,n_{boot}\), and \(n_{boot}\) is the number of bootstrapped
samples. If we are interested in the variation of \(\hat{D}\), we can
use \(\hat{D}_{boot}^{(i)}\) to estimate a confidence interval.

Alternatively, since each \(M_{boot}^{(i)}\) has a set of estimated
coefficients \(\hat{\boldsymbol{\beta}}_{boot}^{(i)}\) and an estimated
variance \(\hat{\sigma^2}_{boot}^{(i)}\), a new approximated null
distribution can be construed and the corresponding 95\% sample quantile
\(Q_{boot}^{(i)}(0.95)\) can be computed. Then, if
\(\hat{D}_{boot}^{(i)} \geq Q_{boot}^{(i)}(0.95)\), \(H_0\) will be
rejected for \(M_{boot}^{(i)}\). The ratio of rejected
\(M_{boot}^{(i)}\) among all the refitted models provides an indication
of how often the assumed regression model are considered to be incorrect
if the data can be obtained repetitively from the same data generating
process. But this approach is computationally very expensive because it
requires \(n_{boot} \times n_{null}\) times of residual plot assessment.
In practice, \(Q_{null}(0.95)\) can be used to replace
\(Q_{boot}^{(i)}(0.95)\) in the computation.

\section{Model Violations Index}\label{sec-model-violations-index}

While statistical testing is a powerful tool for detecting model
violations, it can become cumbersome and time-consuming when quick
decisions are needed, particularly due to the need to evaluate numerous
null plots. In practice, a more convenient and immediate method for
assessing model performance is often required. This is where an index,
such as the Model Violations Index (MVI), becomes valuable. It offers a
straightforward way to quantify deviations from model assumptions,
enabling rapid assessment and easier comparison across models.

The estimator \(\hat{D}\) measures the difference between the true
residual distribution and the reference residual distribution, a
difference primarily arises from deviations in model assumptions. The
magnitude of \(D\) directly reflects the degree of these deviations,
thus making \(\hat{D}\) instrumental in forming a model violations index
(MVI).

Note that if more observations are used for estimating the linear
regression, the result of Equation \ref{eq:kl-1} will increase, as the
integration will be performed with larger \(n\). For a given data
generating process, \(D\) typically increases logarithmically with the
number of observations. This behavior comes from the relationship
\(D = \text{log}(1 + D_{KL})\), where
\(D_{KL} = \sum_{i=1}^{n}D_{KL}^{(i)}\) under the assumption of
independence.

Since \(\hat{D}\) is an estimate of \(D\), it is expected that a larger
number of observations will also lead to a higher \(\hat{D}\). However,
this does not imply that \(\hat{D}\) fails to accurately represent the
extent of model violations. In fact, when examining residual plots with
more observations, we often observe a stronger visual signal strength,
as the underlying patterns are more likely to be revealed, except in
cases of significant overlapping.

Therefore, the Model Violations Index (MVI) can be proposed as

\begin{equation} \label{eq:mvi}
\text{MVI} = C + \hat{D} - \log(n),
\end{equation}

\noindent where \(C\) is a large enough constant keeping the result
positive and the term \(-\log(n)\) is used to offset the increase in
\(D\) due to sample size.

Figure \ref{fig:poly-heter-index} displays the residual plots for fitted
models exhibiting varying degrees of non-linearity and
heteroskedasticity. Each residual plot's MVI is computed using Equation
\ref{eq:mvi} with \(C = 10\). When \(\text{MVI} > 8\), the visual
patterns are notably strong and easily discernible by humans. In the
range \(6 < \text{MVI} < 8\), the visibility of the visual pattern
diminishes as MVI decreases. Conversely, when \(\text{MVI} < 6\), the
visual pattern tends to become relatively faint and challenging to
observe. Table \ref{tab:mvi} provides a summary of the MVI usage and it
is applicable to other linear regression models.

\begin{table}

\caption{\label{tab:mvi}Degree of model violations or the strength of the visual signals according to the Model Violations Index (MVI). The constant $C$ is set to be 10.}
\centering
\begin{tabular}[t]{lc}
\toprule
Degree of model violations & Range ($C$ = 10)\\
\midrule
Strong & $\text{MVI} > 8$\\
Moderate & $6 < \text{MVI} < 8$\\
Weak & $\text{MVI} < 6$\\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}[!h]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/poly-heter-index-1}

}

\caption{Residual plots generated from fitted models exhibiting varying degrees of (A) non-linearity and (B) heteroskedasticity violations. The model violations index (MVI) is displayed atop each residual plot. The non-linearity patterns are relatively strong for $MVI > 8$, and relatively weak for $MVI < 6$, while the heteroskedasticity patterns are relatively strong for $MVI > 8$, and relatively weak for $MVI < 6$.}\label{fig:poly-heter-index}
\end{figure}

\section{Data Generation}\label{sec-model-data-generation}

\subsection{Simulation Scheme}\label{simulation-scheme}

While observational data is frequently employed in training models for
real-world applications, the data generating process of observational
data often remains unknown, making computation for our target variable
\(D\) unattainable. Consequently, the computer vision models developed
in this study were trained using synthetic data, including 80,000
training images and 8,000 test images. This approach provided us with
precise label annotations. Additionally, it ensured a large and diverse
training dataset, as we had control over the data generating process,
and the simulation of the training data was relatively cost-effective.

We have incorporated three types of residual departures of linear
regression model in the training data, including non-linearity,
heteroskedasticity and non-normality. All three departures can be
summarized by the data generating process formulated as

\begin{align} \label{eq:data-sim}
\boldsymbol{y} &= \boldsymbol{1}_n + \boldsymbol{x}_1 + \beta_1\boldsymbol{x}_2 + \beta_2(\boldsymbol{z} + \beta_1\boldsymbol{w}) + \boldsymbol{k} \odot \boldsymbol{\varepsilon}, \\
\boldsymbol{z} &= \text{He}_j(g(\boldsymbol{x}_1, 2)), \\
\boldsymbol{w} &= \text{He}_j(g(\boldsymbol{x}_2, 2)), \\
\boldsymbol{k} &= \left[\boldsymbol{1}_n + b(2 - |a|)(\boldsymbol{x}_1 + \beta_1\boldsymbol{x}_2 - a\boldsymbol{1}_n)^{\circ2}\right]^{\circ1/2},
\end{align}

\noindent where \(\boldsymbol{y}\), \(\boldsymbol{x}_1\),
\(\boldsymbol{x}_2\), \(\boldsymbol{z}\), \(\boldsymbol{w}\),
\(\boldsymbol{k}\) and \(\boldsymbol{\varepsilon}\) are vectors of size
\(n\), \(\boldsymbol{1}_n\) is a vector of ones of size \(n\),
\(\boldsymbol{x}_1\) and \(\boldsymbol{x}_2\) are two independent
predictors, \(\text{He}_j(.)\) is the \(j\)th-order probabilist's
Hermite polynomials \citep{hermite1864nouveau}, \((.)^{\circ2}\) and
\((.)^{\circ1/2}\) are Hadamard square and square root, \(\odot\) is the
Hadamard product, and \(g(\boldsymbol{x}, k)\) is a scaling function to
enforce the support of the random vector to be \([-k, k]^n\) defined as

\[g(\boldsymbol{x}, k) = 2k \cdot \frac{\boldsymbol{x} - x_{\min}\boldsymbol{1}_n}{x_{\max} - x_{\min}} - k\boldsymbol{1}_n,~for~k > 0,\]
\noindent where \(x_{\min} = \underset{i \in \{ 1,...,n\}}{\min} x_i\),
\(x_{\max} = \underset{i \in \{ 1,...,n\}}{\max} x_i\) and \(x_i\) is
the \(i\)-th entry of \(\boldsymbol{x}\).

\begin{table}

\caption{\label{tab:factor}Factors used in the data generating process for synthetic data simulation. Factor $j$ and $a$ controls the non-linearity shape and the heteroskedasticity shape respectively. Factor $b$, $\sigma_\varepsilon$ and $n$ control the signal strength. Factor $\text{dist}_\varepsilon$, $\text{dist}_{x1}$ and $\text{dist}_{x2}$ specifies the distribution of $\varepsilon$, $X_1$ and $X_2$ respectively.}
\centering
\begin{tabular}[t]{ll}
\toprule
Factor & Domain\\
\midrule
j & \{2, 3, ..., 18\}\\
a & {}[-1, 1]\\
b & {}[0, 100]\\
$\beta_1$ & \{0, 1\}\\
$\beta_2$ & \{0, 1\}\\
\addlinespace
$\text{dist}_\varepsilon$ & \{discrete, uniform, normal, lognormal\}\\
$\text{dist}_{x1}$ & \{discrete, uniform, normal, lognormal\}\\
$\text{dist}_{x2}$ & \{discrete, uniform, normal, lognormal\}\\
$\sigma_{\varepsilon}$ & {}[0.0625, 9]\\
$\sigma_{X1}$ & {}[0.3, 0.6]\\
\addlinespace
$\sigma_{X2}$ & {}[0.3, 0.6]\\
n & {}[50, 500]\\
\bottomrule
\end{tabular}
\end{table}

The residuals and fitted values of the fitted model were obtained by
regressing \(\boldsymbol{y}\) on \(\boldsymbol{x}_1\). If
\(\beta_1 \neq 0\), \(\boldsymbol{x}_2\) was also included in the design
matrix. This data generation process was adapted from
\citet{li2024plot}, where it was utilized to simulate residual plots
exhibiting non-linearity and heteroskedasticity visual patterns for
human subject experiments. A summary of the factors utilized in Equation
\ref{eq:data-sim} is provided in Table \ref{tab:factor}.

In Equation \ref{eq:data-sim}, \(\boldsymbol{z}\) and \(\boldsymbol{w}\)
represent higher-order terms of \(\boldsymbol{x}_1\) and
\(\boldsymbol{x}_2\), respectively. If \(\beta_2 \neq 0\), the
regression model will encounter non-linearity issues. Parameter \(j\)
serves as a shape parameter that controls the number of tuning points in
the non-linear pattern. Typically, higher values of \(j\) lead to an
increase in the number of tuning points, as illustrated in Figure
\ref{fig:different-j}.

\begin{figure}[!h]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/different-j-1}

}

\caption{Non-linearity forms generated for the synthetic data simulation. The 17 shapes are generated by varying the order of polynomial given by $j$ in $He_j(.)$.}\label{fig:different-j}
\end{figure}

Additionally, scaling factor \(\boldsymbol{k}\) directly affects the
error distribution and it is correlated with \(\boldsymbol{x}_1\) and
\(\boldsymbol{x}_2\). If \(b \neq 0\) and
\(\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n, \sigma^2\boldsymbol{I}_n)\),
the constant variance assumption will be violated. Parameter \(a\) is a
shape parameter controlling the location of the smallest variance in a
residual plot as shown in Figure \ref{fig:different-a}.

\begin{figure}[!h]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/different-a-1}

}

\caption{Heteroskedasticity forms generated for the synthetic data simulation. Different shapes are controlled by the continuous factor $a$ between -1 and 1. For $a = -1$, the residual plot exhibits a "left-triangle" shape. And for $a = 1$, the residual plot exhibits a "right-triangle" shape. }\label{fig:different-a}
\end{figure}

\begin{figure}[!h]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/different-e-1}

}

\caption{Non-normality forms generated for the synthetic data simulation. Four different error distributions including discrete, lognormal, normal and uniform are considered.}\label{fig:different-e}
\end{figure}

Non-normality violations arise from specifying a non-normal distribution
for \(\boldsymbol{\varepsilon}\). In the synthetic data simulation, four
distinct error distributions are considered, including discrete,
uniform, normal, and lognormal distributions, as presented in Figure
\ref{fig:different-e}. Each distribution imparts unique characteristics
in the residual plot. The discrete error distribution introduces
discrete clusters in residuals, while the lognormal distribution
typically yields outliers. Uniform error distribution may result in
residuals filling the entire space of the residual plot. All of these
distributions exhibit visual distinctions from the normal error
distribution.

\begin{figure}[!h]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/different-j-x2-1}

}

\caption{Residual plots of multiple linear regression models with non-linearity issues. The 17 shapes are generated by varying the order of polynomial given by $j$ in $He_j(.)$. A second predictor $\boldsymbol{x}_2$ is introduced to the regression model to create complex shapes.}\label{fig:different-j-x2}
\end{figure}

\begin{figure}[!h]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/different-j-heter-1}

}

\caption{Residual plots of models violating both the non-linearity and the heteroskedasticity assumptions. The 17 shapes are generated by varying the order of polynomial given by $j$ in $He_j(.)$, and the "left-triangle" shape is introduced by setting $a = -1$.}\label{fig:different-j-heter}
\end{figure}

\begin{figure}[!h]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/different-e-heter-1}

}

\caption{Residual plots of models violating both the non-normality and the heteroskedasticity assumptions. The four shapes are generated by using four different error distributions including discrete, lognormal, normal and uniform, and the "left-triangle" shape is introduced by setting $a = -1$. }\label{fig:different-e-heter}
\end{figure}

Equation \ref{eq:data-sim} accommodates the incorporation of the second
predictor \(\boldsymbol{x}_2\). Introducing it into the data generation
process by setting \(\beta_1 = 1\) significantly enhances the complexity
of the shapes, as illustrated in Figure \ref{fig:different-j-x2}. In
comparison to Figure \ref{fig:different-j}, Figure
\ref{fig:different-j-x2} demonstrates that the non-linear shape
resembles a surface rather than a single curve. This augmentation can
facilitate the computer vision model in learning visual patterns from
residual plots of the multiple linear regression model.

In real-world analysis, it is not uncommon to encounter instances where
multiple model violations coexist. In such cases, the residual plots
often exhibit a mixed pattern of visual anomalies corresponding to
different types of model violations. Figure \ref{fig:different-j-heter}
and Figure \ref{fig:different-e-heter} show the visual patterns of
models with multiple model violations.

\subsection{Balanced Dataset}\label{balanced-dataset}

To train a robust computer vision model, we deliberately controlled the
distribution of the target variable \(D\) in the training data. We
ensured that it followed a uniform distribution between \(0\) and \(7\).
This was achieved by organizing \(50\) buckets, each exclusively
accepting training samples with \(D\) falling within the range
\([7(i - 1)/49, 7i/49)\) for \(i < 50\), where \(i\) represents the
index of the \(i\)-th bucket. For the \(50\)-th bucket, any training
samples with \(D \geq 7\) were accepted.

With 80,000 training images prepared, each bucket accommodated a maximum
of \(80000/ 50 = 1600\) training samples. The simulator iteratively
sampled parameter values from the parameter space, generated residuals
and fitted values using the data generation process, computed the
distance, and checked if the sample fitted within the corresponding
bucket. This process continued until all buckets were filled.

Similarly, we adopted the same methodology to prepare 8,000 test images
for performance evaluation and model diagnostics.

\section{Model Architecture}\label{sec-model-architecture}

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/cnn-diag-1}

}

\caption{Diagram of the architecture of the optimized computer vision model. Numbers at the bottom of each box show the shape of the output of each layer. The band of each box drawn in a darker color indicates the use of the rectified linear unit activation function.  Yellow boxes are 2D convolutional layers, orange boxes are pooling layers, the grey box is the concatenation layer, and the purple boxes are dense layers.}\label{fig:cnn-diag}
\end{figure}

The architecture of the computer vision model is adapted from the
well-established VGG16 architecture \citep{simonyan2014very}. While more
recent architectures like ResNet \citep{he2016deep} and
DenseNet\citep{huang2017densely}, have achieved even greater
performance, VGG16 remains a solid choice for many applications due to
its simplicity and effectiveness. Our decision to use VGG16 aligns with
our goal of starting with a proven and straightforward model. Figure
\ref{fig:cnn-diag} provides a diagram of the architecture. More details
about the neural network layers used in this study are provided in
Appendix A.

The model begins with an input layer of shape
\(n \times h \times w \times 3\), capable of handling \(n\) RGB images.
This is followed by a grayscale conversion layer utilizing the luma
formula under the Rec. 601 standard \citep{series2011studio}, which
converts the color image to grayscale. Grayscale suffices for our task
since data points are plotted in black. We experiment with three
combinations of \(h\) and \(w\): \(32 \times 32\), \(64 \times 64\), and
\(128 \times 128\), aiming to achieve sufficiently high image resolution
for the problem at hand.

The processed image is used as the input for the first convolutional
block. The model comprises at most five consecutive convolutional
blocks, mirroring the original VGG16 architecture. Within each block,
there are two 2D convolutional layers followed by two activation layers,
respectively. Subsequently, a 2D max-pooling layer follows the second
activation layer. The 2D convolutional layer convolves the input with a
fixed number of \(3 \times 3\) convolution filters, while the 2D
max-pooling layer downsamples the input along its spatial dimensions by
taking the maximum value over a \(2 \times 2\) window for each channel
of the input. The activation layer employs the rectified linear unit
(ReLU) activation function, a standard practice in deep learning, which
introduces a non-linear transformation of the output of the 2D
convolutional layer. Additionally, to regularize training, a batch
normalization layer is added after each 2D convolutional layer and
before the activation layer. Finally, a dropout layer is appended at the
end of each convolutional block to randomly set some inputs to zero
during training, further aiding in regularization.

The output of the last convolutional block is summarized by either a
global max pooling layer or a global average pooling layer, resulting in
a two-dimensional tensor. To leverage the information contained in
scagnostics, this tensor is concatenated with an additional
\(n \times 5\) tensor, which contains the ``Monotonic'', ``Sparse'',
``Splines'', and ``Striped'' measures, along with the number of
observations for \(n\) residual plots.

The concatenated tensor is then fed into the final prediction block.
This block consists of two fully-connected layers. The first layer
contains at least \(128\) units, followed by a dropout layer.
Occasionally, a batch normalization layer is inserted between the
fully-connected layer and the dropout layer for regularization purposes.
The second fully-connected layer consists of only one unit, serving as
the output of the model.

The model weights \(\boldsymbol{\theta}\) were randomly initialized and
they were optimized by the Adam optimizer \citep{kingma2014adam} with
the mean square error loss function

\[\hat{\boldsymbol{\theta}} = \underset{\boldsymbol{\theta}}{\text{arg min}}\frac{1}{n_{\text{train}}}\sum_{i=1}^{n_{\text{train}}}(D_i - f_{\boldsymbol{\theta}}(V_i, S_i))^2,\]

\noindent where \(n_{\text{train}}\) is the number of training samples,
\(V_i\) is the \(i\)-th residual plot and \(S_i\) is the additional
information about the \(i\)-th residual plot including four scagnostics
and the number of observations.

\section{Model Training}\label{sec-model-training}

To achieve a near-optimal deep learning model, hyperparameters like the
learning rate often need to be fine-tuned using a tuner. In our study,
we utilized the Bayesian optimization tuner from the \texttt{KerasTuner}
Python library \citep{omalley2019kerastuner} for this purpose. A
comprehensive list of hyperparameters is provided in Table
\ref{tab:hyperparameter}.

The number of base filters determines the number of filters for the
first 2D convolutional layer. In the VGG16 architecture, the number of
filters for the 2D convolutional layer in a block is typically twice the
number in the previous block, except for the last block, which maintains
the same number of convolution filters as the previous one. This
hyperparameter aids in controlling the complexity of the computer vision
model. A higher number of base filters results in more trainable
parameters. Likewise, the number of units for the fully-connected layer
determines the complexity of the final prediction block. Increasing the
number of units enhances model complexity, resulting in more trainable
parameters.

The dropout rate and batch normalization are flexible hyperparameters
that work in conjunction with other parameters to facilitate smooth
training. A higher dropout rate is necessary when the model tends to
overfit the training dataset by learning too much noise
\citep{srivastava2014dropout}. Conversely, a lower dropout rate is
preferred when the model is complex and challenging to converge. Batch
normalization, on the other hand, addresses the internal covariate shift
problem arising from the randomness in weight initialization and input
data \citep{goodfellow2016deep}. It helps stabilize and accelerate the
training process by normalizing the activations of each layer.

Additionally, incorporating additional inputs such as scagnostics and
the number of observations can potentially enhance prediction accuracy.
Therefore, we allow the tuner to determine whether these inputs were
necessary for optimal model performance.

The learning rate is a crucial hyperparameter, as it dictates the step
size of the optimization algorithm. A high learning rate can help the
model avoid local minima but risks overshooting and missing the global
minimum. Conversely, a low learning rate smoothens the training process
but makes the convergence time longer and increases the likelihood of
getting trapped in local minima.

Our model was trained on the MASSIVE M3 high-performance computing
platform \citep{goscinski2014multi}, using TensorFlow
\citep{abadi2016tensorflow} and Keras \citep{chollet2015keras}. During
training, 80\% of the training data was utilized for actual training,
while the remaining 20\% was used as validation data. The Bayesian
optimization tuner conducted 100 trials to identify the best
hyperparameter values based on validation root mean square error. The
tuner then restored the best epoch of the best model from the trials.
Additionally, we applied early stopping, terminating the training
process if the validation root mean square error fails to improve for 50
epochs. The maximum allowed epochs was set at 2,000, although no models
reached this threshold.

\begin{table}

\caption{\label{tab:hyperparameter}Name of hyperparameters and their correspoding domain for the computer vision model.}
\centering
\begin{tabular}[t]{ll}
\toprule
Hyperparameter & Domain\\
\midrule
Number of base filters & \{4, 8, 16, 32, 64\}\\
Dropout rate for convolutional blocks & {}[0.1, 0.6]\\
Batch normalization for convolutional blocks & \{false, true\}\\
Type of global pooling & \{max, average\}\\
Ignore additional inputs & \{false, true\}\\
\addlinespace
Number of units for the fully-connected layer & \{128, 256, 512, 1024, 2048\}\\
Batch normalization for the fully-connected layer & \{false, true\}\\
Dropout rate for the fully-connected layer & {}[0.1, 0.6]\\
Learning rate & {}[$10^{-8}$, $10^{-1}$]\\
\bottomrule
\end{tabular}
\end{table}

Based on the tuning process described above, the optimized
hyperparameter values are presented in Table
\ref{tab:best-hyperparameter}. It was observable that a minimum of
\(32\) base filters was necessary, with the preferable choice being
\(64\) base filters for both the \(64 \times 64\) and \(128 \times 128\)
models, mirroring the original VGG16 architecture. The optimized dropout
rate for convolutional blocks hovered around \(0.4\), and incorporating
batch normalization for convolutional blocks proved beneficial for
performance.

All optimized models chose to retain the additional inputs, contributing
to the reduction of validation error. The number of units required for
the fully-connected layer was \(256\), a relatively modest number
compared to the VGG16 classifier, suggesting that the problem at hand
was less complex. The optimized learning rates were higher for models
with higher resolution input, likely because models with more parameters
are more prone to getting stuck in local minima, requiring a higher
learning rate.

\begin{table}

\caption{\label{tab:best-hyperparameter}Hyperparameters values for the optimized computer vision models with different input sizes.}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{llll}
\toprule
Hyperparameter & $32 \times 32$ & $64 \times 64$ & $128 \times 128$\\
\midrule
Number of base filters & 32 & 64 & 64\\
Dropout rate for convolutional blocks & 0.4 & 0.3 & 0.4\\
Batch normalization for convolutional blocks & true & true & true\\
Type of global pooling & max & average & average\\
Ignore additional inputs & false & false & false\\
\addlinespace
Number of units for the fully-connected layer & 256 & 256 & 256\\
Batch normalization for the fully-connected layer & false & true & true\\
Dropout rate for the fully-connected layer & 0.2 & 0.4 & 0.1\\
Learning rate & 0.0003 & 0.0006 & 0.0052\\
\bottomrule
\end{tabular}}
\end{table}

\section{Results}\label{sec-model-results}

\subsection{Model Performance}\label{model-performance}

The test performance for the optimized models with three different input
sizes are summarized in Table \ref{tab:performance}. Among these models,
the \(32 \times 32\) model consistently exhibited the best test
performance. The mean absolute error of the \(32 \times 32\) model
indicated that the difference between \(\hat{D}\) and \(D\) was
approximately \(0.43\) on the test set, a negligible deviation
considering the normal range of \(D\) typically falls between \(0\) and
\(7\). The high \(R^2\) values also suggested that the predictions were
largely linearly correlated with the target.

Figure \ref{fig:model-performance} presents a hexagonal heatmap for
\(D - \hat{D}\) versus \(D\). The brown smoothing curves, fitted by
generalized additive models \citep{hastie2017generalized}, demonstrate
that all the optimized models perform admirably on the test sets when
\(1.5 < D < 6\), where no structural issues are noticeable. However,
over-predictions occurred when \(D < 1.5\), while under-predictions
occurred predominantly when \(\hat{D} > 6\).

For input images representing null plots where \(D = 0\), it was
expected that the models will over-predict the distance, as explained in
Section \ref{sec-model-lineup-evaluation}. However, it can not explain
the under-prediction issue. Therefore, we analysed the relationship
between residuals and all the factors involved in the data generating
process. We found that most issues actually arose from non-linearity
problems and the presence of a second predictor in the regression model
as illustrated in Figure \ref{fig:over-under}. When the variance for the
error distribution was small, the optimized model tended to
under-predict the distance. Conversely, when the error distribution had
a large variance, the model tended to over-predict the distance.

Since most of the deviation stemmed from the presence of non-linearity
violations, to further investigate this, we split the test set based on
violation types and re-evaluated the performance, as detailed in Table
\ref{tab:performance-sub}. It was evident that metrics for null plots
were notably worse compared to other categories. Furthermore, residual
plots solely exhibiting non-normality issues were the easiest to
predict, with very low test root mean square error (RMSE) at around
\(0.3\). Residual plots with non-linearity issues were more challenging
to assess than those with heteroskedasticity or non-normality issues.
When multiple violations were introduced to a residual plot, the
performance metrics typically lay between the metrics for each
individual violation.

Based on the model performance metrics, we chose to use the
best-performing model evaluated on the test set, namely the
\(32 \times 32\) model, for the subsequent analysis.

\begin{table}

\caption{\label{tab:performance}The test performance of three optimized models with different input sizes.}
\centering
\begin{tabular}[t]{lrrrr}
\toprule
 & RMSE & $R^2$ & MAE & Huber loss\\
\midrule
$32 \times 32$ & 0.660 & 0.901 & 0.434 & 0.18\\
$64 \times 64$ & 0.674 & 0.897 & 0.438 & 0.19\\
$128 \times 128$ & 0.692 & 0.892 & 0.460 & 0.20\\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}[!h]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/model-performance-1}

}

\caption{Hexagonal heatmap for difference in $D$ and $\hat{D}$ vs $D$ on test data for three optimized models with different input sizes. The brown lines are smoothing curves produced by fitting generalized additive models. The area over the zero line in light yellow indicates under-prediction, and the area under the zero line in light green indicates over-prediction.}\label{fig:model-performance}
\end{figure}

\begin{figure}[!h]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/over-under-1}

}

\caption{Scatter plots for difference in $D$ and $\hat{D}$ vs $\sigma$ on test data for the $32 \times 32$ optimized model. The data is grouped by whether the regression has only non-linearity violation, and whether it includes a second predictor in the regression formula. The brown lines are smoothing curves produced by fitting generalized additive models. The area over the zero line in light yellow indicates under-prediction, and the area under the zero line in light green indicates over-prediction.}\label{fig:over-under}
\end{figure}

\begin{table}

\caption{\label{tab:performance-sub}The training and test performance of the $32 \times 32$ model presented with different model violations.}
\centering
\begin{tabular}[t]{lrr}
\toprule
Violations & \#samples & RMSE\\
\midrule
no violations & 155 & 1.267\\
non-linearity & 2218 & 0.787\\
heteroskedasticity & 1067 & 0.602\\
non-linearity + heteroskedasticity & 985 & 0.751\\
non-normality & 1111 & 0.320\\
non-linearity + non-normality & 928 & 0.600\\
heteroskedasticity + non-normality & 819 & 0.489\\
non-linearity + heteroskedasticity + non-normality & 717 & 0.620\\
\bottomrule
\end{tabular}
\end{table}

\subsection{Comparison with Human Visual Inference and Conventional
Tests}\label{comparison-with-human-visual-inference-and-conventional-tests}

\subsubsection{Overview of the Human Subject
Experiment}\label{overview-of-the-human-subject-experiment}

In order to check the validity of the proposed computer vision model,
residual plots presented in the human subject experiment conducted by
\citet{li2024plot} will be assessed.

This study has collected 7,974 human responses to 1,152 lineups. Each
lineup contains one randomly placed true residual plot and 19 null
plots. Among the 1,152 lineups, 24 are attention check lineups in which
the visual patterns are designed to be extremely obvious and very
different from the corresponding to null plots, 36 are null lineups
where all the lineups consist of only null plots, 279 are lineups with
uniform predictor distribution evaluated by 11 participants, and the
remaining 813 are lineups with discrete, skewed or normal predictor
distribution evaluated by 5 participants. Attention check lineups and
null lineups will not be assessed in the following analysis.

In \citet{li2024plot}, the residual plots are simulated from a data
generating process which is a special case of Equation
\ref{eq:data-sim}. The main characteristic is the model violations are
introduced separately, meaning non-linearity and heteroskedasticity will
not co-exist in one lineup but assigned uniformly to all lineups.
Additionally, non-normality and multiple predictors are not considered
in the experimental design.

\subsubsection{Model Performance on the Human-evaluated
Data}\label{model-performance-on-the-human-evaluated-data}

\begin{table}

\caption{\label{tab:experiment-performance}The performance of the $32 \times 32$ model on the data used in the human subject experiment.}
\centering
\begin{tabular}[t]{lrrrr}
\toprule
Violation & RMSE & $R^2$ & MAE & Huber loss\\
\midrule
heteroskedasticity & 0.721 & 0.852 & 0.553 & 0.235\\
non-linearity & 0.738 & 0.770 & 0.566 & 0.246\\
\bottomrule
\end{tabular}
\end{table}

For each lineup used in \citet{li2024plot}, there is one true residual
plot and 19 null plots. While the distance \(D\) for the true residual
plot depends on the underlying data generating process, the distance
\(D\) for the null plots is zero. We have used our optimized computer
vision model to estimate distance for both the true residual plots and
the null plots. To have a fair comparison, \(H_0\) will be rejected if
the true residual plot has the greatest estimated distance among all
plots in a lineup. Additionally, the appropriate conventional tests
including the Ramsey Regression Equation Specification Error Test
(RESET) \citep{ramsey1969tests} for non-linearity and the Breusch-Pagan
test \citep{breusch1979simple} for heteroskedasticity were applied on
the same data for comparison.

The performance metrics of \(\hat{D}\) for true residual plots are
outlined in Table \ref{tab:experiment-performance}. It is notable that
all performance metrics are slightly worse than those evaluated on the
test data. Nevertheless, the mean absolute error remains at a low level,
and the linear correlation between the prediction and the true value
remains very high. Consistent with results in Table
\ref{tab:performance-sub}, lineups with non-linearity issues are more
challenging to predict than those with heteroskedasticity issues.

Table \ref{tab:human-conv-table} provides a summary of the agreement
between decisions made by the computer vision model and conventional
tests. The agreement rates between conventional tests and the computer
vision model are 85.95\% and 79.69\% for residual plots containing
heteroskedasticity and non-linearity patterns, respectively. These
figures are higher than those calculated for visual tests conducted by
human, indicating that the computer vision model exhibits behavior more
akin to the best available conventional tests. However, Figure
\ref{fig:conv-mosaic} shows that the computer vision model does not
always reject when the conventional tests reject. And a small number of
plots will be rejected by computer vision model but not by conventional
tests. This suggests that conventional tests are more sensitive than the
computer vision model.

Figure \ref{fig:pcp} further illustrates the decisions made by visual
tests conducted by human, computer vision models, and conventional
tests, using a parallel coordinate plots. It can be observed that all
three tests will agree with each other for around 50\% of the cases.
When visual tests conducted by human do not reject, there are
substantial amount of cases where computer vision model also do not
reject but conventional tests reject. There are much fewer cases that do
not reject by visual tests and conventional tests, but is rejected by
computer vision models. This indicates computer vision model can behave
like visual tests conducted by human better than conventional tests.
Moreover, there are great proportion of cases where visual tests
conducted by human is the only test who does not reject.

When plotting the decision against the distance, as illustrated in
Figure \ref{fig:power}, several notable observations emerge. Firstly,
compared to conventional tests, the computer vision model tends to have
fewer rejected cases when \(D < 2\) and fewer non-rejected cases when
\(2< D < 4\). This suggests tests based on the computer vision model are
less sensitive to small deviations from model assumptions than
conventional tests but more sensitive to moderate deviations.
Additionally, visual tests demonstrate the lowest sensitivity to
residual plots with small distances where not many residual plots are
rejected when \(D < 2\). Similarly, for large distances where \(D > 4\),
almost all residual plots are rejected by the computer vision model and
conventional tests, but for visual tests conducted by humans, the
threshold is higher with \(D > 5\).

In Figure \ref{fig:power}, rejection decisions are fitted by logistic
regression models with no intercept terms and an offset equals to
\(\text{log}(0.05/0.95)\). The fitted curves for the computer vision
model fall between those of conventional tests and visual tests for both
non-linearity and heteroskedasticity, which means there is still
potential to refine the computer vision model to better align its
behavior with visual tests conducted by humans.

In the experiment conducted in \citet{li2024plot}, participants were
allowed to make multiple selections for a lineup. The weighted detection
rate was computed by assigning weights to each detection. If the
participant selected zero plots, a weight of 0.05 was assigned;
otherwise, if the true residual plot was detected, the weight was 1
divided by the number of selections. This weighted detection rate allow
us to assess the quality of the distance measure purposed in this paper,
by using the \(\delta\)-difference statistic. The \(\delta\)-difference
is originally defined by \citet{chowdhury2018measuring}, is given by

\[
\delta = \bar{d}_{\text{true}} - \underset{j}{\text{max}}\left(\bar{d}_{\text{null}}^{(j)}\right) \quad \text{for}~j = 1,...,m-1,
\]

where \(\bar{d}_{\text{null}}^{(j)}\) is the mean distance between the
\(j\)-th null plot and the other null plots, \(\bar{d}_{\text{true}}\)
is the mean distance between the true residual plot and null plots, and
\(m\) is the number of plots in a lineup. These mean distances are used
because, as noted by \citet{chowdhury2018measuring}, the distances can
vary depending on which data plot is used for comparison. For instance,
with three null plots, A, B and C, the distance between A and B may
differ from the distance between A and C. To obtain a consistent
distance for null plot A, averaging is necessary. However, this approach
is not applicable to the distance proposed in this paper, as we only
compare the residual plot against a theoretically good residual plot.
Consequently, the statistic must be adjusted to evaluate our distance
measure effectively.

One important aspect that the \(\delta\)-difference was designed to
capture is the empirical distribution of distances for null plot. If we
were to replace the mean distances \(\bar{d}_{\text{null}}^{(j)}\)
directly with \(D_{\text{null}}^{(j)}\), the distance of the \(j\)-th
null plot, the resulting distribution would be degenerate, since
\(D_{null}\) equals zero by definition. Additionally, \(D\) can not be
derived from an image, meaning it falls outside the scope of the
distances considered by \citet{chowdhury2018measuring}. Instead, the
focus should be on the empirical distribution of \(\hat{D}\), as it
influences decision-making. Therefore, the adjusted \(\delta\)-different
is defined as

\[
\delta_{\text{adj}} = \hat{D} - \underset{j}{\text{max}}\left(\hat{D}_{\text{null}}^{(j)}\right) \quad \text{for}~j = 1,...,m-1,
\]

\noindent where \(\hat{D}_{\text{null}}^{(j)}\) is the estimated
distance for the \(j\)-th null plot, and \(m\) is the number of plots in
a lineup.

Figure \ref{fig:delta} displays the scatter plot of the weighted
detection rate vs the adjusted \(\delta\)-difference. It indicates that
the weighted detection rate increases as the adjusted
\(\delta\)-difference increases, particularly when the adjusted
\(\delta\)-difference is greater than zero. A negative adjusted
\(\delta\)-difference suggests that there is at least one null plot in
the lineup with a stronger visual signal than the true residual plot. In
some instances, the weighted detection rate is close to one, yet the
adjusted \(\delta\)-difference is negative. This discrepancy implies
that the distance measure, or the estimated distance, may not perfectly
reflect actual human behavior.

\begin{table}

\caption{\label{tab:human-conv-table}Summary of the comparison of decisions made by computer vision model with decisions made by conventional tests and visual tests conducted by human.}
\centering
\begin{tabular}[t]{lrrr}
\toprule
Violations & \#Samples & \#Agreements & Agreement rate\\
\midrule
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Compared with conventional tests}}\\
\hspace{1em}heteroskedasticity & 540 & 464 & 0.8593\\
\hspace{1em}non-linearity & 576 & 459 & 0.7969\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Compared with visual tests conducted by human}}\\
\hspace{1em}heteroskedasticity & 540 & 367 & 0.6796\\
\hspace{1em}non-linearity & 576 & 385 & 0.6684\\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}[!h]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/conv-mosaic-1}

}

\caption{Rejection rate ($p$-value $\leq0.05$) of computer vision models conditional on conventional tests on non-linearity (left) and heteroskedasticity (right) lineups displayed using a mosaic plot. When the conventional test fails to reject, the computer vision mostly fails to reject the same plot as well as indicated by the height of the top right yellow rectangle, but there are non negliable amount of plots where the conventional test rejects but the computer vision model fails to reject as indicated by the width of the top left yellow rectangle.}\label{fig:conv-mosaic}
\end{figure}

\begin{figure}[!h]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/power-1}

}

\caption{Comparison of power of visual tests, conventional tests and the computer vision model. Marks along the x-axis at the bottom of the plot represent rejections made by each type of test. Marks at the top of the plot represent acceptances. Power curves are fitted by logistic regression models with no intercept but an offset equals to $\text{log}(0.05/0.95)$.}\label{fig:power}
\end{figure}

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/pcp-1}

}

\caption{Parallel coordinate plots of decisions made by computer vision model, conventional tests and visual tests made by human.}\label{fig:pcp}
\end{figure}

\begin{figure}[!h]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/delta-1}

}

\caption{A weighted detection rate vs adjusted $\delta$-difference plot. The brown line is smoothing curve produced by fitting generalized additive models.}\label{fig:delta}
\end{figure}

\section{Examples}\label{sec-examples}

In this section, we present the performance of trained computer vision
model on three example datasets. These include the dataset associated
with the residual plot displaying a ``left-triangle'' shape, as
displayed in Figure \ref{fig:false-finding}, along with the Boston
housing dataset \citep{harrison1978hedonic}, and the ``dino'' datasets
from the \texttt{datasauRus} R package \citep{datasaurus}.

The first example illustrates a scenario where both the computer vision
model and human visual inspection successfully avoid rejecting \(H_0\)
when \(H_0\) is true, contrary to conventional tests. This underscores
the necessity of visually examining the residual plot.

In the second example, we encounter a more pronounced violation of the
model, resulting in rejection of \(H_0\) by all three tests. This
highlights the practicality of the computer vision model, particularly
for less intricate tasks.

The third example presents a situation where the model deviation is
non-typical. Here, the computer vision model and human visual inspection
reject \(H_0\), whereas some commonly used conventional tests do not.
This emphasizes the benefits of visual inspection and the unique
advantage of the computer vision model, which, like humans, makes
decisions based on visual discoveries.

\subsection{Left-triangle}\label{left-triangle}

In Section \ref{sec-model-introduction}, we presented an example
residual plot showcased in Figure \ref{fig:false-finding}, illustrating
how humans might misinterpret the ``left-triangle'' shape as indicative
of heteroskedasticity. Additionally, the Breusch-Pagan test yielded a
rejection with a \(p\)-value of 0.046, despite the residuals originating
from a correctly specified model. Figure \ref{fig:false-lineup} offers a
lineup for this fitted model, showcasing various degrees of
``left-triangle'' shape across all residual plots. This phenomenon is
evidently caused by the skewed distribution of the fitted values.
Notably, if the residual plot were evaluated through a visual test, it
would not be rejected since the true residual plot positioned at 10 can
not be distinguished from the others.

Figure \ref{fig:false-check} presents the results of the assessment by
the computer vision model. Notably, the observed visual signal strength
is considerably lower than the 95\% sample quantile of the null
distribution. Moreover, the bootstrapped distribution suggests that it
is highly improbable for the fitted model to be misspecified as the
majority of bootstrapped fitted models will not be rejected. Thus, for
this particular fitted model, both the visual test and the computer
vision model will not reject \(H_0\). However, the Breusch-Pagan test
will reject \(H_0\) because it can not effectively utilize information
from null plots.

The attention map at Figure \ref{fig:false-check}B suggests that the
estimation is highly influenced by the top-right and bottom-right part
of the residual plot, as it forms two vertices of the triangular shape.
A principal component analysis (PCA) is also performed on the output of
the global pooling layer of the computer vision model. As mentioned in
\citet{simonyan2014very}, a computer vision model built upon the
convolutional blocks can be viewed as a feature extractor. For the
\(32 \times 32\) model, there are 256 features outputted from the global
pooling layer, which can be further used for different visual tasks not
limited to distance prediction. To see if these features can be
effectively used for distinguishing null plots and true residual plot,
we linearly project them into the first and second principal components
space as shown in Figure \ref{fig:false-check}D. It can be observed that
because the bootstrapped plots are mostly similar to the null plots, the
points drawn in different colors are mixed together. The true residual
plot is also covered by both the cluster of null plots and cluster of
bootstrapped plots. This accurately reflects our understanding of Figure
\ref{fig:false-lineup}.

\begin{figure}[!h]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/false-check-1}

}

\caption{A summary of the residual plot assessment evaluted on 200 null plots and 200 bootstrapped plots. (A) The true residual plot exhibiting a "left-triangle" shape. (B) The attention map produced by computing the gradient of the output with respect to the greyscale input.  (C) The density plot of estimated distance for null plots and bootstrapped plots. The green area indicates the distribution of estimated distances for bootstrapped plots, while the yellow area represents the distribution of estimated distances for null plots. The fitted model will not be rejected since $\hat{D} < Q_{null}(0.95)$. (D) plot of first two principal components of features extracted from the global pooling layer of the computer vision model.  }\label{fig:false-check}
\end{figure}

\begin{figure}[!h]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/false-lineup-1}

}

\caption{A lineup of residual plots displaying "left-triangle" visual patterns. The true residual plot occupies position 10, yet there are no discernible visual patterns that distinguish it from the other plots.}\label{fig:false-lineup}
\end{figure}

\subsection{Boston Housing}\label{boston-housing}

The Boston housing dataset, originally published by
\citet{harrison1978hedonic}, offers insights into housing in the Boston,
Massachusetts area. For illustration purposes, we utilize a reduced
version from Kaggle, comprising 489 rows and 4 columns: average number
of rooms per dwelling (RM), percentage of lower status of the population
(LSTAT), pupil-teacher ratio by town (PTRATIO), and median value of
owner-occupied homes in \$1000's (MEDV). In our analysis, MEDV will
serve as the response variable, while the other columns will function as
predictors in a linear regression model. Our primary focus is to detect
non-linearity, because the relationships between RM and MEDV or LSTAT
and MEDV are non-linear.

Figure \ref{fig:boston-check} displays the residual plot and the
assessment conducted by the computer vision model. A clear non-linearity
pattern resembling a ``U'' shape is shown in the plot A. Furthermore,
the RESET test yields a very small \(p\)-value. The estimated distance
\(\hat{D}\) significantly exceeds \(Q_{null}(0.95)\), leading to
rejection of \(H_0\). The bootstrapped distribution also suggests that
almost all the bootstrapped fitted models will be rejected, indicating
that the fitted model is unlikely to be correctly specified. The
attention map in plot B suggests the center of the image has higher
leverage than other areas, and it is the turning point of the ``U''
shape. The CPA provided in plot D shows two distinct clusters of data
points, further underling the visual differences between bootstrapped
plots and null plots. This coincides the findings from Figure
\ref{fig:boston-lineup}, where the true plot exhibiting a ``U'' shape is
visually distinctive from null plots. If a visual test is conducted by
human, \(H_0\) will also be rejected.

\begin{figure}[!h]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/boston-check-1}

}

\caption{A summary of the residual plot assessment for the Boston housing fitted model evaluted on 200 null plots and 200 bootstrapped plots. (A) The true residual plot exhibiting a "U" shape. (B) The attention map produced by computing the gradient of the output with respect to the greyscale input.  (C) The density plot of estimated distance for null plots and bootstrapped plots. The blue area indicates the distribution of estimated distances for bootstrapped plots, while the yellow area represents the distribution of estimated distances for null plots. The fitted model will be rejected since $\hat{D} \geq Q_{null}(0.95)$. (D) plot of first two principal components of features extracted from the global pooling layer of the computer vision model. }\label{fig:boston-check}
\end{figure}

\begin{figure}[!h]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/boston-lineup-1}

}

\caption{A lineup of residual plots for the Boston housing fitted model. The true residual plot is at position 7. It can be easily identified as the most different plot.}\label{fig:boston-lineup}
\end{figure}

\subsection{DatasauRus}\label{datasaurus}

The computer vision model possesses the capability to detect not only
typical issues like non-linearity, heteroskedasticity, and non-normality
but also artifact visual patterns resembling real-world objects, as long
as they do not appear in null plots. These visual patterns can be
challenging to categorize in terms of model violations. Therefore, we
will employ the RESET test, the Breusch-Pagan test, and the Shapiro-Wilk
test \citep{shapiro1965analysis} for comparison.

The ``dino'' dataset within the \texttt{datasauRus} R package
exemplifies this scenario. With only two columns, x and y, fitting a
regression model to this data yields a residual plot resembling a
``dinosaur'', as displayed in Figure \ref{fig:dino-check}A.
Unsurprisingly, this distinct residual plot stands out in a lineup, as
shown in Figure \ref{fig:dino-lineup}. A visual test conducted by humans
would undoubtedly reject \(H_0\).

According to the residual plot assessment by the computer vision model,
\(\hat{D}\) exceeds \(Q_{null}(0.95)\), warranting a rejection of
\(H_0\). Additionally, most of the bootstrapped fitted models will be
rejected, indicating an misspecified model. However, both the RESET test
and the Breusch-Pagan test yield \(p\)-values greater than 0.3, leading
to a non-rejection of \(H_0\). Only the Shapiro-Wilk test rejects the
normality assumption with a small \(p\)-value.

More importantly, the attention map in Figure \ref{fig:dino-check}B
clearly exhibits a ``dinosaur'' shape, strongly suggesting the
prediction of the distance is based on human perceptible visual
patterns. The computer vision model is also capable of capturing the
contour or the outline of the embedded shape, just like human being
reading residual plots. The PCA in Figure \ref{fig:dino-check}D also
shows that the cluster of bootstrapped plots is at the corner of the
cluster of null plots.

More importantly, the attention map in Figure \ref{fig:dino-check}B
clearly exhibits a ``dinosaur'' shape, strongly suggesting that the
distance prediction is based on human-perceptible visual patterns. The
computer vision model effectively captures the contour or outline of the
embedded shape, similar to how humans interpret residual plots.
Additionally, the PCA in Figure \ref{fig:dino-check}D demonstrates that
the cluster of bootstrapped plots is positioned at the corner of the
cluster of null plots.

In practice, without accessing the residual plot, it would be
challenging to identify the artificial pattern of the residuals.
Moreover, conducting a normality test for a fitted regression model is
not always standard practice among analysts. Even when performed,
violating the normality assumption is sometimes deemed acceptable,
especially considering the application of quasi-maximum likelihood
estimation in linear regression. This example underscores the importance
of evaluating residual plots and highlights how the proposed computer
vision model can facilitate this process.

\begin{figure}[!h]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/dino-check-1}

}

\caption{A summary of the residual plot assessment for the datasauRus fitted model evaluated on 200 null plots and 200 bootstrapped plots. (A) The residual plot exhibits a "dinosaur" shape. (B) The attention map produced by computing the gradient of the output with respect to the greyscale input.  (C) The density plot of estimated distance for null plots and bootstrapped plots. The blue area indicates the distribution of estimated distances for bootstrapped plots, while the yellow area represents the distribution of estimated distances for null plots. The fitted model will be rejected since $\hat{D} \geq Q_{null}(0.95)$. (D) plot of first two principal components of features extracted from the global pooling layer of the computer vision model.}\label{fig:dino-check}
\end{figure}

\begin{figure}[!h]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/dino-lineup-1}

}

\caption{A lineup of residual plots for the fitted model on the "dinosaur" dataset. The true residual plot is at position 17. It can be easily identified as the most different plot as the visual pattern is extremely artificial.}\label{fig:dino-lineup}
\end{figure}

\section{Limitations and Future Work}\label{limitations-and-future-work}

Despite the computer vision model performing well with general cases
under the synthetic data generation scheme and the three examples used
in this paper, this study has several limitations that could guide
future work.

The proposed distance measure assumes that the true model is a classical
normal linear regression model, which can be restrictive. Although this
paper does not address the relaxation of this assumption, there are
potential methods to evaluate other types of regression models. The most
comprehensive approach would be to define a distance measure for each
different class of regression model and then train the computer vision
model following the methodology described in this paper. To accelerate
training, one could use the convolutional blocks of our trained model as
a feature extractor and perform transfer learning on top of it, as these
blocks effectively capture shapes in residual plots. Another approach
would be to transform the residuals so they are roughly normally
distributed and have constant variance. If only raw residuals are used,
the distance-based statistical testing compares the difference in
distance to a classical normal linear regression model for the true plot
and null plots. This comparison is meaningful only if the difference can
be identified by the distance measure proposed in this paper.

There are other types of residual plots commonly used in diagnostics,
such as residuals vs.~predictor and quantile-quantile plots. In this
study, we focused on the most commonly used residual plot as a starting
point for exploring the new field of automated visual inference.
Similarly, we did not explore other, more sophisticated computer vision
model architectures and specifications for the same reason. While the
performance of the computer vision model is acceptable, there is still
room for improvement to achieve behavior more closely resembling that of
humans interpreting residual plots. This may require external survey
data or human subject experiment data to understand the fundamental
differences between our implementation and human evaluation.

\section{Conclusions}\label{conclusions}

In this paper, we have introduced a distance measure based on
Kullback-Leibler divergence to quantify the disparity between the
residual distribution of a fitted classical normal linear regression
model and the reference residual distribution assumed under correct
model specification. This distance measure effectively captures the
magnitude of model violations in misspecified models. We propose a
computer vision model to estimate this distance, utilizing the residual
plot of the fitted model as input. The resulting estimated distance
serves as the foundation for constructing a single Model Violations
Index (MVI), facilitating the quantification of various model
violations.

Moreover, the estimated distance enables the development of a formal
statistical testing procedure by evaluating a large number of null plots
generated from the fitted model. Additionally, employing bootstrapping
techniques and refitting the regression model allows us to ascertain how
frequently the fitted model is considered misspecified if data were
repeatedly obtained from the same data generating process.

The trained computer vision model demonstrates strong performance on
both the training and test sets, although it exhibits slightly lower
performance on residual plots with non-linearity visual patterns
compared to other types of violations. The statistical tests relying on
the estimated distance predicted by the computer vision model exhibit
lower sensitivity compared to conventional tests but higher sensitivity
compared to visual tests conducted by humans. While the estimated
distance generally mirrors the strength of the visual signal perceived
by humans, there remains scope for further improvement in its
performance.

Several examples are provided to showcase the effectiveness of the
proposed method across different scenarios, emphasizing the similarity
between visual tests and distance-based tests. Overall, both visual
tests and distance-based tests can be viewed as ensemble of tests,
aiming to assess any violations of model assumptions collectively. In
contrast, individual residual diagnostic tests such as the RESET test
and the Breusch-Pagan test only evaluate specific violations of model
assumptions. In practice, selecting an appropriate set of statistical
tests for regression diagnostics can be challenging, particularly given
the necessity of adjusting the significance level for each test.

Our method holds significant value as it helps alleviate a portion of
analysts' workload associated with assessing residual plots. While we
recommend analysts to continue reading residual plots whenever feasible,
as they offer invaluable insights, our approach serves as a valuable
tool for automating the diagnostic process or for supplementary purposes
when needed.

\section*{Acknowledgement}\label{acknowledgement}
\addcontentsline{toc}{section}{Acknowledgement}

These \texttt{R} packages were used for the work: \texttt{tidyverse}
\citep{tidyverse}, \texttt{lmtest} \citep{lmtest}, \texttt{mpoly}
\citep{mpoly}, \texttt{ggmosaic} \citep{ggmosaic}, \texttt{kableExtra}
\citep{kableextra}, \texttt{patchwork} \citep{patchwork},
\texttt{rcartocolor} \citep{rcartocolor}, \texttt{glue} \citep{glue},
\texttt{ggpcp} \citep{ggpcp}, \texttt{here} \citep{here},
\texttt{magick} \citep{magick}, \texttt{yardstick} \citep{yardstick} and
\texttt{reticulate} \citep{reticulate}.

The article was created with R packages \texttt{rticles}
\citep{rticles}, \texttt{knitr} \citep{knitr} and \texttt{rmarkdown}
\citep{rmarkdown}. The project's GitHub repository
(\url{https://github.com/TengMCing/auto_residual_reading_paper})
contains all materials required to reproduce this article.

\bibliographystyle{tfcad}
\bibliography{bibliography.bib}

\end{document}

### 2025-07-28 cover_letter/letter.qmd

---
format: quarto-anu-letter-pdf
author:
  - name: "\u200B" # I need to sign this
    position: Postdoctoral Research Fellow 
    affiliation:
      - name: Australian National University
        department: Research School of Finance, Actuarial Studies and Statistics
        address: "26C Kingsley Street"  
        city: Canberra
        country: Australia
        postal-code: 2600
address:
  - Editors,
  - Journal of Computational and Graphical Statistics
title: Letter from ANU
opening: "Dear Prof. Chen and Prof. Sangalli,"
closing: "Sincerely,"
keep-tex: true
---

Please consider our manuscript "Automated Assessment of Residual Plots with Computer Vision Models" for publication in the Journal of Computational and Graphical Statistics.

This paper describes a new computer vision model that can automatically read residual plots, it's architecture and performance relative to human evaluation and classical hypothesis testing. Our previous work, "A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol" (JCGS, 2024), described interesting findings about human evaluations of residual plots relative to classical hypothesis testing, discovered when building training samples for the computer vision model described in our new paper. The computer vision model described in this new submission was our primary research goal.

We specifically trained a custom model to predict a specified distance measure that quantifies the discrepancy between residuals and theoretically "good" distributions. Building on these predictions, we construct a statistical testing framework aligned with the original lineup protocol, ensuring a valid visual test. Despite substantial experimentation, we could not manage to get a computer vision model to perform quite as well as the human evaluations. It is slightly more sensitive, flagging more residual plots as bad. However, as a computer model, it can process substantially more residual plots than is possible through human evaluation. Relative to statistical tests (Breusch-Pagan, Ramsey RESET, Shapiro-Wilk) it performs substantially better, is less sensitive to small departures, and jointly assesses a variety of departures (heteroscedasticity, non-linearity, non-normality).

The paper also shows that the model, like human observers, relies on visual patterns and shapes, making it a practical tool for reducing manual effort in residual plot diagnostics. This approach would be of particular interest to readers concerned with model diagnostics and the automation of visual data analysis. It represents a new direction for statistical graphics research that engages with new computational tools. There is a shiny app and R package available to support access and use of the computer vision model, and this is documented in a special issue of the Australian and New Zealand Journal of Statistics that celebrates 25 years of R, to appear later this year.

Thank you for the consideration of this manuscript. We believe that it is a good fit for the Journal of Computational and Graphical Statistics.

### 2025-07-28 cover_letter/letter.tex

--- 2025-07-25 cover_letter/template.tex
@@
   pdfauthor={\hspace{0pt}},
-  pdfsubject={Submission of Manuscript: Automated Assessment of Residual
-Plots with Computer Vision Models},
   colorlinks=true,
@@
 \author{\hspace{0pt}}
-\date{July 25, 2025}
+\date{July 28, 2025}
 \begin{document}
@@
 Street\\Canberra 2600\\Australia\\[2mm]}
-\begin{letter}{Editors\\Journal of Computational and Graphical
-Statistics\\ ~ \\Subject: Submission of Manuscript: Automated Assessment
-of Residual Plots with Computer Vision Models}
+\begin{letter}{Editors,\\Journal of Computational and Graphical
+Statistics}
 \opening{Dear Prof.~Chen and Prof.~Sangalli,}
@@
 
-This paper builds on our previous work, ``A Plot is Worth a Thousand
-Tests: Assessing Residual Diagnostics with the Lineup Protocol'' (JCGS),
-which provided empirical evidence for the effectiveness of visual
-evaluation of residual plots through crowd-sourced lineup experiments.
-While that study highlighted the cost and effort of human evaluations,
-this work advances automation of residual plot diagnostics through an
-application of a computer vision model. We specifically trained a model
-to predict a custom distance measure that quantifies the discrepancy
-between residuals and theoretically ``good'' distributions. Building on
-these predictions, we construct a statistical testing framework aligned
-with the original lineup protocol, ensuring a valid visual test. When
-evaluated on the earlier experimenal data, the model performs slightly
-worse than humans but surpasses conventional statistical tests due to
-its reduced sensitivity to minor deviations from model assumptions. We
-further demonstrate that the model, like human observers, relies on
+This paper describes a new computer vision model that can automatically
+read residual plots, it's architecture and performance relative to human
+evaluation and classical hypothesis testing. Our previous work, ``A Plot
+is Worth a Thousand Tests: Assessing Residual Diagnostics with the
+Lineup Protocol'' (JCGS, 2024), described interesting findings about
+human evaluations of residual plots relative to classical hypothesis
+testing, discovered when building training samples for the computer
+vision model described in our new paper. The computer vision model
+described in this new submission was our primary research goal.
+
+We specifically trained a custom model to predict a specified distance
+measure that quantifies the discrepancy between residuals and
+theoretically ``good'' distributions. Building on these predictions, we
+construct a statistical testing framework aligned with the original
+lineup protocol, ensuring a valid visual test. Despite substantial
+experimentation, we could not manage to get a computer vision model to
+perform quite as well as the human evaluations. It is slightly more
+sensitive, flagging more residual plots as bad. However, as a computer
+model, it can process substantially more residual plots than is possible
+through human evaluation. Relative to statistical tests (Breusch-Pagan,
+Ramsey RESET, Shapiro-Wilk) it performs substantially better, is less
+sensitive to small departures, and jointly assesses a variety of
+departures (heteroscedasticity, non-linearity, non-normality).
+
+The paper also shows that the model, like human observers, relies on
 visual patterns and shapes, making it a practical tool for reducing
@@
 particular interest to readers concerned with model diagnostics and the
-automation of visual data analysis.
+automation of visual data analysis. It represents a new direction for
+statistical graphics research that engages with new computational tools.
+There is a shiny app and R package available to support access and use
+of the computer vision model, and this is documented in a special issue
+of the Australian and New Zealand Journal of Statistics that celebrates
+25 years of R, to appear later this year.
 

### 2025-07-28 paper.Rmd

@@
 
+The concept of visual inference, as proposed by @buja2009statistical, provides an inferential framework to assess whether residual plots indeed contain visual patterns inconsistent with the model assumptions.
+The fundamental idea involves testing whether the true residual plot visually differs significantly from null plots, where null plots are plotted with residuals generated from the residual rotation distribution [@langsrud2005rotation], which is a distribution consistent with the null hypothesis $H_0$ that the linear regression model is correctly specified.
+Typically, the visual test is accomplished through the lineup protocol, where the true residual plot is embedded within a lineup alongside several null plots.
+If the true residual plot can be distinguished from the lineup, it provides evidence for rejecting $H_0$.
+
+The practice of delivering a residual plot as a lineup is generally regarded as a valuable approach.
+Beyond its application in residual diagnostics, the lineup protocol has been integrated into the analysis of diverse subjects.
+For instance, Loy and Hofmann [-@loy2013diagnostic; -@loy2014hlmdiag; -@loy2015you] illustrated its applicability in diagnosing hierarchical linear models.
+Additionally, @widen2016graphical and @fieberg2024using demonstrated its utility in geographical and ecology research respectively, while @krishnan2021hierarchical explored its effectiveness in forensic examinations.
+
+A practical limitation of the lineup protocol lies in its reliance on human judgements [see @li2024plot about the practical limitations].
+Unlike conventional statistical tests that can be performed computationally in statistical software, the lineup protocol requires human evaluation of images.
+This characteristic makes it less suitable for large-scale applications, given the associated high labour costs and time requirements.
+There is a substantial need to develop an approach to substitute these human judgements with an automated reading of data plots using machines.
+
+Efforts to automate plot interpretation date back to early work like Tukey’s "Scagnostics" [@tukey1985computer], which is a set of numerical statistics that summarize features of scatter plots.
+@wilkinson2005graph expanded on this work, introducing scagnostics based on computable measures applied to planar proximity graphs.
+These measures, including, but not limited to, "Outlying", "Skinny", "Stringy", "Straight", "Monotonic", "Skewed", "Clumpy", and "Striated", aimed to characterize outliers, shape, density, trend, coherence and other characteristics of the data.
+However, as noted by @buja2009statistical, such measures may not capture the full range of visual patterns necessary for residual diagnostics.
+A more promising direction lies in enabling computers to learn the relevant visual features directly mirroring how humans intuitively interpret plots.
+
+Modern computer vision models are well-suited for addressing this challenge.
+They rely on deep neural networks with convolutional layers [@fukushima1982neocognitron].
+These layers use small, sliding windows to scan the image, performing a dot product to extract local features and patterns.
+Numerous studies have demonstrated the efficacy of convolutional layers in addressing various vision tasks, including image recognition [@rawat2017deep].
+Despite the widespread use of computer vision models in fields like computer-aided diagnosis [@lee2015image], pedestrian detection [@brunetti2018computer], and facial recognition [@emami2012facial], their application in reading data plots remains limited.
+While some studies have explored the use of computer vision models for tasks such as reading recurrence plots for time series regression [@ojeda2020multivariate], time series classification [@chu2019automatic; @hailesilassie2019financial; @hatami2018classification; @zhang2020encoding], anomaly detection [@chen2020convolutional], and pairwise causality analysis [@singh2017deep], the application of reading residual plots with computer vision models is a new field of study.
+
+In this paper, we develop computer vision models and integrate them into the residual plots diagnostics workflow, addressing the need for an automated visual inference.
+The paper is structured as follows.
+Section \ref{sec-model-specifications} discusses various specifications of the computer vision models.
+Section \ref{sec-model-distance-between-residual-plots} defines the distance measure used to detect model violations, while Section \ref{sec-model-distance-estimation} explains how the computer vision models estimate this distance measure.
+Section \ref{sec-model-statistical-testing} covers the statistical tests based on the estimated distance.
+Sections \ref{sec-model-architecture} and \ref{sec-model-data-generation} describe the model architecture and the corresponding data generation and training process, respectively.
+The results are presented in Section \ref{sec-model-results}.
+Example dataset applications are discussed in Section \ref{sec-examples}.
+Finally, we conclude with a discussion of our findings and propose ideas for future research directions.
+
 # Model Specifications {#sec-model-specifications}

### 2025-07-28 paper.tex

@@
 
+The concept of visual inference, as proposed by
+\citet{buja2009statistical}, provides an inferential framework to assess
+whether residual plots indeed contain visual patterns inconsistent with
+the model assumptions. The fundamental idea involves testing whether the
+true residual plot visually differs significantly from null plots, where
+null plots are plotted with residuals generated from the residual
+rotation distribution \citep{langsrud2005rotation}, which is a
+distribution consistent with the null hypothesis \(H_0\) that the linear
+regression model is correctly specified. Typically, the visual test is
+accomplished through the lineup protocol, where the true residual plot
+is embedded within a lineup alongside several null plots. If the true
+residual plot can be distinguished from the lineup, it provides evidence
+for rejecting \(H_0\).
+
+The practice of delivering a residual plot as a lineup is generally
+regarded as a valuable approach. Beyond its application in residual
+diagnostics, the lineup protocol has been integrated into the analysis
+of diverse subjects. For instance, Loy and Hofmann
+\citetext{\citeyear{loy2013diagnostic}; \citeyear{loy2014hlmdiag}; \citeyear{loy2015you}}
+illustrated its applicability in diagnosing hierarchical linear models.
+Additionally, \citet{widen2016graphical} and \citet{fieberg2024using}
+demonstrated its utility in geographical and ecology research
+respectively, while \citet{krishnan2021hierarchical} explored its
+effectiveness in forensic examinations.
+
+A practical limitation of the lineup protocol lies in its reliance on
+human judgements \citep[see][ about the practical
+limitations]{li2024plot}. Unlike conventional statistical tests that can
+be performed computationally in statistical software, the lineup
+protocol requires human evaluation of images. This characteristic makes
+it less suitable for large-scale applications, given the associated high
+labour costs and time requirements. There is a substantial need to
+develop an approach to substitute these human judgements with an
+automated reading of data plots using machines.
+
+Efforts to automate plot interpretation date back to early work like
+Tukey's ``Scagnostics'' \citep{tukey1985computer}, which is a set of
+numerical statistics that summarize features of scatter plots.
+\citet{wilkinson2005graph} expanded on this work, introducing
+scagnostics based on computable measures applied to planar proximity
+graphs. These measures, including, but not limited to, ``Outlying'',
+``Skinny'', ``Stringy'', ``Straight'', ``Monotonic'', ``Skewed'',
+``Clumpy'', and ``Striated'', aimed to characterize outliers, shape,
+density, trend, coherence and other characteristics of the data.
+However, as noted by \citet{buja2009statistical}, such measures may not
+capture the full range of visual patterns necessary for residual
+diagnostics. A more promising direction lies in enabling computers to
+learn the relevant visual features directly mirroring how humans
+intuitively interpret plots.
+
+Modern computer vision models are well-suited for addressing this
+challenge. They rely on deep neural networks with convolutional layers
+\citep{fukushima1982neocognitron}. These layers use small, sliding
+windows to scan the image, performing a dot product to extract local
+features and patterns. Numerous studies have demonstrated the efficacy
+of convolutional layers in addressing various vision tasks, including
+image recognition \citep{rawat2017deep}. Despite the widespread use of
+computer vision models in fields like computer-aided diagnosis
+\citep{lee2015image}, pedestrian detection \citep{brunetti2018computer},
+and facial recognition \citep{emami2012facial}, their application in
+reading data plots remains limited. While some studies have explored the
+use of computer vision models for tasks such as reading recurrence plots
+for time series regression \citep{ojeda2020multivariate}, time series
+classification
+\citep{chu2019automatic, hailesilassie2019financial, hatami2018classification, zhang2020encoding},
+anomaly detection \citep{chen2020convolutional}, and pairwise causality
+analysis \citep{singh2017deep}, the application of reading residual
+plots with computer vision models is a new field of study.
+
+In this paper, we develop computer vision models and integrate them into
+the residual plots diagnostics workflow, addressing the need for an
+automated visual inference. The paper is structured as follows. Section
+\ref{sec-model-specifications} discusses various specifications of the
+computer vision models. Section
+\ref{sec-model-distance-between-residual-plots} defines the distance
+measure used to detect model violations, while Section
+\ref{sec-model-distance-estimation} explains how the computer vision
+models estimate this distance measure. Section
+\ref{sec-model-statistical-testing} covers the statistical tests based
+on the estimated distance. Sections \ref{sec-model-architecture} and
+\ref{sec-model-data-generation} describe the model architecture and the
+corresponding data generation and training process, respectively. The
+results are presented in Section \ref{sec-model-results}. Example
+dataset applications are discussed in Section \ref{sec-examples}.
+Finally, we conclude with a discussion of our findings and propose ideas
+for future research directions.
+
 \section{Model Specifications}\label{sec-model-specifications}
@@
 \(\hat{D}\) above the 95\% quantile of the null distribution
-(\(p-\text{value} = 0.00498\)), leading to rejection of \(H_0\). The
+(\(p\text{-value} = 0.00498\)), leading to rejection of \(H_0\). The
 bootstrapped distribution further supports this, with most fitted models

### 2025-07-28 cover_letter/template.tex
Auto stash before merge of "master" and "origin/master"

@@
-% Options for packages loaded elsewhere
 % Options for packages loaded elsewhere
@@
   letterpaper]{letter}
-\usepackage{xcolor}
-\usepackage[margin=1in,bottom=1in,left=1in,right=1in]{geometry}
+
 \usepackage{amsmath,amssymb}
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
 \usepackage{iftex}
@@
 \else % if luatex or xetex
-  \usepackage{unicode-math} % this also loads fontspec
+  \usepackage{unicode-math}
   \defaultfontfeatures{Scale=MatchLowercase}
@@
 \makeatother
+\usepackage{xcolor}
+\usepackage[margin=1in,bottom=1in,left=1in,right=1in]{geometry}
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
 % Make \paragraph and \subparagraph free-standing
@@
 
-\usepackage{longtable,booktabs,array}
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
 \usepackage{calc} % for calculating minipage widths
@@
 \makeatletter
-\newsavebox\pandoc@box
-\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
-  \sbox\pandoc@box{#1}%
-  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
-  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
-  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
-  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
-  \else\usebox{\pandoc@box}%
-  \fi%
-}
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
 % Set default figure placement to htbp
+\makeatletter
 \def\fps@figure{htbp}
 \makeatother
-
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 
@@
 \makeatother
+
+\ifLuaTeX
+  \usepackage{selnolig}  % disable illegal ligatures
+\fi
 \usepackage{bookmark}
+
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same}
+\urlstyle{same} % disable monospaced font for URLs
 \hypersetup{
@@
 \date{July 25, 2025}
+
 \begin{document}
-\AddToShipoutPicture{\ifthenelse{\value{page}<2}{\includegraphics[width=\paperwidth,height=\paperheight]{\_extensions/anuopensci/quarto-anu-letter/assets/images/coverpage.jpg}}{\includegraphics[width=\paperwidth,height=\paperheight]{\_extensions/anuopensci/quarto-anu-letter/assets/images/page-background.jpg}}}
+\AddToShipoutPicture{\ifthenelse{\value{page}<2}{\includegraphics[width=\paperwidth,height=\paperheight]{\_extensions/quarto-anu-letter/assets/images/coverpage.jpg}}{\includegraphics[width=\paperwidth,height=\paperheight]{\_extensions/quarto-anu-letter/assets/images/page-background.jpg}}}
 
@@
 While that study highlighted the cost and effort of human evaluations,
-this work advances automation of residual plot diagnostics through an
-application of a computer vision model. We specifically trained a model
-to predict a custom distance measure that quantifies the discrepancy
-between residuals and theoretically ``good'' distributions. Building on
-these predictions, we construct a statistical testing framework aligned
-with the original lineup protocol, ensuring a valid visual test. When
-evaluated on the earlier experimenal data, the model performs slightly
-worse than humans but surpasses conventional statistical tests due to
-its reduced sensitivity to minor deviations from model assumptions. We
-further demonstrate that the model, like human observers, relies on
-visual patterns and shapes, making it a practical tool for reducing
-manual effort in residual plot diagnostics. This approach would be of
-particular interest to readers concerned with model diagnostics and the
-automation of visual data analysis.
+this work advances automation through computer vision. We introduce a
+model trained to predict a custom distance measure that quantifies the
+discrepancy between residuals and theoretically ``good'' distributions.
+Building on these predictions, we construct a statistical testing
+framework aligned with the original lineup protocol, ensuring a valid
+visual test. When evaluated on the earlier experiment data, the model
+performs slightly worse than humans but surpasses conventional
+statistical tests due to its reduced sensitivity to minor deviations
+from model assumptions. We further demonstrate that the model, like
+human observers, relies on visual patterns and shapes, making it a
+practical tool for reducing manual effort in residual plot diagnostics.
+This approach should be of particular interest to readers concerned with
+model diagnostics and the automation of visual data analysis.
 

### 2025-07-29 cover_letter/letter.tex
Revert main paper back to commit #4b6fdd, and compile all files

@@
 \author{\hspace{0pt}}
-\date{July 28, 2025}
+\date{July 29, 2025}
+
 \begin{document}
-\AddToShipoutPicture{\ifthenelse{\value{page}<2}{\includegraphics[width=\paperwidth,height=\paperheight]{\_extensions/anuopensci/quarto-anu-letter/assets/images/coverpage.jpg}}{\includegraphics[width=\paperwidth,height=\paperheight]{\_extensions/anuopensci/quarto-anu-letter/assets/images/page-background.jpg}}}
+\AddToShipoutPicture{\ifthenelse{\value{page}<2}{\includegraphics[width=\paperwidth,height=\paperheight]{\_extensions/quarto-anu-letter/assets/images/coverpage.jpg}}{\includegraphics[width=\paperwidth,height=\paperheight]{\_extensions/quarto-anu-letter/assets/images/page-background.jpg}}}
 

### 2025-07-29 paper.Rmd
Add a data availability statement in the main paper

--- 2025-07-25 paper.Rmd
@@
 Yellow boxes are 2D convolutional layers, orange boxes are pooling layers, the grey box is the concatenation layer, and the purple boxes are dense layers.", out.width = "100%"}
-magick::image_read_pdf("figures/cnn.pdf")
+knitr::include_graphics("figures/cnn.pdf")
 ```
@@
   select(res, RMSE, R2, MAE, HUBER) %>%
-  mutate(res = glue::glue("${res} \\times {res}$")) %>%
+  mutate(res = glue("${res} \\times {res}$")) %>%
   mutate(across(RMSE:MAE, ~format(.x, digits = 3))) %>%
@@
 define_gradient <- function() {
-  reticulate::py_run_string(glue::glue(
+  reticulate::py_run_string(glue(
 "
@@
 The cluster also covers the true residual plot.', fig.height = 8, out.width = "80%"}
-p1 <- my_vi$plot_resid(theme = theme_light()) + ggtitle(glue::glue("(A) Residual plot")) 
+p1 <- my_vi$plot_resid(theme = theme_light()) + ggtitle(glue("(A) Residual plot")) 
 p2 <- readRDS(here::here("cached_data/fig1_attention_map.rds")) + ggtitle("(B) Attention map")
 p3 <- my_vi$summary_plot() +
-  annotate("text", x = 2.8, y = 1.25, label = glue::glue("p-value = {format(my_vi$p_value(), digits = 3)}")) +
-  ggtitle("(C) Density plot of estimated D", subtitle = glue::glue("Conventional p-value = {format(HETER_MODEL$test(dat)$p_value, digits = 3)}")) +
+  annotate("text", x = 2.8, y = 1.25, label = glue("p-value = {format(my_vi$p_value(), digits = 3)}")) +
+  ggtitle("(C) Density plot of estimated D", subtitle = glue("Conventional p-value = {format(HETER_MODEL$test(dat)$p_value, digits = 3)}")) +
   theme(legend.position = "bottom") +
@@
   scale_color_brewer(palette = "Dark2") +
-  xlab(glue::glue("PC1 ({scales::percent(pc1_per)})")) +
-  ylab(glue::glue("PC2 ({scales::percent(pc2_per)})")) +
+  xlab(glue("PC1 ({scales::percent(pc1_per)})")) +
+  ylab(glue("PC2 ({scales::percent(pc2_per)})")) +
   labs(col = "") +
@@
 (D) Bootstrapped and null plots form distinct clusters in the space of the first two principal components from the global pooling layer, highlighting their visual separability.', fig.height = 8, out.width = "80%"}
-p1 <- my_vi$plot_resid(theme = theme_light()) + ggtitle(glue::glue("(A) Residual plot")) 
+p1 <- my_vi$plot_resid(theme = theme_light()) + ggtitle(glue("(A) Residual plot")) 
 p2 <- readRDS(here::here("cached_data/boston_attention_map.rds")) + ggtitle("(B) Attention map")
 p3 <- my_vi$summary_plot() +
-    annotate("text", x = 6, y = 2.25, label = glue::glue("p-value = {format(my_vi$p_value(), digits = 3)}")) +
-  ggtitle("(C) Density plot of estimated D", subtitle = glue::glue("Conventional p-value = {format(lmtest::resettest(mod)$p.value, digits = 3)}")) +
+    annotate("text", x = 6, y = 2.25, label = glue("p-value = {format(my_vi$p_value(), digits = 3)}")) +
+  ggtitle("(C) Density plot of estimated D", subtitle = glue("Conventional p-value = {format(lmtest::resettest(mod)$p.value, digits = 3)}")) +
   theme(legend.position = "bottom") +
@@
 
-p1 <- my_vi$plot_resid(theme = theme_light()) + ggtitle(glue::glue("(A) Residual plot")) 
+p1 <- my_vi$plot_resid(theme = theme_light()) + ggtitle(glue("(A) Residual plot")) 
 p2 <- readRDS(here::here("cached_data/dino_attention_map.rds")) + ggtitle("(B) Attention map")
 p3 <- my_vi$summary_plot() +
-    annotate("text", x = 6, y = 1.25, label = glue::glue("p-value = {format(my_vi$p_value(), digits = 3)}")) +
-  ggtitle("(C) Density plot of estimated D", subtitle = glue::glue("RESET test p-value = {format(lmtest::resettest(mod)$p.value, digits = 3)}\n Breusch-Pagan test p-value = {format(lmtest::bptest(mod)$p.value, digits = 3)} \n Shapiro-Wilk test p-value = {format(shapiro.test(mod$residuals)$p.value, digits = 3)}")) +
+    annotate("text", x = 6, y = 1.25, label = glue("p-value = {format(my_vi$p_value(), digits = 3)}")) +
+  ggtitle("(C) Density plot of estimated D", subtitle = glue("RESET test p-value = {format(lmtest::resettest(mod)$p.value, digits = 3)}\n Breusch-Pagan test p-value = {format(lmtest::bptest(mod)$p.value, digits = 3)} \n Shapiro-Wilk test p-value = {format(shapiro.test(mod$residuals)$p.value, digits = 3)}")) +
   theme(legend.position = "bottom") +
@@
   scale_color_brewer(palette = "Dark2") +
-  xlab(glue::glue("PC1 ({scales::percent(pc1_per)})")) +
-  ylab(glue::glue("PC2 ({scales::percent(pc2_per)})")) +
+  xlab(glue("PC1 ({scales::percent(pc1_per)})")) +
+  ylab(glue("PC2 ({scales::percent(pc2_per)})")) +
   labs(col = "", shape = "") +
@@
 
-These `R` packages were used for the work: `tidyverse` [@tidyverse], `lmtest` [@lmtest], `mpoly` [@mpoly], `ggmosaic` [@ggmosaic], `kableExtra` [@kableextra], `patchwork` [@patchwork], `rcartocolor` [@rcartocolor], `glue` [@glue], `ggpcp` [@ggpcp], `here` [@here], `magick` [@magick], and `reticulate` [@reticulate].
+These `R` packages were used for the work: `tidyverse` [@tidyverse], `lmtest` [@lmtest], `mpoly` [@mpoly], `ggmosaic` [@ggmosaic], `kableExtra` [@kableextra], `patchwork` [@patchwork], `glue` [@glue], `ggpcp` [@ggpcp], `here` [@here], and `reticulate` [@reticulate].
 
@@
 
+# Data Availability Statement {-}
+
+The data that support the findings of this study are openly available in the project's GitHub repository (https://github.com/TengMCing/auto_residual_reading_paper).
+
 # Disclosure Statement {-}

### 2025-07-29 paper.tex
Revert main paper back to commit #4b6fdd, and compile all files
Recompile

--- 2025-07-11 paper.tex
@@
 
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/cnn-diag-1}
+{\centering \includegraphics[width=1\linewidth]{figures/cnn}
 
@@
 \citep{mpoly}, \texttt{ggmosaic} \citep{ggmosaic}, \texttt{kableExtra}
-\citep{kableextra}, \texttt{patchwork} \citep{patchwork},
-\texttt{rcartocolor} \citep{rcartocolor}, \texttt{glue} \citep{glue},
-\texttt{ggpcp} \citep{ggpcp}, \texttt{here} \citep{here},
-\texttt{magick} \citep{magick}, and \texttt{reticulate}
-\citep{reticulate}.
+\citep{kableextra}, \texttt{patchwork} \citep{patchwork}, \texttt{glue}
+\citep{glue}, \texttt{ggpcp} \citep{ggpcp}, \texttt{here} \citep{here},
+and \texttt{reticulate} \citep{reticulate}.
 
@@
 
+\section*{Data Availability
+Statement}\label{data-availability-statement}
+\addcontentsline{toc}{section}{Data Availability Statement}
+
+The data that support the findings of this study are openly available in
+the project's GitHub repository
+(\url{https://github.com/TengMCing/auto_residual_reading_paper}).
+
 \section*{Disclosure Statement}\label{disclosure-statement}

### 2025-07-29 cover_letter/letter.qmd
Add a conflict of interest statement and a original work statement in the cover letter
Add co-author info and sign

@@
 opening: "Dear Prof. Chen and Prof. Sangalli,"
-closing: "Sincerely,"
+closing: ""
 keep-tex: true
@@
 Please consider our manuscript "Automated Assessment of Residual Plots with Computer Vision Models" for publication in the Journal of Computational and Graphical Statistics.
+
+We confirm that this work is original and has not been published elsewhere, nor is it currently under consideration for publication elsewhere.
 
@@
 
+We have no conflict of interest to disclose.
+
+Please address all corresponding concerning this manuscript to me at patrick.li@anu.edu.au.
+
 Thank you for the consideration of this manuscript. We believe that it is a good fit for the Journal of Computational and Graphical Statistics.
+
+Sincerely,
+
+\bigskip
+\bigskip
+\bigskip
+\bigskip
+\bigskip
+\bigskip
+
+Dr. Weihao Li\newline
+Postdoctoral Research Fellow, Research School of Finance, Actuarial Studies and Statistics\newline
+Australian National University
+
+Co-author
+
+Prof. Dianne Cook\newline
+Professor, Department of Econometrics and Business Statistics\newline
+Monash University
+
+Dr. Emi Tanaka\newline
+Senior Lecturer, Research School of Finance, Actuarial Studies and Statistics\newline
+Australian National University
+
+Dr. Susan VanderPlas\newline
+Associate Professor, Department of Statistics\newline
+University of Nebraska
+
+Dr. Klaus Ackermann\newline
+Associate Professor, Department of Econometrics and Business Statistics\newline
+Monash University

### 2025-07-29 cover_letter/letter.tex
Recompile
Add co-author info and sign

@@
 
+We confirm that this work is original and has not been published
+elsewhere, nor is it currently under consideration for publication
+elsewhere.
+
 This paper describes a new computer vision model that can automatically
@@
 
+We have no conflict of interest to disclose.
+
+Please address all corresponding concerning this manuscript to me at
+patrick.li@anu.edu.au.
+
 Thank you for the consideration of this manuscript. We believe that it
@@
 
-\closing{Sincerely,}
+Sincerely,
+
+\bigskip
+\bigskip
+\bigskip
+\bigskip
+\bigskip
+\bigskip
+
+Dr.~Weihao Li\newline Postdoctoral Research Fellow, Research School of
+Finance, Actuarial Studies and Statistics\newline Australian National
+University
+
+Co-author
+
+Prof.~Dianne Cook\newline Professor, Department of Econometrics and
+Business Statistics\newline Monash University
+
+Dr.~Emi Tanaka\newline Senior Lecturer, Research School of Finance,
+Actuarial Studies and Statistics\newline Australian National University
+
+Dr.~Susan VanderPlas\newline Associate Professor, Department of
+Statistics\newline University of Nebraska
+
+Dr.~Klaus Ackermann\newline Associate Professor, Department of
+Econometrics and Business Statistics\newline Monash University
+
 \vfill

### 2025-07-29 cover_letter/letter.qmd
Include a statement in cover letter stating that we want to do single-anonymous

@@
 
+We understand that the Journal of Computational and Graphical Statistics now uses a single-anonymous peer review system by default, with an option for authors to opt into double-anonymous review.
+We intend to proceed with the single-anonymous review, and thus our manuscript includes full author information.
+Since the submission system requires uploading two PDF files, one with author details and one anonymized, please note that the anonymized version is identical to the version containing author information.
+
 Please address all corresponding concerning this manuscript to me at patrick.li@anu.edu.au.
